---
title: "00 - Onboarding: dataset metadata"
subtitle: "Fetch the pinned per-deposit metadata and standardise it to one per-sample schema for the pipeline"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    df-print: paged
execute:
  warning: false
  message: false
---

This notebook onboards the deposits **rescued from the discard hunt** — candidates
that MASST flagged and that survived the confirmation filters — plus the two Jarmusch
CYP-probe accessions. Each metadata source is fetched from a **pinned** location (a
tagged GitHub commit, a Pan-ReDU export, or a MetaboLights ISA-Tab), standardised to a
single per-sample schema, and written to `artifacts/` for `01`–`05` to consume.

Nothing here is committed to the repo: the raw per-deposit metadata under
`dataset_metadata/` is git-ignored (regenerated by this notebook + `R/metadata.R`), and
the outputs live in the ignored `artifacts/` tree. Run this **first**, before `01`.

**Location semantics.** `participant_origin` = location describes where participants are
*from* (only MSV000082433); `study_site` = location is where the study was run, with
participant origin/ethnicity **not** verified — treat as a study-site covariate.

```{r setup}
suppressPackageStartupMessages(library(dplyr))
MD <- "dataset_metadata"
dir.create("artifacts", showWarnings = FALSE)
miss <- function(x) { x[x %in% c("", "missing value", "not applicable", "not collected")] <- NA; x }
```

## Dataset-level constants

Fields the metadata sources do not carry (chromatography, polarity, acquisition,
instrument, provenance) come from each study's Methods, pinned here so `01` can join them
by `deposit_id`.

```{r constants}
constants <- tibble::tribble(
  ~deposit_id,      ~repository, ~chromatography, ~polarity, ~acquisition, ~instrument,        ~location,               ~location_type,        ~health_stratifier,   ~cyp2c19_ground_truth, ~paper,          ~metadata_source,
  "MSV000082493",   "MassIVE",   "RP (C18)",      "pos",     "DDA",        "maXis Impact QTOF","USA:CA:San Diego",      "study_site",          "cyp2c19_phenotype",  TRUE,                  "PMC7485946",    "GitHub alan-jarmusch/UntargetedMS-DrugMetabolism@95586620",
  "MSV000084008",   "MassIVE",   "RP (C18)",      "pos",     "DDA",        "maXis Impact QTOF","USA:CA:San Diego",      "study_site",          "cyp2c19_phenotype",  TRUE,                  "PMC7485946",    "GitHub alan-jarmusch/UntargetedMS-DrugMetabolism@95586620 (same study as 082493)",
  "MSV000086975",   "MassIVE",   "RP",            "pos",     "DDA",        "Q Exactive",       "USA (UCSD)",            "study_site",          "health_status",      FALSE,                 "PMID 34741130", "deposit metadata/Metabolomics_Metadata_Massive.tsv",
  "MSV000085256",   "MassIVE",   "RP (C18)",      "pos",     "DDA",        "Q Exactive",       "USA",                   "study_site",          "health_status",      FALSE,                 "PMID 34244695", "Pan-ReDU redu.gnps2.org",
  "MTBLS12231",     "MetaboLights","RP",          "pos+neg", "DDA",        "Q Exactive",       "Japan",                 "study_site",          "health_status",      FALSE,                 "PMC12019401",   "ISA-Tab s_/a_",
  "MSV000082433",   "MassIVE",   "RP (C18)",      "pos",     "DDA",        "Q Exactive",       "Venezuela",             "participant_origin",  "none (all healthy)", FALSE,                 NA,              "Pan-ReDU redu.gnps2.org"
)
write.csv(constants, "artifacts/rescue_dataset_constants.csv", row.names = FALSE)
constants
```

## Per-sample standardiser

Every deposit is mapped to one schema. `timepoint` = `ATTRIBUTE_Study_Day`;
`timepoint_mins` = minutes post-dose within that day (the serial PK draws);
`pharmkinetics_cyp2c19` = the source study's own published per-subject CYP2C19 PK value,
the ground truth `05` validates against. Every field comes from a declared metadata
attribute — nothing is parsed out of filenames.

```{r std-schema}
std_cols <- c("deposit_id","file","subject_id","sex","age","health_status",
              "specimen","timepoint","timepoint_mins","pharmkinetics_cyp2c19",
              "cyp2c19_phenotype","cyp2c19_genotype","include")
num_cols <- c("age", "timepoint_mins", "pharmkinetics_cyp2c19")
blank <- function(n) setNames(as.data.frame(matrix(NA_character_, n, length(std_cols))), std_cols)
```

### Jarmusch CYP cohort — MSV000082493 (urine/feces/skin) + MSV000084008 (plasma)

The two accessions share one metadata table; plasma rows are onboarded under 084008,
the rest under 082493.

```{r jarmusch}
jm <- read.delim(file.path(MD, "MSV000082493_metadata.tsv"), check.names = FALSE,
                 stringsAsFactors = FALSE, na.strings = c("", "not applicable"))
jm_std <- data.frame(
  file = jm$filename, subject_id = jm$ATTRIBUTE_Subject, sex = jm$ATTRIBUTE_sex,
  age = suppressWarnings(as.numeric(jm$age)), health_status = "healthy",
  specimen = dplyr::case_when(grepl("plasma", jm$ATTRIBUTE_Body_Site, TRUE) ~ "plasma",
                              grepl("^UA", jm$ATTRIBUTE_Body_Site) ~ "urine",
                              grepl("fecal", jm$ATTRIBUTE_Body_Site, TRUE) ~ "feces",
                              grepl("forehead|skin", jm$ATTRIBUTE_Body_Site, TRUE) ~ "skin",
                              TRUE ~ NA_character_),
  timepoint = jm$ATTRIBUTE_Study_Day,
  timepoint_mins = suppressWarnings(as.numeric(jm$ATTRIBUTE_Time_Point_Mins)),
  pharmkinetics_cyp2c19 = suppressWarnings(as.numeric(jm$PharmKinetics_CYP2C19)),
  cyp2c19_phenotype = jm$PharmGeno_Phenotype_2C19,
  cyp2c19_genotype = jm$PharmGeno_CYP2C19, include = jm$Include_Exclude,
  stringsAsFactors = FALSE)
jm_std <- jm_std[jm_std$include %in% "include" & !is.na(jm_std$specimen), ]
d84008 <- cbind(deposit_id = "MSV000084008", jm_std[jm_std$specimen == "plasma", std_cols[-1]])
d82493 <- cbind(deposit_id = "MSV000082493", jm_std[jm_std$specimen != "plasma", std_cols[-1]])
```

### MSV000086975 — schizophrenia (MassIVE submitter TSV)

```{r sz}
sz <- read.delim(file.path(MD, "MSV000086975_metadata.tsv"), check.names = FALSE, stringsAsFactors = FALSE)
sid <- strsplit(sz$`#SampleID`, "_", fixed = TRUE)
d86975 <- blank(nrow(sz)); d86975$deposit_id <- "MSV000086975"
d86975$file <- sz$Metabolomics_FileName; d86975$subject_id <- sz$`#SampleID`
d86975$sex <- vapply(sid, function(p) if (length(p) >= 2) p[2] else NA_character_, character(1))
d86975$age <- suppressWarnings(as.numeric(sz$AGE))
d86975$health_status <- sz$Metadata          # HC / SZ
d86975$specimen <- "plasma"; d86975$include <- "include"
```

### MSV000085256 — Alzheimer's (Pan-ReDU JSON)

```{r az}
pj <- function(f) {
  txt <- paste(readLines(file.path(MD, f), warn = FALSE), collapse = "")
  objs <- regmatches(txt, gregexpr("\\{[^{}]*\\}", txt))[[1]]
  g <- function(o, k) { m <- regmatches(o, regexpr(sprintf('"%s": "[^"]*"', k), o))
    if (length(m)) sub('.*: "', "", sub('"$', "", m)) else NA_character_ }
  data.frame(file = miss(vapply(objs, g, character(1), "filename")),
             subject_id = miss(vapply(objs, g, character(1), "SubjectIdentifierAsRecorded")),
             sex = miss(vapply(objs, g, character(1), "BiologicalSex")),
             age = suppressWarnings(as.numeric(miss(vapply(objs, g, character(1), "AgeInYears")))),
             health_status = miss(vapply(objs, g, character(1), "HealthStatus")),
             specimen = miss(vapply(objs, g, character(1), "UBERONBodyPartName")),
             row.names = NULL, stringsAsFactors = FALSE) }
a <- pj("MSV000085256_panredu.json"); d85256 <- blank(nrow(a)); d85256$deposit_id <- "MSV000085256"
d85256[names(a)] <- a; d85256$specimen <- ifelse(grepl("plasma", d85256$specimen, TRUE), "plasma",
                                          ifelse(grepl("cerebro|CSF", d85256$specimen, TRUE), "CSF", d85256$specimen))
d85256$include <- "include"
```

### MSV000082433 — Amerindians (Pan-ReDU JSON)

```{r amerindians}
b <- pj("MSV000082433_panredu.json"); d82433 <- blank(nrow(b)); d82433$deposit_id <- "MSV000082433"
d82433[names(b)] <- b; d82433$include <- "include"
```

### MTBLS12231 — COVID (MetaboLights ISA-Tab, `s_` joined to `a_`)

ISA-Tab has duplicate column names; the first occurrence of each is selected by index.

```{r covid}
col1 <- function(d, nm) d[[ which(names(d) == nm)[1] ]]
s <- read.delim(file.path(MD, "s_MTBLS12231.txt"), check.names = FALSE, stringsAsFactors = FALSE)
read_a <- function(p) { d <- read.delim(
    file.path(MD, sprintf("a_MTBLS12231_LC-MS_%s_reverse-phase_metabolite_profiling.txt", p)),
    check.names = FALSE, stringsAsFactors = FALSE)
  # the raw column is empty here; pick whichever *Spectral Data File column is populated
  fcols <- grep("Spectral Data File", names(d))
  fcol  <- fcols[which.max(vapply(fcols, function(i)
             sum(grepl("\\.(mzML|mzXML|raw|d|wiff)$", d[[i]])), integer(1)))]
  data.frame(sample = col1(d, "Sample Name"), raw = basename(d[[fcol]]),
             stringsAsFactors = FALSE) }
amap <- rbind(read_a("positive"), read_a("negative"))
sref <- data.frame(sample = col1(s, "Sample Name"),
                   genetype = col1(s, "Factor Value[genetype]"),
                   part = col1(s, "Characteristics[Organism part]"), stringsAsFactors = FALSE)
mm <- dplyr::left_join(amap, sref, by = "sample")
d12231 <- blank(nrow(mm)); d12231$deposit_id <- "MTBLS12231"
d12231$file <- basename(mm$raw); d12231$subject_id <- mm$sample
d12231$health_status <- mm$genetype; d12231$specimen <- mm$part
d12231$include <- "include"
```

### Combine and write `rescue_samples.csv`

```{r combine}
frames <- lapply(list(d82493, d84008, d86975, d85256, d82433, d12231), function(f) {
  f$timepoint <- as.character(f$timepoint)
  for (cc in setdiff(std_cols, num_cols)) f[[cc]] <- as.character(f[[cc]])
  for (cc in num_cols) f[[cc]] <- suppressWarnings(as.numeric(f[[cc]]))
  f[, std_cols]
})
all_std <- dplyr::bind_rows(frames)
write.csv(all_std, "artifacts/rescue_samples.csv", row.names = FALSE)

cat("Rescue samples per deposit x specimen:\n"); print(table(all_std$deposit_id, all_std$specimen, useNA = "ifany"))
cat("\nHealth stratifier availability per deposit:\n")
for (d in unique(all_std$deposit_id)) {
  x <- all_std[all_std$deposit_id == d, ]
  cat(sprintf("  %-14s n=%-4d sex=%-3s age=%-3s health={%s}\n", d, nrow(x),
      ifelse(any(!is.na(x$sex)),"Y","-"), ifelse(any(!is.na(x$age)),"Y","-"),
      paste(sort(unique(na.omit(x$health_status))), collapse="/")))
}
```

## MSV000082493 — pinned fetch + detailed onboarding

**Study.** Jarmusch et al. 2020, *Clin Transl Sci*, doi:10.1111/cts.12785 (PMC7485946).
Healthy volunteers given a modified Cooperstown CYP-probe cocktail (omeprazole 40 mg =
the CYP2C19 probe) on study day 1 and day 9, with a 7-day cefprozil course in between;
plasma/urine/feces, untargeted LC-MS/MS. It carries the **genotyped CYP2C19 metabolizer
phenotype** (Poor..Ultrarapid) per subject — the ground truth the omeprazole/5-OH ratio
should predict — so it is the validation cohort, not just extra data.

**Provenance (pinned).** The authors' GitHub repo at commit `95586620`,
`Materials/Metadata.txt`. Cached at `dataset_metadata/MSV000082493_metadata.tsv`;
re-downloaded from the pinned URL if absent, so a fresh clone reproduces it.

```{r fetch-82493}
SRC_COMMIT <- "95586620ead4ac5241dc1e99a98e047c1f99b17b"
SRC_URL <- sprintf(paste0("https://raw.githubusercontent.com/alan-jarmusch/",
                          "UntargetedMS-DrugMetabolism/%s/Materials/Metadata.txt"),
                   SRC_COMMIT)
LOCAL <- "dataset_metadata/MSV000082493_metadata.tsv"
if (!file.exists(LOCAL))
  utils::download.file(SRC_URL, LOCAL, quiet = TRUE, mode = "wb")

m <- read.delim(LOCAL, check.names = FALSE, stringsAsFactors = FALSE,
                na.strings = c("", "not applicable"))
# Map the raw columns down to the fields the pipeline needs, one row per file.
samples <- data.frame(
  file              = m$filename,
  subject_id        = m$ATTRIBUTE_Subject,
  sex               = m$ATTRIBUTE_sex,
  age               = suppressWarnings(as.numeric(m$age)),
  body_site         = m$ATTRIBUTE_Body_Site,           # blood_plasma / UA* (urine) / fecal / L.forehead (skin)
  study_day         = m$ATTRIBUTE_Study_Day,           # 1 = pre-cefprozil, 9 = post-cefprozil
  antibiotic_exposure = m$ATTRIBUTE_antibiotic_exposure,
  omeprazole_dosed  = m$omeprazole,
  cyp2c19_phenotype = m$PharmGeno_Phenotype_2C19,      # ground truth: Poor..Ultrarapid
  cyp2c19_pk        = m$PharmKinetics_CYP2C19,         # measured PK activity
  sample_type       = m$MS_SampleType,
  include           = m$Include_Exclude,
  stringsAsFactors  = FALSE)

# Biological samples only (drop blanks / QC / excluded).
bio <- samples[samples$include %in% "include" & !is.na(samples$body_site), ]
bio$specimen <- dplyr::case_when(
  grepl("plasma", bio$body_site, ignore.case = TRUE) ~ "plasma",
  grepl("^UA",    bio$body_site)                      ~ "urine",
  grepl("fecal",  bio$body_site, ignore.case = TRUE)  ~ "feces",
  grepl("forehead|skin", bio$body_site, ignore.case = TRUE) ~ "skin",
  TRUE ~ tolower(bio$body_site))
write.csv(bio, "artifacts/MSV000082493_samples.csv", row.names = FALSE)

cat("MSV000082493:", nrow(bio), "biological samples (of", nrow(samples), "files)\n")
cat("\nspecimen x study_day:\n");        print(table(bio$specimen, bio$study_day, useNA = "ifany"))
cat("\nCYP2C19 genotyped phenotype:\n"); print(table(bio$cyp2c19_phenotype, useNA = "ifany"))
cat("\nsubjects:", length(unique(bio$subject_id)),
    " sex:", paste(names(table(bio$sex)), table(bio$sex), collapse = " "), "\n")
```

## MSV000084008 — derive the plasma-only metadata

MSV000084008 has no Pan-ReDU record and its hit files are 100% blood plasma, so its
metadata is written as the plasma subset of the shared 082493 table. Files actually
hosted under 082493 simply fail to sync under 084008 and are skipped (no double count).

```{r derive-84008}
is_plasma <- grepl("plasma", m$ATTRIBUTE_Body_Site, ignore.case = TRUE)
plasma <- m[is_plasma, , drop = FALSE]
write.table(plasma, "dataset_metadata/MSV000084008_metadata.tsv",
            sep = "\t", quote = FALSE, row.names = FALSE)
cat("MSV000084008_metadata.tsv <- plasma rows:", nrow(plasma), "of", nrow(m), "\n")
cat("body sites kept:", paste(sort(unique(plasma$ATTRIBUTE_Body_Site)), collapse = ", "), "\n")
```
