---
title: "01 - MASST curation"
subtitle: "Compound pair · MASST exports · dataset selection"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
params:
  run_confirmations: false
execute:
  warning: false
  message: false
---

# Overview

First step of the pipeline: load the StructureMASST hit set for omeprazole
and its CYP2C19 metabolite (5-hydroxyomeprazole), keep datasets where both
compounds appear, and curate that intersection down to the (dataset, assay)
combinations suitable for a parent/metabolite ratio analysis.

A *dataset* means **one assay of one deposit**; every downstream step joins
on `dataset_id`, which encodes the deposit + assay pair.

The search:

- **Query spectra** are the GNPS reference-library MS/MS spectra of omeprazole
  and 5-hydroxyomeprazole, in both polarities. The discovery chunk below
  resolves each compound to its library spectra by InChIKey first block.
- **Search space** is every public MS/MS spectrum indexed by StructureMASST
  across MassIVE, MetaboLights and Metabolomics Workbench.
- **Match threshold** is cosine ≥ 0.7 and ≥ 3 matched peaks, analog search off.

Hits are read from **FASSTrecords** (a DOI-versioned SQLite of repository-scale
MASST annotations, DOI 10.5281/zenodo.18199544), so the discovery step is
scripted and pinned rather than a one-off web query.

**Reads:**

- `StructureMASST_output/fasstrecords_hits.csv` (written by the discovery chunk
  below from FASSTrecords; committed, so normally read as-is)

**Writes:**

- `artifacts/hits_curated.csv` - MASST hits restricted to curated keeps,
  with `dataset_id`, `deposit_id`, `assay` and `adduct_inferred` columns
  populated. Read by the three SIRIUS confirmation notebooks.
- `artifacts/dataset_metadata_table.csv` - one row per kept (deposit,
  assay) with repository, chromatography, paper-derived constants and
  the col_/const_ mappings consumed by `load_metadata()`.

```{r setup}
source("R/_setup.R")
source("R/masst.R")
source("R/metadata.R")
source("check-versions.R")
```

# Compound pair

The omeprazole / 5-hydroxyomeprazole pair is a CYP2C19 phenotyping readout:
CYP2C19 hydroxylates omeprazole at the 5-position, so the parent/5-OH ratio
reflects CYP2C19 metaboliser status.

Both compounds were queried in both polarities; downstream code picks the
adduct with the most confirmed hits per dataset and collapses adducts to one
feature per compound per sample.

```{r compound-spec}
compound_spec
```

# MASST hits (FASSTrecords)

**Provenance and reproduction.** FASSTrecords is a repository-scale snapshot of
public MASST annotations (a ~1.2-billion-row SQLite of every GNPS/MASST spectrum
match), archived on Zenodo at **DOI 10.5281/zenodo.18199544**
(sha256 `95a11bbe…`). We downloaded that archive and extracted the hit set on
**2026-07-10**. To reproduce, download the Zenodo archive to the `FASST_DB` path in
the discovery chunk below and delete `StructureMASST_output/fasstrecords_hits.csv`;
the chunk then resolves each compound to its GNPS library spectra by InChIKey first
block (omeprazole `SUBDBMMJDZJVOS`, 5-OH `CMZHQFXXAAIBKE`), pulls every `masst_table`
match at cosine ≥ 0.7 and ≥ 3 matched peaks, joins the per-file ReDU fields, and
writes `StructureMASST_output/fasstrecords_hits.csv`.

Because the archive is immutable (DOI + checksum), this discovery step is fully
pinned. The chunk **skips itself** when the ~1.2B-row DB is absent or the CSV
already exists, so a fresh clone (and every normal re-render) uses the committed
output unchanged. Each row of the hit table carries:

- `ATTRIBUTE_DatasetAccession`, `file_path`, `scan_id`, `role` (parent /
  metabolite): where the hit is;
- `Precursor_MZ` and the inferred `Adduct`: what was matched;
- the per-file ReDU fields the hard stops key on: `ChromatographyAndPhase`,
  `IonizationSourceAndPolarity`, `NCBITaxonomy`, `SampleType`.

```{r discovery-fasstrecords}
# Resolve the probe pair to its GNPS library spectra and pull every StructureMASST match
# from the FASSTrecords DB. That ~1.2B-row SQLite (Zenodo DOI 10.5281/zenodo.18199544,
# immutable) is not in the repo, so this runs only where the DB is present and the output is
# missing; otherwise the committed CSV is used unchanged. Delete the CSV to force a rebuild.
FASST_DB  <- "D:/FASSTrecords/db/masst_records.sqlite"
FASST_OUT <- "StructureMASST_output/fasstrecords_hits.csv"
if (file.exists(FASST_DB) && !file.exists(FASST_OUT)) {
  suppressPackageStartupMessages(library(RSQLite))
  IKEY <- c(parent = "SUBDBMMJDZJVOS", metabolite = "CMZHQFXXAAIBKE")  # omeprazole / 5-OH
  COSINE_MIN <- 0.7; PEAKS_MIN <- 3
  INVALID_SCAN <- 4294967295                                            # 32-bit uint sentinel
  con <- dbConnect(SQLite(), FASST_DB)
  unq <- function(x) trimws(gsub('"', '', x))          # strip stray quotes in Dataset/mri
  inl <- function(v) paste(v, collapse = ",")          # integer IN-list
  lib_ids <- lapply(IKEY, function(k) dbGetQuery(con, sprintf(
    "SELECT spectrum_id_int, spectrum_id, Adduct, Precursor_MZ
       FROM library_table WHERE InChIKey_smiles_firstBlock = '%s'", k)))
  get_role <- function(role) {
    lib <- lib_ids[[role]]; if (!nrow(lib)) return(NULL)
    h <- dbGetQuery(con, sprintf(
      "SELECT m.cosine AS Cosine, m.matching_peaks AS matching_peaks, m.scan_id AS scan_id,
              m.spectrum_id_int AS spectrum_id_int, m.mri_id_int AS mri_id_int,
              d.Dataset AS ATTRIBUTE_DatasetAccession, mr.mri AS mri
         FROM masst_table m
         JOIN dataset_table d ON m.dataset_id_int = d.dataset_id_int
         JOIN mri_table    mr ON m.mri_id_int     = mr.mri_id_int
        WHERE m.spectrum_id_int IN (%s) AND m.cosine >= %f AND m.matching_peaks >= %d",
      inl(lib$spectrum_id_int), COSINE_MIN, PEAKS_MIN))
    if (!nrow(h)) return(NULL)
    h <- dplyr::left_join(h, lib[, c("spectrum_id_int","spectrum_id","Adduct","Precursor_MZ")],
                          by = "spectrum_id_int")
    h$ATTRIBUTE_DatasetAccession <- unq(h$ATTRIBUTE_DatasetAccession); h$mri <- unq(h$mri)
    h$role <- role; h$inchikey_first_block <- IKEY[[role]]
    h$USI <- paste0(h$mri, ":scan:", h$scan_id)          # resolvable spectrum reference
    h$file_path <- sub("^mzspec:[^:]*:", "", h$mri)       # repo-relative path; basename used downstream
    h$`Matching Peaks` <- h$matching_peaks; h$lib_spectrum <- h$spectrum_id
    h }
  hits <- dplyr::bind_rows(get_role("parent"), get_role("metabolite"))
  hits <- hits[hits$scan_id != INVALID_SCAN | is.na(hits$scan_id), ]   # drop 32-bit sentinel scans
  mids <- unique(hits$mri_id_int); chunks <- split(mids, ceiling(seq_along(mids) / 5000))
  redu <- dplyr::bind_rows(lapply(chunks, function(ch) dbGetQuery(con, sprintf(
    "SELECT mri_id_int, ChromatographyAndPhase, IonizationSourceAndPolarity, NCBITaxonomy, SampleType
       FROM redu_table WHERE mri_id_int IN (%s)", inl(ch)))))
  cln <- function(v) { v[is.na(v) | v %in% c("missing value","not applicable","")] <- NA; v }
  for (cc in c("ChromatographyAndPhase","IonizationSourceAndPolarity","NCBITaxonomy","SampleType"))
    redu[[cc]] <- cln(redu[[cc]])
  redu <- redu[!duplicated(redu$mri_id_int), ]
  hits <- dplyr::left_join(hits, redu, by = "mri_id_int")
  dbDisconnect(con)
  dir.create("StructureMASST_output", showWarnings = FALSE)
  keep <- c("ATTRIBUTE_DatasetAccession","USI","file_path","scan_id","role","Cosine","Matching Peaks",
            "Adduct","Precursor_MZ","inchikey_first_block","lib_spectrum","ChromatographyAndPhase",
            "IonizationSourceAndPolarity","NCBITaxonomy","SampleType")
  write.csv(hits[, keep], FASST_OUT, row.names = FALSE)
  message("FASSTrecords discovery: wrote ", nrow(hits), " hits -> ", FASST_OUT)
} else {
  message("FASSTrecords discovery: using committed ", FASST_OUT,
          if (!file.exists(FASST_DB)) " (DB not present)" else " (already built)")
}
```

```{r load-hits}
all_hits <- read.csv("StructureMASST_output/fasstrecords_hits.csv",
                     check.names = FALSE, stringsAsFactors = FALSE)
# Normalise ReDU / MIxS placeholders ("missing value", …) to NA.
for (j in seq_along(all_hits))
  if (is.character(all_hits[[j]])) all_hits[[j]][is_missing_val(all_hits[[j]])] <- NA
all_hits$scan_id <- as.integer(all_hits$scan_id)
```

```{r load-hits-summary}
#| echo: false
cat("Total MASST hits loaded:", nrow(all_hits), "\n")
cat("  parent    :", sum(all_hits$role == "parent"), "\n")
cat("  metabolite:", sum(all_hits$role == "metabolite"), "\n")
```

## Remove corrupted USIs

The scan number `4294967295` (max 32-bit unsigned int) is written by some
mzML converters when the scan index is unresolved; such hits can't locate a
spectrum, so we drop them.

```{r filter-corrupt-usi}
INVALID_SCAN <- 4294967295   # 32-bit uint sentinel; > .Machine$integer.max, so keep numeric
n_corrupt <- sum(all_hits$scan_id == INVALID_SCAN, na.rm = TRUE)
all_hits  <- all_hits[is.na(all_hits$scan_id) | all_hits$scan_id != INVALID_SCAN, ]
```

```{r filter-corrupt-usi-summary}
#| echo: false
cat("Corrupted USIs removed:", n_corrupt,
    "(", round(100 * n_corrupt / (nrow(all_hits) + n_corrupt), 1),
    "% of all hits)\n")
```

# Dataset selection

## D_both: datasets where both compounds appear

We restrict to the intersection of datasets containing at least one MASST
hit for each compound. Datasets with only one compound cannot provide a
parent/metabolite ratio and are out of scope.

```{r d-both}
D_parent     <- unique(all_hits$ATTRIBUTE_DatasetAccession[all_hits$role == "parent"])
D_metabolite <- unique(all_hits$ATTRIBUTE_DatasetAccession[all_hits$role == "metabolite"])
D_both       <- intersect(D_parent, D_metabolite)
```

```{r d-both-summary}
#| echo: false
cat("parent datasets:", length(D_parent),
    "| metabolite datasets:", length(D_metabolite),
    "| intersection:", length(D_both), "\n")
```

# Per-deposit metadata

```{r metadata-setup}
# Carried over with this section: the pinned metadata directory and the placeholder
# normaliser the per-deposit blocks below rely on.
MD <- "dataset_metadata"
miss <- function(x) {
  x[x %in% c("", "missing value", "not applicable", "not collected")] <- NA
  x
}
```

The search above leaves the deposits that carry both compounds. Before any of them
can be curated further, their metadata has to be retrieved and put on a common
schema: which files are biological samples, which subject each belongs to, and what
was recorded about that subject. Each repository serves this differently, so the
material below fetches from the pinned per-deposit sources and standardises them to
one per-sample table, written to `artifacts/rescue_samples.csv`.

### Dataset-level constants

Fields the metadata sources do not carry (chromatography, polarity, acquisition,
instrument, provenance) come from each study's Methods, pinned here so `01` can join them
by `deposit_id`.

```{r constants}
constants <- tibble::tribble(
  ~deposit_id,      ~repository, ~chromatography, ~polarity, ~acquisition, ~instrument,        ~location,               ~location_type,        ~health_stratifier,   ~cyp2c19_ground_truth, ~paper,          ~metadata_source,
  "MSV000082493",   "MassIVE",   "RP (C18)",      "pos",     "DDA",        "maXis Impact QTOF","USA:CA:San Diego",      "study_site",          "cyp2c19_phenotype",  TRUE,                  "PMC7485946",    "GitHub alan-jarmusch/UntargetedMS-DrugMetabolism@95586620",
  "MSV000084008",   "MassIVE",   "RP (C18)",      "pos",     "DDA",        "maXis Impact QTOF","USA:CA:San Diego",      "study_site",          "cyp2c19_phenotype",  TRUE,                  "PMC7485946",    "GitHub alan-jarmusch/UntargetedMS-DrugMetabolism@95586620 (same study as 082493)",
  "MSV000086975",   "MassIVE",   "RP",            "pos",     "DDA",        "Q Exactive",       "USA (UCSD)",            "study_site",          "health_status",      FALSE,                 "PMID 34741130", "deposit metadata/Metabolomics_Metadata_Massive.tsv",
  "MSV000085256",   "MassIVE",   "RP (C18)",      "pos",     "DDA",        "Q Exactive",       "USA",                   "study_site",          "health_status",      FALSE,                 "PMID 34244695", "Pan-ReDU redu.gnps2.org",
  "MTBLS12231",     "MetaboLights","RP",          "pos+neg", "DDA",        "Q Exactive",       "Japan",                 "study_site",          "health_status",      FALSE,                 "PMC12019401",   "ISA-Tab s_/a_",
  "MSV000082433",   "MassIVE",   "RP (C18)",      "pos",     "DDA",        "Q Exactive",       "Venezuela",             "participant_origin",  "none (all healthy)", FALSE,                 NA,              "Pan-ReDU redu.gnps2.org"
)
write.csv(constants, "artifacts/rescue_dataset_constants.csv", row.names = FALSE)
constants
```

### Per-sample standardiser

Every deposit is mapped to one schema. `timepoint` = `ATTRIBUTE_Study_Day`;
`timepoint_mins` = minutes post-dose within that day (the serial PK draws);
`pharmkinetics_cyp2c19` = the source study's own published per-subject CYP2C19 PK value,
the ground truth `04` validates against. Every field comes from a declared metadata
attribute; nothing is parsed out of filenames.

```{r std-schema}
std_cols <- c("deposit_id","file","subject_id","sex","age","health_status",
              "specimen","timepoint","timepoint_mins","pharmkinetics_cyp2c19",
              "cyp2c19_phenotype","cyp2c19_genotype","include")
num_cols <- c("age", "timepoint_mins", "pharmkinetics_cyp2c19")
blank <- function(n) setNames(as.data.frame(matrix(NA_character_, n, length(std_cols))), std_cols)
```

#### Jarmusch CYP cohort: MSV000082493 (urine/feces/skin) + MSV000084008 (plasma)

The two accessions share one metadata table; plasma rows are onboarded under 084008,
the rest under 082493.

```{r jarmusch}
jm <- read.delim(file.path(MD, "MSV000082493_metadata.tsv"), check.names = FALSE,
                 stringsAsFactors = FALSE, na.strings = c("", "not applicable"))
jm_std <- data.frame(
  file = jm$filename, subject_id = jm$ATTRIBUTE_Subject, sex = jm$ATTRIBUTE_sex,
  age = suppressWarnings(as.numeric(jm$age)), health_status = "healthy",
  specimen = dplyr::case_when(grepl("plasma", jm$ATTRIBUTE_Body_Site, TRUE) ~ "plasma",
                              grepl("^UA", jm$ATTRIBUTE_Body_Site) ~ "urine",
                              grepl("fecal", jm$ATTRIBUTE_Body_Site, TRUE) ~ "feces",
                              grepl("forehead|skin", jm$ATTRIBUTE_Body_Site, TRUE) ~ "skin",
                              TRUE ~ NA_character_),
  timepoint = jm$ATTRIBUTE_Study_Day,
  timepoint_mins = suppressWarnings(as.numeric(jm$ATTRIBUTE_Time_Point_Mins)),
  pharmkinetics_cyp2c19 = suppressWarnings(as.numeric(jm$PharmKinetics_CYP2C19)),
  cyp2c19_phenotype = jm$PharmGeno_Phenotype_2C19,
  cyp2c19_genotype = jm$PharmGeno_CYP2C19, include = jm$Include_Exclude,
  stringsAsFactors = FALSE)
jm_std <- jm_std[jm_std$include %in% "include" & !is.na(jm_std$specimen), ]
d84008 <- cbind(deposit_id = "MSV000084008", jm_std[jm_std$specimen == "plasma", std_cols[-1]])
d82493 <- cbind(deposit_id = "MSV000082493", jm_std[jm_std$specimen != "plasma", std_cols[-1]])
```

#### MSV000086975: schizophrenia (MassIVE submitter TSV)

```{r sz}
sz <- read.delim(file.path(MD, "MSV000086975_metadata.tsv"), check.names = FALSE, stringsAsFactors = FALSE)
sid <- strsplit(sz$`#SampleID`, "_", fixed = TRUE)
d86975 <- blank(nrow(sz)); d86975$deposit_id <- "MSV000086975"
d86975$file <- sz$Metabolomics_FileName; d86975$subject_id <- sz$`#SampleID`
d86975$sex <- vapply(sid, function(p) if (length(p) >= 2) p[2] else NA_character_, character(1))
d86975$age <- suppressWarnings(as.numeric(sz$AGE))
d86975$health_status <- sz$Metadata          # HC / SZ
d86975$specimen <- "plasma"; d86975$include <- "include"
```

#### MSV000085256: Alzheimer's (Pan-ReDU JSON)

```{r az}
pj <- function(f) {
  txt <- paste(readLines(file.path(MD, f), warn = FALSE), collapse = "")
  objs <- regmatches(txt, gregexpr("\\{[^{}]*\\}", txt))[[1]]
  g <- function(o, k) { m <- regmatches(o, regexpr(sprintf('"%s": "[^"]*"', k), o))
    if (length(m)) sub('.*: "', "", sub('"$', "", m)) else NA_character_ }
  data.frame(file = miss(vapply(objs, g, character(1), "filename")),
             subject_id = miss(vapply(objs, g, character(1), "SubjectIdentifierAsRecorded")),
             sex = miss(vapply(objs, g, character(1), "BiologicalSex")),
             age = suppressWarnings(as.numeric(miss(vapply(objs, g, character(1), "AgeInYears")))),
             health_status = miss(vapply(objs, g, character(1), "HealthStatus")),
             specimen = miss(vapply(objs, g, character(1), "UBERONBodyPartName")),
             row.names = NULL, stringsAsFactors = FALSE) }
a <- pj("MSV000085256_panredu.json"); d85256 <- blank(nrow(a)); d85256$deposit_id <- "MSV000085256"
d85256[names(a)] <- a; d85256$specimen <- ifelse(grepl("plasma", d85256$specimen, TRUE), "plasma",
                                          ifelse(grepl("cerebro|CSF", d85256$specimen, TRUE), "CSF", d85256$specimen))
d85256$include <- "include"
```

#### MSV000082433: Amerindians (Pan-ReDU JSON)

```{r amerindians}
b <- pj("MSV000082433_panredu.json"); d82433 <- blank(nrow(b)); d82433$deposit_id <- "MSV000082433"
d82433[names(b)] <- b; d82433$include <- "include"
```

#### MTBLS12231: COVID (MetaboLights ISA-Tab, `s_` joined to `a_`)

ISA-Tab has duplicate column names; the first occurrence of each is selected by index.

```{r covid}
col1 <- function(d, nm) d[[ which(names(d) == nm)[1] ]]
s <- read.delim(file.path(MD, "s_MTBLS12231.txt"), check.names = FALSE, stringsAsFactors = FALSE)
read_a <- function(p) { d <- read.delim(
    file.path(MD, sprintf("a_MTBLS12231_LC-MS_%s_reverse-phase_metabolite_profiling.txt", p)),
    check.names = FALSE, stringsAsFactors = FALSE)
  # the raw column is empty here; pick whichever *Spectral Data File column is populated
  fcols <- grep("Spectral Data File", names(d))
  fcol  <- fcols[which.max(vapply(fcols, function(i)
             sum(grepl("\\.(mzML|mzXML|raw|d|wiff)$", d[[i]])), integer(1)))]
  data.frame(sample = col1(d, "Sample Name"), raw = basename(d[[fcol]]),
             stringsAsFactors = FALSE) }
amap <- rbind(read_a("positive"), read_a("negative"))
sref <- data.frame(sample = col1(s, "Sample Name"),
                   genetype = col1(s, "Factor Value[genetype]"),
                   part = col1(s, "Characteristics[Organism part]"), stringsAsFactors = FALSE)
mm <- dplyr::left_join(amap, sref, by = "sample")
d12231 <- blank(nrow(mm)); d12231$deposit_id <- "MTBLS12231"
d12231$file <- basename(mm$raw); d12231$subject_id <- mm$sample
d12231$health_status <- mm$genetype; d12231$specimen <- mm$part
d12231$include <- "include"
```

#### Combine and write `rescue_samples.csv`

```{r combine}
frames <- lapply(list(d82493, d84008, d86975, d85256, d82433, d12231), function(f) {
  f$timepoint <- as.character(f$timepoint)
  for (cc in setdiff(std_cols, num_cols)) f[[cc]] <- as.character(f[[cc]])
  for (cc in num_cols) f[[cc]] <- suppressWarnings(as.numeric(f[[cc]]))
  f[, std_cols]
})
all_std <- dplyr::bind_rows(frames)
write.csv(all_std, "artifacts/rescue_samples.csv", row.names = FALSE)

cat("Rescue samples per deposit x specimen:\n"); print(table(all_std$deposit_id, all_std$specimen, useNA = "ifany"))
cat("\nHealth stratifier availability per deposit:\n")
for (d in unique(all_std$deposit_id)) {
  x <- all_std[all_std$deposit_id == d, ]
  cat(sprintf("  %-14s n=%-4d sex=%-3s age=%-3s health={%s}\n", d, nrow(x),
      ifelse(any(!is.na(x$sex)),"Y","-"), ifelse(any(!is.na(x$age)),"Y","-"),
      paste(sort(unique(na.omit(x$health_status))), collapse="/")))
}
```

### MSV000082493: pinned fetch + detailed onboarding

**Study.** Jarmusch et al. 2020, *Clin Transl Sci*, doi:10.1111/cts.12785 (PMC7485946).
Healthy volunteers given a modified Cooperstown CYP-probe cocktail (omeprazole 40 mg =
the CYP2C19 probe) on study day 1 and day 9, with a 7-day cefprozil course in between;
plasma/urine/feces, untargeted LC-MS/MS. It carries the **genotyped CYP2C19 metabolizer
phenotype** (Poor..Ultrarapid) per subject, the ground truth the omeprazole/5-OH ratio
should predict, so it is the validation cohort rather than just extra data.

**Provenance (pinned).** The authors' GitHub repo at commit `95586620`,
`Materials/Metadata.txt`. Cached at `dataset_metadata/MSV000082493_metadata.tsv`;
re-downloaded from the pinned URL if absent, so a fresh clone reproduces it.

```{r fetch-82493}
SRC_COMMIT <- "95586620ead4ac5241dc1e99a98e047c1f99b17b"
SRC_URL <- sprintf(paste0("https://raw.githubusercontent.com/alan-jarmusch/",
                          "UntargetedMS-DrugMetabolism/%s/Materials/Metadata.txt"),
                   SRC_COMMIT)
LOCAL <- "dataset_metadata/MSV000082493_metadata.tsv"
if (!file.exists(LOCAL))
  utils::download.file(SRC_URL, LOCAL, quiet = TRUE, mode = "wb")

m <- read.delim(LOCAL, check.names = FALSE, stringsAsFactors = FALSE,
                na.strings = c("", "not applicable"))
# Map the raw columns down to the fields the pipeline needs, one row per file.
samples <- data.frame(
  file              = m$filename,
  subject_id        = m$ATTRIBUTE_Subject,
  sex               = m$ATTRIBUTE_sex,
  age               = suppressWarnings(as.numeric(m$age)),
  body_site         = m$ATTRIBUTE_Body_Site,           # blood_plasma / UA* (urine) / fecal / L.forehead (skin)
  study_day         = m$ATTRIBUTE_Study_Day,           # 1 = pre-cefprozil, 9 = post-cefprozil
  antibiotic_exposure = m$ATTRIBUTE_antibiotic_exposure,
  omeprazole_dosed  = m$omeprazole,
  cyp2c19_phenotype = m$PharmGeno_Phenotype_2C19,      # ground truth: Poor..Ultrarapid
  cyp2c19_pk        = m$PharmKinetics_CYP2C19,         # measured PK activity
  sample_type       = m$MS_SampleType,
  include           = m$Include_Exclude,
  stringsAsFactors  = FALSE)

# Biological samples only (drop blanks / QC / excluded).
bio <- samples[samples$include %in% "include" & !is.na(samples$body_site), ]
bio$specimen <- dplyr::case_when(
  grepl("plasma", bio$body_site, ignore.case = TRUE) ~ "plasma",
  grepl("^UA",    bio$body_site)                      ~ "urine",
  grepl("fecal",  bio$body_site, ignore.case = TRUE)  ~ "feces",
  grepl("forehead|skin", bio$body_site, ignore.case = TRUE) ~ "skin",
  TRUE ~ tolower(bio$body_site))
write.csv(bio, "artifacts/MSV000082493_samples.csv", row.names = FALSE)

cat("MSV000082493:", nrow(bio), "biological samples (of", nrow(samples), "files)\n")
cat("\nspecimen x study_day:\n");        print(table(bio$specimen, bio$study_day, useNA = "ifany"))
cat("\nCYP2C19 genotyped phenotype:\n"); print(table(bio$cyp2c19_phenotype, useNA = "ifany"))
cat("\nsubjects:", length(unique(bio$subject_id)),
    " sex:", paste(names(table(bio$sex)), table(bio$sex), collapse = " "), "\n")
```

### MSV000084008: derive the plasma-only metadata

MSV000084008 has no Pan-ReDU record and its hit files are 100% blood plasma, so its
metadata is written as the plasma subset of the shared 082493 table. Files actually
hosted under 082493 simply fail to sync under 084008 and are skipped (no double count).

```{r derive-84008}
is_plasma <- grepl("plasma", m$ATTRIBUTE_Body_Site, ignore.case = TRUE)
plasma <- m[is_plasma, , drop = FALSE]
write.table(plasma, "dataset_metadata/MSV000084008_metadata.tsv",
            sep = "\t", quote = FALSE, row.names = FALSE)
cat("MSV000084008_metadata.tsv <- plasma rows:", nrow(plasma), "of", nrow(m), "\n")
cat("body sites kept:", paste(sort(unique(plasma$ATTRIBUTE_Body_Site)), collapse = ", "), "\n")
```

## Hard-stop exclusions

Reusable safety filter, derived from columns in the MASST export itself (no
network calls). A dataset is dropped only if **all** its hits fail the rule; one
LC hit in a mixed GC/LC deposit keeps it.

On *this* corpus these rules remove **nothing**, and that is expected: a
FASSTrecords hit set is already restricted to the GNPS ecosystem
(`MSV`/`MTBLS`/`ST`) and to LC/ESI acquisition. The rules earn their place when
the pipeline is pointed at a different, less pre-filtered query. Non-human
deposits whose taxonomy is missing from the export are caught in the manual
exclusions below instead.

| rule | source column | reason |
|---|---|---|
| repository prefix | `ATTRIBUTE_DatasetAccession` | The pipeline only supports direct raw-file download from MetaboLights (`MTBLS*`), Metabolomics Workbench (`ST*`) and MassIVE (`MSV*`). |
| GC-only chromatography | `ChromatographyAndPhase` | The pipeline targets LC-MS for omeprazole; GC ionisation profiles and retention behaviour are incompatible. |
| imaging MS (MALDI / DESI / SIMS) | `IonizationSourceAndPolarity` | No LC dimension, no per-file retention time to match against. |
| non-human taxa | `NCBITaxonomy` | Optional. The ratio analysis is a CYP2C19 phenotyping readout in humans. Flip `drop_non_human <- FALSE` to disable. |

```{r hard-stops}
#| code-fold: true
drop_non_human <- TRUE

hits_by_ds <- split(all_hits, all_hits$ATTRIBUTE_DatasetAccession)

all_match <- function(d, col, pattern) {
  v <- hits_by_ds[[d]][[col]]
  v <- v[!is.na(v) & v != "missing value"]
  length(v) > 0 && all(grepl(pattern, v, perl = TRUE, ignore.case = TRUE))
}

fail_prefix  <- D_both[!grepl("^(MSV|MTBLS|ST)[0-9]", D_both)]
fail_gc      <- D_both[vapply(D_both, all_match, logical(1),
                              "ChromatographyAndPhase", "gas chromatography|^GC")]
fail_imaging <- D_both[vapply(D_both, all_match, logical(1),
                              "IonizationSourceAndPolarity",
                              "MALDI|DESI|SIMS|imaging")]
fail_taxa    <- if (drop_non_human) {
  bad <- vapply(D_both, function(d) {
    tx <- hits_by_ds[[d]]$NCBITaxonomy
    tx <- tx[!is.na(tx) & tx != "missing value"]
    length(tx) > 0 && !any(grepl("Homo sapiens|\\b9606\\b", tx))
  }, logical(1))
  D_both[bad]
} else character(0)

hard_stops <- list(
  repo_prefix = fail_prefix,
  gc_only     = fail_gc,
  imaging_ms  = fail_imaging,
  non_human   = fail_taxa
)
D_both <- setdiff(D_both, unlist(hard_stops, use.names = FALSE))
```

```{r hard-stops-summary}
#| echo: false
cat("Hard-stop exclusions:\n")
for (nm in names(hard_stops)) {
  n <- length(hard_stops[[nm]])
  cat(sprintf("  %-12s: %2d", nm, n))
  if (n) cat("  ->  ", paste(hard_stops[[nm]], collapse = ", "), sep = "")
  cat("\n")
}
cat("\nSurviving:", length(D_both), "datasets\n")
```

## Other manual exclusions

Datasets that survive the hard stops but still have to go for repository- or
study-specific reasons. The `reason` vocabulary:

| reason | what it means |
|---|---|
| `no_or_limited_metadata` | no usable biological metadata anywhere (deposit + Pan-ReDU + publication), or only a single value for every stratifier |
| `access_restricted` | demographics gated behind a controlled-access portal (e.g. NCRAD); only body_site / country recoverable |
| `dia` | DIA acquisition; single-scan MASST matching is unreliable |
| `in_vitro` | cell culture or reference-library deposit; MASST hits are spike artefacts, not biological exposure |
| `non_human` | non-human taxon. Caught here rather than by the taxonomy hard stop because `redu_table` carries no `NCBITaxonomy` for these deposits, so the hard stop cannot fire |
| `reference_material` | QA/QC reference-material suite (e.g. NIST RM), not a phenotyped study cohort |

```{r other-exclusions}
other_exclusions <- tibble::tribble(
  ~dataset_id,    ~reason,                  ~notes,
  "MSV000085870", "no_or_limited_metadata", "DDA acquisition confirmed but no Pan-ReDU or deposit metadata",
  "MSV000093072", "no_or_limited_metadata", "",
  "MSV000095726", "no_or_limited_metadata", "",
  "MSV000096359", "no_or_limited_metadata", "",
  "MSV000096884", "no_or_limited_metadata", "",
  "MSV000097110", "no_or_limited_metadata", "",
  "MSV000094515", "no_or_limited_metadata", "stool / Alzheimer's (U19 ADRC NCRAD); demographics NCRAD-controlled",
  "ST002238",     "no_or_limited_metadata", "no metadata file or linked publication found",
  "MSV000094395", "no_or_limited_metadata", "only DOIDCommonName=Alzheimer's, no other strata in MASST or Pan-ReDU",
  "MSV000095143", "no_or_limited_metadata", "submitter TSV documents 264 files with rich metadata but those filenames don't exist on MassIVE; the 1077 actually-hosted files (Pan-ReDU naming, used by MASST + SIRIUS) have all canonical fields as 'missing value' in every source. Deposit cannot be stratified.",
  "MSV000097935", "access_restricted",      "demographics access-restricted ('Partial Public' MassIVE deposit, Rancho Bernardo Study policy)",
  "MTBLS417",     "dia",                    "DIA; duplicate of a DIA MassIVE deposit, poor metadata",
  "MTBLS2746",    "in_vitro",               "drug-treatment cell culture; MASST hits are spike artefacts",
  "ST001715",     "in_vitro",               "in vitro xenobiotic reference library",
  "MSV000082382", "non_human",              "Pseudomonas / Bacillus on melon plants",
  "MSV000086015", "non_human",              "San Diego Zoo rhinoceros serum (Diceros / Ceratotherium)",
  "MSV000087018", "reference_material",     "NIST RM 8231 frozen-plasma reference-material QA/QC suite, not a study cohort",
  "MSV000090544", "reference_material",     "GNPS Matrix-tubes extraction-method validation on high-biomass fecal+saliva (DOI 10.25345/C5QJ7831B); plate/well layout, no per-sample metadata, not a dosed cohort; omeprazole is incidental",
  "MSV000083065", "no_or_limited_metadata", "NOBEL twin plasma cohort; no Pan-ReDU, no accessible per-sample strata",
  "MSV000083593", "no_or_limited_metadata", "S. aureus infection plasma; no accessible per-sample metadata",
  "MSV000083955", "no_or_limited_metadata", "cerebral cavernous malformation plasma; no per-sample metadata",
  "MSV000084716", "no_or_limited_metadata", "Undiagnosed Disease Network; paper (PMID 33883556) confirms no medication/CYP data, part GC-MS, not in Pan-ReDU",
  "MSV000085508", "no_or_limited_metadata", "UDN, same cohort as MSV000084716 (duplicate), no metadata",
  "MSV000086270", "no_or_limited_metadata", "U19 ADNI-Go 1000-plasma cohort; no accessible per-sample strata",
  "MSV000088645", "no_or_limited_metadata", "U19 T1000 serum; no accessible per-sample strata",
  "MSV000090580", "no_or_limited_metadata", "osteoarthritis plasma (Guma); no accessible per-sample metadata",
  "MSV000092926", "no_or_limited_metadata", "plasma cohort (Knight ADC); no accessible per-sample metadata",
  "MSV000093037", "no_or_limited_metadata", "unidentified deposit: no title/summary/keywords/ReDU",
  "MSV000095383", "no_or_limited_metadata", "U19 Wisconsin serum (Alzheimer's); no accessible per-sample strata"
)

D_both <- setdiff(D_both, other_exclusions$dataset_id)
```

```{r other-exclusions-summary}
#| echo: false
cat("Other manual exclusions:", nrow(other_exclusions), "datasets\n")
print(table(other_exclusions$reason))
cat("\nSurviving after manual exclusions:", length(D_both), "datasets\n")
```

## Auto-derived dataset metadata

For each surviving deposit we harvest **technical** metadata (per-assay:
chromatography, ionisation, instrument) and **biological** metadata
(per-deposit: which native columns hold subject_id, sex, age). Sources and
cache locations are repository-specific:

- **MetaboLights (`MTBLS*`)**: technical via
  `mtbls_assay_data()` (the assay `a_*.txt`
  files); biological via `mtbls_sample_data()` (the `s_*.txt` sample
  file). Both auto-cached from EBI FTP at render time and stored as
  `dataset_metadata/{a_,s_}*.txt`.
- **Metabolomics Workbench (`ST*`)**: technical via
  `mwb_metadata()$MS_run`; biological
  via `mwb_metadata()$sample_annotation`. Both auto-cached from the
  Workbench API into `dataset_metadata/{ds}_AN*.txt`.
- **MassIVE (`MSV*`)**: technical from the submitter deposit TSV/CSV
  (`dataset_metadata/{ds}_metadata.*`) first, with Pan-ReDU
  (`fetch_panredu_metadata()` from `redu.gnps2.org`, cached as
  `dataset_metadata/{ds}_panredu.tsv`) as per-field fallback.
  Biological metadata comes from the same two sources, in the same
  order.

ReDU / MIxS missing-value placeholders (`"missing value"`,
`"not applicable"`, …) are normalised to NA everywhere via the shared
`is_missing_val()` helper in `R/metadata.R`.

```{r auto-assess}
#| code-fold: true
tech_raw <- fetch_assay_metadata(D_both)
coverage <- tibble::as_tibble(
  dplyr::bind_rows(lapply(D_both, harvest_metadata_coverage))
)
write.csv(tech_raw, "artifacts/assay_metadata.csv", row.names = FALSE)
```

```{r auto-assess-display}
#| echo: false
coverage[, c("dataset_id", "n_samples", "n_canonical_mapped",
             grep("^col_", names(coverage), value = TRUE))]
```

## Per-assay refinement (MTBLS / ST)

MetaboLights and Workbench deposits often package several assays under one
accession (MTBLS1866: LC-MS pos, LC-MS neg, GCxGC-MS pos; ST002044: four
HILIC/RP analyses). Since the ratio needs **both** compounds in the same
assay, we treat each `(deposit, assay)` pair as a separate entity and keep
only entities with at least one parent AND one metabolite hit. A deposit
whose every assay fails is appended to `other_exclusions` with reason
`no_assay_with_both_roles`.

MassIVE deposits have no assay concept and stay at deposit granularity.

```{r assay-assignment}
mtbls_st_keeps <- D_both[grepl("^(MTBLS|ST)", D_both)]
assay_lookups  <- setNames(lapply(mtbls_st_keeps, assay_lookup_for),
                           mtbls_st_keeps)

all_hits$assay <- mapply(function(ds, fp) {
  if (startsWith(ds, "MSV")) return(NA_character_)
  file_to_assay(basename(fp), assay_lookups[[ds]])
}, all_hits$ATTRIBUTE_DatasetAccession, all_hits$file_path,
USE.NAMES = FALSE)

all_hits$deposit_id <- all_hits$ATTRIBUTE_DatasetAccession
all_hits$dataset_id <- make_dataset_id(all_hits$deposit_id, all_hits$assay)

# MASST hits whose file basename doesn't resolve to any known assay
mtbls_st_hits <- all_hits$deposit_id %in% mtbls_st_keeps
unresolved    <- mtbls_st_hits & is.na(all_hits$assay)
all_hits$dataset_id[unresolved] <- paste0(all_hits$deposit_id[unresolved],
                                          "_unmatched")
```

```{r entity-coverage}
entity_coverage <- all_hits[all_hits$deposit_id %in% D_both, ] |>
  dplyr::group_by(dataset_id, deposit_id, assay) |>
  dplyr::summarise(has_parent = any(role == "parent"),
                   has_metab  = any(role == "metabolite"),
                   n_hits     = dplyr::n(),
                   .groups    = "drop") |>
  dplyr::mutate(keep = has_parent & has_metab &
                       !grepl("_unmatched$", dataset_id))

datasets_with_kept_entity <- unique(
  entity_coverage$deposit_id[entity_coverage$keep])
deposits_lost <- setdiff(D_both, datasets_with_kept_entity)

if (length(deposits_lost)) {
  new_excl <- tibble::tibble(
    dataset_id = deposits_lost,
    reason     = "no_assay_with_both_roles",
    notes      = "parent + metabolite MASST hits land in different assays of the same deposit")
  other_exclusions <- dplyr::bind_rows(other_exclusions, new_excl)
  D_both <- setdiff(D_both, deposits_lost)
}

entities_kept <- entity_coverage$dataset_id[entity_coverage$keep]
all_hits      <- all_hits[all_hits$dataset_id %in% entities_kept, ]
```

```{r entity-coverage-summary}
#| echo: false
if (length(deposits_lost)) {
  cat("Deposits dropped (no assay carries both parent + metabolite):\n  ",
      paste(deposits_lost, collapse = "\n  "), "\n", sep = "")
}
cat("Surviving entities:", length(entities_kept),
    " | surviving deposits:", length(datasets_with_kept_entity), "\n")
entity_coverage
```

## Paper-derived constants

Dataset-wide values from the linked publication or deposit description
(country of recruitment, body site, chromatography). Not an exclusion
mechanism; provenance lives in the inline code-comment citation. `constants`
is keyed by deposit_id and broadcast to every kept assay; `custom_col_map`
overrides the auto-mapping where native column names miss the
`DRAFT_CANDIDATES` heuristic.

**Location semantics.** `location_type` marks whether `country` is a
*study site* (where the study ran, population origin unverified) or a
*participant origin* (the population itself is the subject). Only
**MSV000082433** (Venezuelan Amerindian communities) is participant-origin;
every other `country` is a study site and must be read as a recruitment
covariate, **not** participant ethnicity.

```{r manual-annotations}
constants <- list(
  # --- study-site cohorts: `country` = recruitment location, not participant origin ---
  MTBLS1866    = list(country = "Italy",          location_type = "study_site"),   # Barberis et al. 2020
  MTBLS11656   = list(country = "China", body_site = "blood plasma",
                      location_type = "study_site"),                                # Xinxiang/Zhengzhou 2017 (PMC11823611)
  ST002044     = list(country = "India", body_site = "blood serum",
                      location_type = "study_site"),                                # Kumar et al.; mwTab has no body_site col
  ST002722     = list(country = "United States of America", body_site = "blood serum",
                      location_type = "study_site"),                                # UCSD NAFLD Center 2014-2017
  MSV000088255 = list(country = "United Kingdom", location_type = "study_site"),    # 100 spot urine, Glasgow, NHS 14/LO/188

  # chromatography companions (prior curation)
  MSV000092754 = list(chromatography = "HILIC"),                                    # HILIC companion of 092757
  MSV000092757 = list(chromatography = "RP"),                                       # RP companion of 092754

  # 082493/084008: omeprazole CYP2C19-probe cohort; genotype in rescue_samples.csv
  MSV000082493 = list(country = "United States of America", location_type = "study_site",
                      chromatography = "RP", health_status = "healthy"),
  MSV000084008 = list(country = "United States of America", location_type = "study_site",
                      chromatography = "RP", health_status = "healthy",
                      body_site = "blood plasma"),
  MSV000086975 = list(country = "United States of America", location_type = "study_site",
                      chromatography = "RP", body_site = "blood plasma"),           # schizophrenia (PMID 34741130)
  MSV000085256 = list(country = "United States of America", location_type = "study_site"),  # Alzheimer's (PMID 34244695)

  # the ONE deposit whose location genuinely describes participant origin:
  MSV000082433 = list(country = "Venezuela", location_type = "participant_origin",
                      chromatography = "RP", health_status = "healthy")             # Amerindian communities
)

custom_col_map <- list(
  MSV000092754 = list(
    subject_id    = "codigo_num",
    sex           = "sex",
    age           = "age_at_earliest_visit",
    health_status = "sero_final"),
  MSV000092757 = list(
    subject_id    = "codigo_num",
    sex           = "sex",
    age           = "age_at_earliest_visit",
    health_status = "sero_final"),
  # native columns the DRAFT_CANDIDATES heuristic misses (082493/084008 need none)
  MSV000086975 = list(subject_id = "#SampleID", age = "AGE", health_status = "Metadata")
)
```

## Consolidate and write the metadata table

`dataset_metadata` is the single source of truth for downstream code:
one row per kept `(deposit, assay)` entity, with all repository, paper,
and column-mapping information attached.

```{r build-metadata-table}
#| code-fold: true
kept_entities <- entity_coverage[entity_coverage$keep,
                                 c("dataset_id", "deposit_id", "assay")]

tech_renamed <- tech_raw |>
  dplyr::rename(deposit_id     = dataset_id,
                chromatography = chrom_type) |>
  dplyr::select(deposit_id, assay, repository, chromatography)

is_msv <- startsWith(kept_entities$deposit_id, "MSV")
msv_meta <- kept_entities[is_msv, ] |>
  dplyr::left_join(tech_renamed |> dplyr::select(-assay),
                   by = "deposit_id")
nonmsv_meta <- kept_entities[!is_msv, ] |>
  dplyr::left_join(tech_renamed, by = c("deposit_id", "assay"))

dataset_metadata <- dplyr::bind_rows(msv_meta, nonmsv_meta) |>
  dplyr::left_join(coverage |> dplyr::rename(deposit_id = dataset_id),
                   by = "deposit_id")

const_fields <- unique(unlist(lapply(constants, names)))
for (canon in const_fields) {
  col <- paste0("const_", canon)
  dataset_metadata[[col]] <- vapply(
    dataset_metadata$deposit_id,
    function(d) {
      v <- constants[[d]][[canon]]
      if (is.null(v)) NA_character_ else as.character(v)
    },
    character(1))
}

if ("const_chromatography" %in% names(dataset_metadata)) {
  use_const <- !is.na(dataset_metadata$const_chromatography) &
               (is.na(dataset_metadata$chromatography) |
                dataset_metadata$chromatography == "" |
                dataset_metadata$chromatography == "Unknown")
  dataset_metadata$chromatography[use_const] <-
    dataset_metadata$const_chromatography[use_const]
}

for (ds in names(custom_col_map)) {
  ix <- which(dataset_metadata$deposit_id == ds)
  if (!length(ix)) next
  for (canon in names(custom_col_map[[ds]])) {
    col_key <- paste0("col_", canon)
    if (col_key %in% names(dataset_metadata)) {
      dataset_metadata[[col_key]][ix] <- custom_col_map[[ds]][[canon]]
    }
  }
}

dataset_metadata <- dataset_metadata[
  , c("dataset_id", "deposit_id", "assay", "repository", "chromatography",
      "n_samples", "n_canonical_mapped",
      grep("^col_",   names(dataset_metadata), value = TRUE),
      grep("^const_", names(dataset_metadata), value = TRUE))]
```

```{r build-metadata-table-display}
#| echo: false
dataset_metadata
```

# Adduct inference

We do not trust MASST's `Adduct` column: it is inherited from the matched
*library* spectrum and need not reflect how the compound ionised in the query
dataset. Instead we recompute the adduct ourselves:

- compute the theoretical m/z of each candidate adduct
  (`r paste(ADDUCT_CANDIDATES, collapse = ", ")`) and assign the one closest to
  the observed `Precursor_MZ`, within 25 ppm;
- keep any hit that resolves to a candidate adduct, and drop the rest.

An earlier version additionally required parent and metabolite to *share* an adduct
within an entity, dropping any hit whose adduct was not in that shared set. That rule is
removed. The metabolite is only ever seen as `[M+H]+`, so the shared set was always
`[M+H]+` alone and the rule silently discarded every parent hit at another adduct —
`[M+Na]+`, `[M-H]-` and the in-source water loss `[M+H-H2O]+`. That produced a
parent-only shortfall in the hit set which had nothing to do with the data: a file in
which omeprazole was fragmented as its sodium adduct still contains omeprazole. The
shared-adduct argument applies to the *ratio*, where the two compounds must ionise
comparably, and it is enforced there rather than by deleting evidence of presence.

```{r infer-adduct}
role_formula <- setNames(compound_spec$formula, compound_spec$role)

all_hits$adduct_inferred <- mapply(
  infer_adduct,
  all_hits$Precursor_MZ,
  role_formula[all_hits$role],
  USE.NAMES = FALSE
)

shared_per_entity <- all_hits |>
  dplyr::filter(!is.na(adduct_inferred)) |>
  dplyr::group_by(dataset_id) |>
  dplyr::summarise(
    shared = list(intersect(
      adduct_inferred[role == "parent"],
      adduct_inferred[role == "metabolite"])),
    .groups = "drop")

shared_lookup <- setNames(shared_per_entity$shared,
                          shared_per_entity$dataset_id)

entities_with_shared <- shared_per_entity$dataset_id[
  vapply(shared_per_entity$shared, length, integer(1)) > 0]

kept_mask <- !is.na(all_hits$adduct_inferred)

dropped_entities <- setdiff(unique(all_hits$dataset_id),
                            entities_with_shared)
all_hits         <- all_hits[kept_mask, ]
dataset_metadata <- dataset_metadata[
  dataset_metadata$dataset_id %in% unique(all_hits$dataset_id), ]
```

```{r infer-adduct-summary}
#| echo: false
print(table(role   = all_hits$role,
            adduct = all_hits$adduct_inferred,
            useNA  = "ifany"))
if (length(dropped_entities)) {
  cat("\nEntities with no shared parent/metabolite adduct (kept; reported only):\n  ",
      paste(dropped_entities, collapse = "\n  "), "\n", sep = "")
}
cat("\nSurviving entities:", length(unique(all_hits$dataset_id)),
    " | hits retained:", nrow(all_hits), "\n")
```

# Write curated hits

```{r write-artifacts}
write.csv(dataset_metadata, "artifacts/dataset_metadata_table.csv",
          row.names = FALSE)
write.csv(all_hits,         "artifacts/hits_curated.csv",
          row.names = FALSE)
```

```{r write-artifacts-summary}
#| echo: false
cat("artifacts/dataset_metadata_table.csv :",
    nrow(dataset_metadata), "rows\n")
cat("artifacts/hits_curated.csv           :",
    nrow(all_hits), "rows /",
    length(unique(all_hits$dataset_id)), "entities\n")
```

# Confirmation (opt-in)

The three confirmation notebooks in `confirmation/` (`massIVE-hit-confirmation.qmd`,
`metaboLights-confirmation.qmd`, `metabolomics-workbench-confirmation.qmd`, one per
repository) decide which curated candidates are *really* omeprazole / 5-OH. For
each (dataset, assay, compound) they pull the MS2 scans behind the MASST hits, run
SIRIUS for the molecular formula and CSI:FingerID for the structure, and
assign a confidence tier (strict / standard / liberal / not_confirmed) from how
well formula and structure match the target. Only passing features reach the
`hits-confirmed-*` artifacts that `02` merges. This confirmation is the step that
removes isobaric look-alikes such as MSV000086975 and MSV000085256.

They run SIRIUS (slow, CSI:FingerID-queued, one shared instance), so they are
skipped on a normal render. Render with `-P run_confirmations:true` to chain them
here; each runs in a fresh quarto process, sequentially, so SIRIUS is never
concurrent.

```{r trigger-confirmations, eval=params$run_confirmations}
for (q in c("confirmation/massIVE-hit-confirmation.qmd",
            "confirmation/metaboLights-confirmation.qmd",
            "confirmation/metabolomics-workbench-confirmation.qmd")) {
  message("Rendering ", q, " …")
  system2("quarto", c("render", shQuote(q)))
}
```

# Appendix: dataset-specific notes

Quirks recorded as data so downstream layers can act on them; not part of the
main curation flow.

## Related deposits: what is not independent

A deposit accession is not the same thing as a study. Two of the accessions we
keep come from a single study, and any analysis that treats each dataset as its
own cohort (pooling, meta-analysis, counting subjects, weighting by precision)
would silently pseudo-replicate it.

**`MSV000082493` and `MSV000084008` are two accessions of one study** (the same
14-subject crossover, Jarmusch / Tsunoda, PMC7485946): a cytochrome P450 probe
cocktail including omeprazole was dosed on days 1 and 9, with cefprozil in between.
The same physical files appear under both accessions, with different paths and
extensions:

```
mzspec:MSV000082493:updates/2019-03-27_alan_jarmusch_6d15f636/peak/blood/DM003_1_360_RG10_01_43808.mzXML
mzspec:MSV000084008:ccms_peak/blood/DM013_1_360_RD1_01_43865.mzML
```

We keep both accessions but must never treat them as independent cohorts. The
de-duplication rule is applied downstream in `05` (collapse on chromatography +
file name); here we only record the relationship so those layers can act on it.

```{r related-deposits}
related_deposits <- tibble::tribble(
  ~dataset_id,    ~related_to,     ~relation,                  ~notes,
  "MSV000084008", "MSV000082493",  "same_study_different_arm",
  paste("Same 14-subject study (PMC7485946). 084008 re-deposits 082493's blood",
        "arm: 171 of 172 blood files are shared. 082493 contributes feces/skin/",
        "urine here, 084008 blood. Do not treat as independent cohorts and do",
        "not count the subjects twice.")
)
write.csv(related_deposits, "artifacts/related_deposits.csv", row.names = FALSE)
related_deposits
```


# Session info

Recorded so the rendered HTML captures the exact package versions that
produced the artifacts above.

```{r session-info}
sessionInfo()
```
