---
title: "01 - MASST curation"
subtitle: "Compound pair · MASST exports · dataset selection"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
params:
  run_confirmations: false
---

# Overview

First step of the pipeline: load the StructureMASST hit set for omeprazole and
its CYP2C19 metabolite (5-hydroxyomeprazole), keep datasets where both compounds
appear, and curate that intersection down to the (dataset, assay) combinations
suitable for a parent/metabolite ratio analysis.

A *dataset* means **one assay of one deposit**; every downstream step joins on
`dataset_id`, which encodes the deposit + assay pair.

The search:

- **Query spectra** are the GNPS reference-library MS/MS spectra of omeprazole
  and 5-hydroxyomeprazole, in both polarities. The discovery chunk below
  resolves each compound to its library spectra by InChIKey first block.
- **Search space** is every public MS/MS spectrum indexed by StructureMASST
  across MassIVE, MetaboLights and Metabolomics Workbench.
- **Match threshold** is cosine ≥ 0.7 and ≥ 3 matched peaks, analog search off.

Hits are read from **FASSTrecords** (a DOI-versioned SQLite of repository-scale
MASST annotations, DOI 10.5281/zenodo.18199544), so the discovery step is
scripted and pinned rather than a one-off web query.

**Reads:**

- `StructureMASST_output/fasstrecords_hits.csv` (written by the discovery chunk
  below from FASSTrecords; committed, so normally read as-is)

**Writes:**

- `artifacts/hits_curated.csv` - MASST hits restricted to curated keeps, with
  `dataset_id`, `deposit_id`, `assay` and `adduct_inferred` columns populated.
  Read by the three SIRIUS confirmation notebooks.
- `artifacts/dataset_metadata_table.csv` - one row per kept (deposit, assay)
  with repository, chromatography, paper-derived constants and the col_/const_
  mappings consumed by `load_metadata()`.

```{r setup}
source("R/_setup.R")
source("R/masst.R")
source("R/metadata.R")
source("check-versions.R")
```

# Compound pair

The omeprazole / 5-hydroxyomeprazole pair is a CYP2C19 phenotyping readout:
CYP2C19 hydroxylates omeprazole at the 5-position, so the parent/5-OH ratio
reflects CYP2C19 metaboliser status.

Both compounds were queried in both polarities; downstream code picks the adduct
with the most confirmed hits per dataset and collapses adducts to one feature
per compound per sample.

```{r compound-spec}
compound_spec
```

# MASST hits (FASSTrecords)

**Provenance and reproduction.** FASSTrecords is a repository-scale snapshot of
public MASST annotations (a ~1.2-billion-row SQLite of every GNPS/MASST spectrum
match), archived on Zenodo at **DOI 10.5281/zenodo.18199544** (sha256
`95a11bbe…`). We downloaded that archive and extracted the hit set on
**2026-07-10**. To reproduce, download the Zenodo archive to the `FASST_DB` path
in the discovery chunk below and delete
`StructureMASST_output/fasstrecords_hits.csv`; the chunk then resolves each
compound to its GNPS library spectra by InChIKey first block (omeprazole
`SUBDBMMJDZJVOS`, 5-OH `CMZHQFXXAAIBKE`), pulls every `masst_table` match at
cosine ≥ 0.7 and ≥ 3 matched peaks, joins the per-file ReDU fields, and writes
`StructureMASST_output/fasstrecords_hits.csv`.

Because the archive is immutable (DOI + checksum), this discovery step is fully
pinned. The chunk **skips itself** when the ~1.2B-row DB is absent or the CSV
already exists, so a fresh clone (and every normal re-render) uses the committed
output unchanged. Each row of the hit table carries:

- `ATTRIBUTE_DatasetAccession`, `file_path`, `scan_id`, `role` (parent /
  metabolite): where the hit is;
- `Precursor_MZ` and the inferred `Adduct`: what was matched;
- the per-file ReDU fields the hard stops key on: `ChromatographyAndPhase`,
  `IonizationSourceAndPolarity`, `NCBITaxonomy`, `SampleType`.

One class of row is dropped as the set is built rather than downstream. Some
mzML converters write the scan number `4294967295`, the maximum 32-bit unsigned
integer, when the scan index is unresolved; such a hit cannot locate a spectrum,
so `INVALID_SCAN` removes it below. The committed CSV therefore contains none,
and nothing further down needs to check again.

```{r discovery-fasstrecords}
# Delete the CSV to force a rebuild.
FASST_DB  <- "D:/FASSTrecords/db/masst_records.sqlite"
FASST_OUT <- "StructureMASST_output/fasstrecords_hits.csv"
if (file.exists(FASST_DB) && !file.exists(FASST_OUT)) {
  suppressPackageStartupMessages(library(RSQLite))
  IKEY <- c(parent = "SUBDBMMJDZJVOS", metabolite = "CMZHQFXXAAIBKE")  # omeprazole / 5-OH
  COSINE_MIN <- 0.7; PEAKS_MIN <- 3
  INVALID_SCAN <- 4294967295                                            # 32-bit uint sentinel
  con <- dbConnect(SQLite(), FASST_DB)
  unq <- function(x) trimws(gsub('"', '', x))          # strip stray quotes in Dataset/mri
  inl <- function(v) paste(v, collapse = ",")          # integer IN-list
  lib_ids <- lapply(IKEY, function(k) dbGetQuery(con, sprintf(
    "SELECT spectrum_id_int, spectrum_id, Adduct, Precursor_MZ
       FROM library_table WHERE InChIKey_smiles_firstBlock = '%s'", k)))
  get_role <- function(role) {
    lib <- lib_ids[[role]]; if (!nrow(lib)) return(NULL)
    h <- dbGetQuery(con, sprintf(
      "SELECT m.cosine AS Cosine, m.matching_peaks AS matching_peaks, m.scan_id AS scan_id,
              m.spectrum_id_int AS spectrum_id_int, m.mri_id_int AS mri_id_int,
              d.Dataset AS ATTRIBUTE_DatasetAccession, mr.mri AS mri
         FROM masst_table m
         JOIN dataset_table d ON m.dataset_id_int = d.dataset_id_int
         JOIN mri_table    mr ON m.mri_id_int     = mr.mri_id_int
        WHERE m.spectrum_id_int IN (%s) AND m.cosine >= %f AND m.matching_peaks >= %d",
      inl(lib$spectrum_id_int), COSINE_MIN, PEAKS_MIN))
    if (!nrow(h)) return(NULL)
    h <- dplyr::left_join(h, lib[, c("spectrum_id_int","spectrum_id","Adduct","Precursor_MZ")],
                          by = "spectrum_id_int")
    h$ATTRIBUTE_DatasetAccession <- unq(h$ATTRIBUTE_DatasetAccession); h$mri <- unq(h$mri)
    h$role <- role; h$inchikey_first_block <- IKEY[[role]]
    h$USI <- paste0(h$mri, ":scan:", h$scan_id)          # resolvable spectrum reference
    h$file_path <- sub("^mzspec:[^:]*:", "", h$mri)       # repo-relative path; basename used downstream
    h$`Matching Peaks` <- h$matching_peaks; h$lib_spectrum <- h$spectrum_id
    h }
  hits <- dplyr::bind_rows(get_role("parent"), get_role("metabolite"))
  hits <- hits[hits$scan_id != INVALID_SCAN | is.na(hits$scan_id), ]   # drop 32-bit sentinel scans
  mids <- unique(hits$mri_id_int); chunks <- split(mids, ceiling(seq_along(mids) / 5000))
  redu <- dplyr::bind_rows(lapply(chunks, function(ch) dbGetQuery(con, sprintf(
    "SELECT mri_id_int, ChromatographyAndPhase, IonizationSourceAndPolarity, NCBITaxonomy, SampleType
       FROM redu_table WHERE mri_id_int IN (%s)", inl(ch)))))
  cln <- function(v) { v[is.na(v) | v %in% c("missing value","not applicable","")] <- NA; v }
  for (cc in c("ChromatographyAndPhase","IonizationSourceAndPolarity","NCBITaxonomy","SampleType"))
    redu[[cc]] <- cln(redu[[cc]])
  redu <- redu[!duplicated(redu$mri_id_int), ]
  hits <- dplyr::left_join(hits, redu, by = "mri_id_int")
  dbDisconnect(con)
  dir.create("StructureMASST_output", showWarnings = FALSE)
  keep <- c("ATTRIBUTE_DatasetAccession","USI","file_path","scan_id","role","Cosine","Matching Peaks",
            "Adduct","Precursor_MZ","inchikey_first_block","lib_spectrum","ChromatographyAndPhase",
            "IonizationSourceAndPolarity","NCBITaxonomy","SampleType")
  write.csv(hits[, keep], FASST_OUT, row.names = FALSE)
  message("FASSTrecords discovery: wrote ", nrow(hits), " hits -> ", FASST_OUT)
} else {
  message("FASSTrecords discovery: using committed ", FASST_OUT,
          if (!file.exists(FASST_DB)) " (DB not present)" else " (already built)")
}
```

```{r load-hits}
all_hits <- read.csv("StructureMASST_output/fasstrecords_hits.csv",
                     check.names = FALSE, stringsAsFactors = FALSE)
# Normalise ReDU / MIxS placeholders ("missing value", …) to NA.
for (j in seq_along(all_hits))
  if (is.character(all_hits[[j]])) all_hits[[j]][is_missing_val(all_hits[[j]])] <- NA
all_hits$scan_id <- as.integer(all_hits$scan_id)
```

```{r load-hits-summary}
#| echo: false
cat("Total MASST hits loaded:", nrow(all_hits), "\n")
cat("  parent    :", sum(all_hits$role == "parent"), "\n")
cat("  metabolite:", sum(all_hits$role == "metabolite"), "\n")
```

# Dataset selection

## D_both: datasets where both compounds appear

We restrict to the intersection of datasets containing at least one MASST hit
for each compound. Datasets with only one compound cannot provide a
parent/metabolite ratio and are out of scope.

```{r d-both}
D_parent     <- unique(all_hits$ATTRIBUTE_DatasetAccession[all_hits$role == "parent"])
D_metabolite <- unique(all_hits$ATTRIBUTE_DatasetAccession[all_hits$role == "metabolite"])
D_both       <- intersect(D_parent, D_metabolite)
```

```{r d-both-summary}
#| echo: false
cat("parent datasets:", length(D_parent),
    "| metabolite datasets:", length(D_metabolite),
    "| intersection:", length(D_both), "\n")
```

# Per-deposit metadata

A deposit that contains the drug pair is only useful if its files can be tied to
the people they came from, and that information is not in the spectra. It sits
in whatever the depositor and the repository happened to record, in whatever
format each repository uses, at whatever level of completeness the submitter
chose. Reconciling that is the single largest piece of manual work in this
project, and it is worth being explicit about the cost rather than presenting
the harmonised table as though it fell out of an API.

Four sources are consulted, and none of them is currently sufficient : the
pinned per-deposit metadata directory, Pan-ReDU, the MetaboLights ISA-Tab, and
the Metabolomics Workbench mwTab. Where all four fail, the metadata has to be
found by hand. Six deposits needed that, and the sections below are the record
of it --- `MSV000085256` and `MSV000082433` from Pan-ReDU after the deposit
itself came up short, `MSV000086975` from a submitter TSV inside the deposit,
`MTBLS12231` from the ISA-Tab assay files, and the probe cohort (`MSV000082493`
/ `MSV000084008`) from **a GitHub repository belonging to the original
authors**, at a pinned commit. That last one is not a footnote: the CYP2C19
genotypes the entire validation rests on are in that repository and in no
metadata catalogue, so without someone reading the paper, finding the repository
and pinning a commit, the central result of this work does not exist.

The mapping is per deposit because there is no shared schema to map to. Each
block below translates one deposit's own column names into the fields the
pipeline needs, and each was written by reading that deposit's documentation.
That is the bottleneck: not compute, not storage, but a person working out what
`#SampleID` means in one study and `SubjectIdentifierAsRecorded` in another, and
whether either identifies a person. Efforts such as MetabolomicsHub are
addressing exactly this, and until something like it is universal, corpus-scale
reuse will keep paying this cost per deposit.

```{r metadata-setup}
MD <- "dataset_metadata"
miss <- function(x) {
  x[x %in% c("", "missing value", "not applicable", "not collected")] <- NA
  x
}
```


## Where each deposit's metadata came from

Six deposits needed their metadata found by hand, and each block below
translates one deposit's own column names into the fields the pipeline needs.
The source and the reading behind it, per deposit:

- **`MSV000082493` and `MSV000084008`** --- the omeprazole probe cohort
  (Jarmusch et al. 2020, *Clin Transl Sci*, PMC7485946). Healthy volunteers
  given a modified Cooperstown CYP-probe cocktail on study days 1 and 9 with a
  7-day cefprozil course in between; plasma, urine, feces and skin. The metadata
  is in neither deposit and in no catalogue --- it is in the authors' GitHub
  repository, `Materials/Metadata.txt` at commit `95586620`, and it carries the
  genotyped CYP2C19 phenotype per subject that `04` validates against. The two
  accessions share one table: plasma rows onboard under 084008, the rest under
  082493.
- **`MSV000086975`** --- schizophrenia (PMID 34741130). A submitter TSV inside
  the deposit.
- **`MSV000085256`** --- Alzheimer's (PMID 34244695). Pan-ReDU JSON, the deposit
  itself having come up short.
- **`MSV000082433`** --- Amerindian communities. Pan-ReDU JSON. The one deposit
  whose recorded country describes participant origin rather than a study site.
- **`MTBLS12231`** --- COVID (PMC12019401). MetaboLights ISA-Tab, the `s_`
  sample file joined to the two `a_` assay files. ISA-Tab permits duplicate
  column names, so the first occurrence of each is selected by index.

The probe cohort is fetched first because everything else reads from
`dataset_metadata/`, and its file is not in the repository until this runs.

```{r fetch-82493}
SRC_COMMIT <- "95586620ead4ac5241dc1e99a98e047c1f99b17b"
SRC_URL <- sprintf(paste0("https://raw.githubusercontent.com/alan-jarmusch/",
                          "UntargetedMS-DrugMetabolism/%s/Materials/Metadata.txt"),
                   SRC_COMMIT)
LOCAL <- "dataset_metadata/MSV000082493_metadata.tsv"
if (!file.exists(LOCAL))
  utils::download.file(SRC_URL, LOCAL, quiet = TRUE, mode = "wb")

jm <- read.delim(LOCAL, check.names = FALSE, stringsAsFactors = FALSE,
                na.strings = c("", "not applicable"))
```

## One schema for all of them

The blocks below write one row per file with the same columns whatever the
source, so the deposits can be combined. Three of those columns exist only
because the probe cohort has them: `timepoint` is its study day,
`timepoint_mins` the minutes post-dose within that day, and
`pharmkinetics_cyp2c19` the source study's own published per-subject CYP2C19
value. Every field is taken from a declared metadata attribute; nothing is
parsed out of a filename. The combined table is written to
`artifacts/rescue_samples.csv`.

```{r std-schema}
std_cols <- c("deposit_id","file","subject_id","sex","age","health_status",
              "specimen","timepoint","timepoint_mins","pharmkinetics_cyp2c19",
              "cyp2c19_phenotype","cyp2c19_genotype","include")
num_cols <- c("age", "timepoint_mins", "pharmkinetics_cyp2c19")
blank <- function(n) setNames(as.data.frame(matrix(NA_character_, n, length(std_cols))), std_cols)
```

```{r jarmusch}
jm_std <- data.frame(
  file = jm$filename, subject_id = jm$ATTRIBUTE_Subject, sex = jm$ATTRIBUTE_sex,
  age = as_num(jm$age), health_status = "healthy",
  specimen = dplyr::case_when(grepl("plasma", jm$ATTRIBUTE_Body_Site, TRUE) ~ "plasma",
                              grepl("^UA", jm$ATTRIBUTE_Body_Site) ~ "urine",
                              grepl("fecal", jm$ATTRIBUTE_Body_Site, TRUE) ~ "feces",
                              grepl("forehead|skin", jm$ATTRIBUTE_Body_Site, TRUE) ~ "skin",
                              TRUE ~ NA_character_),
  timepoint = jm$ATTRIBUTE_Study_Day,
  timepoint_mins = as_num(jm$ATTRIBUTE_Time_Point_Mins),
  pharmkinetics_cyp2c19 = as_num(jm$PharmKinetics_CYP2C19),
  cyp2c19_phenotype = jm$PharmGeno_Phenotype_2C19,
  cyp2c19_genotype = jm$PharmGeno_CYP2C19, include = jm$Include_Exclude,
  stringsAsFactors = FALSE)
jm_std <- jm_std[jm_std$include %in% "include" & !is.na(jm_std$specimen), ]
d84008 <- cbind(deposit_id = "MSV000084008", jm_std[jm_std$specimen == "plasma", std_cols[-1]])
d82493 <- cbind(deposit_id = "MSV000082493", jm_std[jm_std$specimen != "plasma", std_cols[-1]])
```

```{r probe-summary}
#| echo: false
cat("MSV000082493 + 084008:", nrow(jm_std), "biological samples of", nrow(jm), "files
")
cat("
specimen x study day:
");       print(table(jm_std$specimen, jm_std$timepoint, useNA = "ifany"))
cat("
CYP2C19 genotyped phenotype:
"); print(table(jm_std$cyp2c19_phenotype, useNA = "ifany"))
cat("
subjects:", length(unique(jm_std$subject_id)), "
")
```

MSV000084008 has no Pan-ReDU record and its hit files are all blood plasma, so
its metadata is written as the plasma subset of the shared table. Files hosted
under 082493 simply fail to sync under 084008 and are skipped, so nothing is
double counted.

```{r derive-84008}
is_plasma <- grepl("plasma", jm$ATTRIBUTE_Body_Site, ignore.case = TRUE)
plasma <- jm[is_plasma, , drop = FALSE]
write.table(plasma, "dataset_metadata/MSV000084008_metadata.tsv",
            sep = "\t", quote = FALSE, row.names = FALSE)
cat("MSV000084008_metadata.tsv <- plasma rows:", nrow(plasma), "of", nrow(jm), "\n")
cat("body sites kept:", paste(sort(unique(plasma$ATTRIBUTE_Body_Site)), collapse = ", "), "\n")
```

```{r sz}
sz <- read.delim(file.path(MD, "MSV000086975_metadata.tsv"), check.names = FALSE, stringsAsFactors = FALSE)
sid <- strsplit(sz$`#SampleID`, "_", fixed = TRUE)
d86975 <- blank(nrow(sz)); d86975$deposit_id <- "MSV000086975"
d86975$file <- sz$Metabolomics_FileName; d86975$subject_id <- sz$`#SampleID`
d86975$sex <- vapply(sid, function(p) if (length(p) >= 2) p[2] else NA_character_, character(1))
d86975$age <- as_num(sz$AGE)
d86975$health_status <- sz$Metadata          # HC / SZ
d86975$specimen <- "plasma"; d86975$include <- "include"
```

```{r az}
pj <- function(f) {
  txt <- paste(readLines(file.path(MD, f), warn = FALSE), collapse = "")
  objs <- regmatches(txt, gregexpr("\\{[^{}]*\\}", txt))[[1]]
  g <- function(o, k) { m <- regmatches(o, regexpr(sprintf('"%s": "[^"]*"', k), o))
    if (length(m)) sub('.*: "', "", sub('"$', "", m)) else NA_character_ }
  data.frame(file = miss(vapply(objs, g, character(1), "filename")),
             subject_id = miss(vapply(objs, g, character(1), "SubjectIdentifierAsRecorded")),
             sex = miss(vapply(objs, g, character(1), "BiologicalSex")),
             age = as_num(miss(vapply(objs, g, character(1), "AgeInYears"))),
             health_status = miss(vapply(objs, g, character(1), "HealthStatus")),
             specimen = miss(vapply(objs, g, character(1), "UBERONBodyPartName")),
             row.names = NULL, stringsAsFactors = FALSE) }
a <- pj("MSV000085256_panredu.json"); d85256 <- blank(nrow(a)); d85256$deposit_id <- "MSV000085256"
d85256[names(a)] <- a; d85256$specimen <- ifelse(grepl("plasma", d85256$specimen, TRUE), "plasma",
                                          ifelse(grepl("cerebro|CSF", d85256$specimen, TRUE), "CSF", d85256$specimen))
d85256$include <- "include"
```

```{r amerindians}
b <- pj("MSV000082433_panredu.json"); d82433 <- blank(nrow(b)); d82433$deposit_id <- "MSV000082433"
d82433[names(b)] <- b; d82433$include <- "include"
```

```{r covid}
col1 <- function(d, nm) d[[ which(names(d) == nm)[1] ]]
s <- read.delim(file.path(MD, "s_MTBLS12231.txt"), check.names = FALSE, stringsAsFactors = FALSE)
read_a <- function(p) { d <- read.delim(
    file.path(MD, sprintf("a_MTBLS12231_LC-MS_%s_reverse-phase_metabolite_profiling.txt", p)),
    check.names = FALSE, stringsAsFactors = FALSE)
  # the raw column is empty here; pick whichever *Spectral Data File column is
  # populated
  fcols <- grep("Spectral Data File", names(d))
  fcol  <- fcols[which.max(vapply(fcols, function(i)
             sum(grepl("\\.(mzML|mzXML|raw|d|wiff)$", d[[i]])), integer(1)))]
  data.frame(sample = col1(d, "Sample Name"), raw = basename(d[[fcol]]),
             stringsAsFactors = FALSE) }
amap <- rbind(read_a("positive"), read_a("negative"))
sref <- data.frame(sample = col1(s, "Sample Name"),
                   genetype = col1(s, "Factor Value[genetype]"),
                   part = col1(s, "Characteristics[Organism part]"), stringsAsFactors = FALSE)
mm <- dplyr::left_join(amap, sref, by = "sample")
d12231 <- blank(nrow(mm)); d12231$deposit_id <- "MTBLS12231"
d12231$file <- basename(mm$raw); d12231$subject_id <- mm$sample
d12231$health_status <- mm$genetype; d12231$specimen <- mm$part
d12231$include <- "include"
```

```{r combine}
frames <- lapply(list(d82493, d84008, d86975, d85256, d82433, d12231), function(f) {
  f$timepoint <- as.character(f$timepoint)
  for (cc in setdiff(std_cols, num_cols)) f[[cc]] <- as.character(f[[cc]])
  for (cc in num_cols) f[[cc]] <- as_num(f[[cc]])
  f[, std_cols]
})
all_std <- dplyr::bind_rows(frames)
write.csv(all_std, "artifacts/rescue_samples.csv", row.names = FALSE)

cat("Rescue samples per deposit x specimen:\n"); print(table(all_std$deposit_id, all_std$specimen, useNA = "ifany"))
cat("\nHealth stratifier availability per deposit:\n")
for (d in unique(all_std$deposit_id)) {
  x <- all_std[all_std$deposit_id == d, ]
  cat(sprintf("  %-14s n=%-4d sex=%-3s age=%-3s health={%s}\n", d, nrow(x),
      ifelse(any(!is.na(x$sex)),"Y","-"), ifelse(any(!is.na(x$age)),"Y","-"),
      paste(sort(unique(na.omit(x$health_status))), collapse="/")))
}
```


## Manual exclusions

Nothing about the acquisition removes a deposit at this point. All 44 come from
the three repositories the pipeline can download from (36 MassIVE, 4
MetaboLights, 4 Metabolomics Workbench), and every hit was acquired by liquid
chromatography with electrospray ionisation: reverse phase C18, HILIC or reverse
phase unspecified, positive or negative mode. No GC, no MALDI, no DESI. Species
is recorded properly where it is recorded at all --- `NCBITaxonomy` carries
`9606|Homo sapiens` throughout --- and the gap is coverage rather than content,
unevenly by repository: both MetaboLights and both Workbench deposits populate
it on every hit, only 6 of 8 MassIVE deposits populate it at all, and at this
stage it is present for 25 of the 44. The wider corpus is not all human, since
across the 128 deposits with a hit the same column also carries mouse, bacteria
and three *Banisteriopsis* species, and two accessions are NORMAN identifiers
the pipeline cannot download --- but those leave at the both-roles filter, so no
taxonomic gate is load-bearing here. It would be, on a query that returned them.

What does remove deposits is repository- or study-specific and decided by hand.
The `reason` vocabulary:

| reason | what it means |
|---|---|
| `no_or_limited_metadata` | no usable biological metadata anywhere (deposit + Pan-ReDU + publication), or only a single value for every stratifier |
| `access_restricted` | demographics gated behind a controlled-access portal (e.g. NCRAD); only body_site / country recoverable |
| `dia` | DIA acquisition; single-scan MASST matching is unreliable |
| `in_vitro` | cell culture or reference-library deposit; MASST hits are spike artefacts, not biological exposure |
| `non_human` | non-human taxon. Caught here rather than by the taxonomy hard stop because `redu_table` carries no `NCBITaxonomy` for these deposits, so the hard stop cannot fire |
| `reference_material` | QA/QC reference-material suite (e.g. NIST RM), not a phenotyped study cohort |

```{r other-exclusions}
other_exclusions <- tibble::tribble(
  ~dataset_id,    ~reason,                  ~notes,
  "MSV000085870", "no_or_limited_metadata", "DDA acquisition confirmed but no Pan-ReDU or deposit metadata",
  "MSV000093072", "no_or_limited_metadata", "",
  "MSV000095726", "no_or_limited_metadata", "",
  "MSV000096359", "no_or_limited_metadata", "",
  "MSV000096884", "no_or_limited_metadata", "",
  "MSV000097110", "no_or_limited_metadata", "",
  "MSV000094515", "no_or_limited_metadata", "stool / Alzheimer's (U19 ADRC NCRAD); demographics NCRAD-controlled",
  "ST002238",     "no_or_limited_metadata", "no metadata file or linked publication found",
  "MSV000094395", "no_or_limited_metadata", "only DOIDCommonName=Alzheimer's, no other strata in MASST or Pan-ReDU",
  "MSV000095143", "no_or_limited_metadata", "submitter TSV documents 264 files with rich metadata but those filenames don't exist on MassIVE; the 1077 actually-hosted files (Pan-ReDU naming, used by MASST + SIRIUS) have all canonical fields as 'missing value' in every source. Deposit cannot be stratified.",
  "MSV000097935", "access_restricted",      "demographics access-restricted ('Partial Public' MassIVE deposit, Rancho Bernardo Study policy)",
  "MTBLS417",     "dia",                    "DIA; duplicate of a DIA MassIVE deposit, poor metadata",
  "MTBLS2746",    "in_vitro",               "drug-treatment cell culture; MASST hits are spike artefacts",
  "ST001715",     "in_vitro",               "in vitro xenobiotic reference library",
  "MSV000082382", "non_human",              "Pseudomonas / Bacillus on melon plants",
  "MSV000086015", "non_human",              "San Diego Zoo rhinoceros serum (Diceros / Ceratotherium)",
  "MSV000087018", "reference_material",     "NIST RM 8231 frozen-plasma reference-material QA/QC suite, not a study cohort",
  "MSV000090544", "reference_material",     "GNPS Matrix-tubes extraction-method validation on high-biomass fecal+saliva (DOI 10.25345/C5QJ7831B); plate/well layout, no per-sample metadata, not a dosed cohort; omeprazole is incidental",
  "MSV000083065", "no_or_limited_metadata", "NOBEL twin plasma cohort; no Pan-ReDU, no accessible per-sample strata",
  "MSV000083593", "no_or_limited_metadata", "S. aureus infection plasma; no accessible per-sample metadata",
  "MSV000083955", "no_or_limited_metadata", "cerebral cavernous malformation plasma; no per-sample metadata",
  "MSV000084716", "no_or_limited_metadata", "Undiagnosed Disease Network; paper (PMID 33883556) confirms no medication/CYP data, part GC-MS, not in Pan-ReDU",
  "MSV000085508", "no_or_limited_metadata", "UDN, same cohort as MSV000084716 (duplicate), no metadata",
  "MSV000086270", "no_or_limited_metadata", "U19 ADNI-Go 1000-plasma cohort; no accessible per-sample strata",
  "MSV000088645", "no_or_limited_metadata", "U19 T1000 serum; no accessible per-sample strata",
  "MSV000090580", "no_or_limited_metadata", "osteoarthritis plasma (Guma); no accessible per-sample metadata",
  "MSV000092926", "no_or_limited_metadata", "plasma cohort (Knight ADC); no accessible per-sample metadata",
  "MSV000093037", "no_or_limited_metadata", "unidentified deposit: no title/summary/keywords/ReDU",
  "MSV000095383", "no_or_limited_metadata", "U19 Wisconsin serum (Alzheimer's); no accessible per-sample strata"
)

D_both <- setdiff(D_both, other_exclusions$dataset_id)
```

```{r other-exclusions-summary}
#| echo: false
cat("Other manual exclusions:", nrow(other_exclusions), "datasets\n")
print(table(other_exclusions$reason))
cat("\nSurviving after manual exclusions:", length(D_both), "datasets\n")
```

## Auto-derived dataset metadata

The rescue above covers the six deposits no automated route reaches. This step
is the automated route, run over every surviving deposit --- including those
six, whose files it can now read because the blocks above wrote them into
`dataset_metadata/`. For each deposit we harvest **technical** metadata
(per-assay: chromatography, ionisation, instrument) and **biological** metadata
(per-deposit: which native columns hold subject_id, sex, age). Sources and cache
locations are repository-specific:

- **MetaboLights (`MTBLS*`)**: technical via `mtbls_assay_data()` (the assay
  `a_*.txt` files); biological via `mtbls_sample_data()` (the `s_*.txt` sample
  file). Both auto-cached from EBI FTP at render time and stored as
  `dataset_metadata/{a_,s_}*.txt`.
- **Metabolomics Workbench (`ST*`)**: technical via `mwb_metadata()$MS_run`;
  biological via `mwb_metadata()$sample_annotation`. Both auto-cached from the
  Workbench API into `dataset_metadata/{ds}_AN*.txt`.
- **MassIVE (`MSV*`)**: technical from the submitter deposit TSV/CSV
  (`dataset_metadata/{ds}_metadata.*`) first, with Pan-ReDU
  (`fetch_panredu_metadata()` from `redu.gnps2.org`, cached as
  `dataset_metadata/{ds}_panredu.tsv`) as per-field fallback. Biological
  metadata comes from the same two sources, in the same order.

ReDU / MIxS missing-value placeholders (`"missing value"`, `"not applicable"`,
…) are normalised to NA everywhere via the shared `is_missing_val()` helper in
`R/metadata.R`.

```{r auto-assess}
#| code-fold: true
tech_raw <- fetch_assay_metadata(D_both)
coverage <- tibble::as_tibble(
  dplyr::bind_rows(lapply(D_both, harvest_metadata_coverage))
)
write.csv(tech_raw, "artifacts/assay_metadata.csv", row.names = FALSE)
```

```{r auto-assess-display}
#| echo: false
coverage[, c("dataset_id", "n_samples", "n_canonical_mapped",
             grep("^col_", names(coverage), value = TRUE))]
```

## Per-assay refinement (MTBLS / ST)

MetaboLights and Workbench deposits often package several assays under one
accession (MTBLS1866: LC-MS pos, LC-MS neg, GCxGC-MS pos; ST002044: four
HILIC/RP analyses). Since the ratio needs **both** compounds in the same assay,
we treat each `(deposit, assay)` pair as a separate entity and keep only
entities with at least one parent AND one metabolite hit. A deposit whose every
assay fails is appended to `other_exclusions` with reason
`no_assay_with_both_roles`.

MassIVE deposits have no assay concept and stay at deposit granularity.

```{r assay-assignment}
mtbls_st_keeps <- D_both[grepl("^(MTBLS|ST)", D_both)]
assay_lookups  <- setNames(lapply(mtbls_st_keeps, assay_lookup_for),
                           mtbls_st_keeps)

all_hits$assay <- mapply(function(ds, fp) {
  if (startsWith(ds, "MSV")) return(NA_character_)
  file_to_assay(basename(fp), assay_lookups[[ds]])
}, all_hits$ATTRIBUTE_DatasetAccession, all_hits$file_path,
USE.NAMES = FALSE)

all_hits$deposit_id <- all_hits$ATTRIBUTE_DatasetAccession
all_hits$dataset_id <- make_dataset_id(all_hits$deposit_id, all_hits$assay)

# MASST hits whose file basename doesn't resolve to any known assay
mtbls_st_hits <- all_hits$deposit_id %in% mtbls_st_keeps
unresolved    <- mtbls_st_hits & is.na(all_hits$assay)
all_hits$dataset_id[unresolved] <- paste0(all_hits$deposit_id[unresolved],
                                          "_unmatched")
```

```{r entity-coverage}
entity_coverage <- all_hits[all_hits$deposit_id %in% D_both, ] |>
  dplyr::group_by(dataset_id, deposit_id, assay) |>
  dplyr::summarise(has_parent = any(role == "parent"),
                   has_metab  = any(role == "metabolite"),
                   n_hits     = dplyr::n(),
                   .groups    = "drop") |>
  dplyr::mutate(keep = has_parent & has_metab &
                       !grepl("_unmatched$", dataset_id))

datasets_with_kept_entity <- unique(
  entity_coverage$deposit_id[entity_coverage$keep])
deposits_lost <- setdiff(D_both, datasets_with_kept_entity)

if (length(deposits_lost)) {
  new_excl <- tibble::tibble(
    dataset_id = deposits_lost,
    reason     = "no_assay_with_both_roles",
    notes      = "parent + metabolite MASST hits land in different assays of the same deposit")
  other_exclusions <- dplyr::bind_rows(other_exclusions, new_excl)
  D_both <- setdiff(D_both, deposits_lost)
}

entities_kept <- entity_coverage$dataset_id[entity_coverage$keep]
all_hits      <- all_hits[all_hits$dataset_id %in% entities_kept, ]
```

```{r entity-coverage-summary}
#| echo: false
if (length(deposits_lost)) {
  cat("Deposits dropped (no assay carries both parent + metabolite):\n  ",
      paste(deposits_lost, collapse = "\n  "), "\n", sep = "")
}
cat("Surviving entities:", length(entities_kept),
    " | surviving deposits:", length(datasets_with_kept_entity), "\n")
entity_coverage
```

## Paper-derived constants

Dataset-wide values from the linked publication or deposit description (country
of recruitment, body site, chromatography). Not an exclusion mechanism;
provenance lives in the inline code-comment citation. `constants` is keyed by
deposit_id and broadcast to every kept assay; `custom_col_map` overrides the
auto-mapping where native column names miss the `DRAFT_CANDIDATES` heuristic.

**Location semantics.** `location_type` marks whether `country` is a *study
site* (where the study ran, population origin unverified) or a *participant
origin* (the population itself is the subject). Only **MSV000082433**
(Venezuelan Amerindian communities) is participant-origin; every other `country`
is a study site and must be read as a recruitment covariate, **not** participant
ethnicity.

```{r manual-annotations}
constants <- list(
  # --- study-site cohorts: `country` = recruitment location, not participant
  # origin ---
  MTBLS1866    = list(country = "Italy",          location_type = "study_site"),   # Barberis et al. 2020
  MTBLS11656   = list(country = "China", body_site = "blood plasma",
                      location_type = "study_site"),                                # Xinxiang/Zhengzhou 2017 (PMC11823611)
  ST002044     = list(country = "India", body_site = "blood serum",
                      location_type = "study_site"),                                # Kumar et al.; mwTab has no body_site col
  ST002722     = list(country = "United States of America", body_site = "blood serum",
                      location_type = "study_site"),                                # UCSD NAFLD Center 2014-2017
  MSV000088255 = list(country = "United Kingdom", location_type = "study_site"),    # 100 spot urine, Glasgow, NHS 14/LO/188

  # chromatography companions (prior curation)
  MSV000092754 = list(chromatography = "HILIC"),                                    # HILIC companion of 092757
  MSV000092757 = list(chromatography = "RP"),                                       # RP companion of 092754

  # 082493/084008: omeprazole CYP2C19-probe cohort; genotype in
  # rescue_samples.csv
  MSV000082493 = list(country = "United States of America", location_type = "study_site",
                      chromatography = "RP", health_status = "healthy"),
  MSV000084008 = list(country = "United States of America", location_type = "study_site",
                      chromatography = "RP", health_status = "healthy",
                      body_site = "blood plasma"),
  MSV000086975 = list(country = "United States of America", location_type = "study_site",
                      chromatography = "RP", body_site = "blood plasma"),           # schizophrenia (PMID 34741130)
  MSV000085256 = list(country = "United States of America", location_type = "study_site"),  # Alzheimer's (PMID 34244695)

  # the ONE deposit whose location genuinely describes participant origin:
  MSV000082433 = list(country = "Venezuela", location_type = "participant_origin",
                      chromatography = "RP", health_status = "healthy")             # Amerindian communities
)

custom_col_map <- list(
  MSV000092754 = list(
    subject_id    = "codigo_num",
    sex           = "sex",
    age           = "age_at_earliest_visit",
    health_status = "sero_final"),
  MSV000092757 = list(
    subject_id    = "codigo_num",
    sex           = "sex",
    age           = "age_at_earliest_visit",
    health_status = "sero_final"),
  # native columns the DRAFT_CANDIDATES heuristic misses (082493/084008 need
  # none)
  MSV000086975 = list(subject_id = "#SampleID", age = "AGE", health_status = "Metadata")
)
```

## Consolidate and write the metadata table

`dataset_metadata` is the single source of truth for downstream code: one row
per kept `(deposit, assay)` entity, with all repository, paper, and
column-mapping information attached.

```{r build-metadata-table}
#| code-fold: true
kept_entities <- entity_coverage[entity_coverage$keep,
                                 c("dataset_id", "deposit_id", "assay")]

tech_renamed <- tech_raw |>
  dplyr::rename(deposit_id     = dataset_id,
                chromatography = chrom_type) |>
  dplyr::select(deposit_id, assay, repository, chromatography)

is_msv <- startsWith(kept_entities$deposit_id, "MSV")
msv_meta <- kept_entities[is_msv, ] |>
  dplyr::left_join(tech_renamed |> dplyr::select(-assay),
                   by = "deposit_id")
nonmsv_meta <- kept_entities[!is_msv, ] |>
  dplyr::left_join(tech_renamed, by = c("deposit_id", "assay"))

dataset_metadata <- dplyr::bind_rows(msv_meta, nonmsv_meta) |>
  dplyr::left_join(coverage |> dplyr::rename(deposit_id = dataset_id),
                   by = "deposit_id")

const_fields <- unique(unlist(lapply(constants, names)))
for (canon in const_fields) {
  col <- paste0("const_", canon)
  dataset_metadata[[col]] <- vapply(
    dataset_metadata$deposit_id,
    function(d) {
      v <- constants[[d]][[canon]]
      if (is.null(v)) NA_character_ else as.character(v)
    },
    character(1))
}

if ("const_chromatography" %in% names(dataset_metadata)) {
  use_const <- !is.na(dataset_metadata$const_chromatography) &
               (is.na(dataset_metadata$chromatography) |
                dataset_metadata$chromatography == "" |
                dataset_metadata$chromatography == "Unknown")
  dataset_metadata$chromatography[use_const] <-
    dataset_metadata$const_chromatography[use_const]
}

for (ds in names(custom_col_map)) {
  ix <- which(dataset_metadata$deposit_id == ds)
  if (!length(ix)) next
  for (canon in names(custom_col_map[[ds]])) {
    col_key <- paste0("col_", canon)
    if (col_key %in% names(dataset_metadata)) {
      dataset_metadata[[col_key]][ix] <- custom_col_map[[ds]][[canon]]
    }
  }
}

dataset_metadata <- dataset_metadata[
  , c("dataset_id", "deposit_id", "assay", "repository", "chromatography",
      "n_samples", "n_canonical_mapped",
      grep("^col_",   names(dataset_metadata), value = TRUE),
      grep("^const_", names(dataset_metadata), value = TRUE))]
```

```{r build-metadata-table-display}
#| echo: false
dataset_metadata
```

# Adduct inference

We do not trust MASST's `Adduct` column: it is inherited from the matched
*library* spectrum and need not reflect how the compound ionised in the query
dataset. Instead we recompute the adduct ourselves:

- compute the theoretical m/z of each candidate adduct
  (`r paste(ADDUCT_CANDIDATES, collapse = ", ")`) and assign the one closest to
  the observed `Precursor_MZ`, within 25 ppm;
- keep any hit that resolves to a candidate adduct, and drop the rest.

No requirement is placed on parent and metabolite sharing an adduct. The
metabolite is only ever seen as `[M+H]+`, so such a rule would reduce to
`[M+H]+` alone and discard every parent hit at another adduct --- `[M+Na]+`,
`[M-H]-` and the in-source water loss `[M+H-H2O]+` --- producing a parent-only
shortfall with nothing to do with the data: a file in which omeprazole was
fragmented as its sodium adduct still contains omeprazole. The shared-adduct
argument belongs to the *ratio*, where the two compounds must ionise comparably,
and it is enforced there rather than by deleting evidence of presence.

```{r infer-adduct}
role_formula <- setNames(compound_spec$formula, compound_spec$role)

all_hits$adduct_inferred <- mapply(
  infer_adduct,
  all_hits$Precursor_MZ,
  role_formula[all_hits$role],
  USE.NAMES = FALSE
)

shared_per_entity <- all_hits |>
  dplyr::filter(!is.na(adduct_inferred)) |>
  dplyr::group_by(dataset_id) |>
  dplyr::summarise(
    shared = list(intersect(
      adduct_inferred[role == "parent"],
      adduct_inferred[role == "metabolite"])),
    .groups = "drop")

shared_lookup <- setNames(shared_per_entity$shared,
                          shared_per_entity$dataset_id)

entities_with_shared <- shared_per_entity$dataset_id[
  vapply(shared_per_entity$shared, length, integer(1)) > 0]

kept_mask <- !is.na(all_hits$adduct_inferred)

dropped_entities <- setdiff(unique(all_hits$dataset_id),
                            entities_with_shared)
all_hits         <- all_hits[kept_mask, ]
dataset_metadata <- dataset_metadata[
  dataset_metadata$dataset_id %in% unique(all_hits$dataset_id), ]
```

```{r infer-adduct-summary}
#| echo: false
print(table(role   = all_hits$role,
            adduct = all_hits$adduct_inferred,
            useNA  = "ifany"))
if (length(dropped_entities)) {
  cat("\nEntities with no shared parent/metabolite adduct (kept; reported only):\n  ",
      paste(dropped_entities, collapse = "\n  "), "\n", sep = "")
}
cat("\nSurviving entities:", length(unique(all_hits$dataset_id)),
    " | hits retained:", nrow(all_hits), "\n")
```

# Write curated hits

```{r write-artifacts}
write.csv(dataset_metadata, "artifacts/dataset_metadata_table.csv",
          row.names = FALSE)
write.csv(all_hits,         "artifacts/hits_curated.csv",
          row.names = FALSE)
```

```{r write-artifacts-summary}
#| echo: false
cat("artifacts/dataset_metadata_table.csv :",
    nrow(dataset_metadata), "rows\n")
cat("artifacts/hits_curated.csv           :",
    nrow(all_hits), "rows /",
    length(unique(all_hits$dataset_id)), "entities\n")
```

# Confirmation (opt-in)

The three confirmation notebooks in `confirmation/`
(`massIVE-hit-confirmation.qmd`, `metaboLights-confirmation.qmd`,
`metabolomics-workbench-confirmation.qmd`, one per repository) decide which
curated candidates are *really* omeprazole / 5-OH. For each (dataset, assay,
compound) they pull the MS2 scans behind the MASST hits, run SIRIUS for the
molecular formula and CSI:FingerID for the structure, and assign a confidence
tier (strict / standard / liberal / not_confirmed) from how well formula and
structure match the target. Only passing features reach the `hits-confirmed-*`
artifacts that `02` merges. This confirmation is the step that removes isobaric
look-alikes such as MSV000086975 and MSV000085256.

They run SIRIUS (slow, CSI:FingerID-queued, one shared instance), so they are
skipped on a normal render. Render with `-P run_confirmations:true` to chain
them here; each runs in a fresh quarto process, sequentially, so SIRIUS is never
concurrent.

```{r trigger-confirmations, eval=params$run_confirmations}
for (q in c("confirmation/massIVE-hit-confirmation.qmd",
            "confirmation/metaboLights-confirmation.qmd",
            "confirmation/metabolomics-workbench-confirmation.qmd")) {
  message("Rendering ", q, " …")
  system2("quarto", c("render", shQuote(q)))
}
```

# Appendix: related deposits

A deposit accession is not the same thing as a study. Two pairs of the
accessions we keep are related, in opposite ways, and both would corrupt an
analysis that treated each accession as its own cohort. They are recorded here
as data so downstream layers can act on them rather than as part of the main
curation flow, and the distinction between them matters: one pair must be
collapsed, the other must not.

**`MSV000082493` and `MSV000084008` are two accessions of one study** (the same
14-subject crossover, Jarmusch / Tsunoda, PMC7485946): a cytochrome P450 probe
cocktail including omeprazole was dosed on days 1 and 9, with cefprozil in
between. The same physical files appear under both accessions, with different
paths and extensions:

```
mzspec:MSV000082493:updates/2019-03-27_alan_jarmusch_6d15f636/peak/blood/DM003_1_360_RG10_01_43808.mzXML
mzspec:MSV000084008:ccms_peak/blood/DM013_1_360_RD1_01_43865.mzML
```

We keep both accessions but must never treat them as independent cohorts. Of
their files, 333 are the same physical sample under both accessions.

**`MSV000092754` and `MSV000092757` are the opposite case.** They share all 164
file names and cover the same 79 subjects, but they are not duplicates: one is
the HILIC run and the other reverse phase, so each file name refers to two
genuinely different measurements of one specimen. Collapsing them on file name,
which is the right move for the pair above, would here average two
chromatographies and destroy the only within-subject comparison in the corpus
that holds biology fixed while changing the method. They count as one study for
pooling and as two measurements for everything else.

The de-duplication rules are applied downstream in `05`; here we only record the
relationships so those layers can act on them.

```{r related-deposits}
related_deposits <- tibble::tribble(
  ~dataset_id,    ~related_to,     ~relation,                  ~notes,
  "MSV000084008", "MSV000082493",  "same_study_different_arm",
  paste("Same 14-subject study (PMC7485946). 084008 re-deposits 082493's blood",
        "arm: 333 files are shared. 082493 contributes feces/skin/urine here,",
        "084008 blood. Do not treat as independent cohorts and do not count the",
        "subjects twice: collapse on file name."),

  "MSV000092757", "MSV000092754",  "same_subjects_two_chromatographies",
  paste("Same 79-subject Chagas serum panel measured twice, HILIC (092754) and",
        "reverse phase (092757). All 164 file names are shared but the files are",
        "different measurements, not copies. One study for pooling; never merge",
        "on file name.")
)
write.csv(related_deposits, "artifacts/related_deposits.csv", row.names = FALSE)
related_deposits
```


# Session info

Recorded so the rendered HTML captures the exact package versions that produced
the artifacts above.

```{r session-info}
sessionInfo()
```
