---
title: "02 - Confirmed features and biological files"
subtitle: "Filter SIRIUS hits · RT/mz diagnostics · per-dataset bio-file lists"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
execute:
  warning: false
  message: false
---

# Overview

This document combines the three SIRIUS confirmation notebooks, narrows to the
highest-tier confirmations where both parent and metabolite are present in the
same assay, and produces the cross-dataset RT and m/z diagnostics for the EIC
windows used next. It then builds, per surviving dataset, the list of
biological MS files (excluding QCs / blanks / pooled standards) the EIC
pipeline will load.

**Reads:**

- `artifacts/hits-confirmed-massive.csv`
- `artifacts/hits-confirmed-metaboLights.csv`
- `artifacts/hits-confirmed-metabolomics-workbench.csv`
- `artifacts/dataset_metadata_table.csv` (from `01`)
- `dataset_metadata/` (cached repository metadata)

**Writes:**

- `artifacts/hits-confirmed-all.csv` - combined, filtered confirmed features
  (one row per surviving feature, keyed by combined `dataset_id`)
- `artifacts/datasets-confirmed.csv` - one row per `dataset_id`
- `artifacts/bio_files_per_ds.rds` - named list: `dataset_id` -> character
  vector of biological MS file paths
- `artifacts/bio_assays_per_ds.rds` - named list: `dataset_id` -> assay identifier

```{r setup}
source("R/_setup.R")
source("R/masst.R")
source("R/metadata.R")
```

# Load confirmed features

Each confirmation notebook writes a self-contained CSV (`dataset_id`
deposit-level, `assay`, `role`, `rt`, `mz`, `adduct_inferred`, `source_file`,
`confidence`, plus SIRIUS scoring columns). We bind them, rename the
deposit-level key to `deposit_id`, and attach the combined `dataset_id` and
chromatography from `dataset_metadata_table.csv` via a `(deposit_id, assay)`
join.

The count here is **12**, not the 14 curated in `01`: SIRIUS confirmed the
omeprazole / 5-OH pair in 12 datasets and rejected two (MSV000086975 and
MSV000085256) where MASST flagged features at the target masses but the MS2
supported a different, isobaric formula. Both are correct rejections: SIRIUS's
top formula for the parent is fluorinated and sulfur-free (C17H16FN3O4 for
086975, C14H16F3N4O3 for 085256), not omeprazole's C17H19N3O3S, so these are
isobaric look-alikes rather than the target.

```{r load-confirmed}
conf_all <- rbind(
  read.csv("artifacts/hits-confirmed-massive.csv",                check.names = FALSE),
  read.csv("artifacts/hits-confirmed-metaboLights.csv",           check.names = FALSE),
  read.csv("artifacts/hits-confirmed-metabolomics-workbench.csv", check.names = FALSE)
)

names(conf_all)[names(conf_all) == "dataset_id"] <- "deposit_id"

ds_meta_tbl <- read.csv("artifacts/dataset_metadata_table.csv",
                        check.names = FALSE)

conf_all <- dplyr::left_join(
  conf_all,
  ds_meta_tbl[, c("dataset_id", "deposit_id", "assay", "chromatography")],
  by = c("deposit_id", "assay"))

conf_all <- conf_all[!is.na(conf_all$dataset_id), ]
conf_all$confidence <- factor(conf_all$confidence,
                              levels = c("strict", "standard",
                                         "liberal", "not_confirmed"))
```

```{r load-confirmed-summary}
#| echo: false
cat("Confirmed features loaded :", nrow(conf_all), "\n")
cat("Datasets (combined key)   :", length(unique(conf_all$dataset_id)), "\n")
cat("\nChromatography across confirmed datasets:\n")
print(table(unique(conf_all[, c("dataset_id", "chromatography")])$chromatography,
            useNA = "ifany"))
```

# Filter cascade

Both parent/metabolite hits per assay and a shared adduct were already
required upstream, so the remaining filters narrow to what is *interpretable*
after SIRIUS: highest tier per dataset+role, both roles still present after
annotation.

## Highest confidence tier per (dataset, role)

SIRIUS returns a formula rank and a structure rank per feature; we bin them into
tiers (set in the confirmation notebooks):

- **strict** — formula rank 1 and structure rank 1;
- **standard** — formula rank ≤ 3 and structure rank ≤ 5;
- **liberal** — formula rank ≤ 15 and structure rank ≤ 15;
- **not_confirmed** — beyond liberal; dropped before the confirmed CSVs.

A feature can match at several tiers, so we keep only the highest tier reached
per (dataset, role).

```{r filter-best-confidence}
CONF_ORDER <- c(strict = 1, standard = 2, liberal = 3, not_confirmed = 4)

conf_best <- conf_all |>
  dplyr::group_by(dataset_id, role) |>
  dplyr::filter(CONF_ORDER[as.character(confidence)] ==
                  min(CONF_ORDER[as.character(confidence)])) |>
  dplyr::ungroup()
```

## Both roles confirmed (any tier above not_confirmed)

A `dataset_id` is kept only if both `parent` and `metabolite` have at
least one feature at `strict`, `standard`, or `liberal`. The
`confidence` column flows through unchanged so downstream code can
stratify or weight by tier in the meta-analysis.

```{r filter-both-roles}
conf_final <- conf_best |>
  dplyr::filter(confidence %in% c("strict", "standard", "liberal")) |>
  dplyr::group_by(dataset_id) |>
  dplyr::filter(all(c("parent", "metabolite") %in% role)) |>
  dplyr::ungroup() |>
  as.data.frame()

dropped_ds <- setdiff(unique(conf_all$dataset_id),
                      unique(conf_final$dataset_id))
```

```{r filter-summary}
#| echo: false
cat("Surviving datasets :", length(unique(conf_final$dataset_id)), "\n")
cat("Surviving features :", nrow(conf_final), "\n")
if (length(dropped_ds))
  cat("Dropped (one role not confirmed at any tier):\n  ",
      paste(dropped_ds, collapse = "\n  "), "\n", sep = "")
```

# RT diagnostics

Median and range of the confirmed retention time per role and dataset, a check
that parent and metabolite elute where the EIC windows in `03` will look.

## Per-dataset RT windows

Coloured line is the min-to-max range; point is the median.

```{r plot-rt-windows-raw, fig.width = 9, fig.height = 5}
plot_rt_windows <- function(df, title) {
  df |>
    dplyr::filter(!is.na(rt)) |>
    dplyr::group_by(dataset_id, role, chromatography) |>
    dplyr::summarise(rt_lo  = min(rt),
                     rt_hi  = max(rt),
                     rt_med = median(rt),
                     .groups = "drop") |>
    ggplot(aes(y = dataset_id, colour = role)) +
    geom_linerange(aes(xmin = rt_lo, xmax = rt_hi),
                   linewidth = 2, alpha = 0.4,
                   position = position_dodge(width = 0.5)) +
    geom_point(aes(x = rt_med),
               size = 3, position = position_dodge(width = 0.5)) +
    facet_grid(chromatography ~ ., scales = "free_y", space = "free_y") +
    scale_colour_manual(values = c(parent = "#2166AC", metabolite = "#D6604D")) +
    labs(x = "Retention time (s)", y = NULL, colour = "Role",
         title = title) +
    theme_bw() +
    theme(legend.position  = "bottom",
          strip.background = element_rect(fill = "grey92"))
}

plot_rt_windows(conf_final, "Per-dataset RT windows - all confirmed features")
```

Chromatography labels come from the deposit / ReDU metadata and paper constants
(`01`), not from RT. Absolute RT is not comparable across datasets (it depends on
each method's gradient), so the spread within a phase facet is expected, not
evidence of mislabelling. Two datasets are RT outliers worth checking against their
source methods:

- **MTBLS11656** — confirmed RT ≈ 523 s, very late for RP (every other RP dataset
  sits at 88–203 s);
- **MSV000092754** — confirmed RT ≈ 107 s, unusually early for HILIC (every other
  HILIC dataset here sits at 279–507 s).

## MS2 trigger RT versus MS1 chromatographic apex

The confirmed retention time above is the RT of the MS2 scan that matched, i.e.
where data-dependent acquisition (DDA) fired. DDA triggers as soon as a precursor
is intense enough, so the MS2 scan can sit anywhere on the chromatographic peak
(rising edge, apex, or far down the tail) and scatters from file to file. The RT
that the `03` extraction windows must centre on is the MS1 chromatographic **apex**,
which is stable.

MSV000088255's metabolite makes the point. Its nine confirmed MS2 scans span 81 s
(256–337 s), which would force a 91 s extraction window, yet the real
5-hydroxyomeprazole peak sits in a 13 s band in every file. Each MS1 apex below was
read off the file's own EIC and **checked manually**:

| Confirmed MS2 RT (s) | MS1 apex (s) | DDA offset (s) |
|---:|---:|---:|
| 256 | 267 | +11 |
| 261 | 269 |  +8 |
| 265 | 270 |  +5 |
| 269 | 258 | −11 |
| 290 | 271 | −19 |
| 293 | 268 | −25 |
| 306 | 270 | −36 |
| 310 | 270 | −40 |
| 337 | 270 | −67 |

The DDA offset reaches 67 s (one scan fired on the tail while the 62-million-count
apex sat at 270 s), yet every apex lands within 258–271 s. `03` therefore anchors
each extraction window on the MS1 apex rather than the MS2 RT: for every confirmed
hit it finds the MS1 peak nearest the MS2 scan and takes that apex, then brackets
the robust consensus of those apexes. Hits where DDA fired too far for a reliable
automatic snap are corrected by hand in `curation/rt_apex_overrides.csv`,
audited against the per-hit EIC plots saved under `rt_snap/`.

# m/z deviation

The observed precursor m/z should match the inferred adduct's theoretical
m/z within the instrument's mass accuracy; deviation in ppm is
`(observed - theoretical) / theoretical * 1e6`. A well-calibrated Orbitrap is
within +/- 5 ppm; above +/- 10 ppm suggests a calibration issue or a
mis-assigned adduct.

```{r mz-deviation, fig.width = 9, fig.height = 4}
role_formula <- setNames(compound_spec$formula, compound_spec$role)

conf_final$mz_theo <- mapply(
  function(f, a) {
    if (is.na(f) || is.na(a)) return(NA_real_)
    mass2mz(calculateMass(f), adduct = a)[[1]]
  },
  role_formula[conf_final$role],
  conf_final$adduct_inferred,
  USE.NAMES = FALSE)

conf_final$mz_ppm_dev <- (conf_final$mz - conf_final$mz_theo) /
                         conf_final$mz_theo * 1e6

conf_final |>
  dplyr::filter(!is.na(mz_ppm_dev)) |>
  ggplot(aes(y = dataset_id, x = mz_ppm_dev, colour = adduct_inferred)) +
  geom_vline(xintercept = 0,        linetype = 2, colour = "grey40") +
  geom_vline(xintercept = c(-5, 5), linetype = 3, colour = "grey70") +
  geom_jitter(size = 2.5, height = 0.15, alpha = 0.85) +
  facet_grid(. ~ role) +
  labs(y = NULL, x = "m/z deviation (ppm)", colour = "Adduct",
       title = "Precursor m/z deviation vs theoretical adduct mass",
       caption = "Dashed: 0 ppm  ·  Dotted: +/- 5 ppm") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Per-dataset summary

```{r dataset-summary-table}
ds_summary <- conf_final |>
  dplyr::group_by(dataset_id, deposit_id, assay, chromatography) |>
  dplyr::summarise(
    repository      = dplyr::case_when(
      startsWith(deposit_id[1], "MSV")   ~ "MassIVE",
      startsWith(deposit_id[1], "MTBLS") ~ "MetaboLights",
      startsWith(deposit_id[1], "ST")    ~ "Metabolomics Workbench"),
    best_confidence = levels(conf_final$confidence)[
      min(CONF_ORDER[as.character(confidence)])],
    n_parent        = sum(role == "parent"),
    n_metabolite    = sum(role == "metabolite"),
    rt_parent_s     = round(median(rt[role == "parent"],     na.rm = TRUE), 1),
    rt_metabolite_s = round(median(rt[role == "metabolite"], na.rm = TRUE), 1),
    .groups         = "drop") |>
  dplyr::left_join(
    ds_meta_tbl[, intersect(c("dataset_id", "const_body_site", "const_country",
                              "const_location_type"),
                            names(ds_meta_tbl))],
    by = "dataset_id")

ds_summary
```

# Biological file selection

For each surviving `dataset_id` we build the list of biological MS files
(excluding QC injections, blanks, pooled QCs, and reference standards).
The three repositories carry this information differently:

- **MassIVE (GNPS / Pan-ReDU)**: per-deposit TSV from Pan-ReDU
  (auto-cached from `redu.gnps2.org`) or a submitter-provided CSV.
  Both carry a `SampleType` column and a `filename` column. The TSV
  has full relative paths; the submitter CSV has bare stems and is
  reconstructed against the MASST `file_path` directory + extension.
- **MetaboLights (EBI, ISA-Tab)**: the study file `s_{id}.txt` lists
  sample names and factors; the spectral file names live in the assay
  file `a_{id}_*.txt`. We use `mtbls_assay_data()`
  for one assay at a time (each `dataset_id` is already scoped to one
  assay) and restrict to sample rows that are biological per `s_*.txt`.
- **Metabolomics Workbench (mwTab)**: each analysis file
  `<deposit>_AN*.txt` carries the `SUBJECT_SAMPLE_FACTORS` block. Each
  subject's `RAW_FILE_NAME` is a space-separated list, one position per
  analysis in the deposit's canonical `MS_run` order. We use
  `mwb_metadata()$MS_run` to find the
  current analysis's position and pick the matching file from each
  subject.

`EXCL_ST` in `R/metadata.R` defines the regex of sample-type strings
treated as non-biological across all three loops.

```{r init-lists}
bio_files_per_ds  <- list()
bio_assays_per_ds <- list()
# Per file, does a metadata source cover it? Files without metadata are extracted
# for the record (only if they carry a confirmed hit) but excluded downstream.
bio_meta_per_ds   <- list()
```

## MassIVE

The extraction set comes from the **deposit's own file inventory**
(`massive_list_files` via `select_bio_files_massive()`), de-duplicated by
normalised name across all of the deposit's file trees. This matters because a
metadata catalogue alone under-covers a deposit: MSV000082493's 341-file blood
collection was added in a later deposit update that Pan-ReDU never indexed, and
MSV000082433's confirmed exposure sits on a plate Pan-ReDU never received.
Sourcing from the inventory keeps the extraction space equal to the MASST search
space.

A file enters the set if it is **either**:

- covered by a metadata source (Pan-ReDU ∪ submitter table, reconciled in `01`),
  with QC injections, blanks and reference standards removed by the *declared*
  sample-type column — never by file name (`EXCL_ST` on `SampleType` /
  `MS_SampleType`); **or**
- carrying a confirmed MASST hit, whatever tree it sits in, so no confirmed
  exposure is ever lost.

Files that are neither are not downloaded. `bio_meta_per_ds` flags each kept file:
`has_metadata = TRUE` is analysable downstream; a confirmed-hit-only file
(`FALSE`) is extracted for the presence record but excluded from stratified
analysis.

```{r bio-files-massive}
# Loadable basename of a confirmed-hit file (strip the MsBackend cache prefix).
bare_hit_name <- function(x) sub("^MSV[0-9]+[_-]", "", basename(as.character(x)))
msv_rows <- ds_summary[startsWith(ds_summary$deposit_id, "MSV"), ]

for (i in seq_len(nrow(msv_rows))) {
  ds  <- msv_rows$dataset_id[i]
  dep <- msv_rows$deposit_id[i]

  meta_keys <- tryCatch({
    nat <- msv_native_sample_table(dep)
    for (st in intersect(c("SampleType", "MS_SampleType"), names(nat)))
      nat <- nat[!grepl(EXCL_ST, nat[[st]], perl = TRUE) | is.na(nat[[st]]), ]
    fcol <- intersect(MSV_FILENAME_COLS, names(nat))[1]
    if (is.na(fcol)) character() else norm_file_key(nat[[fcol]])
  }, error = function(e) character())
  hit_files <- unique(bare_hit_name(
    conf_final$source_file[conf_final$dataset_id == ds &
                           !is.na(conf_final$source_file)]))

  sel <- tryCatch(select_bio_files_massive(dep, meta_keys, hit_files),
                  error = function(e) {
                    warning("[", ds, "] ", conditionMessage(e)); NULL })
  if (is.null(sel) || !nrow(sel)) {
    warning("[", ds, "] no biological files resolved - skipped"); next
  }

  bio_files_per_ds [[ds]] <- sel$file
  bio_meta_per_ds  [[ds]] <- stats::setNames(sel$has_metadata, sel$file)
  bio_assays_per_ds[[ds]] <- NA_character_   # MassIVE backend needs no assay name
  message(sprintf("[%s] %d files to extract (%d with metadata, %d record-only)",
                  ds, nrow(sel), sum(sel$has_metadata), sum(!sel$has_metadata)))
}
```

## MetaboLights

```{r bio-files-mtbls}
mtbls_rows <- ds_summary[startsWith(ds_summary$deposit_id, "MTBLS"), ]

for (i in seq_len(nrow(mtbls_rows))) {
  ds  <- mtbls_rows$dataset_id[i]
  dep <- mtbls_rows$deposit_id[i]
  an  <- mtbls_rows$assay[i]

  s_path <- file.path(META_DIR, paste0("s_", dep, ".txt"))
  if (!file.exists(s_path)) {
    warning("[", ds, "] no s_*.txt - skipped"); next
  }

  samples <- read.delim(s_path, check.names = FALSE)
  st_col  <- grep("sample type", names(samples), value = TRUE,
                  ignore.case = TRUE)[1]
  if (!is.na(st_col))
    samples <- samples[!grepl(EXCL_ST, samples[[st_col]], perl = TRUE), ]
  bio_names <- unique(samples[["Sample Name"]])

  adf <- mtbls_assay_data(dep, assayName = an)
  if (is.null(adf) || !nrow(adf)) {
    warning("[", ds, "] empty assay table - skipped"); next
  }
  fc <- if ("Derived Spectral Data File" %in% names(adf) &&
            any(nzchar(trimws(as.character(adf[["Derived Spectral Data File"]])))))
          "Derived Spectral Data File"
        else "Raw Spectral Data File"

  rows  <- adf[adf[["Sample Name"]] %in% bio_names, , drop = FALSE]
  files <- unique(trimws(as.character(rows[[fc]])))
  files <- files[!is.na(files) & nchar(files) > 0]

  bio_files_per_ds [[ds]] <- files
  # MetaboLights files come from the deposit's own ISA-Tab manifest, so every one
  # already carries metadata (QC excluded above): all TRUE.
  bio_meta_per_ds  [[ds]] <- stats::setNames(rep(TRUE, length(files)), files)
  bio_assays_per_ds[[ds]] <- an
}
```

## Metabolomics Workbench

```{r bio-files-mwb}
mwb_rows <- ds_summary[startsWith(ds_summary$deposit_id, "ST"), ]

for (i in seq_len(nrow(mwb_rows))) {
  ds  <- mwb_rows$dataset_id[i]
  dep <- mwb_rows$deposit_id[i]
  an_id <- sub(paste0("^", dep, "_"), "",
               sub("\\.txt$", "", mwb_rows$assay[i]))

  mwf <- file.path(META_DIR, mwb_rows$assay[i])
  if (!file.exists(mwf)) {
    warning("[", ds, "] no ", basename(mwf), " - skipped"); next
  }

  ln  <- readLines(mwf, warn = FALSE)
  ssf <- ln[startsWith(ln, "SUBJECT_SAMPLE_FACTORS") & !startsWith(ln, "#")]
  if (!length(ssf)) {
    warning("[", ds, "] no SUBJECT_SAMPLE_FACTORS - skipped"); next
  }

  ssf_parsed <- lapply(strsplit(ssf, "\t"), function(p) {
    raw_str   <- if (length(p) >= 5)
                   regmatches(p[5], regexpr("RAW_FILE_NAME=[^\t]+", p[5]))
                 else character(0)
    raw_str   <- sub("RAW_FILE_NAME=", "",
                     if (length(raw_str)) raw_str else "")
    raw_files <- trimws(unlist(strsplit(raw_str, "\\s+")))
    list(factors   = if (length(p) >= 4) p[4] else NA_character_,
         raw_files = raw_files[nchar(raw_files) > 0])
  })
  keep_qc    <- !vapply(ssf_parsed,
                        function(r) grepl(EXCL_ST, r$factors, perl = TRUE),
                        logical(1))
  ssf_parsed <- ssf_parsed[keep_qc]

  # RAW_FILE_NAME is a positional list, one entry per analysis in the deposit's
  # MS_run order, so we need this analysis's position in that order.
  ms_run <- mwb_metadata(dep)$MS_run
  an_idx <- match(an_id, as.character(ms_run$ANALYSIS_ID))
  if (is.na(an_idx))
    stop("[", ds, "] ", an_id, " not in the deposit's MS_run analysis list")

  files <- unique(vapply(ssf_parsed, function(r) {
    if (length(r$raw_files) >= an_idx) r$raw_files[an_idx] else NA_character_
  }, character(1)))
  files <- files[!is.na(files) & nchar(files) > 0]

  bio_files_per_ds [[ds]] <- files
  # Workbench files come from the deposit's own mwTab manifest (QC excluded), so
  # every one carries metadata: all TRUE.
  bio_meta_per_ds  [[ds]] <- stats::setNames(rep(TRUE, length(files)), files)
  bio_assays_per_ds[[ds]] <- an_id
}
```

## Summary

```{r bio-summary}
#| echo: false
bio_counts <- data.frame(
  dataset_id  = names(bio_files_per_ds),
  n_bio_files = vapply(bio_files_per_ds, length, integer(1)),
  assay       = vapply(names(bio_files_per_ds),
                       function(d) as.character(bio_assays_per_ds[[d]]),
                       character(1)),
  row.names   = NULL
) |>
  dplyr::left_join(ds_summary[, c("dataset_id", "deposit_id", "repository",
                                  "chromatography")],
                   by = "dataset_id")
bio_counts
```

# Write artifacts

What `03` needs: the confirmed features (`hits-confirmed-all.csv`), the
per-dataset summary (`datasets-confirmed.csv`), and the three per-dataset lists
that drive extraction — the biological files to load (`bio_files_per_ds.rds`),
their assay (`bio_assays_per_ds.rds`), and the has-metadata flag per file
(`bio_meta_per_ds.rds`).

```{r write-artifacts}
write.csv(conf_final, "artifacts/hits-confirmed-all.csv", row.names = FALSE)
write.csv(ds_summary, "artifacts/datasets-confirmed.csv", row.names = FALSE)
saveRDS(bio_files_per_ds,  "artifacts/bio_files_per_ds.rds")
saveRDS(bio_assays_per_ds, "artifacts/bio_assays_per_ds.rds")
saveRDS(bio_meta_per_ds,   "artifacts/bio_meta_per_ds.rds")
```

# Session info

```{r session-info}
sessionInfo()
```
