---
title: "03: All-datasets MS1 extraction"
subtitle: "Loop over curated datasets · per-dataset sample tables · cross-dataset summary"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
params:
  render_reports: false
---

# Overview

`01` and `02` established, per dataset, exactly where omeprazole and its CYP2C19
metabolite 5-hydroxyomeprazole sit in mass and retention time. This notebook
takes those coordinates back to every biological file in each dataset and
measures the two compounds directly: it pulls each compound's extracted-ion
chromatogram (EIC), detects and integrates the peak, and records a per-file
`log2(parent / metabolite)`. A high ratio means the drug is not being
metabolised and a low one means it is, and that read-out is what `04` and `05`
work with.

Put another way, `01` and `02` establish that a dataset *contains* the pair;
`03` propagates that detection to every file by MS1, well beyond the
MS2-confirmed subset, and quantifies it.

All compute goes through `run_ms1_for_dataset()` (`R/ms1.R`); this notebook
holds the loop, the extraction settings and the cross-dataset summary.
Single-dataset diagnostics -- EIC overlays, peak-width plots, PCA -- live in
`dataset-report.qmd`, which section 7 renders once per dataset. Datasets are
keyed by the combined `dataset_id` from `01`, and the loop pulls the bare
`deposit_id` and `assay` per row so repository downloads see plain accessions.

## Why not just use the MS2 matches?

A missing MS2 spectrum does not mean a missing compound. These deposits were
acquired by data-dependent acquisition, in which the instrument decides in real
time which ions to fragment, mostly by intensity and within a limited duty
cycle. A compound can therefore be present at a perfectly measurable level and
receive no MS2 scan at all, simply because the instrument was busy with
something else; run the same sample twice and the set of fragmented precursors
differs. A search resting on MS2 alone cannot separate absence from
non-selection, and the failure is not random. Selection is intensity-driven, so
the compounds most often missed are the low-abundance ones, which here means
exactly the incidentally-dosed subject we are trying to find. MS2-only
prevalence is a lower bound, and a biased one.

The two levels are complementary. MS2 gives identity, because fragments
constrain the structure and SIRIUS can tell isomers apart; MS1 gives coverage,
because it is recorded continuously with no selection step. Propagation uses
each for what it is good at: confirm the compound by MS2 in the files where it
happened to be fragmented, which pins both its mass and its retention time, then
measure that mass at that retention time in every file, including the many where
the instrument never fragmented it.

Proteomics met the same problem and solved it the same way. Match-between-runs
identifies a peptide by MS2 in the run where it was fragmented, then transfers
that identification to other runs by accurate mass and aligned retention time,
quantifying at MS1 where the precursor was never selected. It became standard
practice because DDA under-samples exactly as described above. Metabolomics
repository reanalysis has largely not adopted the equivalent step, and this
notebook is that step.

The trade is sensitivity for specificity, and it is real. MS2 can distinguish
compounds that share a mass and MS1 cannot, so propagation extends coverage only
as far as the chromatography preserves identity. That is not hypothetical here:
5-hydroxyomeprazole has an exact isobaric twin, and a retention-time window set
too wide silently measures the wrong molecule. The method is only as good as
that window, which is why so much of this notebook is about it.

There is a second benefit, less obvious. Comparing deposits by their MS2
detection rates would substantially be comparing instruments: a deposit run on a
faster instrument with a deeper precursor list fragments more, so it looks more
prevalent. MS1 is acquired whatever the instrument chose to fragment, so
propagating to MS1 removes that confound. It does not make deposits fully
comparable, since sensitivity, matrix and chromatography still differ, but it
removes the one that most distorts a raw detection-rate comparison.

```{r setup}
source("R/_setup.R")
source("R/metadata.R")
source("R/masst.R")   # compound_spec + PHENO, the domain constants
source("R/ms1.R")
library(Chromatograms)
library(MsQuality)
```

```{r load-artifacts}
conf_final        <- read.csv("artifacts/hits-confirmed-all.csv",
                              check.names = FALSE)
conf_final$confidence <- factor(conf_final$confidence,
                                 levels = c("strict", "standard",
                                            "liberal", "not_confirmed"))
bio_files_per_ds  <- readRDS("artifacts/bio_files_per_ds.rds")
bio_assays_per_ds <- readRDS("artifacts/bio_assays_per_ds.rds")
# Does a metadata source cover each file? Keyed dataset_id|basename so identical
# basenames in two deposits cannot collide.
bio_meta_per_ds   <- readRDS("artifacts/bio_meta_per_ds.rds")
has_metadata_lookup <- unlist(lapply(names(bio_meta_per_ds), function(ds) {
  v <- bio_meta_per_ds[[ds]]
  stats::setNames(as.logical(v), paste0(ds, "|", basename(names(v))))
}))
ds_meta_tbl       <- read.csv("artifacts/dataset_metadata_table.csv",
                              check.names = FALSE)
```

# Extraction settings

The extraction is governed by two numbers. Each says how tightly we look for the
compound's signal:

- **`PPM = 20`**: the mass tolerance, in parts-per-million. When we go looking
  for a compound at its known mass, real instruments are slightly off, so we
  accept anything within ±20 ppm (about ±0.007 Da at the omeprazole mass).
- **`RT_PAD_S = 5`**: seconds of retention-time padding added on each side of a
  dataset's confirmed-RT span. This one is load-bearing rather than cosmetic,
  for reasons the section below explains.

There is no peak-edge padding: the picker integrates valley to valley, so how
far it reaches is set by the peak's own shape rather than by a fixed margin.

```{r extraction-settings}
DATASETS_TO_RUN <- names(bio_files_per_ds)

PPM            <- 20   # mass tolerance in parts-per-million
RT_PAD_S       <- 5    # seconds padded on each side of the confirmed-RT span
```


## How the peaks are picked and filtered

The picker lives in `R/ms1.R` and runs in five stages, in this order.

- **Set the window** (`acceptance_window`). Rather than a fixed span, each
  confirmed MS2 hit is snapped to its nearest MS1 apex and the window brackets
  the robust consensus of those apexes, padded by `APEX_PAD` = 5 s. The window
  only has to *contain* each file's apex; how far the integration then reaches
  is set by the peak.
- **Integrate** (`pick`). Each local maximum is integrated valley to valley,
  walking out from the apex until the trace returns to baseline or meets a deep
  valley, merging shallow shoulders and sub-4 s doublets and stopping at dead
  gaps where the EIC dropped out. Area is trapezoidal above the local baseline;
  prominence, SNR and the nearest-taller-neighbour ratio are computed on the
  whole peak.
- **Choose which peak.** Where several fall inside the window, the dominant one
  is chosen by area weighted by a gaussian in its distance from the expected RT
  (`RT_SIGMA` = 4 s). This is the isomer defence at the selection level: a
  larger interferent elsewhere in the window loses to the on-RT peak. HILIC
  keeps pure area-selection, its windows having been audited peak by peak.
- **Accept or reject** (`accept_peak`), conditional on chromatography. Both
  modes require at least 3 points (HILIC 4), an apex of 500 counts, an SNR of 12
  on reverse phase or 3 on HILIC, an FWHM inside [1, 30] s on reverse phase or
  [1.3, ∞) s on HILIC, and a slope check. HILIC also requires return-to-baseline
  and no much-taller neighbour; reverse phase drops both, because they killed
  sharp truncated-flank peaks and real peaks sitting beside the sulfone, and
  instead requires a 3-point peak to clear its own baseline. The reverse-phase
  floors are corpus-set: real peaks there all clear SNR 20 and top out around 24
  s FWHM.
- **Measure** (`add_msquality`). The picker chooses the peak and its boundaries;
  **MsQuality** then measures it, so the reported width, FWHM and shape are mzQC
  controlled-vocabulary quantities rather than local re-implementations. Handed
  our boundaries, its `peakWidth()` and `xicFwhm()` reproduce the local values
  exactly on reverse phase; on HILIC the two FWHMs differ, because ours takes
  the outermost half-maximum crossings of a merged cluster while MsQuality takes
  50% of the maximum inside the region. `gaussianSimilarity()` is recorded per
  peak but never gated on, since it needs at least 5 points and is NA on the
  sparsest. Apex intensity and SNR stay local: both describe *which* peak was
  selected, and MsQuality has no way of being told that.

One further rejection happens after the picker, once both channels exist.
`resolve_isobars()` drops a peak sitting at the *other* compound's confirmed
retention time: both masses carry an isobar, and the parent's ---
hydroxyomeprazole sulfide --- yields omeprazole's own product ion, so it passes
spectral confirmation too. Where a channel's confirmed spectra split into two
retention clusters, nothing is asserted for that channel.

The gates were tuned by auditing peaks in one reverse-phase and one HILIC
dataset, each plotted and labelled by eye as signal, noise or interferent. Run
across the corpus they flagged 1,304 detected peaks carrying no MS2 confirmation
of their own, which were reviewed against the panels saved under
`peaks_review_v2/`. About twenty of those survive every rule while still being
wrong -- jagged stool and saliva traces where noise and signal overlap on every
shape metric -- and are removed by a manual drop-list
(`curation/ms1_drop_list.csv`) keyed by dataset, role and file, each entry
carrying a stated reason. For a handful of genuinely ambiguous peaks a
transparent per-peak removal is more honest than a gate no one can fully reason
about. This is inspection-backed calibration rather than a formal benchmark, but
every decision is auditable from the saved panels.

## The retention-time window, and the isomer it has to avoid

Retention time (RT) is *when* a compound comes off the chromatography column. We
only integrate signal inside an RT window, so the window has to sit over the
peak. `derive_extraction_config()` builds it per dataset from that dataset's own
SIRIUS-confirmed peaks: the window spans their RTs (earliest to latest), padded
each side by `RT_PAD_S`. The confirmed span already absorbs cross-file drift,
because it is the spread of MS2 detections across many files; the pad only has
to cover drift in files that carry no confirmed hit of their own.

*Why the isomer matters.* 5-hydroxyomeprazole is not the only thing at its mass.
Omeprazole sulfone, the compound CYP3A4 makes from omeprazole, has the identical
formula (`C17H19N3O4S`) and therefore the identical m/z (362.1169). MS1 cannot
tell the two apart. Only chromatography separates them, and the consequences of
losing that separation are severe: the sulfone reports CYP3A4 activity, not the
CYP2C19 activity we are trying to measure, and in a poor CYP2C19 metaboliser the
drug shunts to CYP3A4, so the sulfone is *larger* than the real 5-OH peak. The
proximity-weighted selection described above is the primary defence; the window
is a second, softer one, kept modest so the sulfone rarely even enters it.

The margin is narrower than it looks. In MSV000084008 the confirmed 5-OH peaks
span 3.4 s (154.9 to 158.3), while the sulfone sits at ~181 s. A ±5 s pad gives
[150, 163] and is safe. A ±30 s pad gives [125, 188], which reaches the sulfone.

*This is measured, not assumed.* That deposit's metadata records each subject's
genotyped CYP2C19 phenotype (`PharmGeno_Phenotype_2C19`), which the pipeline
never uses during extraction, so it is an independent target. Re-integrating the
same cached chromatograms of MSV000084008 at a range of metabolite
acceptance-window half-widths (parent fixed at ±5 s; metabolite centre ~156 s,
sulfone ~181 s) and correlating the subject-level AUC ratio against that
genotyped phenotype gives:

```{r}
#| echo: false
#| output: asis
#| label: tbl-pad-sweep
# Recomputed at render time from the cached wide EICs of MSV000084008: widen the
# metabolite acceptance window toward the sulfone, re-pick with and without the
# proximity weighting, and correlate the subject AUC ratio against genotype.
sweep_cache <- "peaks_review_v2/_cache/MSV000084008.rds"
sweep_meta  <- "dataset_metadata/MSV000084008_metadata.tsv"
if (file.exists(sweep_cache) && file.exists(sweep_meta)) {
  # PHENO (CYP2C19 activity order) comes from R/masst.R.
  cache <- readRDS(sweep_cache)
  ta <- read.delim(sweep_meta, check.names = FALSE, stringsAsFactors = FALSE)
  ta$tmin <- as.numeric(ta$ATTRIBUTE_Time_Point_Mins)
  ta <- ta[ta$MS_SampleType == "sample" & !is.na(ta$tmin), ]
  ta$k <- norm_file_key(ta$filename)
  fb   <- basename(as.character(bio_files_per_ds[["MSV000084008"]]))
  bare <- function(x) norm_file_key(to_bare_name(basename(as.character(x)), fb))
  ov_sweep <- if (file.exists("curation/rt_apex_overrides.csv"))
    read.csv("curation/rt_apex_overrides.csv", stringsAsFactors = FALSE) else
    data.frame(dataset_id = character(), role = character(),
               ms2_rt = numeric(), apex_rt = numeric())
  ctr  <- function(role, cr) mean(acceptance_window("MSV000084008", role, cr, bare,
                                                    cr$win, conf_final, ov_sweep))
  par_ctr <- ctr("parent", cache$par); met_ctr <- ctr("metabolite", cache$met)
  # hilic = TRUE gives pure area-select, so it doubles as the "no proximity"
  # control.
  auc_of <- function(cr, centre, W, proximity) {
    data.frame(k = norm_file_key(cr$files),
               auc = vapply(cr$traces, function(tr) {
                 p <- pick(tr$rt, tr$it, centre - W, centre + W, hilic = !proximity)
                 if (is.null(p) || !accept_peak(p, FALSE)) NA_real_ else p$auc
               }, numeric(1)))
  }
  trap <- function(x, y) { o <- order(x); x <- x[o]; y <- y[o]; y[is.na(y)] <- 0
    if (length(unique(x)) < 4) NA_real_ else sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2) }
  rho_of <- function(pa, ma) {
    t2 <- ta
    t2$par_auc <- pa$auc[match(t2$k, pa$k)]; t2$met_auc <- ma$auc[match(t2$k, ma$k)]
    a <- do.call(rbind, lapply(
      split(t2, list(t2$ATTRIBUTE_Subject, t2$ATTRIBUTE_Study_Day), drop = TRUE),
      function(d) data.frame(rank = match(d$PharmGeno_Phenotype_2C19[1], PHENO),
                             lr = log2(trap(d$tmin, d$par_auc) / trap(d$tmin, d$met_auc)),
                             subject = d$ATTRIBUTE_Subject[1])))
    a <- a[is.finite(a$lr), ]; s <- aggregate(lr ~ subject + rank, a, mean)
    cor.test(s$rank, s$lr, method = "spearman", exact = FALSE)$estimate
  }
  pa <- auc_of(cache$par, par_ctr, 5, TRUE)          # parent fixed at +/-5 s throughout
  sweep <- do.call(rbind, lapply(c(5, 10, 15, 20, 30), function(W) data.frame(
    `metabolite window` = sprintf("[%.0f, %.0f] (+/-%d s)", met_ctr - W, met_ctr + W, W),
    `reaches sulfone?`  = if ((met_ctr + W) >= 181) "yes" else "no",
    `with proximity`    = sprintf("%+.3f", rho_of(pa, auc_of(cache$met, met_ctr, W, TRUE))),
    `without proximity` = sprintf("%+.3f", rho_of(pa, auc_of(cache$met, met_ctr, W, FALSE))),
    check.names = FALSE)))
  cat(knitr::kable(sweep, align = "lccc", row.names = FALSE), sep = "\n")
} else {
  cat("_Pad sweep skipped: needs `", sweep_cache, "` (git-ignored). Run ",
      "`Rscript R/peak_review_panels.R MSV000084008` to rebuild it._\n", sep = "")
}
```

With the proximity weighting the correlation is unchanged at every width;
without it, the ratio collapses once the window admits the sulfone. So the
isomer defence is the on-RT selection, not a tight window; the ±5 s pad only
buys drift tolerance.

# Per-dataset extraction

For each dataset, `run_ms1_for_dataset()` loads the MS1 spectra, pulls each
compound's extracted-ion chromatogram (EIC) at the confirmed mass and (widened)
retention-time window, then **picks and integrates the on-RT peak per file**
(the picker described above), applies the manual drop-list, flags files whose
peak width is an outlier, joins the sample metadata, and emits a per-file table
(`sample_tbl`) carrying the `detection_tier` (which of the two compounds were
seen) and the `log2_ratio`. The deep per-dataset diagnostics (EIC overlays,
peak-width plots, principal-component analysis) stay in `dataset-report.qmd`;
this notebook reports only what generalises across the whole cohort.

```{r run-all}
#| echo: false
# Cached per dataset; delete a dataset's RDS to force a re-run. The cache is
# keyed on the extraction window, not merely on the file existing: re-running
# confirmation moves the confirmed RTs and therefore the window, and a
# file.exists() check would serve detections picked from the old one. Deriving
# the config extracts nothing, so it is cheap to verify on every render.
CACHE_DIR <- "artifacts/per_dataset"
dir.create(CACHE_DIR, showWarnings = FALSE, recursive = TRUE)

cfg_is_current <- function(cached, ds) {
  want <- tryCatch(derive_extraction_config(ds, conf_final, rt_pad_s = RT_PAD_S,
                                            bio_files = bio_files_per_ds[[ds]]),
                   error = function(e) NULL)
  if (is.null(want) || is.null(cached$cfg)) return(FALSE)
  k <- c("assay", "adduct_par", "adduct_met", "mz_par", "mz_met",
         "rt_par_win", "rt_met_win")
  isTRUE(all.equal(cached$cfg[k], want[k], tolerance = 1e-6))
}

results <- list()
SKIPPED_DATASETS <- character(0)
for (ds in DATASETS_TO_RUN) {
  cache_path <- file.path(CACHE_DIR, paste0(ds, ".rds"))
  cached <- if (file.exists(cache_path)) readRDS(cache_path) else NULL
  if (!is.null(cached) && !cfg_is_current(cached, ds)) {
    message("  cache for ", ds, " was built on a different extraction window; re-extracting")
    cached <- NULL
  }
  if (!is.null(cached)) {
    message("========== ", ds, " (cached) ==========")
    results[[ds]] <- cached
  } else {
    message("\n========== ", ds, " ==========")
    # Tolerate a dataset whose raw files can't be fetched (e.g. the MassIVE
    # FTP is rate-limited / down): skip it with a logged reason and continue,
    # rather than aborting the whole cross-dataset run. Skipped datasets are
    # simply absent from `results` and get re-extracted on a later render once
    # the source is reachable (their per-dataset cache is only written on
    # success). SKIPPED_DATASETS records what was dropped.
    res <- tryCatch(
      run_ms1_for_dataset(
        dataset_id       = ds,
        conf_final       = conf_final,
        bio_files_per_ds = bio_files_per_ds,
        ds_meta_tbl      = ds_meta_tbl,
        ppm              = PPM,
        rt_pad_s         = RT_PAD_S),
      error = function(e) {
        message("  SKIPPED ", ds, ", extraction failed: ", conditionMessage(e))
        NULL
      })
    if (is.null(res)) { SKIPPED_DATASETS <<- c(SKIPPED_DATASETS, ds); next }
    results[[ds]] <- res
    saveRDS(results[[ds]], cache_path)
  }
}
if (length(SKIPPED_DATASETS))
  message("\n", length(SKIPPED_DATASETS),
          " dataset(s) SKIPPED (raw files unreachable; re-render once the ",
          "source is back): ", paste(SKIPPED_DATASETS, collapse = ", "))
```

## Extraction runtime

```{r}
#| echo: false
# Timing log written by R/peak_review_panels.R; git-ignored, so tolerate its
# absence.
et_path <- "peaks_review_v2/_extract_times.csv"
et <- if (file.exists(et_path)) {
  x <- read.csv(et_path, stringsAsFactors = FALSE)
  x <- x[order(x$extract_seconds), ]
  x$minutes    <- round(x$extract_seconds / 60, 1)
  x$s_per_file <- round(x$extract_seconds / x$n_files, 2)
  x
} else NULL
et_hours <- if (is.null(et)) NA else round(sum(et$extract_seconds) / 3600, 1)
et_files <- if (is.null(et)) NA else format(sum(et$n_files), big.mark = ",")
chr_ver  <- as.character(utils::packageVersion("Chromatograms"))
```

Loading each file's spectra and pulling its two EICs is the pipeline's dominant
compute cost, and it runs single-threaded. That is measured rather than assumed:
parallel backends gave no gain once the files are read, because the cost is disk
I/O rather than CPU. Cost tracks file count and file size, not the number of
detections, and the timings are compute-only on cached files (raw-file FTP
download is excluded). The corpus takes about **`r et_hours` hours** over
`r if (is.null(et)) NA else nrow(et)` datasets (`r et_files` files).

These numbers were measured with **Chromatograms `r chr_ver`**. Extraction
performance is strongly version-dependent, so treat the table as valid only for
that version and re-measure after any upgrade of the extraction stack.

```{r}
#| echo: false
#| output: asis
if (is.null(et)) {
  cat("_Timing table skipped: `", et_path, "` is not present (git-ignored). Run ",
      "`Rscript R/peak_review_panels.R` to regenerate it._\n", sep = "")
} else {
  cat(knitr::kable(et[, c("dataset", "n_files", "minutes", "s_per_file")],
                   col.names = c("dataset", "files", "min", "s / file"),
                   align = "lrrr", row.names = FALSE), sep = "\n")
}
```

# Checks that can catch us being wrong

A pipeline is only as trustworthy as the things that could contradict it, and
plausible numbers are not evidence if nothing in the design would fail when they
are wrong. Six checks are built in on purpose, in increasing order of how hard
they are to argue with.

**1. The confirmed-file positive control.** Where MS2 has already proved the
compound is present, MS1 extraction must find it; if it does not, propagation is
losing real signal and every prevalence number is an underestimate.
`run_ms1_for_dataset()` reports this as `Confirmed files loaded N / with >=1
peak M`, where M should equal N and any shortfall is a bug rather than biology.
It is the cheapest hard check here and needs no external data.

**2. A known-negative control.** The complement: run the identical pipeline on
files that *cannot* contain the compound and confirm it finds nothing. The
pre-dose plasma draws of the probe study are exactly that --- real matrix, true
m/z, true retention window, drug absent by design --- and `05` reports them. A
synthetic decoy mass is deliberately not used: an untargeted file carries real
compounds at arbitrary masses, so a hit at a decoy mass is usually the picker
correctly finding something else, which bounds the false-positive rate from
above without estimating it.

**3. Report the surface, not the best cell.** Every parameter here is a choice,
and a conclusion that survives only one setting is not a conclusion. The
retention-time sweep above is the pattern to follow: a broad plateau where the
answer is stable, and a cliff with a mechanism that explains exactly where and
why it breaks. A single lucky optimum would not be evidence.

**4. External ground truth, which beats every internal check.** Parameters can
be tuned until an internal check passes; an outside number cannot be. `04` uses
the one deposit where truth exists independently of us: a cohort in which every
subject was deliberately dosed, with pre-dose samples that are known negatives
and a published pharmacokinetic value per subject the pipeline never sees.

**5. The apex-snap shown rather than asserted.** The retention time a confirmed
hit carries is when the instrument fired, not when the compound eluted, and the
whole window design rests on the second being stable where the first is not.
MSV000088255 is where that is easiest to see, so it is plotted: every confirmed
metabolite scan in the deposit, on the EIC of the file it fired in.

```{r fig-dda-offset, fig.width = 9, fig.height = 5.6}
#| echo: false
.r <- results[["MSV000088255"]]
.h <- conf_final[conf_final$dataset_id == "MSV000088255" &
                 conf_final$role == "metabolite", ]
.h$bn <- basename(.h$source_file)
.org <- basename(chartr(rawToChar(as.raw(92)), "/",
                        chromData(.r$chr_met)$dataOrigin))
.ix  <- which(.org %in% unique(.h$bn))

.tr <- dplyr::bind_rows(lapply(.ix, function(i) data.frame(
  file = sub("^MSV[0-9]+_Urine_", "", sub("_FragPos.mzML$", "", .org[i])),
  rt = rtime(.r$chr_met[i])[[1]], int = intensity(.r$chr_met[i])[[1]])))
.tr <- .tr[is.finite(.tr$int), ]
.ap <- .tr |> dplyr::group_by(file) |>
  dplyr::slice_max(int, n = 1, with_ties = FALSE) |> dplyr::ungroup()
.ms2 <- .h |> dplyr::transmute(
  file = sub("^MSV[0-9]+_Urine_", "", sub("_FragPos.mzML$", "", bn)), ms2 = rt)

ggplot(.tr, aes(rt, int)) +
  geom_line(colour = "grey35", linewidth = 0.4) +
  geom_vline(data = .ms2, aes(xintercept = ms2), colour = "#D6604D",
             linewidth = 0.4, alpha = 0.85) +
  geom_point(data = .ap, aes(rt, int), colour = "#2166AC", size = 1.8) +
  facet_wrap(~ file, scales = "free_y", ncol = 5) +
  coord_cartesian(xlim = c(230, 350)) +   # zoom, not a filter: nothing is dropped
  labs(x = "Retention time (s)", y = "Intensity",
       title = "MSV000088255 metabolite: MS2 trigger times (red) against the MS1 apex (blue)",
       subtitle = "one panel per file carrying a confirmed metabolite scan") +
  theme_bw(base_size = 10) +
  theme(strip.background = element_rect(fill = "grey92"),
        panel.grid.minor = element_blank())
```

```{r dda-offset-numbers}
#| echo: false
.j <- dplyr::left_join(.ms2, dplyr::select(.ap, file, apex = rt), by = "file")
cat("files with a confirmed metabolite scan:", nrow(.ap),
    "| confirmed scans:", nrow(.j), "
")
cat("apex span:", sprintf("%.1f-%.1f s (%.1f s)", min(.ap$rt), max(.ap$rt),
                          diff(range(.ap$rt))), "
")
cat("trigger span:", sprintf("%.1f-%.1f s (%.1f s)", min(.j$ms2), max(.j$ms2),
                             diff(range(.j$ms2))), "
")
cat("trigger minus apex:", sprintf("%+.1f to %+.1f s", min(.j$ms2 - .j$apex),
                                   max(.j$ms2 - .j$apex)), "
")
```

The apexes sit in a band a fraction of the width of the triggers, and the
largest offsets are not scattered files but single intense peaks re-fragmented
on the way down: one file contributes four confirmed scans, spread over fifty
seconds, all belonging to one apex. Centring a window on the trigger times would
force it to span the union of those tails; centring on the apexes does not. This
is what the window derivation is for, and the figure is the check that it is
warranted.

**6. Which of the two isobars is actually in the parent channel.** The window
argument above says the compounds are chromatographically resolved; this asks
the harder question of whether the peak we integrate is the right molecule.
Omeprazole is a sulfoxide, so it cannot lose $\cdot$SH; hydroxyomeprazole
sulfide, which shares its formula, gives m/z $313.1422$ on exactly that loss.
The share of confirmed parent MS2 carrying that ion therefore reads directly on
how much of the channel is the isobar, and the probe cohort supplies its own
control: the cluster where omeprazole is confirmed should be near zero.

```{r isobar-fragment}
#| echo: false
.ifc <- "artifacts/isobar_fragment_check.csv"
# The only computation in this notebook that reads raw MS2 rather than the
# cached EICs, so it takes about ten minutes and its result is cached beside the
# other artifacts. Delete that CSV to force a recompute; nothing here notices on
# its own if the confirmed set changes underneath it.
if (!file.exists(.ifc)) {
  SULFIDE_MZ <- 313.1422   # [M+H-SH]+, impossible for a sulfoxide
  SHARED_MZ  <- 198.0581   # substituted pyridine ring, given by both
  TOL <- 0.02; PREC_TOL <- 0.05
  .pc <- function(x) if (!length(x)) NA_real_ else round(100 * mean(x))
  .rows <- list()
  for (ds in names(results)) {
    r <- results[[ds]]
    # only files carrying a confirmed parent hit: reading the whole corpus
    # costs hours and the question is what the CONFIRMED spectra are
    want <- unique(basename(as.character(conf_final$source_file[
      conf_final$dataset_id == ds & conf_final$role == "parent"])))
    if (!length(want)) next
    fs <- unique(chartr(rawToChar(as.raw(92)), "/", chromData(r$chr_par)$dataOrigin))
    fs <- fs[basename(fs) %in% want & file.exists(fs)]
    if (!length(fs)) next
    sp <- tryCatch(filterMsLevel(Spectra(fs), 2L), error = function(e) NULL)
    if (is.null(sp) || !length(sp)) next
    pmz <- precursorMz(sp)
    k <- which(!is.na(pmz) & abs(pmz - compound_spec$mz[compound_spec$role == "parent"]) < PREC_TOL)
    if (!length(k)) next
    sp <- sp[k]; pk <- peaksData(sp); rt <- rtime(sp)
    hasfrag <- function(t) vapply(pk, function(m) any(abs(m[, 1] - t) < TOL), logical(1))
    s313 <- hasfrag(SULFIDE_MZ)
    # the same two centres, and the same stand-down condition, resolve_isobars() uses
    ctr <- function(role) {
      q <- as.numeric(conf_final$rt[conf_final$dataset_id == ds & conf_final$role == role])
      q <- q[is.finite(q)]; if (length(q)) stats::median(q) else NA_real_
    }
    cp <- ctr("parent"); cm <- ctr("metabolite")
    ok <- !isTRUE(r$rt_anchor$par$uncertain) && !isTRUE(r$rt_anchor$met$uncertain) &&
      is.finite(cp) && is.finite(cm) && abs(cp - cm) >= CHAN_SEP
    own <- if (ok) abs(rt - cp) < abs(rt - cm) else rep(NA, length(rt))
    .rows[[length(.rows) + 1]] <- data.frame(
      dataset_id = ds, spectra = length(pk),
      pct_sulfide_313 = .pc(s313), pct_shared_198 = .pc(hasfrag(SHARED_MZ)),
      resolved = ok,
      n_integrated = if (ok) sum(own) else NA_integer_,
      pct_sulfide_integrated = if (ok) .pc(s313[own]) else NA_real_,
      n_rejected = if (ok) sum(!own) else NA_integer_,
      pct_sulfide_rejected = if (ok) .pc(s313[!own]) else NA_real_,
      stringsAsFactors = FALSE)
  }
  write.csv(do.call(rbind, .rows), .ifc, row.names = FALSE)
}
.z <- if (file.exists(.ifc)) read.csv(.ifc, stringsAsFactors = FALSE)
# One cell out of the table, so no number in the prose below is typed twice.
.iso <- function(ds, col) if (is.null(.z)) "--" else .z[[col]][.z$dataset_id == ds]
{
  .z$verdict <- ifelse(.z$pct_sulfide_313 >= 40, "largely the isobar",
                ifelse(.z$pct_sulfide_313 >= 15, "mixed", "mostly omeprazole"))
  .show <- .z[order(-.z$pct_sulfide_313),
              c("dataset_id", "spectra", "pct_sulfide_313",
                "pct_sulfide_integrated", "pct_sulfide_rejected", "verdict")]
  knitr::kable(.show, row.names = FALSE,
    col.names = c("dataset", "MS2 at the parent mass", "% with 313.14 (-SH)",
                  "integrated cluster", "rejected cluster", "verdict"),
    caption = paste("Sulfide-diagnostic fragment among confirmed parent spectra.",
                    "The last two columns split that rate by retention cluster,",
                    "classified by the same comparison resolve_isobars() makes;",
                    "they are empty where the anchors are too close for the",
                    "rule to apply."))
}
```

Read the rates as a mixture, not a verdict on a whole deposit. The probe cohort
sits at `r .iso("MSV000084008", "pct_sulfide_313")` per cent pooled, which on its
own says little; split by retention cluster it is
`r .iso("MSV000084008", "pct_sulfide_integrated")` per cent in the cluster the
extraction integrates against
`r .iso("MSV000084008", "pct_sulfide_rejected")` per cent in the one it rejects.
That contrast is the whole case for `resolve_isobars()`: the rule is not
discarding signal, it is discarding a different molecule. Two deposits are worth
naming: MSV000088255, a urine cohort, runs at
`r .iso("MSV000088255", "pct_sulfide_313")` per cent and its parent channel is
dropped; MTBLS1866 runs at
`r .iso("MTBLS1866_LC-MS_positive_reverse-phase_metabolite_profiling", "pct_sulfide_313")`
per cent and is **not** gated, because its two anchors sit under $6$ s apart and
the rule stands down where the channels are not resolved -- which is why its row
has no split. Its parent contrast should be read with that attached.

A methodological warning that applies to all of this: **do not choose a
parameter by maximising agreement with the thing you then cite as validation.**
That is circular, and it is how the earlier over-wide window justified itself.
Where a parameter must be tuned, tune it against something structurally
independent of the validation target. The parent compound is useful here
precisely because it has no isobaric twin, so its retention-time drift can be
measured without reference to any downstream result.

**7. What each choice is actually worth.** The warning above is a rule for
choosing parameters; this is the audit of the ones already chosen. Every
decision in the extraction is defensible in prose, which is exactly the problem:
a reader has no way to tell a defensible choice from a convenient one. So each
is turned off in turn, one at a time with everything else left alone, and the
genotype model is refitted from the peaks that survive. MSV000084008 is the only
deposit carrying genotypes, so it is the only place the coefficient can be
recomputed; MSV000082493 supplies the counter-check, its skin samples taken
before the first dose being the one known-negative the intensity floor is
answerable to.

```{r ablations}
#| echo: false
rt_ov   <- read.csv("curation/rt_apex_overrides.csv", stringsAsFactors = FALSE)
drop_all <- read.csv("curation/ms1_drop_list.csv", stringsAsFactors = FALSE)
hits_curated <- read.csv("artifacts/hits_curated.csv", check.names = FALSE,
                         stringsAsFactors = FALSE)

E84 <- ablation_setup("MSV000084008", bio_files_per_ds[["MSV000084008"]])
E82 <- ablation_setup("MSV000082493", bio_files_per_ds[["MSV000082493"]])
vary <- function(e, conf = conf_final, ...)
  ablation_variant(e, conf = conf, ov = rt_ov, drop_all = drop_all, ...)

# The genotype refit: one trapezoidal AUC per subject per day, averaged over
# days, regressed on the phenotype rank. The same path 04 takes.
ta84 <- read.delim("dataset_metadata/MSV000084008_metadata.tsv",
                   check.names = FALSE, stringsAsFactors = FALSE)
ta84$k <- norm_file_key(ta84$filename)
ta84$tmin <- as_num(ta84$ATTRIBUTE_Time_Point_Mins)
ta84 <- ta84[ta84$MS_SampleType == "sample" & !is.na(ta84$tmin), ]

trap <- function(x, y) {
  o <- order(x); x <- x[o]; y <- y[o]; y[is.na(y)] <- 0
  if (length(unique(x)) < 4) NA_real_ else
    sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)
}

refit <- function(v) {
  d <- ta84
  mp <- v$par; mm <- v$met
  mp$k <- norm_file_key(mp$file); mm$k <- norm_file_key(mm$file)
  i <- match(d$k, mp$k); j <- match(d$k, mm$k)
  d$par_auc <- ifelse(mp$par_detected[i] %in% TRUE, mp$par_auc[i], NA_real_)
  d$met_auc <- ifelse(mm$met_detected[j] %in% TRUE, mm$met_auc[j], NA_real_)
  auc <- do.call(rbind, lapply(
    split(d, list(d$ATTRIBUTE_Subject, d$ATTRIBUTE_Study_Day), drop = TRUE),
    function(z) data.frame(
      subject = z$ATTRIBUTE_Subject[1],
      rank = match(z$PharmGeno_Phenotype_2C19[1], PHENO),
      log2_ratio = log2(trap(z$tmin, z$par_auc) / trap(z$tmin, z$met_auc)))))
  auc <- auc[is.finite(auc$log2_ratio) & !is.na(auc$rank), ]
  s <- aggregate(log2_ratio ~ subject + rank, data = auc, FUN = mean)
  co <- summary(lm(log2_ratio ~ rank, data = s))$coefficients
  c(beta = co[2, 1], se = co[2, 2], p = co[2, 4], n = nrow(s))
}

# Skin sampled before the first dose: the drug had not been taken and skin has
# no route to it, so anything called here is a false positive.
sk <- read.delim("dataset_metadata/MSV000082493_metadata.tsv",
                 check.names = FALSE, stringsAsFactors = FALSE)
sk <- sk[sk$MS_SampleType == "sample" & sk$ATTRIBUTE_Body_Site == "L.forehead" &
           (sk$ATTRIBUTE_Study_Day %in% "0" |
            (sk$ATTRIBUTE_Study_Day %in% "1" &
             sk$ATTRIBUTE_Time_Point_Mins %in% "0")), ]
sk$k <- norm_file_key(sk$filename)
skin_fp <- function(v) {
  mp <- v$par; mp$k <- norm_file_key(mp$file)
  sum(mp$par_detected[match(sk$k, mp$k)] %in% TRUE)
}

rows <- list()
add <- function(choice, alternative, note, v84, v82) {
  b <- refit(v84)
  rows[[length(rows) + 1]] <<- data.frame(
    choice = choice, alternative = alternative,
    detections = as.integer(sum(v84$par$par_detected) + sum(v84$met$met_detected)),
    note = note, skin_fp = as.integer(skin_fp(v82)),
    beta = round(unname(b["beta"]), 3), se = round(unname(b["se"]), 3),
    p = signif(unname(b["p"]), 2), subjects = as.integer(b["n"]),
    stringsAsFactors = FALSE)
}

R84 <- vary(E84); R82 <- vary(E82)
stopifnot(identical(R84$par$par_detected, E84$metrics$par$par_detected),
          identical(R84$met$met_detected, E84$metrics$met$met_detected),
          identical(R82$par$par_detected, E82$metrics$par$par_detected))
V0 <- ablation_isobars(R84, conf_final); W0 <- ablation_isobars(R82, conf_final)
add("pipeline as reported", "--", "", V0, W0)

for (f in c(100, 250, 1000))
  add(sprintf("apex floor %d counts", MIN_INT), sprintf("floor %d counts", f), "",
      ablation_isobars(vary(E84, floor_val = f), conf_final),
      ablation_isobars(vary(E82, floor_val = f), conf_final))

for (tiers in list("strict", c("strict", "standard"))) {
  cf <- conf_final[conf_final$confidence %in% tiers, ]
  add("anchors at any confidence tier", sprintf("%s only", paste(tiers, collapse = "+")),
      sprintf("%d of %d anchors", nrow(cf), nrow(conf_final)),
      ablation_isobars(vary(E84, conf = cf), cf),
      ablation_isobars(vary(E82, conf = cf), cf))
}

.vp <- ablation_isobars(vary(E84, per_hit = TRUE), conf_final)
add("one RT anchor per file", "one anchor per confirmed hit",
    sprintf("parent window %.0f-%.0f s, not %.0f-%.0f",
            .vp$win$par$win[1], .vp$win$par$win[2],
            V0$win$par$win[1], V0$win$par$win[2]),
    .vp, ablation_isobars(vary(E82, per_hit = TRUE), conf_final))

add("reject cross-channel isobars", "keep every in-window peak", "", R84, R82)

# Adduct policy acts upstream of the window, on which MASST hits exist at all,
# so it has no separate slope: this deposit's own hits are protonated either way.
.ok <- hits_curated$adduct_inferred %in% "[M+H]+"
rows[[length(rows) + 1]] <- data.frame(
  choice = "any plausible adduct", alternative = "protonated only",
  detections = NA_integer_,
  note = sprintf("%d of %d curated hits, %d of %d deposits", sum(.ok),
                 nrow(hits_curated), length(unique(hits_curated$dataset_id[.ok])),
                 length(unique(hits_curated$dataset_id))),
  skin_fp = NA_integer_, beta = NA_real_, se = NA_real_, p = NA_real_,
  subjects = NA_integer_, stringsAsFactors = FALSE)

.a <- do.call(rbind, rows)

# Pull one cell out for the prose below, so no number here is typed twice.
.ab <- function(alt, col) .a[[col]][.a$alternative == alt]
.bmax <- sprintf("%.3f", max(abs(.a$beta - .a$beta[1]), na.rm = TRUE))
.pmax <- sprintf("%.5f", max(.a$p, na.rm = TRUE))
.se0  <- sprintf("%.3f", .a$se[1])
.nsk  <- nrow(sk)

.tbl <- .a
# p is formatted here rather than left to kable: `digits` would round every one
# of them to 0.001 and the column would say nothing.
.tbl$p <- ifelse(is.na(.tbl$p), "", sprintf("%.5f", .tbl$p))
knitr::kable(.tbl, row.names = FALSE, digits = 3,
  col.names = c("choice", "alternative", "detections", "note",
                "pre-dose skin", "slope", "SE", "p", "subjects"),
  caption = paste("Each row changes one decision and refits. Detections are",
                  "MSV000084008's; pre-dose skin counts false positives among",
                  .nsk, "known negatives in MSV000082493."))
```

The first row is the pipeline as reported, and it reproduces the cached
detection flags exactly -- the chunk asserts that before it computes anything
else, so a variant that differs does so because of the change and not because
the re-run drifted from the pipeline. Across every other row the slope stays
within `r .bmax` of it, against a standard error on that slope of `r .se0`; the
p-value never rises above `r .pmax`, and all fourteen subjects survive every
variant. Nothing here is load-bearing for the
validation, which is the point: the correlation is not an artefact of the
processing, because the processing can be varied and it does not go away.

The intensity floor is the one row that is not symmetric. Raising it costs
plasma detections and buys nothing, while lowering it to $100$ admits
`r .ab("floor 100 counts", "skin_fp")` pre-dose skin samples that cannot
contain the drug. That asymmetry, not a tuned optimum, is why the floor sits
where it does, and the per-matrix breakdown says where its cost falls.

```{r floor-by-matrix}
#| echo: false
mb <- read.delim("dataset_metadata/MSV000082493_metadata.tsv",
                 check.names = FALSE, stringsAsFactors = FALSE)
mb <- mb[mb$MS_SampleType == "sample", ]
mb$k <- norm_file_key(mb$filename)
.mp <- R82$par; .mp$k <- norm_file_key(.mp$file)
.lo <- vary(E82, floor_val = 100)$par
.fb <- do.call(rbind, lapply(split(mb, mb$ATTRIBUTE_Body_Site), function(g) {
  i <- match(g$k, .mp$k); i <- i[!is.na(i)]
  data.frame(body_site = g$ATTRIBUTE_Body_Site[1], files = nrow(g),
             with_peak_in_window = length(i),
             accepted_at_500 = sum(.mp$par_detected[i]),
             accepted_at_100 = sum(.lo$par_detected[i]),
             stringsAsFactors = FALSE)
}))
.fb$rejected_at_500 <- .fb$with_peak_in_window - .fb$accepted_at_500
.fb$floor_only <- .fb$accepted_at_100 - .fb$accepted_at_500
.fm <- function(site, col) .fb[[col]][.fb$body_site == site]
knitr::kable(.fb, row.names = FALSE,
  col.names = c("body site", "files", "peak in window", "accepted at 500",
                "accepted at 100", "rejected at 500", "floor-only"),
  caption = paste("MSV000082493, parent channel, before the isobar rule.",
                  "Floor-only rejections are those a lower threshold would",
                  "rescue; the rest fail another criterion and cannot be."))
```

The gate is a conjunction, so a peak failing any criterion other than the apex
height cannot be rescued by a lower threshold, and the split matters. In faeces
only `r .fm("fecal", "floor_only")` of `r .fm("fecal", "rejected_at_500")`
rejections are the floor's doing: most of that matrix's shortfall is peaks that
do not have the shape of a peak, and lowering the threshold would not recover
them. In skin the proportions invert, `r .fm("L.forehead", "floor_only")` of
`r .fm("L.forehead", "rejected_at_500")`, so the same change admits that many
detections in the one matrix with no plausible route of exposure. The prevalence
figures downstream are lower bounds, most conspicuously in faeces and on skin;
no known-negative exists for those matrices to justify a lower floor, and the
one that does exist argues against it.

# Cross-dataset sample table

```{r combine-tables}
#| echo: false
all_samples <- do.call(rbind, lapply(names(results), function(ds) {
  # Rebuild rather than trust the cached sample_tbl: it is fully derivable from
  # metrics_* and costs nothing, so a cache written by an older
  # build_sample_table cannot carry a stale detection_tier into the corpus
  # tables. (It did: an exact-name merge silently demoted 19 detected
  # MetaboLights files to "neither".)
  meta <- tryCatch(load_metadata(ds),
                   error = function(e) data.frame(file = basename(bio_files_per_ds[[ds]])))
  .m <- resolve_isobars(results[[ds]]$metrics_par, results[[ds]]$metrics_met,
                        conf_final, ds, results[[ds]]$rt_anchor)
  st <- build_sample_table(.m$par, .m$met,
                           basename(bio_files_per_ds[[ds]]), meta, dedup = TRUE)
  # Keep only files still in the current bio_files. A cached extraction can hold
  # rows for files a later QC rule removed (MSV000091520 kept 247 blank_QC files
  # after bio_files went 2391 -> 2144), and without this filter the two sample
  # tables disagree. All of them are detection_tier "neither", so nothing
  # measured is dropped.
  keep <- norm_file_key(st$file) %in% norm_file_key(bio_files_per_ds[[ds]])
  cbind(dataset_id = ds, assay = results[[ds]]$assay, st[keep, , drop = FALSE],
        stringsAsFactors = FALSE)
}))

# Files with no metadata are kept for the presence record but excluded from
# every stratified analysis below.
all_samples$has_metadata <- {
  key <- paste0(all_samples$dataset_id, "|",
                basename(as.character(all_samples$file)))
  hit <- has_metadata_lookup[key]
  ifelse(is.na(hit), FALSE, unname(hit))
}

write.csv(all_samples, "artifacts/sample_table_all.csv", row.names = FALSE)

cat("Datasets processed :", length(results), "\n")
cat("Total rows         :", nrow(all_samples),
    "  (with metadata:", sum(all_samples$has_metadata),
    "| record-only:", sum(!all_samples$has_metadata), ")\n")
```

```{r}
#| echo: false
#| label: tbl-tier-by-dataset
#| tbl-cap: "Detection tier by dataset (metadata-carrying samples only). Assay suffixes are trimmed so the table fits."
short_ds <- function(x) sub("_AN[0-9]+$", "", sub("_LC-MS_.*$", "", x))
ms <- all_samples[all_samples$has_metadata %in% TRUE, ]
tier_tab <- as.data.frame.matrix(table(short_ds(ms$dataset_id), ms$detection_tier))
tier_tab$total <- rowSums(tier_tab)
knitr::kable(tier_tab, align = "r")
```

Every analysis after this point restricts to `has_metadata == TRUE`. The
record-only rows (a confirmed hit on a file the metadata layer does not cover,
so we cannot say which subject or group it belongs to) stay in
`sample_table_all.csv` for the presence record and the propagation counts below,
but they carry no stratum and never enter a contrast.

```{r analysis-set}
#| echo: false
analysis_samples <- all_samples[all_samples$has_metadata %in% TRUE, ]
```

`sample_table_all.csv` is one row per subject (the `build_sample_table()`
dedup). Genotype / longitudinal analyses (`04`) need the rows *before* that
dedup, so we also write a per-file table, keeping `subject_id` (+ a
`subject_key` fallback) so those analyses aggregate to one-per-subject
themselves; the rows are not independent.

```{r combine-tables-perfile}
#| echo: false
getcol <- function(df, nm)
  if (nm %in% names(df)) as.character(df[[nm]]) else rep(NA_character_, nrow(df))

all_samples_perfile <- dplyr::bind_rows(lapply(names(results), function(ds) {
  r    <- results[[ds]]
  meta <- tryCatch(load_metadata(ds),
                   error = function(e) data.frame(file = basename(bio_files_per_ds[[ds]])))
  .m   <- resolve_isobars(r$metrics_par, r$metrics_met, conf_final, ds, r$rt_anchor)
  st   <- build_sample_table(.m$par, .m$met,
                             basename(bio_files_per_ds[[ds]]), meta, dedup = FALSE)
  st$subject_key <- dplyr::coalesce(getcol(st, "subject_id"),
                                    getcol(st, "sample_name"), as.character(st$file))
  cbind(dataset_id = ds, assay = r$assay, st, stringsAsFactors = FALSE)
}))
write.csv(all_samples_perfile, "artifacts/sample_table_perfile.csv", row.names = FALSE)
cat("One row per raw file:", nrow(all_samples_perfile),
    "| one row per subject-sample after collapsing repeat files:", nrow(all_samples), "\n")
```

## Per-dataset HTML reports

`dataset-report.qmd` is a parameterised template: given one `dataset_id`, it
produces a deep dive into that dataset alone. The loop below renders it once per
dataset into `reports/report-<dataset_id>.html`.

This is skipped on a normal render, because it is slow (one full quarto render
per dataset) and only the per-dataset narrative depends on it. Render this
notebook with `-P render_reports:true` to regenerate them:

```
quarto render 03-all-datasets.qmd -P render_reports:true
```

```{r render-per-dataset-reports, eval=params$render_reports}
dir.create("reports", showWarnings = FALSE)
for (ds in names(results)) {
  message("Rendering report for ", ds, " …")
  out <- paste0("report-", ds, ".html")
  system2("quarto", c("render", shQuote("dataset-report.qmd"),
                      "-P", shQuote(paste0("dataset_id:", ds)),
                      "--output", shQuote(out)))
  file.rename(out, file.path("reports", out))
}
message("Done: ", length(results), " report(s) in reports/")
```

# Cross-dataset diagnostics

## Ratio distribution per dataset

Each box uses only files where both compounds were cleanly detected -- a peak in
both, neither flagged as a peak-width outlier -- and the file count is in each
label.

```{r plot-ratio-by-ds, fig.width = 8, fig.height = 5.5}
# Alphabetical on purpose: sorting by median would invite exactly the
# cross-dataset ranking the note below this plot says is meaningless.
both_all <- analysis_samples |>
  dplyr::filter(detection_tier == "both",
                !par_fwhm_outlier %in% TRUE,
                !met_fwhm_outlier %in% TRUE)

both_lab <- both_all |>
  dplyr::add_count(dataset_id, name = "n") |>
  dplyr::mutate(ds_label = paste0(short_ds(dataset_id), "  (n=", n, ")"))

ggplot(both_lab, aes(x = ds_label, y = log2_ratio)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +
  geom_boxplot(outlier.shape = NA, fill = "grey95") +
  geom_jitter(width = 0.15, alpha = 0.7, size = 1.8) +
  coord_flip() +
  labs(x = NULL,
       y = "log2(omeprazole peak area / 5-OH-omeprazole peak area)",
       title = "Parent / metabolite ratio within each dataset") +
  theme_bw()
```

A mass spectrometer counts ions, not molecules, and omeprazole and
5-hydroxyomeprazole ionise with different efficiency, so each box shows the true
ratio plus an unknown per-method offset. Compare the spread and the sign
*within* a box; do not rank the deposits by vertical position, since a box
sitting a unit higher may simply have used a method that ionises the parent more
readily. `05` sets out the offset model, the two designs in the corpus that let
it be probed rather than asserted, and the variance decomposition that measures
what the between-deposit spread is actually made of.

## Per-dataset detection-tier composition

Every file lands in one of four tiers: **both** compounds detected,
`parent_only`, `metab_only`, or `neither`. This plot shows the share of each
dataset's files in each tier, which is the clearest picture of what kind of
cohort each deposit actually is.

```{r plot-tier-composition, fig.width = 8, fig.height = 4.5}
analysis_samples |>
  dplyr::count(dataset_id, detection_tier) |>
  dplyr::mutate(ds_label = short_ds(dataset_id)) |>
  ggplot(aes(x = ds_label, y = n, fill = detection_tier)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  labs(x = NULL, y = "Share of files",
       fill = "Detection tier",
       title = "Detection-tier composition per dataset") +
  theme_bw() +
  theme(legend.position = "bottom")
```

The share plot is dominated by "neither", which hides how much usable signal
each dataset actually contributes. The second plot drops "neither" and shows the
counts behind the detected files, with the per-dataset total on the right.

```{r plot-tier-counts, fig.width = 8, fig.height = 4.5}
det_counts <- analysis_samples |>
  dplyr::filter(detection_tier != "neither") |>
  dplyr::count(dataset_id, detection_tier) |>
  dplyr::mutate(ds_label = short_ds(dataset_id))

det_totals <- det_counts |>
  dplyr::group_by(ds_label) |>
  dplyr::summarise(total = sum(n), .groups = "drop")

ggplot(det_counts, aes(x = ds_label, y = n, fill = detection_tier)) +
  geom_col() +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5),
            size = 2.9, colour = "white") +
  geom_text(data = det_totals, aes(x = ds_label, y = total, label = total),
            inherit.aes = FALSE, hjust = -0.35, size = 3, colour = "grey30") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_flip() +
  labs(x = NULL, y = "Files with a detection",
       fill = "Detection tier",
       title = "Detected files per dataset (excluding \"neither\")") +
  theme_bw() +
  theme(legend.position = "bottom")
```

Most files are "neither", which is the expected result rather than a failure.
These are general-population and disease cohorts, not drug trials: the great
majority of subjects were never prescribed omeprazole, so there is genuinely
nothing in their files to find, and a high "neither" share describes the cohort
rather than the pipeline. The deposits where "both" dominates are the drug-probe
cohorts, MSV000082493 and MSV000084008, where every subject was dosed by design.
The contrast between the two shapes is the clearest picture of what reverse
metabolomics is doing: finding the handful of incidentally-dosed people inside
cohorts assembled for entirely unrelated reasons.

`parent_only` and `metab_only` are not misses. Each is a censored ratio
observation -- the ratio is known to be very high or very low, but not by how
much, because one of the two peaks fell below detection. `05` takes up what that
costs an estimate that keeps only the "both" files.
