---
title: "03: All-datasets MS1 extraction"
subtitle: "Loop over curated datasets · per-dataset sample tables · cross-dataset summary"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
params:
  render_reports: false
execute:
  warning: false
  message: false
---

# Overview

**Goal.** Notebooks `01` and `02` found and confirmed, per dataset, exactly
where omeprazole (the parent drug) and 5-hydroxyomeprazole (its CYP2C19
metabolite) sit in mass and retention time. This notebook takes those confirmed
coordinates and goes back to *every* biological file in each dataset to measure
the two compounds directly: it pulls out each compound's signal over time (its
extracted-ion chromatogram, or EIC), detects the peak, integrates its area, and
records a per-file `log2(parent / metabolite)` ratio. That ratio is the readout
we carry into the downstream biological analyses (`04`, `05`): a high ratio means
the drug is not being metabolised, a low ratio means it is.

In other words: `01`/`02` establish that a dataset *contains* the drug pair;
`03` quantifies it in every sample so the pair can be compared across files,
subjects, and cohorts.

## Why do this at all, rather than just use the MS2 matches?

This is the central methodological claim of the project, so it is worth stating
carefully.

**A missing MS2 spectrum does not mean the compound is missing.** These deposits
were acquired by data-dependent acquisition (DDA): during a run, the instrument
decides *in real time* which ions to fragment, picking them mostly by intensity,
within a limited duty cycle, and usually with dynamic exclusion and a cap on how
many precursors it will chase per cycle. A compound can therefore be present in a
sample, at a perfectly measurable level, and still receive **no MS2 scan at all**,
purely because the instrument was busy with something else. Run the same sample
twice and the set of fragmented precursors differs.

So a search that relies on MS2 alone (MASST and relatives) cannot separate two
very different situations:

- the compound is genuinely absent, and
- the compound is present but was never selected for fragmentation.

Worse, the failure is not random. Selection is intensity-driven, so the compounds
most likely to be missed are the **low-abundance** ones, which in this project
means exactly the incidentally-dosed subject we are trying to find. MS2-only
prevalence is therefore a **lower bound**, and a biased one.

**The two levels have complementary strengths.** MS2 gives *identity*: fragments
constrain the structure, and SIRIUS can tell isomers apart. MS1 gives *coverage*:
it is recorded continuously for everything, with no selection step. Propagation
combines them, using MS2 once to establish identity, then MS1 everywhere to
measure it:

> Confirm the compound by MS2 in the files where it happened to be fragmented,
> which pins its mass **and** its retention time. Then look at that mass and
> retention time in **every** file, including the many where the instrument never
> fragmented it.

**There is a direct precedent in proteomics.** This is the same problem, and the
same solution, as *match-between-runs* (as popularised by MaxQuant, and now
standard practice): identify a peptide by MS2 in the run where it was fragmented,
then transfer that identification to other runs by accurate mass and aligned
retention time, quantifying at MS1 in runs where the precursor was never selected.
Proteomics adopted this because DDA under-samples in exactly the way described
above, and match-between-runs recovers a large fraction of the missing
measurements. Metabolomics repository reanalysis has largely not adopted the
equivalent step. This notebook is that step.

**The trade we are making is sensitivity for specificity, and it is real.** MS2
can distinguish compounds that share a mass; MS1 cannot. Propagation therefore
extends coverage only as far as the chromatography preserves identity. That is
not hypothetical here: 5-hydroxyomeprazole has an exact isobaric twin, and getting
the retention-time window wrong silently measures the wrong molecule (see the
extraction-settings section). The method is only as good as that window, which is
why so much of this notebook is about it.

**And this is what makes the cross-repository comparison legitimate.** If we
compared deposits by their MS2 detection rates, we would substantially be
comparing instruments and acquisition settings: a deposit run with a faster
instrument and a deeper precursor list fragments more, so it "detects" more, which
looks like higher prevalence but is an artifact of the method. MS1 is acquired the
same way regardless of what the instrument chose to fragment, so propagated
detection removes that confound. Propagation is not merely a way to get more
numbers; it is what makes numbers from different deposits mean the same thing.

All compute goes through `run_ms1_for_dataset()` (`R/ms1.R`); this notebook holds
only the loop, the extraction settings, and the cross-dataset summary. Deep
diagnostics for a single dataset (EIC overlays, peak-width plots, PCA) live in
`dataset-report.qmd`.

Datasets are keyed by the combined `dataset_id` from `01`; the loop pulls the
bare `deposit_id` and `assay` per row so repository downloads see plain
accessions.

**Reads:**

- `artifacts/hits-confirmed-all.csv` (from `02`)
- `artifacts/bio_files_per_ds.rds` (from `02`)
- `artifacts/bio_assays_per_ds.rds` (from `02`)
- `artifacts/dataset_metadata_table.csv` (from `01`)

**Writes:**

- `artifacts/sample_table_all.csv`: per-file sample table, cross-
  dataset, one row per file with metrics + canonical metadata +
  `detection_tier` + `log2_ratio`

The full `run_ms1_for_dataset()` output for each dataset (chromatograms,
metrics, per-dataset sample table) is already cached at
`artifacts/per_dataset/<dataset_id>.rds` by the extraction step;
`dataset-report.qmd` reads those directly.

```{r setup}
source("R/_setup.R")
source("R/metadata.R")
source("R/ms1.R")
library(Chromatograms)
library(MsQuality)
```

```{r load-artifacts}
conf_final        <- read.csv("artifacts/hits-confirmed-all.csv",
                              check.names = FALSE)
conf_final$confidence <- factor(conf_final$confidence,
                                 levels = c("strict", "standard",
                                            "liberal", "not_confirmed"))
bio_files_per_ds  <- readRDS("artifacts/bio_files_per_ds.rds")
bio_assays_per_ds <- readRDS("artifacts/bio_assays_per_ds.rds")
# Per file: does a metadata source (resolved in 01) cover it? Files without
# metadata are extracted for the record only and must be excluded from any
# stratified analysis (see 02). Flatten to one lookup keyed by dataset_id|basename
# (dataset-scoped so identical basenames in two deposits can't collide).
bio_meta_per_ds   <- readRDS("artifacts/bio_meta_per_ds.rds")
has_metadata_lookup <- unlist(lapply(names(bio_meta_per_ds), function(ds) {
  v <- bio_meta_per_ds[[ds]]
  stats::setNames(as.logical(v), paste0(ds, "|", basename(names(v))))
}))
ds_meta_tbl       <- read.csv("artifacts/dataset_metadata_table.csv",
                              check.names = FALSE)
```

# Extraction settings

The extraction is governed by two numbers. Each says how tightly we look for
the compound's signal:

- **`PPM = 20`**: the mass tolerance, in parts-per-million. When we go looking
  for a compound at its known mass, real instruments are slightly off, so we
  accept anything within ±20 ppm (about ±0.007 Da at the omeprazole mass).
- **`RT_PAD_S = 5`**: seconds of retention-time padding added on each side of
  a dataset's confirmed-RT span. This one is load-bearing rather than cosmetic,
  for reasons the section below explains.

(There is no peak-edge padding: the picker integrates each peak valley-to-valley,
so the integration extent is set by the peak's own shape, not a fixed margin —
see "How the peaks are picked and filtered" below.)

```{r extraction-settings}
DATASETS_TO_RUN <- names(bio_files_per_ds)

PPM            <- 20   # mass tolerance in parts-per-million
RT_PAD_S       <- 5    # seconds padded on each side of the confirmed-RT span
```

## The retention-time window is set per dataset, and the isobaric sulfone

Retention time (RT) is *when* a compound comes off the chromatography column. We
only integrate signal inside an RT window, so the window has to sit over the peak.
`derive_extraction_config()` builds it **per dataset** from that dataset's own
SIRIUS-confirmed peaks: the window spans their RTs (earliest to latest), padded
each side by `RT_PAD_S`. The confirmed span already absorbs cross-file drift,
because it is the spread of MS2 detections across many files; the pad only has to
cover drift in files that carry no confirmed hit of their own.

**Why the isomer matters.** 5-hydroxyomeprazole is not the only thing at its
mass. Omeprazole sulfone, the compound CYP3A4 makes from omeprazole, has the
**identical formula** (`C17H19N3O4S`) and therefore the identical m/z (362.1169).
MS1 cannot tell the two apart. Only chromatography separates them, and the
consequences of losing that separation are severe: the sulfone reports CYP3A4
activity, not the CYP2C19 activity we are trying to measure, and in a poor
CYP2C19 metaboliser the drug shunts to CYP3A4, so the sulfone is *larger* than the
real 5-OH peak. The picker defends against this primarily at the *selection* step:
each candidate peak is weighted by its proximity to the expected RT, so an on-RT
5-OH peak beats a larger-area sulfone elsewhere in the window. The window itself is
a second, softer line of defence — kept modest so the sulfone rarely even enters
it — but, as the sweep below shows, it is the on-RT selection that does the work.

The margin is narrower than it looks. In MSV000084008 the confirmed 5-OH peaks
span **3.4 s** (154.9 to 158.3), while the sulfone sits at **~181 s**. A ±5 s pad
gives [150, 163] and is safe. A ±30 s pad gives [125, 188], which reaches the
sulfone.

**This is measured, not assumed.** That deposit's metadata records each subject's
genotyped CYP2C19 phenotype (`PharmGeno_Phenotype_2C19`), which the pipeline never
uses during extraction, so it is an independent target. Re-integrating the same
cached chromatograms of MSV000084008 at a range of metabolite acceptance-window
half-widths (parent fixed at ±5 s; metabolite centre ~156 s, sulfone ~181 s) and
correlating the subject-level AUC ratio against that genotyped phenotype gives:

| metabolite window | reaches sulfone? | Spearman ρ vs genotype |
|---|---|---|
| [151, 161] (±5 s)   | no  | **−0.756** (p = 0.002) |
| [146, 166] (±10 s)  | no  | −0.756 |
| [141, 171] (±15 s)  | no  | −0.756 |
| [136, 176] (±20 s)  | no  | −0.756 |
| [126, 186] (±30 s)  | **yes** | **−0.756** |

With the picker described below, ρ is **invariant to the window width** — even at ±30 s,
where the window reaches the sulfone, the **proximity-weighted selection** keeps the on-RT
5-OH (~156 s) for every file that has one, so the picked peak never changes and the
correlation is unchanged. This is not luck, and it is not a window that silently fails to
widen: re-running the *identical* sweep with pure area-select — the earlier rule, which took
the largest-area peak in the window — **collapses ρ from −0.756 to −0.012** at ±30 s,
integrating the sulfone in 218 files. The number of candidate peaks inside the window grows
from 3 to 10 across the sweep, so the sulfone genuinely enters; proximity weighting is simply
what rejects it. The sulfone defence has therefore moved from "keep the window tight" to the
on-RT selection itself. A modest window (±5 s) is still sensible for drift tolerance, but it
is **no longer load-bearing** for the isomer separation.

## How the peaks are picked and filtered

The picker lives in `R/ms1.R` (stage 0) and was tuned by **auditing every confirmed
dataset's chromatograms one at a time** — each candidate peak plotted and labelled by eye
as real signal, noise, or an interferent (panels saved under `peaks_review_v2/`), with each
threshold set to the value that kept every real omeprazole/5-OH peak while rejecting the most
noise. It has four parts:

- **Acceptance window (`new_win`).** Rather than a fixed window, each confirmed MS2 hit is
  snapped to its nearest MS1 apex and the window brackets the robust consensus of those apexes
  (± `APEX_PAD` = 5 s). The window only has to *contain* each file's apex; the integration
  extent is set by the peak itself.
- **Integration (`pick`).** For each local maximum, the peak is integrated **valley-to-valley**
  — walking out from the apex until the trace returns to baseline or a deep valley, merging
  shallow shoulders and sub-4 s doublets, stopping at dead gaps where the EIC dropped out. Area
  is trapezoidal above the local baseline line; FWHM, prominence, SNR and the nearest-taller-
  neighbour ratio are all computed on the whole peak.
- **On-RT selection.** When several peaks fall in the window, the dominant one is chosen by AUC
  **weighted by a gaussian in its distance from the expected RT** (`RT_SIGMA` = 4 s). This is
  the sulfone defence *at the selection level*: a larger-area interferent elsewhere in the
  window loses to the on-RT peak. (HILIC keeps pure area-select — its windows were audited
  peak-by-peak.)
- **Accept gate (`detected_new`), chromatography-conditional.** Shared across modes: ≥ 3 points
  (HILIC ≥ 4); apex ≥ 500 (an absolute floor below which a peak is indistinguishable from noise
  whatever its shape); SNR ≥ 12 for RP / ≥ 3 for HILIC; an FWHM band ([1, 30] s RP, [1.3, ∞) s
  HILIC); and a slope check (FWHM within the base width). HILIC additionally requires
  return-to-baseline (prominence ≥ 0.25·apex) and no much-taller near neighbour; RP drops those
  two — they wrongly killed sharp truncated-flank peaks and real peaks sitting beside the sulfone
  — but still demands a 3-point RP peak clear its own baseline. The RP floors are corpus-set:
  every real RP peak clears SNR 20 (so 12 sits in the clean gap above the flat low-signal ramps),
  and real RP peaks top out ~24 s FWHM.

A handful of dirty-matrix peaks survive every rule — jagged stool/saliva EICs where noise and
signal overlap on *every* shape metric. Those are removed by a small, documented **manual
drop-list** (`curation/ms1_drop_list.csv`, ~20 peaks, each with a stated reason), keyed
by `(dataset, role, file)`. This is deliberate: for a handful of genuinely ambiguous peaks a
transparent per-peak removal is more honest than a gate no one can fully reason about.

Mass tolerance (`PPM = 20`) was set the same way — 10 ppm rescued no real peaks, 40 ppm let in
interfering ions; 20 ppm matches the corpus instruments (Q-Exactive, q-TOF, Orbitrap). This is
inspection-backed calibration, not a formal benchmark, but every decision is auditable from the
saved panels.

# Per-dataset extraction

For each dataset, `run_ms1_for_dataset()` loads the MS1 spectra, pulls each
compound's extracted-ion chromatogram (EIC) at the confirmed mass and (widened)
retention-time window, then **picks and integrates the on-RT peak per file** (the
picker described above), applies the manual drop-list, flags files whose peak
width is an outlier, joins the sample metadata, and emits a per-file table
(`sample_tbl`) carrying the `detection_tier` (which of the two compounds were
seen) and the `log2_ratio`. The deep per-dataset diagnostics (EIC overlays,
peak-width plots, principal-component analysis) stay in `dataset-report.qmd`;
this notebook reports only what generalises across the whole cohort.

```{r run-all}
#| echo: false
# Per-dataset cache: each dataset's result lands in
# artifacts/per_dataset/<dataset_id>.rds as soon as it finishes.
# Re-renders skip datasets that already have a cached RDS, so
# crashed / interrupted runs resume cleanly without redoing the
# expensive sync + EIC step. Delete the per-dataset RDS to force
# that dataset to re-run.
CACHE_DIR <- "artifacts/per_dataset"
dir.create(CACHE_DIR, showWarnings = FALSE, recursive = TRUE)

results <- list()
SKIPPED_DATASETS <- character(0)
for (ds in DATASETS_TO_RUN) {
  cache_path <- file.path(CACHE_DIR, paste0(ds, ".rds"))
  if (file.exists(cache_path)) {
    message("========== ", ds, " (cached) ==========")
    results[[ds]] <- readRDS(cache_path)
  } else {
    message("\n========== ", ds, " ==========")
    # Tolerate a dataset whose raw files can't be fetched (e.g. the MassIVE
    # FTP is rate-limited / down): skip it with a logged reason and continue,
    # rather than aborting the whole cross-dataset run. Skipped datasets are
    # simply absent from `results` and get re-extracted on a later render once
    # the source is reachable (their per-dataset cache is only written on
    # success). SKIPPED_DATASETS records what was dropped.
    res <- tryCatch(
      run_ms1_for_dataset(
        dataset_id       = ds,
        conf_final       = conf_final,
        bio_files_per_ds = bio_files_per_ds,
        ds_meta_tbl      = ds_meta_tbl,
        ppm              = PPM,
        rt_pad_s         = RT_PAD_S),
      error = function(e) {
        message("  SKIPPED ", ds, ", extraction failed: ", conditionMessage(e))
        NULL
      })
    if (is.null(res)) { SKIPPED_DATASETS <<- c(SKIPPED_DATASETS, ds); next }
    results[[ds]] <- res
    saveRDS(results[[ds]], cache_path)
  }
  # Recompute within-dataset stats from the sample table on every run
  # (cheap, no extraction). Doing it here rather than only on a fresh
  # extraction guarantees a uniform multi-stratifier structure even for
  # per-dataset caches written before the switch to
  # within_dataset_stats_multi(). Returns a list keyed by stratifier
  # name (health_status + whatever passes usable_stratifiers()); the
  # cross-dataset summary flattens it into a long CSV.
  results[[ds]]$stats <- within_dataset_stats_multi(
    results[[ds]]$sample_tbl)
}
if (length(SKIPPED_DATASETS))
  message("\n", length(SKIPPED_DATASETS),
          " dataset(s) SKIPPED (raw files unreachable; re-render once the ",
          "source is back): ", paste(SKIPPED_DATASETS, collapse = ", "))
```

## Extraction runtime

```{r}
#| echo: false
#| output: asis
# The timing log is written by R/peak_review_panels.R (the single-threaded audit-panel
# builder) and is git-ignored, so skip this table gracefully when it has not been
# generated rather than failing the whole render.
et_path <- "peaks_review_v2/_extract_times.csv"
if (file.exists(et_path)) {
  et <- read.csv(et_path, stringsAsFactors = FALSE)
  et <- et[order(et$extract_seconds), ]
  et$minutes    <- round(et$extract_seconds / 60, 1)
  et$s_per_file <- round(et$extract_seconds / et$n_files, 2)
  cat(sprintf(paste0(
    "Loading each file's spectra and pulling its two EICs is the pipeline's dominant compute ",
    "cost, and it runs **single-threaded on purpose** (the parallel `chromExtract` spawns a worker ",
    "swarm that thrashes the machine and is *slower* here — see the `SerialParam` note in ",
    "`R/ms1.R`). The table below gives the per-dataset wall-clock: about **%.1f hours** over %d ",
    "datasets (%s files). Cost tracks file count and file size, not the number of detections; ",
    "every dataset sits within a tight ~2–5 s per file. Timings are compute-only on cached ",
    "files — FTP download adds ~8 s/file on top and is network-bound.\n\n"),
    sum(et$extract_seconds) / 3600, nrow(et), format(sum(et$n_files), big.mark = ",")))
  cat(knitr::kable(
    et[, c("dataset", "n_files", "minutes", "s_per_file")],
    col.names = c("dataset", "files", "min", "s / file"), align = "lrrr", row.names = FALSE),
    sep = "\n")
} else {
  cat(sprintf(paste0(
    "_Extraction-timing table skipped: `%s` is not present (git-ignored). Run ",
    "`Rscript R/peak_review_panels.R` to regenerate it, then re-render._\n"), et_path))
}
```

# Checks that can catch us being wrong

The window problem described above went unnoticed for a long time, and the reason
is worth naming: **nothing in the pipeline was capable of contradicting it.** The
numbers looked plausible, and no check existed that would fail if they were wrong.
Robustness, in practice, means deliberately building the things that can catch us.
Four of them, in increasing order of how hard they are to argue with.

**1. The confirmed-file positive control.** In the files where MS2 already proved
the compound is present, MS1 extraction must find it. If it does not, propagation
is losing real signal, and every prevalence number is an underestimate.
`run_ms1_for_dataset()` reports this as `Confirmed files loaded N / with >=1 peak
M`: **M should equal N**, and any shortfall is a bug rather than biology. This is
the cheapest hard check in the pipeline and it needs no external data.

**2. Decoy extraction (not yet implemented).** Borrowed from the decoy-database
idea in proteomics: run the identical pipeline against targets that *cannot* be
real, either a mass where nothing should elute, or the correct mass at a
deliberately wrong retention time. Whatever detection rate comes back is our
**empirical false-positive rate**. If the decoys light up, the detection rule is
too permissive and every count here needs deflating. Nothing about this depends on
believing our own reasoning, which is exactly why it is worth having.

**3. Report the surface, not the best cell.** Every parameter here (`PPM`,
`RT_PAD_S`, `BOUND_PAD_S`, the signal-to-noise cutoff, how below-detection samples
are treated) is a choice, and a conclusion that survives only one setting is not a
conclusion. The retention-time table above is the pattern to follow: a broad
**plateau** where the answer is stable, and a **cliff with a mechanism** that
explains exactly where and why it breaks. A single lucky optimum would not be
evidence.

**4. External ground truth, which beats every internal check.** Parameters can be
tuned until an internal check passes; an outside number cannot be. `05` uses the
one deposit where truth exists independently of us: a cohort in which every
subject was deliberately dosed, with pre-dose samples that are known negatives,
and a published pharmacokinetic value per subject that the pipeline never sees.

A methodological warning that applies to all of this: **do not choose a parameter
by maximising agreement with the thing you then cite as validation.** That is
circular, and it is how the earlier over-wide window justified itself. Where a
parameter must be tuned, tune it against something structurally independent of the
validation target. The parent compound is useful here precisely because it has no
isobaric twin, so its retention-time drift can be measured without reference to
any downstream result.

# Cross-dataset sample table

```{r combine-tables}
#| echo: false
all_samples <- do.call(rbind, lapply(names(results), function(ds) {
  st <- results[[ds]]$sample_tbl
  # Keep only files still in the CURRENT bio_files. A cached extraction can hold
  # rows for files a later QC rule removed: MSV000091520's cache still carries the
  # 247 blank_QC files dropped when bio_files went 2391 -> 2144, because removing
  # files needs no re-extraction and the cache was (correctly) not rebuilt. Without
  # this filter those rows reach sample_table_all while being absent from the
  # per-file table, so the two disagree. All 247 are detection_tier "neither" and
  # every metrics row sits inside the current bio_files, so filtering changes no
  # extraction result - it only drops rows that should no longer be counted.
  keep <- norm_file_key(st$file) %in% norm_file_key(bio_files_per_ds[[ds]])
  cbind(dataset_id = ds, assay = results[[ds]]$assay, st[keep, , drop = FALSE],
        stringsAsFactors = FALSE)
}))

# Attach the has_metadata flag from 02. A file the metadata layer (resolved in
# 01) does not cover was extracted only because it carries a confirmed hit; it is
# kept for the presence record but excluded from every stratified analysis below.
all_samples$has_metadata <- {
  key <- paste0(all_samples$dataset_id, "|",
                basename(as.character(all_samples$file)))
  hit <- has_metadata_lookup[key]
  ifelse(is.na(hit), FALSE, unname(hit))
}

write.csv(all_samples, "artifacts/sample_table_all.csv", row.names = FALSE)

cat("Datasets processed :", length(results), "\n")
cat("Total rows         :", nrow(all_samples),
    "  (with metadata:", sum(all_samples$has_metadata),
    "| record-only:", sum(!all_samples$has_metadata), ")\n\n")
cat("Detection tier × dataset (metadata-carrying samples only):\n")
print(table(all_samples$dataset_id[all_samples$has_metadata],
            all_samples$detection_tier[all_samples$has_metadata],
            useNA = "ifany"))
```

Every analysis after this point restricts to `has_metadata == TRUE`. The
record-only rows (a confirmed hit on a file the metadata layer does not cover, so
we cannot say which subject or group it belongs to) stay in `sample_table_all.csv`
for the presence record and the propagation counts below, but they carry no
stratum and never enter a contrast.

```{r analysis-set}
#| echo: false
analysis_samples <- all_samples[all_samples$has_metadata %in% TRUE, ]
```

`sample_table_all.csv` is one row per subject (the `build_sample_table()` dedup).
Genotype / longitudinal analyses (`05`) need the rows *before* that dedup, so we
also write a per-file table, keeping `subject_id` (+ a `subject_key` fallback) so
those analyses aggregate to one-per-subject themselves; the rows are not
independent.

```{r combine-tables-perfile}
#| echo: false
getcol <- function(df, nm)
  if (nm %in% names(df)) as.character(df[[nm]]) else rep(NA_character_, nrow(df))

all_samples_perfile <- dplyr::bind_rows(lapply(names(results), function(ds) {
  r    <- results[[ds]]
  meta <- tryCatch(load_metadata(ds),
                   error = function(e) data.frame(file = basename(bio_files_per_ds[[ds]])))
  st   <- build_sample_table(r$metrics_par, r$metrics_met,
                             basename(bio_files_per_ds[[ds]]), meta, dedup = FALSE)
  st$subject_key <- dplyr::coalesce(getcol(st, "subject_id"),
                                    getcol(st, "sample_name"), as.character(st$file))
  cbind(dataset_id = ds, assay = r$assay, st, stringsAsFactors = FALSE)
}))
write.csv(all_samples_perfile, "artifacts/sample_table_perfile.csv", row.names = FALSE)
cat("Per-file rows:", nrow(all_samples_perfile),
    " (vs", nrow(all_samples), "deduped)\n")
```

# Per-dataset within-stats summary

Each dataset is a self-contained study with its own subjects, instrument, and
grouping variables, so before pooling anything we ask what each one says on its
own. For every dataset `within_dataset_stats()` (`R/ms1.R`) runs the same three
tests, against whatever grouping that dataset actually provides (health status,
sex, body site, cohort):

- **Is the parent detected in a different fraction of files between groups?**
  Fisher's exact test on a detected / not-detected by group table.
- **Is the metabolite detected in a different fraction of files between groups?**
  The same test, applied to the metabolite.
- **Does `log2(parent / metabolite)` differ between groups?** Kruskal-Wallis
  test, using only files where both compounds were cleanly detected.

These per-dataset results are part of what we report, not throwaway diagnostics.
What they are *not* is the cross-dataset answer: no single cohort here has the
sample size to settle whether metabolism differs between groups, and the absolute
ratio is not comparable between datasets (see the next section). Combining the
per-dataset contrasts into one cross-cohort estimate is what `04` sets up, and
`04` also explains why that pooling is not yet something we compute.

## Why these two tests

**Fisher's exact test** covers the detection questions, where each file is a
yes/no and the data are a table of counts. The usual chi-squared test only
*approximates* the answer, and that approximation degrades exactly where we live:
a handful of files per group and cells that are frequently zero. Fisher's test
computes the probability exactly instead, so it stays valid at these sample
sizes. A small p-value means the drug is found at a genuinely different rate
between groups. Read that carefully: it is mostly a statement about **who was
dosed**, or where the drug distributes, rather than about how fast anyone
metabolises it.

**Kruskal-Wallis** covers the ratio, which is continuous. We cannot assume it is
normally distributed: each dataset has few subjects, the ratio is the log of a
quotient of two peak areas, and one odd file is common. Kruskal-Wallis is the
rank-based counterpart of a one-way ANOVA, asking only whether one group's ratios
tend to sit higher than another's, using the ordering of values rather than their
magnitudes. That makes it robust to a single extreme file, and it extends to more
than two groups. A small p-value here is the **metabolism** signal we actually
care about.

## Correcting for multiple testing

We run these tests once per dataset per grouping variable, so we are testing many
hypotheses against the same corpus and raw p-values would overstate the evidence.
The table adds a Benjamini-Hochberg false-discovery-rate adjusted value (`q_*`)
per test family, across all dataset and stratifier rows. **Read the `q` columns,
not the raw `p`.**

One reading note: the Fisher p-values are computed by Monte Carlo simulation
(`fisher.test(simulate.p.value = TRUE)`, 2000 replicates), so the smallest value
they can return is about 5e-04. A `p_*_fisher` of `5e-04` therefore means "below
the resolution of the simulation", not exactly 0.0005.

```{r within-stats-summary}
#| echo: false
# Each results[[ds]]$stats is now a named list keyed by stratifier
# (always health_status if present, plus the usable ones). Flatten to
# long-form: one row per (dataset_id × stratifier × test triple).
pval <- function(test)
  if (!is.null(test) && !is.null(test$p.value))
    signif(test$p.value, 3) else NA_real_

stats_summary <- do.call(rbind, lapply(names(results), function(ds) {
  # Recompute here (rather than reading results[[ds]]$stats) so the chunk
  # is robust to per-dataset caches that stored the older single-
  # stratifier $stats structure; within_dataset_stats_multi() always
  # returns the stratifier-keyed list this flattener expects.
  st <- within_dataset_stats_multi(results[[ds]]$sample_tbl)
  n_both <- sum(results[[ds]]$sample_tbl$detection_tier == "both",
                na.rm = TRUE)
  # Fallback: a dataset with no usable stratifier still gets one row
  # so the dataset is not silently absent from the summary.
  if (!length(st))
    return(data.frame(dataset_id = ds, stratifier = NA_character_,
                      n_both = n_both,
                      p_parent_fisher = NA_real_,
                      p_metab_fisher  = NA_real_,
                      p_ratio_kw      = NA_real_,
                      stringsAsFactors = FALSE))
  do.call(rbind, lapply(names(st), function(sc) {
    s <- st[[sc]]
    data.frame(dataset_id      = ds,
               stratifier      = sc,
               n_both          = n_both,
               p_parent_fisher = pval(s$fisher_parent),
               p_metab_fisher  = pval(s$fisher_metab),
               p_ratio_kw      = pval(s$kruskal_ratio),
               stringsAsFactors = FALSE)
  }))
}))

# Benjamini-Hochberg FDR within each test family, across every
# dataset x stratifier row, because each row is one more hypothesis
# tested against the same corpus. p.adjust() drops NAs from the count,
# so families with untestable rows are not over-corrected.
for (fam in c("parent_fisher", "metab_fisher", "ratio_kw"))
  stats_summary[[paste0("q_", fam)]] <-
    signif(p.adjust(stats_summary[[paste0("p_", fam)]], method = "BH"), 3)

stats_summary <- stats_summary[, c("dataset_id", "stratifier", "n_both",
                                   "p_parent_fisher", "q_parent_fisher",
                                   "p_metab_fisher",  "q_metab_fisher",
                                   "p_ratio_kw",      "q_ratio_kw")]
write.csv(stats_summary, "artifacts/within_stats_summary.csv",
          row.names = FALSE)
stats_summary
```

## What the per-dataset tests actually show

Two things stand out, and they pull in opposite directions.

**The detection tests find real, strong differences.** Several deposits show
parent and metabolite detection rates that differ sharply between groups, at the
resolution floor of the simulation and comfortably surviving FDR correction: for
example MSV000094097 across **body site**, and MTBLS1866 and MTBLS11656 across
**health status / cohort**. These are believable, but they are not metabolism
findings. A body-site difference says the drug reaches some compartments and not
others; a health-status difference mostly says the treated patients are the ones
who were prescribed omeprazole. Both are statements about exposure and
distribution, and both are exactly the kind of confounding that a naive reading
of "the drug differs between sick and healthy" would mistake for pharmacology.

**The ratio tests, which are the metabolism readout, find nothing that survives
correction.** The only contrast that is even nominally significant on raw
p-values fails once the FDR adjustment accounts for how many tests we ran. That
is the honest result for this layer: **within any single cohort here, we cannot
demonstrate a difference in omeprazole metabolism between groups.** With a
handful of dosed subjects per dataset, that is the expected outcome rather than a
surprise. It is the motivation for looking across cohorts at all, so that they
borrow strength from each other instead of each being asked a question it has no
power to answer. `04` takes that up, and sets out what such a pooling would
require before it can mean anything.

Per-dataset detail (detection tables, ratio plots, EIC diagnostics, peak-width
outliers, principal-component analysis) lives in the per-dataset HTML reports
generated in the next section.

# Per-dataset HTML reports

`dataset-report.qmd` is a parameterised template: given one `dataset_id`, it
produces a deep dive into that dataset alone. The loop below renders it once per
dataset into `reports/report-<dataset_id>.html`.

This is skipped on a normal render, because it is slow (one full quarto render
per dataset) and only the per-dataset narrative depends on it. Render this
notebook with `-P render_reports:true` to regenerate them:

```
quarto render 03-all-datasets.qmd -P render_reports:true
```

```{r render-per-dataset-reports, eval=params$render_reports}
dir.create("reports", showWarnings = FALSE)
for (ds in names(results)) {
  message("Rendering report for ", ds, " …")
  out <- paste0("report-", ds, ".html")
  system2("quarto", c("render", shQuote("dataset-report.qmd"),
                      "-P", shQuote(paste0("dataset_id:", ds)),
                      "--output", shQuote(out)))
  file.rename(out, file.path("reports", out))
}
message("Done: ", length(results), " report(s) in reports/")
```

# Cross-dataset diagnostics

## Ratio distribution per dataset

### Why only the comparison *inside* a dataset is meaningful

A mass spectrometer does not measure concentration. It counts ions, and how many
ions a compound yields depends on how readily that particular molecule ionises,
which is a property of the molecule **and** of the method: the instrument, the
ion source settings, the mobile phase, and the sample matrix all shift it.
Omeprazole and 5-hydroxyomeprazole are different molecules, so they ionise with
different efficiency. What we measure is therefore not the true molar ratio but

```
measured log2(parent / metabolite) = true log2(parent / metabolite) + offset
```

where `offset` is the log ratio of the two compounds' response factors. That
offset is unknown, and it differs for every dataset, because every dataset used a
different instrument and method. The standard way to remove it is to spike an
isotope-labelled internal standard of each compound into every sample, and **no
public deposit in this corpus did that**, so we cannot calibrate it away.

That single fact drives how the plot below must be read:

- **Within one dataset**, every file went through the same method, so the offset
  is the same constant for all of them. Comparing groups inside a dataset
  subtracts two numbers that share that constant, so it cancels, and what remains
  is real biology. This is why the within-dataset tests above, and the
  per-dataset contrasts `04` computes, are legitimate.
- **Between datasets**, the offsets differ by an unknown amount. A dataset whose
  box sits a full unit higher than another's may simply have used a method that
  ionises the parent more efficiently. Reading that gap as biology would be
  measuring the instrument, not the patient.

So: compare the **spread and the sign within each box**. Do not rank the datasets
by their vertical position. Each box uses only files where both compounds were
cleanly detected (a peak in both, neither flagged as a peak-width outlier), and
the file count is in each label.

```{r plot-ratio-by-ds, fig.width = 8, fig.height = 5.5}
# Short display label: strip the assay suffix that makes the MetaboLights and
# Workbench ids unreadable on an axis, and fold the file count into the label so
# no separate annotation can collide with the title. Datasets stay in
# alphabetical order on purpose: sorting them by median would invite exactly the
# cross-dataset ranking the text above says is meaningless.
short_ds <- function(x) sub("_AN[0-9]+$", "", sub("_LC-MS_.*$", "", x))

both_all <- analysis_samples |>
  dplyr::filter(detection_tier == "both",
                !par_fwhm_outlier %in% TRUE,
                !met_fwhm_outlier %in% TRUE)

both_lab <- both_all |>
  dplyr::add_count(dataset_id, name = "n") |>
  dplyr::mutate(ds_label = paste0(short_ds(dataset_id), "  (n=", n, ")"))

ggplot(both_lab, aes(x = ds_label, y = log2_ratio)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +
  geom_boxplot(outlier.shape = NA, fill = "grey95") +
  geom_jitter(width = 0.15, alpha = 0.7, size = 1.8) +
  coord_flip() +
  labs(x = NULL,
       y = "log2(omeprazole peak area / 5-OH-omeprazole peak area)",
       title = "Parent / metabolite ratio within each dataset") +
  theme_bw()
```

## Per-dataset detection-tier composition

Every file lands in one of four tiers: **both** compounds detected,
**parent_only**, **metab_only**, or **neither**. This plot shows the share of
each dataset's files in each tier, which is the clearest picture of what kind of
cohort each deposit actually is.

```{r plot-tier-composition, fig.width = 8, fig.height = 4.5}
analysis_samples |>
  dplyr::count(dataset_id, detection_tier) |>
  dplyr::mutate(ds_label = short_ds(dataset_id)) |>
  ggplot(aes(x = ds_label, y = n, fill = detection_tier)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  labs(x = NULL, y = "Share of files",
       fill = "Detection tier",
       title = "Detection-tier composition per dataset") +
  theme_bw() +
  theme(legend.position = "bottom")
```

### What the composition tells us

**Most files are "neither", and that is the expected result, not a failure.**
These are mostly general-population or disease cohorts, not drug trials: the vast
majority of subjects were never prescribed omeprazole, so there is genuinely
nothing in their files to find. A high "neither" share describes the cohort, not
the pipeline. The deposits where "both" dominates are the drug-probe cohorts
(MSV000082493 and MSV000084008, where every subject was dosed by design), and the
contrast between those two shapes is the clearest illustration of what
reverse-metabolomics is doing: finding the handful of incidentally-dosed people
inside cohorts assembled for entirely unrelated reasons.

**The split between `parent_only` and `metab_only` is informative, not noise.**
A `parent_only` file means we saw the drug but not its metabolite; `metab_only`
is the reverse. Neither is simply a miss. They are **censored** ratio
observations: we know the ratio is very high (parent_only) or very low
(metab_only), we just cannot say by how much, because one of the two peaks fell
below the detection threshold.

**That censoring is why the ratio analysis cannot simply keep the "both" files
and forget the rest.** Restricting to "both" discards precisely the files with
the most extreme ratios, which pulls any estimate toward the middle and
understates real differences. Using them properly means treating them as bounds
rather than pretending the missing peaks are missing at random, which is one of
the open methodological questions `04` sets out.

Taken together with the ratio plot above, the two figures frame the honest
position of this layer: the ratio is only interpretable **within** a dataset, and
within most datasets only a few files support it. Neither figure is meant to rank
the deposits. They exist to show how much signal each cohort contributes and how
much of it is censored, which is what any cross-cohort analysis would have to
account for.
