---
title: "03: All-datasets MS1 extraction"
subtitle: "Loop over curated datasets · per-dataset sample tables · cross-dataset summary"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
params:
  render_reports: false
execute:
  warning: false
  message: false
---

# Overview

**Goal.** Notebooks `01` and `02` found and confirmed, per dataset, exactly
where omeprazole (the parent drug) and 5-hydroxyomeprazole (its CYP2C19
metabolite) sit in mass and retention time. This notebook takes those confirmed
coordinates and goes back to *every* biological file in each dataset to measure
the two compounds directly: it pulls out each compound's signal over time (its
extracted-ion chromatogram, or EIC), detects the peak, integrates its area, and
records a per-file `log2(parent / metabolite)` ratio. That ratio is the readout
we carry into the downstream biological analyses (`04`, `05`): a high ratio means
the drug is not being metabolised, a low ratio means it is.

In other words: `01`/`02` establish that a dataset *contains* the drug pair;
`03` propagates that detection to every file by MS1 (well beyond the MS2-confirmed
subset) and quantifies the pair, so it can be compared across files, subjects, and
cohorts.

## Why do this at all, rather than just use the MS2 matches?

This is the central methodological claim of the project, so it is worth stating
carefully.

**A missing MS2 spectrum does not mean the compound is missing.** These deposits
were acquired by data-dependent acquisition (DDA): during a run, the instrument
decides *in real time* which ions to fragment, picking them mostly by intensity,
within a limited duty cycle, and usually with dynamic exclusion and a cap on how
many precursors it will chase per cycle. A compound can therefore be present in a
sample, at a perfectly measurable level, and still receive **no MS2 scan at all**,
purely because the instrument was busy with something else. Run the same sample
twice and the set of fragmented precursors differs.

So a search that relies on MS2 alone (MASST and relatives) cannot separate two
very different situations:

- the compound is genuinely absent, and
- the compound is present but was never selected for fragmentation.

Worse, the failure is not random. Selection is intensity-driven, so the compounds
most likely to be missed are the **low-abundance** ones, which in this project
means exactly the incidentally-dosed subject we are trying to find. MS2-only
prevalence is therefore a **lower bound**, and a biased one.

**The two levels have complementary strengths.** MS2 gives *identity*: fragments
constrain the structure, and SIRIUS can tell isomers apart. MS1 gives *coverage*:
it is recorded continuously for everything, with no selection step. Propagation
combines them, using MS2 once to establish identity, then MS1 everywhere to
measure it:

> Confirm the compound by MS2 in the files where it happened to be fragmented,
> which pins its mass **and** its retention time. Then look at that mass and
> retention time in **every** file, including the many where the instrument never
> fragmented it.

**There is a direct precedent in proteomics.** This is the same problem, and the
same solution, as *match-between-runs* (as popularised by MaxQuant, and now
standard practice): identify a peptide by MS2 in the run where it was fragmented,
then transfer that identification to other runs by accurate mass and aligned
retention time, quantifying at MS1 in runs where the precursor was never selected.
Proteomics adopted this because DDA under-samples in exactly the way described
above, and match-between-runs recovers a large fraction of the missing
measurements. Metabolomics repository reanalysis has largely not adopted the
equivalent step. This notebook is that step.

**The trade we are making is sensitivity for specificity, and it is real.** MS2
can distinguish compounds that share a mass; MS1 cannot. Propagation therefore
extends coverage only as far as the chromatography preserves identity. That is
not hypothetical here: 5-hydroxyomeprazole has an exact isobaric twin, and getting
the retention-time window wrong silently measures the wrong molecule (see the
extraction-settings section). The method is only as good as that window, which is
why so much of this notebook is about it.

**This also makes the cross-repository comparison fairer.** If we compared deposits
by their MS2 detection rates, we would substantially be comparing instruments and
acquisition settings: a deposit run with a faster instrument and a deeper precursor
list fragments more, so it "detects" more, which looks like higher prevalence but is
an artifact of the method. MS1 is acquired regardless of what the instrument chose to
fragment, so propagating detection to MS1 removes that DDA-selection confound. It does
not make every deposit comparable (instrument sensitivity, matrix and chromatography
still differ), but it removes the confound that most distorts a raw MS2 detection-rate
comparison.

All compute goes through `run_ms1_for_dataset()` (`R/ms1.R`); this notebook holds
only the loop, the extraction settings, and the cross-dataset summary. Deep
diagnostics for a single dataset (EIC overlays, peak-width plots, PCA) live in
`dataset-report.qmd`, a parameterised template that takes one `dataset_id`, reads
that dataset's cached result, and renders a per-dataset HTML report. Section 7
below renders it once per dataset.

Datasets are keyed by the combined `dataset_id` from `01`; the loop pulls the
bare `deposit_id` and `assay` per row so repository downloads see plain
accessions.

**Reads:**

- `artifacts/hits-confirmed-all.csv` (from `02`)
- `artifacts/bio_files_per_ds.rds` (from `02`)
- `artifacts/bio_assays_per_ds.rds` (from `02`)
- `artifacts/dataset_metadata_table.csv` (from `01`)

**Writes:**

- `artifacts/sample_table_all.csv`: cross-dataset sample table, de-duplicated to one
  row per subject where a subject identifier exists
- `artifacts/sample_table_perfile.csv`: the same measurements one row per file, which
  is what every per-file analysis downstream reads

The full `run_ms1_for_dataset()` output for each dataset (chromatograms,
metrics, per-dataset sample table) is already cached at
`artifacts/per_dataset/<dataset_id>.rds` by the extraction step;
`dataset-report.qmd` reads those directly.

```{r setup}
source("R/_setup.R")
source("R/metadata.R")
source("R/masst.R")   # compound_spec + PHENO, the domain constants
source("R/ms1.R")
library(Chromatograms)
library(MsQuality)
```

```{r load-artifacts}
conf_final        <- read.csv("artifacts/hits-confirmed-all.csv",
                              check.names = FALSE)
conf_final$confidence <- factor(conf_final$confidence,
                                 levels = c("strict", "standard",
                                            "liberal", "not_confirmed"))
bio_files_per_ds  <- readRDS("artifacts/bio_files_per_ds.rds")
bio_assays_per_ds <- readRDS("artifacts/bio_assays_per_ds.rds")
# Does a metadata source cover each file? Keyed dataset_id|basename so identical
# basenames in two deposits cannot collide.
bio_meta_per_ds   <- readRDS("artifacts/bio_meta_per_ds.rds")
has_metadata_lookup <- unlist(lapply(names(bio_meta_per_ds), function(ds) {
  v <- bio_meta_per_ds[[ds]]
  stats::setNames(as.logical(v), paste0(ds, "|", basename(names(v))))
}))
ds_meta_tbl       <- read.csv("artifacts/dataset_metadata_table.csv",
                              check.names = FALSE)
```

# Extraction settings

The extraction is governed by two numbers. Each says how tightly we look for
the compound's signal:

- **`PPM = 20`**: the mass tolerance, in parts-per-million. When we go looking
  for a compound at its known mass, real instruments are slightly off, so we
  accept anything within ±20 ppm (about ±0.007 Da at the omeprazole mass).
- **`RT_PAD_S = 5`**: seconds of retention-time padding added on each side of
  a dataset's confirmed-RT span. This one is load-bearing rather than cosmetic,
  for reasons the section below explains.

(There is no peak-edge padding: the picker integrates each peak valley-to-valley,
so the integration extent is set by the peak's own shape, not a fixed margin.
See "How the peaks are picked and filtered" below.)

```{r extraction-settings}
DATASETS_TO_RUN <- names(bio_files_per_ds)

PPM            <- 20   # mass tolerance in parts-per-million
RT_PAD_S       <- 5    # seconds padded on each side of the confirmed-RT span
```


## How the peaks are picked and filtered

The picker lives in `R/ms1.R` (stage 0). Its gates were tuned by auditing a handful of
peaks in one RP and one HILIC dataset (each plotted and labelled by eye as signal, noise,
or interferent). Run across the corpus, the gates then flagged **1,304 detected peaks with
no MS2 confirmation of their own** (the MS1-propagated detections); these were reviewed
against the panels saved under `peaks_review_v2/`, with ~20 rejected to the drop-list below.
It has four parts:

- **Acceptance window (`acceptance_window`).** Rather than a fixed window, each confirmed MS2 hit is
  snapped to its nearest MS1 apex and the window brackets the robust consensus of those apexes
  (± `APEX_PAD` = 5 s). The window only has to *contain* each file's apex; the integration
  extent is set by the peak itself.
- **Integration (`pick`).** For each local maximum, the peak is integrated valley-to-valley
  (walking out from the apex until the trace returns to baseline or a deep valley), merging
  shallow shoulders and sub-4 s doublets, stopping at dead gaps where the EIC dropped out. Area
  is trapezoidal above the local baseline line; prominence, SNR and the nearest-taller-neighbour
  ratio are computed on the whole peak.
- **Peak metrics (`add_msquality`).** The picker chooses the peak and its boundaries;
  **MsQuality** then measures it, so the reported width, FWHM and shape are the mzQC
  controlled-vocabulary quantities rather than local re-implementations. Handed our
  boundaries, `peakWidth()` is bit-identical to the local `pkwidth` on every trace tested and
  `xicFwhm()` is bit-identical to the local FWHM on reverse phase; on HILIC the two differ,
  because ours takes the outermost half-maximum crossings of a merged cluster while MsQuality
  takes 50% of the maximum inside the region. `gaussianSimilarity()` is recorded per peak but
  never gated on: it needs ≥ 5 points in the peak region and is NA on the sparsest.
  `maxIntensity` and `signalToNoiseRatio` stay local — both describe *the peak the picker
  selected*, and MsQuality has no way of being told which of several peaks in the window that
  is (its SNR is also NA on a tight acceptance window, where the MAD noise estimate is 0).
- **On-RT selection.** When several peaks fall in the window, the dominant one is chosen by AUC
  **weighted by a gaussian in its distance from the expected RT** (`RT_SIGMA` = 4 s). This is
  the sulfone defence *at the selection level*: a larger-area interferent elsewhere in the
  window loses to the on-RT peak. (HILIC keeps pure area-select; its windows were audited
  peak-by-peak.)
- **Accept gate (`accept_peak`), chromatography-conditional.** Shared: ≥ 3 points (HILIC ≥ 4),
  apex ≥ 500, SNR ≥ 12 (RP) / ≥ 3 (HILIC), an FWHM band ([1, 30] s RP, [1.3, ∞) s HILIC), and a
  slope check. HILIC also requires return-to-baseline and no much-taller neighbour; RP drops
  those two (they killed sharp truncated-flank peaks and real peaks beside the sulfone) but
  still requires a 3-point peak to clear its own baseline. The RP floors are corpus-set: real RP
  peaks all clear SNR 20 and top out ~24 s FWHM.

A handful of dirty-matrix peaks survive every rule: jagged stool/saliva EICs where noise and
signal overlap on *every* shape metric. Those are removed by a small, documented **manual
drop-list** (`curation/ms1_drop_list.csv`, ~20 peaks, each with a stated reason), keyed
by `(dataset, role, file)`. This is deliberate: for a handful of genuinely ambiguous peaks a
transparent per-peak removal is more honest than a gate no one can fully reason about.

The gate thresholds are inspection-backed calibration, not a formal benchmark, but every
decision is auditable from the saved panels.

## The retention-time window is set per dataset, and the isobaric sulfone

Retention time (RT) is *when* a compound comes off the chromatography column. We
only integrate signal inside an RT window, so the window has to sit over the peak.
`derive_extraction_config()` builds it per dataset from that dataset's own
SIRIUS-confirmed peaks: the window spans their RTs (earliest to latest), padded
each side by `RT_PAD_S`. The confirmed span already absorbs cross-file drift,
because it is the spread of MS2 detections across many files; the pad only has to
cover drift in files that carry no confirmed hit of their own.

*Why the isomer matters.* 5-hydroxyomeprazole is not the only thing at its mass.
Omeprazole sulfone, the compound CYP3A4 makes from omeprazole, has the identical
formula (`C17H19N3O4S`) and therefore the identical m/z (362.1169). MS1 cannot tell
the two apart. Only chromatography separates them, and the consequences of losing
that separation are severe: the sulfone reports CYP3A4 activity, not the CYP2C19
activity we are trying to measure, and in a poor CYP2C19 metaboliser the drug shunts
to CYP3A4, so the sulfone is *larger* than the real 5-OH peak. The proximity-weighted
selection described above is the primary defence; the window is a second, softer one,
kept modest so the sulfone rarely even enters it.

The margin is narrower than it looks. In MSV000084008 the confirmed 5-OH peaks span
3.4 s (154.9 to 158.3), while the sulfone sits at ~181 s. A ±5 s pad gives [150, 163]
and is safe. A ±30 s pad gives [125, 188], which reaches the sulfone.

*This is measured, not assumed.* That deposit's metadata records each subject's
genotyped CYP2C19 phenotype (`PharmGeno_Phenotype_2C19`), which the pipeline never
uses during extraction, so it is an independent target. Re-integrating the same
cached chromatograms of MSV000084008 at a range of metabolite acceptance-window
half-widths (parent fixed at ±5 s; metabolite centre ~156 s, sulfone ~181 s) and
correlating the subject-level AUC ratio against that genotyped phenotype gives:

```{r}
#| echo: false
#| output: asis
#| label: tbl-pad-sweep
# Recomputed at render time from the cached wide EICs of MSV000084008: widen the
# metabolite acceptance window toward the sulfone, re-pick with and without the
# proximity weighting, and correlate the subject AUC ratio against genotype.
sweep_cache <- "peaks_review_v2/_cache/MSV000084008.rds"
sweep_meta  <- "dataset_metadata/MSV000084008_metadata.tsv"
if (file.exists(sweep_cache) && file.exists(sweep_meta)) {
  # PHENO (CYP2C19 activity order) comes from R/masst.R.
  cache <- readRDS(sweep_cache)
  ta <- read.delim(sweep_meta, check.names = FALSE, stringsAsFactors = FALSE)
  ta$tmin <- suppressWarnings(as.numeric(ta$ATTRIBUTE_Time_Point_Mins))
  ta <- ta[ta$MS_SampleType == "sample" & !is.na(ta$tmin), ]
  ta$k <- norm_file_key(ta$filename)
  fb   <- basename(as.character(bio_files_per_ds[["MSV000084008"]]))
  bare <- function(x) norm_file_key(to_bare_name(basename(as.character(x)), fb))
  ov_sweep <- if (file.exists("curation/rt_apex_overrides.csv"))
    read.csv("curation/rt_apex_overrides.csv", stringsAsFactors = FALSE) else
    data.frame(dataset_id = character(), role = character(),
               ms2_rt = numeric(), apex_rt = numeric())
  ctr  <- function(role, cr) mean(acceptance_window("MSV000084008", role, cr, bare,
                                                    cr$win, conf_final, ov_sweep))
  par_ctr <- ctr("parent", cache$par); met_ctr <- ctr("metabolite", cache$met)
  # hilic = TRUE gives pure area-select, so it doubles as the "no proximity" control.
  auc_of <- function(cr, centre, W, proximity) {
    data.frame(k = norm_file_key(cr$files),
               auc = vapply(cr$traces, function(tr) {
                 p <- pick(tr$rt, tr$it, centre - W, centre + W, hilic = !proximity)
                 if (is.null(p) || !accept_peak(p, FALSE)) NA_real_ else p$auc
               }, numeric(1)))
  }
  trap <- function(x, y) { o <- order(x); x <- x[o]; y <- y[o]; y[is.na(y)] <- 0
    if (length(unique(x)) < 4) NA_real_ else sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2) }
  rho_of <- function(pa, ma) {
    t2 <- ta
    t2$par_auc <- pa$auc[match(t2$k, pa$k)]; t2$met_auc <- ma$auc[match(t2$k, ma$k)]
    a <- do.call(rbind, lapply(
      split(t2, list(t2$ATTRIBUTE_Subject, t2$ATTRIBUTE_Study_Day), drop = TRUE),
      function(d) data.frame(rank = match(d$PharmGeno_Phenotype_2C19[1], PHENO),
                             lr = log2(trap(d$tmin, d$par_auc) / trap(d$tmin, d$met_auc)),
                             subject = d$ATTRIBUTE_Subject[1])))
    a <- a[is.finite(a$lr), ]; s <- aggregate(lr ~ subject + rank, a, mean)
    suppressWarnings(cor.test(s$rank, s$lr, method = "spearman"))$estimate
  }
  pa <- auc_of(cache$par, par_ctr, 5, TRUE)          # parent fixed at +/-5 s throughout
  sweep <- do.call(rbind, lapply(c(5, 10, 15, 20, 30), function(W) data.frame(
    `metabolite window` = sprintf("[%.0f, %.0f] (+/-%d s)", met_ctr - W, met_ctr + W, W),
    `reaches sulfone?`  = if ((met_ctr + W) >= 181) "yes" else "no",
    `with proximity`    = sprintf("%+.3f", rho_of(pa, auc_of(cache$met, met_ctr, W, TRUE))),
    `without proximity` = sprintf("%+.3f", rho_of(pa, auc_of(cache$met, met_ctr, W, FALSE))),
    check.names = FALSE)))
  cat(knitr::kable(sweep, align = "lccc", row.names = FALSE), sep = "\n")
} else {
  cat("_Pad sweep skipped: needs `", sweep_cache, "` (git-ignored). Run ",
      "`Rscript R/peak_review_panels.R MSV000084008` to rebuild it._\n", sep = "")
}
```

With the proximity weighting the correlation is unchanged at every width; without it,
the ratio collapses once the window admits the sulfone. So the isomer defence is the
on-RT selection, not a tight window; the ±5 s pad only buys drift tolerance.

# Per-dataset extraction

For each dataset, `run_ms1_for_dataset()` loads the MS1 spectra, pulls each
compound's extracted-ion chromatogram (EIC) at the confirmed mass and (widened)
retention-time window, then **picks and integrates the on-RT peak per file** (the
picker described above), applies the manual drop-list, flags files whose peak
width is an outlier, joins the sample metadata, and emits a per-file table
(`sample_tbl`) carrying the `detection_tier` (which of the two compounds were
seen) and the `log2_ratio`. The deep per-dataset diagnostics (EIC overlays,
peak-width plots, principal-component analysis) stay in `dataset-report.qmd`;
this notebook reports only what generalises across the whole cohort.

```{r run-all}
#| echo: false
# Cached per dataset; delete a dataset's RDS to force a re-run.
# The cache is keyed on the extraction window, not merely on the file existing: re-running
# confirmation moves the confirmed RTs and therefore the window, and a file.exists() check
# would serve detections picked from the old one. Deriving the config extracts nothing, so
# it is cheap to verify on every render.
CACHE_DIR <- "artifacts/per_dataset"
dir.create(CACHE_DIR, showWarnings = FALSE, recursive = TRUE)

cfg_is_current <- function(cached, ds) {
  want <- tryCatch(derive_extraction_config(ds, conf_final, rt_pad_s = RT_PAD_S,
                                            bio_files = bio_files_per_ds[[ds]]),
                   error = function(e) NULL)
  if (is.null(want) || is.null(cached$cfg)) return(FALSE)
  k <- c("assay", "adduct_par", "adduct_met", "mz_par", "mz_met",
         "rt_par_win", "rt_met_win")
  isTRUE(all.equal(cached$cfg[k], want[k], tolerance = 1e-6))
}

results <- list()
SKIPPED_DATASETS <- character(0)
for (ds in DATASETS_TO_RUN) {
  cache_path <- file.path(CACHE_DIR, paste0(ds, ".rds"))
  cached <- if (file.exists(cache_path)) readRDS(cache_path) else NULL
  if (!is.null(cached) && !cfg_is_current(cached, ds)) {
    message("  cache for ", ds, " was built on a different extraction window; re-extracting")
    cached <- NULL
  }
  if (!is.null(cached)) {
    message("========== ", ds, " (cached) ==========")
    results[[ds]] <- cached
  } else {
    message("\n========== ", ds, " ==========")
    # Tolerate a dataset whose raw files can't be fetched (e.g. the MassIVE
    # FTP is rate-limited / down): skip it with a logged reason and continue,
    # rather than aborting the whole cross-dataset run. Skipped datasets are
    # simply absent from `results` and get re-extracted on a later render once
    # the source is reachable (their per-dataset cache is only written on
    # success). SKIPPED_DATASETS records what was dropped.
    res <- tryCatch(
      run_ms1_for_dataset(
        dataset_id       = ds,
        conf_final       = conf_final,
        bio_files_per_ds = bio_files_per_ds,
        ds_meta_tbl      = ds_meta_tbl,
        ppm              = PPM,
        rt_pad_s         = RT_PAD_S),
      error = function(e) {
        message("  SKIPPED ", ds, ", extraction failed: ", conditionMessage(e))
        NULL
      })
    if (is.null(res)) { SKIPPED_DATASETS <<- c(SKIPPED_DATASETS, ds); next }
    results[[ds]] <- res
    saveRDS(results[[ds]], cache_path)
  }
}
if (length(SKIPPED_DATASETS))
  message("\n", length(SKIPPED_DATASETS),
          " dataset(s) SKIPPED (raw files unreachable; re-render once the ",
          "source is back): ", paste(SKIPPED_DATASETS, collapse = ", "))
```

## Extraction runtime

```{r}
#| echo: false
# Timing log written by R/peak_review_panels.R; git-ignored, so tolerate its absence.
et_path <- "peaks_review_v2/_extract_times.csv"
et <- if (file.exists(et_path)) {
  x <- read.csv(et_path, stringsAsFactors = FALSE)
  x <- x[order(x$extract_seconds), ]
  x$minutes    <- round(x$extract_seconds / 60, 1)
  x$s_per_file <- round(x$extract_seconds / x$n_files, 2)
  x
} else NULL
et_hours <- if (is.null(et)) NA else round(sum(et$extract_seconds) / 3600, 1)
et_files <- if (is.null(et)) NA else format(sum(et$n_files), big.mark = ",")
chr_ver  <- as.character(utils::packageVersion("Chromatograms"))
```

Loading each file's spectra and pulling its two EICs is the pipeline's dominant compute
cost, and it runs single-threaded. That is measured rather than assumed: parallel
backends gave no gain once the files are read, because the cost is disk I/O rather than
CPU. Cost tracks file count and file size, not the number of detections, and the timings
are compute-only on cached files (raw-file FTP download is excluded). The corpus takes
about **`r et_hours` hours** over `r if (is.null(et)) NA else nrow(et)` datasets
(`r et_files` files).

These numbers were measured with **Chromatograms `r chr_ver`**. Extraction performance is
strongly version-dependent, so treat the table as valid only for that version and
re-measure after any upgrade of the extraction stack.

```{r}
#| echo: false
#| output: asis
if (is.null(et)) {
  cat("_Timing table skipped: `", et_path, "` is not present (git-ignored). Run ",
      "`Rscript R/peak_review_panels.R` to regenerate it._\n", sep = "")
} else {
  cat(knitr::kable(et[, c("dataset", "n_files", "minutes", "s_per_file")],
                   col.names = c("dataset", "files", "min", "s / file"),
                   align = "lrrr", row.names = FALSE), sep = "\n")
}
```

# Checks that can catch us being wrong

The window problem described above went unnoticed for a long time, and the reason
is worth naming: **nothing in the pipeline was capable of contradicting it.** The
numbers looked plausible, and no check existed that would fail if they were wrong.
Robustness, in practice, means deliberately building the things that can catch us.
Four of them, in increasing order of how hard they are to argue with.

**1. The confirmed-file positive control.** In the files where MS2 already proved
the compound is present, MS1 extraction must find it. If it does not, propagation
is losing real signal, and every prevalence number is an underestimate.
`run_ms1_for_dataset()` reports this as `Confirmed files loaded N / with >=1 peak
M`: **M should equal N**, and any shortfall is a bug rather than biology. This is
the cheapest hard check in the pipeline and it needs no external data.

**2. A known-negative control.** The complement of the check above: run the
identical pipeline on files that *cannot* contain the compound and confirm it finds
nothing. The pre-dose plasma draws of the probe study are exactly that — real
matrix, true m/z, true retention window, drug absent by design — and `05` reports
them. A synthetic decoy mass is deliberately **not** used: an untargeted file has
real compounds at arbitrary masses, so a hit at a decoy mass is usually the picker
correctly finding something else, which bounds the false-positive rate from above
without estimating it.

**3. Report the surface, not the best cell.** Every parameter here (`PPM`,
`RT_PAD_S`, `BOUND_PAD_S`, the signal-to-noise cutoff, how below-detection samples
are treated) is a choice, and a conclusion that survives only one setting is not a
conclusion. The retention-time table above is the pattern to follow: a broad
a plateau where the answer is stable, and a cliff with a mechanism that
explains exactly where and why it breaks. A single lucky optimum would not be
evidence.

**4. External ground truth, which beats every internal check.** Parameters can be
tuned until an internal check passes; an outside number cannot be. `04` uses the
one deposit where truth exists independently of us: a cohort in which every
subject was deliberately dosed, with pre-dose samples that are known negatives,
and a published pharmacokinetic value per subject that the pipeline never sees.

A methodological warning that applies to all of this: **do not choose a parameter
by maximising agreement with the thing you then cite as validation.** That is
circular, and it is how the earlier over-wide window justified itself. Where a
parameter must be tuned, tune it against something structurally independent of the
validation target. The parent compound is useful here precisely because it has no
isobaric twin, so its retention-time drift can be measured without reference to
any downstream result.

# Cross-dataset sample table

```{r combine-tables}
#| echo: false
all_samples <- do.call(rbind, lapply(names(results), function(ds) {
  # Rebuild rather than trust the cached sample_tbl: it is fully derivable from
  # metrics_* and costs nothing, so a cache written by an older build_sample_table
  # cannot carry a stale detection_tier into the corpus tables. (It did: an
  # exact-name merge silently demoted 19 detected MetaboLights files to "neither".)
  meta <- tryCatch(load_metadata(ds),
                   error = function(e) data.frame(file = basename(bio_files_per_ds[[ds]])))
  st <- build_sample_table(results[[ds]]$metrics_par, results[[ds]]$metrics_met,
                           basename(bio_files_per_ds[[ds]]), meta, dedup = TRUE)
  # Keep only files still in the current bio_files. A cached extraction can hold rows for
  # files a later QC rule removed (MSV000091520 kept 247 blank_QC files after bio_files went
  # 2391 -> 2144), and without this filter the two sample tables disagree. All of them are
  # detection_tier "neither", so nothing measured is dropped.
  keep <- norm_file_key(st$file) %in% norm_file_key(bio_files_per_ds[[ds]])
  cbind(dataset_id = ds, assay = results[[ds]]$assay, st[keep, , drop = FALSE],
        stringsAsFactors = FALSE)
}))

# Files with no metadata are kept for the presence record but excluded from every
# stratified analysis below.
all_samples$has_metadata <- {
  key <- paste0(all_samples$dataset_id, "|",
                basename(as.character(all_samples$file)))
  hit <- has_metadata_lookup[key]
  ifelse(is.na(hit), FALSE, unname(hit))
}

write.csv(all_samples, "artifacts/sample_table_all.csv", row.names = FALSE)

cat("Datasets processed :", length(results), "\n")
cat("Total rows         :", nrow(all_samples),
    "  (with metadata:", sum(all_samples$has_metadata),
    "| record-only:", sum(!all_samples$has_metadata), ")\n")
```

```{r}
#| echo: false
#| label: tbl-tier-by-dataset
#| tbl-cap: "Detection tier by dataset (metadata-carrying samples only). Assay suffixes are trimmed so the table fits."
short_ds <- function(x) sub("_AN[0-9]+$", "", sub("_LC-MS_.*$", "", x))
ms <- all_samples[all_samples$has_metadata %in% TRUE, ]
tier_tab <- as.data.frame.matrix(table(short_ds(ms$dataset_id), ms$detection_tier))
tier_tab$total <- rowSums(tier_tab)
knitr::kable(tier_tab, align = "r")
```

Every analysis after this point restricts to `has_metadata == TRUE`. The
record-only rows (a confirmed hit on a file the metadata layer does not cover, so
we cannot say which subject or group it belongs to) stay in `sample_table_all.csv`
for the presence record and the propagation counts below, but they carry no
stratum and never enter a contrast.

```{r analysis-set}
#| echo: false
analysis_samples <- all_samples[all_samples$has_metadata %in% TRUE, ]
```

`sample_table_all.csv` is one row per subject (the `build_sample_table()` dedup).
Genotype / longitudinal analyses (`04`) need the rows *before* that dedup, so we
also write a per-file table, keeping `subject_id` (+ a `subject_key` fallback) so
those analyses aggregate to one-per-subject themselves; the rows are not
independent.

```{r combine-tables-perfile}
#| echo: false
getcol <- function(df, nm)
  if (nm %in% names(df)) as.character(df[[nm]]) else rep(NA_character_, nrow(df))

all_samples_perfile <- dplyr::bind_rows(lapply(names(results), function(ds) {
  r    <- results[[ds]]
  meta <- tryCatch(load_metadata(ds),
                   error = function(e) data.frame(file = basename(bio_files_per_ds[[ds]])))
  st   <- build_sample_table(r$metrics_par, r$metrics_met,
                             basename(bio_files_per_ds[[ds]]), meta, dedup = FALSE)
  st$subject_key <- dplyr::coalesce(getcol(st, "subject_id"),
                                    getcol(st, "sample_name"), as.character(st$file))
  cbind(dataset_id = ds, assay = r$assay, st, stringsAsFactors = FALSE)
}))
write.csv(all_samples_perfile, "artifacts/sample_table_perfile.csv", row.names = FALSE)
cat("One row per raw file:", nrow(all_samples_perfile),
    "| one row per subject-sample after collapsing repeat files:", nrow(all_samples), "\n")
```

# Per-dataset contrasts live in `05`

This notebook used to run its own per-dataset Fisher and Kruskal tests here, with a
Benjamini-Hochberg correction, and report the survivors as findings. They are removed.

They tested the **file** as the unit, and a file is not an independent observation: the
probe cohort is 1,137 files from 14 people, so one person could enter a contingency table
as 27 observations. A p-value computed that way does not estimate anything, and no
footnote repairs it. The presence definition was wrong too -- it counted any integrated
peak rather than one that passed the gate -- which inflated every rate.

`05` asks the same questions properly, under the analysis-unit rule (one row per subject
per level of the contrast), records which deposits can support a subject-level unit at
all, and flags the rest as design-unknown rather than testing them anyway. What this
notebook produces is the per-file measurement table; the contrasts belong downstream.

# Per-dataset HTML reports

`dataset-report.qmd` is a parameterised template: given one `dataset_id`, it
produces a deep dive into that dataset alone. The loop below renders it once per
dataset into `reports/report-<dataset_id>.html`.

This is skipped on a normal render, because it is slow (one full quarto render
per dataset) and only the per-dataset narrative depends on it. Render this
notebook with `-P render_reports:true` to regenerate them:

```
quarto render 03-all-datasets.qmd -P render_reports:true
```

```{r render-per-dataset-reports, eval=params$render_reports}
dir.create("reports", showWarnings = FALSE)
for (ds in names(results)) {
  message("Rendering report for ", ds, " …")
  out <- paste0("report-", ds, ".html")
  system2("quarto", c("render", shQuote("dataset-report.qmd"),
                      "-P", shQuote(paste0("dataset_id:", ds)),
                      "--output", shQuote(out)))
  file.rename(out, file.path("reports", out))
}
message("Done: ", length(results), " report(s) in reports/")
```

# Cross-dataset diagnostics

## Ratio distribution per dataset

### Why only the comparison *inside* a dataset is meaningful

A mass spectrometer does not measure concentration. It counts ions, and how many
ions a compound yields depends on how readily that particular molecule ionises,
which is a property of the molecule **and** of the method: the instrument, the
ion source settings, the mobile phase, and the sample matrix all shift it.
Omeprazole and 5-hydroxyomeprazole are different molecules, so they ionise with
different efficiency. What we measure is therefore not the true molar ratio but

```
measured log2(parent / metabolite) = true log2(parent / metabolite) + offset
```

where `offset` is the log ratio of the two compounds' response factors. That offset
is unknown, and it differs for every dataset, because every dataset used a different
instrument and method. The standard way to remove it is to spike an isotope-labelled
internal standard of each compound into every sample, and no public deposit in this
corpus did that, so we cannot calibrate it away.

This is not our assumption; it is measured, established analytical chemistry. Oss et
al. built an experimental electrospray ionization-efficiency scale for 62 organic
compounds and found it spans roughly **six orders of magnitude** between compounds
under identical conditions ([Anal. Chem. 2010, 82(7), 2865-2872](https://doi.org/10.1021/ac902856t)).
Two different molecules therefore do not give comparable signal for equal
concentration, which is exactly the `offset` term. Follow-up work from the same group
shows that scale is also condition- and instrument-dependent (mobile phase, source,
analyser), which is why the offset cannot be assumed constant across deposits
([Rapid Commun. Mass Spectrom. 2021, 35, e9178](https://doi.org/10.1002/rcm.9178)).

Does that six-orders-of-magnitude figure sink the ratio? No, for two reasons. It is
the spread across 62 *unrelated* compounds; for our single fixed pair the offset is
one unknown constant, not a wide range. And the ratio is itself the normalization: it
cancels everything the two compounds share in a sample (injected amount, dilution,
most matrix suppression), leaving only that one constant, which the within-dataset
comparison below then removes.

That drives how the plot below must be read:

- *Within one dataset*, every file went through the same method, so the offset is the
  same constant for all of them. Comparing groups inside a dataset subtracts two
  numbers that share that constant, so it cancels and what remains is real biology.
  This is why the within-dataset tests above, and the per-dataset contrasts `05`
  computes, are legitimate; a rank test like the genotype validation in `04` is
  invariant to the offset entirely, and the HILIC-vs-RP agreement `05` reports (the
  same ratio recovered from two columns whose absolute areas differ up to 117x) is the
  empirical version of the same statement.
- *Between datasets*, the offsets differ by an unknown amount. A dataset whose box
  sits a full unit higher than another's may simply have used a method that ionises
  the parent more efficiently. Reading that gap as biology would be measuring the
  instrument, not the patient.

So compare the spread and the sign within each box; do not rank the datasets by their
vertical position. Each box uses only files where both compounds were cleanly detected
(a peak in both, neither flagged as a peak-width outlier), and the file count is in
each label.

```{r plot-ratio-by-ds, fig.width = 8, fig.height = 5.5}
# Alphabetical on purpose: sorting by median would invite exactly the
# cross-dataset ranking the text above says is meaningless. (short_ds defined above.)
both_all <- analysis_samples |>
  dplyr::filter(detection_tier == "both",
                !par_fwhm_outlier %in% TRUE,
                !met_fwhm_outlier %in% TRUE)

both_lab <- both_all |>
  dplyr::add_count(dataset_id, name = "n") |>
  dplyr::mutate(ds_label = paste0(short_ds(dataset_id), "  (n=", n, ")"))

ggplot(both_lab, aes(x = ds_label, y = log2_ratio)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +
  geom_boxplot(outlier.shape = NA, fill = "grey95") +
  geom_jitter(width = 0.15, alpha = 0.7, size = 1.8) +
  coord_flip() +
  labs(x = NULL,
       y = "log2(omeprazole peak area / 5-OH-omeprazole peak area)",
       title = "Parent / metabolite ratio within each dataset") +
  theme_bw()
```

## Per-dataset detection-tier composition

Every file lands in one of four tiers: **both** compounds detected,
`parent_only`, `metab_only`, or `neither`. This plot shows the share of
each dataset's files in each tier, which is the clearest picture of what kind of
cohort each deposit actually is.

```{r plot-tier-composition, fig.width = 8, fig.height = 4.5}
analysis_samples |>
  dplyr::count(dataset_id, detection_tier) |>
  dplyr::mutate(ds_label = short_ds(dataset_id)) |>
  ggplot(aes(x = ds_label, y = n, fill = detection_tier)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  labs(x = NULL, y = "Share of files",
       fill = "Detection tier",
       title = "Detection-tier composition per dataset") +
  theme_bw() +
  theme(legend.position = "bottom")
```

The share plot is dominated by "neither", which hides how much usable signal each
dataset actually contributes. The second plot drops "neither" and shows the counts
behind the detected files, with the per-dataset total on the right.

```{r plot-tier-counts, fig.width = 8, fig.height = 4.5}
det_counts <- analysis_samples |>
  dplyr::filter(detection_tier != "neither") |>
  dplyr::count(dataset_id, detection_tier) |>
  dplyr::mutate(ds_label = short_ds(dataset_id))

det_totals <- det_counts |>
  dplyr::group_by(ds_label) |>
  dplyr::summarise(total = sum(n), .groups = "drop")

ggplot(det_counts, aes(x = ds_label, y = n, fill = detection_tier)) +
  geom_col() +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5),
            size = 2.9, colour = "white") +
  geom_text(data = det_totals, aes(x = ds_label, y = total, label = total),
            inherit.aes = FALSE, hjust = -0.35, size = 3, colour = "grey30") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_flip() +
  labs(x = NULL, y = "Files with a detection",
       fill = "Detection tier",
       title = "Detected files per dataset (excluding \"neither\")") +
  theme_bw() +
  theme(legend.position = "bottom")
```

### What the composition tells us

**Most files are "neither", and that is the expected result, not a failure.**
These are mostly general-population or disease cohorts, not drug trials: the vast
majority of subjects were never prescribed omeprazole, so there is genuinely
nothing in their files to find. A high "neither" share describes the cohort, not
the pipeline. The deposits where "both" dominates are the drug-probe cohorts
(MSV000082493 and MSV000084008, where every subject was dosed by design), and the
contrast between those two shapes is the clearest illustration of what
reverse-metabolomics is doing: finding the handful of incidentally-dosed people
inside cohorts assembled for entirely unrelated reasons.

**The split between `parent_only` and `metab_only` is informative, not noise.**
A `parent_only` file means we saw the drug but not its metabolite; `metab_only`
is the reverse. Neither is simply a miss. They are **censored** ratio
observations: we know the ratio is very high (parent_only) or very low
(metab_only), we just cannot say by how much, because one of the two peaks fell
below the detection threshold.

**That censoring is why the ratio analysis cannot simply keep the "both" files
and forget the rest.** Restricting to "both" discards precisely the files with
the most extreme ratios, which pulls any estimate toward the middle and
understates real differences. Using them properly means treating them as bounds
rather than pretending the missing peaks are missing at random, which is one of
the open methodological questions `05` sets out.

Taken together with the ratio plot above, the two figures frame the honest
position of this layer: the ratio is only interpretable **within** a dataset, and
within most datasets only a few files support it. Neither figure is meant to rank
the deposits. They exist to show how much signal each cohort contributes and how
much of it is censored, which is what any cross-cohort analysis would have to
account for.
