---
title: "04: CYP2C19 ground-truth validation"
subtitle: "Does the mined omeprazole / 5-OH ratio track genotyped CYP2C19 status?"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
execute:
  warning: false
  message: false
---

# Overview

MSV000082493 and MSV000084008 (Jarmusch et al., *Clin Transl Sci* 2020,
PMC7485946) dosed healthy volunteers with omeprazole as a CYP2C19 probe and carry
the **genotyped** CYP2C19 metaboliser phenotype per subject in
`artifacts/rescue_samples.csv`. The authors reported the identical OME / 5-OH-OME
ratio but never tested it against that genotype, so this is a new use of the
data, and a ground-truth check that the whole cascade
(MASST → SIRIUS → MS1 EIC → ratio) measures real pharmacology.

Expected: low CYP2C19 activity (Poor) → less 5-OH → **high** log2(parent/metab);
Ultrarapid → low. Run after `03` has cached the two datasets.

Joins between the metrics and the submitter table go through `norm_file_key()`, never a
bare basename: the metrics carry cache-prefixed names while the TSV carries bare ones, and
that mismatch has silently produced empty joins several times in this project.

```{r setup}
suppressPackageStartupMessages({ library(dplyr); library(tidyr); library(ggplot2) })
source("R/metadata.R")
source("R/masst.R")   # compound_spec + PHENO, the domain constants
```

# Per-file table + genotype join

`sample_table_all.csv` is deduped to one file per subject, which collapses the
timepoint × specimen structure this readout needs. We rebuild a per-file table
from the cached per-dataset metrics (before dedup), replicating
`build_sample_table()`'s detection rule (shape pass AND robust-SNR ≥ 3), and join
the genotype by filename stem (metrics carry `.mzXML`/prefixed cache names, the
genotype table `.mzXML`, identical stems).

```{r dup-overlap}
#| echo: false
# Counted rather than written into the paragraph below: an earlier hand-written pair
# (222 and 513) survived a rebuild that changed both.
.d <- read.csv("artifacts/sample_table_perfile.csv", check.names = FALSE)
.k93 <- norm_file_key(.d$file[.d$dataset_id == "MSV000082493"])
.k08 <- norm_file_key(.d$file[.d$dataset_id == "MSV000084008"])
N_SHARED <- length(intersect(.k93, .k08)); N_84008 <- length(.k08)
```

The two accessions are one study: MSV000082493 re-deposits `r N_SHARED` of
MSV000084008's `r N_84008` plasma files, so a per-file table built from both carries each
of those twice and the join propagates the duplicate. Every rate would then be computed on
a set biased toward the re-deposited files. Plasma is therefore attributed to
MSV000084008 alone; those files carry no metadata under the other accession in any case.

```{r per-file-join}
cyp_ds <- c("MSV000082493", "MSV000084008")
rs <- read.csv("artifacts/rescue_samples.csv", check.names = FALSE, stringsAsFactors = FALSE)

# CYP2C19 activity order (low -> high); the ratio should DECREASE along it
# PHENO (CYP2C19 activity order) comes from R/masst.R.
rs$pheno_rank <- match(rs$cyp2c19_phenotype, PHENO)

bare <- function(x)  # strip MSV000XXXXXX_ cache prefix + extension, lowercase
  sub("^msv[0-9]{9}_", "", tolower(tools::file_path_sans_ext(basename(x))))
rs$stem <- bare(rs$file)

per_file_tier <- function(d) {
  r <- readRDS(sprintf("artifacts/per_dataset/%s.rds", d))
  m <- merge(r$metrics_par, r$metrics_met, by = "file", all = TRUE)
  # Detection is decided in the picker (accept_peak + drop-list); metrics_* carry it as a flag.
  par_det <- m$par_detected %in% TRUE
  met_det <- m$met_detected %in% TRUE
  m$detection_tier <- ifelse( par_det &  met_det, "both",
                       ifelse( par_det & !met_det, "parent_only",
                       ifelse(!par_det &  met_det, "metab_only", "neither")))
  m$log2_ratio <- ifelse(m$detection_tier == "both",
                         log2(m$par_auc / m$met_auc), NA_real_)
  m$dataset_id <- d
  m$stem <- bare(m$file)
  m
}
pf <- bind_rows(lapply(cyp_ds, per_file_tier))

j <- pf |>
  inner_join(rs[, c("stem", "deposit_id", "cyp2c19_phenotype", "cyp2c19_genotype",
                    "pheno_rank", "subject_id", "timepoint", "specimen")],
             by = "stem")

.dup <- j$dataset_id == "MSV000082493" & grepl("plasma", tolower(j$specimen))
if (any(.dup)) {
  cat("dropped", sum(.dup), "re-deposited plasma rows from MSV000082493\n")
  j <- j[!.dup, ]
}
stopifnot(!any(duplicated(j$stem)))
both <- j |> filter(detection_tier == "both", is.finite(log2_ratio),
                    !par_fwhm_outlier %in% TRUE, !met_fwhm_outlier %in% TRUE)
cat("joined files:", nrow(j), " | with both compounds + clean ratio:", nrow(both), "\n")
table(j$dataset_id, j$detection_tier)
```

# Ratio vs genotyped phenotype

Per cohort: median ratio by phenotype, file-level Spearman/Kruskal-Wallis, and the
valid **subject-level** Spearman (one ratio per subject, no pseudoreplication).

```{r phenotype-tests}
test_cohort <- function(d, label) {
  x <- both |> filter(if (is.null(d)) TRUE else deposit_id == d, !is.na(pheno_rank))
  if (nrow(x) < 8 || dplyr::n_distinct(x$cyp2c19_phenotype) < 3) {
    cat("\n[", label, "] too few points (n=", nrow(x), ")\n"); return(invisible())
  }
  cat("\n----------", label, " (n =", nrow(x), "files) ----------\n")
  ph <- x |> group_by(cyp2c19_phenotype) |>
    summarise(n = n(), med = round(median(log2_ratio), 2), .groups = "drop") |>
    arrange(match(cyp2c19_phenotype, PHENO))
  print(as.data.frame(ph))
  # SUBJECT level only. A per-file Spearman used to be printed here too, and it read
  # p = 5e-17 on 14 people -- a file is not an independent observation, so that number
  # measured how often each volunteer was injected, not their genotype. The per-file
  # medians printed above stay, as description.
  subj <- x |> group_by(subject_id, cyp2c19_phenotype, pheno_rank) |>
    summarise(ratio = median(log2_ratio), .groups = "drop")
  if (nrow(subj) >= 8 && dplyr::n_distinct(subj$cyp2c19_phenotype) >= 3) {
    sp2 <- suppressWarnings(cor.test(subj$ratio, subj$pheno_rank, method = "spearman"))
    kw2 <- kruskal.test(ratio ~ factor(cyp2c19_phenotype), subj)
    cat(sprintf("subject level (n = %d): Spearman rho = %.3f  p = %.3g  (expect NEGATIVE)\n",
                nrow(subj), sp2$estimate, sp2$p.value))
    cat(sprintf("Kruskal-Wallis across phenotypes p = %.3g\n", kw2$p.value))
  } else cat("too few subjects for a subject-level test\n")
  invisible(x)
}
x84 <- test_cohort("MSV000084008", "084008 PLASMA (classic CYP readout)")
x82 <- test_cohort("MSV000082493", "082493 urine/feces/skin")
invisible(test_cohort(NULL, "POOLED CYP cohorts"))
```

# Per-subject AUC metabolic ratio (the clinical readout)

Everything above uses **one ratio per subject** taken from single files. That is not
how CYP2C19 phenotyping is done clinically: the accepted metric is the **ratio of
areas under the concentration-time curves** over the dosing interval,
AUC(omeprazole) / AUC(5-OH-omeprazole). Integrating removes the dependence on *when*
the sample happened to be drawn, which in a single-timepoint ratio is pure noise:
one subject sampled near their peak and another past it differ for reasons that have
nothing to do with genotype.

MSV000084008 supports this directly: 461 plasma files, 14 subjects, 8-9 timepoints
per subject on each of two study days. The timepoint comes from the submitter's
`ATTRIBUTE_Time_Point_Mins` (recovered by the metadata union in `01`; Pan-ReDU does
not carry it).

Two conventions worth stating because they change the answer:

- **Days are integrated separately.** Day 1 and day 9 straddle the antibiotic
  exposure, so they are different physiological states and pooling them would blur
  the contrast the study was designed around.
- **A non-detected timepoint contributes 0** to the AUC, the standard
  pharmacokinetic below-limit-of-detection convention. "Non-detected" means the
  timepoint fails the picker's detection gate (`detection_tier`), not merely that no
  peak was integrated: a timepoint where only noise was picked is set to 0, not
  allowed to contribute its raw integrated area. Dropping those points instead would
  bias the AUC upward exactly for the poor metabolisers whose metabolite is genuinely
  absent.

The direction is predicted before looking: poor metabolisers clear omeprazole slowly,
so a lower activity rank should give a higher parent/metabolite ratio, i.e.
Spearman rho < 0.

```{r auc-ratio}
auc_tsv <- "dataset_metadata/MSV000084008_metadata.tsv"
if (!file.exists(auc_tsv)) {
  cat("submitter TSV not present - AUC section skipped\n")
} else {
  r_auc <- readRDS("artifacts/per_dataset/MSV000084008.rds")
  mp <- as.data.frame(r_auc$metrics_par); mm <- as.data.frame(r_auc$metrics_met)
  ta <- read.delim(auc_tsv, check.names = FALSE, stringsAsFactors = FALSE)
  # normalised key, never a bare basename: the metrics carry cache-prefixed names
  ta$k <- norm_file_key(ta$filename)
  mp$k <- norm_file_key(mp$file); mm$k <- norm_file_key(mm$file)
  # Gate to detection: a timepoint that fails the picker's detection gate is below-LOD
  # (NA here -> 0 in trap()), not its raw integrated noise-area. Detected points keep their AUC.
  mp_i <- match(ta$k, mp$k); mm_i <- match(ta$k, mm$k)
  ta$par_auc <- ifelse(mp$par_detected[mp_i] %in% TRUE, mp$par_auc[mp_i], NA_real_)
  ta$met_auc <- ifelse(mm$met_detected[mm_i] %in% TRUE, mm$met_auc[mm_i], NA_real_)
  # apex retention times, kept for the isomer restrictions below
  ta$par_rt <- mp$par_apex_rt[mp_i]; ta$met_rt <- mm$met_apex_rt[mm_i]
  ta$tmin <- suppressWarnings(as.numeric(ta$ATTRIBUTE_Time_Point_Mins))
  ta <- ta[ta$MS_SampleType == "sample" & !is.na(ta$tmin), ]

  trap <- function(x, y) {              # trapezoidal AUC; NA -> 0 (below LOD)
    o <- order(x); x <- x[o]; y <- y[o]; y[is.na(y)] <- 0
    if (length(unique(x)) < 4) return(NA_real_)
    sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)
  }
  auc <- do.call(rbind, lapply(
    split(ta, list(ta$ATTRIBUTE_Subject, ta$ATTRIBUTE_Study_Day), drop = TRUE),
    function(d) data.frame(
      subject = d$ATTRIBUTE_Subject[1], day = d$ATTRIBUTE_Study_Day[1],
      n_tp = length(unique(d$tmin)),
      auc_par = trap(d$tmin, d$par_auc), auc_met = trap(d$tmin, d$met_auc),
      phenotype = d$PharmGeno_Phenotype_2C19[1], stringsAsFactors = FALSE)))
  auc$log2_ratio <- log2(auc$auc_par / auc$auc_met)
  auc <- auc[is.finite(auc$log2_ratio), ]
  auc$rank <- match(auc$phenotype, PHENO)

  cat("timepoints per subject-day: median", median(auc$n_tp),
      "range", min(auc$n_tp), "-", max(auc$n_tp), "\n\n")
  print(auc[order(auc$rank, auc$subject),
            c("subject","day","n_tp","log2_ratio","phenotype")],
        row.names = FALSE, digits = 3)

  cat("\nmean log2(AUC parent / AUC metabolite) by phenotype:\n")
  subj <- aggregate(log2_ratio ~ subject + phenotype + rank, data = auc, FUN = mean)
  print(tapply(subj$log2_ratio, factor(subj$phenotype, levels = PHENO), mean))

  cat("\n")
  for (dd in sort(unique(auc$day))) {
    s <- auc[auc$day == dd, ]
    ct <- suppressWarnings(cor.test(s$rank, s$log2_ratio, method = "spearman"))
    cat(sprintf("day %-3s n=%2d | Spearman rho = %+.3f  p = %.4f\n",
                dd, nrow(s), ct$estimate, ct$p.value))
  }
  ct <- suppressWarnings(cor.test(subj$rank, subj$log2_ratio, method = "spearman"))
  cat(sprintf("both days, per subject  n=%2d | Spearman rho = %+.3f  p = %.4f\n",
              nrow(subj), ct$estimate, ct$p.value))
}
```

## Is it one subject doing the work?

Three phenotype classes have a single subject each, so a strong correlation could be
carried entirely by one person. Two checks: drop each subject in turn, and then throw
away all three singleton classes and test only the two populated ones.

```{r auc-robustness}
if (exists("subj") && nrow(subj) > 5) {
  full <- suppressWarnings(cor.test(subj$rank, subj$log2_ratio, method = "spearman"))
  cat(sprintf("full            n=%2d rho=%+.3f p=%.4f\n\n",
              nrow(subj), full$estimate, full$p.value))
  for (i in seq_len(nrow(subj))) {
    x <- subj[-i, ]
    ct <- suppressWarnings(cor.test(x$rank, x$log2_ratio, method = "spearman"))
    cat(sprintf("drop %-7s (%-24s) rho=%+.3f p=%.4f%s\n",
                subj$subject[i], substr(subj$phenotype[i], 1, 24),
                ct$estimate, ct$p.value,
                if (ct$p.value > 0.05) "   <-- loses significance" else ""))
  }
  two <- subj[subj$phenotype %in% c("Intermediate Metabolizer", "Normal Metabolizer"), ]
  if (dplyr::n_distinct(two$phenotype) == 2) {
    w <- suppressWarnings(wilcox.test(log2_ratio ~ phenotype, data = two))
    cat(sprintf("\nsingletons removed - Intermediate (n=%d) vs Normal (n=%d): p = %.3f\n",
                sum(two$phenotype == "Intermediate Metabolizer"),
                sum(two$phenotype == "Normal Metabolizer"), w$p.value))
  }
}
```

What to take from this: the AUC and the single-timepoint ratio use **the same
subjects, the same files and the same extraction**. Only the summary statistic
differs. If the AUC separates phenotypes where the single-timepoint ratio did not,
the earlier null was a measurement-noise problem, not evidence that the readout
fails - and the honest conclusion is about the metric, not about a bigger cohort.

Limits that remain: n = 14 subjects; three classes have one subject each; the
Intermediate-vs-Normal comparison alone is marginal. MSV000082493 holds plasma files
for these **same** 14 subjects, so checking there is a batch-level consistency check,
not independent replication.

## Is an isomer being integrated instead?

Fragmentation does not separate 5-OH-omeprazole from omeprazole sulfone, and an isomer
of omeprazole itself is reported in these samples. Chromatography does separate them:
each mass carries two RT clusters (parent ~157 and ~167 s, metabolite ~157 and ~182 s)
and the confirmed anchors place omeprazole in the *later* parent cluster and
5-OH-omeprazole in the *earlier* metabolite one. If an isomer were being integrated,
restricting to the correct cluster should change the answer.

A restriction can only remove a **detection**, never a file: an apex in the wrong
cluster makes that channel unusable at that timepoint, which is the below-LOD case the
AUC already handles by contributing 0.

```{r isomer-restrictions}
if (!exists("ta") || !"par_rt" %in% names(ta)) {
  cat("AUC section did not run - isomer restrictions skipped\n")
} else {
  PC <- 162; MC <- 163                      # cluster split, parent / metabolite
  a <- ta[!is.na(ta$par_auc) & !is.na(ta$met_auc), ]
  pmed <- median(a$par_rt, na.rm = TRUE); mmed <- median(a$met_rt, na.rm = TRUE)

  fit_restricted <- function(keep_par, keep_met, label) {
    d <- ta
    d$par_auc[!keep_par] <- NA_real_
    d$met_auc[!keep_met] <- NA_real_
    ac <- do.call(rbind, lapply(
      split(d, list(d$ATTRIBUTE_Subject, d$ATTRIBUTE_Study_Day), drop = TRUE),
      function(z) data.frame(
        subject = z$ATTRIBUTE_Subject[1],
        l2 = log2(trap(z$tmin, z$par_auc) / trap(z$tmin, z$met_auc)),
        phenotype = z$PharmGeno_Phenotype_2C19[1], stringsAsFactors = FALSE)))
    ac <- ac[is.finite(ac$l2), ]
    s <- aggregate(l2 ~ subject + phenotype, data = ac, FUN = mean)
    s$rank <- match(s$phenotype, PHENO)
    ct <- suppressWarnings(cor.test(s$rank, s$l2, method = "spearman"))
    m <- summary(lm(l2 ~ rank, data = s))
    cat(sprintf("  %-40s files %3d subj %2d | rho %+.3f p %.4f | slope %+.3f\n",
                label, sum(!is.na(d$par_auc) & !is.na(d$met_auc)), nrow(s),
                ct$estimate, ct$p.value, m$coefficients[2, 1]))
  }
  ok <- function(x, cond) is.na(x) | cond   # undetected apexes are already NA

  fit_restricted(TRUE, TRUE, "all accepted")
  fit_restricted(ok(ta$par_rt, ta$par_rt >= PC), TRUE, "parent from the late cluster only")
  fit_restricted(TRUE, ok(ta$met_rt, ta$met_rt <= MC), "metabolite inside its window only")
  fit_restricted(ok(ta$par_rt, ta$par_rt >= PC),
                 ok(ta$met_rt, ta$met_rt <= MC), "both channels unambiguous")
  fit_restricted(ok(ta$par_rt, abs(ta$par_rt - pmed) <= 2),
                 ok(ta$met_rt, abs(ta$met_rt - mmed) <= 2),
                 "both within 2 s of the cohort median")

  cat(sprintf("\nparent apex in the EARLY cluster: %d of %d\n",
              sum(a$par_rt < PC, na.rm = TRUE), nrow(a)))
  cat(sprintf("metabolite apex outside its window: %d\n", sum(a$met_rt > MC, na.rm = TRUE)))
  cat(sprintf("per-subject median apex spread: parent %.1f s, metabolite %.1f s\n",
              diff(range(tapply(a$par_rt, a$ATTRIBUTE_Subject, median), na.rm = TRUE)),
              diff(range(tapply(a$met_rt, a$ATTRIBUTE_Subject, median), na.rm = TRUE))))
}
```

# The linear model against genotype

Everything above uses Spearman, which asks only whether the ratio decreases as activity
increases. The chapter reports a linear model instead, because a rank correlation carries
no units and a claim about a measurement needs an effect size in interpretable ones.

The classes are ranked poor = 1 to ultrarapid = 5 and the subject-level log2 ratio is
regressed on that rank, one row per volunteer. The coefficient is a change **per genotype
class**, not per hour. Treating the rank as a number assumes the classes are equally
spaced, which nothing guarantees; that is the price of an interpretable coefficient.

```{r lm-genotype}
subj_auc <- aggregate(log2_ratio ~ subject + phenotype + rank, data = auc, FUN = mean)
m_auc <- lm(log2_ratio ~ rank, data = subj_auc)
summary(m_auc)
cat(sprintf("\nfold change per class: %.3f\n", 2^coef(m_auc)[2]))
```

## Does the summary statistic matter?

Replacing the integral with a plain mean of each volunteer's file ratios should give the
same answer if the conclusion is not an artefact of how the repeated measurements were
collapsed.

```{r lm-mean-of-files}
mof <- both |>
  filter(deposit_id == "MSV000084008", !is.na(pheno_rank)) |>
  group_by(subject_id, pheno_rank) |>
  summarise(v = mean(log2_ratio), .groups = "drop")
summary(lm(v ~ pheno_rank, data = mof))$coefficients
```

## Does timing produce the result?

The ratio decays across the dosing interval, so a difference between volunteers could in
principle be a difference in when their blood was drawn. Refitting the same model at one
fixed draw at a time removes that: every volunteer is then compared at the same point
after dosing.

The submitter TSV carries duplicate column names, which dplyr refuses to transform, so a
minimal frame is built from it once and reused.

```{r lm-draws-frame}
draws <- data.frame(
  subject = ta$ATTRIBUTE_Subject,
  day     = ta$ATTRIBUTE_Study_Day,
  rank    = match(ta$PharmGeno_Phenotype_2C19, PHENO),
  tmin    = ta$tmin,
  l2      = log2(ta$par_auc / ta$met_auc),
  stringsAsFactors = FALSE)
draws <- draws[is.finite(draws$l2) & !is.na(draws$rank), ]
```

```{r lm-fixed-draw}
fixed_draw <- function(tp) {
  x <- draws |> filter(tmin == tp) |>
    group_by(subject, rank) |> summarise(v = mean(l2), .groups = "drop")
  co <- summary(lm(v ~ rank, data = x))$coefficients
  data.frame(draw_min = tp, n = nrow(x), beta = co[2, 1], se = co[2, 2], p = co[2, 4])
}
do.call(rbind, lapply(c(120, 240, 300, 360), fixed_draw))
```

## Each dosing occasion separately

Day 9 follows the seven-day antibiotic course, so it is worth showing that it still
reports the same phenotype rather than assuming it.

```{r lm-per-occasion}
per_day <- do.call(rbind, lapply(sort(unique(auc$day)), function(d) {
  x <- auc[auc$day == d & is.finite(auc$log2_ratio), ]
  co <- summary(lm(log2_ratio ~ rank, data = x))$coefficients
  data.frame(occasion = d, n = nrow(x), beta = co[2, 1], p = co[2, 4])
}))
both_days <- summary(lm(log2_ratio ~ rank, data = subj_auc))$coefficients
rbind(per_day, data.frame(occasion = "mean of both", n = nrow(subj_auc),
                          beta = both_days[2, 1], p = both_days[2, 4]))

w <- auc |> select(subject, day, log2_ratio) |> filter(is.finite(log2_ratio)) |>
  tidyr::pivot_wider(names_from = day, values_from = log2_ratio)
w <- w[complete.cases(w), ]
cat(sprintf("\nwithin-subject change across the antibiotic course: median %+.2f log2, n = %d\n",
            median(w[[3]] - w[[2]]), nrow(w)))
print(wilcox.test(w[[3]], w[[2]], paired = TRUE))
```

## Leave-one-volunteer-out

Three classes contain a single volunteer, so the model is refitted fourteen times, each
fit omitting one.

```{r lm-jackknife}
jk <- do.call(rbind, lapply(seq_len(nrow(subj_auc)), function(i) {
  co <- summary(lm(log2_ratio ~ rank, data = subj_auc[-i, ]))$coefficients
  data.frame(dropped = subj_auc$subject[i], phenotype = subj_auc$phenotype[i],
             beta = co[2, 1], p = co[2, 4])
}))
jk[order(jk$beta), ]
cat(sprintf("\nbeta range %+.3f to %+.3f | worst p %.4f\n",
            min(jk$beta), max(jk$beta), max(jk$p)))
```

## The elimination rate as a second read-out

How fast the ratio falls is itself genotype-dependent: a poor metaboliser clears
omeprazole slowly, so their ratio should decay more gently. This is a natural second
endpoint, and this cohort cannot support it.

```{r lm-elimination-rate}
slopes <- draws |>
  group_by(subject, day, rank) |> filter(n_distinct(tmin) >= 3) |>
  summarise(b = coef(lm(l2 ~ I(tmin / 60)))[2], .groups = "drop") |>
  group_by(subject, rank) |> summarise(b = mean(b), .groups = "drop")
summary(lm(b ~ rank, data = slopes))$coefficients
slopes |> group_by(rank) |> summarise(n = n(), median_slope = round(median(b), 3))
```

# Genotype dosage (star alleles)

CYP2C19 allele activity: *1 = 1 (normal), *2/*3 = 0 (no function), *17 = 1.5
(increased). The activity score respects functional spacing the equal-spaced rank
ignores; the loss-of-function dose isolates the strong *2/*3 effect from the noisy
*17. Subject level.

```{r genotype-dosage}
allele_act <- function(a) ifelse(a %in% c("2", "3"), 0,
                          ifelse(a == "17", 1.5, ifelse(a == "1", 1, NA_real_)))
lof_allele <- function(a) as.integer(a %in% c("2", "3"))
add_geno <- function(df) {
  parts <- strsplit(as.character(df$cyp2c19_genotype), "_")
  df$act_score <- vapply(parts, function(p) sum(allele_act(p)), numeric(1))
  df$lof_dose  <- vapply(parts, function(p) sum(lof_allele(p)), integer(1))
  df
}
gp <- add_geno(both |> filter(deposit_id == "MSV000084008", !is.na(cyp2c19_genotype)))
gsub <- gp |> group_by(subject_id, cyp2c19_genotype, act_score, lof_dose) |>
  summarise(ratio = median(log2_ratio), n_files = n(), .groups = "drop") |>
  arrange(act_score)
if (nrow(gsub) >= 8) {
  print(as.data.frame(gsub))
  s1 <- suppressWarnings(cor.test(gsub$ratio, gsub$act_score, method = "spearman"))
  cat(sprintf("Spearman(subject ratio, activity_score) rho = %+.3f  p = %.3g  (expect NEG)\n",
              s1$estimate, s1$p.value))
  gsub$carrier <- gsub$lof_dose > 0
  if (dplyr::n_distinct(gsub$carrier) == 2) {
    w <- suppressWarnings(wilcox.test(ratio ~ carrier, gsub))
    cat(sprintf("LoF carriers *2/*3 (n=%d) median = %.2f  vs  non-carriers (n=%d) median = %.2f;  Wilcoxon p = %.3g\n",
                sum(gsub$carrier), median(gsub$ratio[gsub$carrier]),
                sum(!gsub$carrier), median(gsub$ratio[!gsub$carrier]), w$p.value))
  }
}
```

# Specimen-resolved phenotyping (082493)

Hepatic CYP2C19 is classically read from plasma: can a non-invasive matrix
(urine / skin / feces) recover the genotype? Subject-level is the honest test.

```{r specimen-resolved}
sp82 <- both |> filter(deposit_id == "MSV000082493", !is.na(pheno_rank))
if (nrow(sp82) > 0) for (sp in sort(unique(sp82$specimen))) {
  z <- sp82 |> filter(specimen == sp)
  if (nrow(z) >= 8 && dplyr::n_distinct(z$cyp2c19_phenotype) >= 3) {
    s_ <- suppressWarnings(cor.test(z$log2_ratio, z$pheno_rank, method = "spearman"))
    zs <- z |> group_by(subject_id, pheno_rank) |>
      summarise(r = median(log2_ratio), .groups = "drop")
    sub_txt <- if (nrow(zs) >= 8 && dplyr::n_distinct(zs$pheno_rank) >= 3) {
      ss <- suppressWarnings(cor.test(zs$r, zs$pheno_rank, method = "spearman"))
      sprintf("subj-level n=%d rho=%+.3f p=%.3g", nrow(zs), ss$estimate, ss$p.value)
    } else sprintf("subj-level n=%d (underpowered)", nrow(zs))
    cat(sprintf("  %-6s file n=%3d rho=%+.3f p=%.3g   |   %s\n",
                sp, nrow(z), s_$estimate, s_$p.value, sub_txt))
  } else cat(sprintf("  %-6s n=%3d  (too few / <3 phenotypes)\n", sp, nrow(z)))
}
```

# Controls the probe cohort makes possible

The dosed cohort supports three checks that no other deposit can offer, because it is
the only place where we know what the answer must be before looking.

## Pre-dose samples: biological negatives

Every subject was drawn at minute 0, before the dose. The drug cannot be present. These
are real plasma from real people, with the full complexity of the matrix, so they are a
harder test of specificity than any synthetic decoy.

```{r pre-dose}
#| echo: false
ta_f <- "dataset_metadata/MSV000084008_metadata.tsv"
if (!file.exists(ta_f)) { cat("submitter TSV not present\n") } else {
  ta <- read.delim(ta_f, check.names = FALSE, stringsAsFactors = FALSE)
  ta$k <- norm_file_key(ta$filename)
  r84 <- readRDS("artifacts/per_dataset/MSV000084008.rds")
  mp <- as.data.frame(r84$metrics_par); mm <- as.data.frame(r84$metrics_met)
  mp$k <- norm_file_key(mp$file); mm$k <- norm_file_key(mm$file)
  ta$par_det <- mp$par_detected[match(ta$k, mp$k)] %in% TRUE
  ta$met_det <- mm$met_detected[match(ta$k, mm$k)] %in% TRUE
  ta$tmin <- suppressWarnings(as.numeric(ta$ATTRIBUTE_Time_Point_Mins))
  z <- ta[ta$MS_SampleType == "sample" & !is.na(ta$tmin), ]
  tb <- data.frame(
    group = c("pre-dose (t = 0)", "post-dose (t > 0)"),
    files = c(sum(z$tmin == 0), sum(z$tmin > 0)),
    parent = c(sum(z$par_det[z$tmin == 0]), sum(z$par_det[z$tmin > 0])),
    metabolite = c(sum(z$met_det[z$tmin == 0]), sum(z$met_det[z$tmin > 0])))
  tb$parent_pct <- sprintf("%.1f%%", 100 * tb$parent / tb$files)
  knitr::kable(tb, row.names = FALSE, caption = "Detections before and after dosing.")
  ft <- fisher.test(matrix(c(tb$parent[1], tb$files[1] - tb$parent[1],
                             tb$parent[2], tb$files[2] - tb$parent[2]), nrow = 2))
  cat(sprintf("\nFisher pre vs post (parent): p = %.3g\n", ft$p.value))
}
```


**Zero detections in 49 pre-dose files, for either compound**, against 60% after dosing
(p = 2e-18). The picker does not invent the drug in samples that cannot contain it.

## The ratio moves with time since dose

A metabolic ratio is not constant over a dosing interval; that is precisely why clinical
protocols fix the draw at 3 h. If our ratio is measuring pharmacology rather than noise,
it must fall as the parent clears faster than its metabolite. Nothing in the pipeline was
tuned to produce this, and the test is run **within subject**, so between-subject
phenotype differences cannot generate it.

```{r ratio-vs-time}
#| echo: false
# Build a minimal frame rather than mutating the submitter TSV: that file carries
# duplicate column names, which dplyr refuses to transform.
if (exists("z")) {
  mpa <- as.data.frame(r84$metrics_par); mma <- as.data.frame(r84$metrics_met)
  mpa$k <- norm_file_key(mpa$file); mma$k <- norm_file_key(mma$file)
  tt <- data.frame(
    subject = as.character(z$ATTRIBUTE_Subject),
    day     = as.character(z$ATTRIBUTE_Study_Day),
    tmin    = z$tmin,
    par_det = z$par_det, met_det = z$met_det,
    par_auc = mpa$par_auc[match(z$k, mpa$k)],
    met_auc = mma$met_auc[match(z$k, mma$k)],
    stringsAsFactors = FALSE)
  tt <- tt[tt$par_det & tt$met_det & tt$tmin > 0, ]
  tt$l2 <- log2(tt$par_auc / tt$met_auc)
  tt <- tt[is.finite(tt$l2), ]
  tt$bin <- cut(tt$tmin, c(0, 30, 60, 120, 240, 480, Inf),
                labels = c("<=30m", "30-60m", "1-2h", "2-4h", "4-8h", ">8h"))
  agg <- do.call(rbind, lapply(levels(tt$bin), function(b) {
    q <- tt[tt$bin == b, ]; if (!nrow(q)) return(NULL)
    data.frame(bin = b, files = nrow(q), subjects = length(unique(q$subject)),
               median_log2 = round(median(q$l2), 2))
  }))
  print(agg, row.names = FALSE)
  rho <- vapply(split(tt, tt$subject), function(q)
    if (nrow(q) >= 4 && length(unique(q$tmin)) >= 4)
      suppressWarnings(cor(q$tmin, q$l2, method = "spearman")) else NA_real_, numeric(1))
  rho <- rho[is.finite(rho)]
  bt <- binom.test(sum(rho < 0), length(rho))
  cat(sprintf("
within-subject Spearman(time, ratio): median rho = %+.3f
", median(rho)))
  cat(sprintf("%d of %d subjects negative, sign test p = %.4g
",
              sum(rho < 0), length(rho), bt$p.value))
}
```


The ratio falls monotonically across the dosing interval and
**`r if (exists("rho")) sum(rho < 0) else NA` of `r if (exists("rho")) length(rho) else NA`
subjects show a negative slope individually**
(sign test p = `r if (exists("bt")) signif(bt$p.value, 2) else NA`). This is ordinary pharmacokinetics
recovered from public files, and it is the clearest evidence that the number behaves like
a drug measurement rather than an artefact of the extraction.

It also explains why the single-timepoint ratio was null against genotype while the AUC
works: a ratio that moves this much with sampling time carries the draw time as noise.

## Does the number agree with the published one?

The source paper measured the same quantity by targeted MS/MS. We measure it from
untargeted MS1 with no internal standard, so the two should agree in magnitude and
direction, not exactly.

```{r vs-published}
#| echo: false
if (exists("tt") && nrow(tt) > 10) {
  out <- lapply(sort(unique(tt$day)), function(dd) {
    q <- tt[tt$day == dd, ]
    sj <- do.call(rbind, lapply(split(q, q$subject), function(u)
      data.frame(l = median(u$l2))))
    data.frame(study_day = dd, subjects = nrow(sj), ours = round(2^mean(sj$l), 2))
  })
  o <- do.call(rbind, out); o$published <- c(1.4, 1.2)[seq_len(nrow(o))]
  knitr::kable(o, row.names = FALSE,
               caption = "Geometric-mean OME / 5-OH-OME ratio: ours (MS1) vs published (targeted MS/MS).")
}
```

Our values sit about 40% above theirs and move in the same direction between periods.
That offset is what an uncalibrated response factor looks like: without an isotope-labelled
internal standard the absolute level is not expected to match, while the ordering and the
within-study contrast are. This is a method-transfer agreement, not a reproduction of
their exact number.

# Why the cefprozil period is not analysed here

The deposit is a two-period design with cefprozil given between day 1 and day 9, so it
looks like a free reproduction target. It is not, and the reason is worth recording.

The source paper defines its CYP2C19 readout the same way we do, parent over metabolite
(Methods: "parent:metabolite area under the curve (AUC) ratios"; Table 1 row
"OME/5-OH OME (AUC)"). Its reported values are **1.4 before the antibiotic and 1.2 after,
LS-GMR 0.89, 90% CI 0.63 to 1.17**. Two things follow. The confidence interval **contains
1**, so no change in CYP2C19 was established; the interval is merely wider than the
0.8 to 1.25 bioequivalence window, which says the data cannot exclude an effect, not that
there is one. And the ratio moved **down**, which for a parent/metabolite ratio means
relatively more metabolite, i.e. *higher* apparent activity, while the paper concludes a
*decrease* in activity. That reading is right for their CYP1A2 and CYP3A4 endpoints, which
are oral clearances, but a metabolic ratio inverts the direction.

Our own paired test gives +0.20 log2 (11 subjects, p = 0.52): also not significant.

So neither side has an effect to compare, and stating agreement or disagreement would mean
adjudicating an inconsistency in someone else's table on the strength of 11 subjects. The
comparison is dropped rather than reported as a null.


# MS1 propagation gain

How many extra genotype-linked measurements the MS1 EIC recovers beyond what the
targeted MS2 data already provided.

One caution about the denominator, because it is easy to inflate this number. The
**SIRIUS-confirmed seed files are not the MS2 baseline**: they are a capped, curated
subset sent for confirmation, so dividing MS1 detections by seeds compares two
different populations and overstates the gain several-fold. The honest baseline is
the set of files MASST matched, which is what `05` Part A reports corpus-wide. Those
figures are quoted there rather than repeated here, because a number copied between
notebooks is a number that will eventually disagree with itself. Seeds are reported here
only to show how few confirmed anchors the propagation starts from.

```{r propagation-gain}
for (d in cyp_ds) {
  r    <- readRDS(sprintf("artifacts/per_dataset/%s.rds", d))
  seed <- length(unique(basename(c(r$cfg$confirmed_files_parent,
                                   r$cfg$confirmed_files_metabolite))))
  detf  <- sum(pf$dataset_id == d & pf$detection_tier != "neither", na.rm = TRUE)
  bothf <- sum(pf$dataset_id == d & pf$detection_tier == "both",    na.rm = TRUE)
  cat(sprintf("  %s: %d confirmed anchors -> %d files with a detection (%d with both)\n",
              d, seed, detf, bothf))
}
```

# Figures

Boxplot of the ratio by genotyped phenotype (both cohorts), plus the two
deck figures: the plasma trend line and the propagation lollipop. Also writes
`artifacts/cyp_validation_data.csv`.

```{r figure-data}
if (!dir.exists("figures")) dir.create("figures")
PH_SHORT <- c("Poor", "Inter.", "Normal", "Rapid", "Ultra")
pal <- c("Poor Metabolizer" = "#B2182B", "Intermediate Metabolizer" = "#EF8A62",
         "Normal Metabolizer" = "#D9D9D9", "Rapid Metabolizer" = "#67A9CF",
         "Ultrarapid Metabolizer" = "#2166AC")
# Facet by the MATRIX, not by the deposit. The two accessions are one study, and
# lumping 082493's urine, faeces and skin into a single panel hides that they behave
# differently: the ratio tracks genotype in plasma and not at all in urine.
harm_site <- function(x) {
  y <- tolower(trimws(x))
  dplyr::case_when(grepl("plasma", y) ~ "plasma",
                   grepl("^ua|urine", y) ~ "urine",
                   grepl("fecal|feces|faec", y) ~ "faeces",
                   grepl("forehead|skin", y) ~ "skin",
                   TRUE ~ y)
}
plot_df <- both |>
  filter(deposit_id %in% cyp_ds, !is.na(pheno_rank)) |>
  mutate(site = harm_site(specimen),
         phenotype = factor(cyp2c19_phenotype, levels = PHENO))
# label each panel with the number of subjects behind it, so a panel resting on two
# people is not read as equivalent to one resting on fourteen
.nsub <- plot_df |> dplyr::group_by(site) |>
  dplyr::summarise(ns = dplyr::n_distinct(subject_id), .groups = "drop")
plot_df$cohort <- factor(
  paste0(plot_df$site, "  (", .nsub$ns[match(plot_df$site, .nsub$site)], " subjects)"),
  levels = paste0(.nsub$site, "  (", .nsub$ns, " subjects)")[order(-.nsub$ns)])
write.csv(plot_df, "artifacts/cyp_validation_data.csv", row.names = FALSE)
```

```{r fig-by-phenotype, fig.width=9, fig.height=4}
ggplot(plot_df, aes(phenotype, log2_ratio, fill = phenotype)) +
  geom_boxplot(outlier.shape = NA, width = 0.62, alpha = 0.95) +
  geom_jitter(width = 0.15, height = 0, size = 0.8, alpha = 0.45) +
  facet_wrap(~cohort, scales = "free_y") +
  scale_fill_manual(values = pal, guide = "none") +
  scale_x_discrete(labels = PH_SHORT) +
  labs(x = "Genotyped CYP2C19 metaboliser class  (low → high activity)",
       y = "log2( omeprazole / 5-OH-omeprazole )",
       title = "log2(omeprazole / 5-OH) by genotyped CYP2C19 class, per matrix") +
  theme_bw(base_size = 11) +
  theme(strip.background = element_rect(fill = "grey92"),
        plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 20, hjust = 1))
```

One point per volunteer, not per file. A per-file boxplot pools every draw of a
longitudinal study into one box, mixing differences between people with the decay of the
ratio across the dosing interval, which is about 1.4 log2 within a person. It also gives a
one-volunteer class some thirty-six points and makes it look as solid as the
eight-volunteer class. This plots exactly what the model is fitted on.

```{r fig-ratio-by-volunteer, fig.width=6.4, fig.height=4.4}
# Written straight to thesis/graphics: the chapter figure and the notebook figure must be
# the same object. The earlier PNGs came from scratch scripts that were later deleted, so
# nine of thirteen chapter figures had no reproducible source and silently went stale.
.p_vol <- if (exists("auc") && nrow(auc) > 5) {
  .v <- aggregate(log2_ratio ~ subject + phenotype + rank, data = auc, FUN = mean)
  .v$phenotype <- factor(.v$phenotype, levels = PHENO)
  .fit <- lm(log2_ratio ~ rank, data = .v)
  .line <- data.frame(rank = seq(1, 5, length.out = 50))
  .line$log2_ratio <- predict(.fit, .line)
  .nv <- table(.v$phenotype)

  ggplot(.v, aes(rank, log2_ratio)) +
    geom_line(data = .line, colour = "grey35", linewidth = 0.8) +
    geom_point(aes(fill = phenotype), shape = 21, size = 3.4, stroke = 0.6,
               colour = "grey20") +
    scale_fill_manual(values = pal, guide = "none") +
    scale_x_continuous(breaks = 1:5,
                       labels = sprintf("%s\n(%d)", PH_SHORT, as.integer(.nv))) +
    labs(x = "Genotyped CYP2C19 metaboliser class  (low → high activity), volunteers in brackets",
         y = "log2( AUC omeprazole / AUC 5-OH-omeprazole )",
         title = "One point per volunteer: the plasma ratio against genotype",
         subtitle = sprintf("Least-squares fit: %.3f log2 per class (p = %.4f, R2 = %.2f, n = %d)",
                            coef(.fit)[2], summary(.fit)$coefficients[2, 4],
                            summary(.fit)$r.squared, nrow(.v))) +
    theme_bw(base_size = 11) +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold"))
} else NULL
if (!is.null(.p_vol)) {
  dir.create("thesis/graphics", showWarnings = FALSE, recursive = TRUE)
  ggsave("thesis/graphics/flagship_ratio_volunteer.png", .p_vol,
         width = 6.4, height = 4.4, dpi = 200)
}
.p_vol
```

```{r fig-ratio-and-detection, fig.width=9.4, fig.height=5.4}
# Plasma and urine, side by side, with the ratio next to the detection rate that
# produced it. Without the right-hand column a null in urine is ambiguous: it could
# be a failure to detect. It is not - urine detects BOTH compounds more often than
# plasma, and the metabolite more often than the parent, which is renal clearance
# rather than a sensitivity problem.
KEEP <- c("plasma", "urine")
rat <- plot_df |>
  dplyr::filter(site %in% KEEP) |>
  dplyr::transmute(site, phenotype, value = log2_ratio,
                   panel = "the ratio: log2( omeprazole / 5-OH-omeprazole )")

# The denominator must be every file of that matrix, not only the ones where the picker
# returned a peak. metrics_* holds a row only when pick() found something, so building
# this from `j` alone silently drops the empty files and inflates every rate (plasma read
# 78-94% against a true 53%). Start from the sample metadata and treat a file with no
# metrics row as "neither".
det <- rs |>
  dplyr::filter(deposit_id %in% cyp_ds) |>
  dplyr::mutate(site = harm_site(specimen),
                phenotype = factor(cyp2c19_phenotype, levels = PHENO)) |>
  dplyr::filter(site %in% KEEP, !is.na(phenotype)) |>
  dplyr::left_join(dplyr::distinct(j, stem, detection_tier), by = "stem") |>
  dplyr::mutate(detection_tier = ifelse(is.na(detection_tier), "neither", detection_tier)) |>
  dplyr::group_by(site, phenotype) |>
  dplyr::summarise(files = dplyr::n(),
                   parent     = mean(detection_tier %in% c("both", "parent_only")),
                   metabolite = mean(detection_tier %in% c("both", "metab_only")),
                   .groups = "drop") |>
  tidyr::pivot_longer(c(parent, metabolite), names_to = "compound", values_to = "value") |>
  dplyr::mutate(panel = "the detection behind it: % of files with the compound",
                value = 100 * value,
                compound = factor(compound, levels = c("parent", "metabolite"),
                                  labels = c("omeprazole (parent)",
                                             "5-OH-omeprazole (metabolite)")))

# facet_grid frees scales by ROW only, which would force the percentages onto the
# ratio's axis. One combined facet variable with facet_wrap gives each panel its own.
lv <- c(paste("plasma —", unique(rat$panel)), paste("plasma —", unique(det$panel)),
        paste("urine —",  unique(rat$panel)), paste("urine —",  unique(det$panel)))
rat$pan <- factor(paste(rat$site, "—", rat$panel), levels = lv)
det$pan <- factor(paste(det$site, "—", det$panel), levels = lv)

# Each volunteer contributes ~36 plasma files (9 draws x 2 dosing occasions x duplicate
# aliquots), so a class holding ONE person still shows dozens of files and its bar looks
# as solid as the eight-person Normal class. Put the volunteer count on the axis: it is
# shared by both columns, so it labels the boxplots and the bars at once.
.nvol <- rs |>
  dplyr::filter(deposit_id %in% cyp_ds, !is.na(cyp2c19_phenotype)) |>
  dplyr::distinct(subject_id, cyp2c19_phenotype) |>
  dplyr::count(cyp2c19_phenotype)
.lab_n <- setNames(sprintf("%s\n(%d)", PH_SHORT,
                           .nvol$n[match(PHENO, .nvol$cyp2c19_phenotype)]), PHENO)

ggplot() +
  geom_boxplot(data = rat, aes(phenotype, value, fill = phenotype),
               outlier.shape = NA, width = 0.62, alpha = 0.95) +
  geom_jitter(data = rat, aes(phenotype, value),
              width = 0.15, height = 0, size = 0.7, alpha = 0.4) +
  geom_col(data = det, aes(phenotype, value, fill = phenotype, alpha = compound),
           position = position_dodge(width = 0.72), width = 0.68, colour = "grey30",
           linewidth = 0.25) +
  # "%" on every label: a bare number above a bar reads as a count, and with ~36 files
  # per volunteer a "50" is easily taken for 50 detections rather than 50 per cent.
  geom_text(data = det, aes(phenotype, value, group = compound,
                            label = sprintf("%.0f%%", value)),
            position = position_dodge(width = 0.72), vjust = -0.35, size = 2.5) +
  facet_wrap(~ pan, scales = "free_y", ncol = 2) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.13))) +  # room for the 100% labels
  scale_fill_manual(values = pal, guide = "none") +
  # Colour already carries the genotype class on both sides, so the compound is carried
  # by shading. Overriding the swatch fill is what makes that legend readable at all.
  scale_alpha_manual(values = c(1, 0.42), name = "Bars (right-hand panels):",
                     guide = guide_legend(
                       override.aes = list(fill = "grey35", colour = "grey30"))) +
  scale_x_discrete(labels = .lab_n) +
  labs(x = "Genotyped CYP2C19 metaboliser class  (low → high activity), volunteers in brackets",
       y = NULL,
       title = "Left: the ratio.  Right: the detection rate behind it.",
       subtitle = paste("Plasma on top, urine below. Each volunteer gives ~36 files,",
                        "so a one-person class still fills a bar")) +
  theme_bw(base_size = 11) +
  theme(strip.background = element_rect(fill = "grey92"),
        strip.text = element_text(size = 9),
        plot.title = element_text(face = "bold"),
        legend.position = "top",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r fig-detect-vs-time, fig.width=9.4, fig.height=4.4}
# Detection rate against time since dose, for MS2 and for the MS1 picker. Every
# volunteer received omeprazole, so this is an absorption curve reconstructed from
# detections: it should be empty before the dose AND at 5 min (not yet absorbed),
# rise to a maximum at 4-5 h, and fall by 8 h. The picker should sit at or above MS2
# everywhere, and the gap should widen at the tails where the concentration is low
# and data-dependent acquisition stops triggering.
h_all <- read.csv("artifacts/hits_curated.csv", check.names = FALSE, stringsAsFactors = FALSE)
h_all$k <- norm_file_key(h_all$file_path)
.r  <- readRDS("artifacts/per_dataset/MSV000084008.rds")
.mp <- as.data.frame(.r$metrics_par); .mp$k <- norm_file_key(.mp$file)
.mm <- as.data.frame(.r$metrics_met); .mm$k <- norm_file_key(.mm$file)
.rs <- rs; .rs$k <- norm_file_key(.rs$file)
.s <- .rs[which(.rs$deposit_id == "MSV000084008" &
                grepl("plasma", tolower(.rs$specimen)) &
                is.finite(.rs$timepoint_mins)), ]
.hk <- function(role) h_all$k[h_all$ATTRIBUTE_DatasetAccession == "MSV000084008" &
                              h_all$role == role]
.d <- data.frame(
  tmin = .s$timepoint_mins,
  pheno = .s$cyp2c19_phenotype,
  `omeprazole|MS2`            = .s$k %in% .hk("parent"),
  `omeprazole|MS1 picker`     = .mp$par_detected[match(.s$k, .mp$k)] %in% TRUE,
  `5-OH-omeprazole|MS2`       = .s$k %in% .hk("metabolite"),
  `5-OH-omeprazole|MS1 picker`= .mm$met_detected[match(.s$k, .mm$k)] %in% TRUE,
  check.names = FALSE)
.d$pheno <- NULL   # the genotype split is plotted separately, below

.long <- .d |>
  tidyr::pivot_longer(-tmin, names_to = "series", values_to = "det") |>
  tidyr::separate(series, c("compound", "route"), sep = "\\|") |>
  dplyr::group_by(tmin, compound, route) |>
  dplyr::summarise(pct = 100 * mean(det), n = dplyr::n(), .groups = "drop") |>
  dplyr::mutate(compound = factor(compound, levels = c("omeprazole", "5-OH-omeprazole")))

# Evenly spaced categories, not real time: on a linear axis the 0 and 5 min draws
# overlap, and those two points carry the negative-control argument.
.lv <- sort(unique(.long$tmin))
.lab <- c("0" = "pre-dose", "5" = "5 min", "30" = "30 min", "60" = "1 h",
          "120" = "2 h", "240" = "4 h", "300" = "5 h", "360" = "6 h", "480" = "8 h")
.long$tf <- factor(.lab[as.character(.long$tmin)], levels = unname(.lab[as.character(.lv)]))

# (this is now the top row of the combined figure in the next chunk)
```

```{r fig-detect-by-genotype, fig.width=9.4, fig.height=7.0}
# The same absorption curve, split by genotype. Purely descriptive: three of the five
# classes hold a single volunteer, so the classes are pooled into low / normal / high
# activity and the volunteer count is written into each label. No test is run on this -
# the formal genotype result is the ratio, not presence.
.g <- data.frame(
  tmin  = .s$timepoint_mins,
  pheno = factor(.s$cyp2c19_phenotype, levels = PHENO),
  omeprazole        = .mp$par_detected[match(.s$k, .mp$k)] %in% TRUE,
  `5-OH-omeprazole` = .mm$met_detected[match(.s$k, .mm$k)] %in% TRUE,
  check.names = FALSE)
.g <- .g[!is.na(.g$pheno), ]
.g$grp <- cut(as.integer(.g$pheno), c(0, 2, 3, 5),
              labels = c("low activity", "normal activity", "high activity"))
.n <- tapply(.s$subject_id[!is.na(.s$cyp2c19_phenotype)],
             cut(match(.s$cyp2c19_phenotype[!is.na(.s$cyp2c19_phenotype)], PHENO),
                 c(0, 2, 3, 5), labels = c("low activity", "normal activity",
                                           "high activity")),
             function(x) length(unique(x)))
levels(.g$grp) <- sprintf("%s (%d volunteers)", levels(.g$grp), .n[levels(.g$grp)])

.gl <- .g |>
  tidyr::pivot_longer(c(omeprazole, `5-OH-omeprazole`),
                      names_to = "compound", values_to = "det") |>
  dplyr::group_by(tmin, grp, compound) |>
  dplyr::summarise(pct = 100 * mean(det), .groups = "drop") |>
  dplyr::mutate(compound = factor(compound,
                                  levels = c("omeprazole", "5-OH-omeprazole")))
.gl$tf <- factor(.lab[as.character(.gl$tmin)], levels = unname(.lab[as.character(.lv)]))

# One figure, four panels: compound across the columns, and down the rows the two
# questions asked of the same curve - what each route reaches, and whether genotype
# shows in presence. Sharing the axis makes the two directly comparable.
.top <- .long |> dplyr::transmute(tf, compound, series = route, pct,
                                  row = "by route")
.bot <- .gl   |> dplyr::transmute(tf, compound, series = as.character(grp), pct,
                                  row = "by genotype")
.cmb <- dplyr::bind_rows(.top, .bot)
.cmb$row <- factor(.cmb$row, levels = c("by route", "by genotype"))
.cmb$series <- factor(.cmb$series, levels = c("MS2", "MS1 picker", levels(.gl$grp)))

.p_det <- ggplot(.cmb, aes(tf, pct, colour = series, linetype = series, group = series)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.9) +
  facet_grid(row ~ compound) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_colour_manual(values = setNames(
    c("#B2182B", "#2166AC", "#B2182B", "grey35", "#2166AC"), levels(.cmb$series))) +
  scale_linetype_manual(values = setNames(
    c("solid", "solid", "22", "22", "22"), levels(.cmb$series))) +
  labs(x = NULL, y = "% of files with the compound",
       colour = NULL, linetype = NULL, tag = "A") +
  theme_bw(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "grey92")) +
  guides(colour = guide_legend(nrow = 1))

# Panel B: the ratio over the same axis. Presence and the quantity are two views of the
# same absorption curve, so they belong in one figure rather than two.
.rt <- tt |>
  dplyr::group_by(subject, day, tmin) |>
  dplyr::summarise(l2 = median(l2), .groups = "drop") |>
  dplyr::mutate(series = paste(subject, day))
.rt$tf <- factor(.lab[as.character(.rt$tmin)], levels = unname(.lab[as.character(.lv)]))
.rmed <- .rt |> dplyr::group_by(tf) |>
  dplyr::summarise(l2 = median(l2), n = dplyr::n_distinct(subject), .groups = "drop")

.p_rat <- ggplot(.rt, aes(tf, l2)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey70") +
  geom_line(aes(group = series), colour = "grey78", linewidth = 0.4) +
  geom_point(colour = "grey50", size = 0.9, alpha = 0.7) +
  geom_line(data = .rmed, aes(group = 1), colour = "#111111", linewidth = 1.1) +
  geom_point(data = .rmed, aes(fill = n < 5), shape = 21, size = 2.4,
             colour = "#111111", stroke = 0.8) +
  geom_text(data = .rmed, aes(label = n), vjust = -1.1, size = 2.7) +
  scale_fill_manual(values = c(`FALSE` = "#111111", `TRUE` = "white"), guide = "none") +
  labs(x = "Time since dose", y = "log2( omeprazole / 5-OH )", tag = "B") +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank())

.p_comb <- patchwork::wrap_plots(.p_det, .p_rat, ncol = 1, heights = c(2, 1.1))
dir.create("thesis/graphics", showWarnings = FALSE, recursive = TRUE)
ggsave("thesis/graphics/flagship_detect_combined.png", .p_comb,
       width = 9.4, height = 7.0, dpi = 200)
.p_comb
```

```{r tbl-propagation}
#| echo: false
# Propagation in a cohort where the answer is known: every volunteer received
# omeprazole, so a post-dose file is one where the compound is expected. MS2 reaches
# the files it happened to fragment; the picker reaches those plus the rest.
# Pre-dose and 5 min are shown too: the drug is not yet absorbed, so both routes
# should be empty there.
.hk2 <- function(ds, role) h_all$k[h_all$ATTRIBUTE_DatasetAccession == ds &
                                   h_all$role == role]
.grab <- function(ds, pat) {
  .rr <- readRDS(sprintf("artifacts/per_dataset/%s.rds", ds))
  .a <- as.data.frame(.rr$metrics_par); .a$k <- norm_file_key(.a$file)
  .b <- as.data.frame(.rr$metrics_met); .b$k <- norm_file_key(.b$file)
  .z <- .rs[which(.rs$deposit_id == ds & grepl(pat, tolower(.rs$specimen))), ]
  data.frame(tmin = .z$timepoint_mins,
             par_ms2 = .z$k %in% .hk2(ds, "parent"),
             met_ms2 = .z$k %in% .hk2(ds, "metabolite"),
             par_pick = .a$par_detected[match(.z$k, .a$k)] %in% TRUE,
             met_pick = .b$met_detected[match(.z$k, .b$k)] %in% TRUE)
}
.plas <- .grab("MSV000084008", "plasma")
.urin <- .grab("MSV000082493", "urine")
.skin <- .grab("MSV000082493", "skin|forehead")

.row <- function(d, label) data.frame(
  Stratum = label, Files = nrow(d),
  `MS2 parent` = sum(d$par_ms2), `Picker parent` = sum(d$par_pick),
  `MS2 metabolite` = sum(d$met_ms2), `Picker metabolite` = sum(d$met_pick),
  check.names = FALSE)

.lab <- c("0" = "plasma, pre-dose", "5" = "plasma, 5 min", "30" = "plasma, 30 min",
          "60" = "plasma, 1 h", "120" = "plasma, 2 h", "240" = "plasma, 4 h",
          "300" = "plasma, 5 h", "360" = "plasma, 6 h", "480" = "plasma, 8 h")
.tab <- dplyr::bind_rows(
  lapply(sort(unique(.plas$tmin[is.finite(.plas$tmin)])),
         function(t) .row(.plas[.plas$tmin %in% t, ], .lab[[as.character(t)]])),
  .row(.plas[which(.plas$tmin > 0), ], "plasma, all post-dose"),
  .row(.urin[which(.urin$tmin > 0), ], "urine, all post-dose"),
  .row(.skin[which(.skin$tmin > 0), ], "skin, all post-dose"))
.tab$`Propagation multiplier` <- ifelse(
  .tab$`MS2 parent` + .tab$`MS2 metabolite` == 0, "-",
  sprintf("%.2f", (.tab$`Picker parent` + .tab$`Picker metabolite`) /
                  pmax(1, .tab$`MS2 parent` + .tab$`MS2 metabolite`)))
# The thesis table is built from this, so it is written on every render. It used to be
# a file nothing produced any more, which let the chapter drift six days behind the data.
write.csv(.tab, "artifacts/propagation_dosed_cohort.csv", row.names = FALSE)
knitr::kable(.tab, caption = paste(
  "Files with each compound detected by MS2 and by the MS1 picker, in a cohort where",
  "every volunteer received omeprazole. Pre-dose and 5-minute draws precede absorption",
  "and are empty by both routes. The multiplier is picker detections over MS2",
  "detections, summed across the two compounds; it is undefined where MS2 finds",
  "nothing, which is the case for the parent throughout urine."))
```


Each subject is dosed twice, on study days 1 and 9, with a full time course each time.
Pooling the two collapses two independent dosing occasions into one cloud and makes every
line appear to jump backwards in time, so they are shown side by side; the second panel is
then an internal replicate of the first.

```{r fig-ratio-vs-time, fig.width=9.2, fig.height=4.6}
if (exists("tt") && nrow(tt) > 10) {
  tp <- tt |>
    dplyr::group_by(subject, day, tmin) |>          # one point per draw, not per injection
    dplyr::summarise(l2 = median(l2), .groups = "drop") |>
    dplyr::mutate(series = paste(subject, day),
                  panel  = paste0("Study day ", day,
                                  ifelse(day == min(day), "  (first dose)", "  (second dose)")))
  # A ratio needs BOTH compounds detected, and early in the interval the metabolite
  # often is not there yet, so the panel is badly unbalanced: 30 min rests on 2 subjects
  # against 13 at 4 h. Without the counts a reader takes the left end of the median line
  # for the cohort. Points on <5 subjects are drawn hollow and the line dashed there.
  med <- tp |> dplyr::group_by(panel, tmin) |>
    dplyr::summarise(l2 = median(l2), n = dplyr::n_distinct(subject), .groups = "drop") |>
    dplyr::mutate(sparse = n < 5)

  ggplot(tp, aes(tmin, l2)) +
    geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey70") +
    geom_line(aes(group = series), colour = "grey72", linewidth = 0.45) +
    geom_point(colour = "grey45", size = 1, alpha = 0.8) +
    geom_line(data = med, colour = "#111111", linewidth = 1.1) +
    geom_point(data = med, aes(fill = sparse), colour = "#111111", size = 2.4,
               shape = 21, stroke = 0.9) +
    geom_text(data = med, aes(label = n), vjust = -1.15, size = 2.9, colour = "#111111") +
    scale_fill_manual(values = c(`FALSE` = "#111111", `TRUE` = "white"), guide = "none") +
    facet_wrap(~ panel) +
    scale_x_continuous(trans = "log10",
                       breaks = c(5, 30, 60, 120, 240, 360, 480),
                       labels = c("5 m", "30 m", "1 h", "2 h", "4 h", "6 h", "8 h")) +
    labs(x = "Time since dose (log scale). Pre-dose draws at t = 0 carry no drug, so no ratio exists there",
         y = "log2( omeprazole / 5-OH-omeprazole )",
         title = "log2(omeprazole / 5-OH) across the dosing interval, both doses, MSV000084008",
         subtitle = paste("Numbers are subjects contributing a ratio at that draw;",
                          "hollow points rest on fewer than 5")) +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "grey93", colour = NA),
          strip.text = element_text(face = "bold", size = 10.5),
          plot.title = element_text(face = "bold", size = 12))
}
```

<!-- The anchor fan-out plot was removed. Confirmed anchors are a CAPPED subset (25
per role per biofluid on MassIVE, 50 elsewhere), so plotting anchors beside files
reached makes the spread look like a gain factor when it is an artefact of the cap.
The honest per-compound and per-ratio gains are tabulated in `05` Part A against the
full MASST baseline. -->
