---
title: "04: CYP2C19 ground-truth validation"
subtitle: "Does the mined omeprazole / 5-OH ratio track genotyped CYP2C19 status?"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
---

# Overview

MSV000082493 and MSV000084008 (Jarmusch et al., *Clin Transl Sci* 2020,
PMC7485946) dosed healthy volunteers with omeprazole as a CYP2C19 probe and
carry the **genotyped** CYP2C19 metaboliser phenotype per subject in
`artifacts/rescue_samples.csv`. The authors reported the identical OME /
5-OH-OME ratio but never tested it against that genotype, so this is a new use
of the data, and a ground-truth check that the whole cascade (MASST → SIRIUS →
MS1 EIC → ratio) measures real pharmacology.

Expected: low CYP2C19 activity (Poor) → less 5-OH → **high** log2(parent/metab);
Ultrarapid → low. Run after `03` has cached the two datasets.

Joins between the metrics and the submitter table go through `norm_file_key()`,
never a bare basename: the metrics carry cache-prefixed names while the TSV
carries bare ones, and that mismatch has silently produced empty joins several
times in this project.

```{r setup}
suppressPackageStartupMessages({ library(dplyr); library(tidyr); library(ggplot2) })
source("R/metadata.R")
source("R/masst.R")   # compound_spec + PHENO, the domain constants
source("R/ms1.R")     # resolve_isobars(), so this notebook gates as the corpus does

# shared by every figure below: class order and class colours
PH_SHORT <- c("Poor", "Inter.", "Normal", "Rapid", "Ultra")
pal <- c("Poor Metabolizer" = "#B2182B", "Intermediate Metabolizer" = "#EF8A62",
         "Normal Metabolizer" = "#D9D9D9", "Rapid Metabolizer" = "#67A9CF",
         "Ultrarapid Metabolizer" = "#2166AC")
```

# Per-file table + genotype join

`sample_table_all.csv` is deduped to one file per subject, which collapses the
timepoint × specimen structure this readout needs. We rebuild a per-file table
from the cached per-dataset metrics (before dedup), replicating
`build_sample_table()`'s detection rule (shape pass AND robust-SNR ≥ 3), and
join the genotype by filename stem (metrics carry `.mzXML`/prefixed cache names,
the genotype table `.mzXML`, identical stems).

```{r dup-overlap}
#| echo: false
# Counted rather than written into the paragraph below: an earlier hand-written
# pair (222 and 513) survived a rebuild that changed both.
.d <- read.csv("artifacts/sample_table_perfile.csv", check.names = FALSE)
.k93 <- norm_file_key(.d$file[.d$dataset_id == "MSV000082493"])
.k08 <- norm_file_key(.d$file[.d$dataset_id == "MSV000084008"])
N_SHARED <- length(intersect(.k93, .k08)); N_84008 <- length(.k08)
```

The two accessions are one study: MSV000082493 re-deposits `r N_SHARED` of
MSV000084008's `r N_84008` plasma files, so a per-file table built from both
carries each of those twice and the join propagates the duplicate. Every rate
would then be computed on a set biased toward the re-deposited files. Plasma is
therefore attributed to MSV000084008 alone; those files carry no metadata under
the other accession in any case. The count the join below reports as dropped is
smaller than `r N_SHARED`, because it counts only the rows that survived the
genotype join rather than every shared file.

```{r per-file-join}
cyp_ds <- c("MSV000082493", "MSV000084008")
rs <- read.csv("artifacts/rescue_samples.csv", check.names = FALSE, stringsAsFactors = FALSE)
cf_all <- read.csv("artifacts/hits-confirmed-all.csv", stringsAsFactors = FALSE)

# CYP2C19 activity order (low -> high); the ratio should DECREASE along it
# PHENO (CYP2C19 activity order) comes from R/masst.R.
rs$pheno_rank <- match(rs$cyp2c19_phenotype, PHENO)

bare <- function(x)  # strip MSV000XXXXXX_ cache prefix + extension, lowercase
  sub("^msv[0-9]{9}_", "", tolower(tools::file_path_sans_ext(basename(x))))
rs$stem <- bare(rs$file)

per_file_tier <- function(d) {
  r <- readRDS(sprintf("artifacts/per_dataset/%s.rds", d))
  # Same isobar rejection the corpus tables use. Without it this notebook counts
  # peaks at the other compound's retention time as detections: 86 of the 90
  # urine "omeprazole" peaks are the isobaric hydroxyomeprazole sulfide.
  .g <- resolve_isobars(r$metrics_par, r$metrics_met, cf_all, d, r$rt_anchor)
  m <- merge(.g$par, .g$met, by = "file", all = TRUE)
  # Detection is decided in the picker (accept_peak + drop-list); metrics_*
  # carry it as a flag.
  par_det <- m$par_detected %in% TRUE
  met_det <- m$met_detected %in% TRUE
  m$detection_tier <- ifelse( par_det &  met_det, "both",
                       ifelse( par_det & !met_det, "parent_only",
                       ifelse(!par_det &  met_det, "metab_only", "neither")))
  m$log2_ratio <- ifelse(m$detection_tier == "both",
                         log2(m$par_auc / m$met_auc), NA_real_)
  m$dataset_id <- d
  m$stem <- bare(m$file)
  m
}
pf <- bind_rows(lapply(cyp_ds, per_file_tier))

j <- pf |>
  inner_join(rs[, c("stem", "deposit_id", "cyp2c19_phenotype", "cyp2c19_genotype",
                    "pheno_rank", "subject_id", "timepoint", "timepoint_mins",
                    "specimen", "pharmkinetics_cyp2c19")],
             by = "stem")

.dup <- j$dataset_id == "MSV000082493" & grepl("plasma", tolower(j$specimen))
if (any(.dup)) {
  cat("dropped", sum(.dup), "re-deposited plasma rows from MSV000082493\n")
  j <- j[!.dup, ]
}
stopifnot(!any(duplicated(j$stem)))
both <- j |> filter(detection_tier == "both", is.finite(log2_ratio),
                    !par_fwhm_outlier %in% TRUE, !met_fwhm_outlier %in% TRUE)
cat("joined files:", nrow(j), " | with both compounds + clean ratio:", nrow(both), "\n")
table(j$dataset_id, j$detection_tier)
```

# Ratio vs genotyped phenotype

Per cohort: median ratio by phenotype, file-level Spearman/Kruskal-Wallis, and
the valid **subject-level** Spearman (one ratio per subject, no
pseudoreplication).

```{r phenotype-tests}
test_cohort <- function(d, label) {
  x <- both |> filter(if (is.null(d)) TRUE else deposit_id == d, !is.na(pheno_rank))
  if (nrow(x) < 8 || dplyr::n_distinct(x$cyp2c19_phenotype) < 3) {
    cat("\n[", label, "] too few points (n=", nrow(x), ")\n"); return(invisible())
  }
  cat("\n----------", label, " (n =", nrow(x), "files) ----------\n")
  ph <- x |> group_by(cyp2c19_phenotype) |>
    summarise(n = n(), med = round(median(log2_ratio), 2), .groups = "drop") |>
    arrange(match(cyp2c19_phenotype, PHENO))
  print(as.data.frame(ph))
  # SUBJECT level only: a file is not an independent observation, so a per-file
  # Spearman here would measure how often each volunteer was injected rather
  # than their genotype. The per-file medians printed above are description, not
  # a test.
  subj <- x |> group_by(subject_id, cyp2c19_phenotype, pheno_rank) |>
    summarise(ratio = median(log2_ratio), .groups = "drop")
  if (nrow(subj) >= 8 && dplyr::n_distinct(subj$cyp2c19_phenotype) >= 3) {
    sp2 <- cor.test(subj$ratio, subj$pheno_rank, method = "spearman", exact = FALSE)
    kw2 <- kruskal.test(ratio ~ factor(cyp2c19_phenotype), subj)
    cat(sprintf("subject level (n = %d): Spearman rho = %.3f  p = %.3g  (expect NEGATIVE)\n",
                nrow(subj), sp2$estimate, sp2$p.value))
    cat(sprintf("Kruskal-Wallis across phenotypes p = %.3g\n", kw2$p.value))
  } else cat("too few subjects for a subject-level test\n")
  invisible(x)
}
x84 <- test_cohort("MSV000084008", "084008 PLASMA (classic CYP readout)")
x82 <- test_cohort("MSV000082493", "082493 urine/feces/skin")
invisible(test_cohort(NULL, "POOLED CYP cohorts"))
```

# Per-subject AUC metabolic ratio (the clinical readout)

Everything above uses **one ratio per subject** taken from single files. That is
not how CYP2C19 phenotyping is done clinically: the accepted metric is the
**ratio of areas under the concentration-time curves** over the dosing interval,
AUC(omeprazole) / AUC(5-OH-omeprazole). Integrating removes the dependence on
*when* the sample happened to be drawn, which in a single-timepoint ratio is
pure noise: one subject sampled near their peak and another past it differ for
reasons that have nothing to do with genotype.

MSV000084008 supports this directly: 461 plasma files, 14 subjects, 8-9
timepoints per subject on each of two study days. The timepoint comes from the
submitter's `ATTRIBUTE_Time_Point_Mins` (recovered by the metadata union in
`01`; Pan-ReDU does not carry it).

Two conventions worth stating because they change the answer:

- **Days are integrated separately.** Day 1 and day 9 straddle the antibiotic
  exposure, so they are different physiological states and pooling them would
  blur the contrast the study was designed around.
- **A non-detected timepoint contributes 0** to the AUC, the standard
  pharmacokinetic below-limit-of-detection convention. "Non-detected" means the
  timepoint fails the picker's detection gate (`detection_tier`), not merely
  that no peak was integrated: a timepoint where only noise was picked is set to
  0, not allowed to contribute its raw integrated area. Dropping those points
  instead would bias the AUC upward exactly for the poor metabolisers whose
  metabolite is genuinely absent.

The direction is predicted before looking: poor metabolisers clear omeprazole
slowly, so a lower activity rank should give a higher parent/metabolite ratio,
i.e. Spearman rho < 0.

```{r auc-ratio}
auc_tsv <- "dataset_metadata/MSV000084008_metadata.tsv"
if (!file.exists(auc_tsv)) {
  cat("submitter TSV not present - AUC section skipped\n")
} else {
  r_auc <- readRDS("artifacts/per_dataset/MSV000084008.rds")
  # same isobar rejection the corpus tables use, so this section cannot integrate
  # a peak at the other compound's retention time
  .cf0 <- read.csv("artifacts/hits-confirmed-all.csv", stringsAsFactors = FALSE)
  .mr  <- resolve_isobars(r_auc$metrics_par, r_auc$metrics_met, .cf0,
                          "MSV000084008", r_auc$rt_anchor)
  mp <- as.data.frame(.mr$par); mm <- as.data.frame(.mr$met)
  ta <- read.delim(auc_tsv, check.names = FALSE, stringsAsFactors = FALSE)
  # normalised key, never a bare basename: the metrics carry cache-prefixed
  # names
  ta$k <- norm_file_key(ta$filename)
  mp$k <- norm_file_key(mp$file); mm$k <- norm_file_key(mm$file)
  # Gate to detection: a timepoint that fails the picker's detection gate is
  # below-LOD (NA here -> 0 in trap()), not its raw integrated noise-area.
  # Detected points keep their AUC.
  mp_i <- match(ta$k, mp$k); mm_i <- match(ta$k, mm$k)
  ta$par_auc <- ifelse(mp$par_detected[mp_i] %in% TRUE, mp$par_auc[mp_i], NA_real_)
  ta$met_auc <- ifelse(mm$met_detected[mm_i] %in% TRUE, mm$met_auc[mm_i], NA_real_)
  # apex retention times, kept for the isomer restrictions below
  ta$par_rt <- mp$par_apex_rt[mp_i]; ta$met_rt <- mm$met_apex_rt[mm_i]
  ta$tmin <- as.numeric(ta$ATTRIBUTE_Time_Point_Mins)
  ta <- ta[ta$MS_SampleType == "sample" & !is.na(ta$tmin), ]

  trap <- function(x, y) {              # trapezoidal AUC; NA -> 0 (below LOD)
    o <- order(x); x <- x[o]; y <- y[o]; y[is.na(y)] <- 0
    if (length(unique(x)) < 4) return(NA_real_)
    sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)
  }
  auc <- do.call(rbind, lapply(
    split(ta, list(ta$ATTRIBUTE_Subject, ta$ATTRIBUTE_Study_Day), drop = TRUE),
    function(d) data.frame(
      subject = d$ATTRIBUTE_Subject[1], day = d$ATTRIBUTE_Study_Day[1],
      n_tp = length(unique(d$tmin)),
      auc_par = trap(d$tmin, d$par_auc), auc_met = trap(d$tmin, d$met_auc),
      phenotype = d$PharmGeno_Phenotype_2C19[1], stringsAsFactors = FALSE)))
  auc$log2_ratio <- log2(auc$auc_par / auc$auc_met)
  auc <- auc[is.finite(auc$log2_ratio), ]
  auc$rank <- match(auc$phenotype, PHENO)

  cat("timepoints per subject-day: median", median(auc$n_tp),
      "range", min(auc$n_tp), "-", max(auc$n_tp), "\n\n")
  print(auc[order(auc$rank, auc$subject),
            c("subject","day","n_tp","log2_ratio","phenotype")],
        row.names = FALSE, digits = 3)

  cat("\nmean log2(AUC parent / AUC metabolite) by phenotype:\n")
  subj <- aggregate(log2_ratio ~ subject + phenotype + rank, data = auc, FUN = mean)
  print(tapply(subj$log2_ratio, factor(subj$phenotype, levels = PHENO), mean))

  cat("\n")
  for (dd in sort(unique(auc$day))) {
    s <- auc[auc$day == dd, ]
    ct <- cor.test(s$rank, s$log2_ratio, method = "spearman", exact = FALSE)
    cat(sprintf("day %-3s n=%2d | Spearman rho = %+.3f  p = %.4f\n",
                dd, nrow(s), ct$estimate, ct$p.value))
  }
  ct <- cor.test(subj$rank, subj$log2_ratio, method = "spearman", exact = FALSE)
  cat(sprintf("both days, per subject  n=%2d | Spearman rho = %+.3f  p = %.4f\n",
              nrow(subj), ct$estimate, ct$p.value))
}
```

Every volunteer as one line, on the axis the clinical protocol fixes:

```{r fig-ratio-by-volunteer, fig.width=6.4, fig.height=4.4}
#| echo: false
# Written straight to thesis/graphics: the chapter figure and the notebook
# figure must be the same object. The earlier PNGs came from scratch scripts
# that were later deleted, so nine of thirteen chapter figures had no
# reproducible source and silently went stale.
.p_vol <- if (exists("auc") && nrow(auc) > 5) {
  .v <- aggregate(log2_ratio ~ subject + phenotype + rank, data = auc, FUN = mean)
  .v$phenotype <- factor(.v$phenotype, levels = PHENO)
  .fit <- lm(log2_ratio ~ rank, data = .v)
  .line <- data.frame(rank = seq(1, 5, length.out = 50))
  .line$log2_ratio <- predict(.fit, .line)
  .nv <- table(.v$phenotype)

  ggplot(.v, aes(rank, log2_ratio)) +
    geom_line(data = .line, colour = "grey35", linewidth = 0.8) +
    geom_point(aes(fill = phenotype), shape = 21, size = 3.4, stroke = 0.6,
               colour = "grey20") +
    scale_fill_manual(values = pal, guide = "none") +
    scale_x_continuous(breaks = 1:5,
                       labels = sprintf("%s\n(%d)", PH_SHORT, as.integer(.nv))) +
    labs(x = "Genotyped CYP2C19 metaboliser class  (low → high activity), volunteers in brackets",
         y = "log2( AUC omeprazole / AUC 5-OH-omeprazole )",
         title = "One point per volunteer: the plasma ratio against genotype",
         subtitle = sprintf("Least-squares fit: %.3f log2 per class (p = %.4f, R2 = %.2f, n = %d)",
                            coef(.fit)[2], summary(.fit)$coefficients[2, 4],
                            summary(.fit)$r.squared, nrow(.v))) +
    theme_bw(base_size = 11) +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold"))
} else NULL
if (!is.null(.p_vol)) {
  dir.create("thesis/graphics", showWarnings = FALSE, recursive = TRUE)
  ggsave("thesis/graphics/flagship_ratio_volunteer.png", .p_vol,
         width = 6.4, height = 4.4, dpi = 200)
}
.p_vol
```

## Does the ranking depend on the classes being equally spaced?

The model below treats the activity rank as a number, which assumes the five
classes are evenly spaced --- something nothing guarantees. Spearman assumes
only that the order is right, so it is the assumption-free version of the same
claim, and the singleton classes can be thrown away entirely to check that the
two populated ones still separate. Whether one subject carries the result is
asked once, of the model the chapter reports, in the leave-one-out below.

```{r auc-robustness}
if (exists("subj") && nrow(subj) > 5) {
  full <- cor.test(subj$rank, subj$log2_ratio, method = "spearman", exact = FALSE)
  cat(sprintf("rank-only (Spearman): n=%2d rho=%+.3f p=%.4f\n",
              nrow(subj), full$estimate, full$p.value))
  # Ties make an exact Spearman impossible, so the p above is asymptotic. A
  # permutation test assumes nothing; if the two agree, the fallback is sound.
  set.seed(1)
  .pm <- replicate(20000, cor(sample(subj$rank), subj$log2_ratio,
                              method = "spearman"))
  cat(sprintf("permutation check (20,000 relabellings): p = %.4f
",
              (1 + sum(abs(.pm) >= abs(full$estimate) - 1e-12)) / 20001))
  two <- subj[subj$phenotype %in% c("Intermediate Metabolizer",
                                    "Normal Metabolizer"), ]
  if (dplyr::n_distinct(two$phenotype) == 2) {
    w <- wilcox.test(log2_ratio ~ phenotype, data = two)
    cat(sprintf("singletons removed - Intermediate (n=%d) vs Normal (n=%d): p = %.3f\n",
                sum(two$phenotype == "Intermediate Metabolizer"),
                sum(two$phenotype == "Normal Metabolizer"), w$p.value))
  }
}
```

Ties are the design here --- eight of the fourteen are normal metabolisers ---
so an exact Spearman cannot be computed and `exact = FALSE` takes the asymptotic
p-value deliberately. The permutation line above is the check on that fallback:
it assumes nothing, and its agreeing with the asymptotic p is what licenses
using it.

## Is an isomer being integrated instead?

Fragmentation does not separate 5-OH-omeprazole from omeprazole sulfone, and an
isomer of omeprazole itself is reported in these samples. Chromatography does
separate them: each mass carries two RT clusters (parent ~157 and ~167 s,
metabolite ~157 and ~182 s) and the confirmed anchors place omeprazole in the
*later* parent cluster and 5-OH-omeprazole in the *earlier* metabolite one. If
an isomer were being integrated, restricting to the correct cluster should
change the answer.

A restriction can only remove a **detection**, never a file: an apex in the
wrong cluster makes that channel unusable at that timepoint, which is the
below-LOD case the AUC already handles by contributing 0.

```{r isomer-restrictions}
if (!exists("ta") || !"par_rt" %in% names(ta)) {
  cat("AUC section did not run - isomer restrictions skipped\n")
} else {
  PC <- 162; MC <- 163                      # cluster split, parent / metabolite
  a <- ta[!is.na(ta$par_auc) & !is.na(ta$met_auc), ]
  pmed <- median(a$par_rt, na.rm = TRUE); mmed <- median(a$met_rt, na.rm = TRUE)

  fit_restricted <- function(keep_par, keep_met, label) {
    d <- ta
    d$par_auc[!keep_par] <- NA_real_
    d$met_auc[!keep_met] <- NA_real_
    ac <- do.call(rbind, lapply(
      split(d, list(d$ATTRIBUTE_Subject, d$ATTRIBUTE_Study_Day), drop = TRUE),
      function(z) data.frame(
        subject = z$ATTRIBUTE_Subject[1],
        l2 = log2(trap(z$tmin, z$par_auc) / trap(z$tmin, z$met_auc)),
        phenotype = z$PharmGeno_Phenotype_2C19[1], stringsAsFactors = FALSE)))
    ac <- ac[is.finite(ac$l2), ]
    s <- aggregate(l2 ~ subject + phenotype, data = ac, FUN = mean)
    s$rank <- match(s$phenotype, PHENO)
    ct <- cor.test(s$rank, s$l2, method = "spearman", exact = FALSE)
    m <- summary(lm(l2 ~ rank, data = s))
    cat(sprintf("  %-40s files %3d subj %2d | rho %+.3f p %.4f | slope %+.3f\n",
                label, sum(!is.na(d$par_auc) & !is.na(d$met_auc)), nrow(s),
                ct$estimate, ct$p.value, m$coefficients[2, 1]))
  }
  ok <- function(x, cond) is.na(x) | cond   # undetected apexes are already NA

  fit_restricted(TRUE, TRUE, "all accepted")
  fit_restricted(ok(ta$par_rt, ta$par_rt >= PC), TRUE, "parent from the late cluster only")
  fit_restricted(TRUE, ok(ta$met_rt, ta$met_rt <= MC), "metabolite inside its window only")
  fit_restricted(ok(ta$par_rt, ta$par_rt >= PC),
                 ok(ta$met_rt, ta$met_rt <= MC), "both channels unambiguous")
  fit_restricted(ok(ta$par_rt, abs(ta$par_rt - pmed) <= 2),
                 ok(ta$met_rt, abs(ta$met_rt - mmed) <= 2),
                 "both within 2 s of the cohort median")

  cat(sprintf("\nparent apex in the EARLY cluster: %d of %d\n",
              sum(a$par_rt < PC, na.rm = TRUE), nrow(a)))
  cat(sprintf("metabolite apex outside its window: %d\n", sum(a$met_rt > MC, na.rm = TRUE)))
  cat(sprintf("per-subject median apex spread: parent %.1f s, metabolite %.1f s\n",
              diff(range(tapply(a$par_rt, a$ATTRIBUTE_Subject, median), na.rm = TRUE)),
              diff(range(tapply(a$met_rt, a$ATTRIBUTE_Subject, median), na.rm = TRUE))))
}
```

# The linear model against genotype

Everything above uses Spearman, which asks only whether the ratio decreases as
activity increases. The chapter reports a linear model instead, because a rank
correlation carries no units and a claim about a measurement needs an effect
size in interpretable ones.

The classes are ranked poor = 1 to ultrarapid = 5 and the subject-level log2
ratio is regressed on that rank, one row per volunteer. The coefficient is a
change **per genotype class**, not per hour. Treating the rank as a number
assumes the classes are equally spaced, which nothing guarantees; that is the
price of an interpretable coefficient.

```{r lm-genotype}
subj_auc <- aggregate(log2_ratio ~ subject + phenotype + rank, data = auc, FUN = mean)
m_auc <- lm(log2_ratio ~ rank, data = subj_auc)
summary(m_auc)
.ci <- confint(m_auc)["rank", ]
cat(sprintf("\nslope %.3f log2 per class, 95%% CI %.3f to %.3f\n",
            coef(m_auc)[2], .ci[1], .ci[2]))
cat(sprintf("fold change per class: %.3f (%.3f to %.3f)\n",
            2^coef(m_auc)[2], 2^.ci[1], 2^.ci[2]))
# R-squared has no closed form here, so resample the 14 volunteers for an interval
set.seed(1)
.r2 <- replicate(2000, {
  i <- sample(nrow(subj_auc), replace = TRUE)
  if (length(unique(subj_auc$rank[i])) < 2) return(NA_real_)
  summary(lm(log2_ratio ~ rank, data = subj_auc[i, ]))$r.squared
})
cat(sprintf("R2 %.2f, bootstrap 95%% CI %.2f to %.2f, over %d volunteers\n",
            summary(m_auc)$r.squared,
            quantile(.r2, 0.025, na.rm = TRUE), quantile(.r2, 0.975, na.rm = TRUE),
            nrow(subj_auc)))
```

## Does the summary statistic matter?

Replacing the integral with a plain mean of each volunteer's file ratios should
give the same answer if the conclusion is not an artefact of how the repeated
measurements were collapsed.

```{r lm-mean-of-files}
mof <- both |>
  filter(deposit_id == "MSV000084008", !is.na(pheno_rank)) |>
  group_by(subject_id, pheno_rank) |>
  summarise(v = mean(log2_ratio), .groups = "drop")
summary(lm(v ~ pheno_rank, data = mof))$coefficients
```

## Does timing produce the result?

The ratio decays across the dosing interval, so a difference between volunteers
could in principle be a difference in when their blood was drawn. Refitting the
same model at one fixed draw at a time removes that: every volunteer is then
compared at the same point after dosing.

The submitter TSV carries duplicate column names, which dplyr refuses to
transform, so a minimal frame is built from it once and reused.

```{r lm-draws-frame}
draws <- data.frame(
  subject = ta$ATTRIBUTE_Subject,
  day     = ta$ATTRIBUTE_Study_Day,
  rank    = match(ta$PharmGeno_Phenotype_2C19, PHENO),
  tmin    = ta$tmin,
  l2      = log2(ta$par_auc / ta$met_auc),
  stringsAsFactors = FALSE)
draws <- draws[is.finite(draws$l2) & !is.na(draws$rank), ]
```

```{r lm-fixed-draw}
fixed_draw <- function(tp) {
  x <- draws |> filter(tmin == tp) |>
    group_by(subject, rank) |> summarise(v = mean(l2), .groups = "drop")
  co <- summary(lm(v ~ rank, data = x))$coefficients
  data.frame(draw_min = tp, n = nrow(x), beta = co[2, 1], se = co[2, 2], p = co[2, 4])
}
do.call(rbind, lapply(c(120, 240, 300, 360), fixed_draw))
```

## Each dosing occasion separately

Day 9 follows the seven-day antibiotic course, so it is worth showing that it
still reports the same phenotype rather than assuming it.

```{r lm-per-occasion}
per_day <- do.call(rbind, lapply(sort(unique(auc$day)), function(d) {
  x <- auc[auc$day == d & is.finite(auc$log2_ratio), ]
  co <- summary(lm(log2_ratio ~ rank, data = x))$coefficients
  data.frame(occasion = d, n = nrow(x), beta = co[2, 1], p = co[2, 4])
}))
both_days <- summary(lm(log2_ratio ~ rank, data = subj_auc))$coefficients
rbind(per_day, data.frame(occasion = "mean of both", n = nrow(subj_auc),
                          beta = both_days[2, 1], p = both_days[2, 4]))

w <- auc |> select(subject, day, log2_ratio) |> filter(is.finite(log2_ratio)) |>
  tidyr::pivot_wider(names_from = day, values_from = log2_ratio)
w <- w[complete.cases(w), ]
cat(sprintf("\nwithin-subject change across the antibiotic course: median %+.2f log2, n = %d\n",
            median(w[[3]] - w[[2]]), nrow(w)))
print(wilcox.test(w[[3]], w[[2]], paired = TRUE))
```

## Leave-one-volunteer-out

Three classes contain a single volunteer, so the model is refitted fourteen
times, each fit omitting one.

```{r lm-jackknife}
jk <- do.call(rbind, lapply(seq_len(nrow(subj_auc)), function(i) {
  co <- summary(lm(log2_ratio ~ rank, data = subj_auc[-i, ]))$coefficients
  data.frame(dropped = subj_auc$subject[i], phenotype = subj_auc$phenotype[i],
             beta = co[2, 1], p = co[2, 4])
}))
jk[order(jk$beta), ]
cat(sprintf("\nbeta range %+.3f to %+.3f | worst p %.4f\n",
            min(jk$beta), max(jk$beta), max(jk$p)))
```

## The elimination rate as a second read-out

How fast the ratio falls is itself genotype-dependent: a poor metaboliser clears
omeprazole slowly, so their ratio should decay more gently. This is a natural
second endpoint, and this cohort cannot support it.

```{r lm-elimination-rate}
slopes <- draws |>
  group_by(subject, day, rank) |> filter(n_distinct(tmin) >= 3) |>
  summarise(b = coef(lm(l2 ~ I(tmin / 60)))[2], .groups = "drop") |>
  group_by(subject, rank) |> summarise(b = mean(b), .groups = "drop")
summary(lm(b ~ rank, data = slopes))$coefficients
slopes |> group_by(rank) |> summarise(n = n(), median_slope = round(median(b), 3))
```

# Genotype dosage (star alleles)

CYP2C19 allele activity: *1 = 1 (normal), *2/*3 = 0 (no function), *17 = 1.5
(increased). The activity score respects functional spacing the equal-spaced
rank ignores; the loss-of-function dose isolates the strong *2/*3 effect from
the noisy *17. Subject level.

```{r genotype-dosage}
allele_act <- function(a) ifelse(a %in% c("2", "3"), 0,
                          ifelse(a == "17", 1.5, ifelse(a == "1", 1, NA_real_)))
lof_allele <- function(a) as.integer(a %in% c("2", "3"))
add_geno <- function(df) {
  parts <- strsplit(as.character(df$cyp2c19_genotype), "_")
  df$act_score <- vapply(parts, function(p) sum(allele_act(p)), numeric(1))
  df$lof_dose  <- vapply(parts, function(p) sum(lof_allele(p)), integer(1))
  df
}
gp <- add_geno(both |> filter(deposit_id == "MSV000084008", !is.na(cyp2c19_genotype)))
gsub <- gp |> group_by(subject_id, cyp2c19_genotype, act_score, lof_dose) |>
  summarise(ratio = median(log2_ratio), n_files = n(), .groups = "drop") |>
  arrange(act_score)
if (nrow(gsub) >= 8) {
  print(as.data.frame(gsub))
  s1 <- cor.test(gsub$ratio, gsub$act_score, method = "spearman", exact = FALSE)
  cat(sprintf("Spearman(subject ratio, activity_score) rho = %+.3f  p = %.3g  (expect NEG)\n",
              s1$estimate, s1$p.value))
  gsub$carrier <- gsub$lof_dose > 0
  if (dplyr::n_distinct(gsub$carrier) == 2) {
    w <- wilcox.test(ratio ~ carrier, gsub)
    cat(sprintf("LoF carriers *2/*3 (n=%d) median = %.2f  vs  non-carriers (n=%d) median = %.2f;  Wilcoxon p = %.3g\n",
                sum(gsub$carrier), median(gsub$ratio[gsub$carrier]),
                sum(!gsub$carrier), median(gsub$ratio[!gsub$carrier]), w$p.value))
  }
}
```

# Specimen-resolved phenotyping (082493)

Hepatic CYP2C19 is classically read from plasma: can a non-invasive matrix
(urine / skin / feces) recover the genotype? Subject-level is the honest test.



The same comparison as a figure, one panel per matrix and each panel labelled
with the number of subjects behind it:

```{r figure-data}
#| echo: false
# Facet by the MATRIX, not by the deposit. The two accessions are one study, and
# lumping 082493's urine, faeces and skin into a single panel hides that they
# behave differently: the ratio tracks genotype in plasma and not at all in
# urine.
harm_site <- function(x) {
  y <- tolower(trimws(x))
  dplyr::case_when(grepl("plasma", y) ~ "plasma",
                   grepl("^ua|urine", y) ~ "urine",
                   grepl("fecal|feces|faec", y) ~ "faeces",
                   grepl("forehead|skin", y) ~ "skin",
                   TRUE ~ y)
}
# One value per subject per matrix, integrated over the dosing interval, not one
# per file. The ratio falls across the interval, so pooling draws mixes the time
# course into the between-subject comparison -- which is why the single-timepoint
# version of this test came out null. Trapezoid over the recorded minutes, the
# same estimator the AUC section above validates on; faeces records no timepoint
# and therefore supports no curve.
.trap_site <- function(d, ...) {
  d <- d[is.finite(d$timepoint_mins), ]
  if (nrow(d) < 3 || dplyr::n_distinct(d$timepoint_mins) < 3)
    return(data.frame(par = NA_real_, met = NA_real_))
  d <- d[order(d$timepoint_mins), ]
  ar <- function(y) { y[is.na(y)] <- 0
    sum(diff(d$timepoint_mins) * (head(y, -1) + tail(y, -1)) / 2) }
  data.frame(par = ar(d$par_auc), met = ar(d$met_auc))
}
plot_df <- j |>
  filter(deposit_id %in% cyp_ds, !is.na(pheno_rank)) |>
  mutate(site = harm_site(specimen)) |>
  # Urine is excluded rather than left to come out empty. Its parent channel is
  # 96 per cent the isobaric sulfide, so after the isobar gate a urine ratio
  # rests on a handful of files and would read as a null result instead of as
  # the absence of a usable measurement.
  filter(site != "urine") |>
  mutate(
         par_auc = ifelse(par_detected %in% TRUE, par_auc, NA_real_),
         met_auc = ifelse(met_detected %in% TRUE, met_auc, NA_real_)) |>
  group_by(subject_id, cyp2c19_phenotype, pheno_rank, site, timepoint) |>
  group_modify(.trap_site) |>
  ungroup() |>
  filter(is.finite(par), is.finite(met), par > 0, met > 0) |>
  mutate(log2_ratio = log2(par / met),
         phenotype = factor(cyp2c19_phenotype, levels = PHENO))
# label each panel with the number of subjects behind it, so a panel resting on
# two people is not read as equivalent to one resting on fourteen
.nsub <- plot_df |> dplyr::group_by(site) |>
  dplyr::summarise(ns = dplyr::n_distinct(subject_id), .groups = "drop")
stopifnot(nrow(.nsub) > 0)
plot_df$cohort <- factor(
  paste0(plot_df$site, "  (", .nsub$ns[match(plot_df$site, .nsub$site)], " subjects)"),
  levels = paste0(.nsub$site, "  (", .nsub$ns, " subjects)")[order(-.nsub$ns)])
write.csv(plot_df, "artifacts/cyp_validation_data.csv", row.names = FALSE)
```

```{r fig-by-phenotype, fig.width=9, fig.height=4}
#| echo: false
ggplot(plot_df, aes(phenotype, log2_ratio, fill = phenotype)) +
  geom_boxplot(outlier.shape = NA, width = 0.62, alpha = 0.95) +
  geom_jitter(width = 0.15, height = 0, size = 0.8, alpha = 0.45) +
  facet_wrap(~cohort, scales = "free_y") +
  scale_fill_manual(values = pal, guide = "none") +
  scale_x_discrete(labels = PH_SHORT) +
  labs(x = "Genotyped CYP2C19 metaboliser class  (low → high activity)",
       y = "log2( AUC omeprazole / AUC 5-OH-omeprazole )",
       title = "Per-subject AUC ratio by genotyped CYP2C19 class, per matrix") +
  theme_bw(base_size = 11) +
  theme(strip.background = element_rect(fill = "grey92"),
        plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r specimen-resolved}
# One row per subject, as everywhere else: plot_df carries a value per dosing day,
# and a subject measured twice must not enter the test twice.
.by_subject <- plot_df |>
  dplyr::group_by(site, subject_id, pheno_rank) |>
  dplyr::summarise(log2_ratio = mean(log2_ratio), .groups = "drop")
for (sp in sort(unique(.by_subject$site))) {
  z <- .by_subject[.by_subject$site == sp, ]
  n <- nrow(z)
  if (nrow(z) >= 8 && dplyr::n_distinct(z$pheno_rank) >= 3) {
    ct <- cor.test(z$log2_ratio, z$pheno_rank, method = "spearman", exact = FALSE)
    co <- summary(lm(log2_ratio ~ pheno_rank, data = z))$coefficients
    cat(sprintf("  %-7s %2d subjects  rho=%+.3f p=%.3g  |  beta=%+.3f p=%.3g
",
                sp, n, ct$estimate, ct$p.value, co[2, 1], co[2, 4]))
  } else {
    cat(sprintf("  %-7s %2d subjects  (too few, or under three classes)
", sp, n))
  }
}
```


# Controls the probe cohort makes possible

The dosed cohort supports three checks that no other deposit can offer, because
it is the only place where we know what the answer must be before looking.

## Pre-dose samples: biological negatives

Every subject was drawn at minute 0, before the dose. The drug cannot be
present. These are real plasma from real people, with the full complexity of the
matrix, so they are a harder test of specificity than any synthetic decoy.

```{r pre-dose}
#| echo: false
ta_f <- "dataset_metadata/MSV000084008_metadata.tsv"
if (!file.exists(ta_f)) { cat("submitter TSV not present\n") } else {
  ta <- read.delim(ta_f, check.names = FALSE, stringsAsFactors = FALSE)
  ta$k <- norm_file_key(ta$filename)
  r84 <- readRDS("artifacts/per_dataset/MSV000084008.rds")
  mp <- as.data.frame(r84$metrics_par); mm <- as.data.frame(r84$metrics_met)
  mp$k <- norm_file_key(mp$file); mm$k <- norm_file_key(mm$file)
  ta$par_det <- mp$par_detected[match(ta$k, mp$k)] %in% TRUE
  ta$met_det <- mm$met_detected[match(ta$k, mm$k)] %in% TRUE
  ta$tmin <- as.numeric(ta$ATTRIBUTE_Time_Point_Mins)
  z <- ta[ta$MS_SampleType == "sample" & !is.na(ta$tmin), ]
  tb <- data.frame(
    group = c("pre-dose (t = 0)", "post-dose (t > 0)"),
    files = c(sum(z$tmin == 0), sum(z$tmin > 0)),
    parent = c(sum(z$par_det[z$tmin == 0]), sum(z$par_det[z$tmin > 0])),
    metabolite = c(sum(z$met_det[z$tmin == 0]), sum(z$met_det[z$tmin > 0])))
  tb$parent_pct <- sprintf("%.1f%%", 100 * tb$parent / tb$files)
  knitr::kable(tb, row.names = FALSE, caption = "Detections before and after dosing.")
  ft <- fisher.test(matrix(c(tb$parent[1], tb$files[1] - tb$parent[1],
                             tb$parent[2], tb$files[2] - tb$parent[2]), nrow = 2))
  cat(sprintf("\nFisher pre vs post (parent): p = %.3g\n", ft$p.value))
}
```


**Zero detections in 49 pre-dose files, for either compound**, against 60% after
dosing (p = 2e-18). The picker does not invent the drug in samples that cannot
contain it.

## The ratio moves with time since dose

A metabolic ratio is not constant over a dosing interval; that is precisely why
clinical protocols fix the draw at 3 h. If our ratio is measuring pharmacology
rather than noise, it must fall as the parent clears faster than its metabolite.
Nothing in the pipeline was tuned to produce this, and the test is run **within
subject**, so between-subject phenotype differences cannot generate it.

```{r ratio-vs-time}
#| echo: false
# Build a minimal frame rather than mutating the submitter TSV: that file
# carries duplicate column names, which dplyr refuses to transform.
if (exists("z")) {
  mpa <- as.data.frame(r84$metrics_par); mma <- as.data.frame(r84$metrics_met)
  mpa$k <- norm_file_key(mpa$file); mma$k <- norm_file_key(mma$file)
  tt <- data.frame(
    subject = as.character(z$ATTRIBUTE_Subject),
    day     = as.character(z$ATTRIBUTE_Study_Day),
    tmin    = z$tmin,
    par_det = z$par_det, met_det = z$met_det,
    par_auc = mpa$par_auc[match(z$k, mpa$k)],
    met_auc = mma$met_auc[match(z$k, mma$k)],
    stringsAsFactors = FALSE)
  tt <- tt[tt$par_det & tt$met_det & tt$tmin > 0, ]
  tt$l2 <- log2(tt$par_auc / tt$met_auc)
  tt <- tt[is.finite(tt$l2), ]
  tt$bin <- cut(tt$tmin, c(0, 30, 60, 120, 240, 480, Inf),
                labels = c("<=30m", "30-60m", "1-2h", "2-4h", "4-8h", ">8h"))
  agg <- do.call(rbind, lapply(levels(tt$bin), function(b) {
    q <- tt[tt$bin == b, ]; if (!nrow(q)) return(NULL)
    data.frame(bin = b, files = nrow(q), subjects = length(unique(q$subject)),
               median_log2 = round(median(q$l2), 2))
  }))
  print(agg, row.names = FALSE)
  rho <- vapply(split(tt, tt$subject), function(q)
    if (nrow(q) >= 4 && length(unique(q$tmin)) >= 4)
      cor(q$tmin, q$l2, method = "spearman") else NA_real_, numeric(1))
  rho <- rho[is.finite(rho)]
  bt <- binom.test(sum(rho < 0), length(rho))
  cat(sprintf("
within-subject Spearman(time, ratio): median rho = %+.3f
", median(rho)))
  cat(sprintf("%d of %d subjects negative, sign test p = %.4g
",
              sum(rho < 0), length(rho), bt$p.value))
}
```


The ratio falls monotonically across the dosing interval and
**`r if (exists("rho")) sum(rho < 0) else NA` of
`r if (exists("rho")) length(rho) else NA` subjects show a negative slope
individually** (sign test p =
`r if (exists("bt")) signif(bt$p.value, 2) else NA`). This is ordinary
pharmacokinetics recovered from public files, and it is the clearest evidence
that the number behaves like a drug measurement rather than an artefact of the
extraction.

It also explains why the single-timepoint ratio was null against genotype while
the AUC works: a ratio that moves this much with sampling time carries the draw
time as noise.

```{r fig-ratio-and-detection, fig.width=9.4, fig.height=5.4}
#| echo: false
# Plasma only. Urine used to sit beside it to show that its null ratio was not a
# detection failure -- but after the isobar gate urine has no ratio to defend, and
# the claim that it detected both compounds more often than plasma was measuring
# the sulfide. What urine does say is in the paragraph below, where it is a
# statement about the metabolite rather than about the ratio.
KEEP <- "plasma"
rat <- plot_df |>
  dplyr::filter(site %in% KEEP) |>
  dplyr::transmute(site, phenotype, value = log2_ratio,
                   panel = "the ratio: log2( AUC omeprazole / AUC 5-OH )")

# The denominator must be every file of that matrix, not only the ones where the
# picker returned a peak. metrics_* holds a row only when pick() found
# something, so building this from `j` alone silently drops the empty files and
# inflates every rate (plasma read 78-94% against a true 53%). Start from the
# sample metadata and treat a file with no metrics row as "neither".
det <- rs |>
  dplyr::filter(deposit_id %in% cyp_ds) |>
  dplyr::mutate(site = harm_site(specimen),
                phenotype = factor(cyp2c19_phenotype, levels = PHENO)) |>
  dplyr::filter(site %in% KEEP, !is.na(phenotype)) |>
  dplyr::left_join(dplyr::distinct(j, stem, detection_tier), by = "stem") |>
  dplyr::mutate(detection_tier = ifelse(is.na(detection_tier), "neither", detection_tier)) |>
  dplyr::group_by(site, phenotype) |>
  dplyr::summarise(files = dplyr::n(),
                   parent     = mean(detection_tier %in% c("both", "parent_only")),
                   metabolite = mean(detection_tier %in% c("both", "metab_only")),
                   .groups = "drop") |>
  tidyr::pivot_longer(c(parent, metabolite), names_to = "compound", values_to = "value") |>
  dplyr::mutate(panel = "the detection behind it: % of files with the compound",
                value = 100 * value,
                compound = factor(compound, levels = c("parent", "metabolite"),
                                  labels = c("omeprazole (parent)",
                                             "5-OH-omeprazole (metabolite)")))

# facet_grid frees scales by ROW only, which would force the percentages onto
# the ratio's axis. One combined facet variable with facet_wrap gives each panel
# its own.
lv <- c(paste("plasma —", unique(rat$panel)), paste("plasma —", unique(det$panel)))
rat$pan <- factor(paste(rat$site, "—", rat$panel), levels = lv)
det$pan <- factor(paste(det$site, "—", det$panel), levels = lv)

# Each volunteer contributes ~36 plasma files (9 draws x 2 dosing occasions x
# duplicate aliquots), so a class holding ONE person still shows dozens of files
# and its bar looks as solid as the eight-person Normal class. Put the volunteer
# count on the axis: it is shared by both columns, so it labels the boxplots and
# the bars at once.
.nvol <- rs |>
  dplyr::filter(deposit_id %in% cyp_ds, !is.na(cyp2c19_phenotype)) |>
  dplyr::distinct(subject_id, cyp2c19_phenotype) |>
  dplyr::count(cyp2c19_phenotype)
.lab_n <- setNames(sprintf("%s\n(%d)", PH_SHORT,
                           .nvol$n[match(PHENO, .nvol$cyp2c19_phenotype)]), PHENO)

ggplot() +
  geom_boxplot(data = rat, aes(phenotype, value, fill = phenotype),
               outlier.shape = NA, width = 0.62, alpha = 0.95) +
  geom_jitter(data = rat, aes(phenotype, value),
              width = 0.15, height = 0, size = 0.7, alpha = 0.4) +
  geom_col(data = det, aes(phenotype, value, fill = phenotype, alpha = compound),
           position = position_dodge(width = 0.72), width = 0.68, colour = "grey30",
           linewidth = 0.25) +
  # "%" on every label: a bare number above a bar reads as a count, and with ~36
  # files per volunteer a "50" is easily taken for 50 detections rather than 50
  # per cent.
  geom_text(data = det, aes(phenotype, value, group = compound,
                            label = sprintf("%.0f%%", value)),
            position = position_dodge(width = 0.72), vjust = -0.35, size = 2.5) +
  facet_wrap(~ pan, scales = "free_y", ncol = 2) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.13))) +  # room for the 100% labels
  scale_fill_manual(values = pal, guide = "none") +
  # Colour already carries the genotype class on both sides, so the compound is
  # carried by shading. Overriding the swatch fill is what makes that legend
  # readable at all.
  scale_alpha_manual(values = c(1, 0.42), name = "Bars (right-hand panels):",
                     guide = guide_legend(
                       override.aes = list(fill = "grey35", colour = "grey30"))) +
  scale_x_discrete(labels = .lab_n) +
  labs(x = "Genotyped CYP2C19 metaboliser class  (low → high activity), volunteers in brackets",
       y = NULL,
       title = "Left: the ratio.  Right: the detection rate behind it.",
       subtitle = paste("Each volunteer gives ~36 files,",
                        "so a one-person class still fills a bar")) +
  theme_bw(base_size = 11) +
  theme(strip.background = element_rect(fill = "grey92"),
        strip.text = element_text(size = 9),
        plot.title = element_text(face = "bold"),
        legend.position = "top",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r fig-detect-vs-time, fig.width=9.4, fig.height=4.4}
#| echo: false
# Detection rate against time since dose, for MS2 and for the MS1 picker. Every
# volunteer received omeprazole, so this is an absorption curve reconstructed
# from detections: it should be empty before the dose AND at 5 min (not yet
# absorbed), rise to a maximum at 4-5 h, and fall by 8 h. The picker should sit
# at or above MS2 everywhere, and the gap should widen at the tails where the
# concentration is low and data-dependent acquisition stops triggering.
h_all <- read.csv("artifacts/hits_curated.csv", check.names = FALSE, stringsAsFactors = FALSE)
h_all$k <- norm_file_key(h_all$file_path)
.r  <- readRDS("artifacts/per_dataset/MSV000084008.rds")
.mp <- as.data.frame(.r$metrics_par); .mp$k <- norm_file_key(.mp$file)
.mm <- as.data.frame(.r$metrics_met); .mm$k <- norm_file_key(.mm$file)
.rs <- rs; .rs$k <- norm_file_key(.rs$file)
.s <- .rs[which(.rs$deposit_id == "MSV000084008" &
                grepl("plasma", tolower(.rs$specimen)) &
                is.finite(.rs$timepoint_mins)), ]
.hk <- function(role) h_all$k[h_all$ATTRIBUTE_DatasetAccession == "MSV000084008" &
                              h_all$role == role]
.d <- data.frame(
  tmin = .s$timepoint_mins,
  pheno = .s$cyp2c19_phenotype,
  `omeprazole|MS2`            = .s$k %in% .hk("parent"),
  `omeprazole|MS1 picker`     = .mp$par_detected[match(.s$k, .mp$k)] %in% TRUE,
  `5-OH-omeprazole|MS2`       = .s$k %in% .hk("metabolite"),
  `5-OH-omeprazole|MS1 picker`= .mm$met_detected[match(.s$k, .mm$k)] %in% TRUE,
  check.names = FALSE)
.d$pheno <- NULL   # the genotype split is plotted separately, below

.long <- .d |>
  tidyr::pivot_longer(-tmin, names_to = "series", values_to = "det") |>
  tidyr::separate(series, c("compound", "route"), sep = "\\|") |>
  dplyr::group_by(tmin, compound, route) |>
  dplyr::summarise(pct = 100 * mean(det), n = dplyr::n(), .groups = "drop") |>
  dplyr::mutate(compound = factor(compound, levels = c("omeprazole", "5-OH-omeprazole")))

# Evenly spaced categories, not real time: on a linear axis the 0 and 5 min
# draws overlap, and those two points carry the negative-control argument.
.lv <- sort(unique(.long$tmin))
.lab <- c("0" = "pre-dose", "5" = "5 min", "30" = "30 min", "60" = "1 h",
          "120" = "2 h", "240" = "4 h", "300" = "5 h", "360" = "6 h", "480" = "8 h")
.long$tf <- factor(.lab[as.character(.long$tmin)], levels = unname(.lab[as.character(.lv)]))

# (this is now the top row of the combined figure in the next chunk)
```

```{r fig-detect-by-genotype, fig.width=9.4, fig.height=7.0}
#| echo: false
# The same absorption curve, split by genotype. Purely descriptive: three of the
# five classes hold a single volunteer, so the classes are pooled into low /
# normal / high activity and the volunteer count is written into each label. No
# test is run on this - the formal genotype result is the ratio, not presence.
.g <- data.frame(
  tmin  = .s$timepoint_mins,
  pheno = factor(.s$cyp2c19_phenotype, levels = PHENO),
  omeprazole        = .mp$par_detected[match(.s$k, .mp$k)] %in% TRUE,
  `5-OH-omeprazole` = .mm$met_detected[match(.s$k, .mm$k)] %in% TRUE,
  check.names = FALSE)
.g <- .g[!is.na(.g$pheno), ]
.g$grp <- cut(as.integer(.g$pheno), c(0, 2, 3, 5),
              labels = c("low activity", "normal activity", "high activity"))
.n <- tapply(.s$subject_id[!is.na(.s$cyp2c19_phenotype)],
             cut(match(.s$cyp2c19_phenotype[!is.na(.s$cyp2c19_phenotype)], PHENO),
                 c(0, 2, 3, 5), labels = c("low activity", "normal activity",
                                           "high activity")),
             function(x) length(unique(x)))
levels(.g$grp) <- sprintf("%s (%d volunteers)", levels(.g$grp), .n[levels(.g$grp)])

.gl <- .g |>
  tidyr::pivot_longer(c(omeprazole, `5-OH-omeprazole`),
                      names_to = "compound", values_to = "det") |>
  dplyr::group_by(tmin, grp, compound) |>
  dplyr::summarise(pct = 100 * mean(det), .groups = "drop") |>
  dplyr::mutate(compound = factor(compound,
                                  levels = c("omeprazole", "5-OH-omeprazole")))
.gl$tf <- factor(.lab[as.character(.gl$tmin)], levels = unname(.lab[as.character(.lv)]))

# One figure, four panels: compound across the columns, and down the rows the
# two questions asked of the same curve - what each route reaches, and whether
# genotype shows in presence. Sharing the axis makes the two directly
# comparable.
.top <- .long |> dplyr::transmute(tf, compound, series = route, pct,
                                  row = "by route")
.bot <- .gl   |> dplyr::transmute(tf, compound, series = as.character(grp), pct,
                                  row = "by genotype")
.cmb <- dplyr::bind_rows(.top, .bot)
.cmb$row <- factor(.cmb$row, levels = c("by route", "by genotype"))
.cmb$series <- factor(.cmb$series, levels = c("MS2", "MS1 picker", levels(.gl$grp)))

.p_det <- ggplot(.cmb, aes(tf, pct, colour = series, linetype = series, group = series)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.9) +
  facet_grid(row ~ compound) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_colour_manual(values = setNames(
    c("#B2182B", "#2166AC", "#B2182B", "grey35", "#2166AC"), levels(.cmb$series))) +
  scale_linetype_manual(values = setNames(
    c("solid", "solid", "22", "22", "22"), levels(.cmb$series))) +
  labs(x = NULL, y = "% of files with the compound",
       colour = NULL, linetype = NULL, tag = "A") +
  theme_bw(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "grey92")) +
  guides(colour = guide_legend(nrow = 1))

# Panel B: the ratio over the same axis. Presence and the quantity are two views
# of the same absorption curve, so they belong in one figure rather than two.
.rt <- tt |>
  dplyr::group_by(subject, day, tmin) |>
  dplyr::summarise(l2 = median(l2), .groups = "drop") |>
  dplyr::mutate(series = paste(subject, day))
.rt$tf <- factor(.lab[as.character(.rt$tmin)], levels = unname(.lab[as.character(.lv)]))
.rmed <- .rt |> dplyr::group_by(tf) |>
  dplyr::summarise(l2 = median(l2), n = dplyr::n_distinct(subject), .groups = "drop")

.p_rat <- ggplot(.rt, aes(tf, l2)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey70") +
  geom_line(aes(group = series), colour = "grey78", linewidth = 0.4) +
  geom_point(colour = "grey50", size = 0.9, alpha = 0.7) +
  geom_line(data = .rmed, aes(group = 1), colour = "#111111", linewidth = 1.1) +
  geom_point(data = .rmed, aes(fill = n < 5), shape = 21, size = 2.4,
             colour = "#111111", stroke = 0.8) +
  geom_text(data = .rmed, aes(label = n), vjust = -1.1, size = 2.7) +
  scale_fill_manual(values = c(`FALSE` = "#111111", `TRUE` = "white"), guide = "none") +
  labs(x = "Time since dose", y = "log2( omeprazole / 5-OH )", tag = "B") +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank())

.p_comb <- patchwork::wrap_plots(.p_det, .p_rat, ncol = 1, heights = c(2, 1.1))
dir.create("thesis/graphics", showWarnings = FALSE, recursive = TRUE)
ggsave("thesis/graphics/flagship_detect_combined.png", .p_comb,
       width = 9.4, height = 7.0, dpi = 200)
.p_comb
```

```{r fig-ratio-vs-time, fig.width=9.2, fig.height=4.6}
#| echo: false
if (exists("tt") && nrow(tt) > 10) {
  tp <- tt |>
    dplyr::group_by(subject, day, tmin) |>          # one point per draw, not per injection
    dplyr::summarise(l2 = median(l2), .groups = "drop") |>
    dplyr::mutate(series = paste(subject, day),
                  panel  = paste0("Study day ", day,
                                  ifelse(day == min(day), "  (first dose)", "  (second dose)")))
  # A ratio needs BOTH compounds detected, and early in the interval the
  # metabolite often is not there yet, so the panel is badly unbalanced: 30 min
  # rests on 2 subjects against 13 at 4 h. Without the counts a reader takes the
  # left end of the median line for the cohort. Points on <5 subjects are drawn
  # hollow and the line dashed there.
  med <- tp |> dplyr::group_by(panel, tmin) |>
    dplyr::summarise(l2 = median(l2), n = dplyr::n_distinct(subject), .groups = "drop") |>
    dplyr::mutate(sparse = n < 5)

  ggplot(tp, aes(tmin, l2)) +
    geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey70") +
    geom_line(aes(group = series), colour = "grey72", linewidth = 0.45) +
    geom_point(colour = "grey45", size = 1, alpha = 0.8) +
    geom_line(data = med, colour = "#111111", linewidth = 1.1) +
    geom_point(data = med, aes(fill = sparse), colour = "#111111", size = 2.4,
               shape = 21, stroke = 0.9) +
    geom_text(data = med, aes(label = n), vjust = -1.15, size = 2.9, colour = "#111111") +
    scale_fill_manual(values = c(`FALSE` = "#111111", `TRUE` = "white"), guide = "none") +
    facet_wrap(~ panel) +
    scale_x_continuous(trans = "log10",
                       breaks = c(5, 30, 60, 120, 240, 360, 480),
                       labels = c("5 m", "30 m", "1 h", "2 h", "4 h", "6 h", "8 h")) +
    labs(x = "Time since dose (log scale). Pre-dose draws at t = 0 carry no drug, so no ratio exists there",
         y = "log2( omeprazole / 5-OH-omeprazole )",
         title = "log2(omeprazole / 5-OH) across the dosing interval, both doses, MSV000084008",
         subtitle = paste("Numbers are subjects contributing a ratio at that draw;",
                          "hollow points rest on fewer than 5")) +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "grey93", colour = NA),
          strip.text = element_text(face = "bold", size = 10.5),
          plot.title = element_text(face = "bold", size = 12))
}
```

Urine is worth a sentence of its own, because it fails and succeeds in different
channels. Omeprazole is extensively metabolised before renal excretion, so the
parent is found in $4$ of $140$ urine files while 5-hydroxyomeprazole is found
in $110$ --- and the parent channel there is not merely empty but misleading,
since $96$ per cent of the peaks the picker first accepted at that mass were the
isobaric sulfide. Two things follow for anyone mining archives for exposure.
Which compound to search for depends on the matrix: in urine the metabolite is
the marker and the parent is a trap. And urine is the one matrix where
propagation earns almost nothing --- the tandem search already found the
metabolite in $106$ of $112$ files and the extraction adds three --- because
propagation works at the edge of the measurable range, and an abundant renal
metabolite is not at that edge.

## Does it agree with the authors' own measurement?

The genotype above is a *prediction* of metabolic capacity. The deposit also
carries a *measurement*: `PharmKinetics_CYP2C19`, the per-subject value the
original authors determined by targeted HPLC-MS/MS (MRM 346 $\to$ 198 and 362
$\to$ 214), which our extraction never sees. It is the strongest external check
available here, because it is per-subject and comes from the deposit rather than
from a number transcribed out of a paper.

```{r vs-published-pk}
#| echo: false
.pk <- both |>
  dplyr::filter(grepl("plasma", tolower(specimen)),
                is.finite(pharmkinetics_cyp2c19)) |>
  dplyr::group_by(subject_id, timepoint) |>
  dplyr::summarise(ours = median(log2_ratio),
                   theirs = dplyr::first(pharmkinetics_cyp2c19),
                   rank = dplyr::first(pheno_rank), .groups = "drop")
.gm <- do.call(rbind, lapply(sort(unique(.pk$timepoint)), function(d) {
  q <- .pk[.pk$timepoint == d, ]
  data.frame(`study day` = d, subjects = nrow(q),
             `their geometric mean` = round(exp(mean(log(q$theirs))), 2),
             `published (Table 1)` = c("1" = 1.4, "9" = 1.2)[[as.character(d)]],
             check.names = FALSE)
}))
knitr::kable(.gm, row.names = FALSE, caption = paste(
  "The deposit column reproduces the paper's own cohort values, which is how we",
  "know it is the Table 1 parameter and not some other measure."))

.sub <- .pk |> dplyr::group_by(subject_id) |>
  dplyr::summarise(ours = mean(ours), theirs = mean(theirs),
                   rank = dplyr::first(rank), .groups = "drop")
.ot <- cor.test(.sub$ours, .sub$theirs, method = "spearman", exact = FALSE)
.og <- cor.test(.sub$ours, .sub$rank, method = "spearman", exact = FALSE)
.tg <- cor.test(.sub$theirs, .sub$rank, method = "spearman", exact = FALSE)
knitr::kable(data.frame(
  comparison = c("ours vs theirs", "ours vs genotype rank",
                 "theirs vs genotype rank"),
  rho = sprintf("%+.3f", c(.ot$estimate, .og$estimate, .tg$estimate)),
  p = signif(c(.ot$p.value, .og$p.value, .tg$p.value), 2)),
  row.names = FALSE,
  caption = paste("Spearman over", nrow(.sub), "volunteers."))
```

The three correlations are consistent, but only once the direction of their
parameter is pinned down, and Table 1's label is not a reliable guide to it. The
row reads `OME/5-OH OME (AUC)`, yet the numbers behave as the reciprocal, on
three independent grounds:

- Table 1 reports caffeine and midazolam as oral clearances and lists omeprazole
  between them. The Results then read the fall in all three together --- *"the
  decrease in the LS-GMRs ... suggest a decrease in CYP1A2, CYP2C19, and CYP3A4
  activities"* --- and the Discussion repeats it, *"reflecting decreased
  activities of CYP1A2, CYP2C19, and CYP3A, respectively"*. A clearance falls
  when activity falls; a parent-over-metabolite ratio would *rise*. For either
  sentence to hold, the omeprazole parameter has to move with activity as the
  two clearances do. The paper offers no other reading of its direction.
- Their values rise along the genotype ladder (rho =
  `r sprintf("%+.3f", .tg$estimate)`), which parent-over-metabolite cannot do
  and metabolite-over-parent must.
- Their own star alleles order their own values. DM014 (`*2/*3`, two
  loss-of-function alleles) holds the cohort's two lowest values, 0.28 and 0.34;
  DM010 (`*17/*17`, two gain-of-function alleles) holds the highest, 3.91; the
  `*1/*2` heterozygotes cluster just above DM014. Read as parent-over-metabolite
  that ordering would make the cohort's poorest metaboliser its most efficient.

The paper does describe phenotyping generically as *"oral clearance or
parent:metabolite area under the curve (AUC) ratios"*, so the label is not a
one-off slip --- but the numbers, the genotypes and the activity sentence all
point the other way, and it is the numbers we join to.

Read as metabolite-over-parent, theirs is the reciprocal of ours, so agreement
has to appear as a negative rank correlation --- and it does, at
**`r sprintf("%+.3f", .ot$estimate)`** over `r nrow(.sub)` volunteers (p =
`r signif(.ot$p.value, 2)`), close to what the same ratio achieves against the
genotype itself. Two independent read-outs of one enzyme, one predicted from DNA
and one measured by a targeted assay, land on the same ordering of the same
people from spectra neither of them saw.

What the comparison cannot do is reproduce their absolute level. Without an
isotope-labelled internal standard our response-factor offset is uncalibrated,
so the ranking is what transfers, not the number.

# MS1 propagation gain

How many extra genotype-linked measurements the MS1 EIC recovers beyond what the
targeted MS2 data already provided.

One caution about the denominator, because it is easy to inflate this number.
The **SIRIUS-confirmed seed files are not the MS2 baseline**: they are a capped,
curated subset sent for confirmation, so dividing MS1 detections by seeds
compares two different populations and overstates the gain several-fold. The
honest baseline is the set of files MASST matched, which is what `05` Part A
reports corpus-wide. Those figures are quoted there rather than repeated here,
because a number copied between notebooks is a number that will eventually
disagree with itself. Seeds are reported here only to show how few confirmed
anchors the propagation starts from.

```{r propagation-gain}
for (d in cyp_ds) {
  r    <- readRDS(sprintf("artifacts/per_dataset/%s.rds", d))
  seed <- length(unique(basename(c(r$cfg$confirmed_files_parent,
                                   r$cfg$confirmed_files_metabolite))))
  detf  <- sum(pf$dataset_id == d & pf$detection_tier != "neither", na.rm = TRUE)
  bothf <- sum(pf$dataset_id == d & pf$detection_tier == "both",    na.rm = TRUE)
  cat(sprintf("  %s: %d confirmed anchors -> %d files with a detection (%d with both)\n",
              d, seed, detf, bothf))
}
```

```{r tbl-propagation}
#| echo: false
# Propagation in a cohort where the answer is known: every volunteer received
# omeprazole, so a post-dose file is one where the compound is expected. MS2
# reaches the files it happened to fragment; the picker reaches those plus the
# rest. Pre-dose and 5 min are shown too: the drug is not yet absorbed, so both
# routes should be empty there.
.hk2 <- function(ds, role) h_all$k[h_all$ATTRIBUTE_DatasetAccession == ds &
                                   h_all$role == role]
.grab <- function(ds, pat) {
  .rr <- readRDS(sprintf("artifacts/per_dataset/%s.rds", ds))
  # gate the isobars here too, or this table credits propagation with peaks the
  # corpus tables reject: in urine 86 of 90 parent peaks sit at the metabolite's
  # retention time, not omeprazole's
  .g <- resolve_isobars(.rr$metrics_par, .rr$metrics_met,
                        read.csv("artifacts/hits-confirmed-all.csv",
                                 stringsAsFactors = FALSE), ds, .rr$rt_anchor)
  .a <- as.data.frame(.g$par); .a$k <- norm_file_key(.a$file)
  .b <- as.data.frame(.g$met); .b$k <- norm_file_key(.b$file)
  .z <- .rs[which(.rs$deposit_id == ds & grepl(pat, tolower(.rs$specimen))), ]
  data.frame(tmin = .z$timepoint_mins,
             par_ms2 = .z$k %in% .hk2(ds, "parent"),
             met_ms2 = .z$k %in% .hk2(ds, "metabolite"),
             par_pick = .a$par_detected[match(.z$k, .a$k)] %in% TRUE,
             met_pick = .b$met_detected[match(.z$k, .b$k)] %in% TRUE)
}
.plas <- .grab("MSV000084008", "plasma")
.urin <- .grab("MSV000082493", "urine")
.skin <- .grab("MSV000082493", "skin|forehead")

.row <- function(d, label) data.frame(
  Stratum = label, Files = nrow(d),
  `MS2 parent` = sum(d$par_ms2), `Picker parent` = sum(d$par_pick),
  `MS2 metabolite` = sum(d$met_ms2), `Picker metabolite` = sum(d$met_pick),
  check.names = FALSE)

.lab <- c("0" = "plasma, pre-dose", "5" = "plasma, 5 min", "30" = "plasma, 30 min",
          "60" = "plasma, 1 h", "120" = "plasma, 2 h", "240" = "plasma, 4 h",
          "300" = "plasma, 5 h", "360" = "plasma, 6 h", "480" = "plasma, 8 h")
.tab <- dplyr::bind_rows(
  lapply(sort(unique(.plas$tmin[is.finite(.plas$tmin)])),
         function(t) .row(.plas[.plas$tmin %in% t, ], .lab[[as.character(t)]])),
  .row(.plas[which(.plas$tmin > 0), ], "plasma, all post-dose"),
  .row(.urin[which(.urin$tmin > 0), ], "urine, all post-dose"),
  .row(.skin[which(.skin$tmin > 0), ], "skin, all post-dose"))
.tab$`Propagation multiplier` <- ifelse(
  .tab$`MS2 parent` + .tab$`MS2 metabolite` == 0, "-",
  sprintf("%.2f", (.tab$`Picker parent` + .tab$`Picker metabolite`) /
                  pmax(1, .tab$`MS2 parent` + .tab$`MS2 metabolite`)))
# The thesis table is built from this, so it is written on every render.
write.csv(.tab, "artifacts/propagation_dosed_cohort.csv", row.names = FALSE)
knitr::kable(.tab, caption = paste(
  "Files with each compound detected by MS2 and by the MS1 picker, in a cohort where",
  "every volunteer received omeprazole. Pre-dose and 5-minute draws precede absorption",
  "and are empty by both routes. The multiplier is picker detections over MS2",
  "detections, summed across the two compounds; it is undefined where MS2 finds",
  "nothing, which is the case for the parent throughout urine."))
```

Urine is where propagation is worth least. Omeprazole is largely metabolised
before renal excretion, so the tandem silence there is correct; the isobar gate
of `03` cuts the extraction's parent count from $90$ to $4$, the rest having sat
beside 5-OH-omeprazole rather than $9.6$ s after it.
