---
title: "04: Cross-dataset exploration"
subtitle: "What each dataset contributes, what can be compared across them, and what a meta-analysis would still need."
author: "Philippine Louail"
date: today
format:
  html:
    df-print: paged
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(forcats)
library(scales)
library(ggrepel)
source("R/_setup.R")
# _setup.R attaches Bioconductor packages (S4Vectors via Spectra/MetaboAnnotation)
# whose verbs `first`, `rename`, `filter`, etc. shadow the dplyr ones. Re-attach
# dplyr / tidyr so the tidyverse verbs sit on top of the search path again.
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
})
```

# What this notebook does

`03` measured omeprazole and 5-hydroxyomeprazole in every biological file of every
confirmed dataset. This notebook asks what the corpus as a whole supports, in two
explicitly separated meta-analyses plus the context needed to read them.

- **Context first** (kept from the exploratory pass): what each dataset is, how
  much signal it contributes, and the one hard constraint that shapes everything,
  that an absolute ratio is not comparable across datasets while a within-study
  contrast is.
- **Part A: expanding the classic reverse-metabolomics prevalence analysis.** The
  field reports molecule-by-cohort *detection rates* from MS2 matches. We run the
  same analysis with MS1 propagation and compare MS2 vs MS1 detection per compound
  per cohort. This is presence-based, so it is isomer-immune, and it is where the
  pipeline's value shows most cleanly.
- **Part B: a pipeline-specific meta-analysis.** The quantitative, error-controlled
  layer the field does not have, built on the proteomics match-between-runs
  precedent: a decoy-controlled false-transfer rate for the propagated detections,
  and the within-study contrast as the poolable quantity.

**What it deliberately does not do: pool an absolute ratio into a single
cross-cohort estimate.** An earlier version did, and the result was meaningless
(most datasets only share a `sex` contrast). Pooling is a real option but needs a
shared, biologically meaningful contrast we do not yet have, so it is described
rather than run (final section).

# Load the inputs

```{r load-inputs}
samples  <- read_csv("artifacts/sample_table_all.csv",  show_col_types = FALSE)
datasets <- read_csv("artifacts/datasets-confirmed.csv", show_col_types = FALSE)

# Per-file table (not deduped to subjects) + the raw MASST hits: the prevalence
# meta-analysis in Part A is a per-file rate, so it must compare like with like
# (per-file MS1 detection vs per-file MS2/MASST detection).
samples_perfile <- read_csv("artifacts/sample_table_perfile.csv", show_col_types = FALSE)
masst_hits      <- read.csv("artifacts/hits_curated.csv", check.names = FALSE,
                            stringsAsFactors = FALSE)

# has_metadata flag from 02 (metadata resolved in 01): restrict every analysis to
# files a metadata source covers; record-only files stay out of stratified work.
bio_meta_per_ds <- readRDS("artifacts/bio_meta_per_ds.rds")
has_meta_lookup <- unlist(lapply(names(bio_meta_per_ds), function(ds) {
  v <- bio_meta_per_ds[[ds]]
  stats::setNames(as.logical(v), paste0(ds, "|", basename(names(v))))
}))
# Normalise a file name for matching across sources (drop our accession prefix,
# extension, case): the same rule the pipeline uses everywhere.
fkey <- function(x) tolower(sub("[.][^.]*$", "",
          sub("^(MSV[0-9]+|MTBLS[0-9]+|ST[0-9]+)[_-]", "", basename(as.character(x)))))
```

`sample_table_all.csv` carries one row per subject (`03` de-duplicates files to
subjects where subject metadata exists; where it does not, a row is a file).
Throughout this notebook "sample" means one row of that table, except in Part A
which is explicitly per-file.

```{r short-labels}
# The MetaboLights and Workbench ids carry an assay suffix that makes every axis
# unreadable. Shorten for display only; dataset_id stays the join key.
short_ds <- function(x) sub("_AN[0-9]+$", "", sub("_LC-MS_.*$", "", x))
```

# The datasets

Before any analysis, the cast. Each row is one confirmed (deposit, assay) entity:
where it came from, what it is, how much usable signal it contributes, and which
grouping variables it offers for a contrast.

```{r dataset-summary}
#| echo: false
# A dataset can only support a contrast on a variable that has >= 2 levels with
# >= 3 samples each: the same rule fit_within_study() applies below. Reporting it
# here makes the "what can this dataset answer?" question explicit up front.
usable_contrasts <- function(df) {
  cand <- c("health_status", "group", "sex", "country", "body_site")
  ok <- vapply(cand, function(v) {
    if (!v %in% names(df)) return(FALSE)
    tb <- table(df[[v]][!is.na(df[[v]])])
    length(tb) >= 2 && all(tb >= 3)
  }, logical(1))
  if (!any(ok)) "none" else paste(cand[ok], collapse = ", ")
}

ds_table <- samples |>
  group_by(dataset_id) |>
  summarise(
    n_samples = n(),
    n_both    = sum(detection_tier == "both", na.rm = TRUE),
    both_rate = n_both / n_samples,
    contrasts = usable_contrasts(pick(everything())),
    .groups   = "drop") |>
  left_join(datasets |> select(dataset_id, repository, chromatography,
                               const_country, const_body_site),
            by = "dataset_id") |>
  mutate(dataset = short_ds(dataset_id),
         both_rate = percent(both_rate, accuracy = 0.1)) |>
  select(dataset, repository, body_site = const_body_site,
         country = const_country, chromatography,
         n_samples, n_both, both_rate, contrasts) |>
  arrange(desc(n_both))

ds_table
```

Read this table as the corpus' honest capacity. Three things it makes obvious:

- **`n_both` is what the ratio analysis actually runs on.** Everything else is a
  sample where at least one of the two compounds was not detected, so no ratio can
  be formed. Across the whole corpus this number is small, and it is concentrated
  in a few deposits.
- **The deposits are not the same kind of thing.** A couple are drug-probe
  cohorts where every subject was dosed by design. Most are general-population or
  disease cohorts where a handful of people happened to be on omeprazole. Those
  two categories cannot be read the same way.
- **`contrasts` shows what each dataset can be asked.** Note how few offer a
  biologically meaningful grouping. That single column is the reason the last
  section of this notebook exists.

# Exploratory look at the corpus

This section is descriptive on purpose. Before modelling anything, it is worth
seeing what the data look like.

## Detection-tier composition per dataset

Counts on the left, the same thing proportionally on the right. The left panel
shows where the signal actually is; the right panel shows what fraction of each
cohort was exposed.

```{r exp-tier-bars}
#| fig-width: 6
#| fig-height: 4.5
#| layout-ncol: 2
tier_levels  <- c("both", "parent_only", "metab_only", "neither")
tier_palette <- c(both        = "#1b9e77",
                  parent_only = "#7570b3",
                  metab_only  = "#d95f02",
                  neither     = "grey75")

tier_long <- samples |>
  mutate(detection_tier = factor(detection_tier, levels = tier_levels),
         ds = short_ds(dataset_id)) |>
  count(ds, detection_tier) |>
  group_by(ds) |>
  mutate(p = n / sum(n)) |>
  ungroup()

ggplot(tier_long,
       aes(x = fct_reorder(ds, p * (detection_tier == "both"),
                           .fun = sum, .desc = FALSE),
           y = n, fill = detection_tier)) +
  geom_col() + coord_flip() +
  scale_fill_manual(values = tier_palette) +
  labs(x = NULL, y = "samples", fill = "tier",
       title = "Detection tier, counts") +
  theme_bw() + theme(legend.position = "bottom")

ggplot(tier_long,
       aes(x = fct_reorder(ds, p * (detection_tier == "both"),
                           .fun = sum, .desc = FALSE),
           y = p, fill = detection_tier)) +
  geom_col() + coord_flip() +
  scale_fill_manual(values = tier_palette) +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = "share of samples", fill = "tier",
       title = "Detection tier, proportional") +
  theme_bw() + theme(legend.position = "bottom")
```

The `parent_only` and `metab_only` slices are censored ratio observations, not
failures: we know the ratio is very high or very low, but not by how much,
because one peak fell below detection. Discarding them biases any ratio estimate
toward the middle, which is a point the final section returns to.

## Deposit size vs detection rate

```{r exp-n-vs-both, fig.width = 6.5, fig.height = 5}
ds_summary <- samples |>
  group_by(dataset_id) |>
  summarise(n_samples = n(),
            n_both    = sum(detection_tier == "both"),
            both_rate = n_both / n_samples,
            .groups   = "drop") |>
  left_join(datasets |> select(dataset_id, chromatography,
                               const_country, const_body_site),
            by = "dataset_id")

ggplot(ds_summary, aes(x = n_samples, y = both_rate)) +
  geom_point(aes(size = n_both, colour = chromatography), alpha = 0.85) +
  geom_text_repel(aes(label = short_ds(dataset_id)), size = 3,
                  min.segment.length = 0) +
  scale_x_log10() + scale_y_continuous(labels = percent) +
  scale_size_continuous(range = c(2, 8)) +
  labs(x = "samples in deposit (log10)", y = "'both' detection rate",
       size = "n with both",
       title = "Deposit size vs detection rate") +
  theme_bw()
```

The two axes trade off against each other. The drug-probe cohorts sit top-left
(small but almost fully exposed); the large unselected cohorts sit bottom-right
(many subjects, almost none on the drug). A large deposit with a low rate can
still be worth having, but only if the absolute number of exposed subjects is
large enough to matter, which is the `n_both` point size rather than the position.

## Ratio distribution per dataset

```{r exp-log2ratio-violin, fig.width = 8, fig.height = 5}
both_only <- samples |>
  filter(detection_tier == "both", is.finite(log2_ratio)) |>
  mutate(ds = short_ds(dataset_id))

ggplot(both_only, aes(x = ds, y = log2_ratio)) +
  geom_violin(fill = "#1b9e77", alpha = 0.25, scale = "width") +
  geom_jitter(width = 0.18, height = 0, alpha = 0.45, size = 1.3) +
  geom_hline(yintercept = 0, linetype = 3, colour = "grey50") +
  coord_flip() +
  labs(x = NULL, y = "log2(parent / metabolite)",
       title = "Per-dataset log2(par/met) distribution, both-tier samples") +
  theme_bw()
```

Datasets are in alphabetical order rather than sorted by median, deliberately.
Sorting would invite a ranking, and as the next section explains, the vertical
position of a dataset is not interpretable.

## Parent vs metabolite intensity

```{r exp-auc-scatter, fig.width = 7.5, fig.height = 5.5}
auc_df <- samples |>
  filter(detection_tier == "both", par_auc > 0, met_auc > 0) |>
  mutate(ds = short_ds(dataset_id))

ggplot(auc_df, aes(x = par_auc, y = met_auc, colour = ds)) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, colour = "grey60") +
  geom_point(alpha = 0.7, size = 1.4) +
  scale_x_log10() + scale_y_log10() +
  labs(x = "parent peak area (log10)", y = "metabolite peak area (log10)",
       colour = NULL,
       title = "Per-sample parent vs metabolite intensity") +
  theme_bw() + theme(legend.position = "right")
```

The dashed line is `y = x`. Points below it have more parent than metabolite.
Note the datasets separate into visible clusters: that separation is largely
method, not biology, which is exactly the problem the next section describes.

## Demographic composition per dataset

```{r exp-demog-sex, fig.width = 7, fig.height = 4}
if ("sex" %in% names(samples)) {
  samples |>
    filter(detection_tier == "both") |>
    mutate(ds = short_ds(dataset_id),
           sex = ifelse(is.na(sex), "unknown", sex)) |>
    count(ds, sex) |>
    ggplot(aes(x = ds, y = n, fill = sex)) +
    geom_col(position = "fill") + coord_flip() +
    scale_y_continuous(labels = percent) +
    labs(x = NULL, y = "share", fill = "sex",
         title = "Sex composition of both-tier samples") +
    theme_bw() + theme(legend.position = "bottom")
}
```

```{r exp-demog-health, fig.width = 8, fig.height = 4.5}
if ("health_status" %in% names(samples)) {
  samples |>
    filter(detection_tier == "both") |>
    mutate(ds = short_ds(dataset_id),
           health_status = ifelse(is.na(health_status), "unknown", health_status)) |>
    count(ds, health_status) |>
    ggplot(aes(x = ds, y = n, fill = health_status)) +
    geom_col(position = "fill") + coord_flip() +
    scale_y_continuous(labels = percent) +
    labs(x = NULL, y = "share", fill = "health status",
         title = "Health-status composition of both-tier samples") +
    theme_bw() + theme(legend.position = "bottom")
}
```

## Ratio density overlaid across datasets

```{r exp-ratio-density, fig.width = 7.5, fig.height = 4.5}
ggplot(both_only, aes(x = log2_ratio, colour = ds, fill = ds)) +
  geom_density(alpha = 0.12, linewidth = 0.6) +
  geom_vline(xintercept = 0, linetype = 3, colour = "grey50") +
  labs(x = "log2(parent / metabolite)", y = "density", colour = NULL, fill = NULL,
       title = "Density of log2(par/met) overlaid across datasets") +
  theme_bw() + theme(legend.position = "right")
```

This is the clearest picture of the problem the next section formalises: the
curves are shifted horizontally relative to each other, and that shift is mostly
method rather than metabolism.

## Peak-quality control

```{r exp-fwhm-qc, fig.width = 7.5, fig.height = 4.5}
peak_qc <- samples |>
  filter(detection_tier == "both") |>
  mutate(ds = short_ds(dataset_id)) |>
  select(ds, par_fwhm, met_fwhm) |>
  pivot_longer(c(par_fwhm, met_fwhm), names_to = "role", values_to = "fwhm") |>
  mutate(role = sub("_fwhm", "", role))

ggplot(peak_qc |> filter(is.finite(fwhm)), aes(x = fwhm, y = ds, colour = role)) +
  geom_jitter(width = 0, height = 0.2, alpha = 0.5, size = 1.2) +
  scale_x_continuous(trans = "log10") +
  labs(x = "peak width at half maximum (s, log10)", y = NULL, colour = "role",
       title = "Peak-width distribution per dataset") +
  theme_bw()
```

Peak width is a chromatography property. A dataset whose widths sit far from the
rest, or whose two compounds disagree with each other, is worth a second look in
its per-dataset report before its numbers are trusted.

# What we can and cannot compare across datasets

This is the central constraint of the whole project, so it is worth stating
precisely.

A mass spectrometer counts ions, not molecules. How many ions a compound yields
depends on how readily it ionises, which is a property of the molecule **and** of
the method. Omeprazole and 5-hydroxyomeprazole ionise with different efficiency,
so what we measure is

```
measured log2(parent / metabolite) = true log2(parent / metabolite) + offset
```

where `offset` is the log ratio of the two compounds' response factors. It is
unknown, and it is different in every dataset, because every dataset used a
different instrument and method. The standard fix is to spike an isotope-labelled
internal standard of each compound into every sample. **No deposit in this corpus
did that**, so the offset cannot be calibrated away.

Two consequences follow, and they are the reason this notebook is shaped the way
it is.

**We cannot compare absolute ratios between datasets.** A dataset whose median
sits a unit above another's may simply have used a method that ionises the parent
more efficiently. Any ranking of the deposits by their ratio would be measuring
instruments, not patients. This is why none of the plots above are sorted by
median.

**We can compare a difference measured inside a single dataset.** Within one
dataset every sample went through the same method, so every sample carries the
same `offset`. Take two groups in that dataset and subtract:

```
(true_case + offset) - (true_ref + offset) = true_case - true_ref
```

The offset cancels exactly. That difference, which we will call the
**within-study contrast**, is free of the method term and is therefore the only
quantity in this corpus that means the same thing in every dataset. It underpins
the quantitative analyses in Part B.

# Part A: Expanding the classic reverse-metabolomics prevalence analysis

## What the field does, and how we extend it

The reverse-metabolomics / MASST tools answer one question at repository scale:
in what fraction of files does a molecule appear? Their cross-cohort output is a
**prevalence table or heatmap** of molecule-by-cohort detection rates, built from
MS2 spectral matches, and used strictly for hypothesis generation. They never
quantify intensity across deposits, and their own authors state that a negative
result is not evidence of absence, because a compound present at low abundance may
never be selected for fragmentation.

That last point is exactly the gap MS1 propagation fills. We run the same
prevalence analysis, but replace "did a file get an MS2 match?" with "does the
compound's signal appear at its confirmed mass and retention time?", which is
recorded for every file whether or not the instrument fragmented it. So Part A is
the field's own analysis at higher sensitivity: **per compound, per cohort, the
MS2 detection rate versus the MS1-propagated detection rate.** It rests on
presence, not on a quantity, so it is immune to the response-factor and isobaric
problems that constrain the ratio.

One unit note: prevalence is a **per-file** rate, so this section uses the
per-file table (not the subject-deduped one) and compares it to the per-file
MASST hits, restricted to files a metadata source covers.

```{r prevalence-setup}
#| echo: false
pf <- samples_perfile
pf$has_metadata <- {
  k <- paste0(pf$dataset_id, "|", basename(as.character(pf$file)))
  h <- has_meta_lookup[k]; ifelse(is.na(h), FALSE, unname(h))
}
pf <- pf[pf$has_metadata %in% TRUE, ]
pf$fk <- fkey(pf$file)

# MS2 baseline: metadata-carrying files with a MASST hit of each role. Restrict
# the hits to the same file universe (dataset_id | normalised key) so MS2 and MS1
# rates share a denominator.
valid_dk       <- paste0(pf$dataset_id, "|", pf$fk)
masst_hits$fk  <- fkey(masst_hits$file_path)
masst_hits$dk  <- paste0(masst_hits$dataset_id, "|", masst_hits$fk)
mh_valid       <- masst_hits[masst_hits$dk %in% valid_dk, ]

ms2 <- mh_valid |>
  dplyr::distinct(dataset_id, role, fk) |>
  dplyr::count(dataset_id, role, name = "ms2") |>
  tidyr::pivot_wider(names_from = role, values_from = ms2, values_fill = 0)
if (!"parent"     %in% names(ms2)) ms2$parent     <- 0L
if (!"metabolite" %in% names(ms2)) ms2$metabolite <- 0L

prevalence <- pf |>
  dplyr::group_by(dataset_id) |>
  dplyr::summarise(n_files        = dplyr::n(),
                   ms1_parent     = sum(detection_tier %in% c("both", "parent_only")),
                   ms1_metabolite = sum(detection_tier %in% c("both", "metab_only")),
                   .groups = "drop") |>
  dplyr::left_join(ms2, by = "dataset_id") |>
  dplyr::mutate(dplyr::across(c(parent, metabolite),
                              ~ tidyr::replace_na(as.integer(.), 0L)),
                ds = short_ds(dataset_id))

prev_long <- prevalence |>
  dplyr::transmute(ds, n_files,
    `omeprazole (MS2)`          = parent        / n_files,
    `omeprazole (MS1)`          = ms1_parent    / n_files,
    `5-OH-omeprazole (MS2)`     = metabolite    / n_files,
    `5-OH-omeprazole (MS1)`     = ms1_metabolite / n_files) |>
  tidyr::pivot_longer(-c(ds, n_files), names_to = "panel", values_to = "rate") |>
  tidyr::separate(panel, into = c("compound", "method"), sep = " \\(") |>
  dplyr::mutate(method = sub("\\)", "", method))
```

## The prevalence heatmap: MS2 versus MS1, per compound, per cohort

This is the figure that shows what the pipeline buys. Each row is a cohort, each
column a (compound, detection-method) pair; colour is the fraction of that
cohort's files in which the compound is detected. The MS1 columns should be at
least as dark as their MS2 counterparts, and much darker wherever the instrument
under-fragmented the compound.

```{r prevalence-heatmap, fig.width = 8, fig.height = 5.5}
prev_long |>
  dplyr::mutate(col = paste0(compound, "\n", method),
                col = factor(col, levels = c(
                  "omeprazole\nMS2", "omeprazole\nMS1",
                  "5-OH-omeprazole\nMS2", "5-OH-omeprazole\nMS1"))) |>
  ggplot(aes(x = col, y = reorder(ds, rate, sum), fill = rate)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = scales::percent(rate, accuracy = 1)), size = 2.7) +
  scale_fill_gradient(low = "grey95", high = "#1b6837",
                      labels = scales::percent, name = "detection\nrate") +
  labs(x = NULL, y = NULL,
       title = "Detection prevalence per cohort: MASST (MS2) vs MS1 propagation") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(lineheight = 0.9))
```

## The propagation gain per cohort

The same numbers as a gain: MS1 detections divided by MS2 detections, per compound
per cohort. This is the honest, cohort-resolved version of the propagation-gain
headline (a single corpus-wide multiplier hides that the gain is largest exactly
where MS2 coverage is thinnest).

A cohort whose MS2 baseline is zero inside the analysed files has **no defined
gain**: a ratio needs a denominator. Printing `x / max(0, 1)` would invent one, so
those cells read `n/a` and the counts alone carry the meaning. MSV000082433 is the
case in point: its entire MS2 evidence is a single file (`957.mzML`) that Pan-ReDU
never described, so inside the annotated files the baseline is 0 while propagation
recovers 17 metabolite detections. That is "1 unannotated seed -> 17 annotated
files", not a 16-fold gain.

```{r prevalence-gain}
#| echo: false
gain_str <- function(ms2, ms1) {
  ifelse(ms2 > 0,
         sprintf("%d -> %d  (%.1fx)", ms2, ms1, ms1 / pmax(ms2, 1)),
         sprintf("%d -> %d  (n/a: no MS2 in analysed files)", ms2, ms1))
}
prevalence |>
  dplyr::transmute(cohort = ds, n_files,
    `omeprazole: MS2 -> MS1` = gain_str(parent, ms1_parent),
    `5-OH: MS2 -> MS1`       = gain_str(metabolite, ms1_metabolite)) |>
  dplyr::arrange(dplyr::desc(n_files))
```

## One study, two accessions: MSV000082493 and MSV000084008

These two deposits are **not independent cohorts**. They are the same study,
submitted twice:

- The same **14 subjects**, and on the 333 files present in both, `subject_id`
  agrees for **333 of 333**.
- **333 files are shared between them** - all `blood_plasma`. Where both yield a
  parent detection, the AUC correlation is **1.000** and the median relative
  difference is **0**; the largest single difference is 11%, so most are the same
  raw file and a few differ slightly (most likely re-conversion, not different
  samples). Both deposits are RP.
- 082493 additionally holds the multi-matrix collection (390 skin, 148 fecal,
  139 urine, 8 plasma); 084008 holds 106 plasma files 082493 does not. Neither is a
  subset of the other, so distinct plasma across both is 447.

Nothing in the repository metadata signals this. Different accessions, separate
submissions, no cross-reference. It surfaced only from a duplicate-file audit.

**This inflates any public-data meta-analysis, and it hits an MS2-only analysis
harder than ours**, because the duplicated files are exactly the plasma where MASST
found the parent:

| counted over | summed | distinct | inflation |
|---|---|---|---|
| parent MS2 hit files | 389 | 233 | **+67%** |
| 5-OH MS2 hit files | 494 | 342 | **+44%** |
| all MS2 hit files, corpus | 668 | 472 | **+42%** |

So a classic reverse-metabolomics presence analysis reporting "N public files contain
the drug" would over-report by roughly half on this pair alone. That is a property of
the data ecosystem, not of any particular method, and it is not routinely checked.

### A shared file name does NOT always mean a duplicate

The opposite case exists in this same corpus and is just as dangerous.
**MSV000092754 and MSV000092757 share all 164 file names** (`pos_QC_1.mzXML` and so
on) and the same 79 subjects, but 092754 is **HILIC** and 092757 is **RP**: the same
samples on two columns, submitted as separate accessions. Their extracted values
differ by up to **117x**. De-duplicating those by file name would merge a HILIC run
into an RP run and delete 164 real files.

So file name identity means two opposite things here, and only the analytical
context distinguishes them:

| pair | file names | chromatography | verdict |
|---|---|---|---|
| 082493 / 084008 | 333 shared | RP / RP | same files - **collapse** |
| 092754 / 092757 | 164 shared | HILIC / RP | different runs - **keep apart** |

**What we therefore do**, and it must be stated wherever these numbers appear:

1. **Identify the duplicate pair from subject identity**, not file names: the same 14
   subjects, with subject, timepoint, study day and body site agreeing on 461 of 461
   shared files. File names cannot make this call - here they indicate a duplicate,
   in 092754/092757 they indicate a collision.
2. **Collapse that pair, and only that pair, on file name** (`DUP_STUDY` below).
   Every other dataset stays namespaced by its own accession.
3. **Treat 082493 + 084008 as one study.** Per-cohort tables keep both accessions for
   provenance, but they must never be read as two independent replications - and a
   result "confirmed" in both is confirmed once.
4. **Re-run the collision audit when datasets are added**
   (`scratchpad/file_audit2.R`): it compares extracted AUCs for every shared key
   across every dataset pair, which is the only reliable way to tell a duplicate from
   a name collision.

One thing this does **not** address, and which is worth stating rather than hiding:
each deposit also contains **technical replicates** - 237 rows in each of 082493 and
084008 repeat a `(subject, timepoint, day, body_site)` combination. Those are
multiple injections of one specimen, and they are counted individually in every
file-level number here, including the gain. A specimen-level count would be more
defensible, but only 8 of 12 datasets carry subject IDs, so it cannot be done
corpus-wide without mixing units.

Subject-level analyses were never affected: the project already takes one measure per
subject, and the subject IDs agree across the two accessions.

```{r dedup-corpus}
#| echo: false
fk_all <- fkey(samples_perfile$file)
dk_all <- paste0(samples_perfile$dataset_id, "|", fk_all)
cat("files summed across datasets :", length(dk_all), "\n")
cat("distinct physical files      :", length(unique(fk_all)), "\n")
cat("counted more than once       :", length(dk_all) - length(unique(fk_all)), "\n")
```

## The corpus-wide gain, with and without the metadata restriction

The per-cohort table above counts only files a metadata source covers, because
every stratified analysis downstream needs that metadata. But the pipeline
extracts from more files than that: a file with no metadata still yields a valid
presence call, it just cannot be assigned to a subject or a group. Both numbers
are worth stating, because they answer different questions. The unrestricted row
is what the extraction actually recovers; the restricted row is what survives
into an analysis.

```{r prevalence-gain-total}
#| echo: false
# The ONE deposit pair known to be the same study. Identified from subject identity,
# not from file names: the same 14 subjects, and on every shared file the subject,
# timepoint, study day and body site agree 461/461. Confirmed independently by the
# extracted values (AUC correlation 1.000, median relative difference 0).
#
# Scoping matters. Collapsing on file name GLOBALLY would be wrong, because
# MSV000092754 and MSV000092757 share all 164 file names while being HILIC and RP
# runs of the same 79 subjects - different measurements whose values differ by up to
# 117x. Only this pair collapses; every other dataset keeps its own files even when
# names collide.
DUP_STUDY <- c("MSV000082493", "MSV000084008")

# Rows from the duplicate pair share one namespace; everything else is namespaced by
# its own dataset_id, so identical names across unrelated deposits never merge.
dup_key <- function(dataset_id, fk)
  ifelse(dataset_id %in% DUP_STUDY, paste0("DUPSTUDY|", fk),
         paste0(dataset_id, "|", fk))

prop_totals <- function(d, label, dedup_files = TRUE) {
  d$fk <- fkey(d$file)
  dk   <- unique(paste0(d$dataset_id, "|", d$fk))
  mh   <- masst_hits[masst_hits$dk %in% dk, ]
  if (dedup_files) {
    # Count DISTINCT physical files per role. dup_key must NOT be deduplicated
    # across roles first: a file with both a parent and a metabolite hit would
    # collapse to one row and silently drop a role (this deflated the metabolite
    # MS2 baseline 526 -> 146 and inflated its gain to 8x). Dedup within each role.
    d  <- d[!duplicated(dup_key(d$dataset_id, d$fk)), ]
    nd <- function(role) {
      r <- mh[mh$role == role, ]
      length(unique(dup_key(r$dataset_id, r$fk)))
    }
  } else {
    nd <- function(role) nrow(dplyr::distinct(mh[mh$role == role, c("dataset_id", "fk")]))
  }
  p2 <- nd("parent"); m2 <- nd("metabolite")
  p1 <- sum(d$detection_tier %in% c("both", "parent_only"))
  m1 <- sum(d$detection_tier %in% c("both", "metab_only"))
  fmt <- function(ms2, ms1) if (ms2 > 0)
    sprintf("%d -> %d  (%.1fx)", ms2, ms1, ms1 / ms2) else
    sprintf("%d -> %d  (n/a)", ms2, ms1)
  data.frame(
    scope                    = label,
    files                    = nrow(d),
    `omeprazole: MS2 -> MS1` = fmt(p2, p1),
    `5-OH: MS2 -> MS1`       = fmt(m2, m1),
    check.names = FALSE, stringsAsFactors = FALSE)
}
rbind(
  prop_totals(samples_perfile, "all extracted, DISTINCT files",   dedup_files = TRUE),
  prop_totals(pf,              "metadata-carrying, DISTINCT",     dedup_files = TRUE),
  prop_totals(samples_perfile, "all extracted, summed (INFLATED)", dedup_files = FALSE))
```

The MS2 baseline here is **every MASST hit in `hits_curated`**, not just the
subset we sent to SIRIUS. That matters: the confirmation step was capped per role
per dataset, so dividing by the confirmed count would measure our own cap rather
than the method, and inflates the gain several-fold. The uncapped baseline is the
conservative choice and the one a "classic" MS2-only reanalysis would have had.

## What the spread across cohorts tells us

The spread of detection rate across cohorts is not noise; it is the biology of who
is exposed. Two regimes are visible:

- **Drug-probe cohorts** (every subject dosed by design) sit near 100% for both
  compounds. They are the internal positive control: propagation should, and does,
  recover almost everything there.
- **General-population and disease cohorts** sit low, because only incidentally
  dosed subjects carry the drug. Here the MS1-over-MS2 gain is largest, and the
  absolute number of recovered exposures is the payoff: a few percent of a large
  cohort is still tens of subjects the MS2-only analysis would have missed.

The metabolite frequently has a higher detection rate than the parent (e.g. in
excretory matrices), which is expected pharmacology, not an artefact: omeprazole
is extensively metabolised, so in urine or feces the metabolite persists while the
parent is largely gone. This asymmetry is itself a hypothesis-generating readout
and is exactly the kind of pattern the presence analysis surfaces.

## Prevalence by biofluid: the reverse-metabolomics biology view

The heatmap above is organised by cohort to show the method's value (MS2 vs MS1).
The reverse-metabolomics papers instead organise prevalence by a **biology axis**
(tissue / biofluid / health phenotype); Gentry et al. (Nature 2024, Fig 2b-d) is
the canonical example, a molecule-by-tissue and molecule-by-health-condition
prevalence heatmap. We reproduce that view here, MS1-propagated: pooling files
across cohorts by body site, the detection prevalence of each compound.

This pooling deliberately crosses cohorts, so exposure and body site are partly
confounded (a biofluid seen only in a drug-probe cohort will read high for reasons
of study design, not physiology). It is therefore hypothesis-generating, in
exactly the spirit the field uses it: the expected signal is that the parent
concentrates in blood/plasma while the metabolite persists in excretory matrices
(urine, feces), which if it appears is the pharmacology falling out of public data
with no targeted assay.

```{r prevalence-by-biofluid, fig.width = 7, fig.height = 5}
if ("body_site" %in% names(pf) && any(!is.na(pf$body_site))) {
  bysite <- pf |>
    dplyr::filter(!is.na(body_site), nzchar(body_site)) |>
    dplyr::group_by(body_site) |>
    dplyr::summarise(n_files = dplyr::n(),
                     omeprazole        = mean(detection_tier %in% c("both", "parent_only")),
                     `5-OH-omeprazole` = mean(detection_tier %in% c("both", "metab_only")),
                     .groups = "drop") |>
    dplyr::filter(n_files >= 5) |>
    tidyr::pivot_longer(c(omeprazole, `5-OH-omeprazole`),
                        names_to = "compound", values_to = "rate")

  ggplot(bysite, aes(x = compound,
                     y = reorder(paste0(body_site, "  (n=", n_files, ")"), rate, sum),
                     fill = rate)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = scales::percent(rate, accuracy = 1)), size = 3) +
    scale_fill_gradient(low = "grey95", high = "#762a83",
                        labels = scales::percent, name = "MS1 detection\nrate") +
    labs(x = NULL, y = NULL,
         title = "MS1-propagated detection prevalence by biofluid",
         subtitle = "pooled across cohorts; hypothesis-generating (exposure and body site are confounded)") +
    theme_minimal() +
    theme(panel.grid = element_blank())
} else {
  message("body_site not populated in the per-file table; biofluid view skipped.")
}
```

## Why the ratio is not part of this section

Prevalence is a per-compound presence question, and presence rests on the parent,
which has no isobaric twin. The parent/metabolite **ratio** is a different kind of
object: a quantity, comparable across cohorts only if the response-factor offset
were known (it is not), and vulnerable to the omeprazole-sulfone isomer discussed
in `03`. Pooling ratios across cohorts as if they were a prevalence would import
both problems. So the ratio does not enter Part A. It is legitimate in exactly two
places: as a **within-cohort spread** (descriptive, the offset is constant within
a cohort, shown in the exploratory section), and as a **validated quantity against
an external standard** (the probe cohort versus the published pharmacokinetic
value, in `05`). Part B treats the within-cohort contrast; nowhere do we pool an
absolute ratio across cohorts.

# Part B: A pipeline-specific meta-analysis

Part A is the field's prevalence analysis, made more sensitive. Part B is the
piece the field does not have: turning the propagated detections into a
quantitative, error-controlled cross-cohort analysis, borrowing directly from how
proteomics handles the identical problem.

## The proteomics precedent: propagation needs a false-transfer rate

Our pipeline is the metabolomics analogue of **match-between-runs (MBR)** in
proteomics: identify by MS2 where a feature was fragmented, then transfer the
identity to other files by accurate mass and retention time and quantify at MS1.
Proteomics has measured what this costs. In a controlled two-species experiment,
uncontrolled MBR transferred 44% of one organism's proteins into samples that
provably did not contain it, concentrated in low-abundance one-hit detections
(Lim, Paulo & Gygi 2019). Independent audits find that the common tools do not
actually control the false-discovery rate of propagated identities unless a decoy
procedure is added (PIP-ECHO, 2025).

The lesson transfers exactly: **a propagated detection is a hypothesis until its
false-transfer rate is estimated.** Part A's prevalence gains are only as
trustworthy as our specificity, and specificity has to be measured, not assumed.
The design proteomics uses is a decoy: run the identical pipeline against a target
that cannot be real, and whatever it "detects" estimates the false-positive rate.

```{r decoy-plan}
#| eval: false
# Decoy specificity (to compute in a dedicated extraction step, not inline here):
#   1. Decoy m/z  : extract at a mass with no expected compound (offset the real
#                   m/z beyond the isolation window), same RT window.
#   2. Decoy RT   : extract at the real m/z but a retention time where the compound
#                   does not elute (e.g. confirmed RT + 120 s).
# The detection rate returned by each decoy is an empirical false-transfer rate.
# Report every Part A prevalence alongside it; if the decoys light up, the
# detection rule (shape + robust SNR >= 3) is too permissive and the gains deflate.
# Pre-dose (minute-0) samples in the probe cohort are a second, biological decoy
# (known negatives) and are used in `05`.
```

Until that decoy pass is run, the Part A gains should be read as **upper bounds**
on the true recovery, and the internal positive control (drug-probe cohorts near
100%, in Part A) is the evidence that the rule is not grossly permissive.

## The within-study contrast (the quantitative strand)

Where a cohort supports it, the within-study contrast (defined above: the offset
cancels in a within-cohort difference) is the quantitative unit that *could* be
pooled. We compute it per cohort here; the pooling itself is deferred for the
reasons in the final section.

## Defining the contrast

For each dataset we restrict to samples where both compounds were detected, pick
the grouping variable the dataset can actually support, collapse it to two levels
("case" vs "reference"), and fit a linear model of `log2_ratio` on that grouping.
The coefficient is the contrast.

```{r within-study-fn}
fit_within_study <- function(df) {

  df_both <- df %>% filter(detection_tier == "both", is.finite(log2_ratio))
  if (nrow(df_both) < 6) {
    return(tibble(contrast = NA_character_, estimate = NA_real_,
                  se = NA_real_, n = nrow(df_both), note = "n<6"))
  }

  # Try stratifiers in priority order; first usable one wins. This ordering is
  # the weak point of the whole section: see the final discussion.
  candidates <- c("health_status", "group", "sex", "country", "body_site")
  picked <- NULL
  for (v in candidates) {
    if (!v %in% names(df_both)) next
    counts <- df_both %>% filter(!is.na(.data[[v]])) %>% count(.data[[v]])
    if (nrow(counts) >= 2 && all(counts$n >= 3)) { picked <- v; break }
  }
  if (is.null(picked)) {
    return(tibble(contrast = NA_character_, estimate = NA_real_,
                  se = NA_real_, n = nrow(df_both), note = "no_stratifier"))
  }

  # Collapse to two levels: the smallest group becomes "case", so the estimate
  # reads as "case minus reference".
  df_both <- df_both %>%
    mutate(
      .v    = .data[[picked]],
      .case = .v == names(sort(table(.v)))[1],
      .grp  = factor(ifelse(.case, "case", "ref"), levels = c("ref", "case"))
    )

  m <- lm(log2_ratio ~ .grp, data = df_both)
  coefs <- summary(m)$coefficients
  tibble(
    contrast = paste0(picked, ": ", names(sort(table(df_both$.v)))[1], " vs rest"),
    estimate = coefs[".grpcase", "Estimate"],
    se       = coefs[".grpcase", "Std. Error"],
    n        = nrow(df_both),
    note     = NA_character_
  )
}
```

## The effect-size table

```{r per-study-effects}
effects <- samples %>%
  group_by(dataset_id) %>%
  group_modify(~ fit_within_study(.x)) %>%
  ungroup() %>%
  left_join(datasets %>% select(dataset_id, repository, chromatography,
                                const_country, const_body_site),
            by = "dataset_id") %>%
  arrange(dataset_id)

effects
write_csv(effects, "artifacts/meta_effects.csv")
```

### What each column means

An **effect size** is just a number that says how big a difference is, expressed
so that it can be compared with the same number from another study. Many fields
use a standardised effect size (Cohen's d, Hedges' g), which divides the
difference by its spread so that studies measuring different things on different
scales can be put side by side. **We deliberately do not standardise here.** Our
readout is already on a common and directly interpretable scale: `log2_ratio` is
in doublings, and the response-factor offset cancels in the subtraction. So the
raw difference in log2 units *is* the effect size, and it means the same thing in
every dataset. Standardising would divide by a within-dataset spread that varies
for reasons (cohort heterogeneity, sample size) that have nothing to do with the
biology, and would make the studies less comparable rather than more.

| column | meaning |
|---|---|
| `dataset_id` | the (deposit, assay) entity |
| `contrast` | which grouping variable was used, and which level was treated as "case". `sex: female vs rest` and `health_status: severe vs rest` are **not the same question**, which matters a great deal below |
| `estimate` | the contrast itself: mean `log2(par/met)` in the case group minus the reference group. Units are log2, so `+1` means the parent-to-metabolite ratio is twice as high in the case group; `0` means no difference; negative means the case group has relatively more metabolite (faster apparent metabolism) |
| `se` | standard error of that estimate: how precisely we know it. Small `se` means a tight estimate. A rough 95% interval is `estimate ± 1.96 × se`. Any pooling would weight studies by this |
| `n` | number of both-tier samples the fit used. With `n` in single digits, `se` is large and the estimate is weak |
| `note` | why a dataset has no estimate: `n<6` (too few usable samples) or `no_stratifier` (no grouping variable with two levels of at least three samples) |
| `repository`, `chromatography`, `const_country`, `const_body_site` | dataset-level attributes, available as moderators for a future pooled model |

## Per-study contrast, ordered

```{r exp-effect-bars, fig.width = 7.5, fig.height = 4.5}
fit_eff <- effects |> filter(!is.na(estimate), !is.na(se))

eff_plot <- fit_eff |>
  mutate(ci_lo = estimate - 1.96 * se,
         ci_hi = estimate + 1.96 * se,
         ds    = fct_reorder(short_ds(dataset_id), estimate))

ggplot(eff_plot, aes(x = estimate, y = ds)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi), height = 0.25, colour = "grey40") +
  geom_point(aes(colour = chromatography), size = 3) +
  geom_text(aes(label = paste0("n=", n)), hjust = -0.25, size = 3, colour = "grey30") +
  labs(x = "contrast: case minus reference, log2(parent / metabolite)",
       y = NULL, colour = "chrom.",
       title = "Per-study contrast with 95% confidence interval") +
  theme_bw()
```

Here the intervals are wide, most of them cross zero, and, critically, the
contrasts on the y-axis are **not all asking the same question**. Read this as a
description of what each dataset happens to offer, not as a set of comparable
results.

# Chagas HILIC vs RP: one direct check

MSV000092754 (HILIC) and MSV000092757 (reverse phase) are paired deposits from
the same Chagas serum study: the same subjects and blood draws, run on two
chromatographies, matched by `codigo_num`. It is the only place in the corpus
where we can watch the same biology measured two ways, which makes it the one
direct check on the offset argument above.

```{r chagas-pair}
#| echo: false
HILIC_DS <- "MSV000092754"
RP_DS    <- "MSV000092757"

per_subject_ratio <- function(ds) {
  samples %>%
    filter(dataset_id == ds, detection_tier == "both",
           !par_fwhm_outlier %in% TRUE, !met_fwhm_outlier %in% TRUE,
           is.finite(log2_ratio)) %>%
    group_by(subject_id) %>%
    summarise(log2_ratio    = median(log2_ratio, na.rm = TRUE),
              health_status = dplyr::first(health_status),
              .groups       = "drop")
}

chagas <- dplyr::inner_join(
  per_subject_ratio(HILIC_DS) %>% dplyr::rename(log2_ratio_HILIC = log2_ratio),
  per_subject_ratio(RP_DS)    %>% dplyr::select(subject_id, log2_ratio_RP = log2_ratio),
  by = "subject_id")

cat("Subjects matched across both chromatographies:", nrow(chagas), "\n")
```

```{r chagas-paired, fig.width = 6, fig.height = 4.5}
if (nrow(chagas) >= 3) {
  chagas |>
    pivot_longer(c(log2_ratio_HILIC, log2_ratio_RP),
                 names_to = "chromatography", values_to = "log2_ratio") |>
    mutate(chromatography = sub("log2_ratio_", "", chromatography)) |>
    ggplot(aes(x = chromatography, y = log2_ratio,
               group = subject_id, colour = health_status)) +
    geom_line(alpha = 0.5) +
    geom_point(size = 2.5) +
    labs(x = NULL, y = "log2(parent / metabolite)", colour = "health status",
         title = "Chagas: per-subject paired HILIC vs RP ratio",
         subtitle = "lines connect the same subject across chromatographies") +
    theme_bw()
}
```

**What this can and cannot tell us.** Only a handful of subjects survive the
requirement of a usable ratio in *both* chromatographies, so this is a look, not
a test. Correlation coefficients, Bland-Altman limits of agreement and
heteroscedasticity slopes on this many points would report precision we do not
have, so we do not compute them.

What the picture is good for is direction. If the lines were roughly parallel,
each subject shifting by a similar amount between the two methods, that is the
constant-additive-offset behaviour the previous section assumes, and it would
support the claim that the offset cancels in a within-study contrast. If instead
the lines cross each other freely, the readout depends on the chromatography in a
way a single constant cannot absorb, and every cross-method claim in the project
needs a louder caveat. Judge it by eye, and treat it as one encouraging or
discouraging data point rather than as a validation.

# What a meta-analysis could add, and what it needs first

The obvious next step is to pool the per-study contrasts into a single
cross-cohort estimate: a random-effects meta-analysis, weighting each dataset by
its precision, reporting a pooled contrast with a confidence interval plus τ² and
I² to quantify how much the cohorts genuinely disagree. That is the standard tool
and it is the right shape for this data. **We are not running it yet, on
purpose.**

**Why not.** A meta-analysis pools estimates that answer the same question. Look
again at the `contrast` column: because `fit_within_study()` walks a priority list
and takes the first usable grouping, most datasets end up contributing a **sex**
contrast, simply because sex is the one variable that is populated nearly
everywhere. Pooling "severe versus healthy" from one study together with "male
versus female" from another does not estimate anything. The earlier version of
this notebook did exactly that and produced a tidy-looking pooled value with a
narrow interval and I² near zero, which is worse than no answer, because it looks
like a result. The blocker is not the statistics; it is that we do not yet have a
shared, biologically meaningful contrast across enough datasets.

**What would make it work**, roughly in order of value:

- **Pick one question and only include datasets that can answer it.** For example
  health status or disease severity, dropping every dataset without it. Fewer
  studies, but a pooled number that means something.
- **Handle the censored samples.** `parent_only` and `metab_only` are ratio
  observations with one bound known. An interval-censored (Tobit-style) model uses
  them as bounds instead of discarding them, which matters because they are
  exactly the most extreme ratios, and dropping them shrinks any effect toward
  zero.
- **Decide how to treat the method.** Chromatography, instrument and biofluid all
  reset the offset. With a real pooled model these become moderators or subgroups.
  Note the corpus currently has at least two deposits whose recorded
  chromatography is doubtful given their retention times, so this needs the
  metadata fixed before it can be trusted.
- **Consider the one-stage alternative.** Rather than fitting each study then
  pooling, a hierarchical (mixed-effects) model on the sample-level data with a
  random effect per dataset uses the individual samples directly, which is
  generally better behaved when per-study `n` is small, as it is here.
- **Pool a robust quantity, never raw abundance.** This is settled practice, and
  it converges across fields. Cross-repository proteomics meta-analyses of
  independent public deposits (e.g. the 12-dataset colorectal-cancer study,
  PMC10833860) uniformly re-process the raw files and then pool a
  distribution-free quantity: rank-binned abundances there, Stouffer/Fisher
  p-value combination in MetaMSD, within-study log fold-changes in the
  metabolomics tool MetaHD. If we ever pool a ratio across cohorts, it must be
  rank-transformed or reduced to a per-cohort statistic first, for the same
  response-factor reason that forbids pooling the raw ratio. That these
  cross-repository quantitative meta-analyses exist and work is also the point:
  our contribution is not the pooling, which is standard, but using MS1
  propagation to *produce* the per-cohort quantity that the reverse-metabolomics /
  MASST world does not generate.
- **Do not treat every accession as an independent cohort.** `01` records the
  known exceptions in `artifacts/related_deposits.csv`. Today it contains one:
  **MSV000084008 is a re-deposit of MSV000082493's blood arm**, and both come from
  the same 14 subjects (PMC7485946). They contribute different biofluids here, so
  both are kept, but pooling them as two studies would count those subjects twice
  and hand the pair a double vote in any precision-weighted estimate. A deposit
  accession is not a study, and any pooling has to be keyed on the study.
- **Be honest about country.** Each country is represented by a single dataset,
  so country and dataset are perfectly confounded. Any "country effect" is that
  one deposit's method. At best this is a calibration check against published
  CYP2C19 frequencies, never an estimate.

The prerequisite for all of it is a decision about which question the pool is
supposed to answer. That is a discussion to have as a group, not something to
settle by picking whichever variable happens to be populated.

# References

- Viechtbauer, W. (2010). Conducting Meta-Analyses in R with the metafor
  package. *Journal of Statistical Software*, 36(3).
- Blanz, J. et al. (2017). Evaluation of relative MS response factors of drug
  metabolites for semi-quantitative assessment of chemical liabilities.
  *Journal of Mass Spectrometry*, 52(4).
- Viallon, V. et al. (2021). A new pipeline for the normalization and pooling
  of metabolomics data. *Metabolites*, 11(9).
- Belasimo, A. et al. (2024). MetaHD: a multivariate meta-analysis model for
  metabolomics data. *Bioinformatics*, 40(7).
- Petrović, J. et al. (2020). Meta-analysis of probability estimates of
  worldwide variation of CYP2D6 and CYP2C19. *Translational Psychiatry*, 10.
- Tornio, A., Backman, J. T. (2018). Cytochrome P450 in pharmacogenetics: an
  update. *British Journal of Clinical Pharmacology*, 122 (Suppl. 5).
- Abouir, K. et al. (2024). Improving CYP2C19 phenotyping using
  stereoselective omeprazole and 5-hydroxy-omeprazole metabolic ratios.
  *Basic & Clinical Pharmacology & Toxicology*.
