---
title: "05: CYP2C19 ground-truth validation"
subtitle: "Does the mined omeprazole / 5-OH ratio track genotyped CYP2C19 status?"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
execute:
  warning: false
  message: false
---

# Overview

MSV000082493 and MSV000084008 (Jarmusch et al., *Clin Transl Sci* 2020,
PMC7485946) dosed healthy volunteers with omeprazole as a CYP2C19 probe and carry
the **genotyped** CYP2C19 metaboliser phenotype per subject in
`artifacts/rescue_samples.csv`. The authors reported the identical OME / 5-OH-OME
ratio but never tested it against that genotype, so this is a new use of the
data, and a ground-truth check that the whole cascade
(MASST → SIRIUS → MS1 EIC → ratio) measures real pharmacology.

Expected: low CYP2C19 activity (Poor) → less 5-OH → **high** log2(parent/metab);
Ultrarapid → low. Run after `03` has cached the two datasets.

```{r setup}
suppressPackageStartupMessages({ library(dplyr); library(tidyr); library(ggplot2) })
# norm_file_key() for the AUC section's metrics <-> submitter-TSV join. Use the
# project's own normaliser, never a bare basename: the metrics carry cache-prefixed
# names while the TSV carries bare ones, and that mismatch has silently produced
# empty joins several times in this project.
source("R/metadata.R")
```

# Per-file table + genotype join

`sample_table_all.csv` is deduped to one file per subject, which collapses the
timepoint × specimen structure this readout needs. We rebuild a per-file table
from the cached per-dataset metrics (before dedup), replicating
`build_sample_table()`'s detection rule (shape pass AND robust-SNR ≥ 3), and join
the genotype by filename stem (metrics carry `.mzXML`/prefixed cache names, the
genotype table `.mzXML`, identical stems).

```{r per-file-join}
cyp_ds <- c("MSV000082493", "MSV000084008")
rs <- read.csv("artifacts/rescue_samples.csv", check.names = FALSE, stringsAsFactors = FALSE)

# CYP2C19 activity order (low -> high); the ratio should DECREASE along it
PHENO <- c("Poor Metabolizer", "Intermediate Metabolizer", "Normal Metabolizer",
           "Rapid Metabolizer", "Ultrarapid Metabolizer")
rs$pheno_rank <- match(rs$cyp2c19_phenotype, PHENO)

bare <- function(x)  # strip MSV000XXXXXX_ cache prefix + extension, lowercase
  sub("^msv[0-9]{9}_", "", tolower(tools::file_path_sans_ext(basename(x))))
rs$stem <- bare(rs$file)

per_file_tier <- function(d) {
  r <- readRDS(sprintf("artifacts/per_dataset/%s.rds", d))
  m <- merge(r$metrics_par, r$metrics_met, by = "file", all = TRUE)
  # Detection is decided in the picker (detected_new + drop-list); metrics_* carry it as a flag.
  par_det <- m$par_detected %in% TRUE
  met_det <- m$met_detected %in% TRUE
  m$detection_tier <- ifelse( par_det &  met_det, "both",
                       ifelse( par_det & !met_det, "parent_only",
                       ifelse(!par_det &  met_det, "metab_only", "neither")))
  m$log2_ratio <- ifelse(m$detection_tier == "both",
                         log2(m$par_auc / m$met_auc), NA_real_)
  m$dataset_id <- d
  m$stem <- bare(m$file)
  m
}
pf <- bind_rows(lapply(cyp_ds, per_file_tier))

j <- pf |>
  inner_join(rs[, c("stem", "deposit_id", "cyp2c19_phenotype", "cyp2c19_genotype",
                    "pheno_rank", "subject_id", "timepoint", "specimen")],
             by = "stem")
both <- j |> filter(detection_tier == "both", is.finite(log2_ratio),
                    !par_fwhm_outlier %in% TRUE, !met_fwhm_outlier %in% TRUE)
cat("joined files:", nrow(j), " | with both compounds + clean ratio:", nrow(both), "\n")
table(j$dataset_id, j$detection_tier)
```

# Ratio vs genotyped phenotype

Per cohort: median ratio by phenotype, file-level Spearman/Kruskal-Wallis, and the
valid **subject-level** Spearman (one ratio per subject, no pseudoreplication).

```{r phenotype-tests}
test_cohort <- function(d, label) {
  x <- both |> filter(if (is.null(d)) TRUE else deposit_id == d, !is.na(pheno_rank))
  if (nrow(x) < 8 || dplyr::n_distinct(x$cyp2c19_phenotype) < 3) {
    cat("\n[", label, "] too few points (n=", nrow(x), ")\n"); return(invisible())
  }
  cat("\n----------", label, " (n =", nrow(x), "files) ----------\n")
  ph <- x |> group_by(cyp2c19_phenotype) |>
    summarise(n = n(), med = round(median(log2_ratio), 2), .groups = "drop") |>
    arrange(match(cyp2c19_phenotype, PHENO))
  print(as.data.frame(ph))
  sp <- suppressWarnings(cor.test(x$log2_ratio, x$pheno_rank, method = "spearman"))
  kw <- kruskal.test(log2_ratio ~ factor(cyp2c19_phenotype), x)
  cat(sprintf("Spearman(ratio, activity_rank) rho = %.3f  p = %.3g  (expect NEGATIVE)\n",
              sp$estimate, sp$p.value))
  cat(sprintf("Kruskal-Wallis across phenotypes p = %.3g\n", kw$p.value))
  subj <- x |> group_by(subject_id, cyp2c19_phenotype, pheno_rank) |>
    summarise(ratio = median(log2_ratio), .groups = "drop")
  if (nrow(subj) >= 8 && dplyr::n_distinct(subj$cyp2c19_phenotype) >= 3) {
    sp2 <- suppressWarnings(cor.test(subj$ratio, subj$pheno_rank, method = "spearman"))
    cat(sprintf("  subject-level (n=%d subjects): Spearman rho = %.3f  p = %.3g\n",
                nrow(subj), sp2$estimate, sp2$p.value))
  }
  invisible(x)
}
x84 <- test_cohort("MSV000084008", "084008 PLASMA (classic CYP readout)")
x82 <- test_cohort("MSV000082493", "082493 urine/feces/skin")
invisible(test_cohort(NULL, "POOLED CYP cohorts"))
```

# Per-subject AUC metabolic ratio (the clinical readout)

Everything above uses **one ratio per subject** taken from single files. That is not
how CYP2C19 phenotyping is done clinically: the accepted metric is the **ratio of
areas under the concentration-time curves** over the dosing interval,
AUC(omeprazole) / AUC(5-OH-omeprazole). Integrating removes the dependence on *when*
the sample happened to be drawn, which in a single-timepoint ratio is pure noise:
one subject sampled near their peak and another past it differ for reasons that have
nothing to do with genotype.

MSV000084008 supports this directly: 461 plasma files, 14 subjects, 8-9 timepoints
per subject on each of two study days. The timepoint comes from the submitter's
`ATTRIBUTE_Time_Point_Mins` (recovered by the metadata union in `01`; Pan-ReDU does
not carry it).

Two conventions worth stating because they change the answer:

- **Days are integrated separately.** Day 1 and day 9 straddle the antibiotic
  exposure, so they are different physiological states and pooling them would blur
  the contrast the study was designed around.
- **A non-detected timepoint contributes 0** to the AUC — the standard
  pharmacokinetic below-limit-of-detection convention. "Non-detected" means the
  timepoint fails the picker's detection gate (`detection_tier`), not merely that no
  peak was integrated: a timepoint where only noise was picked is set to 0, not
  allowed to contribute its raw integrated area. Dropping those points instead would
  bias the AUC upward exactly for the poor metabolisers whose metabolite is genuinely
  absent.

The direction is predicted before looking: poor metabolisers clear omeprazole slowly,
so a **lower** activity rank should give a **higher** parent/metabolite ratio, i.e.
Spearman rho < 0.

```{r auc-ratio}
auc_tsv <- "dataset_metadata/MSV000084008_metadata.tsv"
if (!file.exists(auc_tsv)) {
  cat("submitter TSV not present - AUC section skipped\n")
} else {
  r_auc <- readRDS("artifacts/per_dataset/MSV000084008.rds")
  mp <- as.data.frame(r_auc$metrics_par); mm <- as.data.frame(r_auc$metrics_met)
  ta <- read.delim(auc_tsv, check.names = FALSE, stringsAsFactors = FALSE)
  # normalised key, never a bare basename: the metrics carry cache-prefixed names
  ta$k <- norm_file_key(ta$filename)
  mp$k <- norm_file_key(mp$file); mm$k <- norm_file_key(mm$file)
  # Gate to detection: a timepoint that fails the picker's detection gate is below-LOD
  # (NA here -> 0 in trap()), not its raw integrated noise-area. Detected points keep their AUC.
  mp_i <- match(ta$k, mp$k); mm_i <- match(ta$k, mm$k)
  ta$par_auc <- ifelse(mp$par_detected[mp_i] %in% TRUE, mp$par_auc[mp_i], NA_real_)
  ta$met_auc <- ifelse(mm$met_detected[mm_i] %in% TRUE, mm$met_auc[mm_i], NA_real_)
  ta$tmin <- suppressWarnings(as.numeric(ta$ATTRIBUTE_Time_Point_Mins))
  ta <- ta[ta$MS_SampleType == "sample" & !is.na(ta$tmin), ]

  trap <- function(x, y) {              # trapezoidal AUC; NA -> 0 (below LOD)
    o <- order(x); x <- x[o]; y <- y[o]; y[is.na(y)] <- 0
    if (length(unique(x)) < 4) return(NA_real_)
    sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)
  }
  auc <- do.call(rbind, lapply(
    split(ta, list(ta$ATTRIBUTE_Subject, ta$ATTRIBUTE_Study_Day), drop = TRUE),
    function(d) data.frame(
      subject = d$ATTRIBUTE_Subject[1], day = d$ATTRIBUTE_Study_Day[1],
      n_tp = length(unique(d$tmin)),
      auc_par = trap(d$tmin, d$par_auc), auc_met = trap(d$tmin, d$met_auc),
      phenotype = d$PharmGeno_Phenotype_2C19[1], stringsAsFactors = FALSE)))
  auc$log2_ratio <- log2(auc$auc_par / auc$auc_met)
  auc <- auc[is.finite(auc$log2_ratio), ]
  auc$rank <- match(auc$phenotype, PHENO)

  cat("timepoints per subject-day: median", median(auc$n_tp),
      "range", min(auc$n_tp), "-", max(auc$n_tp), "\n\n")
  print(auc[order(auc$rank, auc$subject),
            c("subject","day","n_tp","log2_ratio","phenotype")],
        row.names = FALSE, digits = 3)

  cat("\nmean log2(AUC parent / AUC metabolite) by phenotype:\n")
  subj <- aggregate(log2_ratio ~ subject + phenotype + rank, data = auc, FUN = mean)
  print(tapply(subj$log2_ratio, factor(subj$phenotype, levels = PHENO), mean))

  cat("\n")
  for (dd in sort(unique(auc$day))) {
    s <- auc[auc$day == dd, ]
    ct <- suppressWarnings(cor.test(s$rank, s$log2_ratio, method = "spearman"))
    cat(sprintf("day %-3s n=%2d | Spearman rho = %+.3f  p = %.4f\n",
                dd, nrow(s), ct$estimate, ct$p.value))
  }
  ct <- suppressWarnings(cor.test(subj$rank, subj$log2_ratio, method = "spearman"))
  cat(sprintf("both days, per subject  n=%2d | Spearman rho = %+.3f  p = %.4f\n",
              nrow(subj), ct$estimate, ct$p.value))
}
```

## Is it one subject doing the work?

Three phenotype classes have a single subject each, so a strong correlation could be
carried entirely by one person. Two checks: drop each subject in turn, and then throw
away all three singleton classes and test only the two populated ones.

```{r auc-robustness}
if (exists("subj") && nrow(subj) > 5) {
  full <- suppressWarnings(cor.test(subj$rank, subj$log2_ratio, method = "spearman"))
  cat(sprintf("full            n=%2d rho=%+.3f p=%.4f\n\n",
              nrow(subj), full$estimate, full$p.value))
  for (i in seq_len(nrow(subj))) {
    x <- subj[-i, ]
    ct <- suppressWarnings(cor.test(x$rank, x$log2_ratio, method = "spearman"))
    cat(sprintf("drop %-7s (%-24s) rho=%+.3f p=%.4f%s\n",
                subj$subject[i], substr(subj$phenotype[i], 1, 24),
                ct$estimate, ct$p.value,
                if (ct$p.value > 0.05) "   <-- loses significance" else ""))
  }
  two <- subj[subj$phenotype %in% c("Intermediate Metabolizer", "Normal Metabolizer"), ]
  if (dplyr::n_distinct(two$phenotype) == 2) {
    w <- suppressWarnings(wilcox.test(log2_ratio ~ phenotype, data = two))
    cat(sprintf("\nsingletons removed - Intermediate (n=%d) vs Normal (n=%d): p = %.3f\n",
                sum(two$phenotype == "Intermediate Metabolizer"),
                sum(two$phenotype == "Normal Metabolizer"), w$p.value))
  }
}
```

What to take from this: the AUC and the single-timepoint ratio use **the same
subjects, the same files and the same extraction**. Only the summary statistic
differs. If the AUC separates phenotypes where the single-timepoint ratio did not,
the earlier null was a measurement-noise problem, not evidence that the readout
fails - and the honest conclusion is about the metric, not about a bigger cohort.

Limits that remain: n = 14 subjects; three classes have one subject each; the
Intermediate-vs-Normal comparison alone is marginal. MSV000082493 holds plasma files
for these **same** 14 subjects, so checking there is a batch-level consistency check,
not independent replication.

# Genotype dosage (star alleles)

CYP2C19 allele activity: *1 = 1 (normal), *2/*3 = 0 (no function), *17 = 1.5
(increased). The activity score respects functional spacing the equal-spaced rank
ignores; the loss-of-function dose isolates the strong *2/*3 effect from the noisy
*17. Subject level.

```{r genotype-dosage}
allele_act <- function(a) ifelse(a %in% c("2", "3"), 0,
                          ifelse(a == "17", 1.5, ifelse(a == "1", 1, NA_real_)))
lof_allele <- function(a) as.integer(a %in% c("2", "3"))
add_geno <- function(df) {
  parts <- strsplit(as.character(df$cyp2c19_genotype), "_")
  df$act_score <- vapply(parts, function(p) sum(allele_act(p)), numeric(1))
  df$lof_dose  <- vapply(parts, function(p) sum(lof_allele(p)), integer(1))
  df
}
gp <- add_geno(both |> filter(deposit_id == "MSV000084008", !is.na(cyp2c19_genotype)))
gsub <- gp |> group_by(subject_id, cyp2c19_genotype, act_score, lof_dose) |>
  summarise(ratio = median(log2_ratio), n_files = n(), .groups = "drop") |>
  arrange(act_score)
if (nrow(gsub) >= 8) {
  print(as.data.frame(gsub))
  s1 <- suppressWarnings(cor.test(gsub$ratio, gsub$act_score, method = "spearman"))
  cat(sprintf("Spearman(subject ratio, activity_score) rho = %+.3f  p = %.3g  (expect NEG)\n",
              s1$estimate, s1$p.value))
  gsub$carrier <- gsub$lof_dose > 0
  if (dplyr::n_distinct(gsub$carrier) == 2) {
    w <- suppressWarnings(wilcox.test(ratio ~ carrier, gsub))
    cat(sprintf("LoF carriers *2/*3 (n=%d) median = %.2f  vs  non-carriers (n=%d) median = %.2f;  Wilcoxon p = %.3g\n",
                sum(gsub$carrier), median(gsub$ratio[gsub$carrier]),
                sum(!gsub$carrier), median(gsub$ratio[!gsub$carrier]), w$p.value))
  }
}
```

# Specimen-resolved phenotyping (082493)

Hepatic CYP2C19 is classically read from plasma: can a non-invasive matrix
(urine / skin / feces) recover the genotype? Subject-level is the honest test.

```{r specimen-resolved}
sp82 <- both |> filter(deposit_id == "MSV000082493", !is.na(pheno_rank))
if (nrow(sp82) > 0) for (sp in sort(unique(sp82$specimen))) {
  z <- sp82 |> filter(specimen == sp)
  if (nrow(z) >= 8 && dplyr::n_distinct(z$cyp2c19_phenotype) >= 3) {
    s_ <- suppressWarnings(cor.test(z$log2_ratio, z$pheno_rank, method = "spearman"))
    zs <- z |> group_by(subject_id, pheno_rank) |>
      summarise(r = median(log2_ratio), .groups = "drop")
    sub_txt <- if (nrow(zs) >= 8 && dplyr::n_distinct(zs$pheno_rank) >= 3) {
      ss <- suppressWarnings(cor.test(zs$r, zs$pheno_rank, method = "spearman"))
      sprintf("subj-level n=%d rho=%+.3f p=%.3g", nrow(zs), ss$estimate, ss$p.value)
    } else sprintf("subj-level n=%d (underpowered)", nrow(zs))
    cat(sprintf("  %-6s file n=%3d rho=%+.3f p=%.3g   |   %s\n",
                sp, nrow(z), s_$estimate, s_$p.value, sub_txt))
  } else cat(sprintf("  %-6s n=%3d  (too few / <3 phenotypes)\n", sp, nrow(z)))
}
```

# Cefprozil timepoint (day 1 vs day 9)

The paper reported cefprozil **decreased** CYP2C19 activity, so we expect a higher
day-9 ratio. A raw day-vs-day pool mixes early (parent-dominated) and late
post-dose timepoints, so we also compare dose-time-matched pairs within subject
(the dose-minute is the 3rd field of the file stem).

```{r timepoint}
parse_dt <- function(file) {
  stem <- sub("^MSV[0-9]+_", "", basename(as.character(file)))
  m <- regmatches(stem, regexec("^(DM[0-9]+)_([0-9]+)_([0-9]+)_", stem))
  data.frame(
    p_subj = vapply(m, function(x) if (length(x) >= 2) x[2] else NA_character_, ""),
    p_day  = vapply(m, function(x) if (length(x) >= 3) x[3] else NA_character_, ""),
    p_dose = vapply(m, function(x) if (length(x) >= 4) suppressWarnings(as.integer(x[4])) else NA_integer_, 0L),
    stringsAsFactors = FALSE)
}
tp <- both |> filter(deposit_id == "MSV000084008", timepoint %in% c("1", "9"))
tp <- cbind(tp, parse_dt(tp$file))
if (nrow(tp) > 8) {
  print(tp |> group_by(timepoint) |>
          summarise(n = n(), median_ratio = round(median(log2_ratio), 2), .groups = "drop"))
  w <- tp |> group_by(subject_id, timepoint) |>
    summarise(r = median(log2_ratio), .groups = "drop") |>
    tidyr::pivot_wider(names_from = timepoint, values_from = r, names_prefix = "d")
  w <- w[stats::complete.cases(w), ]
  if (nrow(w) >= 5) {
    wt <- wilcox.test(w$d1, w$d9, paired = TRUE)
    cat(sprintf("(a) pooled: paired Wilcoxon day1 vs day9 (n=%d subjects) p = %.3g  median dd = %.2f\n",
                nrow(w), wt$p.value, median(w$d9 - w$d1)))
  }
  wm <- tp |> filter(!is.na(p_dose)) |>
    group_by(subject_id, p_dose, timepoint) |>
    summarise(r = median(log2_ratio), .groups = "drop") |>
    tidyr::pivot_wider(names_from = timepoint, values_from = r, names_prefix = "d")
  wm <- wm[stats::complete.cases(wm), ]
  if (nrow(wm) >= 6) {
    wt2 <- wilcox.test(wm$d1, wm$d9, paired = TRUE)
    cat(sprintf("(b) dose-matched: paired Wilcoxon (n=%d subject×dose pairs) p = %.3g  median dd(day9-day1) = %.2f\n",
                nrow(wm), wt2$p.value, median(wm$d9 - wm$d1)))
    cat("    (paper predicts dd > 0: cefprozil lowers CYP2C19 -> higher day-9 ratio)\n")
  } else cat("(b) dose-matched: too few paired subject×dose points (n=", nrow(wm), ")\n")
}
```

# MS1 propagation gain

Files SIRIUS confirmed from DDA (seeds) vs files where the MS1 EIC recovered the
compound: the extra genotype-linked measurements the targeted MS2 data missed.

```{r propagation-gain}
for (d in cyp_ds) {
  r    <- readRDS(sprintf("artifacts/per_dataset/%s.rds", d))
  seed <- length(unique(basename(c(r$cfg$confirmed_files_parent,
                                   r$cfg$confirmed_files_metabolite))))
  detf  <- sum(pf$dataset_id == d & pf$detection_tier != "neither", na.rm = TRUE)
  bothf <- sum(pf$dataset_id == d & pf$detection_tier == "both",    na.rm = TRUE)
  cat(sprintf("  %s: MS2 seeds=%d -> MS1 detections=%d (both=%d)  gain x%.1f\n",
              d, seed, detf, bothf, ifelse(seed > 0, detf / seed, NA)))
}
```

# Figures

Boxplot of the ratio by genotyped phenotype (both cohorts), plus the two
deck figures: the plasma trend line and the propagation lollipop. Also writes
`artifacts/cyp_validation_data.csv`.

```{r figure-data}
if (!dir.exists("figures")) dir.create("figures")
PH_SHORT <- c("Poor", "Inter.", "Normal", "Rapid", "Ultra")
pal <- c("Poor Metabolizer" = "#B2182B", "Intermediate Metabolizer" = "#EF8A62",
         "Normal Metabolizer" = "#D9D9D9", "Rapid Metabolizer" = "#67A9CF",
         "Ultrarapid Metabolizer" = "#2166AC")
plot_df <- both |>
  filter(deposit_id %in% cyp_ds, !is.na(pheno_rank)) |>
  mutate(cohort = ifelse(deposit_id == "MSV000084008",
                         "Plasma  (MSV000084008)", "Urine / feces / skin  (MSV000082493)"),
         phenotype = factor(cyp2c19_phenotype, levels = PHENO))
write.csv(plot_df, "artifacts/cyp_validation_data.csv", row.names = FALSE)
```

```{r fig-by-phenotype, fig.width=9, fig.height=4}
ggplot(plot_df, aes(phenotype, log2_ratio, fill = phenotype)) +
  geom_boxplot(outlier.shape = NA, width = 0.62, alpha = 0.95) +
  geom_jitter(width = 0.15, height = 0, size = 0.8, alpha = 0.45) +
  facet_wrap(~cohort, scales = "free_y") +
  scale_fill_manual(values = pal, guide = "none") +
  scale_x_discrete(labels = PH_SHORT) +
  labs(x = "Genotyped CYP2C19 metaboliser class  (low → high activity)",
       y = "log2( omeprazole / 5-OH-omeprazole )",
       title = "Ratio vs. genotyped CYP2C19 class: directional (Poor→Rapid), underpowered",
       subtitle = "Pipeline blinded to genotype; expect a downward trend. n=14 subjects/cohort.") +
  theme_bw(base_size = 11) +
  theme(strip.background = element_rect(fill = "grey92"),
        plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r fig-plasma-trend, fig.width=8.4, fig.height=4.6}
pl <- plot_df |> filter(deposit_id == "MSV000084008")
meds <- pl |> group_by(phenotype) |>
  summarise(med = median(log2_ratio), .groups = "drop") |>
  mutate(x = as.integer(phenotype))
g1 <- ggplot(pl, aes(phenotype, log2_ratio)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey70") +
  geom_boxplot(aes(fill = phenotype), outlier.shape = NA, width = 0.6, alpha = 0.95) +
  geom_jitter(width = 0.13, height = 0, size = 0.7, alpha = 0.35) +
  geom_line(data = meds, aes(x = x, y = med), inherit.aes = FALSE,
            colour = "#111111", linewidth = 0.9) +
  geom_point(data = meds, aes(x = x, y = med), inherit.aes = FALSE,
             colour = "#111111", size = 2.4) +
  scale_fill_manual(values = pal, guide = "none") +
  scale_x_discrete(labels = PH_SHORT) +
  labs(x = "Genotyped CYP2C19 metaboliser class  (low → high activity)",
       y = "log2( omeprazole / 5-OH-omeprazole )",
       title = "Plasma ratio moves with CYP2C19 genotype, in the expected direction",
       subtitle = "MSV000084008. Median falls monotonically Poor→Rapid; Ultrarapid (*17) is the known exception. Directional, not powered (14 subjects).") +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(size = 9.5, colour = "grey30"),
        panel.grid.minor = element_blank())
ggsave("figures/cyp_plasma_trend.png", g1, width = 8.4, height = 4.6, dpi = 200)
g1
```

```{r fig-propagation, fig.width=8.4, fig.height=4.6}
det_logic <- function(m) {              # detection from the picker's flag (see per_file_tier)
  pd <- m$par_detected %in% TRUE
  md <- m$met_detected %in% TRUE
  ifelse(pd&md,"both",ifelse(pd&!md,"parent_only",ifelse(!pd&md,"metab_only","neither")))
}
rows <- lapply(list.files("artifacts/per_dataset", pattern = "[.]rds$", full.names = TRUE), function(f) {
  r <- readRDS(f); m <- merge(r$metrics_par, r$metrics_met, by = "file", all = TRUE)
  data.frame(ds = sub("_.*", "", r$dataset_id),
             seeds = length(unique(basename(c(r$cfg$confirmed_files_parent, r$cfg$confirmed_files_metabolite)))),
             both = sum(det_logic(m) == "both"))
})
p <- bind_rows(rows) |> filter(seeds > 0) |> mutate(gain = both/seeds) |> arrange(both)
p$ds <- factor(p$ds, levels = p$ds)
tot_seed <- sum(p$seeds); tot_both <- sum(p$both)
g2 <- ggplot(p) +
  geom_segment(aes(y = ds, yend = ds, x = seeds, xend = both), colour = "grey65", linewidth = 1) +
  geom_point(aes(y = ds, x = seeds), colour = "#B2182B", size = 2.6) +
  geom_point(aes(y = ds, x = both),  colour = "#2166AC", size = 2.6) +
  geom_text(aes(y = ds, x = both, label = sprintf(" ×%.0f", gain)), hjust = 0, size = 3, colour = "grey25") +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  labs(x = "files  ( ● MS2-confirmed seeds → ● files with a full parent+metabolite ratio )", y = NULL,
       title = "MS1 propagation multiplies the usable readout",
       subtitle = sprintf("Across %d datasets: %d MS2 confirmations → %d files with both compounds (×%.1f).",
                          nrow(p), tot_seed, tot_both, tot_both/tot_seed)) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(size = 9.5, colour = "grey30"),
        panel.grid.major.y = element_blank(), panel.grid.minor = element_blank())
ggsave("figures/propagation_gain.png", g2, width = 8.4, height = 4.6, dpi = 200)
g2
```
