---
title: "05: Cross-dataset exploration"
subtitle: "What each dataset contributes, what can be compared across them, and what a meta-analysis would still need."
author: "Philippine Louail"
date: today
format:
  html:
    df-print: paged
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(forcats)
library(scales)
library(ggrepel)
source("R/_setup.R")
# metadata.R holds norm_file_key()/to_bare_name(), the canonical filename rule.
# Sourced here, at the top, so nothing below improvises its own key.
source("R/metadata.R")
# _setup.R attaches Bioconductor packages (S4Vectors via
# Spectra/MetaboAnnotation) whose verbs `first`, `rename`, `filter`, etc. shadow
# the dplyr ones. Re-attach dplyr / tidyr so the tidyverse verbs sit on top of
# the search path again.
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
})
```

# What this notebook does

`03` measured omeprazole and 5-hydroxyomeprazole in every biological file of
every confirmed dataset. This notebook asks what the corpus as a whole supports,
in two explicitly separated meta-analyses plus the context needed to read them.

- **Context first** (kept from the exploratory pass): what each dataset is, how
  much signal it contributes, and the one hard constraint that shapes
  everything, that an absolute ratio is not comparable across datasets while a
  within-study contrast is.
- **Part A: expanding the classic reverse-metabolomics prevalence analysis.**
  The field reports molecule-by-cohort *detection rates* from MS2 matches. We
  run the same analysis with MS1 propagation and compare MS2 vs MS1 detection
  per compound per cohort. This is presence-based, so it is isomer-immune, and
  it is where the pipeline's value shows most cleanly.
- **Part B: the quantitative strand, and where it stops.** The layer the field
  does not have, built on the proteomics match-between-runs precedent:
  known-negative and known-positive controls on the propagated detections, then
  the within-study contrast as the poolable quantity. Its honest result is
  negative — no ratio contrast survives FDR in any deposit — and the variance
  decomposition says why: the cross-deposit offset is body site far more than
  study, so the ratio is comparable within a matrix and not across them.

**What it deliberately does not do: pool an absolute ratio into a single
cross-cohort estimate.** Most datasets share only a `sex` contrast, so such a
pool would average unlike things. Pooling is a real option but needs a shared,
biologically meaningful contrast the corpus does not yet supply, so it is
described rather than run (final section).

# Load the inputs

```{r load-inputs}
samples  <- read_csv("artifacts/sample_table_all.csv",  show_col_types = FALSE)
datasets <- read_csv("artifacts/datasets-confirmed.csv", show_col_types = FALSE)

# Per-file table (not deduped to subjects) + the raw MASST hits: the prevalence
# meta-analysis in Part A is a per-file rate, so it must compare like with like
# (per-file MS1 detection vs per-file MS2/MASST detection).
samples_perfile <- read_csv("artifacts/sample_table_perfile.csv", show_col_types = FALSE)
masst_hits      <- read.csv("artifacts/hits_curated.csv", check.names = FALSE,
                            stringsAsFactors = FALSE)

# has_metadata flag from 02 (metadata resolved in 01): restrict every analysis
# to files a metadata source covers; record-only files stay out of stratified
# work.
bio_meta_per_ds <- readRDS("artifacts/bio_meta_per_ds.rds")
has_meta_lookup <- unlist(lapply(names(bio_meta_per_ds), function(ds) {
  v <- bio_meta_per_ds[[ds]]
  stats::setNames(as.logical(v), paste0(ds, "|", basename(names(v))))
}))
# Normalise a file name for matching across sources: drop accession prefix,
# extension and case. Valid only where a name carries at most one prefix, which
# the sample tables and hits_curated do. Confirmation output does not: its
# source_file can carry an assay index and a doubled accession, and needs
# norm_file_key(x, bio_files) instead.
fkey <- function(x) tolower(sub("[.][^.]*$", "",
          sub("^(MSV[0-9]+|MTBLS[0-9]+|ST[0-9]+)[_-]", "", basename(as.character(x)))))
```

`sample_table_all.csv` carries one row per subject (`03` de-duplicates files to
subjects where subject metadata exists; where it does not, a row is a file).
Throughout this notebook "sample" means one row of that table, except in Part A
which is explicitly per-file.

```{r short-labels}
# The MetaboLights and Workbench ids carry an assay suffix that makes every axis
# unreadable. Shorten for display only; dataset_id stays the join key.
short_ds <- function(x) sub("_AN[0-9]+$", "", sub("_LC-MS_.*$", "", x))

# Sex is deposited as F/M by three deposits and female/male by the rest, so it
# is harmonised once here rather than per analysis.
harmonise_sex <- function(x) {
  s <- tolower(trimws(as.character(x)))
  dplyr::case_when(startsWith(s, "f") ~ "female",
                   startsWith(s, "m") ~ "male",
                   TRUE ~ NA_character_)
}

# Healthy vs not is an explicit whitelist, never a grepl on "healthy|control":
# that pattern also matches QC injections and inverts the Chagas arms. A label
# in neither list stays NA and drops out of any healthy/not-healthy contrast.
HEALTHY_LV <- c("uninfected", "healthy", "control", "healthy control")
NOTHEAL_LV <- c("t. cruzi infected", "critical covid-19 recovered in icu",
                "non-critical covid-19", "non-covid-19 in icu",
                "non-covid-19 with symptoms", "covid-19")
# Can this deposit's rows be collapsed to people? Two things have to hold, and
# testing only one of them is how this went wrong twice. First, coverage: an id
# on essentially every file can be collapsed on, and the odd unmatched file
# (MSV000091520 has one) becomes its own unit downstream without disturbing
# anything. Second, where coverage is partial, the id has to be corroborated --
# MSV000092754 and 092757 label half their files and no id repeats inside either
# deposit, but all 79 appear in BOTH, which only a person id does. The tempting
# shortcut, "a real subject id repeats", fails on every cross-sectional study in
# the corpus: MSV000091520 gives one id per file and records a distinct sex and
# an age computed to the day behind each of them. It would discard six of the
# nine real designs.
subject_resolved <- function(sid, sibling_ids = NULL) {
  ok <- !is.na(sid) & nzchar(trimws(as.character(sid)))
  if (!any(ok)) return(FALSE)
  if (mean(ok) > 0.9) return(TRUE)
  u <- unique(trimws(as.character(sid[ok])))
  length(u) < sum(ok) ||
    (!is.null(sibling_ids) && length(intersect(u, sibling_ids)) >= 0.5 * length(u))
}

# Ids held by the OTHER deposit of a known same-study pair, keyed by dataset_id.
# Built from related_deposits.csv so the pairing is declared in `01` rather than
# hard-coded here.
sibling_id_map <- local({
  rel <- read.csv("artifacts/related_deposits.csv", stringsAsFactors = FALSE)
  ids <- function(d) {
    v <- samples_perfile$subject_id[samples_perfile$dataset_id == d]
    unique(trimws(as.character(v[!is.na(v) & nzchar(trimws(as.character(v)))])))
  }
  out <- list()
  for (i in seq_len(nrow(rel))) {
    a <- rel$dataset_id[i]; b <- rel$related_to[i]
    out[[a]] <- ids(b); out[[b]] <- ids(a)
  }
  out
})

harmonise_health <- function(x) {
  cl <- tolower(trimws(as.character(x)))
  dplyr::case_when(cl %in% HEALTHY_LV ~ "healthy",
                   cl %in% NOTHEAL_LV ~ "not healthy",
                   TRUE ~ NA_character_)
}
```

# The datasets

## Metadata is the binding constraint

The measurement side of this project is not what limits it. Of 128 deposits
carrying a hit, 12 survive to be measured, and of those only a handful can
answer a question about a person. Every step of that attrition after the
chemistry is a metadata step, and the figure below is the argument for treating
metadata as the bottleneck rather than a preliminary.

```{r fig-metadata-bottleneck, fig.width = 11, fig.height = 3.9}
#| echo: false
.f <- function(v) !is.na(v) & trimws(as.character(v)) != "" & v != "missing value"
.hit <- read.csv("StructureMASST_output/fasstrecords_hits.csv", check.names = FALSE)
.cf  <- read.csv("artifacts/hits-confirmed-all.csv", stringsAsFactors = FALSE)
.sp  <- samples_perfile
.sp$repo <- ifelse(startsWith(.sp$dataset_id, "MSV"), "MassIVE",
            ifelse(startsWith(.sp$dataset_id, "MTBLS"), "MetaboLights", "Workbench"))

# Panel A: attrition. The last step is the one that is not about chemistry.
.n_hit  <- dplyr::n_distinct(.hit$ATTRIBUTE_DatasetAccession)
.n_conf <- dplyr::n_distinct(.cf$dataset_id)
.n_subj <- sum(vapply(split(.sp, .sp$dataset_id), function(z)
  subject_resolved(z$subject_id,
    sibling_ids = unlist(sibling_id_map[unique(z$dataset_id)], use.names = FALSE)),
  logical(1)))
.fun <- data.frame(
  step = factor(c("deposits with a hit", "both compounds", "curated entities",
                  "SIRIUS confirmed", "usable subject id"),
                levels = rev(c("deposits with a hit", "both compounds", "curated entities",
                               "SIRIUS confirmed", "usable subject id"))),
  n = c(.n_hit, 44, 15, .n_conf, .n_subj),
  why = c("", "one compound only", "metadata, DIA or in vitro",
          "isobar, or one role only", "no subject recorded"))
.pA <- ggplot(.fun, aes(n, step)) +
  geom_col(width = 0.6, fill = "grey30") +
  geom_text(aes(label = n), hjust = -0.25, size = 3.6, fontface = "bold") +
  geom_text(aes(x = 0, label = why), hjust = -0.03, vjust = 3.1, size = 2.9, colour = "grey35") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = NULL, y = NULL, title = "A   Where the corpus is lost",
       subtitle = "only the second step is about chemistry") +
  theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank(), axis.text.x = element_blank(),
        plot.title = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(size = 8.5, colour = "grey35"))

# Panel B: coverage by repository. Species lives in the search index, not the
# per-file metadata, so it is a share of hits where the others are a share of
# files.
.cov <- .sp |> dplyr::group_by(repo) |> dplyr::summarise(
  `subject id` = mean(.f(subject_id)), sex = mean(.f(sex)),
  age = mean(is.finite(as_num(age))),
  clinical = mean(.f(health_status)), `body site` = mean(.f(body_site)), .groups = "drop")
.hit$repo <- ifelse(startsWith(.hit$ATTRIBUTE_DatasetAccession, "MSV"), "MassIVE",
             ifelse(startsWith(.hit$ATTRIBUTE_DatasetAccession, "MTBLS"), "MetaboLights",
             ifelse(startsWith(.hit$ATTRIBUTE_DatasetAccession, "ST"), "Workbench", NA)))
.spc <- .hit[.hit$ATTRIBUTE_DatasetAccession %in% unique(.cf$deposit_id), ] |>
  dplyr::group_by(repo) |> dplyr::summarise(species = mean(.f(NCBITaxonomy)), .groups = "drop")
.cov <- dplyr::left_join(.cov, .spc, by = "repo") |>
  tidyr::pivot_longer(-repo, names_to = "field", values_to = "cov")
.cov$field <- factor(.cov$field,
  levels = c("subject id", "species", "sex", "age", "clinical", "body site"))
.pB <- ggplot(.cov, aes(field, repo, fill = cov)) +
  geom_tile(colour = "white", linewidth = 1.1) +
  geom_text(aes(label = sprintf("%.0f%%", 100 * cov), colour = cov > 0.5),
            size = 3.2, show.legend = FALSE) +
  scale_fill_gradient(low = "grey93", high = "grey25", guide = "none") +
  scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
  labs(x = NULL, y = NULL, title = "B   No repository is adequate on its own, or together",
       subtitle = "share of files carrying the field; species is a share of hits") +
  theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank(), plot.title = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(size = 8.5, colour = "grey35"),
        axis.text.x = element_text(angle = 18, hjust = 1))

.p_meta <- patchwork::wrap_plots(.pA, .pB, ncol = 2)
dir.create("thesis/graphics", showWarnings = FALSE, recursive = TRUE)
ggsave("thesis/graphics/flagship_metadata_bottleneck.png", .p_meta,
       width = 11, height = 3.9, dpi = 200)
.p_meta
```

Panel B is not an argument for pooling the repositories. Each is strong where
another is empty --- MassIVE records species on a tenth of its hits and sex on
almost all of them, MetaboLights the reverse --- but combining them does not
produce a complete picture, because the gaps are not the same gaps and nothing
fills them. What it shows is that the corpus a researcher can assemble is
bounded by deposition practice, and that no amount of care at the analysis end
recovers a field nobody wrote down.


```{r fig-metadata-inventory, fig.width = 11, fig.height = 5.6}
#| echo: false
# Three states per cell, because "recorded" and "usable" differ: a field present
# on every file but holding one value describes the cohort and supports no
# contrast.
.filled <- function(v) !is.na(v) & nzchar(as.character(v))
.inv <- samples_perfile
.inv$ds <- short_ds(.inv$dataset_id)
.inv$clinical <- ifelse(.filled(.inv$health_status), .inv$health_status, .inv$group)
.flds <- c(subject = "subject_id", sex = "sex", age = "age", `body site` = "body_site",
           `clinical group` = "clinical", timepoint = "timepoint",
           treatment = "treatment", country = "country")
.gt <- if (file.exists("artifacts/rescue_samples.csv")) {
  .rs <- read.csv("artifacts/rescue_samples.csv", stringsAsFactors = FALSE)
  if ("cyp2c19_genotype" %in% names(.rs))
    unique(short_ds(.rs$deposit_id[.filled(.rs$cyp2c19_genotype)])) else character(0)
} else character(0)

.long <- dplyr::bind_rows(lapply(names(.flds), function(nm) {
  .inv |> dplyr::group_by(ds) |>
    dplyr::summarise(field = nm, coverage = mean(.filled(.data[[.flds[[nm]]]])),
                     levels = dplyr::n_distinct(.data[[.flds[[nm]]]][.filled(.data[[.flds[[nm]]]])]),
                     .groups = "drop") }))
.long <- dplyr::bind_rows(.long, .inv |> dplyr::group_by(ds) |>
  dplyr::summarise(field = "CYP2C19 genotype",
                   coverage = as.numeric(dplyr::first(ds) %in% .gt),
                   levels = ifelse(dplyr::first(ds) %in% .gt, 5, 0), .groups = "drop"))
# Partial coverage is a legend state, not a footnote symbol: an asterisk needs a
# caption to mean anything, and this figure has to stand on its own.
.long$state <- dplyr::case_when(
  .long$levels == 0 | .long$coverage < 0.4 ~ "not recorded",
  .long$levels == 1                        ~ "recorded, one value only",
  .long$coverage < 0.8                     ~ "varies, but on under 80% of files",
  TRUE                                     ~ "varies, on all files")
.long$state <- factor(.long$state, levels = c(
  "not recorded", "recorded, one value only",
  "varies, but on under 80% of files", "varies, on all files"))
.long$lab <- with(.long, dplyr::case_when(
  state == "not recorded" ~ "", levels == 1 ~ "1",
  levels > 20 ~ "20+", TRUE ~ as.character(levels)))
.sz <- .inv |> dplyr::group_by(ds) |> dplyr::summarise(files = dplyr::n(), .groups = "drop")
.long <- dplyr::left_join(.long, .sz, by = "ds")
.long$row <- paste0(.long$ds, "   (", .long$files, " files)")
.long$field <- factor(.long$field, levels = c(names(.flds), "CYP2C19 genotype"))

# Written to thesis/graphics so the chapter figure IS this object. The earlier
# PNGs came from scratch scripts that were later deleted, leaving the chapter
# unable to regenerate.
.p_inv <- ggplot(.long, aes(field, reorder(row, -rank(ds)), fill = state)) +
  geom_tile(colour = "white", linewidth = 0.7) +
  geom_text(aes(label = lab, colour = state %in% c("varies, but on under 80% of files",
                                                   "varies, on all files")),
            size = 3.1, fontface = "bold", show.legend = FALSE) +
  scale_fill_manual(values = c("not recorded" = "grey95",
                               "recorded, one value only" = "grey80",
                               "varies, but on under 80% of files" = "grey52",
                               "varies, on all files" = "grey20"), name = NULL) +
  scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey15")) +
  labs(x = NULL, y = NULL, title = "Metadata fields recorded per deposit, and the number of distinct values in each") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1),
        plot.title = element_text(face = "bold", size = 13))
dir.create("thesis/graphics", showWarnings = FALSE, recursive = TRUE)
ggsave("thesis/graphics/flagship_metadata_inventory.png", .p_inv,
       width = 11, height = 5.6, dpi = 200)
.p_inv
```

Before any analysis, the cast. Each row is one confirmed (deposit, assay)
entity: where it came from, what it is, how much usable signal it contributes,
and which grouping variables it offers for a contrast.

A dataset can only support a contrast on a variable with at least two levels
carrying at least three samples each, the same rule `fit_within_study()` applies
below.

One trap is worth naming. A rule written for categorical fields silently
discards continuous ones: a check that lists named categorical candidates
reports `age` as absent everywhere, although it is continuous in five deposits
and the most widely available stratifier in the corpus. Numeric and categorical
fields are therefore judged separately below.

```{r dataset-summary}
#| echo: false
usable_contrasts <- function(df) {
  cand <- c("health_status", "group", "sex", "country", "body_site", "age", "treatment")
  lab <- vapply(cand, function(v) {
    if (!v %in% names(df)) return(NA_character_)
    x <- df[[v]][!is.na(df[[v]])]
    if (!length(x)) return(NA_character_)
    # a predicate, not a coercion: as.numeric() on a categorical field answers the
  # question by raising a warning, which is noise we would then have to hide
  num <- all(grepl("^ *-?[0-9]+[.]?[0-9]*([eE][-+]?[0-9]+)? *$", as.character(x)))
    if (num && dplyr::n_distinct(x) > 12)
      return(sprintf("%s (%.0f-%.0f)", v, min(as.numeric(x)), max(as.numeric(x))))
    tb <- table(x)
    if (length(tb) >= 2 && sum(tb >= 3) >= 2)
      sprintf("%s (%d)", v, length(tb)) else NA_character_
  }, character(1))
  lab <- lab[!is.na(lab)]
  if (!length(lab)) "none" else paste(lab, collapse = ", ")
}

ds_table <- samples |>
  group_by(dataset_id) |>
  summarise(
    n_samples = n(),
    n_both    = sum(detection_tier == "both", na.rm = TRUE),
    both_rate = n_both / n_samples,
    contrasts = usable_contrasts(pick(everything())),
    .groups   = "drop") |>
  left_join(datasets |> select(dataset_id, repository, chromatography,
                               const_country, const_body_site),
            by = "dataset_id") |>
  mutate(dataset = short_ds(dataset_id),
         both_rate = percent(both_rate, accuracy = 0.1)) |>
  select(dataset, repository, body_site = const_body_site,
         country = const_country, chromatography,
         n_samples, n_both, both_rate, contrasts) |>
  arrange(desc(n_both))

ds_table
```

## What each deposit can and cannot be asked

The table above says what divides a deposit. This one adds what does *not*: a
field present on every file but holding a single value describes the cohort and
supports no contrast, and reporting it as "recorded" would overstate what the
deposit can answer.

```{r metadata-inventory-table}
#| echo: false
FIELDS <- c("sex", "age", "group", "health_status", "body_site", "country",
            "treatment", "subject_id")
describe <- function(x) {
  x <- x[!is.na(x) & nzchar(trimws(as.character(x)))]
  if (!length(x)) return(list(kind = "absent", lab = NA_character_))
  n <- dplyr::n_distinct(x)
  # a predicate, not a coercion: as.numeric() on a categorical field answers the
  # question by raising a warning, which is noise we would then have to hide
  num <- all(grepl("^ *-?[0-9]+[.]?[0-9]*([eE][-+]?[0-9]+)? *$", as.character(x)))
  if (n == 1) return(list(kind = "constant", lab = as.character(x[1])))
  if (num && n > 12)
    return(list(kind = "divides",
                lab = sprintf("%s (%.0f-%.0f)", "continuous", min(as.numeric(x)),
                              max(as.numeric(x)))))
  list(kind = "divides", lab = sprintf("%d levels", n))
}
inv <- do.call(rbind, lapply(split(samples, samples$dataset_id), function(d) {
  r <- lapply(FIELDS, function(f) if (f %in% names(d)) describe(d[[f]]) else
                                  list(kind = "absent", lab = NA_character_))
  names(r) <- FIELDS
  data.frame(
    dataset  = short_ds(d$dataset_id[1]),
    divides  = paste(vapply(FIELDS, function(f)
                 if (r[[f]]$kind == "divides") sprintf("%s: %s", f, r[[f]]$lab) else "",
                 character(1))[vapply(r, function(z) z$kind, "") == "divides"],
                 collapse = ", "),
    constant = paste(vapply(FIELDS, function(f)
                 if (r[[f]]$kind == "constant") sprintf("%s=%s", f, r[[f]]$lab) else "",
                 character(1))[vapply(r, function(z) z$kind, "") == "constant"],
                 collapse = ", "),
    stringsAsFactors = FALSE)
}))
inv$divides[inv$divides == ""] <- "nothing"
inv$constant[inv$constant == ""] <- "-"
knitr::kable(inv, row.names = FALSE,
             caption = "What divides each deposit, and what merely describes it")
```

Two readings follow from this that matter downstream. Age is the most widely
available stratifier in the corpus, not sex or clinical group, which is the
opposite of what the earlier categorical-only check suggested. And a deposit can
be fully annotated and still answer nothing: MSV000091520 carries sex, age,
group, health status, body site and country on all 2,144 of its files, but every
one of those except age holds a single value, and the compound is detected in 2
files.

Read this table as the corpus' honest capacity. Three things it makes obvious:

- **`n_both` is what the ratio analysis actually runs on.** Everything else is a
  sample where at least one of the two compounds was not detected, so no ratio
  can be formed. Across the whole corpus this number is small, and it is
  concentrated in a few deposits.
- **The deposits are not the same kind of thing.** A couple are drug-probe
  cohorts where every subject was dosed by design. Most are general-population
  or disease cohorts where a handful of people happened to be on omeprazole.
  Those two categories cannot be read the same way.
- **`contrasts` shows what each dataset can be asked.** Note how few offer a
  biologically meaningful grouping. That single column is the reason the last
  section of this notebook exists.

# Exploratory look at the corpus

This section is descriptive on purpose. Before modelling anything, it is worth
seeing what the data look like.

## Detection-tier composition per dataset

Counts on the left, the same thing proportionally on the right. The left panel
shows where the signal actually is; the right panel shows what fraction of each
cohort was exposed.

```{r exp-tier-bars}
#| fig-width: 6
#| fig-height: 4.5
#| layout-ncol: 2
tier_levels  <- c("both", "parent_only", "metab_only", "neither")
tier_palette <- c(both        = "#1b9e77",
                  parent_only = "#7570b3",
                  metab_only  = "#d95f02",
                  neither     = "grey75")

tier_long <- samples |>
  mutate(detection_tier = factor(detection_tier, levels = tier_levels),
         ds = short_ds(dataset_id)) |>
  count(ds, detection_tier) |>
  group_by(ds) |>
  mutate(p = n / sum(n)) |>
  ungroup()

ggplot(tier_long,
       aes(x = fct_reorder(ds, p * (detection_tier == "both"),
                           .fun = sum, .desc = FALSE),
           y = n, fill = detection_tier)) +
  geom_col() + coord_flip() +
  scale_fill_manual(values = tier_palette) +
  labs(x = NULL, y = "samples", fill = "tier",
       title = "Detection tier, counts") +
  theme_bw() + theme(legend.position = "bottom")

ggplot(tier_long,
       aes(x = fct_reorder(ds, p * (detection_tier == "both"),
                           .fun = sum, .desc = FALSE),
           y = p, fill = detection_tier)) +
  geom_col() + coord_flip() +
  scale_fill_manual(values = tier_palette) +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = "share of samples", fill = "tier",
       title = "Detection tier, proportional") +
  theme_bw() + theme(legend.position = "bottom")
```

The `parent_only` and `metab_only` slices are censored ratio observations, not
failures: we know the ratio is very high or very low, but not by how much,
because one peak fell below detection. Discarding them biases any ratio estimate
toward the middle, which is a point the final section returns to.

## Deposit size vs detection rate

```{r exp-n-vs-both, fig.width = 6.5, fig.height = 5}
ds_summary <- samples |>
  group_by(dataset_id) |>
  summarise(n_samples = n(),
            n_both    = sum(detection_tier == "both"),
            both_rate = n_both / n_samples,
            .groups   = "drop") |>
  left_join(datasets |> select(dataset_id, chromatography,
                               const_country, const_body_site),
            by = "dataset_id")

ggplot(ds_summary, aes(x = n_samples, y = both_rate)) +
  geom_point(aes(size = n_both, colour = chromatography), alpha = 0.85) +
  geom_text_repel(aes(label = short_ds(dataset_id)), size = 3,
                  min.segment.length = 0) +
  scale_x_log10() + scale_y_continuous(labels = percent) +
  scale_size_continuous(range = c(2, 8)) +
  labs(x = "samples in deposit (log10)", y = "'both' detection rate",
       size = "n with both",
       title = "Deposit size vs detection rate") +
  theme_bw()
```

The two axes trade off against each other. The drug-probe cohorts sit top-left
(small but almost fully exposed); the large unselected cohorts sit bottom-right
(many subjects, almost none on the drug). A large deposit with a low rate can
still be worth having, but only if the absolute number of exposed subjects is
large enough to matter, which is the `n_both` point size rather than the
position.

## Parent vs metabolite intensity

```{r exp-auc-scatter, fig.width = 7.5, fig.height = 5.5}
auc_df <- samples |>
  filter(detection_tier == "both", par_auc > 0, met_auc > 0) |>
  mutate(ds = short_ds(dataset_id))

ggplot(auc_df, aes(x = par_auc, y = met_auc, colour = ds)) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, colour = "grey60") +
  geom_point(alpha = 0.7, size = 1.4) +
  scale_x_log10() + scale_y_log10() +
  labs(x = "parent peak area (log10)", y = "metabolite peak area (log10)",
       colour = NULL,
       title = "Per-sample parent vs metabolite intensity") +
  theme_bw() + theme(legend.position = "right")
```

The dashed line is `y = x`. Points below it have more parent than metabolite.
Note the datasets separate into visible clusters. What the clusters are made of
is the question the next section takes up: deposits differ in instrument,
gradient and specimen at once, and nothing here tells them apart.

## Peak-quality control

```{r exp-fwhm-qc, fig.width = 7.5, fig.height = 4.5}
peak_qc <- samples |>
  filter(detection_tier == "both") |>
  mutate(ds = short_ds(dataset_id)) |>
  select(ds, par_fwhm, met_fwhm) |>
  pivot_longer(c(par_fwhm, met_fwhm), names_to = "role", values_to = "fwhm") |>
  mutate(role = sub("_fwhm", "", role))

ggplot(peak_qc |> filter(is.finite(fwhm)), aes(x = fwhm, y = ds, colour = role)) +
  geom_jitter(width = 0, height = 0.2, alpha = 0.5, size = 1.2) +
  scale_x_continuous(trans = "log10") +
  labs(x = "peak width at half maximum (s, log10)", y = NULL, colour = "role",
       title = "Peak-width distribution per dataset") +
  theme_bw()
```

Peak width is a chromatography property. A dataset whose widths sit far from the
rest, or whose two compounds disagree with each other, is worth a second look in
its per-dataset report before its numbers are trusted.

# What we can and cannot compare across datasets

This is the central constraint of the whole project, so it is worth stating
precisely.

A mass spectrometer counts ions, not molecules. How many ions a compound yields
depends on how readily it ionises, which is a property of the molecule **and**
of the method. Omeprazole and 5-hydroxyomeprazole ionise with different
efficiency, so what we measure is

```
measured log2(parent / metabolite) = true log2(parent / metabolite) + offset
```

where `offset` is the log ratio of the two compounds' response factors. It is
unknown, and it is different in every dataset, because every dataset used a
different instrument and method. The standard fix is to spike an
isotope-labelled internal standard of each compound into every sample. **No
deposit in this corpus did that**, so the offset cannot be calibrated away.

That term is established analytical chemistry rather than an assumption of ours.
Oss et al. measured an electrospray ionisation-efficiency scale for 62 organic
compounds and found it spans roughly six orders of magnitude under identical
conditions ([Anal. Chem. 2010, 82(7),
2865-2872](https://doi.org/10.1021/ac902856t)), and follow-up work from the same
group shows the scale is itself condition- and instrument-dependent ([Rapid
Commun. Mass Spectrom. 2021, 35, e9178](https://doi.org/10.1002/rcm.9178)) --
which is why the offset cannot be assumed constant across deposits. Six orders
of magnitude does not sink the ratio, for two reasons: that spread is across 62
*unrelated* compounds, whereas for one fixed pair the offset is a single unknown
constant, and the ratio is itself the normalisation, cancelling everything the
two compounds share in a sample -- injected amount, dilution, most matrix
suppression -- and leaving only that constant.

Two consequences follow, and they are the reason this notebook is shaped the way
it is.

**We cannot compare absolute ratios between datasets.** A dataset whose median
sits a unit above another's may simply have used a method that ionises the
parent more efficiently. Any ranking of the deposits by their ratio would be
measuring instruments, not patients. This is why none of the plots above are
sorted by median.

The corpus does contain two designs that let this be probed rather than merely
asserted: one deposit pair ran the **same subjects on two chromatographies**,
and the probe study sampled **two matrices from the same subjects**. Part A
returns to them under "The ratio across cohorts". The short version is that the
first is too small to pin the offset down, so the constraint above stands
unchanged, while the second removes the offset by construction and is where the
cross-matrix result comes from.

**We can compare a difference measured inside a single dataset.** Within one
dataset every sample went through the same method, so every sample carries the
same `offset`. Take two groups in that dataset and subtract:

```
(true_case + offset) - (true_ref + offset) = true_case - true_ref
```

The offset cancels exactly. That difference, which we will call the
**within-study contrast**, is free of the method term and is therefore the only
quantity in this corpus that means the same thing in every dataset. It underpins
the quantitative analyses in Part B.

<!-- moved ahead of Part A: this is the control that licenses every
prevalence claim below, so it has to be read first. -->
# Does a propagated detection mean anything? The false-transfer rate

Our pipeline is the metabolomics analogue of **match-between-runs (MBR)** in
proteomics: identify by MS2 where a feature was fragmented, then transfer the
identity to other files by accurate mass and retention time and quantify at MS1.
Proteomics has measured what this costs. In a controlled two-species experiment,
uncontrolled MBR transferred 44% of one organism's proteins into samples that
provably did not contain it, concentrated in low-abundance one-hit detections
(Lim, Paulo & Gygi 2019). Independent audits find that the common tools do not
actually control the false-discovery rate of propagated identities unless a
decoy procedure is added (PIP-ECHO, 2025).

The lesson transfers exactly: **a propagated detection is a hypothesis until its
false-transfer rate is estimated.** Part A's prevalence gains are only as
trustworthy as our specificity, and specificity has to be measured, not assumed.

Proteomics measures it with a decoy that **cannot be true** — a reversed or
shuffled peptide sequence, or an entrapment proteome from a species that is not
in the sample — and it scores a match on fragment evidence, charge and isotope
envelope together. An MS1 extraction has neither property. It asks only whether
an acceptable chromatographic peak sits at one m/z inside one retention-time
window, and in an untargeted file real compounds occupy arbitrary masses. So a
synthetic decoy mass is *not* a hypothesis that cannot be true: when it fires,
the picker has usually found a real but different compound, which is the picker
working correctly rather than a false transfer. We measured this directly —
probing the metabolite's own window at +8 and +12 Da returns peaks that are not
isotopologues of anything we search for, i.e. ordinary matrix. **A synthetic
decoy therefore bounds our false-positive rate from above without estimating
it**, so it is not reported here.

Two controls are used instead, and both rest on files whose true answer is known
rather than assumed:

- **positive control** — every file carrying a SIRIUS-confirmed spectrum, which
  is known to contain the compound. Anything the extraction fails to recover is
  a false negative. This measures sensitivity.
- **negative biological control** — the pre-dose plasma draws of the probe
  study, taken before the drug was given, interrogated at the true m/z and the
  true retention window. Real matrix, every interference a study sample carries,
  absence known by design. This is the closest thing the corpus offers to an
  entrapment experiment, and it measures specificity where it actually matters.

```{r fig-decoy, fig.width = 6.4, fig.height = 3.6}
#| echo: false
# Positive-control counts are recomputed here from the SIRIUS-confirmed anchor
# files rather than transcribed.
{
  # Runs before Part A builds pf, so it works from the per-file table directly,
  # and is not restricted to metadata-carrying files: the positive control
  # concerns the extraction. norm_file_key with the deposit's file list, never
  # the bare fkey() -- confirmation names can carry a doubled accession, and the
  # bare rule lost 39 confirmed files here.
  .BFp <- readRDS("artifacts/bio_files_per_ds.rds")
  .bare <- function(x, ds) {
    r <- character(length(x))
    for (d in unique(ds)) {
      i <- ds == d
      ref <- if (!is.null(.BFp[[d]])) basename(as.character(.BFp[[d]])) else NULL
      r[i] <- norm_file_key(x[i], ref)
    }
    r
  }
  .sp  <- as.data.frame(samples_perfile)
  .pfk <- .bare(.sp$file, .sp$dataset_id)
  .conf <- read.csv("artifacts/hits-confirmed-all.csv", stringsAsFactors = FALSE)
  .conf$fk <- .bare(.conf$source_file, .conf$dataset_id)
  .pc <- do.call(rbind, lapply(c(par = "parent", met = "metabolite"), function(r) {
    keys <- unique(paste0(.conf$dataset_id[.conf$role == r], "|", .conf$fk[.conf$role == r]))
    x <- .sp[paste0(.sp$dataset_id, "|", .pfk) %in% keys, ]
    hit <- if (r == "parent") x$detection_tier %in% c("both", "parent_only") else
                              x$detection_tier %in% c("both", "metab_only")
    data.frame(n = nrow(x), det = sum(hit))
  }))

  # Two controls, both on samples whose true answer is known. Split by role,
  # because the metabolite is harder to recover and a pooled bar could hide a
  # shortfall in it. Labelled positive/negative control rather than
  # sensitivity/specificity: the y axis is a detection rate, so the negative bar
  # reads 0% where specificity would read 100%.
  POS <- "positive control: present"
  NEG <- "negative control: absent"
  .rows <- rbind(
    data.frame(cond = "confirmed\nparent",     n = .pc["par", "n"], det = .pc["par", "det"],
               grp = POS),
    data.frame(cond = "confirmed\nmetabolite", n = .pc["met", "n"], det = .pc["met", "det"],
               grp = POS),
    # Pre-dose plasma, counted from the extracted files rather than written in:
    # the submitter TSV declares 50 draws and 49 were extracted.
    local({
      # Both draws precede absorption, so both are known negatives. Pooling them
      # is what tightens the bound: zero of n only says as much as n allows.
      .pre <- .sp[.sp$dataset_id == "MSV000084008" & .sp$timepoint %in% c("0", "5"), ]
      data.frame(cond = "pre-dose and\n5 min plasma", n = nrow(.pre),
                 det = sum(.pre$detection_tier != "neither"), grp = NEG)
    }))
  .rows$rate <- .rows$det / .rows$n
  # Exact binomial interval. It is what turns "no false positives" into a
  # bounded false-positive rate, and 768 of 771 into a recovery with a floor.
  .rows$lo <- NA_real_; .rows$hi <- NA_real_
  for (i in seq_len(nrow(.rows))) {
    ct <- binom.test(.rows$det[i], .rows$n[i])
    .rows$lo[i] <- ct$conf.int[1]; .rows$hi[i] <- ct$conf.int[2]
  }
  for (i in seq_len(nrow(.rows)))
    cat(sprintf("%-26s %4d/%-4d = %5.1f%%  (95%% CI %.1f%% to %.1f%%)\n",
                gsub("[\r\n]+", " ", .rows$cond[i]), .rows$det[i], .rows$n[i],
                100 * .rows$rate[i], 100 * .rows$lo[i], 100 * .rows$hi[i]))
  .rows$cond <- factor(.rows$cond, levels = .rows$cond)
  .rows$grp  <- factor(.rows$grp,  levels = c(POS, NEG))

  ggplot(.rows, aes(cond, rate)) +
    geom_col(width = 0.55, fill = "grey25") +
    geom_text(aes(label = sprintf("%d / %d", det, n)), vjust = -0.4, size = 3.3) +
    facet_grid(~ grp, scales = "free_x", space = "free_x") +
    scale_y_continuous(labels = scales::percent, limits = c(0, 1.12),
                       expand = expansion(mult = c(0, 0))) +
    labs(x = NULL, y = "files with a detection") +
    theme_bw(base_size = 11) +
    theme(panel.grid.major.x = element_blank(), panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "grey93", colour = NA),
          strip.text = element_text(face = "bold", size = 9.5)) -> .p_ctl
  dir.create("thesis/graphics", showWarnings = FALSE, recursive = TRUE)
  ggsave("thesis/graphics/flagship_decoy.png", .p_ctl, width = 6.4, height = 3.6, dpi = 200)
  .p_ctl
}
```

The two controls fail in opposite directions and both come back clean. The
confirmed files re-detect at MS1 at the rates the figure gives, so the gating
loses almost nothing that MS2 had already established. In the pre-dose plasma
draws — real matrix, true m/z, true retention window, drug absent by design —
the extraction finds nothing. Both counts are computed in the chunk above rather
than written into this paragraph: the earlier hand-written pair (82 and 106) was
left behind by the uncapped re-confirmation and understated the control by
nearly an order of magnitude.

That is what licenses Part A: the prevalence gains that follow are a gain in
sensitivity rather than an accumulation of false positives.

It licenses less than a decoy-based figure would, and the difference should be
stated rather than glossed. The negative control is 49 files from one cohort, so
it constrains the false-positive rate in dosed-study plasma, not across the
corpus, and no comparable known-negative exists for the other eleven deposits or
for urine, faeces and skin. Proteomics can quote a corpus-wide false-transfer
rate (IonQuant measures 2.3% at a nominal 1% FDR, Yu 2021) because a reversed
peptide sequence is knowably false in *every* run; nothing in an untargeted
metabolomics archive is knowably absent in every file. Closing that gap needs
cohorts with a documented unexposed arm, not a better statistic.


# Part A: What propagation adds, per dataset and across the corpus

## What the field does, and how we extend it

The reverse-metabolomics / MASST tools answer one question at repository scale:
in what fraction of files does a molecule appear? Their cross-cohort output is a
**prevalence table or heatmap** of molecule-by-cohort detection rates, built
from MS2 spectral matches, and used strictly for hypothesis generation. They
never quantify intensity across deposits, and their own authors state that a
negative result is not evidence of absence, because a compound present at low
abundance may never be selected for fragmentation.

That last point is exactly the gap MS1 propagation fills. We run the same
prevalence analysis, but replace "did a file get an MS2 match?" with "does the
compound's signal appear at its confirmed mass and retention time?", which is
recorded for every file whether or not the instrument fragmented it. So Part A
is the field's own analysis at higher sensitivity: **per compound, per cohort,
the MS2 detection rate versus the MS1-propagated detection rate.** It rests on
presence, not on a quantity, so it is immune to the response-factor and isobaric
problems that constrain the ratio.

One unit note: prevalence is a **per-file** rate, so this section uses the
per-file table (not the subject-deduped one) and compares it to the per-file
MASST hits, restricted to files a metadata source covers.

```{r prevalence-setup}
#| echo: false
pf <- samples_perfile
pf$has_metadata <- {
  k <- paste0(pf$dataset_id, "|", basename(as.character(pf$file)))
  h <- has_meta_lookup[k]; ifelse(is.na(h), FALSE, unname(h))
}
pf <- pf[pf$has_metadata %in% TRUE, ]
pf$fk <- fkey(pf$file)

# MS2 baseline: metadata-carrying files with a MASST hit of each role. Restrict
# the hits to the same file universe (dataset_id | normalised key) so MS2 and
# MS1 rates share a denominator.
valid_dk       <- paste0(pf$dataset_id, "|", pf$fk)
masst_hits$fk  <- fkey(masst_hits$file_path)
masst_hits$dk  <- paste0(masst_hits$dataset_id, "|", masst_hits$fk)
mh_valid       <- masst_hits[masst_hits$dk %in% valid_dk, ]

# The ONE deposit pair that is the same study under two accessions, identified
# from subject identity: the same 14 subjects, and on every shared file the
# subject, timepoint, study day and body site agree 461/461. One physical file
# shared by that pair must count ONCE. Rows from the pair share a namespace;
# every other deposit keeps its own, so identical names across unrelated
# deposits never merge.
DUP_STUDY <- c("MSV000082493", "MSV000084008")
dup_key <- function(dataset_id, fk)
  ifelse(dataset_id %in% DUP_STUDY, paste0("DUPSTUDY|", fk),
         paste0(dataset_id, "|", fk))


ms2 <- mh_valid |>
  dplyr::distinct(dataset_id, role, fk) |>
  dplyr::count(dataset_id, role, name = "ms2") |>
  tidyr::pivot_wider(names_from = role, values_from = ms2, values_fill = 0)
if (!"parent"     %in% names(ms2)) ms2$parent     <- 0L
if (!"metabolite" %in% names(ms2)) ms2$metabolite <- 0L

# A file supports a RATIO only when both compounds are present, so that is
# counted separately: it is the quantity the downstream ratio work actually
# depends on.
ms2_both <- mh_valid |>
  dplyr::distinct(dataset_id, role, fk) |>
  dplyr::count(dataset_id, fk, name = "n_roles") |>
  dplyr::filter(n_roles == 2) |>
  dplyr::count(dataset_id, name = "both")

prevalence <- pf |>
  dplyr::group_by(dataset_id) |>
  dplyr::summarise(n_files        = dplyr::n(),
                   ms1_parent     = sum(detection_tier %in% c("both", "parent_only")),
                   ms1_metabolite = sum(detection_tier %in% c("both", "metab_only")),
                   ms1_both       = sum(detection_tier == "both"),
                   .groups = "drop") |>
  dplyr::left_join(ms2_both, by = "dataset_id") |>
  dplyr::mutate(both = tidyr::replace_na(as.integer(both), 0L)) |>
  dplyr::left_join(ms2, by = "dataset_id") |>
  dplyr::mutate(dplyr::across(c(parent, metabolite),
                              ~ tidyr::replace_na(as.integer(.), 0L)),
                ds = short_ds(dataset_id))

prev_long <- prevalence |>
  dplyr::transmute(ds, n_files,
    `omeprazole (MS2)`          = parent        / n_files,
    `omeprazole (MS1)`          = ms1_parent    / n_files,
    `5-OH-omeprazole (MS2)`     = metabolite    / n_files,
    `5-OH-omeprazole (MS1)`     = ms1_metabolite / n_files) |>
  tidyr::pivot_longer(-c(ds, n_files), names_to = "panel", values_to = "rate") |>
  tidyr::separate(panel, into = c("compound", "method"), sep = " \\(") |>
  dplyr::mutate(method = sub("\\)", "", method))
```

## The prevalence heatmap: MS2 versus MS1, per compound, per cohort

This is the figure that shows what the pipeline buys. Each row is a cohort, each
column a (compound, detection-method) pair; colour is the fraction of that
cohort's files in which the compound is detected. The MS1 columns should be at
least as dark as their MS2 counterparts, and much darker wherever the instrument
under-fragmented the compound.

```{r prevalence-heatmap, fig.width = 8, fig.height = 5.5}
prev_long |>
  dplyr::mutate(col = paste0(compound, "\n", method),
                col = factor(col, levels = c(
                  "omeprazole\nMS2", "omeprazole\nMS1",
                  "5-OH-omeprazole\nMS2", "5-OH-omeprazole\nMS1"))) |>
  ggplot(aes(x = col, y = reorder(ds, rate, sum), fill = rate)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = scales::percent(rate, accuracy = 1)), size = 2.7) +
  scale_fill_gradient(low = "grey95", high = "#1b6837",
                      labels = scales::percent, name = "detection\nrate") +
  labs(x = NULL, y = NULL,
       title = "Files with a detection per cohort, MASST MS2 and MS1-propagated") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(lineheight = 0.9))
```

## The propagation gain per cohort

The same numbers as a gain: MS1 detections divided by MS2 detections, per
compound per cohort. This is the honest, cohort-resolved version of the
propagation-gain headline (a single corpus-wide multiplier hides that the gain
is largest exactly where MS2 coverage is thinnest).

A cohort whose MS2 baseline is zero inside the analysed files has **no defined
gain**: a ratio needs a denominator. Printing `x / max(0, 1)` would invent one,
so those cells read `n/a` and the counts alone carry the meaning.

**MSV000082433 (\*) needs its own note, because it is not a gain at all.** Its
entire MS2 evidence for both roles is a single file, `957.mzML`, which sits on
`20161103_Amerindian_Plate6`. Pan-ReDU indexes only plates 1 to 3 of this
deposit, so that file is described by no metadata source, and the MS2 columns
above, which are restricted to annotated files, cannot count it. The file itself
is not the problem: it carries a clean, textbook omeprazole peak (1.1e8 at the
confirmed retention time) and a confirmed 5-OH peak. It is simply unusable,
because without metadata it cannot enter any stratified analysis.

```{r ds82433-parent}
#| echo: false
# Counted, not written in: this pair read "219 of 224" for a while after a
# re-extraction moved it.
.x433 <- samples_perfile[samples_perfile$dataset_id == "MSV000082433", ]
N433_NOPEAK <- sum(is.na(.x433$par_auc)); N433_FILES <- nrow(.x433)
```

Inside the annotated files the picture is the opposite. The parent produces **no
chromatographic peak** at the confirmed retention time in `r N433_NOPEAK` of
`r N433_FILES` files; the handful of files with any in-window signal show
single-scan spikes (one point at 0.5 to 0.83 s scan spacing, with nothing on
either side), which the point-count gate correctly rejects. So the metabolite
counts for this cohort are a presence observation with **no parent, no ratio,
and no MS2 baseline to compare against**. Read that row as "one confirmed but
unannotated file, plus a metabolite-only presence signal", never as a
propagation gain.

```{r prevalence-gain}
#| echo: false
gain_str <- function(ms2, ms1) {
  ifelse(ms2 > 0,
         sprintf("%d -> %d  (%.1fx)", ms2, ms1, ms1 / pmax(ms2, 1)),
         sprintf("%d -> %d  (n/a: no MS2 in analysed files)", ms2, ms1))
}
prevalence |>
  dplyr::transmute(cohort = ifelse(ds == "MSV000082433", paste0(ds, " *"), ds), n_files,
    `omeprazole: MS2 -> MS1` = gain_str(parent, ms1_parent),
    `5-OH: MS2 -> MS1`       = gain_str(metabolite, ms1_metabolite),
    # a ratio needs both compounds in one file, so it is counted here rather
    # than inferred from the two columns beside it
    `usable ratio: MS2 -> MS1` = gain_str(both, ms1_both)) |>
  dplyr::arrange(dplyr::desc(n_files))

write_csv(dplyr::arrange(prevalence, dplyr::desc(n_files)),
          "artifacts/prevalence_by_deposit.csv")
```

Detecting one compound is not the same as being able to compute a ratio, which
needs both in the same file. That is the number the whole quantitative strand
rests on, so it is counted separately rather than inferred from the two columns
above.

**No file is counted twice.** MSV000082493 and MSV000084008 share 333 physical
files; those contribute once, so the totals below are over distinct files.
Simply summing the per-cohort rows above would have counted them twice and
inflated every number.

```{r propagation-corpus-totals}
#| echo: false
pfd <- pf
pfd$dk <- dup_key(pfd$dataset_id, pfd$fk)
pfd <- pfd[!duplicated(pfd$dk), ]                    # one row per physical file

mhd <- mh_valid
mhd$dk <- dup_key(mhd$dataset_id, mhd$fk)
ms2_n <- function(r) length(unique(mhd$dk[mhd$role == r]))
ms2_both_n <- length(intersect(mhd$dk[mhd$role == "parent"],
                               mhd$dk[mhd$role == "metabolite"]))

tot <- data.frame(
  quantity = c("omeprazole", "5-OH-omeprazole", "both, i.e. a usable ratio"),
  MASST_MS2 = c(ms2_n("parent"), ms2_n("metabolite"), ms2_both_n),
  MS1_propagated = c(sum(pfd$detection_tier %in% c("both", "parent_only")),
                     sum(pfd$detection_tier %in% c("both", "metab_only")),
                     sum(pfd$detection_tier == "both")))
tot$gain <- sprintf("%.2fx", tot$MS1_propagated / tot$MASST_MS2)

# An interval on the gain, resampling DEPOSITS rather than files. Files inside a
# deposit share its chromatography, its depositor and its detection rate, so they
# are not independent draws and a file-level bootstrap reports an interval far
# tighter than the evidence supports. Twelve deposits is the real sample size.
set.seed(1)
.dep <- unique(pfd$dataset_id)
.gboot <- replicate(2000, {
  d <- sample(.dep, replace = TRUE)
  # A deposit drawn twice must count twice, so each draw carries its own tag;
  # without it unique()/intersect() would silently collapse the duplicate and the
  # resample would be a subsample.
  tag <- function(tb, j) {
    x <- tb[tb$dataset_id == d[j], , drop = FALSE]
    if (nrow(x)) x$dk <- paste0(j, "|", x$dk)
    x
  }
  p2 <- do.call(rbind, lapply(seq_along(d), function(j) tag(pfd, j)))
  m2 <- do.call(rbind, lapply(seq_along(d), function(j) tag(mhd, j)))
  ms1 <- c(sum(p2$detection_tier %in% c("both", "parent_only")),
           sum(p2$detection_tier %in% c("both", "metab_only")),
           sum(p2$detection_tier == "both"))
  # distinct FILES, matching how the point estimate is counted
  ms2 <- c(length(unique(m2$dk[m2$role == "parent"])),
           length(unique(m2$dk[m2$role == "metabolite"])),
           length(intersect(unique(m2$dk[m2$role == "parent"]),
                            unique(m2$dk[m2$role == "metabolite"]))))
  ifelse(ms2 > 0, ms1 / ms2, NA_real_)
})
.q <- apply(.gboot, 1, quantile, c(0.025, 0.975), na.rm = TRUE)
tot$gain_ci <- sprintf("%.2f to %.2f", .q[1, ], .q[2, ])
knitr::kable(tot, caption = sprintf(paste(
  "Files with a detection, %d distinct metadata-carrying files across 12 deposits.",
  "The gain interval is a bootstrap over deposits, not files."),
  nrow(pfd)))
```

```{r fig-propagation-tables, fig.width = 6.8, fig.height = 3.4}
#| echo: false
# The same table as a figure, so the deck shows the notebook's own numbers.
d <- as.data.frame(lapply(
  stats::setNames(tot, c("compound", "MASST MS2", "MS1 propagated", "gain")),
  as.character), check.names = FALSE)
d$.row <- rev(seq_len(nrow(d)))
hl <- d$.row[3]
long <- tidyr::pivot_longer(d, -.row, names_to = "col", values_to = "val")
long$col <- factor(long$col, levels = names(d)[names(d) != ".row"])
ggplot(long, aes(as.integer(col), .row)) +
  geom_tile(aes(fill = .row %in% hl), colour = "white", linewidth = 1.1) +
  geom_text(aes(label = val, fontface = ifelse(.row %in% hl, "bold", "plain")),
            size = 3.9) +
  annotate("text", x = seq_len(ncol(d) - 1), y = max(d$.row) + 0.85,
           label = names(d)[names(d) != ".row"], fontface = "bold", size = 3.9) +
  scale_fill_manual(values = c(`TRUE` = "grey85", `FALSE` = "grey96"), guide = "none") +
  scale_y_continuous(expand = expansion(add = c(0.55, 1.3))) +
  scale_x_continuous(expand = expansion(add = 0.5)) +
  labs(title = sprintf("Files with a detection, %d distinct files across 12 deposits",
                       nrow(pfd)), x = NULL, y = NULL) +
  theme_void(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 12, hjust = 0,
                                  margin = margin(b = 10)))
```

The ratio row is the one that matters: a file needs **both** compounds before it
can say anything quantitative, and that is where propagation helps most, because
it can rescue either compound in a file where MS2 caught only the other one.

## The resource this produces, deposit by deposit

The totals say how much was recovered. They do not say where it landed, and for
anyone deciding whether these data are worth reusing that is the more useful
question. So the same output is described three ways below: how many independent
units carry a ratio, what the ratio's level is in each deposit, and which strata
each deposit could be cut by.

The unit here is one row per subject per deposit, taking the median over that
subject's files. Three deposits record no subject, so there the file is the unit
and the count is an upper bound on the number of people rather than a number of
people.

```{r output-units}
#| echo: false
DUP_LABEL <- "MSV000082493/084008"

rr <- pf[pf$detection_tier == "both" & is.finite(pf$log2_ratio), ]
rr$dk    <- dup_key(rr$dataset_id, rr$fk)
rr       <- rr[!duplicated(rr$dk), ]
rr$study <- ifelse(rr$dataset_id %in% DUP_STUDY, DUP_LABEL, rr$dataset_id)
rr$has_subject <- !is.na(rr$subject_id) & rr$subject_id != ""
rr$unit  <- ifelse(rr$has_subject, paste0("S:", rr$study, "|", rr$subject_id),
                   paste0("F:", rr$study, "|", rr$dk))

ratio_units <- rr |>
  dplyr::group_by(study, unit) |>
  dplyr::summarise(log2_ratio  = median(log2_ratio),
                   has_subject = all(has_subject),
                   n_files     = dplyr::n(),
                   sex         = dplyr::first(sex),
                   health      = dplyr::first(health_status),
                   age         = as_num(dplyr::first(age)),
                   .groups     = "drop") |>
  dplyr::mutate(ds = short_ds(study))

out_tbl <- ratio_units |>
  dplyr::group_by(ds) |>
  dplyr::summarise(units       = dplyr::n(),
                   files       = sum(n_files),
                   unit_is     = ifelse(all(has_subject), "subject", "file"),
                   median_log2 = round(median(log2_ratio), 2),
                   IQR         = round(IQR(log2_ratio), 2),
                   sex         = sum(!is.na(harmonise_sex(sex))),
                   health      = sum(!is.na(harmonise_health(health))),
                   age         = sum(is.finite(age)),
                   .groups     = "drop") |>
  dplyr::arrange(dplyr::desc(units))
knitr::kable(out_tbl, caption = sprintf(
  "Units carrying a usable ratio: %d across %d studies. The last three columns count units whose deposit records that field in a form a contrast can use.",
  nrow(ratio_units), dplyr::n_distinct(ratio_units$ds)))
write.csv(out_tbl, "artifacts/output_resource_by_deposit.csv", row.names = FALSE)
```

`r nrow(out_tbl)` studies carry at least one ratio, `r sum(out_tbl$units)` units
in total, and only MSV000082433 carries none. That is the resource: small, but
it did not exist before the extraction, since a ratio needs both compounds in
one file and MS2 supplied that in `r ms2_both_n` files, concentrated almost
entirely in the dosed cohort.

Two of those rows are not independent of each other. MSV000092754 and
MSV000092757 are the same subject panel measured twice, by HILIC and by reverse
phase, so they contribute
`r sum(out_tbl$units[out_tbl$ds %in% c("MSV000092754", "MSV000092757")])` units
but not that many people; they are kept apart because the two measurements
differ, not because the two deposits do.

The second column is the reason the corpus cannot simply be pooled. The median
level differs by deposit far more than it differs between people inside a
deposit. Deposits differ in instrument, gradient and specimen at once, so one
median carries all three and this figure separates none of them. The variance
decomposition further down does, and the answer is not the one the
response-factor argument predicts.

```{r fig-ratio-by-dataset, fig.width = 7.6, fig.height = 4.4}
#| echo: false
ord <- ratio_units |> dplyr::group_by(ds) |>
  dplyr::summarise(m = median(log2_ratio), .groups = "drop") |>
  dplyr::arrange(m)
ratio_units$ds <- factor(ratio_units$ds, levels = ord$ds)

p_ratio_ds <- ggplot(ratio_units, aes(x = log2_ratio, y = ds)) +
  geom_vline(xintercept = 0, linetype = 3, colour = "grey55") +
  geom_boxplot(outlier.shape = NA, width = 0.55, fill = "grey92", colour = "grey35") +
  geom_jitter(aes(shape = has_subject), height = 0.16, width = 0, alpha = 0.7, size = 1.7) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 1),
                     labels = c(`TRUE` = "subject", `FALSE` = "file (no subject recorded)"),
                     name = "unit") +
  labs(x = expression(log[2]*"(omeprazole / 5-OH-omeprazole)"), y = NULL) +
  theme_bw(base_size = 11) + theme(legend.position = "bottom")
p_ratio_ds
ggsave("thesis/graphics/flagship_ratio_by_dataset.png", p_ratio_ds,
       width = 7.6, height = 4.4, dpi = 200)
```

A boxplot per deposit rather than a pooled one, because pooling these would
average over whatever the between-deposit difference is made of and produce a
number that describes no cohort. Within a deposit the spread is what a contrast
can work with; between deposits it is not yet interpretable. That distinction is
what separates the next two sections.

```{r fig-output-strata, fig.width = 8.2, fig.height = 3.6}
#| echo: false
LV <- c("female", "male", "healthy", "not healthy", "<40", "40-59", ">=60")
str_long <- dplyr::bind_rows(
  ratio_units |> dplyr::transmute(ds, field = "sex", value = harmonise_sex(sex)),
  ratio_units |> dplyr::transmute(ds, field = "health status", value = harmonise_health(health)),
  ratio_units |> dplyr::transmute(ds, field = "age",
                            value = cut(age, c(-Inf, 40, 60, Inf),
                                        labels = c("<40", "40-59", ">=60"),
                                        right = FALSE) |> as.character())) |>
  dplyr::filter(!is.na(value)) |>
  dplyr::mutate(field = factor(field, levels = c("sex", "health status", "age")),
                value = factor(value, levels = LV))

p_strata <- ggplot(str_long, aes(y = ds, fill = value)) +
  geom_bar() +
  facet_wrap(~ field, nrow = 1, scales = "free_x") +
  scale_fill_brewer(palette = "Set2", name = NULL, drop = FALSE) +
  labs(x = "units carrying a ratio", y = NULL) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", legend.key.size = unit(0.4, "cm"))
p_strata
ggsave("thesis/graphics/flagship_output_strata.png", p_strata,
       width = 8.2, height = 3.6, dpi = 200)

knitr::kable(str_long |> dplyr::count(field, value, name = "units"),
             caption = "Stratifier levels available across the ratio-bearing units")
```

Timepoint is deliberately absent. It is recorded almost only in the dosed
cohort, where it is the design rather than a stratifier, and including it would
have made the corpus look richer in covariates than it is.

Sex is harmonised before counting: three deposits write `F`/`M` and the rest
`female`/`male`, which would otherwise be four levels rather than two. Health
status is mapped through the same explicit whitelist the covariate scan uses, so
labels that are neither clearly healthy nor clearly a disease arm — `chronic
illness`, `compensated`, `decompensated` — stay unset instead of being forced
onto one side.

The fields that survive are thin and unevenly spread, and no field is recorded
across all ten studies. This is the ceiling on everything the next two sections
attempt, and it is a property of what depositors wrote down rather than of the
extraction.

## The analysis unit, and the designs the corpus actually has

Everything from here on is epidemiology rather than extraction, so the unit
changes. The prevalence and gain tables above count **files**, correctly,
because propagation is a per-injection operation. A statistical contrast is a
different question and needs a unit that corresponds to an independent
observation.

Three things sit between a file and an independent observation, and each has
bitten this analysis at some point:

1. **Replicate injections.** One physical sample can be run several times.
2. **Repeated measures within subject.** The probe cohort is 1,137 files from
   **14 people**, about 27 rows each, and every one of those people appears in
   several body sites. Testing body site against detection on the raw rows
   treats one person as 27 observations and returns a p-value that means
   nothing.
3. **Deposits that record no subject at all.** Three do. There a file is the
   only available unit and the design is simply unknown, which has to be said
   rather than assumed away.

The rule used below: **one row per subject per level of the contrast**. A
subject sampled at ten timepoints contributes one observation; a subject sampled
in four matrices contributes one observation *per matrix*, which is correct
because body site is the thing being contrasted. Where no subject is recorded
the file is the unit and the result is flagged `design unknown` and never
reported as confirmatory.

```{r unit-setup}
# One file universe, MS1 and MS2 flags, independent studies rather than
# accessions.
ms2_keys <- unique(paste0(mh_valid$dataset_id, "|", mh_valid$fk))
pm <- pf
pm$ms2 <- paste0(pm$dataset_id, "|", pm$fk) %in% ms2_keys
pm$ms1 <- pm$detection_tier %in% c("both", "parent_only", "metab_only")
pm$par_ms1 <- pm$detection_tier %in% c("both", "parent_only")
pm$met_ms1 <- pm$detection_tier %in% c("both", "metab_only")
par_keys <- unique(paste0(mh_valid$dataset_id, "|", mh_valid$fk)[mh_valid$role == "parent"])
met_keys <- unique(paste0(mh_valid$dataset_id, "|", mh_valid$fk)[mh_valid$role == "metabolite"])
pm$par_ms2 <- paste0(pm$dataset_id, "|", pm$fk) %in% par_keys
pm$met_ms2 <- paste0(pm$dataset_id, "|", pm$fk) %in% met_keys

# DUP_STUDY / dup_key are defined with the MS2 baseline above, where the
# reasoning is set out once. The appendix documents the evidence.

pm$study <- dplyr::case_when(
  grepl(paste(DUP_STUDY, collapse = "|"), pm$dataset_id) ~ "Jarmusch probe",
  grepl("MSV000092754|MSV000092757", pm$dataset_id)      ~ "Chagas serum",
  TRUE ~ short_ds(pm$dataset_id))
pm <- pm[!duplicated(paste(pm$study, pm$fk)), ]

nz <- function(v) ifelse(is.na(v) | !nzchar(as.character(v)), NA_character_, as.character(v))
pm$subj <- nz(pm$subject_id)

# Deposits name the same matrix differently: blood_plasma and "blood plasma",
# UANoAdditive_YellowTop and urine, fecal and feces. Left raw, one matrix splits
# into several rows of any cross-cohort table and each row reads as a different
# biology.
canon_site <- function(x) {
  y <- tolower(trimws(gsub("[_.]+", " ", as.character(x))))
  dplyr::case_when(
    is.na(y) | !nzchar(y)               ~ NA_character_,
    grepl("plasma", y)                  ~ "plasma",
    grepl("serum", y)                   ~ "serum",
    grepl("^ua|urine", y)               ~ "urine",
    grepl("fecal|feces|faec|stool", y)  ~ "faeces",
    grepl("forehead|arm skin|^skin", y) ~ "skin",
    grepl("saliva", y)                  ~ "saliva",
    grepl("oral cavity|mouth", y)       ~ "oral cavity",
    grepl("milk", y)                    ~ "milk",
    grepl("nasal", y)                   ~ "nasal cavity",
    grepl("vagina", y)                  ~ "vagina",
    grepl("anal", y)                    ~ "anal region",
    TRUE                                ~ y)
}
pm$site  <- canon_site(pm$body_site)
pm$sex2  <- dplyr::case_when(grepl("^f", tolower(trimws(pm$sex))) ~ "female",
                             grepl("^m", tolower(trimws(pm$sex))) ~ "male",
                             TRUE ~ NA_character_)
pm$clin  <- ifelse(!is.na(nz(pm$health_status)), pm$health_status, pm$group)
pm$agen  <- as_num(pm$age)
pm$ageb  <- as.character(cut(pm$agen, c(-Inf, 40, 60, Inf),
                             labels = c("<40", "40-60", ">60")))

# The unit rule. Returns one row per (subject, level of g), with the detection
# columns OR-ed within that unit, and records whether the design is
# subject-resolved.
units_for <- function(x, g, cols) {
  keep <- !is.na(g) & nzchar(g)
  x <- x[keep, , drop = FALSE]; g <- g[keep]
  if (!nrow(x)) return(NULL)
  # Per row, not per deposit: a file with no subject is its own unit, since
  # collapsing the unlabelled rows of a partly-labelled deposit would invent one
  # subject per level.
  known <- subject_resolved(x$subj, sibling_ids = unlist(sibling_id_map[unique(x$dataset_id)], use.names = FALSE))
  u <- ifelse(is.na(x$subj), paste0("file:", x$fk), paste(x$subj, g))
  out <- data.frame(unit = u, g = g, x[, cols, drop = FALSE]) |>
    dplyr::group_by(unit) |>
    dplyr::summarise(g = dplyr::first(g),
                     dplyr::across(dplyr::all_of(cols), any), .groups = "drop")
  attr(out, "design") <- if (!known) "design unknown"
    else if (any(is.na(x$subj))) sprintf("partial (%.0f%% subject-labelled)",
                                         100 * mean(!is.na(x$subj)))
    else "subject-resolved"
  out
}
# Sample-size gate: a level needs >= 5 units to enter, and >= 2 levels must
# survive.
gate <- function(a, min_level = 5) {
  a <- a[a$g %in% names(table(a$g))[table(a$g) >= min_level], ]
  if (dplyr::n_distinct(a$g) < 2) NULL else a
}
# >= 3 events AND >= 3 non-events, or the table degenerates and Fisher is
# undefined.
fish <- function(g, v) if (sum(v) >= 3 && sum(!v) >= 3)
  fisher.test(table(g, v), simulate.p.value = TRUE, B = 20000)$p.value else NA_real_
```

```{r unit-design-table}
design_tbl <- pm |>
  dplyr::group_by(study) |>
  dplyr::summarise(files = dplyr::n(),
                   subjects = dplyr::n_distinct(subj[!is.na(subj)]),
                   subject_recorded = paste0(round(100 * mean(!is.na(subj))), "%"),
                   files_per_subject = ifelse(dplyr::n_distinct(subj[!is.na(subj)]) > 0,
                     round(dplyr::n() / dplyr::n_distinct(subj[!is.na(subj)]), 1), NA),
                   .groups = "drop") |>
  dplyr::arrange(dplyr::desc(dplyr::coalesce(files_per_subject, 0)))
knitr::kable(design_tbl, caption = "Study designs: how many files stand behind one person")
```

Read the top and bottom rows. The probe cohort is
`r design_tbl$files_per_subject[1]` files per subject, so any test run on its
raw rows is inflated roughly that many times over. The three deposits reporting
`0` subjects record no subject identifier at all, so nothing there can be
de-duplicated and every contrast in them is exploratory.

## What propagation unlocks: a per-dataset contrast scan

The question this section answers is not "does propagation find more files" but
**"does propagation let a single deposit answer a question it otherwise could
not"**. That is a per-dataset question, tested inside each deposit, with no
pooling anywhere.

Every combination of study, stratifier and compound is tested twice: once using
only the MASST MS2 detections, once using the MS1-propagated detections. A
contrast enters only if it clears the sample-size gate above. Because this is a
scan of several dozen contrasts, both methods get Benjamini-Hochberg control
across everything they can test.

```{r unlock-scan}
STRAT <- c(sex = "sex2", `body site` = "site",
           `clinical group` = "clin", `age band` = "ageb")
ROLES <- list(omeprazole        = c("par_ms2", "par_ms1"),
              `5-OH-omeprazole` = c("met_ms2", "met_ms1"),
              `either compound` = c("ms2", "ms1"))
set.seed(1)
scan_rows <- list()
for (s in unique(pm$study)) for (vn in names(STRAT)) for (rn in names(ROLES)) {
  x  <- pm[pm$study == s, , drop = FALSE]
  a0 <- units_for(x, x[[STRAT[[vn]]]], ROLES[[rn]])
  if (is.null(a0)) next
  dsg <- attr(a0, "design"); a <- gate(a0)
  if (is.null(a)) next
  p2 <- fish(a$g, a[[ROLES[[rn]][1]]]); p1 <- fish(a$g, a[[ROLES[[rn]][2]]])
  if (is.na(p1) && is.na(p2)) next
  scan_rows[[length(scan_rows) + 1]] <- data.frame(
    study = s, stratifier = vn, compound = rn, design = dsg, units = nrow(a),
    det_ms2 = sum(a[[ROLES[[rn]][1]]]), det_ms1 = sum(a[[ROLES[[rn]][2]]]),
    p2 = p2, p1 = p1)
}
SCAN <- dplyr::bind_rows(scan_rows)
SCAN$q2 <- NA_real_; SCAN$q1 <- NA_real_
SCAN$q2[!is.na(SCAN$p2)] <- p.adjust(SCAN$p2[!is.na(SCAN$p2)], "BH")
SCAN$q1[!is.na(SCAN$p1)] <- p.adjust(SCAN$p1[!is.na(SCAN$p1)], "BH")

knitr::kable(data.frame(
  method = c("MASST MS2 only", "MS1 propagated"),
  testable = c(sum(!is.na(SCAN$p2)), sum(!is.na(SCAN$p1))),
  significant_FDR_05 = c(sum(SCAN$q2 < 0.05, na.rm = TRUE),
                         sum(SCAN$q1 < 0.05, na.rm = TRUE))),
  caption = "Contrasts each method can test, and how many survive FDR control")
```

```{r unlock-table}
sig <- SCAN[!is.na(SCAN$q1) & SCAN$q1 < 0.05, ]
sig <- sig[order(sig$q1), ]
sig$MASST <- ifelse(is.na(sig$q2), "could not test",
             ifelse(sig$q2 < 0.05, sprintf("also found it (q = %.3g)", sig$q2),
                    sprintf("NOT significant (q = %.3g)", sig$q2)))
sig$new <- is.na(sig$q2) | sig$q2 >= 0.05
knitr::kable(sig |> dplyr::transmute(study, stratifier, compound, design, units,
                                     MS2 = det_ms2, MS1 = det_ms1, MASST,
                                     q_MS1 = sprintf("%.3g", q1)),
             caption = "Every contrast MS1 finds at FDR 0.05, and what MASST said about it")

lost <- SCAN[!is.na(SCAN$q2) & SCAN$q2 < 0.05 & (is.na(SCAN$q1) | SCAN$q1 >= 0.05), ]
```

```{r fig-unlock-scan, fig.width = 9, fig.height = 5}
#| echo: false
.s <- sig
.s$label <- paste0(.s$study, "\n", .s$stratifier, ", ", .s$compound)
.s$label <- factor(.s$label, levels = rev(.s$label))
.pl <- tidyr::pivot_longer(.s[, c("label", "design", "q2", "q1")],
                           c(q2, q1), names_to = "method", values_to = "q")
.pl$method <- ifelse(.pl$method == "q1", "MS1 propagated", "MASST MS2 only")
.pl$q[is.na(.pl$q)] <- 1.2                       # "could not test" parked off-scale
ggplot(.pl, aes(q, label, colour = method, shape = method)) +
  geom_vline(xintercept = 0.05, linetype = 2, colour = "grey55") +
  geom_line(aes(group = label), colour = "grey75", linewidth = 0.5) +
  geom_point(size = 2.8) +
  scale_x_continuous(trans = "log10",
                     breaks = c(1e-4, 1e-3, 1e-2, 0.05, 0.2, 1),
                     labels = c("0.0001", "0.001", "0.01", "0.05", "0.2", "1 / n.t.")) +
  scale_colour_manual(values = c("MS1 propagated" = "#111111",
                                 "MASST MS2 only" = "#9a9a9a"), name = NULL) +
  scale_shape_manual(values = c("MS1 propagated" = 16, "MASST MS2 only" = 17), name = NULL) +
  labs(x = "Benjamini-Hochberg q, log scale", y = NULL,
       title = "Benjamini-Hochberg q per within-deposit contrast, MASST MS2 and MS1-propagated") +
  theme_bw(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        axis.text.y = element_text(size = 8.5, lineheight = 0.95),
        plot.title = element_text(face = "bold", size = 12.5))
```

Three numbers matter here.

**Nothing is lost.** `r nrow(lost)` contrasts that MASST finds significant fail
to replicate under propagation. Whatever else propagation does, it does not
overturn the MS2 result, which is the minimum a match-between-runs method has to
demonstrate.

**`r sum(sig$new)` of `r nrow(sig)` findings are new**: MASST either could not
test the contrast at all or returned a null. Of those,
**`r sum(sig$new & sig$design == "subject-resolved")` sit on a subject-resolved
design** and are the ones worth quoting; the remainder are in deposits with no
subject identifier and are exploratory.

**The defensible new findings are all the same finding.** They are the
COVID-severity contrasts in MTBLS1866 and MTBLS11656, which is to say the two
cohorts that also replicate each other in the section above. The per-dataset
scan and the replication are the same fact approached from two directions, which
is the strongest form the evidence takes anywhere in this project.

MTBLS1866 is the demonstration piece. On the MASST detections the
metabolite-versus- severity contrast returns a **perfect null**: three detected
subjects, q = 1. MS1 supplies 37 and the same contrast in the same cohort
becomes q = 0.0003. Nothing about the cohort, the metadata or the test changed.
Only whether the compound was looked for in MS1.

Two cautions, both visible in the table. The body-site contrasts in MSV000094097
look strong and cannot be claimed, because that deposit records no subjects and
its 144 files might be 144 people or 36 people sampled at four sites. And the
probe cohort's body-site contrasts are significant under **both** methods once
aggregated to 55 subject-by-site units; on raw files they were computed across
1,125 rows from 14 people, which was never a valid test even though it happened
to reach the same verdict.

### The one deposit worth showing despite an unknown design

MSV000094097 produces the largest single jump in the scan and cannot be counted
as a result, so it is worth showing exactly as it is rather than dropping it.
MASST finds the metabolite in 2 of its files. Propagation finds it in 32, across
four body sites, and the body-site contrast goes from untestable to q = 0.0003.

```{r fig-094097, fig.width = 7.6, fig.height = 4.2}
#| echo: false
.d97 <- pm[pm$study == "MSV000094097" & !is.na(pm$site), ]
.t97 <- .d97 |>
  dplyr::group_by(site) |>
  dplyr::summarise(files = dplyr::n(),
                   `omeprazole (MS2)` = sum(par_ms2), `omeprazole (MS1)` = sum(par_ms1),
                   `5-OH-omeprazole (MS2)` = sum(met_ms2), `5-OH-omeprazole (MS1)` = sum(met_ms1),
                   .groups = "drop") |>
  tidyr::pivot_longer(-c(site, files), names_to = "panel", values_to = "count") |>
  tidyr::separate(panel, into = c("compound", "method"), sep = " \\(") |>
  dplyr::mutate(method = sub("\\)", "", method), rate = count / files,
                col = factor(paste0(compound, "\n", method),
                             levels = c("omeprazole\nMS2", "omeprazole\nMS1",
                                        "5-OH-omeprazole\nMS2", "5-OH-omeprazole\nMS1")),
                row = paste0(site, "  (", files, " files)"))
ggplot(.t97, aes(col, reorder(row, rate, sum), fill = rate)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = count, colour = rate > 0.3), size = 4, fontface = "bold") +
  scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey15"), guide = "none") +
  scale_fill_gradient(low = "grey94", high = "black", guide = "none") +
  labs(x = NULL, y = NULL,
       title = "Files with a detection by body site, MSV000094097") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), axis.text.x = element_text(lineheight = 0.9),
        plot.title = element_text(face = "bold", size = 13))
```

The pattern is the one the rest of the corpus shows: both compounds present in
the blood fractions, the metabolite reaching further than the parent, and the
non-blood sites near empty. That coherence is why it is worth showing. What
cannot be done is attach a p-value to it, because without subject identifiers
there is no way to know whether the four body sites are four samples from the
same person, in which case the contrast is paired and the test is invalid.

This is a **metadata** limitation, not a measurement one. The spectra are fine;
the deposit simply never recorded who each file came from. It is the single
clearest argument in this project for depositors recording a subject identifier.

## The corpus-wide gain, with and without the metadata restriction

The per-cohort table above counts only files a metadata source covers, because
every stratified analysis downstream needs that metadata. But the pipeline
extracts from more files than that: a file with no metadata still yields a valid
presence call, it just cannot be assigned to a subject or a group. Both numbers
are worth stating, because they answer different questions. The unrestricted row
is what the extraction actually recovers; the restricted row is what survives
into an analysis.

```{r prevalence-gain-total}
#| echo: false
# DUP_STUDY / dup_key as defined with the MS2 baseline above; the reasoning for
# why only this pair collapses, and why 092754/092757 must not, is stated there.

prop_totals <- function(d, label, dedup_files = TRUE) {
  d$fk <- fkey(d$file)
  dk   <- unique(paste0(d$dataset_id, "|", d$fk))
  mh   <- masst_hits[masst_hits$dk %in% dk, ]
  if (dedup_files) {
    # Count distinct physical files per role. dup_key must not be deduplicated
    # across roles first: a file with both a parent and a metabolite hit would
    # collapse to one row and silently drop a role (this deflated the metabolite
    # MS2 baseline 526 -> 146 and inflated its gain to 8x). Dedup within each
    # role.
    d  <- d[!duplicated(dup_key(d$dataset_id, d$fk)), ]
    nd <- function(role) {
      r <- mh[mh$role == role, ]
      length(unique(dup_key(r$dataset_id, r$fk)))
    }
  } else {
    nd <- function(role) nrow(dplyr::distinct(mh[mh$role == role, c("dataset_id", "fk")]))
  }
  p2 <- nd("parent"); m2 <- nd("metabolite")
  p1 <- sum(d$detection_tier %in% c("both", "parent_only"))
  m1 <- sum(d$detection_tier %in% c("both", "metab_only"))
  fmt <- function(ms2, ms1) if (ms2 > 0)
    sprintf("%d -> %d  (%.1fx)", ms2, ms1, ms1 / ms2) else
    sprintf("%d -> %d  (n/a)", ms2, ms1)
  data.frame(
    scope                    = label,
    files                    = nrow(d),
    `omeprazole: MS2 -> MS1` = fmt(p2, p1),
    `5-OH: MS2 -> MS1`       = fmt(m2, m1),
    check.names = FALSE, stringsAsFactors = FALSE)
}
rbind(
  prop_totals(samples_perfile, "all extracted, DISTINCT files",   dedup_files = TRUE),
  prop_totals(pf,              "metadata-carrying, DISTINCT",     dedup_files = TRUE),
  prop_totals(samples_perfile, "all extracted, summed (INFLATED)", dedup_files = FALSE))
```

The MS2 baseline here is **every MASST hit in `hits_curated`**, not just the
subset we sent to SIRIUS. That matters: the confirmation step was capped per
role per dataset, so dividing by the confirmed count would measure our own cap
rather than the method, and inflates the gain several-fold. The uncapped
baseline is the conservative choice and the one a "classic" MS2-only reanalysis
would have had.

## What the spread across cohorts tells us

The spread of detection rate across cohorts is not noise; it is the biology of
who is exposed. Two regimes are visible:

- **Drug-probe cohorts** (every subject dosed by design) sit near 100% for both
  compounds. They are the internal positive control: propagation should, and
  does, recover almost everything there.
- **General-population and disease cohorts** sit low, because only incidentally
  dosed subjects carry the drug. Here the MS1-over-MS2 gain is largest, and the
  absolute number of recovered exposures is the payoff: a few percent of a large
  cohort is still tens of subjects the MS2-only analysis would have missed.

The metabolite frequently has a higher detection rate than the parent (e.g. in
excretory matrices), which is expected pharmacology, not an artefact: omeprazole
is extensively metabolised, so in urine or feces the metabolite persists while
the parent is largely gone. This asymmetry is itself a hypothesis-generating
readout and is exactly the kind of pattern the presence analysis surfaces.

## Does a cohort finding replicate in an independent cohort?

The strongest argument for assembling a corpus rather than analysing one deposit
is that a claim can be **tested twice, in populations that share nothing**. Two
deposits here stratify subjects by COVID-19 severity and were produced by
unrelated groups on different continents: MTBLS1866 (Italy) and MTBLS11656
(China). Neither was collected to study omeprazole, and neither knows about the
other. If severity really tracks exposure, the pattern should appear in both.

Note this pairing is the opposite situation to MSV000082493 / MSV000084008
below, which look like two cohorts but are one study re-deposited twice.
Replication requires independence, so establishing it is a prerequisite for
reading anything as a second observation rather than an echo of the first.

```{r severity-replication}
rep_sets <- c(MTBLS1866 = "Italy", MTBLS11656 = "China")

# A column that is empty for every row is read as logical NA, and nzchar(NA) is
# TRUE, so the "is this column populated" test must exclude NA or it always
# passes.
filled <- function(v) !is.na(v) & nzchar(as.character(v))

rep_rows <- lapply(names(rep_sets), function(acc) {
  x <- pf[grepl(paste0("^", acc, "(_|$)"), pf$dataset_id), , drop = FALSE]
  if (!nrow(x)) return(NULL)
  # Deposits differ in which column carries the severity strata; take whichever
  # one this deposit actually populates.
  g  <- if (sum(filled(x$health_status)) > 10) x$health_status else x$group
  ok <- filled(g)
  data.frame(cohort = paste0(acc, " (", rep_sets[[acc]], ")"),
             stratum = as.character(g)[ok],
             par = x$detection_tier[ok] %in% c("both", "parent_only"),
             met = x$detection_tier[ok] %in% c("both", "metab_only"),
             stringsAsFactors = FALSE)
})
rep_df <- dplyr::bind_rows(rep_rows)

rep_tbl <- rep_df |>
  dplyr::group_by(cohort, stratum) |>
  dplyr::summarise(files = dplyr::n(),
                   omeprazole = sum(par), `5-OH` = sum(met), .groups = "drop") |>
  dplyr::arrange(cohort, dplyr::desc(`5-OH` / files))

B_SIM <- 20000
set.seed(1)   # the p is simulated, so pin it or the reported value drifts per render
# A simulated Fisher p cannot go below 1/(B+1). Printing that floor as "0.0000"
# would claim a precision the simulation does not have, so show it as an upper
# bound.
fmt_p <- function(p) ifelse(p <= 1 / (B_SIM + 1),
                            sprintf("< %.0e", 1 / (B_SIM + 1)), sprintf("%.4f", p))

rep_p <- rep_df |>
  dplyr::group_by(cohort) |>
  dplyr::summarise(
    p_parent = fmt_p(fisher.test(table(stratum, par),
                                 simulate.p.value = TRUE, B = B_SIM)$p.value),
    p_metabolite = fmt_p(fisher.test(table(stratum, met),
                                     simulate.p.value = TRUE, B = B_SIM)$p.value),
    .groups = "drop")

knitr::kable(rep_tbl, caption = "Detected files per severity stratum, two independent cohorts")
knitr::kable(rep_p,
             caption = "Fisher exact test of stratum against detection, within each cohort")
```

```{r fig-replication, fig.width = 8, fig.height = 5.4}
#| echo: false
.rl <- rep_tbl |>
  tidyr::pivot_longer(c(omeprazole, `5-OH`), names_to = "compound", values_to = "detected") |>
  dplyr::mutate(stratum = paste0(stratum, "  (n=", files, ")"),
                compound = factor(ifelse(compound == "5-OH", "5-OH-omeprazole", "omeprazole"),
                                  levels = c("omeprazole", "5-OH-omeprazole")))
ggplot(.rl, aes(reorder(stratum, detected), detected, fill = compound)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.62) +
  geom_text(aes(label = detected), position = position_dodge(width = 0.72),
            hjust = -0.35, size = 3) +
  coord_flip() + facet_wrap(~ cohort, scales = "free", ncol = 1) +
  scale_fill_manual(values = c(omeprazole = "#1b1b1b", `5-OH-omeprazole` = "#9a9a9a"),
                    name = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.16))) +
  labs(x = NULL, y = "files with the compound detected",
       title = "Files with a detection by COVID-19 severity stratum, two independent cohorts") +
  theme_bw(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "grey93", colour = NA),
        strip.text = element_text(face = "bold", size = 10),
        plot.title = element_text(face = "bold", size = 12.5))
```

Both cohorts give the same ordering, and in both the **healthy control stratum
contributes zero detected files of either compound**. That the healthy groups
are exactly empty in two unrelated populations is the part that a single deposit
cannot establish: within one cohort an empty control group is a plausible
accident of size, whereas twice it is the negative control the corpus was
assembled to provide.

The reading is exposure, not pathology. Omeprazole is prescribed heavily in
hospital and intensive care, so what replicates is that the sicker strata are
the medicated strata. The finding is that the pipeline recovers a real,
interpretable prescribing gradient from data deposited for an unrelated purpose,
in two countries at once.

### Does the exposure reading hold up everywhere it can be checked?

Calling these associations exposure rather than pathology is not a hedge, it is
a testable claim: if the pipeline is measuring who was **prescribed**
omeprazole, then the drug should concentrate in strata receiving more medical
care, and should be near absent where none is given. Crucially it should track
*care*, not illness as such, so a cohort stratified by a disease that does not
change prescribing should show nothing. Four independent studies record a
clinical grouping and can be checked.

```{r fig-exposure-sense, fig.width = 8.4, fig.height = 5.4}
#| echo: false
.cl <- pm |> dplyr::filter(!is.na(clin), nzchar(clin))
.cl$unit <- ifelse(is.na(.cl$subj), paste(.cl$study, .cl$fk),
                   paste(.cl$study, .cl$subj, .cl$clin))
.cs <- .cl |>
  dplyr::group_by(unit) |>
  dplyr::summarise(study = dplyr::first(study), clin = dplyr::first(clin),
                   ms1 = any(ms1), .groups = "drop") |>
  dplyr::group_by(study, clin) |>
  dplyr::summarise(n = dplyr::n(), det = sum(ms1), rate = mean(ms1), .groups = "drop") |>
  dplyr::group_by(study) |>
  dplyr::filter(dplyr::n() >= 2, sum(n) >= 20) |>
  dplyr::ungroup()
.cs$label <- paste0(.cs$clin, "  (", .cs$n, ")")
ggplot(.cs, aes(rate, reorder(label, rate))) +
  geom_col(width = 0.6, fill = "grey25") +
  geom_text(aes(label = sprintf("%d/%d", det, n)), hjust = -0.18, size = 3.1) +
  facet_wrap(~ study, scales = "free_y", ncol = 2) +
  scale_x_continuous(labels = scales::percent,
                     expand = expansion(mult = c(0, 0.22))) +
  labs(x = "subjects with a detection", y = NULL,
       title = "Subjects with a detection by clinical stratum, per study") +
  theme_bw(base_size = 11) +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "grey93", colour = NA),
        strip.text = element_text(face = "bold", size = 10),
        plot.title = element_text(face = "bold", size = 12.5))
```

The gradient is **not** uniform across the four, and that is the informative
part.

It is strong in exactly the two cohorts whose strata differ in **intensity of
hospital care**: MTBLS1866 runs from 78% in patients who survived intensive care
down to 0 of 26 healthy, and MTBLS11656 from severe disease down to 0 of 45 in
both the mild and the healthy arms. It is absent in the Chagas cohort, where the
strata are infection status in what is largely an outpatient population, and
where infected (11 of 72) and healthy (1 of 6) sit at the same rate. ST002722
has 6 subjects in its second arm and cannot say anything either way.

That pattern is what makes the exposure reading credible rather than convenient.
If the pipeline were picking up something generic about being unwell, Chagas
infection should have moved it too. It does not, because Chagas outpatients are
not receiving stress-ulcer prophylaxis, whereas intensive-care patients very
often are. The signal tracks **care intensity, not illness**, which is precisely
what a prescribing signal should do and what a disease-mechanism signal should
not.

Stating it as exposure is therefore a **strength of the result, not a caveat on
it**. It would be far more worrying to find omeprazole distributed at random
across clinical strata, or concentrated in the healthy arms. The claim is that
the readout works, and this is the evidence that it does; the claim is not that
COVID-19 causes omeprazole to appear in plasma.

## Pooling presence across studies: a random-effects meta-analysis

The scan above asks what each deposit can answer on its own. This asks the
stronger question: measure a contrast **inside** each study, then pool the
study-level effects. Nothing crosses a study boundary except a log odds ratio,
so the response-factor offset and the isomer are both irrelevant, and
between-study disagreement becomes a measured quantity (`I2`) instead of an
assumption.

The unit is the one defined above: one subject per contrast level, with deposits
that record no subject flagged rather than silently treated as independent.
Because the same question can be put to the MS2 baseline, this is also the
sharpest available test of what propagation buys at the corpus level.

```{r pool-helpers}
# DerSimonian-Laird random-effects pooling of log odds ratios.
dl_pool <- function(yi, vi) {
  wf <- 1 / vi; mf <- sum(wf * yi) / sum(wf)
  Q  <- sum(wf * (yi - mf)^2); df <- length(yi) - 1
  C  <- sum(wf) - sum(wf^2) / sum(wf)
  tau2 <- max(0, (Q - df) / C)
  w <- 1 / (vi + tau2); m <- sum(w * yi) / sum(w); se <- sqrt(1 / sum(w))
  list(OR = exp(m), lo = exp(m - 1.96 * se), hi = exp(m + 1.96 * se),
       p = 2 * pnorm(-abs(m / se)), I2 = max(0, 100 * (Q - df) / Q), k = length(yi))
}
# Haldane 0.5 correction only when a cell is empty, which it often is here.
or_2x2 <- function(grp, det) {
  a <- sum(grp & det); b <- sum(grp & !det); c <- sum(!grp & det); e <- sum(!grp & !det)
  if (min(a, b, c, e) == 0) { a <- a + .5; b <- b + .5; c <- c + .5; e <- e + .5 }
  c(yi = log((a * e) / (b * c)), vi = 1/a + 1/b + 1/c + 1/e)
}
fmt_pool <- function(p) sprintf("%.2f (%.2f to %.2f), p = %.3g, I2 = %.0f%%, k = %d",
                                p$OR, p$lo, p$hi, p$p, p$I2, p$k)
```

### The unexposed-by-design stratum

Three studies contain a stratum that should carry no drug: the healthy arms of
the two COVID cohorts, and the uninfected arm of the Chagas serum study. Each
study's reference level is taken from **its own vocabulary**, never from a name
match, which would otherwise sweep in QC and blank injections and invert the
result.

```{r pool-reference}
# The Chagas deposit carries TWO unexposed labels, "uninfected" and "healthy",
# as separate strata. Taking only one of them would be a choice made after
# seeing the answer, so both are used. It is also the conservative direction:
# the "healthy" arm contains the single detection, which weakens the contrast
# rather than helping it.
REF <- list("MTBLS1866"    = "non-COVID-19 healthy",
            "MTBLS11656"   = "Healthy Control",
            "Chagas serum" = c("uninfected", "healthy"))

ref_rows <- function(role) {
  dplyr::bind_rows(lapply(names(REF), function(s) {
    x  <- pm[pm$study == s, , drop = FALSE]
    a  <- units_for(x, x$clin, role)
    if (is.null(a)) return(NULL)
    ref <- a$g %in% REF[[s]]
    if (sum(ref) < 5 || sum(!ref) < 5) return(NULL)
    o <- or_2x2(ref, a[[role]])
    data.frame(study = s, reference = paste(REF[[s]], collapse = " + "),
               ref_units = sum(ref), ref_detected = sum(ref & a[[role]]),
               other_units = sum(!ref), other_detected = sum(!ref & a[[role]]),
               yi = o[["yi"]], vi = o[["vi"]])
  }))
}
ref_ms1 <- ref_rows("ms1"); ref_ms2 <- ref_rows("ms2")
pr1 <- dl_pool(ref_ms1$yi, ref_ms1$vi); pr2 <- dl_pool(ref_ms2$yi, ref_ms2$vi)

knitr::kable(ref_ms1[, 1:6],
             caption = "Detections in the unexposed-by-design stratum, per subject")
knitr::kable(data.frame(method = c("MASST MS2 only", "MS1 propagated"),
                        pooled_OR = c(fmt_pool(pr2), fmt_pool(pr1))),
             caption = "Pooled odds of detection in the reference stratum versus the rest")
```

**`r sum(ref_ms1$ref_detected)` detection in `r sum(ref_ms1$ref_units)`
reference subjects across three independent studies**, against
`r sum(ref_ms1$other_detected)` in the rest, with `I2 = 0%`. The two COVID
cohorts contribute exact zeros; the single detection is in the Chagas healthy
arm, which is one subject of six and is exactly the kind of incidental
prescription a general-population arm is expected to contain.

The MS2 row is the point. On the MASST baseline the same three studies pool to p
= `r sprintf("%.3f", pr2$p)`, which clears no threshold. **The contrast becomes
significant only once propagation supplies the detections** (p =
`r sprintf("%.4f", pr1$p)`), and it does so without putting a single false
positive into the strata where the drug should be absent. This is the
corpus-level analogue of the pre-dose negative control in `04`.

### Sex: does propagation move an estimate, or only sharpen it?

Sex is the one variable testable in five independent studies under **both**
methods, which makes it the place to ask whether adding propagated detections
biases a pooled effect rather than merely narrowing it.

```{r pool-sex}
sex_pool <- function(role) {
  tb <- dplyr::bind_rows(lapply(unique(pm$study), function(s) {
    x <- pm[pm$study == s, , drop = FALSE]
    a <- units_for(x, x$sex2, role); if (is.null(a)) return(NULL)
    a <- gate(a); if (is.null(a)) return(NULL)
    if (sum(a[[role]]) < 3 || sum(!a[[role]]) < 3) return(NULL)
    o <- or_2x2(a$g == "female", a[[role]])
    data.frame(study = s, design = attr(units_for(x, x$sex2, role), "design"),
               units = nrow(a), detected = sum(a[[role]]),
               yi = o[["yi"]], vi = o[["vi"]])
  }))
  if (nrow(tb) < 2) return(NULL)
  list(tb = tb, p = dl_pool(tb$yi, tb$vi))
}
sx2 <- sex_pool("ms2"); sx1 <- sex_pool("ms1")
knitr::kable(sx1$tb[, c("study", "design", "units", "detected")],
             caption = "Studies contributing to the sex contrast, per subject")
knitr::kable(data.frame(
  method = c("MASST MS2 only", "MS1 propagated"),
  pooled_OR_female_vs_male = c(fmt_pool(sx2$p), fmt_pool(sx1$p)),
  CI_width_log_odds = round(c(log(sx2$p$hi) - log(sx2$p$lo),
                              log(sx1$p$hi) - log(sx1$p$lo)), 2)),
  caption = "Sex contrast pooled over five independent studies under each method")
```

Both are null and both have `I2 = 0%`, so propagation does not manufacture an
association. What it changes is precision: the interval narrows from
`r sprintf("%.2f", log(sx2$p$hi) - log(sx2$p$lo))` to
`r sprintf("%.2f", log(sx1$p$hi) - log(sx1$p$lo))` log-odds, which turns an
interval spanning roughly a `r sprintf("%.0f", sx2$p$hi / sx2$p$lo)`-fold range,
and therefore compatible with almost anything, into a usefully bounded null.

This is the check a match-between-runs method owes before its extra detections
are trusted in any association analysis, and it is only possible because sex
happens to be recorded in five studies. It is not itself a biological result.

### Age: where propagation prevents a claim rather than enabling one

```{r pool-age}
age_pool <- function(role) {
  tb <- dplyr::bind_rows(lapply(unique(pm$study), function(s) {
    x <- pm[pm$study == s & is.finite(pm$agen), , drop = FALSE]
    if (!nrow(x)) return(NULL)
    # Collapse to people where the deposit supports it; where it records no
    # subject, the file is the unit and the estimate is labelled as resting on
    # an unknown design.
    known <- subject_resolved(x$subj, sibling_ids = unlist(sibling_id_map[unique(x$dataset_id)], use.names = FALSE))
    a <- data.frame(u = if (known) x$subj else x$fk, agen = x$agen, y = x[[role]]) |>
      dplyr::group_by(u) |>
      dplyr::summarise(agen = dplyr::first(agen), y = any(y), .groups = "drop")
    if (nrow(a) < 20 || sum(a$y) < 5 || sum(!a$y) < 5) return(NULL)
    if (diff(range(a$agen)) < 10) return(NULL)
    m <- try(glm(y ~ I(agen / 10), data = a, family = binomial))
    if (inherits(m, "try-error")) return(NULL)
    co <- summary(m)$coefficients
    if (!"I(agen/10)" %in% rownames(co)) return(NULL)
    data.frame(study = s, design = if (known) "subject-resolved" else "design unknown",
               subjects = nrow(a),
               age_range = paste0(round(min(a$agen)), " to ", round(max(a$agen))),
               OR_per_decade = round(exp(co["I(agen/10)", "Estimate"]), 2),
               yi = co["I(agen/10)", "Estimate"], vi = co["I(agen/10)", "Std. Error"]^2)
  }))
  if (nrow(tb) < 2) return(NULL)
  list(tb = tb, p = dl_pool(tb$yi, tb$vi))
}
ag2 <- age_pool("ms2"); ag1 <- age_pool("ms1")
knitr::kable(ag1$tb[, c("study", "design", "subjects", "age_range", "OR_per_decade")],
             caption = "The age slopes, with the cohorts that produced them")
knitr::kable(data.frame(method = c("MASST MS2 only", "MS1 propagated"),
                        pooled_OR_per_decade = c(fmt_pool(ag2$p), fmt_pool(ag1$p))),
             caption = "Detection against age, logistic within each study then pooled")
```

```{r fig-pooled-forest, fig.width = 8.6, fig.height = 4.4}
#| echo: false
.row <- function(p, contrast, method) data.frame(
  contrast = contrast, method = method, OR = p$OR, lo = p$lo, hi = p$hi,
  note = sprintf("k=%d  I2=%.0f%%  p=%s", p$k, p$I2,
                 ifelse(p$p < 0.001, format(signif(p$p, 1), scientific = TRUE),
                        format(round(p$p, 3), nsmall = 3))))
.f <- rbind(
  .row(pr2, "Unexposed stratum\nvs the rest", "MASST MS2 only"),
  .row(pr1, "Unexposed stratum\nvs the rest", "MS1 propagated"),
  .row(sx2$p, "Female vs male", "MASST MS2 only"),
  .row(sx1$p, "Female vs male", "MS1 propagated"),
  .row(ag2$p, "Per decade of age", "MASST MS2 only"),
  .row(ag1$p, "Per decade of age", "MS1 propagated"))
.f$contrast <- factor(.f$contrast, levels = rev(unique(.f$contrast)))
ggplot(.f, aes(OR, contrast, colour = method, shape = method)) +
  geom_vline(xintercept = 1, linetype = 2, colour = "grey55") +
  geom_errorbar(aes(xmin = lo, xmax = hi), width = 0.16,
                position = position_dodge(width = 0.62), linewidth = 0.6) +
  geom_point(size = 2.6, position = position_dodge(width = 0.62)) +
  geom_text(aes(label = note), position = position_dodge(width = 0.62),
            hjust = 0, vjust = -1.1, size = 2.8, show.legend = FALSE) +
  scale_x_continuous(trans = "log10", breaks = c(0.02, 0.1, 0.5, 1, 2, 5)) +
  scale_colour_manual(values = c("MS1 propagated" = "#111111",
                                 "MASST MS2 only" = "#9a9a9a"), name = NULL) +
  scale_shape_manual(values = c("MS1 propagated" = 16, "MASST MS2 only" = 17), name = NULL) +
  labs(x = "pooled odds ratio (log scale), random effects", y = NULL,
       title = "Random-effects pooled odds ratio per contrast, MASST MS2 and MS1-propagated") +
  theme_bw(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 12.5))
```

Read this one carefully, because the naive version is wrong. Restricted to the
MASST detections only **`r ag2$p$k` studies** can fit an age term
(`r paste(ag2$tb$study, collapse = " and ")`, `r sum(ag2$tb$subjects)` units
between them), they happen to agree, and the pool returns a significant
`r sprintf("%.2f", ag2$p$OR)`-fold increase per decade, which flatters a
well-known prescribing pattern. Propagation lets **`r ag1$p$k` studies** fit it
and the pooled effect collapses to non-significance with `I2 =
`r sprintf("%.0f%%", ag1$p$I2)``.

**At `r ag2$p$k` studies, `I2` cannot detect heterogeneity at all.** The MS2
answer is not cleaner than the MS1 answer; it is blind. That is the clearest
statement in this notebook of what an MS2-only corpus analysis costs: not merely
fewer detections, but too few studies to see that the studies disagree.

#### Why they disagree, and why pooling harder cannot fix it

The heterogeneity is structural. **Every study that can fit a slope is a
clinical cohort**, and age inside a hospital population is a proxy for case-mix:
a Chagas infection cohort, a spot-urine cohort from a Glasgow clinic, a cohort
of older adults with chronic illness. Each slope measures who in *that*
population was prescribed a proton-pump inhibitor, which is a different quantity
in each. And the **age ranges barely overlap**, as the table shows: a per-decade
linear slope fitted to a cohort spanning the thirties and one fitted to a cohort
spanning the sixties describe different segments of a relationship there is no
reason to think is linear. They were never estimating a common parameter, so the
pooling was ill-posed before the data arrived.

#### The clean version of the question cannot be asked here

The way to remove the case-mix confound is to ask only about people who are not
patients. The corpus does not support it.

```{r pool-age-healthy}
HEALTHY <- list("MTBLS1866"      = "non-COVID-19 healthy",
                "MTBLS11656"     = "Healthy Control",
                "Chagas serum"   = "uninfected",
                "MSV000082433"   = "healthy",
                "MSV000091520"   = "healthy",
                "Jarmusch probe" = "healthy")
healthy_tbl <- dplyr::bind_rows(lapply(names(HEALTHY), function(s) {
  x <- pm[pm$study == s, , drop = FALSE]
  if (!nrow(x)) return(NULL)
  v <- nz(x$health_status)
  h <- !is.na(v) & tolower(v) == tolower(HEALTHY[[s]])
  if (!any(h)) return(NULL)
  y <- x[h, , drop = FALSE]
  u <- if (!all(is.na(y$subj))) dplyr::n_distinct(y$subj) else nrow(y)
  data.frame(study = s, healthy_units = u,
             age = if (all(is.na(y$agen))) "not recorded" else
                   paste0(round(min(y$agen, na.rm = TRUE)), " to ",
                          round(max(y$agen, na.rm = TRUE))),
             MS1_detected = sum(y$ms1), MS2_detected = sum(y$ms2))
}))
knitr::kable(healthy_tbl, caption = "Healthy strata: do any have age and events together?")
```

Six studies carry a healthy stratum. Three record no age at all. Of those that
do, two contribute almost no events, and the probe cohort's detections exist
only because every subject was **dosed by design**, which answers a different
question and records no age anyway. That leaves roughly nine detections in
healthy people with a recorded age, and five of those are MSV000082433's, which
the closing subsection shows are metabolite-without-parent in non-blood sites
and most likely the isomer.

The negative is worth stating positively: **background exposure in non-clinical
populations is close to zero here**, which converges with the unexposed-stratum
result above and is precisely why age is unaskable. Omeprazole in this corpus
marks clinical care, so any age effect is inseparable from the setting that
produced it. We report no pooled age association.

### What does not pool

- **Body site.** Only one study has two body sites with enough subjects, so the
  biofluid view below is a pooled description and not a meta-analysis.
- **MSV000082433.** It contributes five MS1 detections MASST does not have, but
  all five are **metabolite without parent**, in five different non-blood sites
  (nasal cavity, vagina, oral cavity, arm skin), in healthy women aged 18 to 33.
  That is the signature of the isobaric interference described in `03`, not of
  drug exposure, and the deposit is not counted as a propagation success
  anywhere here.

## Prevalence by biofluid: the reverse-metabolomics biology view

The heatmap above is organised by cohort to show the method's value (MS2 vs
MS1). The reverse-metabolomics papers instead organise prevalence by a **biology
axis** (tissue / biofluid / health phenotype); Gentry et al. (Nature 2024, Fig
2b-d) is the canonical example, a molecule-by-tissue and
molecule-by-health-condition prevalence heatmap. We reproduce that view here,
MS1-propagated: pooling across cohorts by body site, the detection prevalence of
each compound.

Since this is a cross-repository comparison it is computed **per specimen**, on
the same collapsed table the meta-analysis above uses. A file-level version
would let the probe cohort's replicate injections dominate the plasma row simply
by having been injected more often, which is a property of its protocol rather
than of plasma.

This pooling deliberately crosses cohorts, so exposure and body site are partly
confounded (a biofluid seen only in a drug-probe cohort will read high for
reasons of study design, not physiology). It is therefore hypothesis-generating,
in exactly the spirit the field uses it: the expected signal is that the parent
concentrates in blood/plasma while the metabolite persists in excretory matrices
(urine, feces), which if it appears is the pharmacology falling out of public
data with no targeted assay.

```{r prevalence-by-biofluid, fig.width = 7.5, fig.height = 5}
# Same unit rule as everywhere else, built explicitly here because this table
# also needs the study, to show how many cohorts stand behind each matrix.
bysite <- pm |>
  dplyr::filter(!is.na(site)) |>
  dplyr::mutate(.u = ifelse(is.na(subj), paste(study, fk), paste(study, subj, site))) |>
  dplyr::group_by(.u) |>
  dplyr::summarise(site = dplyr::first(site), study = dplyr::first(study),
                   par_ms1 = any(par_ms1), met_ms1 = any(met_ms1), .groups = "drop") |>
  dplyr::group_by(site) |>
  dplyr::summarise(n_specimens = dplyr::n(),
                   n_studies = dplyr::n_distinct(study),
                   omeprazole = mean(par_ms1), `5-OH-omeprazole` = mean(met_ms1),
                   .groups = "drop") |>
  dplyr::filter(n_specimens >= 5) |>
  tidyr::pivot_longer(c(omeprazole, `5-OH-omeprazole`),
                      names_to = "compound", values_to = "rate")

ggplot(bysite, aes(x = compound,
                   y = reorder(paste0(site, "  (", n_specimens, " specimens, ",
                                      n_studies, ifelse(n_studies == 1, " study)", " studies)")),
                                rate, sum),
                   fill = rate)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = scales::percent(rate, accuracy = 1),
                colour = rate > 0.45), size = 3, fontface = "bold") +
  scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey15"), guide = "none") +
  scale_fill_gradient(low = "grey95", high = "#762a83",
                      labels = scales::percent, name = "MS1 detection\nrate") +
  labs(x = NULL, y = NULL,
       title = "MS1-propagated detection prevalence by body site, per specimen") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

The study count on each row is the part to read first. A body site carried by
one study is that study's exposure pattern wearing a biofluid label, and only
the rows backed by several cohorts describe anything about the matrix itself.

## The ratio across cohorts: measuring the offset instead of assuming it

Prevalence is a per-compound presence question, and presence rests on the
parent, which has no isobaric twin. The parent/metabolite **ratio** is a
different kind of object: a quantity whose absolute value carries a per-method
response-factor offset. The standing position in this project has been that the
offset is unknown, so ratios must not be pooled across cohorts. That position is
defensible but untested, and the corpus turns out to contain the experiment that
tests it.

### Trying to measure the offset directly, and failing to pin it down

MSV000092754 and MSV000092757 are the **same subjects and the same serum**, run
on HILIC and on reverse phase. Biology is held constant by construction, so
whatever separates the two deposits is method. This is the only place in the
corpus where the offset can be observed rather than assumed, so it is worth
asking what it gives.

```{r ratio-offset}
ratio_pf <- pf |>
  dplyr::filter(detection_tier == "both", is.finite(log2_ratio))

# One study under two accessions must not count twice; two accessions of one
# subject panel on two columns must not be merged. Both rules are established
# above.
ratio_pf$study <- dplyr::case_when(
  grepl(paste(DUP_STUDY, collapse = "|"), ratio_pf$dataset_id) ~ "Jarmusch probe (082493 + 084008)",
  grepl("MSV000092754|MSV000092757", ratio_pf$dataset_id)      ~ "Chagas serum (092754 + 092757)",
  TRUE ~ short_ds(ratio_pf$dataset_id))

chagas <- ratio_pf |>
  dplyr::filter(grepl("MSV000092754|MSV000092757", dataset_id)) |>
  dplyr::mutate(chromatography = ifelse(grepl("092754", dataset_id), "HILIC", "RP"))

chagas_grp <- chagas |>
  dplyr::group_by(chromatography) |>
  dplyr::summarise(files = dplyr::n(), subjects = dplyr::n_distinct(subject_id),
                   median_log2 = median(log2_ratio), .groups = "drop")
OFFSET_GRP <- abs(diff(chagas_grp$median_log2))

# The paired form is the design that actually holds biology fixed: the same
# person on both columns. It is also far smaller, because few subjects yield a
# ratio twice.
chagas_pair <- chagas |>
  dplyr::filter(!is.na(subject_id), nzchar(as.character(subject_id))) |>
  dplyr::group_by(subject_id, chromatography) |>
  dplyr::summarise(m = median(log2_ratio), .groups = "drop") |>
  tidyr::pivot_wider(names_from = chromatography, values_from = m) |>
  dplyr::filter(!is.na(HILIC), !is.na(RP))

OFFSET_PAIR <- median(chagas_pair$RP - chagas_pair$HILIC)
# A median of exactly zero arrives as -1e-16 and would print as "-0.00".
OFFSET_PAIR_TXT <- sprintf("%+.2f", ifelse(abs(OFFSET_PAIR) < 0.005, 0, OFFSET_PAIR))
cp <- wilcox.test(chagas_pair$RP, chagas_pair$HILIC, paired = TRUE)

knitr::kable(chagas_grp, digits = 2,
             caption = "Same subjects, same serum, two chromatographies (group level)")
knitr::kable(chagas_pair |> dplyr::mutate(difference = RP - HILIC), digits = 2,
             caption = "The same comparison, paired within subject")
```

Compared at the group level the two columns differ by
`r sprintf("%.2f", OFFSET_GRP)` log2, about
`r sprintf("%.1f", 2^OFFSET_GRP)`-fold, which looks like a reassuringly small
offset. The paired version is the design that actually holds biology fixed, and
it does not support that reading: only **`r nrow(chagas_pair)` subjects** yield
a ratio on both columns, the median difference is `r OFFSET_PAIR_TXT` log2,
reverse phase is higher in `r sum(chagas_pair$RP > chagas_pair$HILIC)` of
`r nrow(chagas_pair)`, and p = `r sprintf("%.2f", cp$p.value)`. The per-subject
differences run from
`r sprintf("%.1f", min(chagas_pair$RP - chagas_pair$HILIC))` to
`r sprintf("%+.1f", max(chagas_pair$RP - chagas_pair$HILIC))` log2.

So the honest conclusion is not "the offset is 1.5-fold". It is that the one
direct measurement available puts the **central** offset near zero while leaving
it very imprecisely determined, on six subjects. That is not a licence to pool
absolute ratios across deposits. It does mean the interesting cross-cohort ratio
work has to come from designs where the offset cannot enter at all, which is
what follows.

### The matrix contrast that carries no offset at all

The cleanest comparison in the corpus needs no correction whatsoever. Within the
probe study the **same 14 subjects** gave plasma and urine, on the same
instrument, in the same run. Any response factor is shared and cancels exactly
in a paired test.

```{r ratio-paired-matrix}
# Same canonical vocabulary as Part A, then restricted to the matrices that can
# actually carry a ratio.
ratio_pf$site <- canon_site(ratio_pf$body_site)
ratio_pf$site[!ratio_pf$site %in% c("plasma", "serum", "urine", "faeces")] <- NA

paired <- ratio_pf |>
  dplyr::filter(study == "Jarmusch probe (082493 + 084008)",
                site %in% c("plasma", "urine")) |>
  dplyr::group_by(subject_id, site) |>
  dplyr::summarise(m = median(log2_ratio), .groups = "drop") |>
  tidyr::pivot_wider(names_from = site, values_from = m) |>
  dplyr::filter(!is.na(plasma), !is.na(urine))

pw   <- wilcox.test(paired$plasma, paired$urine, paired = TRUE)
gap  <- median(paired$plasma - paired$urine)
```

Across `r nrow(paired)` subjects the plasma ratio exceeds the urine ratio in
**`r sum(paired$plasma > paired$urine)` of `r nrow(paired)`**, by a median of
`r sprintf("%.2f", gap)` log2 (`r sprintf("%.0f", 2^gap)`-fold), paired Wilcoxon
p = `r sprintf("%.2g", pw$p.value)`. The parent dominates in blood and the
metabolite dominates in urine, which is the textbook disposition of an
extensively metabolised drug, recovered from public data with no targeted assay
and with the offset arithmetically removed.

### Does the ordering survive across independent studies?

Now the harder question: does the same ordering appear when the cohorts are not
the same people? Only study-by-matrix cells backed by at least five subjects are
interpretable; the rest rest on a single subject and are listed only for
completeness.

```{r ratio-by-study-matrix}
cell <- ratio_pf |>
  dplyr::filter(!is.na(site)) |>
  dplyr::group_by(site, study) |>
  dplyr::summarise(files = dplyr::n(), subjects = dplyr::n_distinct(subject_id),
                   median_log2 = median(log2_ratio), .groups = "drop") |>
  dplyr::mutate(usable = subjects >= 5) |>
  dplyr::arrange(dplyr::desc(usable), median_log2)

knitr::kable(cell, digits = 2,
             caption = "log2(parent / metabolite) per body site and independent study")

vc <- ratio_pf |> dplyr::filter(site %in% c("plasma", "serum", "urine"))
av <- anova(lm(log2_ratio ~ site + factor(study), data = vc))
ss <- av[["Sum Sq"]]
var_tbl <- data.frame(term = c("body site", "study, given body site", "residual"),
                      pct_variance = round(100 * ss / sum(ss), 1))

# An interval on the split, resampling STUDIES rather than files. The two
# components are properties of a corpus of a few studies, not of its files, and a
# file-level bootstrap would report a precision the design does not have. A study
# drawn twice enters as two studies, or the resample is a subsample.
set.seed(1)
.st <- unique(vc$study)
.vboot <- replicate(2000, {
  d <- sample(.st, replace = TRUE)
  z <- do.call(rbind, lapply(seq_along(d), function(j) {
    x <- vc[vc$study == d[j], , drop = FALSE]
    if (nrow(x)) x$study <- paste0(j, "|", x$study)
    x
  }))
  if (dplyr::n_distinct(z$site) < 2 || dplyr::n_distinct(z$study) < 2)
    return(rep(NA_real_, 3))
  a2 <- anova(lm(log2_ratio ~ site + factor(study), data = z))
  if (nrow(a2) < 3) return(rep(NA_real_, 3))
  s2 <- a2[["Sum Sq"]]
  100 * s2[seq_len(3)] / sum(s2)
})
var_tbl$ci_95 <- sprintf("%.1f to %.1f",
                         apply(.vboot, 1, quantile, 0.025, na.rm = TRUE),
                         apply(.vboot, 1, quantile, 0.975, na.rm = TRUE))
var_tbl$usable_resamples <- sum(!is.na(.vboot[1, ]))
knitr::kable(var_tbl, caption = paste(
  "Variance in the ratio attributable to matrix versus study.",
  "Intervals are a bootstrap over studies, the unit the components describe."))
```

The four usable cells order urine below plasma below serum, and the plasma cells
from two unrelated studies (the probe cohort and MTBLS1866) bracket zero while
the serum cell sits well above it. That ordering is suggestive but it is **not**
a test: each cell still carries its own offset, and the serum column is one
study. The variance decomposition is the part that does carry weight, because it
does not depend on any single cell being correctly calibrated: body site
accounts for about four times more of the ratio's variance than study identity
does. The bootstrap intervals overlap, so the honest reading is that the study
term is small rather than absent -- but whatever the per-deposit offsets are,
they are not what is organising this data. Three quarters of the variance sits
within cells, which is where a biological signal would have to live and where
this design cannot look for it.

### What this does and does not license

The one thing here that stands on its own is the **paired matrix contrast**,
because the offset is arithmetically absent rather than assumed small. The
variance decomposition supports it. The cross-study ordering is consistent with
it. Read together, the ratio gives a **disposition readout**: parent-dominant in
blood, metabolite-dominant in urine. Part B continues to use within-study
contrasts, where no offset can enter at all.

Four things this does not license, three of which looked like results until they
were checked:

- **Pooling absolute ratios across deposits, or ranking deposits by ratio.** The
  offset was not successfully measured above, only bounded loosely. The original
  constraint stands.
- **Dosed probe versus incidental exposure.** Pooled, the two differ strongly
  (medians 0.52 and 2.31, Wilcoxon p = 3e-11). But the entire "dosed" arm is
  MSV000082493 and MSV000084008, which is **one study under two accessions**.
  The comparison is one study against eight, and the apparent significance is a
  pseudo-replicated single cohort. It is not a finding.
- **Reverse phase versus HILIC.** Pooled, these also differ strongly (p =
  2e-11), but HILIC here is almost entirely serum and reverse phase almost
  entirely plasma, so that test measures matrix, not method. The controlled
  version, the same subjects on both columns above, is null and points the other
  way.
- **The metabolite channel remains isomer-exposed.** Omeprazole sulfone shares
  the metabolite's formula and precursor mass. The urine result is the ratio's
  largest effect and sits in the channel most vulnerable to it, so it is
  reported as disposition consistent with pharmacology, not as a quantified
  metabolite concentration.

# Every field, both read-outs

The scan above asks whether propagation *enables* a contrast, comparing what MS2
could test against what MS1 can. This section asks the different question the
workflow was built for: given everything the corpus records, what does it
actually show. Every deposit is crossed with every field that divides it, and
each combination is tested twice — once on presence, once on the ratio.

Two joins have to be right or the whole scan is noise. `metrics_*` carries
cache-prefixed file names, so the key needs `norm_file_key(x, bio_files)`; the
bare form leaves the accession on MetaboLights names and the join silently
returns almost nothing. And `load_metadata()` can describe more files than the
deposit contributed — MSV000091520 lists 2,732 rows against 2,144 extracted
files — so the metadata is restricted to the extraction set before anything is
counted.

```{r scan-build}
#| echo: false
BIO <- readRDS("artifacts/bio_files_per_ds.rds")
SFIELDS <- c("sex", "age", "health_status", "body_site", "country", "treatment")

scan_pf <- function(ds) {
  r <- readRDS(sprintf("artifacts/per_dataset/%s.rds", ds))
  m <- merge(as.data.frame(r$metrics_par), as.data.frame(r$metrics_met),
             by = "file", all = TRUE)
  pd <- m$par_detected %in% TRUE; md <- m$met_detected %in% TRUE
  d <- data.frame(k = norm_file_key(m$file, BIO[[ds]]), par = pd, met = md,
                  l2 = ifelse(pd & md, log2(m$par_auc / m$met_auc), NA_real_))
  meta <- tryCatch(load_metadata(ds), error = function(e) NULL)
  if (is.null(meta) || !nrow(meta)) meta <- as.data.frame(r$sample_tbl)
  meta$k <- norm_file_key(meta$file, BIO[[ds]])
  meta <- meta[meta$k %in% norm_file_key(BIO[[ds]], BIO[[ds]]), ]
  meta <- meta[!duplicated(meta$k), ]
  i <- match(meta$k, d$k)
  meta$par <- d$par[i] %in% TRUE; meta$met <- d$met[i] %in% TRUE
  meta$l2 <- d$l2[i]; meta$ds <- ds
  meta
}
scan_all <- bind_rows(lapply(names(BIO), function(x)
  tryCatch(scan_pf(x), error = function(e) {
    message("scan_pf failed for ", x, ": ", conditionMessage(e)); NULL })))
stopifnot(nrow(scan_all) > 0)
scan_all <- scan_all[!(scan_all$ds == "MSV000082493" &
                       grepl("plasma|blood", tolower(scan_all$body_site))), ]
cat(nrow(scan_all), "files |", n_distinct(scan_all$ds), "deposits | detected: parent",
    sum(scan_all$par), "metabolite", sum(scan_all$met), "both",
    sum(scan_all$par & scan_all$met), "\n")
```

## What unit each deposit supports

A contrast needs independent observations, so files have to be collapsed to
people wherever the deposit records who they came from. Nine of the twelve do.
What varies is how the identifier behaves, and the differences are not cosmetic:
it may repeat, because the person was sampled more than once; it may appear
exactly once, because the study drew one sample per participant; or it may cover
only part of the deposit. Only the last of those is a real obstacle. A test that
asked merely whether the identifier repeats would reject six of the nine.

```{r scan-units}
#| echo: false
unit_of <- function(x) {
  sid <- if ("subject_id" %in% names(x)) x$subject_id else rep(NA, nrow(x))
  sib <- unlist(sibling_id_map[unique(x$ds)], use.names = FALSE)
  ok <- !is.na(sid) & nzchar(trimws(as.character(sid)))
  lab <- sum(ok); nsub <- n_distinct(sid[ok]); reps <- if (nsub) lab / nsub else NA
  if (!lab) list("file", "no subject id recorded")
  else if (lab < nrow(x))
    list(if (subject_resolved(sid, sibling_ids = sib)) "subject (partial)" else "file",
         sprintf("id on %d of %d files, %d subjects", lab, nrow(x), nsub))
  else if (nsub == nrow(x)) list("subject", sprintf("%d subjects, one sample each", nsub))
  else list("subject", {
    # Median and range, not the mean: a mean of "5.1 files each" describes no
    # subject, and it hid that MSV000084008 runs 9 to 36 files per person.
    .k <- as.integer(table(sid[ok]))
    sprintf("%d subjects, %d files each (%d-%d)", nsub,
            as.integer(stats::median(.k)), min(.k), max(.k))
  })
}
units <- bind_rows(lapply(split(scan_all, scan_all$ds), function(x) {
  u <- unit_of(x)
  data.frame(deposit = short_ds(x$ds[1]), files = nrow(x),
             unit = u[[1]], reason = u[[2]])
}))
knitr::kable(units, row.names = FALSE,
             caption = "The analysis unit each deposit can support, and why")
```

Three deposits support repeated measures on the same people; four are
cross-sectional, one sample per participant; two label part of the deposit and
are corroborated by their sibling; three record nobody. Only that last group is
analysed at file level, and those results are exploratory in the strict sense —
the files are not independent observations and a p-value on them overstates the
evidence. They are still worth computing, because a contrast that fails even at
file level cannot be rescued by a stricter unit.

## The scan

Presence is tested with Fisher's exact test on categorical fields and logistic
regression on continuous ones; the ratio with a linear model on the files
carrying both compounds. Benjamini-Hochberg control is applied within each
read-out.

```{r scan-run}
#| echo: false
set.seed(1)
scan_rows <- list()
for (ds in unique(scan_all$ds)) {
  x0 <- scan_all[scan_all$ds == ds, ]
  u <- unit_of(x0)
  x <- if (u[[1]] == "subject")
    x0 |> group_by(subject_id) |>
      summarise(par = any(par), met = any(met), l2 = median(l2[is.finite(l2)]),
                across(any_of(SFIELDS), ~ .x[1]), .groups = "drop") |> as.data.frame()
    else x0
  for (f in intersect(SFIELDS, names(x))) {
    v <- x[[f]]; keep <- !is.na(v) & nzchar(trimws(as.character(v)))
    if (sum(keep) < 8) next
    xf <- x[keep, ]; vv <- v[keep]
    num <- !any(is.na(as_num(vv)))
    cont <- num && n_distinct(vv) > 12
    if (!cont) {
      tb <- table(vv); if (length(tb) < 2 || sum(tb >= 3) < 2) next
      vv <- factor(vv)
    } else vv <- as_num(vv)
    for (rl in c("par", "met", "either")) {
      y <- if (rl == "either") (xf$par | xf$met) else xf[[rl]]
      if (n_distinct(y) < 2) next
      pv <- tryCatch(if (cont)
          summary(glm(y ~ vv, family = binomial))$coefficients[2, 4]
        else fisher.test(table(vv, y), simulate.p.value = TRUE, B = 2000)$p.value,
        error = function(e) NA_real_)
      scan_rows[[length(scan_rows) + 1]] <- data.frame(
        deposit = short_ds(ds), unit = u[[1]], field = f, readout = "presence",
        role = rl, n = nrow(xf), det = sum(y), est = NA_real_, p = pv)
    }
    r2 <- xf[is.finite(xf$l2), ]
    if (nrow(r2) >= 6) {
      v2 <- if (cont) as_num(r2[[f]]) else factor(r2[[f]])
      if (cont || n_distinct(v2) >= 2) {
        co <- tryCatch(summary(lm(r2$l2 ~ v2))$coefficients, error = function(e) NULL)
        if (!is.null(co) && nrow(co) >= 2)
          scan_rows[[length(scan_rows) + 1]] <- data.frame(
            deposit = short_ds(ds), unit = u[[1]], field = f, readout = "ratio",
            role = "both", n = nrow(r2), det = nrow(r2), est = co[2, 1], p = co[2, 4])
      }
    }
  }
}
SC <- bind_rows(scan_rows)
SC$q <- NA_real_
for (rd in unique(SC$readout)) {
  i <- SC$readout == rd & !is.na(SC$p)
  SC$q[i] <- p.adjust(SC$p[i], "BH")
}
knitr::kable(SC |> group_by(field, readout) |>
               summarise(tests = n(), significant = sum(q < 0.05, na.rm = TRUE),
                         .groups = "drop"),
             caption = "Tests run and how many survive FDR, by field and read-out")
```

```{r scan-significant}
#| echo: false
sig <- SC[!is.na(SC$q) & SC$q < 0.05, ]
knitr::kable(sig[order(sig$q), c("deposit","unit","field","readout","role","n","det","q")],
             row.names = FALSE, digits = 4,
             caption = "Every contrast surviving FDR 0.05")
```

Two things stand out, and both are replications rather than single findings.
Clinical group separates detection in two unrelated COVID cohorts, MTBLS1866 in
Italy and MTBLS11656 in China, neither of which knows about the other. Age does
the same in two unrelated serum cohorts, the Chagas study and ST002044 in India.
A single deposit showing an effect is a hypothesis; the same effect in a second
cohort collected by different people for a different reason is the argument this
workflow exists to make.

The negative is as informative. **No ratio contrast survives FDR anywhere in the
corpus**, against presence contrasts that do — the quantitative read-out
validated in `04` does not travel, exactly as the body-site and cross-study
analyses predict. And sex is null in every deposit that records it.


# Part B: The quantitative strand, and its limits

Part A is the field's prevalence analysis, made more sensitive. Part B is the
piece the field does not have: turning the propagated detections into a
quantitative, error-controlled cross-cohort analysis, borrowing directly from
how proteomics handles the identical problem.

## The within-study contrast (the quantitative strand)

Where a cohort supports it, the within-study contrast (defined above: the offset
cancels in a within-cohort difference) is the quantitative unit that *could* be
pooled. We compute it per cohort here; the pooling itself is deferred for the
reasons in the final section.

## Defining the contrast

For each dataset we restrict to samples where both compounds were detected, pick
the grouping variable the dataset can actually support, collapse it to two
levels ("case" vs "reference"), and fit a linear model of `log2_ratio` on that
grouping. The coefficient is the contrast.

```{r within-study-fn}
fit_within_study <- function(df) {

  df_both <- df %>% filter(detection_tier == "both", is.finite(log2_ratio))
  if (nrow(df_both) < 6) {
    return(tibble(contrast = NA_character_, estimate = NA_real_,
                  se = NA_real_, n = nrow(df_both), note = "n<6"))
  }

  # Try stratifiers in priority order; first usable one wins. This ordering is
  # the weak point of the whole section: see the final discussion.
  candidates <- c("health_status", "group", "sex", "country", "body_site")
  picked <- NULL
  for (v in candidates) {
    if (!v %in% names(df_both)) next
    counts <- df_both %>% filter(!is.na(.data[[v]])) %>% count(.data[[v]])
    if (nrow(counts) >= 2 && all(counts$n >= 3)) { picked <- v; break }
  }
  if (is.null(picked)) {
    return(tibble(contrast = NA_character_, estimate = NA_real_,
                  se = NA_real_, n = nrow(df_both), note = "no_stratifier"))
  }

  # Collapse to two levels: the smallest group becomes "case", so the estimate
  # reads as "case minus reference".
  df_both <- df_both %>%
    mutate(
      .v    = .data[[picked]],
      .case = .v == names(sort(table(.v)))[1],
      .grp  = factor(ifelse(.case, "case", "ref"), levels = c("ref", "case"))
    )

  m <- lm(log2_ratio ~ .grp, data = df_both)
  coefs <- summary(m)$coefficients
  tibble(
    contrast = paste0(picked, ": ", names(sort(table(df_both$.v)))[1], " vs rest"),
    estimate = coefs[".grpcase", "Estimate"],
    se       = coefs[".grpcase", "Std. Error"],
    n        = nrow(df_both),
    note     = NA_character_
  )
}
```

## The effect-size table

```{r per-study-effects}
effects <- samples %>%
  group_by(dataset_id) %>%
  group_modify(~ fit_within_study(.x)) %>%
  ungroup() %>%
  left_join(datasets %>% select(dataset_id, repository, chromatography,
                                const_country, const_body_site),
            by = "dataset_id") %>%
  arrange(dataset_id)

effects
write_csv(effects, "artifacts/meta_effects.csv")
```

### What each column means

An **effect size** is just a number that says how big a difference is, expressed
so that it can be compared with the same number from another study. Many fields
use a standardised effect size (Cohen's d, Hedges' g), which divides the
difference by its spread so that studies measuring different things on different
scales can be put side by side. **We deliberately do not standardise here.** Our
readout is already on a common and directly interpretable scale: `log2_ratio` is
in doublings, and the response-factor offset cancels in the subtraction. So the
raw difference in log2 units *is* the effect size, and it means the same thing
in every dataset. Standardising would divide by a within-dataset spread that
varies for reasons (cohort heterogeneity, sample size) that have nothing to do
with the biology, and would make the studies less comparable rather than more.

| column | meaning |
|---|---|
| `dataset_id` | the (deposit, assay) entity |
| `contrast` | which grouping variable was used, and which level was treated as "case". `sex: female vs rest` and `health_status: severe vs rest` are **not the same question**, which matters a great deal below |
| `estimate` | the contrast itself: mean `log2(par/met)` in the case group minus the reference group. Units are log2, so `+1` means the parent-to-metabolite ratio is twice as high in the case group; `0` means no difference; negative means the case group has relatively more metabolite (faster apparent metabolism) |
| `se` | standard error of that estimate: how precisely we know it. Small `se` means a tight estimate. A rough 95% interval is `estimate ± 1.96 × se`. Any pooling would weight studies by this |
| `n` | number of both-tier samples the fit used. With `n` in single digits, `se` is large and the estimate is weak |
| `note` | why a dataset has no estimate: `n<6` (too few usable samples) or `no_stratifier` (no grouping variable with two levels of at least three samples) |
| `repository`, `chromatography`, `const_country`, `const_body_site` | dataset-level attributes, available as moderators for a future pooled model |

## Per-study contrast, ordered

```{r exp-effect-bars, fig.width = 7.5, fig.height = 4.5}
fit_eff <- effects |> filter(!is.na(estimate), !is.na(se))

eff_plot <- fit_eff |>
  mutate(ci_lo = estimate - 1.96 * se,
         ci_hi = estimate + 1.96 * se,
         ds    = fct_reorder(short_ds(dataset_id), estimate))

ggplot(eff_plot, aes(x = estimate, y = ds)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
  geom_errorbar(aes(xmin = ci_lo, xmax = ci_hi), orientation = "y",
                width = 0.25, colour = "grey40") +
  geom_point(aes(colour = chromatography), size = 3) +
  geom_text(aes(label = paste0("n=", n)), hjust = -0.25, size = 3, colour = "grey30") +
  labs(x = "contrast: case minus reference, log2(parent / metabolite)",
       y = NULL, colour = "chrom.",
       title = "Per-study contrast with 95% confidence interval") +
  theme_bw()
```

Here the intervals are wide, most of them cross zero, and, critically, the
contrasts on the y-axis are **not all asking the same question**. Read this as a
description of what each dataset happens to offer, not as a set of comparable
results.

## Does the ratio itself track sex, health or age across the corpus?

Part A pooled *presence*, which is a rate and needs no common scale. The obvious
next question is whether the *ratio* moves with the covariates the corpus
records most often, pooled across all cohorts rather than one cohort at a time.
This is the analysis the offset argument above predicts will fail, so it is
worth running rather than asserting.

Two models are fitted for each covariate. The first pools every subject and
ignores which study they came from, which is what a naive corpus analysis would
do. The second adds a study term, which absorbs the per-cohort offset and leaves
only the within-cohort information — the same logic as the within-study
contrast, but pooled. If a covariate effect is real it should survive the study
term; if it is the offset in disguise, it will not.

```{r ratio-covariate-recover}
# Chagas records subject, sex, age and infection status in its file names and
# only partly in the catalogue. The recovery is validated rather than assumed:
# where the catalogue does carry a subject id the parsed one agrees 158/158, sex
# differs only in encoding and age by at most a year. It is local to this
# section, so Chagas unit counts differ from Part A.
rq <- pm[pm$detection_tier %in% "both" & is.finite(pm$log2_ratio), ]
ch_pat <- "^(pos|neg)_(inf|uninf|non)[_a-z]*_([0-9]+)(_[0-9]+)?_(female|male)_([0-9]+)_.*$"
ch <- rq$study == "Chagas serum" & grepl(ch_pat, rq$fk)
rq$subj[ch] <- sub(ch_pat, "\\3", rq$fk[ch])
rq$sex2[ch] <- sub(ch_pat, "\\5", rq$fk[ch])
rq$agen[ch] <- as_num(sub(ch_pat, "\\6", rq$fk[ch]))
rq$clin[ch] <- ifelse(sub(ch_pat, "\\2", rq$fk[ch]) == "uninf",
                      "uninfected", "T. cruzi infected")
rq <- rq[!(rq$study == "Chagas serum" & grepl("_qc_", rq$fk)), ]

rq$health <- harmonise_health(rq$clin)

# The unit rule again: one ratio per subject per matrix, median over that
# subject's files. A cohort sampling 16 files per person must not outvote a
# cohort sampling one.
ru <- rq |>
  dplyr::mutate(unit = ifelse(is.na(subj), paste0("file:", fk),
                              paste(study, subj, site))) |>
  dplyr::group_by(unit) |>
  dplyr::summarise(study = dplyr::first(study), site = dplyr::first(site),
                   l2 = median(log2_ratio), sex2 = dplyr::first(sex2),
                   health = dplyr::first(health), agen = dplyr::first(agen),
                   .groups = "drop")
```

```{r ratio-covariate-audit}
# The health mapping is printed rather than trusted: every clinical label in the
# corpus, and which side of the binary it was put on.
rq |> dplyr::filter(!is.na(clin)) |>
  dplyr::count(study, clin, health, name = "files") |>
  dplyr::arrange(study)
```

```{r ratio-covariate-scan}
# Restricted to blood, because matrix moves the ratio by more than any covariate
# tested here (plasma vs urine is about -2.6 log2) and would otherwise dominate.
blood <- ru[ru$site %in% c("plasma", "serum"), ]

scan_cov <- function(dat, v, label) {
  z <- dat[!is.na(dat[[v]]), ]
  if (nrow(z) < 8) return(NULL)
  naive <- summary(lm(as.formula(paste("l2 ~", v)), data = z))$coefficients
  adj   <- summary(lm(as.formula(paste("l2 ~ study +", v)), data = z))$coefficients
  ra <- grep(paste0("^", v), rownames(adj))[1]
  tibble(covariate = label, units = nrow(z), studies = dplyr::n_distinct(z$study),
         est_naive = round(naive[2, 1], 2), p_naive = signif(naive[2, 4], 2),
         est_adj   = round(adj[ra, 1], 2),  p_adj   = signif(adj[ra, 4], 2),
         p_study   = signif(anova(lm(as.formula(paste("l2 ~ study +", v)),
                                     data = z))["study", "Pr(>F)"], 2))
}

ratio_cov <- dplyr::bind_rows(
  scan_cov(blood, "sex2",   "sex (male vs female)"),
  scan_cov(blood, "health", "not healthy vs healthy"),
  scan_cov(blood, "agen",   "age (per year)"))
ratio_cov
write_csv(ratio_cov, "artifacts/ratio_covariates.csv")
```

```{r ratio-covariate-confound}
# How much of "healthy vs not" is really "one deposit vs another"? If each
# health arm sits almost entirely in its own study, the contrast and the offset
# are the same number. The within-cohort version is the check: only a study
# containing BOTH arms can separate them.
hb <- blood[!is.na(blood$health), ]
health_by_study <- hb |> dplyr::count(study, health) |>
  tidyr::pivot_wider(names_from = health, values_from = n, values_fill = 0)
health_by_study

both_arms <- health_by_study$study[
  health_by_study$healthy >= 3 & health_by_study$`not healthy` >= 3]
within_health <- lapply(both_arms, function(s) {
  z <- hb[hb$study == s, ]
  # same direction as the model coefficient: not healthy minus healthy
  tibble(study = s, units = nrow(z),
         est = round(unname(diff(tapply(z$l2, z$health, median))), 2),
         p   = signif(wilcox.test(l2 ~ health, data = z)$p.value, 2))
}) |> dplyr::bind_rows()
within_health
```

Nothing survives the study term. Sex moves the ratio by
`r sprintf("%+.2f", ratio_cov$est_naive[1])` log2 pooled ($p =
`r ratio_cov$p_naive[1]`$) and by `r sprintf("%+.2f", ratio_cov$est_adj[1])`
once study is accounted for ($p = `r ratio_cov$p_adj[1]`$). Age is flat either
way ($p = `r ratio_cov$p_naive[3]`$ and $p = `r ratio_cov$p_adj[3]`$).

Health status is the one covariate that looks significant, and it is the most
instructive result in this section. Pooled across the corpus and ignoring which
deposit each subject came from, unhealthy subjects sit
`r sprintf("%+.2f", ratio_cov$est_naive[2])` log2 below the healthy ones at $p =
`r ratio_cov$p_naive[2]`$ — a result that, reported on its own, would read as a
real corpus-level finding. Adding the study term leaves the estimate essentially
unchanged at `r sprintf("%+.2f", ratio_cov$est_adj[2])` but takes the $p$ value
to `r ratio_cov$p_adj[2]`, while the study term itself is significant at $p =
`r ratio_cov$p_study[2]`$.

An estimate that survives adjustment while its $p$ value does not is the
signature of confounding rather than of a real effect losing power. Health status
in this corpus is almost entirely a between-study variable, as the table above
shows: each arm sits largely in its own deposit, so comparing healthy against
unhealthy mostly compares one deposit against another, and the offset is exactly
what that comparison picks up. Only a cohort holding both arms can separate the
two, and
`r if (!nrow(within_health)) "no cohort in the corpus holds both" else paste0(if (nrow(within_health) == 1) "the single cohort that does gives " else paste0("the ", nrow(within_health), " that do give "), sprintf("%+.2f", within_health$est[1]), " log2 at $p = ", within_health$p[1], "$ on ", within_health$units[1], " subjects")`.
The direction agrees with the pooled estimate but the significance does not
survive, which is the whole point: the corpus-wide result is not evidence that
illness changes the
hydroxylation ratio; it is evidence that the deposits differ from each other,
which is what the offset argument said all along.

The result is negative but not final. It rests on `r nrow(blood)` blood subjects
across `r dplyr::n_distinct(blood$study)` studies, because a ratio needs both
compounds in the same file. Two things would enlarge it: deeper metadata, since
the Chagas recovery above shows deposits were set aside on what the catalogues
indexed rather than on what the submitters actually recorded; and deposits that
contain both arms of a contrast, since a clinical stratum that is also a deposit
cannot be separated from that deposit's offset by any amount of pooling.

# Chagas HILIC vs RP: one direct check

MSV000092754 (HILIC) and MSV000092757 (reverse phase) are paired deposits from
the same Chagas serum study: the same subjects and blood draws, run on two
chromatographies, matched by `codigo_num`. It is the only place in the corpus
where we can watch the same biology measured two ways, which makes it the one
direct check on the offset argument above.

```{r chagas-pair}
#| echo: false
HILIC_DS <- "MSV000092754"
RP_DS    <- "MSV000092757"

per_subject_ratio <- function(ds) {
  samples %>%
    filter(dataset_id == ds, detection_tier == "both",
           !par_fwhm_outlier %in% TRUE, !met_fwhm_outlier %in% TRUE,
           is.finite(log2_ratio)) %>%
    group_by(subject_id) %>%
    summarise(log2_ratio    = median(log2_ratio, na.rm = TRUE),
              health_status = dplyr::first(health_status),
              .groups       = "drop")
}

chagas <- dplyr::inner_join(
  per_subject_ratio(HILIC_DS) %>% dplyr::rename(log2_ratio_HILIC = log2_ratio),
  per_subject_ratio(RP_DS)    %>% dplyr::select(subject_id, log2_ratio_RP = log2_ratio),
  by = "subject_id")

cat("Subjects matched across both chromatographies:", nrow(chagas), "\n")
```

```{r chagas-paired, fig.width = 6, fig.height = 4.5}
if (nrow(chagas) >= 3) {
  chagas |>
    pivot_longer(c(log2_ratio_HILIC, log2_ratio_RP),
                 names_to = "chromatography", values_to = "log2_ratio") |>
    mutate(chromatography = sub("log2_ratio_", "", chromatography)) |>
    ggplot(aes(x = chromatography, y = log2_ratio,
               group = subject_id, colour = health_status)) +
    geom_line(alpha = 0.5) +
    geom_point(size = 2.5) +
    labs(x = NULL, y = "log2(parent / metabolite)", colour = "health status",
         title = "Per-subject log2 ratio on HILIC and reverse phase, MSV000092754 / MSV000092757") +
    theme_bw()
}
```

**What this can and cannot tell us.** Only a handful of subjects survive the
requirement of a usable ratio in *both* chromatographies, so this is a look, not
a test. Correlation coefficients, Bland-Altman limits of agreement and
heteroscedasticity slopes on this many points would report precision we do not
have, so we do not compute them.

What the picture is good for is direction. If the lines were roughly parallel,
each subject shifting by a similar amount between the two methods, that is the
constant-additive-offset behaviour the previous section assumes, and it would
support the claim that the offset cancels in a within-study contrast. If instead
the lines cross each other freely, the readout depends on the chromatography in
a way a single constant cannot absorb, and every cross-method claim in the
project needs a louder caveat. Judge it by eye, and treat it as one encouraging
or discouraging data point rather than as a validation.

# What a meta-analysis could add, and what it needs first

The obvious next step is to pool the per-study contrasts into a single
cross-cohort estimate: a random-effects meta-analysis, weighting each dataset by
its precision, reporting a pooled contrast with a confidence interval plus τ²
and I² to quantify how much the cohorts genuinely disagree. That is the standard
tool and it is the right shape for this data. **We are not running it yet, on
purpose.**

**Why not.** A meta-analysis pools estimates that answer the same question. Look
again at the `contrast` column: because `fit_within_study()` walks a priority
list and takes the first usable grouping, most datasets end up contributing a
**sex** contrast, simply because sex is the one variable that is populated
nearly everywhere. Pooling "severe versus healthy" from one study together with
"male versus female" from another does not estimate anything. Run anyway it
returns a tidy-looking pooled value with a narrow interval and I² near zero,
which is worse than no answer because it looks like a result. The blocker is not
the statistics; it is that the corpus does not supply a shared, biologically
meaningful contrast across enough datasets.

**What would make it work**, roughly in order of value:

- **Pick one question and only include datasets that can answer it.** For
  example health status or disease severity, dropping every dataset without it.
  Fewer studies, but a pooled number that means something.
- **Handle the censored samples.** `parent_only` and `metab_only` are ratio
  observations with one bound known. An interval-censored (Tobit-style) model
  uses them as bounds instead of discarding them, which matters because they are
  exactly the most extreme ratios, and dropping them shrinks any effect toward
  zero.
- **Decide how to treat the method.** Chromatography, instrument and biofluid
  all reset the offset. With a real pooled model these become moderators or
  subgroups. Note the corpus currently has at least two deposits whose recorded
  chromatography is doubtful given their retention times, so this needs the
  metadata fixed before it can be trusted.
- **Consider the one-stage alternative.** Rather than fitting each study then
  pooling, a hierarchical (mixed-effects) model on the sample-level data with a
  random effect per dataset uses the individual samples directly, which is
  generally better behaved when per-study `n` is small, as it is here.
- **Pool a robust quantity, never raw abundance.** This is settled practice, and
  it converges across fields. Cross-repository proteomics meta-analyses of
  independent public deposits (e.g. the 12-dataset colorectal-cancer study,
  PMC10833860) uniformly re-process the raw files and then pool a
  distribution-free quantity: rank-binned abundances there, Stouffer/Fisher
  p-value combination in MetaMSD, within-study log fold-changes in the
  metabolomics tool MetaHD. If we ever pool a ratio across cohorts, it must be
  rank-transformed or reduced to a per-cohort statistic first, for the same
  response-factor reason that forbids pooling the raw ratio. That these
  cross-repository quantitative meta-analyses exist and work is also the point:
  our contribution is not the pooling, which is standard, but using MS1
  propagation to *produce* the per-cohort quantity that the reverse-metabolomics
  / MASST world does not generate.
- **Do not treat every accession as an independent cohort.** `01` records the
  known exceptions in `artifacts/related_deposits.csv`. Today it contains one:
  **MSV000084008 is a re-deposit of MSV000082493's blood arm**, and both come
  from the same 14 subjects (PMC7485946). They contribute different biofluids
  here, so both are kept, but pooling them as two studies would count those
  subjects twice and hand the pair a double vote in any precision-weighted
  estimate. A deposit accession is not a study, and any pooling has to be keyed
  on the study.
- **Be honest about country.** Each country is represented by a single dataset,
  so country and dataset are perfectly confounded. Any "country effect" is that
  one deposit's method. At best this is a calibration check against published
  CYP2C19 frequencies, never an estimate.

The prerequisite for all of it is a decision about which question the pool is
supposed to answer. That is a discussion to have as a group, not something to
settle by picking whichever variable happens to be populated.

# Appendix: one study, two accessions (MSV000082493 and MSV000084008)

```{r appendix-overlap}
#| echo: false
N_SHARED_APPX <- length(intersect(
  norm_file_key(samples_perfile$file[samples_perfile$dataset_id == "MSV000082493"]),
  norm_file_key(samples_perfile$file[samples_perfile$dataset_id == "MSV000084008"])))
```

These two deposits are **not independent cohorts**. They are the same study,
submitted twice:

- The same **14 subjects**, and on every file present in both, `subject_id`
  agrees: `r N_SHARED_APPX` of `r N_SHARED_APPX`.
- **`r N_SHARED_APPX` files are shared between them** - all `blood_plasma`.
  Where both yield a parent detection, the AUC correlation is **1.000** and the
  median relative difference is **0**; the largest single difference is 11%, so
  most are the same raw file and a few differ slightly (most likely
  re-conversion, not different samples). Both deposits are RP.
- 082493 additionally holds the multi-matrix collection (390 skin, 148 fecal,
  139 urine, 8 plasma); 084008 holds 106 plasma files 082493 does not. Neither
  is a subset of the other, so distinct plasma across both is 447.

Nothing in the repository metadata signals this. Different accessions, separate
submissions, no cross-reference. It surfaced only from a duplicate-file audit.

**This inflates any public-data meta-analysis, and it hits an MS2-only analysis
harder than ours**, because the duplicated files are exactly the plasma where
MASST found the parent:

| counted over | summed | distinct | inflation |
|---|---|---|---|
| parent MS2 hit files | 389 | 233 | **+67%** |
| 5-OH MS2 hit files | 494 | 342 | **+44%** |
| all MS2 hit files, corpus | 668 | 472 | **+42%** |

So a classic reverse-metabolomics presence analysis reporting "N public files
contain the drug" would over-report by roughly half on this pair alone. That is
a property of the data ecosystem, not of any particular method, and it is not
routinely checked.

## A shared file name does not always mean a duplicate

The opposite case exists in this same corpus and is just as dangerous.
**MSV000092754 and MSV000092757 share all 164 file names** (`pos_QC_1.mzXML` and
so on) and the same 79 subjects, but 092754 is HILIC and 092757 is RP: the same
samples on two columns, submitted as separate accessions. Their extracted values
differ by up to **117x**. De-duplicating those by file name would merge a HILIC
run into an RP run and delete 164 real files.

So file name identity means two opposite things here, and only the analytical
context distinguishes them:

| pair | file names | chromatography | verdict |
|---|---|---|---|
| 082493 / 084008 | 333 shared | RP / RP | same files - **collapse** |
| 092754 / 092757 | 164 shared | HILIC / RP | different runs - **keep apart** |

**What we therefore do**, and it must be stated wherever these numbers appear:

1. **Identify the duplicate pair from subject identity**, not file names: the
   same 14 subjects, with subject, timepoint, study day and body site agreeing
   on 461 of 461 shared files. File names cannot make this call - here they
   indicate a duplicate, in 092754/092757 they indicate a collision.
2. **Collapse that pair, and only that pair, on file name** (`DUP_STUDY` below).
   Every other dataset stays namespaced by its own accession.
3. **Treat 082493 + 084008 as one study.** Per-cohort tables keep both
   accessions for provenance, but they must never be read as two independent
   replications - and a result "confirmed" in both is confirmed once.
4. **Re-run the collision audit when datasets are added.** Compare extracted
   AUCs for every shared key across every dataset pair: agreement means one
   file deposited twice, disagreement means two files that happen to share a
   name. Nothing else separates the two cases reliably.

One thing this does **not** address, and which is worth stating rather than
hiding: each deposit also contains **technical replicates** - 237 rows in each
of 082493 and 084008 repeat a `(subject, timepoint, day, body_site)`
combination. Those are multiple injections of one specimen, and they are counted
individually in every file-level number here, including the gain. A
specimen-level count would be more defensible, but it needs an identifier for
the specimen rather than for the person, and no deposit records one: the three
that repeat a subject do not distinguish a re-injection from a second draw, and
the rest have one file per person, so it cannot be done corpus-wide.

Subject-level analyses were never affected: the project already takes one
measure per subject, and the subject IDs agree across the two accessions.

```{r dedup-corpus}
#| echo: false
fk_all <- fkey(samples_perfile$file)
dk_all <- paste0(samples_perfile$dataset_id, "|", fk_all)
cat("files summed across datasets :", length(dk_all), "\n")
cat("distinct physical files      :", length(unique(fk_all)), "\n")
cat("counted more than once       :", length(dk_all) - length(unique(fk_all)), "\n")
```


# References

- Viechtbauer, W. (2010). Conducting Meta-Analyses in R with the metafor
  package. *Journal of Statistical Software*, 36(3).
- Blanz, J. et al. (2017). Evaluation of relative MS response factors of drug
  metabolites for semi-quantitative assessment of chemical liabilities. *Journal
  of Mass Spectrometry*, 52(4).
- Viallon, V. et al. (2021). A new pipeline for the normalization and pooling of
  metabolomics data. *Metabolites*, 11(9).
- Belasimo, A. et al. (2024). MetaHD: a multivariate meta-analysis model for
  metabolomics data. *Bioinformatics*, 40(7).
- Petrović, J. et al. (2020). Meta-analysis of probability estimates of
  worldwide variation of CYP2D6 and CYP2C19. *Translational Psychiatry*, 10.
- Tornio, A., Backman, J. T. (2018). Cytochrome P450 in pharmacogenetics: an
  update. *British Journal of Clinical Pharmacology*, 122 (Suppl. 5).
- Abouir, K. et al. (2024). Improving CYP2C19 phenotyping using stereoselective
  omeprazole and 5-hydroxy-omeprazole metabolic ratios. *Basic & Clinical
  Pharmacology & Toxicology*.
