---
title: "Dataset report"
subtitle: "Per-dataset deep dive: extraction config, EIC diagnostics, detection tiers, within-study stats"
date: today
params:
  dataset_id: "MTBLS1866_LC-MS_positive_reverse-phase_metabolite_profiling"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: paged
execute:
  warning: false
  message: false
  echo: false
---

```{r setup}
source("R/_setup.R")
source("R/ms1.R")
library(Chromatograms)
library(ggplot2)

ds <- params$dataset_id
rds <- sprintf("artifacts/per_dataset/%s.rds", ds)
if (!file.exists(rds))
  stop("No cache for ", ds, " at ", rds, "; run 03 first.")

res        <- readRDS(rds)
sample_tbl <- res$sample_tbl
sample_tbl$detection_tier <- factor(
  sample_tbl$detection_tier,
  levels = c("both", "parent_only", "metab_only", "neither"))

tier_bar <- function(df, col) {
  ggplot(df, aes(x = .data[[col]], fill = detection_tier)) +
    geom_bar() +
    labs(x = NULL, y = "files", fill = "Detection tier",
         title = paste("Detection tier by", col)) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 35, hjust = 1),
          legend.position = "bottom")
}
```

# `r params$dataset_id`

```{r overview}
#| results: asis
cat("- **Deposit**: ", res$deposit_id, "\n",
    "- **Assay**:   ", res$assay,      "\n",
    "- **Files**:   ", nrow(sample_tbl), " biological MS files\n",
    "- **m/z**:     parent ", round(res$cfg$mz_par, 4),
    " · metabolite ", round(res$cfg$mz_met, 4), "\n",
    "- **Adducts**: parent ", res$cfg$adduct_par,
    " · metabolite ", res$cfg$adduct_met, "\n\n",
    sep = "")
cat("Detection tiers:\n\n```\n")
print(table(sample_tbl$detection_tier))
cat("```\n")
```

# Extraction config

The m/z and RT windows below come from `derive_extraction_config()`
applied to this dataset's SIRIUS-confirmed features. The window uses the
median observed precursor m/z (not the theoretical mass) to absorb
per-instrument calibration offset and anchor to the adduct SIRIUS resolved.
The RT window is the range of confirmed RTs ± 30 s, conservative against
within-study gradient drift.

```{r config-table}
#| results: asis
fmt_rt <- function(r) sprintf("[%.1f, %.1f] s", r[1], r[2])
cat("| | parent | metabolite |\n")
cat("|---|---|---|\n")
cat("| adduct      |", res$cfg$adduct_par, "|", res$cfg$adduct_met, "|\n")
cat("| m/z         |", round(res$cfg$mz_par, 5), "|",
                       round(res$cfg$mz_met, 5), "|\n")
cat("| RT window   |", fmt_rt(res$cfg$rt_par_win), "|",
                       fmt_rt(res$cfg$rt_met_win), "|\n")
cat("| confirmed source files (positive controls) |",
    length(res$cfg$confirmed_files_parent),     "|",
    length(res$cfg$confirmed_files_metabolite), "|\n")
```

# EIC overlay

`Chromatograms::chromExtract()` extracted EICs in a single pass: one
trace per file, per role, at the m/z window above. The plots overlay all
traces; a tight cluster at the configured RT is expected, while large
vertical spread or shoulders indicate real biology or unresolved isobars.

```{r eic-overlay, fig.width = 9, fig.height = 6}
op <- par(mfrow = c(2, 1), mar = c(3, 4, 2, 1))
plotChromatogramsOverlay(res$chr_par,
  col   = adjustcolor("steelblue", 0.5),
  ylim  = eic_ylim(res$chr_par),
  main  = paste(ds, "(parent EICs)"))
plotChromatogramsOverlay(res$chr_met,
  col   = adjustcolor("tomato", 0.5),
  ylim  = eic_ylim(res$chr_met),
  main  = paste(ds, "(metabolite EICs)"))
par(op)
```

# Peak detection

`MsCoreUtils::localMaxima` looks for a local maximum within a half-
window of scans, above a fraction of the trace maximum. The summary
below reports how many files produced at least one detected peak per
role.

```{r peak-summary}
#| results: asis
cat("- **Parent**     EICs with ≥1 peak: ",
    sum(res$peaks_par > 0), " / ", length(res$peaks_par), "\n",
    "- **Metabolite** EICs with ≥1 peak: ",
    sum(res$peaks_met > 0), " / ", length(res$peaks_met), "\n",
    sep = "")
```

## Positive-control check

SIRIUS-confirmed files are positive controls: the EIC pipeline **must**
detect a peak there. A failure flags a settings problem (PPM tolerance,
RT window, or adduct choice), not a biology result.

```{r positive-control}
par_files_all <- basename(dataOrigin(res$chr_par))
met_files_all <- basename(dataOrigin(res$chr_met))
check_confirmed(par_files_all, res$peaks_par,
                res$cfg$confirmed_files_parent,     "Parent    ")
check_confirmed(met_files_all, res$peaks_met,
                res$cfg$confirmed_files_metabolite, "Metabolite")
```

# "Neither" EIC inspection

Files with zero peaks for **both** roles are either genuinely missing the
compound or had a misplaced detection window. We re-extract a wider RT
window (±60 s) and drop the EICs empty at the target m/z. What remains,
files with *some* signal at the target mass that still failed the peak
threshold, is the informative subset to inspect. The shaded band marks
the configured detection window.

```{r neither-eics, fig.width = 11, fig.height = 4}
files_par_ok <- basename(dataOrigin(res$chr_par))[res$peaks_par > 0]
files_met_ok <- basename(dataOrigin(res$chr_met))[res$peaks_met > 0]
all_files    <- basename(dataOrigin(res$chr_par))
neither_files <- setdiff(all_files, union(files_par_ok, files_met_ok))

WIDE_PAD_S <- 60
rt_wide_lo <- min(res$cfg$rt_par_win[1], res$cfg$rt_met_win[1]) - WIDE_PAD_S
rt_wide_hi <- max(res$cfg$rt_par_win[2], res$cfg$rt_met_win[2]) + WIDE_PAD_S
PPM <- 20

if (length(neither_files)) {
  neither_orig <- dataOrigin(res$chr_par)[all_files %in% neither_files]

  peak_tbl_wide <- rbind(
    data.frame(role = "parent", msLevel = 1L, dataOrigin = neither_orig,
               rtMin = rt_wide_lo, rtMax = rt_wide_hi,
               mzMin = res$cfg$mz_par * (1 - PPM * 1e-6),
               mzMax = res$cfg$mz_par * (1 + PPM * 1e-6),
               stringsAsFactors = FALSE),
    data.frame(role = "metabolite", msLevel = 1L, dataOrigin = neither_orig,
               rtMin = rt_wide_lo, rtMax = rt_wide_hi,
               mzMin = res$cfg$mz_met * (1 - PPM * 1e-6),
               mzMax = res$cfg$mz_met * (1 + PPM * 1e-6),
               stringsAsFactors = FALSE)
  )
  chr_neither_wide <- chromExtract(res$chr_all, peak_tbl_wide,
                                   by = c("msLevel", "dataOrigin"))
  n_total    <- length(chr_neither_wide)
  chr_signal <- filterEmptyChromatograms(chr_neither_wide)
  n_empty    <- n_total - length(chr_signal)

  cat("\"Neither\" files:", length(neither_files), "(", n_total, "EICs)\n")
  cat("  genuinely empty (no ion at target m/z) :", n_empty, "→ dropped\n")
  cat("  some signal, sub-threshold             :", length(chr_signal),
      "→ shown\n")

  if (length(chr_signal)) {
    cd <- chromData(chr_signal)
    keep_files <- unique(basename(cd$dataOrigin))
    if (length(keep_files) > 24) {
      keep_files <- keep_files[seq_len(24)]
      cat("  (capped to first 24 files for the plot)\n")
      chr_signal <- chr_signal[basename(cd$dataOrigin) %in% keep_files]
      cd <- chromData(chr_signal)
    }

    df_neither <- do.call(rbind, lapply(seq_len(length(chr_signal)), function(i) {
      pd <- peaksData(chr_signal)[[i]]
      data.frame(role = cd$role[i], file = basename(cd$dataOrigin[i]),
                 rtime = pd$rtime, intensity = pd$intensity,
                 stringsAsFactors = FALSE)
    }))

    win_rects <- data.frame(
      role = c("parent", "metabolite"),
      xmin = c(res$cfg$rt_par_win[1], res$cfg$rt_met_win[1]),
      xmax = c(res$cfg$rt_par_win[2], res$cfg$rt_met_win[2]),
      stringsAsFactors = FALSE)

    print(
      ggplot(df_neither, aes(x = rtime, y = intensity)) +
        geom_rect(data = win_rects,
                  aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
                  inherit.aes = FALSE, fill = "steelblue", alpha = 0.15) +
        geom_line(colour = "grey30") +
        facet_grid(role ~ file, scales = "free_y") +
        labs(x = "Retention time (s)", y = "Intensity",
             title = "\"Neither\" files with sub-threshold signal",
             subtitle = paste0("Shaded: configured detection window  ·  ",
                               "wide window ±", WIDE_PAD_S, " s")) +
        theme_bw(base_size = 9) +
        theme(strip.text.x     = element_text(size = 7),
              strip.background = element_rect(fill = "grey92"))
    )
  } else {
    cat("All \"neither\" EICs are genuinely empty; nothing to inspect.\n")
  }
} else {
  cat("No \"neither\" files: every loaded file produced ≥1 peak in at least one role.\n")
}
```

# FWHM outliers

Within one dataset peak width should be roughly constant (fixed gradient,
same compound across injections). A file whose FWHM departs strongly from
the median (> 3 × MAD) usually means the EIC integrates two fused peaks, a
baseline-noise lump, or a co-eluting isobar; its AUC is suspect and
flagged before entering the ratio.

```{r fwhm-summary}
#| results: asis
cat("- **Parent**     median FWHM = ",
    round(median(res$metrics_par$par_fwhm, na.rm = TRUE), 1),
    " s, outliers flagged: ",
    sum(res$metrics_par$par_fwhm_outlier %in% TRUE), "\n",
    "- **Metabolite** median FWHM = ",
    round(median(res$metrics_met$met_fwhm, na.rm = TRUE), 1),
    " s, outliers flagged: ",
    sum(res$metrics_met$met_fwhm_outlier %in% TRUE), "\n",
    sep = "")
```

Each flagged EIC is plotted against a reference EIC from the file closest
to the median FWHM. A genuine outlier looks deformed (double apex,
asymmetric tail, broad shoulder, noisy baseline); one that looks
structurally similar to the reference is a false alarm from small FWHM
variance and should stay in the ratio.

```{r fwhm-plots, fig.width = 10, fig.height = 5}
plot_fwhm_outliers(res$chr_par_narrow, res$metrics_par, "par", "Parent")
plot_fwhm_outliers(res$chr_met_narrow, res$metrics_met, "met", "Metabolite")
```

# PCA of EIC metrics

An unsupervised view on the per-file metric matrix (`par_auc`, `met_auc`,
`par_fwhm`, `met_fwhm`, `log2_ratio`), coloured by `health_status` and
labelled by `subject_id`. A cluster split that does **not** track
`health_status` is evidence of a latent covariate the deposit does not
expose: a missing-metadata diagnostic, not just an exploratory plot.

```{r pca, fig.width = 7, fig.height = 5}
both_for_pca <- as.data.frame(sample_tbl[sample_tbl$detection_tier == "both", ])
both_for_pca$.row_id <- seq_len(nrow(both_for_pca))

pc_cols <- c("par_auc", "met_auc", "par_fwhm", "met_fwhm", "log2_ratio")
pc_cols <- intersect(pc_cols, names(both_for_pca))
pc_mat  <- as.matrix(both_for_pca[, pc_cols, drop = FALSE])
rownames(pc_mat) <- as.character(both_for_pca$.row_id)

inf_cols <- apply(pc_mat, 2, function(x) {
  v <- stats::sd(x, na.rm = TRUE); !is.na(v) && v > 0 })
pc_mat <- pc_mat[, inf_cols, drop = FALSE]
pc_mat <- pc_mat[stats::complete.cases(pc_mat), , drop = FALSE]

if (nrow(pc_mat) >= 5 && ncol(pc_mat) >= 2) {
  pc      <- stats::prcomp(pc_mat, scale. = TRUE)
  row_ids <- as.integer(rownames(pc$x))
  pc_df   <- cbind(
    as.data.frame(pc$x[, 1:min(2, ncol(pc$x)), drop = FALSE]),
    both_for_pca[row_ids,
                 intersect(c("subject_id", "health_status", "country"),
                           names(both_for_pca)), drop = FALSE])

  p <- ggplot(pc_df, aes(PC1, PC2))
  if ("health_status" %in% names(pc_df))
    p <- p + aes(colour = health_status)
  if ("subject_id" %in% names(pc_df))
    p <- p + aes(label = subject_id) +
         geom_text(vjust = -0.7, size = 3, show.legend = FALSE)
  print(
    p +
      geom_point(size = 3, alpha = 0.85) +
      labs(title    = paste(ds, "(PCA of EIC metrics)"),
           subtitle = paste("n =", nrow(pc_df), "subjects"),
           caption  = "Cluster splits not tracking health_status hint at unmeasured covariates") +
      theme_bw() +
      theme(legend.position = "bottom")
  )
} else {
  cat("Skipping PCA: need ≥5 `both` samples and ≥2 non-constant metrics ",
      "(have ", nrow(pc_mat), " rows, ", ncol(pc_mat), " informative cols).\n",
      sep = "")
}
```

# Stratified by health_status

```{r hs-bar}
#| fig-width: 6.5
#| fig-height: 4
if ("health_status" %in% names(sample_tbl) &&
    length(unique(na.omit(sample_tbl$health_status))) >= 1) {
  print(tier_bar(sample_tbl, "health_status"))
}
```

```{r hs-stats}
# Description only. The per-dataset p-values that used to appear here tested files rather
# than subjects, so they estimated nothing; `05` runs the contrasts under the unit rule.
ws <- describe_by_group(sample_tbl, group_col = "health_status")
ws$detection_table
ws$per_group_ratio
```

```{r hs-ratio}
#| fig-width: 6.5
#| fig-height: 5
both <- sample_tbl[sample_tbl$detection_tier == "both" &
                   !sample_tbl$par_fwhm_outlier %in% TRUE &
                   !sample_tbl$met_fwhm_outlier %in% TRUE, ]
if (nrow(both) > 0 && "health_status" %in% names(both)) {
  print(
    ggplot(both, aes(x = health_status, y = log2_ratio)) +
      geom_jitter(width = 0.18, alpha = 0.85, size = 2.5) +
      stat_summary(fun = median, geom = "crossbar",
                   width = 0.5, colour = "grey20", linewidth = 0.4) +
      geom_hline(yintercept = 0, linetype = 2, colour = "grey60") +
      labs(x = NULL, y = "log2(omeprazole / 5-OH-omeprazole)",
           title = paste0("Ratio by health_status (n=", nrow(both), ")")) +
      theme_bw() +
      theme(legend.position = "none",
            axis.text.x = element_text(angle = 35, hjust = 1))
  )
}
```

# Additional stratifiers

Sections below are generated only for canonical fields this dataset
actually carries (≥2 levels, ≥3 samples each), beyond `health_status`.

```{r extra-stratifiers}
#| results: asis
extra <- usable_stratifiers(sample_tbl, exclude = "health_status")

if (!length(extra)) {
  cat("_None: `health_status` is the only usable stratifier for this dataset._\n")
} else {
  tmpl <- paste(
    "## Stratified by `%1$s`",
    "",
    "```{r}",
    "ws_x <- describe_by_group(sample_tbl, group_col = \"%1$s\")",
    "ws_x$detection_table",
    "ws_x$per_group_ratio",
    "```",
    "",
    "```{r}",
    "#| fig-width: 6.5",
    "#| fig-height: 4",
    "print(tier_bar(sample_tbl, \"%1$s\"))",
    "```",
    "", sep = "\n")
  for (f in extra)
    cat(knitr::knit_child(text = sprintf(tmpl, f), quiet = TRUE), sep = "\n")
}
```
