---
title: "MASST Hit Confirmation — RuSirius"
subtitle: "Per-dataset SIRIUS annotation of curated MASST hits (MassIVE)"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
execute:
  warning: false
  message: false
---

# Overview

This document takes the curated MASST hit table produced by
`reverse-metabolomics-workflow.qmd` (DIA datasets and datasets without
metadata already removed) and runs SIRIUS annotation for every remaining
MassIVE dataset.

For each dataset two SIRIUS projects are created:

```
sirius_projects/
  <dataset_id>/
    parent/
    metabolite/
results/
  <dataset_id>/
    parent_results.csv
    metabolite_results.csv
```

Results are written to disk for manual review. The classification
helper at the end of this document (`eval: false`) can be run
interactively after inspecting the raw tables.

# Version check

```{r versions}
source("check-versions.R")
```

# Setup

```{r setup}
library(Spectra)
library(MsBackendMassIVE)
library(MetaboCoreUtils)
library(RuSirius)
library(dplyr)
```

# Compound targets

```{r compound-spec}
compound_spec <- data.frame(
  role    = c("parent",       "metabolite"),
  name    = c("omeprazole",   "5-hydroxyomeprazole"),
  formula = c("C17H19N3O3S",  "C17H19N3O4S"),
  ikey1   = c("SUBDBMMJDZJVOS", "YMBXIQCHYXGHAG")
)
```

# Load curated hits

`hits_curated.csv` is written by `reverse-metabolomics-workflow.qmd`
after the DIA and no-metadata filters have been applied.

```{r load-hits}
hits_curated <- read.csv("hits_curated.csv", check.names = FALSE)
```

# Filter to MassIVE datasets

```{r filter-massive}
hits_massive <- hits_curated[
  startsWith(hits_curated$ATTRIBUTE_DatasetAccession, "MSV"), ]

massive_datasets <- unique(hits_massive$ATTRIBUTE_DatasetAccession)
cat(length(massive_datasets), "MassIVE datasets to process\n")
```

# Helper functions

```{r helpers}
ADDUCT_CANDIDATES <- c("[M+H]+", "[M+Na]+", "[M+K]+", "[M+NH4]+",
                       "[M-H]-", "[M+Cl]-")
PPM     <- 20
RT_TOL  <- 3   # seconds around hit RT

infer_adduct <- function(obs_mz, formula, ppm = PPM,
                         candidates = ADDUCT_CANDIDATES) {
  if (is.na(obs_mz) || is.na(formula)) return(NA_character_)
  mass     <- calculateMass(formula)
  theo     <- vapply(candidates, function(a) mass2mz(mass, adduct = a)[[1]],
                     numeric(1))
  diff_ppm <- abs(obs_mz - theo) / theo * 1e6
  best     <- which.min(diff_ppm)
  if (diff_ppm[best] > ppm) NA_character_ else candidates[best]
}

# Returns a list of Spectra slices (one per MASST hit) for one role.
# Each slice contains the top MS1 scan + all MS2 scans within the RT/mz window.
slice_role <- function(hits_ds, sp, sp_file, sp_scan, role) {
  role_formula <- compound_spec$formula[compound_spec$role == role]
  h <- hits_ds[hits_ds$role == role & !is.na(hits_ds$rt), ]
  if (!nrow(h)) return(list())

  lapply(seq_len(nrow(h)), function(i) {
    hi     <- h[i, ]
    mz_rng <- hi$Precursor_MZ * c(1 - PPM * 1e-6, 1 + PPM * 1e-6)
    rt_rng <- hi$rt + c(-RT_TOL, RT_TOL)

    sp_f <- filterRt(sp[sp_file == hi$file], rt_rng)
    ms1  <- filterMsLevel(sp_f, 1L)
    ms2  <- filterPrecursorMzRange(filterMsLevel(sp_f, 2L), mz_rng)

    int_at_mz <- vapply(
      peaksData(ms1),
      function(pm) {
        keep <- pm[, "mz"] >= mz_rng[1] & pm[, "mz"] <= mz_rng[2]
        if (any(keep)) max(pm[keep, "intensity"]) else 0
      },
      numeric(1)
    )
    ms1_top <- if (length(ms1) && max(int_at_mz) > 0) {
      ms1[which.max(int_at_mz)]
    } else {
      ms1[integer(0)]
    }

    out          <- c(ms1_top, ms2)
    out$adduct   <- rep(hi$adduct_inferred, length(out))
    out$group_idx <- i
    out
  })
}

run_sirius_role <- function(sp_slices, adduct_list, project_path) {
  sp_cat <- concatenateSpectra(sp_slices)
  srs <- Sirius(
    projectId = project_path,
    username  = "plouail@eurac.edu",
    password  = Sys.getenv("SIRIUS_PASSWORD")
  )
  srs <- import(
    sirius                 = srs,
    spectra                = sp_cat,
    adduct                 = adduct_list,
    ms_column_name         = "group_idx",
    deleteExistingFeatures = TRUE
  )
  run(
    srs,
    formulaIdParams          = formulaIdParam(),
    structureDbSearchParams  = structureDbSearchParam(structureSearchDbs = "ALL"),
    predictParams            = predictParam(),
    wait                     = TRUE
  )
  res <- results(srs, result.type = "structureDb", return.type = "data.frame")
  shutdown(srs)
  res
}
```

# Per-dataset SIRIUS annotation

```{r per-dataset-loop}
role_formula <- setNames(compound_spec$formula, compound_spec$role)

for (ds in massive_datasets) {
  message("\n--- Processing ", ds, " ---")
  hits_ds <- hits_massive[hits_massive$ATTRIBUTE_DatasetAccession == ds, ]

  # Infer adducts from precursor m/z
  hits_ds$adduct_inferred <- mapply(
    infer_adduct,
    hits_ds$Precursor_MZ,
    role_formula[hits_ds$role],
    USE.NAMES = FALSE
  )

  # Output directories
  dir_par <- file.path("sirius_projects", ds, "parent")
  dir_met <- file.path("sirius_projects", ds, "metabolite")
  dir_res <- file.path("results", ds)
  dir.create(dir_par, recursive = TRUE, showWarnings = FALSE)
  dir.create(dir_met, recursive = TRUE, showWarnings = FALSE)
  dir.create(dir_res, recursive = TRUE, showWarnings = FALSE)

  # Download MASST-hit files for this dataset
  wanted <- unique(basename(hits_ds$file_path))
  massive_sync_data_files(ds, fileName = wanted)

  sp       <- Spectra(backendInitialize(MsBackendMassIVE(), massiveId = ds,
                                        offline = TRUE))
  sp_file  <- basename(dataOrigin(sp))
  sp_scan  <- acquisitionNum(sp)

  hits_ds$file    <- basename(hits_ds$file_path)
  hit_idx         <- match(
    paste(hits_ds$file, hits_ds$scan_id, sep = "::"),
    paste(sp_file,      sp_scan,          sep = "::")
  )
  hits_ds$rt <- rtime(sp)[hit_idx]

  # Slice spectra per role
  sp_par <- slice_role(hits_ds, sp, sp_file, sp_scan, "parent")
  sp_met <- slice_role(hits_ds, sp, sp_file, sp_scan, "metabolite")

  if (!length(sp_par)) {
    message("  No parent spectra — skipping parent SIRIUS run")
  } else {
    adduct_list_par <- lapply(sp_par, function(s) s$adduct[1])
    res_par <- run_sirius_role(sp_par, adduct_list_par, dir_par)
    write.csv(res_par, file.path(dir_res, "parent_results.csv"), row.names = FALSE)
    message("  Parent done: ", nrow(res_par), " rows written")
  }

  if (!length(sp_met)) {
    message("  No metabolite spectra — skipping metabolite SIRIUS run")
  } else {
    adduct_list_met <- lapply(sp_met, function(s) s$adduct[1])
    res_met <- run_sirius_role(sp_met, adduct_list_met, dir_met)
    write.csv(res_met, file.path(dir_res, "metabolite_results.csv"), row.names = FALSE)
    message("  Metabolite done: ", nrow(res_met), " rows written")
  }
}

message("All datasets processed.")
```

# Evaluate SIRIUS matches

The classifier below is moved here from `reverse-metabolomics-workflow.qmd`
and kept `eval: false` for manual review after inspecting the result tables.
Load the CSVs for a dataset, then call `classify_sirius_feature()` on each
feature group.

```{r classify-sirius, eval=false}
classify_sirius_feature <- function(
  df_feature,
  target_formula,
  target_ikey_prefix,
  norm_score_min = 0.5,
  tanimoto_min   = 0.6
) {
  top_formula_rows <- df_feature[df_feature$rank.x == 1, ]
  formula_match    <- any(top_formula_rows$molecularFormula == target_formula)
  norm_score       <- unique(top_formula_rows$siriusScoreNormalized)

  if (!formula_match) {
    any_rank <- df_feature[df_feature$molecularFormula == target_formula, ]
    if (nrow(any_rank) == 0 ||
        max(any_rank$siriusScoreNormalized, na.rm = TRUE) < 0.05) {
      return(data.frame(tier = "D", reason = "target formula not plausible"))
    }
    return(data.frame(tier = "C", reason = "target formula low-ranked"))
  }

  if (norm_score < norm_score_min) {
    return(data.frame(tier = "C",
                      reason = "formula match but low normalized score"))
  }

  structure_rows <- top_formula_rows[
    !is.na(top_formula_rows$inchiKey) & top_formula_rows$rank.y == 1, ]
  if (nrow(structure_rows) == 0) {
    return(data.frame(tier = "B", reason = "no CSI structure returned"))
  }

  ikey_match <- startsWith(structure_rows$inchiKey, target_ikey_prefix)
  tani       <- structure_rows$tanimotoSimilarity

  if (ikey_match && tani >= tanimoto_min) {
    return(data.frame(tier = "A",
                      reason = sprintf("top CSI hit is target (Tanimoto %.2f)", tani)))
  }
  if (ikey_match) {
    return(data.frame(tier = "B",
                      reason = sprintf("top CSI is target but Tanimoto low (%.2f)", tani)))
  }
  data.frame(tier = "B",
             reason = "formula matches but top structure is not target")
}

# Example: load results for one dataset and classify
ds      <- "MSV000085256"
res_par <- read.csv(file.path("results", ds, "parent_results.csv"))
res_met <- read.csv(file.path("results", ds, "metabolite_results.csv"))

tier_par <- lapply(
  split(res_par, res_par$sirius_fts),
  classify_sirius_feature,
  target_formula     = compound_spec$formula[compound_spec$role == "parent"],
  target_ikey_prefix = compound_spec$ikey1[compound_spec$role == "parent"]
)
tier_met <- lapply(
  split(res_met, res_met$sirius_fts),
  classify_sirius_feature,
  target_formula     = compound_spec$formula[compound_spec$role == "metabolite"],
  target_ikey_prefix = compound_spec$ikey1[compound_spec$role == "metabolite"]
)

tier_par
tier_met
```
