---
title: "MASST Hit Confirmation — RuSirius"
subtitle: "Per-dataset SIRIUS annotation of curated MASST hits (MassIVE)"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
execute:
  warning: false
  message: false
---

# Overview

This document takes the curated MASST hit table produced by
`reverse-metabolomics-workflow.qmd` (DIA datasets and datasets without
metadata already removed) and runs SIRIUS annotation for every remaining
MassIVE dataset.

For each dataset two SIRIUS projects are created:

```
sirius_projects/
  <dataset_id>/
    parent/
    metabolite/
results/
  <dataset_id>/
    parent_results.csv
    metabolite_results.csv
```

Results are written to disk for manual review. The classification
helper at the end of this document (`eval: false`) can be run
interactively after inspecting the raw tables.

# Version check

```{r versions}
#source("check-versions.R")
```

# Setup

```{r setup}
library(Spectra)
library(MsBackendMassIVE)
library(MetaboCoreUtils)
library(RuSirius)
library(dplyr)
library(tidyr)
```

# Compound targets

```{r compound-spec}
compound_spec <- data.frame(
  role    = c("parent",       "metabolite"),
  name    = c("omeprazole",   "5-hydroxyomeprazole"),
  formula = c("C17H19N3O3S",  "C17H19N3O4S"),
  ikey1   = c("SUBDBMMJDZJVOS", "CMZHQFXXAAIBKE")  
)
```

# Load curated hits

`hits_curated.csv` is written by `reverse-metabolomics-workflow.qmd`
after the DIA and no-metadata filters have been applied.

```{r load-hits}
hits_curated <- read.csv("hits_curated.csv", check.names = FALSE)
```

# Filter to MassIVE datasets

```{r filter-massive}
hits_massive <- hits_curated[
  startsWith(hits_curated$ATTRIBUTE_DatasetAccession, "MSV"), ]

massive_datasets <- unique(hits_massive$ATTRIBUTE_DatasetAccession)
cat(length(massive_datasets), "MassIVE datasets to process\n")
```

# Helper functions

```{r helpers}
ADDUCT_CANDIDATES <- c(
  "[M+H]+",
  "[M+Na]+",
  "[M+K]+",
  "[M+NH4]+",
  "[M-H]-",
  "[M+Cl]-"
)
PPM <- 20
RT_TOL <- 3 # seconds around hit RT

infer_adduct <- function(
  obs_mz,
  formula,
  ppm = PPM,
  candidates = ADDUCT_CANDIDATES
) {
  if (is.na(obs_mz) || is.na(formula)) {
    return(NA_character_)
  }
  mass <- calculateMass(formula)
  theo <- vapply(
    candidates,
    function(a) mass2mz(mass, adduct = a)[[1]],
    numeric(1)
  )
  diff_ppm <- abs(obs_mz - theo) / theo * 1e6
  best <- which.min(diff_ppm)
  if (diff_ppm[best] > ppm) NA_character_ else candidates[best]
}

# Returns a list of Spectra slices (one per MASST hit) for one role.
# Each slice contains the top MS1 scan + all MS2 scans within the RT/mz window.
slice_role <- function(hits_ds, sp, sp_file, sp_scan, role) {
  role_formula <- compound_spec$formula[compound_spec$role == role]
  h <- hits_ds[hits_ds$role == role & !is.na(hits_ds$rt), ]
  if (!nrow(h)) {
    return(list())
  }

  lapply(seq_len(nrow(h)), function(i) {
    hi <- h[i, ]
    mz_rng <- hi$Precursor_MZ * c(1 - PPM * 1e-6, 1 + PPM * 1e-6)
    rt_rng <- hi$rt + c(-RT_TOL, RT_TOL)

    sp_f <- filterRt(sp[sp_file == hi$file], rt_rng)
    ms1 <- filterMsLevel(sp_f, 1L)
    ms2 <- filterPrecursorMzRange(filterMsLevel(sp_f, 2L), mz_rng)

    int_at_mz <- vapply(
      peaksData(ms1),
      function(pm) {
        keep <- pm[, "mz"] >= mz_rng[1] & pm[, "mz"] <= mz_rng[2]
        if (any(keep)) max(pm[keep, "intensity"]) else 0
      },
      numeric(1)
    )
    ms1_top <- if (length(ms1) && max(int_at_mz) > 0) {
      ms1[which.max(int_at_mz)]
    } else {
      ms1[integer(0)]
    }

    out <- c(ms1_top, ms2)
    out$adduct <- rep(hi$adduct_inferred, length(out))
    out$group_idx <- rep(i, length(out))
    out
  })
}

run_sirius_role <- function(sp_slices, adduct_list, project_path, study_id) {
  # The MassIVE metadata index can be out of sync with the actual mzXML after
  # file re-uploads: the indexed RT/precursorMz looked valid but the file
  # content has no matching scan in that RT window. Drop empty slices.
  nonempty    <- vapply(sp_slices, length, integer(1)) > 0
  n_dropped   <- sum(!nonempty)
  sp_slices   <- sp_slices[nonempty]
  adduct_list <- adduct_list[nonempty]
  if (n_dropped > 0)
    message("  Dropped ", n_dropped,
            " empty slice(s): MassIVE index out of sync with file content")
  if (!length(sp_slices)) return(data.frame())

  srs <- Sirius(
    projectId = study_id,
    username = Sys.getenv("SIRIUS_USERNAME"),
    path = paste0(getwd(), "/", project_path),
    password = Sys.getenv("SIRIUS_PASSWORD")
  )

  # RuSirius does not support importing mixed-polarity spectra in one call.
  # Split slices by polarity (derived from adduct sign) and import sequentially;
  # only the first import clears existing features.
  pol   <- ifelse(endsWith(trimws(unlist(adduct_list)), "+"), "pos", "neg")
  first <- TRUE
  for (p in unique(pol)) {
    idx  <- which(pol == p)
    sp_p <- concatenateSpectra(sp_slices[idx])
    srs  <- import(
      sirius = srs,
      spectra = sp_p,
      adduct = adduct_list[idx],
      ms_column_name = "group_idx",
      deleteExistingFeatures = first
    )
    first <- FALSE
  }
  run(
    srs,
    formulaIdParams = formulaIdParam(),
    structureDbSearchParams = structureDbSearchParam(
      structureSearchDbs = "ALL"
    ),
    predictParams = predictParam(),
    wait = TRUE
  )
  res <- results(
    srs,
    result.type = "structureDb",
    return.type = "data.frame",
    topFormula = 15,
    topStructure = 15
  )
  shutdown(srs)
  res
}
```

# Per-dataset SIRIUS annotation

```{r per-dataset-loop}
role_formula <- setNames(compound_spec$formula, compound_spec$role)

for (ds in massive_datasets) {
  message("\n--- Processing ", ds, " ---")
  hits_ds <- hits_massive[hits_massive$ATTRIBUTE_DatasetAccession == ds, ]

  # Skip datasets already analyzed: every role present in hits_ds has its
  # results CSV on disk.
  dir_res <- file.path("results", ds)
  roles_present <- unique(hits_ds$role)
  expected_csv <- file.path(dir_res, paste0(roles_present, "_results.csv"))
  if (length(expected_csv) && all(file.exists(expected_csv))) {
    message("  Already analyzed — skipping download and SIRIUS run")
    next
  }

  # Infer adducts from precursor m/z
  hits_ds$adduct_inferred <- mapply(
    infer_adduct,
    hits_ds$Precursor_MZ,
    role_formula[hits_ds$role],
    USE.NAMES = FALSE
  )
  # Hits where no candidate adduct matches within PPM cannot be passed to
  # SIRIUS (unknown polarity and mass offset). Record the reason and drop them.
  hits_ds$scan_verified <- NA_character_
  na_adduct <- is.na(hits_ds$adduct_inferred)
  if (any(na_adduct)) {
    hits_ds$scan_verified[na_adduct] <- "no_adduct_match"
    message("  ", sum(na_adduct), " hit(s) dropped: precursor mz unmatched ",
            "to any candidate adduct within ", PPM, " ppm")
    hits_ds <- hits_ds[!na_adduct, ]
  }

  # Output directories
  dir_par <- file.path("sirius_projects", ds, "parent")
  dir_met <- file.path("sirius_projects", ds, "metabolite")
  dir.create(dir_par, recursive = TRUE, showWarnings = FALSE)
  dir.create(dir_met, recursive = TRUE, showWarnings = FALSE)
  dir.create(dir_res, recursive = TRUE, showWarnings = FALSE)

  # Download MASST-hit files for this dataset
  wanted <- unique(basename(hits_ds$file_path))
  massive_sync_data_files(ds, fileName = wanted)
  sp       <- Spectra(backendInitialize(MsBackendMassIVE(), massiveId = ds,
                                        offline = TRUE)) ## true because we don't want to download more files. 
  sp_file  <- basename(dataOrigin(sp))
  sp_scan  <- acquisitionNum(sp)

  hits_ds$file    <- basename(hits_ds$file_path)
  hit_idx         <- match(
    paste(hits_ds$file, hits_ds$scan_id, sep = "::"),
    paste(sp_file,      sp_scan,          sep = "::")
  )
  hits_ds$rt <- rtime(sp)[hit_idx]

  # Verify that the matched scan's precursorMz agrees with the MASST hit.
  # Some files are re-uploaded with different scan numbering after indexing,
  # making acquisitionNum unreliable. Fall back to mz-based search when needed.
  obs_pmz  <- precursorMz(sp)[hit_idx]
  ppm_err  <- abs(obs_pmz - hits_ds$Precursor_MZ) / hits_ds$Precursor_MZ * 1e6
  mismatch <- is.na(hit_idx) | is.na(obs_pmz) | (!is.na(ppm_err) & ppm_err > PPM)
  hits_ds$scan_verified <- ifelse(mismatch, "mismatch", "ok")

  if (any(mismatch)) {
    message("  ", sum(mismatch), " hit(s) with scan_id/precursorMz mismatch — ",
            "attempting mz fallback")
    extra_rows <- list()
    drop_rows  <- integer(0)
    for (j in which(mismatch)) {
      hi_j  <- hits_ds[j, ]
      sp_j  <- sp[sp_file == hi_j$file]
      pmz_j <- precursorMz(sp_j)
      cands <- which(!is.na(pmz_j) &
                     abs(pmz_j - hi_j$Precursor_MZ) / hi_j$Precursor_MZ * 1e6 < PPM)
      if (length(cands) == 1) {
        hits_ds$rt[j]            <- rtime(sp_j)[cands]
        hits_ds$scan_verified[j] <- "mz_fallback"
        message("    row ", j, ": mz fallback OK (rt=",
                round(rtime(sp_j)[cands], 1), "s)")
      } else if (length(cands) == 0) {
        hits_ds$rt[j]            <- NA_real_
        hits_ds$scan_verified[j] <- "mz_not_found"
        message("    row ", j, ": no scan with matching mz — skipping")
      } else {
        # Multiple elution events for the same precursor mz: expand to one row
        # per candidate so each gets its own RT window and SIRIUS feature.
        drop_rows <- c(drop_rows, j)
        exp <- hits_ds[rep(j, length(cands)), ]
        exp$rt            <- rtime(sp_j)[cands]
        exp$scan_verified <- "mz_expanded"
        extra_rows[[length(extra_rows) + 1]] <- exp
        message("    row ", j, ": ", length(cands),
                " candidates — expanded to ", length(cands), " slices")
      }
    }
    if (length(drop_rows))
      hits_ds <- rbind(hits_ds[-drop_rows, ],
                       do.call(rbind, extra_rows))
  }

  # Slice spectra per role
  sp_par <- slice_role(hits_ds, sp, sp_file, sp_scan, "parent")
  sp_met <- slice_role(hits_ds, sp, sp_file, sp_scan, "metabolite")

  if (!length(sp_par)) {
    message("  No parent spectra — skipping parent SIRIUS run")
  } else {
    adduct_list_par <- lapply(sp_par, function(s) s$adduct[1])
    res_par <- run_sirius_role(sp_par, adduct_list_par, dir_par, study_id = "parent")
    write.csv(res_par, file.path(dir_res, "parent_results.csv"), row.names = FALSE)
    message("  Parent done: ", nrow(res_par), " rows written")
  }

  if (!length(sp_met)) {
    message("  No metabolite spectra — skipping metabolite SIRIUS run")
  } else {
    adduct_list_met <- lapply(sp_met, function(s) s$adduct[1])
    res_met <- run_sirius_role(sp_met, adduct_list_met, dir_met,study_id = "metabolite")
    write.csv(res_met, file.path(dir_res, "metabolite_results.csv"), row.names = FALSE)
    message("  Metabolite done: ", nrow(res_met), " rows written")
  }
}

message("All datasets processed.")
```

this mistmatch i have between reported scan and [precursor mz is somethign we could probably discuss in a paper too.. 

# Evaluate SIRIUS matches

## Helper functions

Each SIRIUS feature is assigned a **confidence level** based on where the
target formula and structure rank in the SIRIUS output:

| Confidence | Formula rank | Structure rank |
|------------|-------------|----------------|
| strict     | 1           | 1              |
| standard   | ≤ 3         | ≤ 5            |
| liberal    | ≤ 15        | ≤ 15           |
| not_confirmed | target formula absent or ranks beyond liberal thresholds | |

Features that don't meet even the liberal thresholds are `not_confirmed`.

```{r resolve-helpers}
library(dplyr)
library(ggplot2)

CONFIDENCE_LEVELS <- c("strict", "standard", "liberal", "not_confirmed")
CONFIDENCE_COLOURS <- c(
  strict      = "#1A7837",
  standard    = "#74C476",
  liberal     = "#FEC44F",
  not_confirmed = "#D7191C"
)

#' Resolve target features in one SIRIUS results table (one dataset, one role)
resolve_features <- function(res,
                             target_formula,
                             target_ikey_prefix) {

  target_rows <- res |>
    filter(molecularFormula == target_formula,
           !is.na(inchiKey),
           startsWith(inchiKey, target_ikey_prefix)) |>
    group_by(xcms_fts) |>
    slice_min(rank.y, n = 1, with_ties = FALSE) |>
    ungroup() |>
    transmute(
      xcms_fts,
      formula_rank   = rank.x,
      formula_score  = round(siriusScoreNormalized, 3),
      structure_rank = rank.y,
      tanimoto       = round(tanimotoSimilarity, 3),
      csi_score      = round(csiScore, 1),
      adduct,
      structure_name = structureName,
      target_present = TRUE
    )

  res |>
    distinct(xcms_fts) |>
    left_join(target_rows, by = "xcms_fts") |>
    mutate(
      target_present = !is.na(formula_rank),
      confidence = case_when(
        formula_rank == 1  & structure_rank == 1  ~ "strict",
        formula_rank <= 3  & structure_rank <= 5  ~ "standard",
        formula_rank <= 15 & structure_rank <= 15 ~ "liberal",
        TRUE                                       ~ "not_confirmed"
      ),
      confidence = factor(confidence, levels = CONFIDENCE_LEVELS)
    ) |>
    arrange(confidence, formula_rank, structure_rank)
}

#' Run resolve_features across all (dataset, role) result CSVs
resolve_all <- function(results_dir = "results", compound_spec) {

  datasets <- list.dirs(results_dir, recursive = FALSE, full.names = FALSE)
  all_rows <- list()

  for (ds in datasets) {
    for (i in seq_len(nrow(compound_spec))) {
      role     <- compound_spec$role[i]
      csv_path <- file.path(results_dir, ds, paste0(role, "_results.csv"))
      if (!file.exists(csv_path)) next
      res <- read.csv(csv_path, check.names = FALSE)
      if (!nrow(res)) next

      resolve_features(
        res,
        target_formula     = compound_spec$formula[i],
        target_ikey_prefix = compound_spec$ikey1[i]
      ) |>
        mutate(dataset_id = ds, role = role, .before = xcms_fts) ->
        all_rows[[paste(ds, role, sep = "::")]]
    }
  }

  features <- bind_rows(all_rows)

  summary_tbl <- features |>
    group_by(dataset_id, role) |>
    summarise(
      n_total          = n(),
      n_target_present = sum(target_present, na.rm = TRUE),
      n_strict         = sum(confidence == "strict",      na.rm = TRUE),
      n_standard       = sum(confidence == "standard",    na.rm = TRUE),
      n_liberal        = sum(confidence == "liberal",     na.rm = TRUE),
      .groups = "drop"
    )

  list(features = features, summary = summary_tbl)
}
```

## Run evaluation

```{r resolve-run}
resolved    <- resolve_all(results_dir = "results", compound_spec = compound_spec)
features    <- resolved$features
summary_tbl <- resolved$summary
summary_tbl
```

## Plots

### Confidence levels per dataset and role

```{r plot-confidence, fig.width=9, fig.height=4}
features |>
  count(dataset_id, role, confidence) |>
  ggplot(aes(x = dataset_id, y = n, fill = confidence)) +
  geom_col() +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5),
            size = 3, colour = "white", fontface = "bold") +
  facet_wrap(~role) +
  scale_fill_manual(values = CONFIDENCE_COLOURS, drop = FALSE) +
  labs(x = NULL, y = "Number of features", fill = "Confidence",
       title = "SIRIUS confirmation confidence per dataset") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

### Formula and structure rank coloured by confidence

```{r plot-ranks, fig.width=6, fig.height=5}
features |>
  filter(target_present) |>
  ggplot(aes(x = formula_rank, y = structure_rank,
             colour = confidence, shape = role)) +
  annotate("rect", xmin = 0.5, xmax = 1.5, ymin = 0.5, ymax = 1.5,
           fill = "#1A7837", alpha = 0.08) +
  annotate("rect", xmin = 0.5, xmax = 3.5, ymin = 0.5, ymax = 5.5,
           fill = "#74C476", alpha = 0.06) +
  annotate("rect", xmin = 0.5, xmax = 6, ymin = 0.5, ymax = 5,
           fill = "#FEC44F", alpha = 0.04) +
  geom_jitter(size = 3, width = 0.2, height = 0.2, alpha = 0.9) +
  scale_colour_manual(values = CONFIDENCE_COLOURS, drop = FALSE) +
  scale_x_continuous(breaks = c(1, 3, 5)) +
  scale_y_continuous(breaks = c(1, 5)) +
  labs(x = "Formula rank", y = "Structure rank",
       colour = "Confidence", shape = "Role",
       title = "Rank of target among confirmed features") +
  theme_bw() +
  theme(legend.position = "bottom")
```

### Score scatter coloured by confidence

```{r plot-scores, fig.width=7, fig.height=4}
features |>
  filter(target_present) |>
  ggplot(aes(x = tanimoto, y = formula_score,
             colour = confidence, shape = role)) +
  geom_point(size = 3, alpha = 0.85) +
  scale_colour_manual(values = CONFIDENCE_COLOURS, drop = FALSE) +
  labs(x = "Tanimoto similarity", y = "Normalised formula score",
       colour = "Confidence", shape = "Role",
       title = "Score distribution (target-present features)") +
  theme_bw() +
  theme(legend.position = "bottom")
```


I'm not very happy to get rid of one dataset. but i could be because of the matching spectrum quality ? i do not know why the cosine match is so good but SIrius cannotanntoate it.. i could feed it the formula butthis would be biased.

# Write confirmed hits

Hits whose SIRIUS confidence is `strict` or `standard` are kept as
MassIVE-confirmed hits.

```{r write-confirmed}
hits_confirmed_massive <- features |>
  filter(confidence %in% c("strict", "standard"))

write.csv(hits_confirmed_massive, "hits-confirmed-massive.csv",
          row.names = FALSE)
```


