---
title: "Reverse Metabolomics — Part 1: Dataset Selection & Single-Dataset Prototype"
subtitle: "Omeprazole / 5-hydroxyomeprazole · CYP2C19 ratio analysis"
author: "Your Name"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
execute:
  eval: false
  warning: false
  message: false
---

# Overview

This document covers the first half of the pipeline: loading the two
MASST exports, selecting datasets with both compounds, tiering them by
study-design suitability, picking one prototype, and running the full
single-dataset analysis (EIC extraction, peak detection, and
parent/metabolite ratio statistics).

Part 2 (`all-datasets-analysis.qmd`) scales the same analysis to all
Tier-B datasets.

```{mermaid}
%%| label: fig-part1
%%| fig-cap: "Part 1 pipeline — from search results to single-dataset ratio statistics."
flowchart LR
  A["Search results\n2 compound files"] --> B["Shared datasets\n26 found"]
  B --> C["Summary table\npick prototype manually"]
  C --> D["Access raw spectra\nno full download"]
  D --> E["Verify compound\nidentity"]
  E --> F["Extract signal\nboth compounds"]
  F --> G["Detect peaks · join samples\nparent / metabolite ratio"]
```

## why Omeprazol
Omeprazole is widely used for CYP2C19 phenotyping. When administered as a
 racemic mixture, (R-S)-omeprazole undergoes stereoselective metabolism and
 is primarily metabolized to (R)-5-hydroxyomeprazole via CYP2C19/CYP3A and
 (S)-5-hydroxyomeprazole via CYP2C19. Cytochrome P450 2C19 (CYP2C19) enzyme is estimated to metabolize approximately 15% of all marketed drugs, among which proton pump inhibitors, antiplatelet agents, antidepressants, beta-blockers, benzodiazepines and many other compounds
https://onlinelibrary.wiley.com/doi/full/10.1111/bcpt.14095

could also add a second probe pair (e.g., dextromethorphan/dextrorphan for CYP2D6. to show that it can be applied ot others.

title such as: Quantitative post-MASST reverse metabolomics enables pharmacoepidemiology of CYP2C19 activity across public repositories

# Setup

```{r setup}
library(Spectra)
library(MsBackendMassIVE)
library(xcms)
library(dplyr)
library(ggplot2)

# remotes::install_github("<your-org>/RuSirius")   # in review at JOSS
library(RuSirius)
```

# Compound pair

| role | compound | formula | monoisotopic mass | InChIKey block 1 |
|---|---|---|---|---|
| parent | omeprazole | C17H19N3O3S | 345.1147 | SUBDBMMJDZJVOS |
| metabolite | 5-hydroxyomeprazole | C17H19N3O4S | 361.1096 | CMZHQFXXAAIBKE |

The parent/5-OH ratio is a classical CYP2C19 phenotyping readout.
Both compounds are queried in positive and negative mode; downstream
code picks the adduct with the most hits per dataset and collapses
adducts so each compound is a single feature per sample.

```{r compound-spec}
compound_spec <- data.frame(
  role    = c("parent",       "metabolite"),
  name    = c("omeprazole",   "5-hydroxyomeprazole"),
  formula = c("C17H19N3O3S",  "C17H19N3O4S"),
  mz_pos  = c(346.1220,       362.1169)   # [M+H]+, are these rigth
)
```

# MASST exports

Two StructureMASST queries on GNPS2 with identical settings
(cosine ≥ 0.7, matched peaks ≥ 3, analog search off, both polarities).

| compound | GNPS2 session |
|---|---|
| omeprazole | `2399487d18d6` |
| 5-hydroxyomeprazole | `4c9e1b910599` |

```{r load-hits}
read_masst <- function(path, role) {
  x <- read.csv(path, check.names = FALSE)
  x[x == "missing value"] <- NA
  parts        <- strsplit(x$USI, ":", fixed = TRUE)
  x$file_path  <- vapply(parts, `[`, character(1), 3)
  x$scan_str   <- vapply(parts, \(p) p[length(p)], character(1))  # keep as character
  x$role       <- role
  x
}

all_hits <- rbind(
  read_masst("Omeoprazole.csv",          "parent"),
  read_masst("5-hydroxyomeprazole.csv",  "metabolite")
)
```

## Remove corrupted USIs

Some USIs carry the scan number `4294967295` — the maximum value of a
32-bit unsigned integer. This value is written by certain mzML converters when
the scan index is missing or could not be resolved. We remove them as they
cannot give us the position of the spectrum of interest in the raw file.

```{r filter-corrupt-usi}
INVALID_SCAN_STR <- "4294967295"

n_corrupt <- sum(all_hits$scan_str == INVALID_SCAN_STR, na.rm = TRUE)
cat("Corrupted USIs removed:", n_corrupt,
    "(", round(100 * n_corrupt / nrow(all_hits), 1), "% of all hits)\n")

all_hits           <- all_hits[all_hits$scan_str != INVALID_SCAN_STR, ]
all_hits$scan_id   <- as.integer(all_hits$scan_str)   # now safe to convert
all_hits$scan_str  <- NULL                             # drop the staging column
```

## Infer adduct from observed precursor m/z

MASST reports an `Adduct` column, but it's inherited from whichever library
spectrum matched and does not always reflect how *our* compound ionised in
the query data. We know the target formula per role, so for each hit we
compute the theoretical m/z of a set of common adducts and pick the one
closest to the observed `Precursor_MZ` within a ppm tolerance. This
ground-truthed adduct is what we feed to SIRIUS later.

```{r infer-adduct}
library(MetaboCoreUtils)

ADDUCT_CANDIDATES <- c(
  "[M+H]+", "[M+Na]+", "[M+K]+", "[M+NH4]+",
  "[M-H]-", "[M+Cl]-"
)

infer_adduct <- function(obs_mz, formula, ppm = 20,
                         candidates = ADDUCT_CANDIDATES) {
  if (is.na(obs_mz) || is.na(formula)) return(NA_character_)
  mass <- calculateMass(formula)
  theo <- vapply(candidates, function(a) mass2mz(mass, adduct = a)[[1]],
                 numeric(1))
  diff_ppm <- abs(obs_mz - theo) / theo * 1e6
  best <- which.min(diff_ppm)
  if (diff_ppm[best] > ppm) NA_character_ else candidates[best]
}

role_formula <- setNames(compound_spec$formula, compound_spec$role)

all_hits$adduct_inferred <- mapply(
  infer_adduct,
  all_hits$Precursor_MZ,
  role_formula[all_hits$role],
  USE.NAMES = FALSE
)

# Coverage and adduct breakdown per role
table(role   = all_hits$role,
      adduct = all_hits$adduct_inferred,
      useNA  = "ifany")
```

# Dataset selection

## D_both — datasets where both compounds appear

```{r d-both}
D_parent     <- unique(all_hits$ATTRIBUTE_DatasetAccession[all_hits$role == "parent"])
D_metabolite <- unique(all_hits$ATTRIBUTE_DatasetAccession[all_hits$role == "metabolite"])
D_both       <- intersect(D_parent, D_metabolite)

cat("parent:", length(D_parent), "| metabolite:", length(D_metabolite),
    "| both:", length(D_both), "\n")
```

Pre-computed: 53 parent datasets, 30 metabolite datasets, **26 in both**
(26/30 metabolite datasets also contain the parent — high co-occurrence).

## Summary table for prototype selection

For each overlapping dataset, compute the number of hits per compound,
the number of unique files (a proxy for sample count), and how
completely the key Pan-ReDU metadata fields are filled. Sort by file
count ascending so small, manageable datasets appear first.

```{r dataset-summary, eval = FALSE}
dataset_overlap <- all_hits |>
  filter(ATTRIBUTE_DatasetAccession %in% D_both)
write.csv(dataset_overlap, "overlap_dataset.xlsx", row.names = FALSE)
```

After that do some metadata work. add columns, do not remove any data.

## remove datasets without metadata

```{r}
## there's a column for this
dataset_overlap <- readxl::read_xlsx("overlap_dataset.xlsx")
```


## remove DIA/ SWATH Datasets.

```{r}
## from manual curation.
dataset_overlap <- dataset_overlap[dataset_overlap$`DIA/SWATH` == FALSE, ]
```

## Write curated hits

The curated table is read by `masst-hit-confirmation.qmd` for SIRIUS
annotation.

```{r write-curated}
hits_curated <- dataset_overlap[dataset_overlap$extra_metadata == TRUE, ]
write.csv(hits_curated, "hits_curated.csv", row.names = TRUE)

unique(hits_curated$ATTRIBUTE_DatasetAccession) |> length()
```

We end up with 13 datasets for now. Let's see after pooling of metadata and
hity confirmation.

# Metadata coverage map

Build a *(dataset × variable)* feasibility matrix from the per-dataset
files in `dataset_metadata/`, restricted to the curated datasets. For
each canonical biological/clinical variable (sex, age, health status,
body site, country, …), the cell holds either
`"source_column (coverage%)"` or `FALSE` if the variable is absent.
This is the feasibility view for downstream meta-analysis.

The folder mixes three formats:

* MassIVE (`MSV*.tsv|.txt|.csv`) — flat PanReDU-style table.
* MetaboLights (`s_MTBLS*.txt`) — flat ISA-Tab; column headers wrap
  the variable name in `Characteristics[...]` / `Factor Value[...]`.
* Metabolomics Workbench (`ST*_AN*.txt`) — multi-section mwTab; per-sample
  factors live on `SUBJECT_SAMPLE_FACTORS` lines as `Key:Value | Key:Value`.
  Multiple AN files per study are collapsed to one row.

Assay-only files (`a_MTBLS*.txt`) are not needed — technical metadata
is already in the MASST hit table.

```{r metadata-helpers}
# Parse one metadata file → list(format, source_columns = data.frame(name, coverage_pct))
parse_metadata <- function(path) {
  fn <- basename(path)

  if (startsWith(fn, "MSV")) {
    sep <- if (grepl("\\.csv$", fn)) "," else "\t"
    df  <- read.table(path, header = TRUE, sep = sep, quote = "\"",
                      check.names = FALSE, comment.char = "",
                      fill = TRUE, stringsAsFactors = FALSE)
    cov <- vapply(df, function(x) {
      v <- as.character(x)
      mean(!is.na(v) & nzchar(v) & !v %in% c("not specified", "not applicable", "NA")) * 100
    }, numeric(1))
    return(list(format = "MassIVE",
                cols = data.frame(name = names(cov), coverage_pct = round(cov, 1),
                                  row.names = NULL)))
  }

  if (startsWith(fn, "s_MTBLS")) {
    df <- read.table(path, header = TRUE, sep = "\t", quote = "\"",
                     check.names = FALSE, comment.char = "",
                     fill = TRUE, stringsAsFactors = FALSE)
    keep <- !names(df) %in% c("Term Source REF", "Term Accession Number",
                              "Protocol REF", "Unit")
    df <- df[, keep, drop = FALSE]
    cov <- vapply(df, function(x) {
      v <- as.character(x)
      mean(!is.na(v) & nzchar(v)) * 100
    }, numeric(1))
    return(list(format = "MetaboLights",
                cols = data.frame(name = names(cov), coverage_pct = round(cov, 1),
                                  row.names = NULL)))
  }

  if (startsWith(fn, "ST")) {
    lines <- readLines(path, warn = FALSE)
    fac_lines <- grep("^SUBJECT_SAMPLE_FACTORS\\s", lines, value = TRUE)
    if (!length(fac_lines)) {
      return(list(format = "MetabolomicsWorkbench",
                  cols = data.frame(name = character(0), coverage_pct = numeric(0))))
    }
    factors_field <- vapply(strsplit(fac_lines, "\t"), function(p) {
      if (length(p) >= 4) p[4] else NA_character_
    }, character(1))
    pairs    <- unlist(strsplit(factors_field, "\\s*\\|\\s*"))
    keys     <- sub(":.*$", "", pairs)
    values   <- sub("^[^:]*:\\s*", "", pairs)
    keep     <- nzchar(keys) & !is.na(keys)
    keys     <- trimws(keys[keep]); values <- trimws(values[keep])
    n_samples <- length(fac_lines)
    cov <- tapply(values, keys, function(v) {
      mean(nzchar(v) & !v %in% c("-", "NA", "not specified")) * 100
    }) * (table(keys) / n_samples)
    cov <- pmin(cov, 100)
    return(list(format = "MetabolomicsWorkbench",
                cols = data.frame(name = names(cov),
                                  coverage_pct = round(as.numeric(cov), 1),
                                  row.names = NULL)))
  }

  list(format = NA_character_,
       cols   = data.frame(name = character(0), coverage_pct = numeric(0)))
}

CANONICAL_PATTERNS <- list(
  # BiologicalSex, sex, ATTRIBUTE_sex (MSV); Gender (MWB ST002044)
  sex           = "sex|gender",
  # AgeInYears, age, age_at_* (MSV); LifeStage (PanReDU)
  age           = "^age|life.?stage",
  # HealthStatus, DOIDCommonName (PanReDU); Factor Value[Cohort] (MTBLS11656);
  # Comment[Patient status] (MTBLS1866); ATTRIBUTE_status (MSV092757);
  # Phenotype (MWB ST002722)
  health_status = "healthstatus|\\bhealth\\b|disease|diagnos|condition|cohort|patient.?status|doid|phenotype|attribute_status",
  # UBERONBodyPartName (PanReDU); Characteristics[Organism part] (MTBLS);
  # SampleType, SampleTypeSub1 (PanReDU) — biological matrix / anatomical site
  body_site     = "uberon|organism.?part|sampletype",
  country       = "country|geograph",
  ethnicity     = "ethnic|race|ancestry",
  bmi           = "\\bbmi\\b",
  weight        = "\\bweight\\b",
  height        = "\\bheight\\b",
  treatment     = "treatment|intervention|factor.?value\\[treatment\\]|factor.?value\\[cohort\\]"
)

map_canonical <- function(col_name) {
  nm <- tolower(col_name)
  for (canon in names(CANONICAL_PATTERNS)) {
    if (grepl(CANONICAL_PATTERNS[[canon]], nm, perl = TRUE)) return(canon)
  }
  NA_character_
}

# MSV000085256_metadata.tsv → MSV000085256
# s_MTBLS11656.txt          → MTBLS11656
# ST001715_AN002792.txt     → ST001715
dataset_id_from_file <- function(fn) {
  fn <- basename(fn)
  if (startsWith(fn, "MSV"))     return(sub("_.*$", "", fn))
  if (startsWith(fn, "s_MTBLS")) return(sub("^s_", "", sub("\\.txt$", "", fn)))
  if (startsWith(fn, "ST"))      return(sub("_.*$", "", fn))
  tools::file_path_sans_ext(fn)
}
```

```{r metadata-inventory}
curated_ids <- unique(hits_curated$ATTRIBUTE_DatasetAccession)

meta_files <- list.files("dataset_metadata", full.names = TRUE)
meta_files <- meta_files[!startsWith(basename(meta_files), "a_MTBLS")]
meta_files <- meta_files[vapply(meta_files,
                                function(f) dataset_id_from_file(f) %in% curated_ids,
                                logical(1))]

inventory <- do.call(rbind, lapply(meta_files, function(f) {
  parsed <- parse_metadata(f)
  if (!nrow(parsed$cols)) return(NULL)
  data.frame(
    dataset_id    = dataset_id_from_file(f),
    source_file   = basename(f),
    format        = parsed$format,
    source_column = parsed$cols$name,
    coverage_pct  = parsed$cols$coverage_pct,
    canonical     = vapply(parsed$cols$name, map_canonical, character(1)),
    row.names     = NULL,
    stringsAsFactors = FALSE
  )
}))

# Collapse multiple files per dataset (e.g. ST*_AN* analyses) — pick the
# best-covered source column per (dataset, canonical) pair.
inv_canon <- inventory[!is.na(inventory$canonical), ]
inv_best  <- inv_canon |>
  group_by(dataset_id, canonical) |>
  slice_max(coverage_pct, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(cell = sprintf("%s (%.0f%%)", source_column, coverage_pct))

write.csv(inventory, "metadata_inventory.csv", row.names = FALSE)
```

```{r metadata-coverage-wide}
canonical_vars <- names(CANONICAL_PATTERNS)

coverage_wide <- inv_best |>
  select(dataset_id, canonical, cell) |>
  tidyr::pivot_wider(names_from = canonical, values_from = cell) |>
  tidyr::complete(dataset_id = curated_ids)

for (v in canonical_vars) {
  if (!v %in% names(coverage_wide)) coverage_wide[[v]] <- NA_character_
}
coverage_wide <- coverage_wide[, c("dataset_id", canonical_vars)]
for (v in canonical_vars) {
  coverage_wide[[v]][is.na(coverage_wide[[v]])] <- "FALSE"
}

write.csv(coverage_wide, "metadata_coverage.csv", row.names = FALSE)
coverage_wide
```

# MASST hits confirmation

SIRIUS confirmation is done in `masst-hit-confirmation.qmd`.
Results per dataset are written to `results/<dataset_id>/`.

Load back and se ehow many dataset we are left with. 


# Single-dataset prototype

## Stage 2: download remaining files for quantification

Once both gates have cleared, fetch the rest of the dataset — every
file needed for EIC peak detection — minus anything that looks like a
blank or QC. This is the only point where the full per-dataset
bandwidth cost is paid, and only for datasets that already passed
SIRIUS confirmation and the metadata check.

The MASST-hit files were synced in stage 1 and remain in the cache, so
this call only downloads the *new* files. Re-initialising the backend
picks up everything currently cached for the dataset.

```{r stage2-download}
all_files   <- massive_list_files(dataset_proto)
quant_files <- basename(all_files)
quant_files <- quant_files[!grepl("BLANK|blank", quant_files)]

massive_sync_data_files(dataset_proto, fileName = quant_files)

sp_full <- Spectra(backendInitialize(
  MsBackendMassIVE(), massiveId = dataset_proto, offline = TRUE
))
cat(length(unique(dataOrigin(sp_full))), "files (stage 2)\n")
```

## EIC extraction and peak detection

Both compounds are extracted in a single MS1 read pass over the full
sample set fetched in stage 2.

```{r eic-peaks}
mz_win <- function(hits_sub, ppm = 10) {
  mz <- median(hits_sub$Precursor_MZ, na.rm = TRUE)
  mz * c(1 - ppm * 1e-6, 1 + ppm * 1e-6)
}

rt_win <- function(conf, tol = 9) {
  c(min(conf$rt_confirmed) - tol, max(conf$rt_confirmed) + tol)
}

sp_ms1 <- filterMsLevel(sp_full, 1L)

# below use C
peaks_par <- extract_peaks(conf_par)
peaks_met <- extract_peaks(conf_met)
```

## Sample table

```{r sample-table}
make_tier <- function(peaks, conf) {
  data.frame(file = basename(files)) |>
    left_join(peaks |> select(file, area = into), by = "file") |>
    mutate(tier = case_when(
      file %in% conf$confirmed_files & !is.na(area) ~ "Tier1",
      !is.na(area)                                   ~ "Tier2",
      TRUE                                            ~ "Tier3"
    ))
}

meta <- hits_proto |>
  distinct(file = basename(file_path), HealthStatus, UBERONBodyPartName, Country)

sample_tbl <- make_tier(peaks_par, conf_par) |>
  rename(parent_area = area, parent_tier = tier) |>
  full_join(
    make_tier(peaks_met, conf_met) |> rename(metab_area = area, metab_tier = tier),
    by = "file"
  ) |>
  left_join(meta, by = "file") |>
  mutate(log2_ratio = log2(parent_area / metab_area))

count(sample_tbl, parent_tier, metab_tier)
```

## Statistics

```{r stats}
# Detection rates — Fisher's exact per compound
fisher.test(table(sample_tbl$HealthStatus, !is.na(sample_tbl$parent_area)))
fisher.test(table(sample_tbl$HealthStatus, !is.na(sample_tbl$metab_area)))

# Ratio — Wilcoxon on samples where both are detected
both <- sample_tbl[!is.na(sample_tbl$log2_ratio), ]
wilcox.test(log2_ratio ~ HealthStatus, data = both)
```

## Plots

```{r plots}
ggplot(both, aes(HealthStatus, log2_ratio)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(colour = parent_tier), width = 0.1, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey40") +
  labs(y = "log2(omeprazole / 5-OH-omeprazole)",
       title = dataset_proto, colour = "parent tier") +
  theme_bw()

ggplot(sample_tbl, aes(parent_area, metab_area, colour = HealthStatus)) +
  geom_point(alpha = 0.75) +
  scale_x_log10() + scale_y_log10() +
  labs(x = "omeprazole area", y = "5-OH-omeprazole area") +
  theme_bw()
```

# Sanity check

Does the result match the pharmacology literature? For this compound
pair, CYP2C19-extensive metabolisers show a lower parent/5-OH ratio
than poor metabolisers. If the prototype dataset is not CYP2C19-
stratified or PPI-intervention designed, re-select `dataset_proto`.

# References {.unnumbered}

::: {#refs}
:::
