---
title: "Reverse Metabolomics — Part 1: Dataset Selection & Single-Dataset Prototype"
subtitle: "Omeprazole / 5-hydroxyomeprazole · CYP2C19 ratio analysis"
author: "Your Name"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
execute:
  warning: false
  message: false
---

# Overview

This document covers the first half of the pipeline: loading the two
MASST exports, selecting datasets with both compounds, tiering them by
study-design suitability, picking one prototype, and running the full
single-dataset analysis (EIC extraction, peak detection, and
parent/metabolite ratio statistics).

Part 2 (`all-datasets-analysis.qmd`) scales the same analysis to all
Tier-B datasets.

```{mermaid}
%%| label: fig-part1
%%| fig-cap: "Part 1 pipeline — from search results to single-dataset ratio statistics."
flowchart LR
  A["Search results\n2 compound files"] --> B["Shared datasets\n26 found"]
  B --> C["Summary table\npick prototype manually"]
  C --> D["Access raw spectra\nno full download"]
  D --> E["Verify compound\nidentity"]
  E --> F["Extract signal\nboth compounds"]
  F --> G["Detect peaks · join samples\nparent / metabolite ratio"]
```

## why Omeprazol
Omeprazole is widely used for CYP2C19 phenotyping. When administered as a
 racemic mixture, (R-S)-omeprazole undergoes stereoselective metabolism and
 is primarily metabolized to (R)-5-hydroxyomeprazole via CYP2C19/CYP3A and
 (S)-5-hydroxyomeprazole via CYP2C19. Cytochrome P450 2C19 (CYP2C19) enzyme is estimated to metabolize approximately 15% of all marketed drugs, among which proton pump inhibitors, antiplatelet agents, antidepressants, beta-blockers, benzodiazepines and many other compounds
https://onlinelibrary.wiley.com/doi/full/10.1111/bcpt.14095

could also add a second probe pair (e.g., dextromethorphan/dextrorphan for CYP2D6. to show that it can be applied ot others.

title such as: Quantitative post-MASST reverse metabolomics enables pharmacoepidemiology of CYP2C19 activity across public repositories

# Setup

```{r setup}
library(Spectra)
library(MsBackendMassIVE)
library(xcms)
library(dplyr)
library(ggplot2)

library(RuSirius)
```

# Compound pair

| role | compound | formula | monoisotopic mass | InChIKey block 1 |
|---|---|---|---|---|
| parent | omeprazole | C17H19N3O3S | 345.1147 | SUBDBMMJDZJVOS |
| metabolite | 5-hydroxyomeprazole | C17H19N3O4S | 361.1096 | CMZHQFXXAAIBKE |

The parent/5-OH ratio is a classical CYP2C19 phenotyping readout.
Both compounds are queried in positive and negative mode; downstream
code picks the adduct with the most hits per dataset and collapses
adducts so each compound is a single feature per sample.

```{r compound-spec}
compound_spec <- data.frame(
  role    = c("parent",       "metabolite"),
  name    = c("omeprazole",   "5-hydroxyomeprazole"),
  formula = c("C17H19N3O3S",  "C17H19N3O4S"),
  mz_pos  = c(346.1220,       362.1169)   # [M+H]+, are these rigth
)
```

# MASST exports

Two StructureMASST queries on GNPS2 with identical settings
(cosine ≥ 0.7, matched peaks ≥ 3, analog search off, both polarities).

| compound | GNPS2 session |
|---|---|
| omeprazole | `2399487d18d6` |
| 5-hydroxyomeprazole | `4c9e1b910599` |

```{r load-hits}
read_masst <- function(path, role) {
  x <- read.csv(path, check.names = FALSE)
  x[x == "missing value"] <- NA
  parts        <- strsplit(x$USI, ":", fixed = TRUE)
  x$file_path  <- vapply(parts, `[`, character(1), 3)
  x$scan_str   <- vapply(parts, \(p) p[length(p)], character(1))  # keep as character
  x$role       <- role
  x
}

all_hits <- rbind(
  read_masst("Omeoprazole.csv",          "parent"),
  read_masst("5-hydroxyomeprazole.csv",  "metabolite")
)
```

## Remove corrupted USIs

Some USIs carry the scan number `4294967295` — the maximum value of a
32-bit unsigned integer. This value is written by certain mzML converters when
the scan index is missing or could not be resolved. We remove them as they
cannot give us the position of the spectrum of interest in the raw file.

```{r filter-corrupt-usi}
INVALID_SCAN_STR <- "4294967295"

n_corrupt <- sum(all_hits$scan_str == INVALID_SCAN_STR, na.rm = TRUE)
cat("Corrupted USIs removed:", n_corrupt,
    "(", round(100 * n_corrupt / nrow(all_hits), 1), "% of all hits)\n")

all_hits           <- all_hits[all_hits$scan_str != INVALID_SCAN_STR, ]
all_hits$scan_id   <- as.integer(all_hits$scan_str)   # now safe to convert
all_hits$scan_str  <- NULL                             # drop the staging column
```

## Infer adduct from observed precursor m/z

MASST reports an `Adduct` column, but it's inherited from whichever library
spectrum matched and does not always reflect how *our* compound ionised in
the query data. We know the target formula per role, so for each hit we
compute the theoretical m/z of a set of common adducts and pick the one
closest to the observed `Precursor_MZ` within a ppm tolerance. This
ground-truthed adduct is what we feed to SIRIUS later.

```{r infer-adduct}
library(MetaboCoreUtils)

ADDUCT_CANDIDATES <- c(
  "[M+H]+", "[M+Na]+", "[M+K]+", "[M+NH4]+",
  "[M-H]-", "[M+Cl]-"
)

infer_adduct <- function(obs_mz, formula, ppm = 20,
                         candidates = ADDUCT_CANDIDATES) {
  if (is.na(obs_mz) || is.na(formula)) return(NA_character_)
  mass <- calculateMass(formula)
  theo <- vapply(candidates, function(a) mass2mz(mass, adduct = a)[[1]],
                 numeric(1))
  diff_ppm <- abs(obs_mz - theo) / theo * 1e6
  best <- which.min(diff_ppm)
  if (diff_ppm[best] > ppm) NA_character_ else candidates[best]
}

role_formula <- setNames(compound_spec$formula, compound_spec$role)

all_hits$adduct_inferred <- mapply(
  infer_adduct,
  all_hits$Precursor_MZ,
  role_formula[all_hits$role],
  USE.NAMES = FALSE
)

# Coverage and adduct breakdown per role
table(role   = all_hits$role,
      adduct = all_hits$adduct_inferred,
      useNA  = "ifany")
```

# Dataset selection

## D_both — datasets where both compounds appear

```{r d-both}
D_parent     <- unique(all_hits$ATTRIBUTE_DatasetAccession[all_hits$role == "parent"])
D_metabolite <- unique(all_hits$ATTRIBUTE_DatasetAccession[all_hits$role == "metabolite"])
D_both       <- intersect(D_parent, D_metabolite)

cat("parent:", length(D_parent), "| metabolite:", length(D_metabolite),
    "| both:", length(D_both), "\n")
```

Pre-computed: 53 parent datasets, 30 metabolite datasets, **26 in both**
(26/30 metabolite datasets also contain the parent — high co-occurrence).

## Initial screening

13 of the 26 D_both datasets were excluded at initial inspection before
detailed curation. Two exclusion categories:

* **DIA/SWATH** — acquisition mode incompatible with single-scan MASST matching
* **no_metadata** — no sample-level metadata available in the public deposit

```{r initial-screening}
EXCLUDE_INITIAL <- c(
  MSV000085870 = "DIA/SWATH",
  MSV000091520 = "DIA/SWATH",
  MSV000093072 = "DIA/SWATH",
  MSV000094395 = "DIA/SWATH",
  MSV000094515 = "DIA/SWATH",
  MSV000095726 = "no_metadata",
  MSV000096359 = "no_metadata",
  MSV000096884 = "no_metadata",
  MSV000097110 = "no_metadata",
  MSV000097935 = "no_metadata",
  MTBLS12231   = "no_metadata",
  MTBLS12372   = "no_metadata",
  MTBLS417     = "no_metadata"
)

dataset_overlap <- all_hits[
  all_hits$ATTRIBUTE_DatasetAccession %in% D_both &
  !all_hits$ATTRIBUTE_DatasetAccession %in% names(EXCLUDE_INITIAL), ]

cat("After initial screening:", length(unique(dataset_overlap$ATTRIBUTE_DatasetAccession)),
    "datasets\n")
```

## Manual dataset curation
T
Remaining datasets were reviewed individually. Decisions are recorded in
`dataset_curation_table.csv` (`keep_for_ratio` column). Two further
exclusion categories apply here:

* **In vitro / artefact** — MASST hits arise from compound-spiked experiments,
  not biological patient exposure.
* **Insufficient metadata** — phenotype data is in the publication supplementary
  only and cannot be linked to file names without additional manual work.

```{r manual-curation}
curation <- read.csv("dataset_curation_table.csv", check.names = FALSE)
curation[, c("dataset_id", "keep_for_ratio", "reason")]
```

```{r apply-curation}
keep_ids   <- curation$dataset_id[curation$keep_for_ratio == TRUE]
before_ids <- unique(dataset_overlap$ATTRIBUTE_DatasetAccession)

hits_curated <- dataset_overlap[
  dataset_overlap$ATTRIBUTE_DatasetAccession %in% keep_ids, ]

removed <- setdiff(before_ids, keep_ids)
cat("Datasets kept   :", length(unique(hits_curated$ATTRIBUTE_DatasetAccession)), "\n")
cat("Datasets removed:", length(removed), "\n")
cat(paste(" ", removed, "<-",
          curation$reason[match(removed, curation$dataset_id)],
          collapse = "\n"), "\n")

write.csv(hits_curated, "hits_curated.csv", row.names = FALSE)
```

# MASST hits confirmation

SIRIUS confirmation is run in three separate notebooks, one per repository:

* `masst-hit-confirmation.qmd` → `hits-confirmed-massive.csv`
* `metaboLights-confirmation.qmd` → `hits-confirmed-metaboLights.csv`
* `metabolomics-workbench-confirmation.qmd` → `hits-confirmed-metabolomics-workbench.csv`

## Load and combine

Each file is filtered to its own accession prefix before combining to
avoid cross-contamination (the `resolve_all` helper reads the whole
`results/` folder regardless of which notebook called it).

```{r load-confirmed}
conf_massive <- read.csv("hits-confirmed-massive.csv",    check.names = FALSE)
conf_mtbls   <- read.csv("hits-confirmed-metaboLights.csv", check.names = FALSE)
conf_mwb     <- read.csv("hits-confirmed-metabolomics-workbench.csv", check.names = FALSE)

conf_massive <- conf_massive[startsWith(conf_massive$dataset_id, "MSV"),  ]
conf_mtbls   <- conf_mtbls  [startsWith(conf_mtbls$dataset_id,   "MTBLS"),]
conf_mwb     <- conf_mwb    [startsWith(conf_mwb$dataset_id,     "ST"),   ]

conf_all <- rbind(conf_massive, conf_mtbls, conf_mwb)

# Ensure all expected columns exist (older confirmed CSVs lack mz/adduct_inferred)
for (col in c("mz", "adduct_inferred", "source_file")) {
  if (!col %in% names(conf_all)) conf_all[[col]] <- NA
}

# Patch rt / mz / adduct_inferred / source_file from per-dataset hits_rt.csv.
# source_file is always re-patched (it is never written to the confirmed CSV
# itself, so it is always NA there regardless of whether rt/mz are present).
need_patch_ds <- unique(conf_all$dataset_id[
  is.na(conf_all$rt) | is.na(conf_all$mz) |
  is.na(conf_all$adduct_inferred) | is.na(conf_all$source_file)
])
if (length(need_patch_ds)) {
  cat("Patching rt/mz/adduct from hits_rt.csv for",
      length(need_patch_ds), "dataset(s):\n")
  for (ds in need_patch_ds) {
    rt_path <- file.path("results", ds, "hits_rt.csv")
    if (!file.exists(rt_path)) {
      cat("  [", ds, "] no hits_rt.csv — skipped\n"); next
    }
    rt_map <- read.csv(rt_path, check.names = FALSE)
    idx    <- conf_all$dataset_id == ds
    keys_l <- paste(conf_all$role[idx], conf_all$xcms_fts[idx])
    keys_r <- paste(rt_map$role,         rt_map$xcms_fts)
    m      <- match(keys_l, keys_r)
    if (is.null(conf_all$rt[idx]) || any(is.na(conf_all$rt[idx])))
      conf_all$rt[idx] <- ifelse(is.na(conf_all$rt[idx]),
                                  rt_map$rt[m], conf_all$rt[idx])
    if ("mz" %in% names(rt_map))
      conf_all$mz[idx] <- ifelse(is.na(conf_all$mz[idx]),
                                  rt_map$mz[m], conf_all$mz[idx])
    if ("adduct_inferred" %in% names(rt_map))
      conf_all$adduct_inferred[idx] <- ifelse(
        is.na(conf_all$adduct_inferred[idx]),
        rt_map$adduct_inferred[m], conf_all$adduct_inferred[idx])
    if ("file" %in% names(rt_map))
      conf_all$source_file[idx] <- ifelse(
        is.na(conf_all$source_file[idx]),
        rt_map$file[m], conf_all$source_file[idx])
    cat("  [", ds, "] patched (non-NA mz:",
        sum(!is.na(conf_all$mz[idx])), "/", sum(idx),
        ", non-NA adduct_inferred:",
        sum(!is.na(conf_all$adduct_inferred[idx])), "/", sum(idx), ")\n")
  }
}

conf_all$confidence <- factor(conf_all$confidence,
                               levels = c("strict", "standard", "liberal",
                                          "not_confirmed"))

# Join chromatography type from curation table
# (curation was loaded in the manual-curation chunk above)
if (!exists("curation"))
  curation <- read.csv("dataset_curation_table.csv", check.names = FALSE)
curation$chrom_type <- dplyr::case_when(
  grepl("HILIC|ZIC|pHILIC", curation$chromatography, ignore.case = TRUE) ~ "HILIC",
  grepl("RP|C18|reverse", curation$chromatography, ignore.case = TRUE)   ~ "RP",
  TRUE                                                                     ~ "Unknown"
)
conf_all <- dplyr::left_join(
  conf_all,
  curation[, c("dataset_id", "chrom_type")],
  by = "dataset_id"
)

cat("Total confirmed features:", nrow(conf_all), "\n")
cat("Datasets             :", length(unique(conf_all$dataset_id)), "\n")
cat("Chromatography types :\n")
print(table(unique(conf_all[, c("dataset_id", "chrom_type")])$chrom_type))
```

## Resolve assay per confirmed feature

Multi-assay studies (mostly MetaboLights, sometimes Metabolomics Workbench)
host several assays per dataset — different polarities, columns, or
matrices. Confirmed features from different assays must NEVER be mixed:
they map to different chromatography and ionisation chemistry, and
therefore to different RT and m/z windows. We resolve the assay for each
confirmed feature from its `source_file` (the MS file where SIRIUS matched
the compound) and store the result in a new `assay` column. Every
downstream filter, RT/mz summary, and EIC extraction then groups by
(`dataset_id`, `assay`) so multi-assay studies stay separated.

```{r resolve-assay}
META_DIR <- "dataset_metadata"

# --- MetaboLights ---------------------------------------------------------
# Build basename(file) → assay_file_name lookup by scanning every
# a_{ds}_*.txt assay table cached in dataset_metadata/.
mtbls_assay_lookup <- function(ds) {
  fs <- list.files(META_DIR, pattern = paste0("^a_", ds, "_.*\\.txt$"),
                   full.names = TRUE)
  if (!length(fs)) return(character(0))
  out <- character(0)
  for (fp in fs) {
    adf <- tryCatch(read.delim(fp, check.names = FALSE),
                    error = function(e) NULL)
    if (is.null(adf)) next
    file_cols <- grep("Spectral Data File", names(adf),
                      value = TRUE, ignore.case = TRUE)
    files <- unlist(lapply(file_cols, function(c) adf[[c]]), use.names = FALSE)
    files <- basename(trimws(files))
    files <- files[!is.na(files) & nchar(files) > 0]
    if (!length(files)) next
    out <- c(out, setNames(rep(basename(fp), length(files)), files))
  }
  out
}

# --- Metabolomics Workbench -----------------------------------------------
# Build per-feature AN assignment. RAW_FILE_NAME is shared across AN files
# (SUBJECT_SAMPLE_FACTORS is identical), so we disambiguate by ION_MODE
# matched against the confirmed adduct polarity, then narrow by filename
# polarity tokens when several AN files match the same mode.
mwb_assay_for_rows <- function(ds, conf_rows) {
  fs <- list.files(META_DIR, pattern = paste0("^", ds, "_AN.*\\.txt$"),
                   full.names = TRUE)
  if (!length(fs)) return(rep(NA_character_, nrow(conf_rows)))
  an_pol <- vapply(fs, function(fp) {
    ln <- readLines(fp, warn = FALSE)
    trimws(sub("^MS:ION_MODE\\s+", "",
               grep("^MS:ION_MODE", ln, value = TRUE)[1]))
  }, character(1))
  out <- rep(NA_character_, nrow(conf_rows))
  for (i in seq_len(nrow(conf_rows))) {
    adu <- conf_rows$adduct_inferred[i]
    if (is.na(adu)) next
    pol <- if (endsWith(trimws(adu), "+")) "positive" else "negative"
    candidates <- fs[grepl(pol, an_pol, ignore.case = TRUE)]
    if (!length(candidates)) next
    sf <- basename(conf_rows$source_file[i])
    if (length(candidates) > 1 && !is.na(sf) && nchar(sf)) {
      hit <- vapply(candidates, function(fp) {
        any(grepl(sf, readLines(fp, warn = FALSE), fixed = TRUE))
      }, logical(1))
      if (any(hit)) candidates <- candidates[hit]
    }
    out[i] <- basename(candidates[1])
  }
  out
}

conf_all$assay <- NA_character_

# MassIVE: a single metadata file per dataset → assay collapsed to dataset_id
msv_idx <- startsWith(conf_all$dataset_id, "MSV")
conf_all$assay[msv_idx] <- conf_all$dataset_id[msv_idx]

# MetaboLights: lookup per source_file basename
for (ds in unique(conf_all$dataset_id[startsWith(conf_all$dataset_id, "MTBLS")])) {
  lookup <- mtbls_assay_lookup(ds)
  idx    <- conf_all$dataset_id == ds
  src    <- basename(conf_all$source_file[idx])
  conf_all$assay[idx] <- unname(lookup[src])
}

# Metabolomics Workbench: polarity-based AN selection
for (ds in unique(conf_all$dataset_id[startsWith(conf_all$dataset_id, "ST")])) {
  idx <- conf_all$dataset_id == ds
  conf_all$assay[idx] <- mwb_assay_for_rows(ds, conf_all[idx, , drop = FALSE])
}

cat("Assay resolved for", sum(!is.na(conf_all$assay)), "/", nrow(conf_all),
    "confirmed features\n")
print(table(dataset_id = conf_all$dataset_id, assay = conf_all$assay,
            useNA = "ifany"))

# Datasets where more than one assay carries confirmed features
ma_ds <- vapply(split(conf_all$assay, conf_all$dataset_id),
                function(x) length(unique(na.omit(x))) > 1, logical(1))
multi_assay_ds <- names(ma_ds)[ma_ds]
if (length(multi_assay_ds))
  cat("Multi-assay datasets (kept separate downstream):\n  ",
      paste(multi_assay_ds, collapse = ", "), "\n")
```

## Adduct consistency rule

For the parent/metabolite ratio to be interpretable, both compounds must
be extracted using the **same adduct** (same ionisation chemistry, same
nominal m/z offset). A dataset where omeprazole is confirmed as [M+H]+
but 5-OH-omeprazole only as [M+Na]+ is not a problem per se — both
adducts are real — but the EIC windows will target different m/z values
and mixing them in the ratio is confusing.

**Decision rule:** we restrict all diagnostic plots and downstream EIC
windows to adducts that are observed in **both** parent and metabolite
confirmed features. Adducts that appear in only one role are dropped.
This does not discard entire datasets; it only drops individual features
whose adduct was not replicated on the other side of the ratio.

```{r shared-adducts}
adducts_parent <- unique(conf_all$adduct_inferred[
  conf_all$role == "parent" & !is.na(conf_all$adduct_inferred)])
adducts_metab  <- unique(conf_all$adduct_inferred[
  conf_all$role == "metabolite" & !is.na(conf_all$adduct_inferred)])
shared_adducts <- intersect(adducts_parent, adducts_metab)

cat("Adducts in parent    :", paste(adducts_parent, collapse = ", "), "\n")
cat("Adducts in metabolite:", paste(adducts_metab,  collapse = ", "), "\n")
cat("Shared adducts kept  :", paste(shared_adducts, collapse = ", "), "\n")

conf_all_shared <- conf_all[
  !is.na(conf_all$adduct_inferred) &
  conf_all$adduct_inferred %in% shared_adducts, ]
cat("Features retained:", nrow(conf_all_shared), "of", nrow(conf_all), "\n")
```

## RT across datasets

Retention times vary widely between datasets because each study used a
different LC gradient. This is expected and will be resolved by
per-dataset RT windows in the EIC step. What matters here is
**within-dataset consistency**: features of the same compound in the
same dataset should cluster tightly.

Note: only adducts present in both parent and metabolite are displayed
(see adduct consistency rule above).

```{r plot-rt-discrepancy, fig.width = 11, fig.height = 5}
CONFIDENCE_COLOURS <- c(strict = "#1A7837", standard = "#74C476",
                         liberal = "#FEC44F", not_confirmed = "#D7191C")

conf_all_shared |>
  filter(!is.na(rt)) |>
  ggplot(aes(x = dataset_id, y = rt,
             colour = confidence, shape = adduct_inferred)) +
  geom_jitter(size = 2.5, width = 0.15, alpha = 0.85) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5,
               linewidth = 0.4, colour = "grey20") +
  facet_grid(chrom_type ~ role, scales = "free", space = "free_x") +
  scale_colour_manual(values = CONFIDENCE_COLOURS, drop = FALSE) +
  scale_shape_discrete(drop = TRUE) +
  labs(x = NULL, y = "Retention time (s)",
       colour = "Confidence", shape = "Adduct",
       title = "RT of confirmed features — cross-dataset comparison") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom",
        strip.background = element_rect(fill = "grey92"))
```

## m/z deviation across datasets

The observed precursor m/z (from the actual MS2 scan in the raw file)
should match the theoretical m/z of the inferred adduct within the
instrument's mass accuracy. We compute deviation in ppm:
`(observed − theoretical) / theoretical × 10⁶`.

A well-calibrated Orbitrap is typically within ±5 ppm; anything above
±10 ppm suggests either a calibration issue in that dataset or a
mis-assigned adduct. The ±5 ppm dashed lines serve as a visual guide.
This plot is also used to sanity-check whether our adduct inference
(from `MetaboCoreUtils::infer_adduct`) is consistent with what SIRIUS
independently resolved in the `adduct` column of the confirmed CSV.

```{r mz-deviation, fig.width = 11, fig.height = 5}
library(MetaboCoreUtils)

role_formula <- setNames(compound_spec$formula, compound_spec$role)

conf_all_shared$mz_theo <- mapply(
  function(f, a) {
    if (is.na(f) || is.na(a)) return(NA_real_)
    mass2mz(calculateMass(f), adduct = a)[[1]]
  },
  role_formula[conf_all_shared$role],
  conf_all_shared$adduct_inferred,
  USE.NAMES = FALSE
)
conf_all_shared$mz_ppm_dev <- (conf_all_shared$mz - conf_all_shared$mz_theo) /
                               conf_all_shared$mz_theo * 1e6

# Also propagate to conf_all for use in proto-config
conf_all$mz_theo    <- conf_all_shared$mz_theo[
  match(paste(conf_all$dataset_id, conf_all$role, conf_all$xcms_fts),
        paste(conf_all_shared$dataset_id, conf_all_shared$role,
              conf_all_shared$xcms_fts))]
conf_all$mz_ppm_dev <- conf_all_shared$mz_ppm_dev[
  match(paste(conf_all$dataset_id, conf_all$role, conf_all$xcms_fts),
        paste(conf_all_shared$dataset_id, conf_all_shared$role,
              conf_all_shared$xcms_fts))]

cat("m/z deviation summary (ppm):\n")
print(summary(conf_all_shared$mz_ppm_dev))

conf_all_shared |>
  filter(!is.na(mz_ppm_dev)) |>
  ggplot(aes(x = dataset_id, y = mz_ppm_dev,
             colour = adduct_inferred, shape = adduct_inferred)) +
  geom_hline(yintercept = 0,        linetype = 2, colour = "grey40") +
  geom_hline(yintercept = c(-5, 5), linetype = 3, colour = "grey70") +
  geom_jitter(size = 2.5, width = 0.15, alpha = 0.85) +
  facet_grid(chrom_type ~ role, scales = "free_x", space = "free_x") +
  scale_shape_discrete(drop = TRUE) +
  labs(x = NULL, y = "m/z deviation (ppm)",
       colour = "Adduct", shape = "Adduct",
       title = "Precursor m/z deviation vs theoretical adduct mass",
       caption = "Dashed: 0 ppm  ·  Dotted: ±5 ppm") +
  theme_bw() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        legend.position = "bottom",
        strip.background = element_rect(fill = "grey92"))
```

## Adduct distribution per role

```{r adduct-distribution}
table(role   = conf_all_shared$role,
      adduct = conf_all_shared$adduct_inferred,
      useNA  = "ifany")
```

## Filter: keep highest confidence level per dataset

SIRIUS can return multiple candidate structures at different confidence
tiers for the same feature. We keep only the highest tier reached per
(dataset, role) pair. For a dataset that has one strict and two standard
features for the parent, we keep only the strict one — it provides the
most reliable RT anchor and m/z for EIC extraction.

**Why this matters for the ratio:** if we kept both strict and standard
features for the same compound in a dataset, we would be double-counting
that dataset in downstream statistics. One confirmed m/z window per
compound per dataset is what we need.

```{r filter-best-confidence}
CONF_ORDER <- c(strict = 1, standard = 2, liberal = 3, not_confirmed = 4)

conf_best <- conf_all_shared |>
  group_by(dataset_id, assay, role) |>
  filter(CONF_ORDER[as.character(confidence)] ==
           min(CONF_ORDER[as.character(confidence)])) |>
  ungroup()

cat("Features after confidence filter:", nrow(conf_best), "\n")
cat("Features removed               :", nrow(conf_all_shared) - nrow(conf_best), "\n")
```

## Filter: both parent and metabolite must be confirmed

The CYP2C19 ratio requires **both** omeprazole (parent) and
5-OH-omeprazole (metabolite) to be detectable in the same dataset.
A dataset with only the parent confirmed cannot contribute a ratio — it
can tell us that omeprazole is present (consistent with PPI use) but
not how efficiently CYP2C19 is metabolising it.

Datasets confirmed for only one role are dropped here. The diagnostic
table below identifies the reason for each dropout by checking the full
SIRIUS output (including liberal and not_confirmed tiers), so we can
distinguish between "SIRIUS ran and found nothing convincing" vs "SIRIUS
never ran for this role" (e.g., no MASST hits for metabolite in that
dataset to begin with).

```{r diagnose-missing-roles}
# Read ALL SIRIUS results (not filtered to strict/standard) to see full picture
read_all_results <- function(results_dir = "results") {
  CONFIDENCE_LEVELS <- c("strict", "standard", "liberal", "not_confirmed")
  CONF_ORDER        <- c(strict = 1, standard = 2, liberal = 3, not_confirmed = 4)

  datasets <- list.dirs(results_dir, recursive = FALSE, full.names = FALSE)
  rows <- lapply(datasets, function(ds) {
    lapply(c("parent", "metabolite"), function(role) {
      p <- file.path(results_dir, ds, paste0(role, "_results.csv"))
      if (!file.exists(p)) return(NULL)
      res <- read.csv(p, check.names = FALSE)
      if (!nrow(res)) return(NULL)
      # just need the best confidence reached per dataset × role
      data.frame(dataset_id = ds, role = role,
                 n_features  = length(unique(res$xcms_fts)),
                 stringsAsFactors = FALSE)
    })
  })
  do.call(rbind, unlist(rows, recursive = FALSE))
}

all_results <- read_all_results()

# What is confirmed (strict/standard) vs what was run
role_status <- all_results |>
  left_join(
    conf_best |>
      group_by(dataset_id, role) |>
      summarise(best_conf = as.character(confidence[1]), .groups = "drop"),
    by = c("dataset_id", "role")
  ) |>
  mutate(
    status = case_when(
      best_conf %in% c("strict", "standard") ~ best_conf,
      !is.na(n_features)                     ~ "ran — below threshold",
      TRUE                                   ~ "no results"
    )
  ) |>
  tidyr::pivot_wider(names_from = role,
                     values_from = c(n_features, status, best_conf))

role_status |>
  filter(startsWith(dataset_id, "MTBLS") | startsWith(dataset_id, "ST")) |>
  dplyr::arrange(dataset_id)
```

```{r filter-both-roles}
# Per (dataset, assay): an assay must carry both parent and metabolite
# for its ratio to be computable. Multi-assay datasets can therefore
# survive even if only one of their assays passes both-roles.
both_roles <- conf_best |>
  group_by(dataset_id, assay) |>
  filter(all(c("parent", "metabolite") %in% role)) |>
  ungroup()

ds_keys_before <- unique(paste(conf_best$dataset_id, conf_best$assay, sep = "::"))
ds_keys_after  <- unique(paste(both_roles$dataset_id, both_roles$assay, sep = "::"))
dropped_keys   <- setdiff(ds_keys_before, ds_keys_after)
cat("Dataset×assay units with both roles confirmed:",
    length(ds_keys_after), "\n")
cat("Datasets with both roles confirmed (>=1 assay):",
    length(unique(both_roles$dataset_id)), "\n")
if (length(dropped_keys)) {
  cat("Dropped dataset×assay (one role missing strict/standard):\n",
      paste("  ", dropped_keys, collapse = "\n"), "\n")
}

conf_final <- both_roles
```

## RT window summary

```{r rt-window-summary}
rt_windows <- conf_final |>
  filter(!is.na(rt)) |>
  group_by(dataset_id, assay, role) |>
  summarise(
    confidence   = as.character(unique(confidence)),
    n_features   = n(),
    rt_median_s  = round(median(rt), 1),
    rt_min_s     = round(min(rt), 1),
    rt_max_s     = round(max(rt), 1),
    rt_range_s   = round(max(rt) - min(rt), 1),
    .groups      = "drop"
  ) |>
  tidyr::pivot_wider(
    names_from  = role,
    values_from = c(confidence, n_features, rt_median_s,
                    rt_min_s, rt_max_s, rt_range_s)
  )

rt_windows
```

```{r plot-rt-windows, fig.width = 9, fig.height = 5}
# Per (dataset, assay) so multi-assay studies render as separate rows;
# the y-axis label collapses to dataset_id when there is only one assay.
conf_final |>
  filter(!is.na(rt)) |>
  group_by(dataset_id, assay, role, chrom_type) |>
  summarise(rt_lo = min(rt), rt_hi = max(rt),
            rt_med = median(rt), .groups = "drop") |>
  group_by(dataset_id) |>
  mutate(ds_label = if (length(unique(assay)) > 1)
                      paste(dataset_id, assay, sep = " · ")
                    else dataset_id) |>
  ungroup() |>
  ggplot(aes(y = ds_label, colour = role)) +
  geom_linerange(aes(xmin = rt_lo, xmax = rt_hi),
                 linewidth = 2, alpha = 0.4,
                 position = position_dodge(width = 0.5)) +
  geom_point(aes(x = rt_med),
             size = 3, position = position_dodge(width = 0.5)) +
  facet_grid(chrom_type ~ ., scales = "free_y", space = "free_y") +
  scale_colour_manual(values = c(parent = "#2166AC", metabolite = "#D6604D")) +
  labs(x = "Retention time (s)", y = NULL, colour = "Role",
       title = "Per-dataset RT windows (confirmed features)") +
  theme_bw() +
  theme(legend.position = "bottom",
        strip.background = element_rect(fill = "grey92"))
```

## Dataset summary

```{r dataset-summary-table}
ds_summary <- conf_final |>
  group_by(dataset_id, assay) |>
  summarise(
    repository      = case_when(
      startsWith(dataset_id[1], "MSV")   ~ "MassIVE",
      startsWith(dataset_id[1], "MTBLS") ~ "MetaboLights",
      startsWith(dataset_id[1], "ST")    ~ "Metabolomics Workbench"
    ),
    roles_confirmed = paste(sort(unique(role)), collapse = " + "),
    best_confidence = levels(conf_final$confidence)[
      min(CONF_ORDER[as.character(confidence)])],
    n_parent        = sum(role == "parent"),
    n_metabolite    = sum(role == "metabolite"),
    rt_parent_s     = round(median(rt[role == "parent"],  na.rm = TRUE), 1),
    rt_metabolite_s = round(median(rt[role == "metabolite"], na.rm = TRUE), 1),
    has_rt          = any(!is.na(rt)),
    .groups         = "drop"
  ) |>
  left_join(
    curation[, c("dataset_id", "chrom_type", "matrix", "design", "n_subjects")],
    by = "dataset_id"
  ) |>
  arrange(repository, dataset_id, assay)

ds_summary
```

```{r save-conf-final}
write.csv(conf_final,  "hits-confirmed-all.csv",  row.names = FALSE)
write.csv(ds_summary,  "datasets-confirmed.csv",  row.names = FALSE)
```

# Biological file selection

Before any download, we derive the list of biological sample files per
confirmed dataset from local metadata alone — excluding QC injections,
blanks, pooled QC, reference standards, and instrument controls.

## Metadata format differences across repositories

The three repositories store sample-level metadata in very different
structures, and **the location of the spectral file name differs in each**:

**MassIVE (GNPS / Pan-ReDU)**
: Each dataset has a metadata TSV downloaded from Pan-ReDU (or a
custom CSV from the data submitter). Both contain a `SampleType` column
(values: `"Sample"`, `"QC"`, `"Blank"`, etc.) and a `filename` column.
*TSV (Pan-ReDU):* `filename` holds the full relative path within the
MassIVE file tree, e.g.
`peak/05092023-pos-pc18-peaklist/pos_NA_8990_NA_NA_193_G_B5.mzXML`.
*CSV (submitter-provided):* `filename` holds only the bare sample stem
without extension or subdirectory, e.g. `pos_inf_non_10124_male_67_267_R_C8`.
For the CSV case, we reconstruct the full path by extracting the
directory prefix and extension pattern from the MASST hit `file_path`
entries already present in `hits_curated` for that dataset.

**MetaboLights (EBI)**
: MetaboLights uses the ISA-Tab format. The study sample file
(`s_{id}.txt`) lists sample names and phenotype factors
(`Characteristics[Sample type]`) but contains **no spectral file names**.
File names live in the assay file(s) (`a_{id}_*.txt`), which map
`Sample Name` to `Raw Spectral Data File` or `Derived Spectral Data File`.
A study may have **multiple assay files** — typically one per
chromatographic mode (positive, negative) or matrix.  We use the
`MsBackendMetaboLights` helper `mtbls_assay_data()` (with an FTP fallback)
to load each assay table, then match the confirmed MASST source files
to determine which assay(s) are relevant.  When confirmed files span
more than one assay, `bio_files_per_ds[[ds]]` becomes a **named list**
(keyed by assay file name) rather than a flat character vector.

**Metabolomics Workbench**
: Data are stored in mwtab plain-text files, one per analysis
(e.g., `ST002044_AN003325.txt`). Each contains `SUBJECT_SAMPLE_FACTORS`
lines with tab-delimited fields: tag, subject ID, sample ID, factor
string, and an optional extra field that may include `RAW_FILE_NAME=...`.
We parse these lines across all analysis files for the study and again
use confirmed source files to restrict to only the analysis/analyses
that contain real data for the compounds of interest.

```{r bio-file-selection}
library(MsBackendMetaboLights)   # mtbls_assay_data(), used in fetch_assay_table()
META_DIR <- "dataset_metadata"
EXCL_ST  <- "(?i)^(blank|qc|pool|reference|standard|control|empty)"

# ── MetaboLights helpers ──────────────────────────────────────────────────────

# Get list of assay file names from the ISA investigation file.
# Caches the investigation file locally; returns character vector of
# assay file names like "a_MTBLS1866_LC-MS_positive.txt".
mtbls_inv_assay_names <- function(ds) {
  inv_path <- file.path(META_DIR, paste0(ds, "_i_Investigation.txt"))
  if (!file.exists(inv_path)) {
    inv_url <- paste0("https://ftp.ebi.ac.uk/pub/databases/metabolights/",
                      "studies/public/", ds, "/i_Investigation.txt")
    inv_lines <- tryCatch(
      readLines(url(inv_url), warn = FALSE),
      error = function(e) { message("FTP unreachable: ", e$message); NULL }
    )
    if (is.null(inv_lines)) return(NULL)
    writeLines(inv_lines, inv_path)
  } else {
    inv_lines <- readLines(inv_path, warn = FALSE)
  }
  assay_line  <- inv_lines[startsWith(inv_lines, "Study Assay File Name")]
  if (!length(assay_line)) return(NULL)
  an <- trimws(unlist(strsplit(sub("^[^\t]+\t", "", assay_line[1]), "\t")))
  an[nchar(an) > 0]
}

# Fetch (or load cached) assay table for one assay of a MetaboLights study.
# Tries mtbls_assay_data() first (MsBackendMetaboLights); falls back to FTP.
fetch_assay_table <- function(ds, assay_name) {
  cache_path <- file.path(META_DIR, assay_name)
  if (file.exists(cache_path)) return(read.delim(cache_path, check.names = FALSE))
  adf <- tryCatch(
    mtbls_assay_data(ds, assay_name),
    error = function(e) NULL
  )
  if (!is.null(adf)) {
    write.table(adf, cache_path, sep = "\t", row.names = FALSE, quote = FALSE)
    message("  Assay file cached (mtbls_assay_data) → ", cache_path)
    return(adf)
  }
  assay_url <- paste0("https://ftp.ebi.ac.uk/pub/databases/metabolights/",
                      "studies/public/", ds, "/", assay_name)
  adf <- tryCatch(
    read.delim(url(assay_url), check.names = FALSE),
    error = function(e) { message("FTP fetch failed for ", assay_name, ": ", e$message); NULL }
  )
  if (!is.null(adf)) {
    write.table(adf, cache_path, sep = "\t", row.names = FALSE, quote = FALSE)
    message("  Assay file cached (FTP) → ", cache_path)
  }
  adf
}

# ── Main loop ─────────────────────────────────────────────────────────────────
bio_files_per_ds  <- list()   # ds → file vector, or named list per assay
bio_assays_per_ds <- list()   # ds → assay/analysis identifiers used

for (ds in unique(conf_final$dataset_id)) {
  cat("[", ds, "]\n")

  if (startsWith(ds, "MSV")) {
    # ── MassIVE ──────────────────────────────────────────────────────────────
    f <- list.files(META_DIR,
                    pattern    = paste0("^", ds, "_metadata.*\\.(tsv|csv)$"),
                    full.names = TRUE)
    if (!length(f)) { cat("  no metadata file\n"); next }

    is_tsv <- grepl("\\.tsv$", f[1])
    meta   <- if (is_tsv) read.delim(f[1], check.names = FALSE)
              else        read.csv(f[1],   check.names = FALSE)

    if ("SampleType" %in% names(meta))
      meta <- meta[!grepl(EXCL_ST, meta$SampleType, perl = TRUE), ]

    if (!is_tsv) {
      # Custom CSV: filename = bare sample stem (no extension, no subdirectory).
      # Reconstruct full relative paths using the directory prefix + extension
      # pattern observed in the MASST hit file_path entries for this dataset.
      masst_paths <- unique(hits_curated$file_path[
        hits_curated$ATTRIBUTE_DatasetAccession == ds])
      dir_prefix  <- names(which.max(table(dirname(masst_paths))))
      exts        <- tools::file_ext(masst_paths)
      ext_str     <- names(which.max(table(exts[nchar(exts) > 0])))
      ext_str     <- if (nchar(ext_str) > 0) paste0(".", ext_str) else ""
      cat("  CSV metadata — reconstructing paths: dir =", dir_prefix,
          "| ext =", ext_str, "\n")
      bio_files_per_ds[[ds]] <- file.path(dir_prefix,
                                          paste0(unique(meta$filename), ext_str))
    } else {
      bio_files_per_ds[[ds]] <- unique(meta$filename)
    }
    bio_assays_per_ds[[ds]] <- basename(f[1])

  } else if (startsWith(ds, "MTBLS")) {
    # ── MetaboLights ─────────────────────────────────────────────────────────
    s_path <- file.path(META_DIR, paste0("s_", ds, ".txt"))
    if (!file.exists(s_path)) { cat("  no sample file\n"); next }

    samples <- read.delim(s_path, check.names = FALSE)
    st_col  <- grep("sample type", names(samples), value = TRUE, ignore.case = TRUE)[1]
    if (!is.na(st_col))
      samples <- samples[!grepl(EXCL_ST, samples[[st_col]], perl = TRUE), ]
    bio_names <- unique(samples[["Sample Name"]])

    # Get all assay names for this study from the investigation file
    assay_names <- mtbls_inv_assay_names(ds)
    if (is.null(assay_names) || !length(assay_names)) {
      cat("  could not retrieve assay list\n"); next
    }
    cat("  assay files found:", paste(assay_names, collapse = "; "), "\n")

    # Load all assay tables and extract bio file lists.
    # fcol_pick prefers "Derived Spectral Data File" over "Raw Spectral Data File"
    # because in MetaboLights the Raw column is often all-NA while Derived has paths.
    # Also filter NA explicitly — nchar(NA) returns NA not 0, so nchar(fv)>0 alone
    # would pass NA through and cause downstream unlist() to produce NA elements.
    fcol_pick <- function(adf) {
      derived <- grep("Derived Spectral Data File", names(adf),
                      value = TRUE, ignore.case = TRUE)[1]
      if (!is.na(derived) && any(!is.na(adf[[derived]]))) return(derived)
      grep("Raw Spectral Data File", names(adf),
           value = TRUE, ignore.case = TRUE)[1]
    }
    assay_bio_files <- lapply(setNames(assay_names, assay_names), function(an) {
      adf <- fetch_assay_table(ds, an)
      if (is.null(adf)) return(character(0))
      fc <- fcol_pick(adf)
      if (is.na(fc)) return(character(0))
      rows <- adf[adf[["Sample Name"]] %in% bio_names, , drop = FALSE]
      fv   <- unique(trimws(rows[[fc]]))
      fv[!is.na(fv) & nchar(fv) > 0]
    })
    assay_bio_files <- Filter(function(x) length(x) > 0, assay_bio_files)

    # Select assays that contain confirmed source files for BOTH roles.
    # Requiring both roles in the same assay guarantees the ratio is computable.
    conf_ds_mtbls <- conf_final[conf_final$dataset_id == ds & !is.na(conf_final$source_file), ]
    src_par_mtbls <- unique(conf_ds_mtbls$source_file[conf_ds_mtbls$role == "parent"])
    src_met_mtbls <- unique(conf_ds_mtbls$source_file[conf_ds_mtbls$role == "metabolite"])

    if (length(src_par_mtbls) > 0 && length(src_met_mtbls) > 0) {
      in_par <- sapply(assay_bio_files, function(fv) any(basename(fv) %in% src_par_mtbls))
      in_met <- sapply(assay_bio_files, function(fv) any(basename(fv) %in% src_met_mtbls))
      relevant_assays <- names(assay_bio_files)[in_par & in_met]
      if (!length(relevant_assays)) {
        par_ans <- names(assay_bio_files)[in_par]
        met_ans <- names(assay_bio_files)[in_met]
        cat("  WARNING: parent confirmed in [", paste(par_ans, collapse=", "),
            "] but metabolite in [", paste(met_ans, collapse=", "),
            "] — role/assay mismatch; using all assays\n")
        relevant_assays <- names(assay_bio_files)
      }
    } else if (length(c(src_par_mtbls, src_met_mtbls)) > 0) {
      # Only one role has source_file info — best effort
      conf_src_any    <- c(src_par_mtbls, src_met_mtbls)
      in_any          <- sapply(assay_bio_files, function(fv) any(basename(fv) %in% conf_src_any))
      relevant_assays <- names(assay_bio_files)[in_any]
      cat("  NOTE: only", if (length(src_par_mtbls)) "parent" else "metabolite",
          "has confirmed source_file — assay selection partial\n")
      if (!length(relevant_assays)) relevant_assays <- names(assay_bio_files)
    } else {
      relevant_assays <- names(assay_bio_files)
      cat("  NOTE: no confirmed source_file available — using all assays\n")
    }

    bio_assays_per_ds[[ds]] <- relevant_assays

    if (length(relevant_assays) == 1) {
      bio_files_per_ds[[ds]] <- assay_bio_files[[relevant_assays]]
    } else {
      # Multi-assay: store as named list; handled downstream per assay
      bio_files_per_ds[[ds]] <- assay_bio_files[relevant_assays]
      cat("  MULTI-ASSAY:", length(relevant_assays),
          "assays contain confirmed files — bio_files_per_ds[[ds]] is a named list\n")
    }

  } else if (startsWith(ds, "ST")) {
    # ── Metabolomics Workbench — mwtab format ─────────────────────────────────
    # Each AN file has SUBJECT_SAMPLE_FACTORS lines whose tag is padded with
    # spaces before the first tab — NOT a bare tag+tab. The regex fix is to
    # match on startsWith("SUBJECT_SAMPLE_FACTORS") only (no \t requirement).
    #
    # Additionally, RAW_FILE_NAME in multi-analysis studies lists ALL raw files
    # for a sample (space-separated, one per analysis). We expand these into
    # individual rows, then use confirmed adduct polarity to restrict to the
    # correct analysis.
    mwf <- list.files(META_DIR,
                      pattern    = paste0("^", ds, "_AN.*\\.txt$"),
                      full.names = TRUE)
    if (!length(mwf)) { cat("  no mwtab file\n"); next }

    # Read ION_MODE and CHROMATOGRAPHY_TYPE per analysis file
    an_ids  <- sub(".*_(AN[0-9]+)\\.txt$", "\\1", basename(mwf))
    an_meta <- setNames(lapply(mwf, function(path) {
      ln <- readLines(path, warn = FALSE)
      c(ion_mode   = trimws(sub("^MS:ION_MODE\\s+", "",
                    grep("^MS:ION_MODE",            ln, value = TRUE)[1])),
        chrom_type = trimws(sub("^CH:CHROMATOGRAPHY_TYPE\\s+", "",
                    grep("^CH:CHROMATOGRAPHY_TYPE", ln, value = TRUE)[1])))
    }), an_ids)
    cat("  analyses:", paste(mapply(function(id, m)
          paste0(id, "(", m["chrom_type"], "/", m["ion_mode"], ")"),
          names(an_meta), an_meta), collapse = "; "), "\n")

    # Parse SSF from the first file (lines are identical across all AN files).
    # Match by prefix only — the tag is space-padded before the tab.
    ln  <- readLines(mwf[1], warn = FALSE)
    ssf <- ln[startsWith(ln, "SUBJECT_SAMPLE_FACTORS") & !startsWith(ln, "#")]
    if (!length(ssf)) { cat("  no SUBJECT_SAMPLE_FACTORS\n"); next }

    # Fields after splitting on tab:
    #   p[1] = "SUBJECT_SAMPLE_FACTORS   " (with trailing spaces — trimmed by strsplit)
    #   p[2] = subject (often "-")
    #   p[3] = sample_id
    #   p[4] = factors
    #   p[5] = extra string (may contain "RAW_FILE_NAME=a.mzML b.mzML c.mzML ...")
    # RAW_FILE_NAME may list multiple space-separated files for multi-analysis studies.
    ssf_rows <- lapply(strsplit(ssf, "\t"), function(p) {
      raw_str <- if (length(p) >= 5)
                   regmatches(p[5], regexpr("RAW_FILE_NAME=[^\t]+", p[5]))
                 else character(0)
      raw_str   <- sub("RAW_FILE_NAME=", "", if (length(raw_str)) raw_str else "")
      raw_files <- trimws(unlist(strsplit(raw_str, "\\s+")))
      raw_files <- raw_files[nchar(raw_files) > 0]
      if (!length(raw_files)) raw_files <- NA_character_
      data.frame(sample_id = p[3], factors = p[4],
                 raw_file  = raw_files,
                 stringsAsFactors = FALSE)
    })
    ssf_df <- do.call(rbind, ssf_rows)
    if (is.null(ssf_df) || nrow(ssf_df) == 0L) { cat("  no SUBJECT_SAMPLE_FACTORS\n"); next }
    ssf_df <- ssf_df[!grepl(EXCL_ST, ssf_df$factors, perl = TRUE), ]

    all_bio_raw <- unique(ssf_df$raw_file[!is.na(ssf_df$raw_file)])

    # Restrict to the confirmed analysis using:
    # (a) confirmed adduct polarity → target ION_MODE
    # (b) direct filename match against confirmed source files
    conf_ds_st  <- conf_final[conf_final$dataset_id == ds, ]
    src_par_st  <- unique(conf_ds_st$source_file[conf_ds_st$role == "parent"   & !is.na(conf_ds_st$source_file)])
    src_met_st  <- unique(conf_ds_st$source_file[conf_ds_st$role == "metabolite" & !is.na(conf_ds_st$source_file)])
    conf_src_st <- c(src_par_st, src_met_st)

    if (length(conf_src_st) > 0) {
      # Step 1: determine target ion mode from confirmed adducts
      conf_adducts_st <- unique(conf_ds_st$adduct_inferred[!is.na(conf_ds_st$adduct_inferred)])
      pos_mode_st     <- any(endsWith(trimws(conf_adducts_st), "+"))
      target_mode     <- if (pos_mode_st) "positive" else "negative"

      # Step 2: find which AN files match the target ion mode
      an_with_conf <- names(an_meta)[sapply(an_meta, function(m) {
        grepl(target_mode, m["ion_mode"], ignore.case = TRUE)
      })]
      if (!length(an_with_conf)) an_with_conf <- names(an_meta)

      # Step 3: if multiple files per sample (multi-analysis), restrict by naming.
      # Exclude files whose names contain the wrong polarity token (NEG for POS mode,
      # POS for NEG mode). Files with no polarity token in their name pass through.
      excl_pol <- if (pos_mode_st) "(?i)(^|[_-])NEG([_-]|\\.[a-zA-Z])"
                  else             "(?i)(^|[_-])POS([_-]|\\.[a-zA-Z])"
      bio_raw  <- all_bio_raw[!grepl(excl_pol, basename(all_bio_raw), perl = TRUE)]

      # Step 4: if confirmed source files directly match some bio files, try to
      # narrow further by the file-name suffix pattern of those confirmed files
      # (e.g. all confirmed files end in "_POS.mzML" → keep only *_POS.mzML).
      direct <- all_bio_raw[basename(all_bio_raw) %in% conf_src_st]
      if (length(direct) > 0) {
        sfx <- unique(regmatches(basename(direct),
                                 regexpr("_[^_]+\\.[^.]+$", basename(direct))))
        if (length(sfx) == 1 && nchar(sfx) > 2) {
          pat     <- paste0(gsub("([.+*?\\[\\]()|])", "\\\\\\1", sfx), "$")
          sub_raw <- all_bio_raw[grepl(pat, basename(all_bio_raw))]
          if (length(sub_raw) > 0) {
            bio_raw <- sub_raw
            cat("  file suffix from conf_src:", sfx, "\n")
          }
        }
      }

      # Both-roles check within the selected files
      has_par_st <- length(src_par_st) == 0 || any(basename(bio_raw) %in% src_par_st)
      has_met_st <- length(src_met_st) == 0 || any(basename(bio_raw) %in% src_met_st)
      if (!has_par_st || !has_met_st) {
        miss <- c(if (!has_par_st) "parent", if (!has_met_st) "metabolite")
        cat("  WARNING: selected files lack confirmed", paste(miss, collapse=" + "),
            "— ratio not computable for this dataset\n")
      }
    } else {
      bio_raw      <- all_bio_raw
      an_with_conf <- names(an_meta)
      cat("  NOTE: no confirmed source_file — using all analyses\n")
    }

    bio_assays_per_ds[[ds]] <- an_with_conf
    bio_files_per_ds[[ds]]  <- bio_raw

    if (length(an_with_conf) > 1)
      cat("  NOTE: confirmed files span", length(an_with_conf),
          "analyses:", paste(an_with_conf, collapse = ", "), "\n")
  }

  # ── Both-roles check (all dataset types) ─────────────────────────────────
  if (!is.null(bio_files_per_ds[[ds]])) {
    ds_roles <- unique(conf_final$role[conf_final$dataset_id == ds])
    if (!all(c("parent", "metabolite") %in% ds_roles))
      cat("  WARNING: only role(s) [", paste(ds_roles, collapse=", "),
          "] confirmed — ratio not computable\n")
  }

  n_files <- if (is.list(bio_files_per_ds[[ds]]) && !is.character(bio_files_per_ds[[ds]]))
               sum(sapply(bio_files_per_ds[[ds]], length))
             else
               length(bio_files_per_ds[[ds]])
  cat("  →", n_files, "biological files |",
      length(bio_assays_per_ds[[ds]]), "assay(s)\n")
}
```

```{r bio-file-counts}
# Summary table: biological files + assay fractionation per confirmed dataset.
# Datasets where n_assays > 1 will need separate Spectra objects and
# separate RT/mz windows when scaling to all confirmed datasets.
bio_counts <- data.frame(
  dataset_id  = names(bio_files_per_ds),
  n_bio_files = sapply(bio_files_per_ds, function(x) {
    if (is.list(x) && !is.character(x)) sum(sapply(x, length)) else length(x)
  }),
  n_assays    = sapply(names(bio_files_per_ds), function(ds) {
    length(bio_assays_per_ds[[ds]])
  }),
  multi_assay = sapply(names(bio_files_per_ds), function(ds) {
    length(bio_assays_per_ds[[ds]]) > 1
  }),
  assay_ids   = sapply(names(bio_files_per_ds), function(ds) {
    paste(basename(bio_assays_per_ds[[ds]]), collapse = "; ")
  }),
  row.names = NULL
)
bio_counts <- dplyr::left_join(bio_counts,
  ds_summary[, c("dataset_id", "repository", "matrix", "n_subjects")],
  by = "dataset_id")
bio_counts
```

# Single-dataset prototype — MTBLS1866

MTBLS1866 is the cleanest candidate: Q Exactive Plus, RP-C18, DDA,
5-group COVID severity design, both parent and 5-OH metabolite confirmed
at strict / standard. We walk through the full EIC → peak detection
pipeline on a 10-sample subset, then decide on filtering rules before
scaling to all confirmed datasets.

### m/z and RT window strategy

**m/z:** We use the **median observed precursor m/z** of the confirmed
SIRIUS features for this dataset (stored in `conf_final$mz`, extracted
from the actual MS2 scan headers during confirmation). We do not use
the theoretical mass. Reasons:

1. The instrument's mass accuracy may be systematically shifted (e.g.,
   a few ppm offset due to recalibration). The observed m/z absorbs
   this shift automatically.
2. Different adducts ([M+H]+, [M+Na]+, ...) give very different m/z
   values. Using the observed m/z means the EIC window is anchored to
   whichever adduct SIRIUS confirmed, without having to hard-code adduct
   logic here.
3. Only adducts present in both roles are retained (see shared-adducts
   rule above), so the m/z we extract here is guaranteed to be
   chemically consistent across parent and metabolite.

**RT window:** Confirmed RT ± `RT_TOL_S` seconds. The tolerance is set
conservatively (30 s) to account for run-to-run gradient drift within
the same study. It can be tightened if multiple peaks appear within the
window.

```{r proto-config}
dataset_proto <- "MTBLS1866"
PPM       <- 20
RT_TOL_S  <- 30          # ± seconds around the confirmed RT window

proto_conf_all <- conf_final[conf_final$dataset_id == dataset_proto, ]

# --- Resolve a single assay for the prototype --------------------------
# Multi-assay studies host several assays per dataset (e.g. LC-MS positive
# / LC-MS negative / GCxGC-MS); each carries its own RT and m/z space.
# The prototype runs on ONE assay — pick the one with the most confirmed
# features that still has BOTH parent and metabolite present.
assay_tab <- proto_conf_all |>
  dplyr::group_by(assay) |>
  dplyr::summarise(
    n_par = sum(role == "parent"),
    n_met = sum(role == "metabolite"),
    n_tot = dplyr::n(),
    .groups = "drop") |>
  dplyr::filter(n_par > 0 & n_met > 0) |>
  dplyr::arrange(dplyr::desc(n_tot))
if (!nrow(assay_tab))
  stop("No assay in ", dataset_proto, " has both parent and metabolite confirmed")
assay_proto <- assay_tab$assay[1]
cat("Available assays for", dataset_proto, ":\n"); print(assay_tab)
cat("Selected prototype assay:", assay_proto, "\n\n")

proto_conf <- proto_conf_all[proto_conf_all$assay == assay_proto, ]

# --- m/z anchor from the observed precursor m/z of confirmed features ---
# IMPORTANT: do NOT average across adducts. [M+H]+ (~346.12) and [M+Na]+
# (~368.10) for omeprazole sit ~22 Da apart; taking a median across them
# lands between two real peaks and points the EIC at empty space. We
# therefore pick the dominant adduct per role (most confirmed features)
# and median the observed m/z only within that adduct.
#
# conf_final$mz comes from precursorMz() of the matched MS2 scan in the
# confirmation step. The within-adduct median absorbs scan-level drift
# without crossing chemistry.
dominant_adduct <- function(df) {
  tab <- table(df$adduct_inferred[!is.na(df$adduct_inferred)])
  if (!length(tab)) return(NA_character_)
  names(tab)[which.max(tab)]
}

ADDUCT_PARENT <- dominant_adduct(proto_conf[proto_conf$role == "parent",     ])
ADDUCT_METAB  <- dominant_adduct(proto_conf[proto_conf$role == "metabolite", ])

MZ_PARENT <- median(
  proto_conf$mz[proto_conf$role == "parent" &
                proto_conf$adduct_inferred == ADDUCT_PARENT],
  na.rm = TRUE)
MZ_METAB  <- median(
  proto_conf$mz[proto_conf$role == "metabolite" &
                proto_conf$adduct_inferred == ADDUCT_METAB],
  na.rm = TRUE)

# Warn loudly if more than one adduct is present per role — we will only
# extract one m/z window; the other adduct's features are dropped.
all_par_adducts <- unique(proto_conf$adduct_inferred[proto_conf$role == "parent"])
all_met_adducts <- unique(proto_conf$adduct_inferred[proto_conf$role == "metabolite"])
if (length(setdiff(all_par_adducts, NA)) > 1)
  cat("NOTE: parent has multiple adducts —", paste(all_par_adducts, collapse=", "),
      "— EIC will use only", ADDUCT_PARENT, "\n")
if (length(setdiff(all_met_adducts, NA)) > 1)
  cat("NOTE: metabolite has multiple adducts —", paste(all_met_adducts, collapse=", "),
      "— EIC will use only", ADDUCT_METAB, "\n")

# --- RT window from the confirmation step (already in seconds) ---
# Restrict to the same adduct as the m/z anchor — different adducts can
# elute at the same time, but mixing their RTs into the window is still
# noisy when one adduct dominates.
rt_par <- range(proto_conf$rt[proto_conf$role == "parent" &
                              proto_conf$adduct_inferred == ADDUCT_PARENT],
                na.rm = TRUE)
rt_met <- range(proto_conf$rt[proto_conf$role == "metabolite" &
                              proto_conf$adduct_inferred == ADDUCT_METAB],
                na.rm = TRUE)
rt_par_win <- rt_par + c(-RT_TOL_S, RT_TOL_S)
rt_met_win <- rt_met + c(-RT_TOL_S, RT_TOL_S)

cat("parent     adduct =", ADDUCT_PARENT,
    "| mz =", round(MZ_PARENT, 5), "\n")
cat("metabolite adduct =", ADDUCT_METAB,
    "| mz =", round(MZ_METAB,  5), "\n")
cat("parent     mz window:", MZ_PARENT * (1 + PPM*1e-6 * c(-1, 1)), "\n")
cat("metabolite mz window:", MZ_METAB  * (1 + PPM*1e-6 * c(-1, 1)), "\n")
cat("parent     rt window:", round(rt_par_win, 1), "s\n")
cat("metabolite rt window:", round(rt_met_win, 1), "s\n")
cat("\nMASST hit source files used for confirmation (selected assay only):\n")
print(unique(proto_conf[, c("role", "xcms_fts", "confidence", "rt", "mz",
                             "adduct_inferred", "source_file")]))
```

## Stage 2a — pick 10 biological files

File selection comes from the metadata-based `bio_files_per_ds` built in the
section above — no filename heuristics, no need to download first.

```{r proto-file-list}
# bio_files_per_ds[[ds]] is a flat character vector for single-assay datasets
# and a named list (keyed by assay file name) for multi-assay datasets.
# Use the assay resolved in proto-config (assay_proto) to pull the exact
# file list for that assay — no name-pattern heuristics.
tmp_bio <- bio_files_per_ds[[dataset_proto]]
if (is.list(tmp_bio) && !is.character(tmp_bio)) {
  if (assay_proto %in% names(tmp_bio)) {
    cat("Multi-assay study: selecting assay '", assay_proto,
        "' (resolved from confirmed source files)\n", sep = "")
    all_bio <- tmp_bio[[assay_proto]]
  } else {
    cat("WARNING: assay_proto '", assay_proto,
        "' not in bio_files_per_ds — falling back to union of all assays\n",
        sep = "")
    all_bio <- unlist(tmp_bio, use.names = FALSE)
  }
} else {
  all_bio <- tmp_bio
}

# MsBackendMetaboLights expects bare filenames (no path prefix).
# The assay table stores full relative paths like FILES/DERIVED_FILES/pos/B-107.mzXML;
# mtbls_sync_data_files and backendInitialize both take basename only.
all_bio <- basename(all_bio)

cat("Biological files available for", dataset_proto, "/", assay_proto, ":",
    length(all_bio), "\n")

# --- Confirmed source files (positive-control "test" files) -------------
# These are the files where SIRIUS actually matched parent and/or metabolite
# in this assay. We add them on top of the random 10-sample subset so that
# (a) we can verify our EIC/peak-detection settings find signal where
#     SIRIUS already saw it, and
# (b) the random subset itself stays unbiased (10 truly random samples,
#     plus N test files appended).
test_files_proto <- unique(basename(
  proto_conf$source_file[!is.na(proto_conf$source_file)]))
test_files_proto <- test_files_proto[test_files_proto %in% all_bio]
cat("Confirmed test files in this assay:", length(test_files_proto), "\n")
if (length(test_files_proto)) print(test_files_proto)

# --- Random 10-sample subset, drawn from the REMAINING pool -------------
# Excluding test files from the random pool prevents double-counting and
# keeps the random draw representative of "unseen" samples.
random_pool <- setdiff(all_bio, test_files_proto)
set.seed(1)
random_files <- sample(random_pool, min(10, length(random_pool)))

proto_files <- unique(c(test_files_proto, random_files))
cat("Selected (", length(test_files_proto), "test +",
    length(random_files), "random =", length(proto_files), "files):\n", sep = " ")
print(proto_files)
```

## Stage 2b — sync and load (on disk)

```{r proto-sync-load}
mtbls_sync_data_files(mtblsId = dataset_proto, fileName = basename(proto_files))

sp <- Spectra(backendInitialize(
  MsBackendMetaboLights(),
  mtblsId = dataset_proto,
  fileName = basename(proto_files),
  offline = TRUE
))
ms1 <- filterMsLevel(sp, 1L)

cat("Files loaded :", length(unique(dataOrigin(ms1))), "\n")
cat("MS1 spectra  :", length(ms1), "\n")
```

## EIC extraction

We use `Chromatograms::chromExtract()` to extract EICs in a single pass
over the Spectra object. A *peak table* defines one extraction window per
(role × file): `rtMin`/`rtMax` from the confirmed-RT windows and
`mzMin`/`mzMax` from the target mass ± PPM tolerance. The `by` argument
tells `chromExtract` which peak-table columns to use as grouping keys when
matching spectra — here `msLevel` (MS1 only) and `dataOrigin` (one file per
row). The resulting `Chromatograms` object carries the full peak table as
`chromData()`, so subsetting by role is straightforward.

```{r proto-chromatograms, fig.width = 9, fig.height = 6}
library(Chromatograms)

# One extraction window per role × file
all_orig <- unique(dataOrigin(ms1))

peak_tbl <- rbind(
  data.frame(
    role       = "parent",
    msLevel    = 1L,
    dataOrigin = all_orig,
    rtMin      = rt_par_win[1],  rtMax = rt_par_win[2],
    mzMin      = MZ_PARENT * (1 - PPM * 1e-6),
    mzMax      = MZ_PARENT * (1 + PPM * 1e-6),
    stringsAsFactors = FALSE
  ),
  data.frame(
    role       = "metabolite",
    msLevel    = 1L,
    dataOrigin = all_orig,
    rtMin      = rt_met_win[1],  rtMax = rt_met_win[2],
    mzMin      = MZ_METAB * (1 - PPM * 1e-6),
    mzMax      = MZ_METAB * (1 + PPM * 1e-6),
    stringsAsFactors = FALSE
  )
)

chr_all  <- Chromatograms(ms1)
chr_eics <- chromExtract(chr_all, peak_tbl, by = c("msLevel", "dataOrigin"))

chr_par  <- chr_eics[chromData(chr_eics)$role == "parent"]
chr_met  <- chr_eics[chromData(chr_eics)$role == "metabolite"]

# plotChromatogramsOverlay sets ylim from the first chromatogram it draws,
# so any file with higher intensity gets clipped.  Pre-compute ylim across
# all EICs for each role and pass it explicitly.
eic_ylim <- function(chr) {
  ints <- unlist(lapply(peaksData(chr), `[[`, "intensity"), use.names = FALSE)
  c(0, max(ints, na.rm = TRUE) * 1.05)   # 5 % headroom
}

# Call with package qualifier — xcms masks plotChromatogramsOverlay
par(mfrow = c(2, 1), mar = c(3, 4, 2, 1))
Chromatograms::plotChromatogramsOverlay(
  chr_par,
  main = paste(dataset_proto, "— omeprazole (parent)"),
  ylim = eic_ylim(chr_par))
Chromatograms::plotChromatogramsOverlay(
  chr_met,
  main = paste(dataset_proto, "— 5-OH-omeprazole (metabolite)"),
  ylim = eic_ylim(chr_met))
```

## Peak detection

`MsCoreUtils::localMaxima` is applied to the full intensity vector
(including NAs; the C code never returns NA positions as maxima).
`hws` sets the half-window in scans; a peak must be the maximum within
±`hws` neighbours. RT and AUC are computed separately via MsQuality.

```{r proto-peaks}
count_peaks <- function(chr, hws = 3L, pct_max = 0.1) {
  ints <- intensity(chr)
  vapply(ints, function(x) {
    mx <- max(x, na.rm = TRUE)
    if (!is.finite(mx)) return(0L)
    x_s <- x[!is.na(x)]
    lm  <- MsCoreUtils::localMaxima(x_s, hws = hws)
    sum(lm & x_s >= pct_max * mx)
  }, integer(1L))
}

peaks_par <- count_peaks(chr_par)
peaks_met <- count_peaks(chr_met)

cat("Parent     — files with ≥1 peak:", sum(peaks_par > 1), "/",
    length(peaks_par), "\n")
cat("Metabolite — files with ≥1 peak:", sum(peaks_met > 1), "/",
    length(peaks_met), "\n")
```

Here check that we have not more than one peak.
For now, on the small subset it looks great. and the sanity check look very  good.
I would do another check on the full dataset before we go on to all the data.


## Positive-control check

SIRIUS-confirmed files are positive controls: our EIC pipeline must
detect a peak there. Failures flag a settings problem (PPM, RT window,
or adduct choice). We run this check BEFORE dropping empty EICs so we
notice if a confirmed file would be silently discarded.

```{r proto-verify-signal}
conf_par_files <- unique(basename(proto_conf$source_file[
  proto_conf$role == "parent" & !is.na(proto_conf$source_file)]))
conf_met_files <- unique(basename(proto_conf$source_file[
  proto_conf$role == "metabolite" & !is.na(proto_conf$source_file)]))

par_files_all <- basename(chromData(chr_par)[["dataOrigin"]])
met_files_all <- basename(chromData(chr_met)[["dataOrigin"]])

check_confirmed <- function(files, n_peaks, conf_files, role) {
  idx <- which(files %in% conf_files)
  cat(role, "— confirmed files loaded:", length(idx),
      "| with >=1 peak:", sum(n_peaks[idx] > 0), "\n")
  missed <- files[idx[n_peaks[idx] == 0]]
  if (length(missed)) cat("  WARNING missed:", paste(missed, collapse=", "), "\n")
}

check_confirmed(par_files_all, peaks_par, conf_par_files, "Parent    ")
check_confirmed(met_files_all, peaks_met, conf_met_files, "Metabolite")
```

## Drop empty EICs

A chromatogram with zero detected peaks carries no usable signal — its
trapezoidal integral is just baseline drift, and `peakBoundary()` has
nothing to anchor to. We drop empty EICs per-role (NOT per-file) so
asymmetric cases (parent visible / metabolite not, or vice versa) are
still represented in the sample table; the *file* survives even though
one of its two EICs has been dropped.

```{r proto-drop-empty}
keep_par <- peaks_par > 0
keep_met <- peaks_met > 0

chr_par_ok <- chr_par[keep_par]
chr_met_ok <- chr_met[keep_met]

cat("Parent     EICs kept :", sum(keep_par), "/", length(keep_par),
    " (dropped:", sum(!keep_par), ")\n")
cat("Metabolite EICs kept :", sum(keep_met), "/", length(keep_met),
    " (dropped:", sum(!keep_met), ")\n")

# Track which files retain each role — used by the sample table to
# assign a detection tier (both / parent_only / metab_only / neither).
files_par_ok <- basename(chromData(chr_par_ok)[["dataOrigin"]])
files_met_ok <- basename(chromData(chr_met_ok)[["dataOrigin"]])
```

## Peak boundaries and narrow re-extraction

`Chromatograms::peakBoundary()` returns the rtime boundaries of the
dominant peak per EIC, derived from a baseline-quantile threshold. We
re-extract a narrower EIC over `[left_boundary - PAD, right_boundary +
PAD]` from the original `chr_all` Spectra so trapezoidal AUC and FWHM
are computed strictly on the peak region — not on the broad RT window
we used for detection. The small PAD captures the tails the
baseline-cut clips.

```{r proto-peakbounds}
BOUND_PAD_S <- 5  # seconds of padding on each side of the boundary

# Re-extract narrow EICs.  We feed peakBoundary back into chromExtract
# via a fresh peak_tbl so the result has the same `dataOrigin` ordering
# as the input — important for downstream joins.
narrow_extract <- function(chr, role, mz_centre, ppm) {
  bnd <- Chromatograms::peakBoundary(chr)            # n x 2 matrix
  cd  <- chromData(chr)
  ok  <- is.finite(bnd[, "left_boundary"]) &
         is.finite(bnd[, "right_boundary"])
  if (!any(ok)) return(chr[FALSE])
  pt  <- data.frame(
    role       = role,
    msLevel    = 1L,
    dataOrigin = cd[["dataOrigin"]][ok],
    rtMin      = bnd[ok, "left_boundary"]  - BOUND_PAD_S,
    rtMax      = bnd[ok, "right_boundary"] + BOUND_PAD_S,
    mzMin      = mz_centre * (1 - ppm * 1e-6),
    mzMax      = mz_centre * (1 + ppm * 1e-6),
    stringsAsFactors = FALSE)
  chromExtract(chr_all, pt, by = c("msLevel", "dataOrigin"))
}

chr_par_narrow <- narrow_extract(chr_par_ok, "parent",     MZ_PARENT, PPM)
chr_met_narrow <- narrow_extract(chr_met_ok, "metabolite", MZ_METAB,  PPM)

cat("Narrow EICs — parent:", length(chr_par_narrow),
    "| metabolite:", length(chr_met_narrow), "\n")
```

## EIC metrics via MsQuality

`MsQuality::calculateMetrics()` is the high-level dispatcher: it
inspects the class of `object` and forwards to the right backend
(`calculateMetricsFromChromatograms()` for our `Chromatograms` input,
`…FromSpectra` for raw spectra, `…FromMsExperiment` for full
experiments). Using the dispatcher keeps this chunk class-agnostic —
if we later switch to processing `MsExperiment` objects, the call
doesn't change. It returns one row per `dataOrigin` with columns named
after the requested metrics. We pull the integrated area
(`areaUnderTic`), the chromatographic peak width at half-max
(`xicFwhm`, computed internally from a fresh `peakBoundary`), the apex
prominence (`peakProminence`), and the raw `peakWidth` (RT range
covered by data) for diagnostics. The `peakCount` metric here counts
*MS scans*, not chromatographic peaks — we already established those
above with the local-maxima detector.

```{r proto-msquality}
library(MsQuality)

EIC_METRICS <- c("areaUnderTic", "xicFwhm", "peakProminence", "peakWidth")

tidy_metrics <- function(chr, prefix) {
  if (!length(chr))
    return(data.frame(file = character(0)))
  m <- calculateMetrics(chr, metrics = EIC_METRICS, msLevel = 1L)
  data.frame(
    file       = basename(rownames(m)),
    auc        = m[["areaUnderTic"]],
    fwhm       = m[["xicFwhm"]],
    prominence = m[["peakProminence"]],
    pkwidth    = m[["peakWidth"]],
    row.names  = NULL,
    stringsAsFactors = FALSE) |>
    setNames(c("file", paste0(prefix, "_",
                              c("auc","fwhm","prominence","pkwidth"))))
}

metrics_par <- tidy_metrics(chr_par_narrow, "par")
metrics_met <- tidy_metrics(chr_met_narrow, "met")

metrics_par
metrics_met
```

## FWHM outlier check

Within the same dataset the chromatographic peak width should be
roughly constant — the gradient is fixed, the same compound elutes the
same way across injections. A file whose FWHM departs strongly from
the median (here: > 3 × MAD) usually means the EIC is integrating two
fused peaks, baseline noise, or a co-eluting isobar — i.e., the AUC
for that file is suspect and should be flagged before it enters the
ratio.

```{r proto-fwhm-outliers}
flag_fwhm <- function(df, col) {
  v   <- df[[col]]
  med <- median(v, na.rm = TRUE)
  mad <- stats::mad(v, na.rm = TRUE)
  if (!is.finite(mad) || mad == 0)
    return(rep(FALSE, length(v)))
  abs(v - med) > 3 * mad
}

metrics_par$par_fwhm_outlier <- flag_fwhm(metrics_par, "par_fwhm")
metrics_met$met_fwhm_outlier <- flag_fwhm(metrics_met, "met_fwhm")

cat("Parent     FWHM (s): median =",
    round(median(metrics_par$par_fwhm, na.rm = TRUE), 1),
    "| outliers (>3 MAD):", sum(metrics_par$par_fwhm_outlier), "\n")
cat("Metabolite FWHM (s): median =",
    round(median(metrics_met$met_fwhm, na.rm = TRUE), 1),
    "| outliers (>3 MAD):", sum(metrics_met$met_fwhm_outlier), "\n")
```

### Visualise the flagged peaks

Plot each FWHM-outlier EIC alongside a reference EIC drawn from the
file closest to the median FWHM. A genuine outlier will look obviously
deformed against the reference — double apex, asymmetric tail, broad
shoulder, or noisy baseline. If the outlier looks structurally similar
to the reference, the flag is a false alarm (typically driven by a
small absolute FWHM in a dataset with low FWHM variance) and the file
should stay in the ratio.

```{r proto-fwhm-outliers-plot, fig.width = 10, fig.height = 5}
plot_fwhm_outliers <- function(chr, metrics, prefix, role_label) {
  oc <- paste0(prefix, "_fwhm_outlier")
  fc <- paste0(prefix, "_fwhm")
  out_files <- metrics$file[metrics[[oc]] %in% TRUE]
  if (!length(out_files) || !length(chr)) {
    cat("No FWHM outliers for ", role_label, " — nothing to plot.\n", sep = "")
    return(invisible())
  }
  files_chr <- basename(chromData(chr)[["dataOrigin"]])
  med_fwhm  <- median(metrics[[fc]], na.rm = TRUE)
  ref_file  <- metrics$file[which.min(abs(metrics[[fc]] - med_fwhm))]
  to_show   <- unique(c(ref_file, out_files))

  rows <- lapply(seq_along(files_chr), function(i) {
    f <- files_chr[i]
    if (!(f %in% to_show)) return(NULL)
    pd <- peaksData(chr)[[i]]
    fwhm_v <- metrics[[fc]][match(f, metrics$file)]
    data.frame(
      file      = f,
      rtime     = pd$rtime,
      intensity = pd$intensity,
      panel     = paste0(f, "\nFWHM = ", round(fwhm_v, 1), " s"),
      kind      = ifelse(f == ref_file,
                         "reference (median FWHM)", "outlier"),
      stringsAsFactors = FALSE)
  })
  df <- do.call(rbind, rows)
  # Stable panel order: reference first, then outliers sorted by FWHM
  fwhm_lookup <- setNames(metrics[[fc]], metrics$file)
  df$panel <- factor(df$panel,
    levels = unique(df$panel[order(df$file != ref_file,
                                   fwhm_lookup[df$file])]))

  ggplot(df, aes(rtime, intensity, colour = kind)) +
    geom_line(linewidth = 0.7) +
    facet_wrap(~ panel, scales = "free_y") +
    scale_colour_manual(
      values = c("reference (median FWHM)" = "grey40",
                 "outlier"                  = "#D7191C")) +
    labs(title    = paste(dataset_proto, "—", role_label, "FWHM outliers"),
         subtitle = paste0("Median FWHM = ", round(med_fwhm, 1),
                           " s | outlier rule: |FWHM - median| > 3 * MAD"),
         x = "Retention time (s)", y = "Intensity", colour = NULL) +
    theme_bw() +
    theme(legend.position = "bottom",
          strip.background = element_rect(fill = "grey92"),
          strip.text       = element_text(size = 8))
}

plot_fwhm_outliers(chr_par_narrow, metrics_par, "par", "parent")
plot_fwhm_outliers(chr_met_narrow, metrics_met, "met", "metabolite")
```

## Canonical metadata resolution

Three sources contribute per-file metadata, with a fixed priority:

1. **Repository per-sample file** (`s_*.txt` for MetaboLights,
   `*_metadata.tsv` for MassIVE/Pan-ReDU, `SUBJECT_SAMPLE_FACTORS` in
   mwtab files for Metabolomics Workbench). Read once per dataset, joined
   to file basenames via the assay table's `Sample Name → Derived
   Spectral Data File` map.
2. **`dataset_curation_table.csv`** — the human-readable curation
   notebook. Names the raw column per canonical field with coverage
   tags (`"Yes - Comment[Patient status] (100%)"`) and stores
   per-dataset constants (`country = "Italy (Novara University
   Hospital, …)"` for MTBLS1866, taken from the paper).
3. **Linked paper** — only where the deposit is silent. For MTBLS1866
   the paper (Barberis et al. 2020, PMC7696386) supplies `country` as
   a per-dataset constant, and gives per-arm aggregates for sex and
   age (Table 1) that are NOT mappable to individual files.

Each canonical field is therefore one of three granularities:

| granularity              | example field      | how it lands in `meta`            |
|--------------------------|--------------------|------------------------------------|
| per-file                 | `health_status`    | row-level join from the s-file     |
| per-dataset constant     | `country`          | broadcast scalar from curation     |
| per-group aggregate only | `sex`, `age`       | NA at per-file level; summarised separately |

A **`subject_id`** is also tracked per file. For MTBLS1866 the
deposit is 1:1 (`Source Name == Sample Name`, no replicates), so it
collapses trivially. For datasets with technical replicate injections
or multi-timepoint sampling, this column is what we will deduplicate
on before stats — counting one subject once, not once per injection.

```{r resolve-metadata}
# This block is hard-wired to the prototype (MTBLS1866); the
# all-datasets pass turns it into a function that branches on dataset_id
# prefix and reads the curation row's column-name strings.

curation_row <- curation[curation$dataset_id == dataset_proto, ]

s_tbl <- read.delim(
  file.path("dataset_metadata", paste0("s_", dataset_proto, ".txt")),
  check.names = FALSE)
a_tbl <- read.delim(
  file.path("dataset_metadata", assay_proto),
  check.names = FALSE)

# file basename -> Sample Name via assay's Derived Spectral Data File
file_to_sample <- setNames(
  a_tbl[["Sample Name"]],
  basename(a_tbl[["Derived Spectral Data File"]]))

meta <- data.frame(
  file       = proto_files,
  subject_id = unname(file_to_sample[proto_files]),
  stringsAsFactors = FALSE)

# Per-file from repo s-file (MTBLS1866-specific column names — in the
# generalised version, parse `curation_row$health_status` for the
# bracketed column name)
sidx <- match(meta$subject_id, s_tbl[["Sample Name"]])
meta$health_status <- s_tbl[["Comment[Patient status]"]][sidx]
meta$body_site     <- s_tbl[["Characteristics[Organism part]"]][sidx]

# Per-dataset constant from curation/paper
meta$country       <- "Italy"   # Barberis 2020 — Novara University Hospital

# Not mappable per file for this deposit (paper Table 1 is per-arm only)
meta$sex           <- NA_character_
meta$age           <- NA_real_

cat("Canonical metadata for", dataset_proto, "(", nrow(meta), "files):\n")
cat("  subject_id    : ", length(unique(meta$subject_id)),
    "unique subjects (", nrow(meta), "files = ",
    round(nrow(meta)/length(unique(meta$subject_id)),1), "files/subject)\n")
cat("  health_status : filled =", sum(!is.na(meta$health_status)),
    "/", nrow(meta), "\n")
cat("  body_site     : filled =", sum(!is.na(meta$body_site)), "/", nrow(meta), "\n")
cat("  country       : filled =", sum(!is.na(meta$country)),
    "/", nrow(meta), "  (constant from paper)\n")
cat("  sex / age     : NA per file (paper Table 1 is per-arm aggregate only)\n\n")
print(meta)
```

## Sample table — detection tiers

A sample's *detection tier* records which roles produced a peak:

| tier         | parent peak | metabolite peak | ratio computable                       |
|--------------|-------------|-----------------|----------------------------------------|
| `both`       | yes         | yes             | yes — `log2(par/met)`                  |
| `parent_only`| yes         | no              | no — censored high (low CYP2C19)       |
| `metab_only` | no          | yes             | no — censored low (high CYP2C19)       |
| `neither`    | no          | no              | sample contributes nothing             |

The ratio is computed *only* for `both`. Asymmetric cases stay in the
table as categorical readouts — we do not impute the missing AUC and we
do not set the ratio to 1 (which would imply equal amounts, the wrong
conclusion).

```{r sample-table}
# Outer-join the two metric tables so files in only one role survive.
sample_tbl <- merge(metrics_par, metrics_met, by = "file", all = TRUE)
sample_tbl <- merge(data.frame(file = proto_files), sample_tbl,
                    by = "file", all.x = TRUE)
sample_tbl <- merge(sample_tbl, meta, by = "file", all.x = TRUE)

sample_tbl$detection_tier <- with(sample_tbl, ifelse(
  !is.na(par_auc) & !is.na(met_auc), "both",
  ifelse(!is.na(par_auc) & is.na(met_auc), "parent_only",
  ifelse( is.na(par_auc) & !is.na(met_auc), "metab_only",
                                            "neither"))))

# Ratio only where both detected
sample_tbl$log2_ratio <- with(sample_tbl, ifelse(
  detection_tier == "both", log2(par_auc / met_auc), NA_real_))

# Subject-level deduplication: if a subject contributed multiple
# injections (not the case for MTBLS1866, but routine elsewhere), keep
# the row with the highest summed AUC so a single subject contributes a
# single ratio.  For 1:1 datasets this is a no-op.
n_before <- nrow(sample_tbl)
sample_tbl <- sample_tbl |>
  dplyr::group_by(subject_id) |>
  dplyr::slice_max(
    order_by = dplyr::coalesce(par_auc, 0) + dplyr::coalesce(met_auc, 0),
    n = 1, with_ties = FALSE) |>
  dplyr::ungroup()
n_after <- nrow(sample_tbl)
if (n_before != n_after)
  cat("Subject-level dedup: collapsed", n_before, "->", n_after,
      "(replicates per subject:",
      round(n_before / length(unique(sample_tbl$subject_id)), 2), ")\n")

cat("Detection tier counts:\n"); print(table(sample_tbl$detection_tier))
sample_tbl
```

## Ratio handling — the parent-only file

In this prototype run we have at least one file with parent detected
but no metabolite. Biologically, for CYP2C19 phenotyping, that profile
is **informative**: it points to a poor metabolizer (the substrate is
present, the enzyme is not turning it over). Three options for scoring
it:

1. **Drop the sample.** Loses real signal — poor metabolizers are
   exactly the phenotype we want to detect.
2. **Set ratio = 1.** Wrong: implies parent ≈ metabolite, which is the
   profile of an *extensive* metabolizer. This would invert the
   biological meaning.
3. **Treat as right-censored.** Keep the sample, note `detection_tier
   == "parent_only"`, and in the stats stage use either (a) a
   detection-status logistic test (parent_only count vs both count per
   group) or (b) impute `met_auc` to the per-dataset noise floor (e.g.
   half the smallest detected `met_auc`) and label the resulting ratio
   as a lower bound.

The prototype uses option 3a: the ratio stays NA for `parent_only` and
the file enters the categorical analysis. We revisit imputation when
scaling to all datasets, because the noise floor is dataset-specific.

```{r stats}
both <- sample_tbl[sample_tbl$detection_tier == "both" &
                   !sample_tbl$par_fwhm_outlier %in% TRUE &
                   !sample_tbl$met_fwhm_outlier %in% TRUE, ]

cat("Samples entering ratio stats:", nrow(both),
    " | parent_only (censored):",
    sum(sample_tbl$detection_tier == "parent_only"),
    " | metab_only:",
    sum(sample_tbl$detection_tier == "metab_only"),
    " | neither:",
    sum(sample_tbl$detection_tier == "neither"), "\n")

# wilcox.test(log2_ratio ~ HealthStatus, data = both)   # enable when metadata joined
```

## Plots

```{r plots}
# log2 ratio distribution (only `both` samples).  Censored cases
# (parent_only / metab_only) are shown as a side panel count for context
# — they don't have a numeric ratio.
ggplot(both, aes(x = "samples", y = log2_ratio)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey40") +
  labs(x = NULL, y = "log2(omeprazole / 5-OH-omeprazole)",
       title = paste0(dataset_proto, " — ratio distribution (n=", nrow(both), ")"),
       subtitle = paste0("Censored: parent_only=",
                         sum(sample_tbl$detection_tier == "parent_only"),
                         ", metab_only=",
                         sum(sample_tbl$detection_tier == "metab_only"))) +
  theme_bw()

# Parent vs metabolite area, with detection tier as colour so the
# censored cases are visible on the axes (along x=NA or y=NA strip).
ggplot(sample_tbl,
       aes(par_auc, met_auc, colour = detection_tier,
           shape = par_fwhm_outlier | met_fwhm_outlier)) +
  geom_point(alpha = 0.85, size = 2.5) +
  scale_x_log10() + scale_y_log10() +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 4),
                     name = "FWHM outlier") +
  labs(x = "omeprazole area", y = "5-OH-omeprazole area",
       colour = "Detection tier", title = dataset_proto) +
  theme_bw()
```

# Metadata limitations and unsupervised exploration

What we resolved per file for MTBLS1866 is the rule, not the
exception. Across the tier-B curated datasets, only a minority
of public deposits ship the biological covariates (sex, age, country,
treatment, BMI, ethnicity) that a phenotyping analysis would normally
adjust for. The MASST/Pan-ReDU `ATTRIBUTE_*` columns are usable when
the depositor populated them but are empty for most MetaboLights and
Metabolomics Workbench studies, and the linked papers — when they
exist — typically report demographics as per-arm aggregates (Table 1
sex/age) rather than as per-sample tables joinable to file names.

For our analysis this has two concrete consequences:

1. **Unadjustable confounding.** A ratio difference we observe between
   two datasets could reflect a real biological effect or simply
   a difference in sex / age / ancestry composition that we cannot
   model. Within-dataset contrasts (e.g. the 5-level COVID severity
   here) are insulated from this; cross-dataset meta-analysis is not.
2. **Limited stratification.** Within most deposits the only available
   stratifier is the study's primary design factor (case vs control,
   severity grade, …). Secondary effects we would normally investigate
   — e.g. is the CYP2C19 ratio sex-dependent in this cohort? — are not
   testable from the public data alone.

## Unsupervised structure as a metadata-gap diagnostic

A practical workaround is to run an unsupervised analysis on the
per-file metric matrix (`par_auc`, `met_auc`, `par_fwhm`, `met_fwhm`,
`log2_ratio`) and check whether the resulting structure aligns with
the *available* metadata. If a clear cluster split emerges that does
**not** track `health_status` (the only per-file factor we have), that
split is evidence of a latent covariate the deposit does not expose —
which is itself a finding worth reporting. The unsupervised view
becomes a missing-metadata diagnostic, not just an exploratory plot.

```{r unsup-pca, fig.width = 7, fig.height = 5}
# PCA over the per-file metric matrix, restricted to the `both` tier
# (numeric ratio available). Need at least 5 samples and at least 2
# non-constant columns to be meaningful.
#
# Implementation notes:
#   - sample_tbl is a tibble (dplyr pipeline above). as.matrix() on a
#     tibble drops row names, so we cannot rely on rownames(pc$x) to map
#     PC rows back to metadata.  We add an explicit `.row_id` column,
#     copy it into rownames(pc_mat), and read it back as integer.
#   - complete.cases() before prcomp is required: scale. = TRUE divides
#     by per-column sd, which is NaN if any NA is present.

both_for_pca <- as.data.frame(
  sample_tbl[sample_tbl$detection_tier == "both", ])
both_for_pca$.row_id <- seq_len(nrow(both_for_pca))

pc_cols <- c("par_auc","met_auc","par_fwhm","met_fwhm","log2_ratio")
pc_mat  <- as.matrix(both_for_pca[, pc_cols])
rownames(pc_mat) <- as.character(both_for_pca$.row_id)

# Keep non-constant columns
inf_cols <- apply(pc_mat, 2, function(x) {
  v <- stats::sd(x, na.rm = TRUE); !is.na(v) && v > 0 })
pc_mat <- pc_mat[, inf_cols, drop = FALSE]

# Drop rows with any NA across the kept columns
pc_mat <- pc_mat[stats::complete.cases(pc_mat), , drop = FALSE]

if (nrow(pc_mat) >= 5 && ncol(pc_mat) >= 2) {
  pc      <- stats::prcomp(pc_mat, scale. = TRUE)
  row_ids <- as.integer(rownames(pc$x))         # safely tracked
  pc_df   <- cbind(
    as.data.frame(pc$x[, 1:min(2, ncol(pc$x)), drop = FALSE]),
    both_for_pca[row_ids,
                 c("subject_id","health_status","country"),
                 drop = FALSE])

  ggplot(pc_df, aes(PC1, PC2,
                    colour = health_status,
                    label  = subject_id)) +
    geom_point(size = 3, alpha = 0.85) +
    geom_text(vjust = -0.7, size = 3, show.legend = FALSE) +
    labs(title    = paste(dataset_proto, "— PCA of EIC metrics"),
         subtitle = "Cluster splits not tracking health_status hint at unmeasured covariates",
         caption  = paste("n =", nrow(pc_df), "subjects | available factors:",
                          "health_status, body_site, country (constant)")) +
    theme_bw() + theme(legend.position = "bottom")
} else {
  cat("Skipping PCA — need >=5 `both` samples and >=2 non-constant metrics",
      "(have", nrow(pc_mat), "rows,", ncol(pc_mat), "informative cols)\n")
}
```

## A note for the paper

The same diagnostic, run dataset by dataset, becomes a quantitative
argument for richer mandatory metadata in public metabolomics
repositories: how many datasets show unsupervised structure that the
available metadata cannot explain? A handful would already justify
proposing a minimum-metadata standard (at least `BiologicalSex`,
`AgeInYears`, `Country`, and a structured `treatment_history` field
where ethical) at the level of MetaboLights / Workbench submission
templates and Pan-ReDU re-annotation. This finding is a free side
benefit of the pipeline — we get it from the same metric matrix that
feeds the ratio analysis, no extra computation.

# Sanity check

Does the result match the pharmacology literature? For this compound
pair, CYP2C19-extensive metabolisers show a lower parent/5-OH ratio
than poor metabolisers. If the prototype dataset is not CYP2C19-
stratified or PPI-intervention designed, re-select `dataset_proto`.

# References {.unnumbered}

::: {#refs}
:::
