---
title: "01 - Search"
subtitle: "Step 1: one structure query per target"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
---

# Overview

Each target is resolved to its reference MS/MS spectra, and their matches
against the public repositories are read from FASSTrecords. Library spectra that
are not a plausible ion of their target are set aside, and only deposits that
carry every target are kept.

**Input:** `compound.csv` (the targets). The hits come from the FASSTrecords
database, or from its cached extract `StructureMASST_output/fasstrecords_hits.csv`,
which the search writes when the database is available.

**Output:** `artifacts/hits_search.csv`, with the adduct of each hit, read by
step 2.

```{r setup}
source("../../R/setup.R")
source("../../R/masst.R")
source("../../R/metadata.R")
```

# Targets

```{r compound-spec}
compound_spec
```

Tryptophan and kynurenine: indoleamine 2,3-dioxygenase converts tryptophan to
kynurenine, and the balance between the two is the standard read-out of its
activity, which rises with age.

The search covers every deposit that carries both compounds. Which of them can
test the hypothesis (human blood from adults, with an age) is decided from their
metadata: step 2 first removes the deposits already known not to fit, then looks
for the rest in every source, and step 3 keeps those that fit.

# The search

FASSTrecords (DOI 10.5281/zenodo.18199544) holds a reference library of MS/MS
spectra and their precomputed matches against every public MS/MS spectrum in
MassIVE, MetaboLights, the Metabolomics Workbench and NORMAN. For each target we take
all library spectra with its InChIKey first block and keep their matches with a
cosine of at least 0.7 over at least 3 peaks.

The database is a fixed release, so the search is reproducible. The chunk below
runs only when the database is available and the cached
`StructureMASST_output/fasstrecords_hits.csv` is absent, or was queried for
other targets than those in `compound.csv`; otherwise the cached file is read.
The search writes the InChIKeys it queried next to it, in
`fasstrecords_targets.csv`. The roles are taken from `compound.csv`, so renaming
a target needs no new query. Delete the cached file to rebuild it. The search
queries only the deposits that carry every target, so it also writes each
target's deposits with a hit over the whole archive, in
`fasstrecords_deposits.csv`, for the step summary.

```{r discovery-fasstrecords}
FASST_DB   <- Sys.getenv("FASST_DB", "E:/FASSTrecords/db/masst_records.sqlite")
FASST_OUT  <- "StructureMASST_output/fasstrecords_hits.csv"
FASST_KEYS <- "StructureMASST_output/fasstrecords_targets.csv"  # InChIKeys it was queried for
FASST_DEPS <- "StructureMASST_output/fasstrecords_deposits.csv" # each target's deposits with a hit
same_targets <- function(keys) setequal(keys, compound_spec$inchikey_block1)
rebuild <- !file.exists(FASST_OUT) ||
  (file.exists(FASST_KEYS) && !same_targets(read.csv(FASST_KEYS)$inchikey_first_block))
IKEY <- stats::setNames(compound_spec$inchikey_block1, compound_spec$role)
COSINE_MIN <- 0.7; PEAKS_MIN <- 3
INVALID_SCAN <- 4294967295                             # "no scan" sentinel
unq <- function(x) trimws(gsub('"', '', x))
inl <- function(v) paste(v, collapse = ",")
# Each target's library spectra.
library_ids <- function(con) lapply(IKEY, function(k) dbGetQuery(con, sprintf(
  "SELECT spectrum_id_int, spectrum_id, Adduct, Precursor_MZ
     FROM library_table WHERE InChIKey_smiles_firstBlock = '%s'", k)))
# Each target's deposits with a hit over the whole archive (step summary),
# filtered as the hits below: plausible ions only, no "no scan" sentinel.
deposits_per_target <- function(con, lib_ids) {
  dplyr::bind_rows(lapply(compound_spec$role, function(role) {
    lib <- lib_ids[[role]]
    lib <- lib[!is.na(mapply(infer_adduct, lib$Precursor_MZ, role_formula[[role]],
                             USE.NAMES = FALSE)), ]
    if (!nrow(lib)) return(NULL)
    dep <- unique(unq(dbGetQuery(con, sprintf(
      "SELECT DISTINCT d.Dataset AS deposit
         FROM masst_table m JOIN dataset_table d ON m.dataset_id_int = d.dataset_id_int
        WHERE m.spectrum_id_int IN (%s) AND m.cosine >= %f AND m.matching_peaks >= %d
          AND (m.scan_id IS NULL OR m.scan_id != %.0f)",
      inl(lib$spectrum_id_int), COSINE_MIN, PEAKS_MIN, INVALID_SCAN))$deposit))
    if (!length(dep)) return(NULL)
    data.frame(inchikey_first_block = IKEY[[role]], deposit = dep)
  }))
}
if (file.exists(FASST_DB) && rebuild) {
  suppressPackageStartupMessages(library(RSQLite))
  con <- dbConnect(SQLite(), FASST_DB)
  lib_ids <- library_ids(con)
  # Query only the deposits carrying every target: the others are dropped below
  # anyway and would make the extract much larger.
  scope <- Reduce(intersect, lapply(lib_ids, function(lib) {
    if (!nrow(lib)) return(character(0))
    dbGetQuery(con, sprintf(
      "SELECT DISTINCT d.Dataset AS dataset
         FROM masst_table m JOIN dataset_table d ON m.dataset_id_int = d.dataset_id_int
        WHERE m.spectrum_id_int IN (%s) AND m.cosine >= %f AND m.matching_peaks >= %d",
      inl(lib$spectrum_id_int), COSINE_MIN, PEAKS_MIN))$dataset
  }))
  if (!length(scope)) {
    dbDisconnect(con)
    stop("No deposit carries every target (library spectra per target: ",
         paste(names(lib_ids), sapply(lib_ids, nrow), collapse = ", "), ")")
  }
  deposits <- deposits_per_target(con, lib_ids)
  get_role <- function(role) {
    lib <- lib_ids[[role]]; if (!nrow(lib)) return(NULL)
    h <- dbGetQuery(con, sprintf(
      "SELECT m.cosine AS Cosine, m.matching_peaks AS matching_peaks, m.scan_id AS scan_id,
              m.spectrum_id_int AS spectrum_id_int, m.mri_id_int AS mri_id_int,
              d.Dataset AS ATTRIBUTE_DatasetAccession, mr.mri AS mri
         FROM masst_table m
         JOIN dataset_table d ON m.dataset_id_int = d.dataset_id_int
         JOIN mri_table    mr ON m.mri_id_int     = mr.mri_id_int
        WHERE m.spectrum_id_int IN (%s) AND m.cosine >= %f AND m.matching_peaks >= %d%s",
      inl(lib$spectrum_id_int), COSINE_MIN, PEAKS_MIN,
      sprintf(" AND d.Dataset IN ('%s')", paste(scope, collapse = "','"))))
    if (!nrow(h)) return(NULL)
    h <- dplyr::left_join(h, lib[, c("spectrum_id_int","spectrum_id","Adduct","Precursor_MZ")],
                          by = "spectrum_id_int")
    h$ATTRIBUTE_DatasetAccession <- unq(h$ATTRIBUTE_DatasetAccession); h$mri <- unq(h$mri)
    h$role <- role; h$inchikey_first_block <- IKEY[[role]]
    h$USI <- paste0(h$mri, ":scan:", h$scan_id)
    h$file_path <- sub("^mzspec:[^:]*:", "", h$mri)
    h$`Matching Peaks` <- h$matching_peaks; h$lib_spectrum <- h$spectrum_id
    h }
  hits <- dplyr::bind_rows(lapply(compound_spec$role, get_role))
  hits <- hits[hits$scan_id != INVALID_SCAN | is.na(hits$scan_id), ]
  mids <- unique(hits$mri_id_int); chunks <- split(mids, ceiling(seq_along(mids) / 5000))
  redu <- dplyr::bind_rows(lapply(chunks, function(ch) dbGetQuery(con, sprintf(
    "SELECT mri_id_int, ChromatographyAndPhase, IonizationSourceAndPolarity, NCBITaxonomy, SampleType
       FROM redu_table WHERE mri_id_int IN (%s)", inl(ch)))))
  cln <- function(v) { v[is.na(v) | v %in% c("missing value","not applicable","")] <- NA; v }
  for (cc in c("ChromatographyAndPhase","IonizationSourceAndPolarity","NCBITaxonomy","SampleType"))
    redu[[cc]] <- cln(redu[[cc]])
  redu <- redu[!duplicated(redu$mri_id_int), ]
  hits <- dplyr::left_join(hits, redu, by = "mri_id_int")
  dbDisconnect(con)
  dir.create("StructureMASST_output", showWarnings = FALSE)
  keep <- c("ATTRIBUTE_DatasetAccession","USI","file_path","scan_id","role","Cosine","Matching Peaks",
            "Adduct","Precursor_MZ","inchikey_first_block","lib_spectrum","ChromatographyAndPhase",
            "IonizationSourceAndPolarity","NCBITaxonomy","SampleType")
  write.csv(hits[, keep], FASST_OUT, row.names = FALSE)
  write.csv(data.frame(inchikey_first_block = compound_spec$inchikey_block1),
            FASST_KEYS, row.names = FALSE)
  write.csv(deposits, FASST_DEPS, row.names = FALSE)
}
```

An extract cached before the search wrote each target's deposits gets them here,
with the same queries, without being rebuilt.

```{r deposits-per-target}
if (file.exists(FASST_DB) && !file.exists(FASST_DEPS)) {
  suppressPackageStartupMessages(library(RSQLite))
  con <- dbConnect(SQLite(), FASST_DB)
  write.csv(deposits_per_target(con, library_ids(con)), FASST_DEPS, row.names = FALSE)
  dbDisconnect(con)
}
```

```{r load-hits}
all_hits <- read.csv(FASST_OUT, check.names = FALSE, stringsAsFactors = FALSE)
for (j in seq_along(all_hits))
  if (is.character(all_hits[[j]])) all_hits[[j]][is_missing_val(all_hits[[j]])] <- NA
all_hits$scan_id <- as.integer(all_hits$scan_id)
# The extract must be for the targets in compound.csv. One written before its list
# of targets was kept is checked on the InChIKeys of its hits instead.
keys <- if (file.exists(FASST_KEYS)) read.csv(FASST_KEYS)$inchikey_first_block else
  all_hits$inchikey_first_block
if (!same_targets(keys))
  stop(FASST_OUT, " was queried for other targets than compound.csv lists; delete it ",
       "and render again with the FASSTrecords database (FASST_DB) available")
# Each hit takes its role from compound.csv, matched on the InChIKey.
all_hits$role <- compound_spec$role[match(all_hits$inchikey_first_block,
                                          compound_spec$inchikey_block1)]
n_hits <- nrow(all_hits)
cat("Total MASST hits loaded:", n_hits, "\n")
print(table(factor(all_hits$role, compound_spec$role)))
```

# Library spectra of each target

MASST reports, for each hit, the adduct and precursor m/z of the library
spectrum it matched, not of the deposited spectrum. A library spectrum is kept
only if its precursor m/z is a plausible ion of the target: the closest adduct
among `r paste(ADDUCT_CANDIDATES, collapse = ", ")` within 25 ppm. That adduct,
taken from the m/z rather than from the library label, is the one used from
here on. A library spectrum that fits no adduct is set aside with all its hits,
for example an isotope-labelled standard, which shares the target's InChIKey
first block but not its mass.

```{r library-spectra}
libs <- unique(all_hits[, c("role", "lib_spectrum", "Adduct", "Precursor_MZ")])
libs$adduct_inferred <- mapply(infer_adduct, libs$Precursor_MZ,
                               role_formula[libs$role], USE.NAMES = FALSE)
libs$hits <- as.vector(table(all_hits$lib_spectrum)[libs$lib_spectrum])

# The library writes "[M+H]1+" where we write "[M+H]+".
label <- sub("\\]1([+-])$", "]\\1", libs$Adduct)
label <- sub("[M-H2O+H]", "[M+H-H2O]", label, fixed = TRUE)
label <- sub("[M+NH3+H]", "[M+NH4]", label, fixed = TRUE)
libs$status <- ifelse(is.na(libs$adduct_inferred), "set aside",
               ifelse(label == libs$adduct_inferred, "label agrees",
                      "label corrected"))

all_hits$adduct_inferred <- libs$adduct_inferred[
  match(all_hits$lib_spectrum, libs$lib_spectrum)]
all_hits <- all_hits[!is.na(all_hits$adduct_inferred), ]
```

```{r library-spectra-summary}
#| echo: false
print(table(role = factor(libs$role, compound_spec$role), status = libs$status))
cat("\nHits set aside:", sum(libs$hits[libs$status == "set aside"]),
    "| hits with a corrected label:",
    sum(libs$hits[libs$status == "label corrected"]), "\n")
libs[libs$status != "label agrees", ]
```

`r sum(libs$status == "label corrected")` GNPS library spectra of tryptophan,
`r paste(libs$lib_spectrum[libs$status == "label corrected"], collapse = " and ")`,
are labelled [M+H]+, but their precursor m/z,
`r paste(unique(sprintf("%.3f", libs$Precursor_MZ[libs$status == "label corrected"])), collapse = " and ")`,
is that of
`r paste(unique(libs$adduct_inferred[libs$status == "label corrected"]), collapse = " and ")`.
Their matches lie in
`r length(unique(all_hits$ATTRIBUTE_DatasetAccession[all_hits$lib_spectrum %in% libs$lib_spectrum[libs$status == "label corrected"]]))`
deposits, and they now carry that ion. The `r sum(libs$status == "set aside")` library spectra set aside
(listed above) are a tryptophan spectrum whose precursor m/z is too far from [M+H]+,
which carries most of their hits, spectra of the dimers, spectra of the labelled
standard tryptophan-d5, spectra recorded at nominal mass, and one spectrum at an
unrelated mass.

# Deposits carrying every target

A deposit that carries only some of the targets cannot be used, so it is
dropped here.

```{r d-every}
D_by_role <- lapply(stats::setNames(compound_spec$role, compound_spec$role), function(r)
  unique(all_hits$ATTRIBUTE_DatasetAccession[all_hits$role == r]))
D_every   <- Reduce(intersect, D_by_role)
if (!length(D_every))
  stop("No deposit carries every target once the library spectra are checked ",
       "(deposits per target: ",
       paste(names(D_by_role), lengths(D_by_role), collapse = ", "), ")")
all_hits  <- all_hits[all_hits$ATTRIBUTE_DatasetAccession %in% D_every, ]
cat(paste(names(D_by_role), "deposits:", lengths(D_by_role), collapse = " | "),
    "| intersection:", length(D_every), "\n")
```

```{r write-hits}
write.csv(all_hits, "artifacts/hits_search.csv", row.names = FALSE)
```

# Step summary

The search returned `r format(n_hits, big.mark = ",")` hits, matching
`r knitr::combine_words(paste(table(factor(libs$role, compound_spec$role)), c("library spectra of the", rep("of the", nrow(compound_spec) - 1)), compound_spec$role))`.
Library spectra set aside: `r sum(libs$status == "set aside")`, with
`r format(sum(libs$hits[libs$status == "set aside"]), big.mark = ",")` hits. Library
spectra whose label was corrected: `r sum(libs$status == "label corrected")`, with
`r format(sum(libs$hits[libs$status == "label corrected"]), big.mark = ",")` hits.

The chapter's table counts deposits only, one unit per table.

```{r step-summary}
# Deposits with a hit, of any target: over the whole archive from FASST_DEPS,
# else only those the cached extract was queried for.
n_with_hit <- if (file.exists(FASST_DEPS))
  length(unique(read.csv(FASST_DEPS)$deposit)) else length(Reduce(union, D_by_role))
summary_step <- data.frame(
  item  = c("deposits with a hit",
            paste("deposits with", targets_phrase(nrow(compound_spec)))),
  value = c(n_with_hit,
            length(D_every)))
# For the discussion's recap only (collect-tables.R keeps "recap:" items out of the
# step table): the files with a hit, all in the deposits that carry every target.
summary_step <- rbind(summary_step, data.frame(
  item  = "recap: files",
  value = length(unique(paste(all_hits$ATTRIBUTE_DatasetAccession,
                              basename(all_hits$file_path))))))
write.csv(summary_step, "artifacts/summary_step1.csv", row.names = FALSE)
summary_step
```

# Session info

```{r session-info}
sessionInfo()
```
