---
title: "02 - Metadata"
subtitle: "Step 2: sample descriptions retrieved and harmonised"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
---

# Overview

Each deposit that step 1 kept is described from its repository's own metadata,
and the descriptions are mapped onto one per-sample schema. Each hit is also
assigned to the assay it belongs to: from here on the unit is the *assay* (one
analysis of a deposit; a MassIVE deposit counts as one), and later steps join on
its identifier, `dataset_id`.

**Input:** `artifacts/hits_search.csv`, plus per-deposit rules written in this
notebook where the automatic route falls short, marked as manual work.

**Output:** `artifacts/hits_assigned.csv`, `artifacts/metadata_entities.csv`,
`artifacts/assay_metadata.csv`, `artifacts/metadata_inventory.csv` and
`artifacts/related_deposits.csv`, plus `artifacts/removed_step2.csv` (the deposits
removed in this step).

```{r setup}
source("../../R/setup.R")
source("../../R/masst.R")
source("../../R/metadata.R")
```

```{r load-hits}
all_hits <- read.csv("artifacts/hits_search.csv", check.names = FALSE,
                     stringsAsFactors = FALSE)
D_both <- unique(all_hits$ATTRIBUTE_DatasetAccession)
D_entering <- D_both
cat("Deposits:", length(D_both), "\n")
```

# What is already known

Before anything is fetched, the deposits that the search index (ReDU) already
describes are checked. ReDU often describes only part of a deposit, so a deposit is
judged here only when ReDU describes every file that carries a hit. The check then
reads every file ReDU describes in the deposit, blanks and QCs set aside. The
deposit is removed when none of these files is human and at least one is of
another species (a file with no species recorded is ignored), or when every one of
them is from a body site other than blood. A deposit in a repository the
`MsBackend` packages cannot read is removed too. Everything else, including the
deposits ReDU describes only in part or not at all, goes on to the other sources.

::: {.callout-warning title="Manual work"}
`known_by_hand` removes a deposit whose species or body site is stated on its
study page but in no source read here (ST003302: rat brain dialysate).
:::

```{r known-wrong}
# Known by hand: the study page says what ReDU does not.
known_by_hand <- tibble::tribble(
  ~dataset_id, ~reason,     ~notes,
  "ST003302",  "not_human", "rat brain dialysate (Rattus norvegicus), from the study page")
# Species and body site (blood) are checked: the question is on adult blood.
pass1 <- redu_first_pass(
  all_hits,
  fasst_db  = Sys.getenv("FASST_DB", "E:/FASSTrecords/db/masst_records.sqlite"),
  redu_out  = "StructureMASST_output/redu_files.csv",
  body_site = "blood|plasma|serum",
  by_hand   = known_by_hand)
removed_step2 <- pass1$removed
D_both        <- pass1$deposits
all_hits      <- pass1$hits
write.csv(removed_step2, "artifacts/removed_step2.csv", row.names = FALSE)
cat("ReDU describes every hit file of", pass1$n_described,
    "deposits | removed:", nrow(removed_step2), "| deposits left:", length(D_both), "\n")
# Per reason, the deposits ReDU judged and those removed by hand (known_by_hand).
table(reason  = removed_step2$reason,
      by_hand = removed_step2$dataset_id %in% known_by_hand$dataset_id)
```

# Sources

The repositories share no metadata schema. MassIVE deposits are described by the
submitter's own sample table, when there is one, and by Pan-ReDU; MetaboLights
deposits by their ISA-Tab files and Metabolomics Workbench deposits by their
mwTab files, both fetched with functions of `MsBackendMetaboLights` and
`MsBackendMetabolomicsWorkbench`. Most fetched files are cached in
`dataset_metadata/`. Placeholders such as "missing value" are read as missing.

Each deposit left after the check above is looked up in every source that exists:
Pan-ReDU as it is today, the submitter's own table in the deposit (MassIVE), the
ISA-Tab files (MetaboLights) and the mwTab files (Workbench). A MassIVE submitter
table is any table file of the deposit, in any folder, with a column that names the
deposit's own MS files; only the rows of those files are kept.

```{r submitter-tables}
msv   <- D_both[startsWith(D_both, "MSV")]
found <- vapply(msv, function(d) !is.null(fetch_massive_submitter_table(d)), logical(1))
cat("MassIVE deposits with a submitter table:", sum(found), "of", length(msv), "\n")
```

# Automatic harvest

For every deposit, technical metadata (chromatography, ionisation, instrument,
per assay) and biological metadata (which columns hold the subject, sex, age and
so on) are read from the sources above. `harvest_metadata_coverage()` maps each
deposit's own column names onto the shared fields.

```{r auto-assess}
#| code-fold: true
tech_raw <- fetch_assay_metadata(D_both)
coverage <- tibble::as_tibble(
  dplyr::bind_rows(lapply(D_both, harvest_metadata_coverage))
)
write.csv(tech_raw, "artifacts/assay_metadata.csv", row.names = FALSE)
```

```{r auto-assess-display}
#| echo: false
coverage[, c("dataset_id", "n_samples", "n_canonical_mapped",
             grep("^col_", names(coverage), value = TRUE))]
```

# Assays

MetaboLights and Workbench deposits often hold several assays of the same
samples under one accession. Each hit is assigned to its assay from the file it
came from; MassIVE has no assay concept, so its deposits stay whole. A hit whose
file matches no known assay is marked `_unmatched`.

::: {.callout-warning title="Manual work"}
The Workbench lists each sample's files without saying which analysis each one
belongs to, so the first name is taken to belong to the first analysis, and so
on. Where a deposit lists them in another order, the order is declared by hand in
`MWB_FILE_ORDER` (`R/metadata.R`), checked against the deposit's folders and the
retention times in its own results tables. File names found in two folders of a
deposit cannot be told apart and are left out.
:::

```{r mwb-file-order}
MWB_FILE_ORDER[intersect(names(MWB_FILE_ORDER), D_both)]
```

A deposit whose list of assays cannot be read cannot have its hits assigned. It is
removed here, with the error as its note. A repository that does not answer
stops the render instead, so that no deposit is removed for it.

```{r assay-assignment}
mtbls_st <- D_both[grepl("^(MTBLS|ST)", D_both)]
unread   <- character()
assay_lookups <- setNames(lapply(mtbls_st, function(d)
  tryCatch(assay_lookup_for(d), error = function(e) {
    if (is_network_error(e)) stop(e)
    unread[[d]] <<- sub("\n.*", "", conditionMessage(e))
    NULL
  })), mtbls_st)
if (length(unread)) {
  removed_step2 <- dplyr::bind_rows(removed_step2, tibble::tibble(
    dataset_id = names(unread), reason = "metadata_unreadable", notes = unname(unread)))
  write.csv(removed_step2, "artifacts/removed_step2.csv", row.names = FALSE)
  D_both   <- setdiff(D_both, names(unread))
  all_hits <- all_hits[!all_hits$ATTRIBUTE_DatasetAccession %in% names(unread), ]
}
unread

all_hits$assay <- mapply(function(ds, fp) {
  if (startsWith(ds, "MSV")) return(NA_character_)
  file_to_assay(basename(fp), assay_lookups[[ds]])
}, all_hits$ATTRIBUTE_DatasetAccession, all_hits$file_path,
USE.NAMES = FALSE)

all_hits$deposit_id <- all_hits$ATTRIBUTE_DatasetAccession
all_hits$dataset_id <- make_dataset_id(all_hits$deposit_id, all_hits$assay)

# MASST hits whose file basename doesn't resolve to any known assay
mtbls_st_hits <- all_hits$deposit_id %in% mtbls_st
unresolved    <- mtbls_st_hits & is.na(all_hits$assay)
all_hits$dataset_id[unresolved] <- paste0(all_hits$deposit_id[unresolved],
                                          "_unmatched")
```

```{r write-hits}
write.csv(all_hits, "artifacts/hits_assigned.csv", row.names = FALSE)
```

# Manual: deposit-level facts and column names

::: {.callout-warning title="Manual work"}
Values that no per-file column carries, taken from the publication or the deposit
description (`constants`), and column names that the automatic mapping misses
(`custom_col_map`).
One fact comes from file names: MSV000094097 has no subject column in any of
its metadata, and its file names are the only place its submitters say which
person a file comes from (`OAD_001_Basal_P`, `OAD_001_Day15_S`), so the person is
read from the start of the file name (`subject_from_file`). Without it, each
person's plasma and serum at two time points would count as four people.
:::

```{r deposit-constants}
#| code-fold: true
# Deposit-level facts missing from the per-file columns. No body site here:
# each deposit records it per file, and a constant would overwrite that.
constants <- list(
  # One subject id per blood draw (15NAVY01.1.plasma.met, 15NAVY01.2.plasma.met):
  # the draw number is removed so that one person is one subject.
  MSV000082630 = list(location_type = "study_site",
                      subject_suffix = "[._][0-9]+[._]plasma[.]met$"),
  # The same plasma samples, measured again on another instrument: the same ids.
  MSV000086158 = list(location_type = "study_site",
                      subject_suffix = "[._][0-9]+[._]plasma[.]met$"),
  MSV000085607 = list(location_type = "study_site"),
  # One serum panel on two columns, both modes. No chromatography is recorded,
  # and HILIC vs reverse phase sets which peak gate 05 applies.
  MSV000092754 = list(location_type = "study_site", chromatography = "HILIC"),
  MSV000092756 = list(location_type = "study_site", chromatography = "HILIC"),
  MSV000092757 = list(location_type = "study_site", chromatography = "RP"),
  MSV000092758 = list(location_type = "study_site", chromatography = "RP"),
  # No subject column anywhere; the file name starts with the person
  # (OAD_001_Basal_P, OAD_001_Day15_S).
  MSV000094097 = list(location_type = "study_site",
                      subject_from_file = "^([A-Z]*OAD_[0-9]+)_.*$"),
  MSV000085256 = list(location_type = "study_site"),
  # The subject id carries the visit (FD001_T1, FD001_T2, FD001_T3): the suffix is
  # removed so that one person is one subject.
  MSV000084556 = list(location_type = "study_site", subject_suffix = "_T[0-9]+$")
)

# Columns the automatic mapping misses (Chagas panel, MSV000096564). Every other
# deposit is left to the heuristic, so a bad mapping stays visible.
chagas_cols <- list(subject_id = "patient_id1", sex = "ATTRIBUTE_sex",
                    age = "age", health_status = "ATTRIBUTE_status")
# MSV000096564: Pan-ReDU records no subject and no disease; the submitter table has
# host_subject_id (same id in plasma and serum), disease_state, sex, and host_age in years.
ibd_cols <- list(subject_id = "host_subject_id", health_status = "disease_state",
                 sex = "sex")
custom_col_map <- list(MSV000092754 = chagas_cols, MSV000092756 = chagas_cols,
                       MSV000092757 = chagas_cols, MSV000092758 = chagas_cols,
                       MSV000096564 = ibd_cols)
```

# One table per assay

One row per assay, with the repository, the chromatography, the column
mappings (`col_`) and the deposit-level facts (`const_`). Step 3 keeps the rows
that survive curation.

```{r build-metadata-table}
#| code-fold: true
entities <- unique(all_hits[!grepl("_unmatched$", all_hits$dataset_id),
                            c("dataset_id", "deposit_id", "assay")])
entities <- entities[order(entities$dataset_id, method = "radix"), ]

tech_renamed <- tech_raw |>
  dplyr::rename(deposit_id     = dataset_id,
                chromatography = chrom_type) |>
  dplyr::select(deposit_id, assay, repository, chromatography)

is_msv <- startsWith(entities$deposit_id, "MSV")
msv_meta <- entities[is_msv, ] |>
  dplyr::left_join(tech_renamed |> dplyr::select(-assay),
                   by = "deposit_id")
# A MassIVE deposit can report a real chromatography for some runs and "Unknown"
# for others, which gives it two rows here. Keep the real one.
unknown  <- is.na(msv_meta$chromatography) | msv_meta$chromatography %in% c("Unknown", "")
msv_meta <- msv_meta[order(msv_meta$dataset_id, unknown, method = "radix"), ]
msv_meta <- msv_meta[!duplicated(msv_meta$dataset_id), ]
nonmsv_meta <- entities[!is_msv, ] |>
  dplyr::left_join(tech_renamed, by = c("deposit_id", "assay"))

dataset_metadata <- dplyr::bind_rows(msv_meta, nonmsv_meta) |>
  dplyr::left_join(coverage |> dplyr::rename(deposit_id = dataset_id),
                   by = "deposit_id")

const_fields <- unique(unlist(lapply(constants, names)))
for (canon in const_fields) {
  col <- paste0("const_", canon)
  dataset_metadata[[col]] <- vapply(
    dataset_metadata$deposit_id,
    function(d) {
      v <- constants[[d]][[canon]]
      if (is.null(v)) NA_character_ else as.character(v)
    },
    character(1))
}

if ("const_chromatography" %in% names(dataset_metadata)) {
  use_const <- !is.na(dataset_metadata$const_chromatography) &
               (is.na(dataset_metadata$chromatography) |
                dataset_metadata$chromatography == "" |
                dataset_metadata$chromatography == "Unknown")
  dataset_metadata$chromatography[use_const] <-
    dataset_metadata$const_chromatography[use_const]
}

for (ds in names(custom_col_map)) {
  ix <- which(dataset_metadata$deposit_id == ds)
  if (!length(ix)) next
  for (canon in names(custom_col_map[[ds]])) {
    col_key <- paste0("col_", canon)
    if (col_key %in% names(dataset_metadata)) {
      dataset_metadata[[col_key]][ix] <- custom_col_map[[ds]][[canon]]
    }
  }
}

dataset_metadata <- dataset_metadata[
  , c("dataset_id", "deposit_id", "assay", "repository", "chromatography",
      "n_samples", "n_canonical_mapped",
      grep("^col_",   names(dataset_metadata), value = TRUE),
      grep("^const_", names(dataset_metadata), value = TRUE))]
write.csv(dataset_metadata, "artifacts/metadata_entities.csv", row.names = FALSE)
dataset_metadata
```

# What the metadata says

The question is on the blood of adults, so this section counts, per assay, what the
question needs: blood samples, samples with an age, and blood samples from adults.

```{r adult-age}
ADULT_MIN_AGE <- 18   # the same threshold as step 6
```

For each assay, once every source has been read: how many samples its metadata
describes, their species, how many are blood, how many have an age, and how many
are blood samples from adults (`r ADULT_MIN_AGE` or older). An assay whose
metadata could not be read is marked, so that "unreadable" is not read as
"nothing found" (a repository that does not answer stops the render instead).
Step 3 decides from this table.

```{r metadata-inventory}
# Ages as years, from free text such as "45", "45 years", "3 months" or "10 days".
age_in_years <- function(x) {
  x <- tolower(as.character(x))
  n <- suppressWarnings(as.numeric(sub("^[^0-9]*([0-9]+([.][0-9]+)?).*$", "\\1", x)))
  n[!grepl("[0-9]", x)] <- NA
  n / ifelse(grepl("month", x), 12, ifelse(grepl("week", x), 52,
                                          ifelse(grepl("day", x), 365, 1)))
}

inventory <- do.call(rbind, lapply(seq_len(nrow(dataset_metadata)), function(i) {
  spec <- as.list(dataset_metadata[i, ])
  s <- tryCatch(native_sample_table(spec$deposit_id), error = function(e) {
    if (is_network_error(e)) stop(e)
    NULL
  })
  if (is.null(s) || !nrow(s))
    return(data.frame(dataset_id = spec$dataset_id, samples = 0L, human = 0L,
                      other_species = 0L, blood = 0L, with_age = 0L, adult_blood = 0L,
                      adult_blood_people = NA_integer_))
  x  <- apply_spec_to_native(s, spec)
  sp <- if (!is.na(spec$col_species)) as.character(s[[spec$col_species]]) else rep(NA, nrow(s))
  human <- grepl("9606|homo sapiens|human", sp, ignore.case = TRUE)
  blood <- grepl("blood|plasma|serum", x$body_site, ignore.case = TRUE)
  age   <- age_in_years(x$age)
  adult <- blood & !is.na(age) & age >= ADULT_MIN_AGE
  # People behind those samples, where the metadata names them (NA where it does not).
  people <- unique(stats::na.omit(x$subject_id[adult]))
  data.frame(dataset_id = spec$dataset_id, samples = nrow(s), human = sum(human),
             other_species = sum(!is_missing_val(sp) & !human),
             blood = sum(blood), with_age = sum(!is.na(age)),
             adult_blood = sum(adult),
             adult_blood_people = if (length(people)) length(people) else NA_integer_)
}))
inventory$reachable <- !is.na(dataset_metadata$n_samples)
write.csv(inventory, "artifacts/metadata_inventory.csv", row.names = FALSE)
inventory
```

# Manual: related deposits

::: {.callout-warning title="Manual work"}
An accession is not always a study, and the analysis has to know when two are
related. **MSV000092754, MSV000092756, MSV000092757 and MSV000092758** are one
Chagas serum panel, measured on HILIC and on reverse phase, each in positive and
negative mode. They share their file names and subjects, but the same file name
refers to a different measurement of one serum sample in each deposit. They count
as one study for pooling and must never be merged on file name. **MSV000086158**
measured the plasma samples of MSV000082630 again, on a maXis instead of a
Q Exactive (its table names each file's Q Exactive counterpart). (MSV000085724
re-deposits MSV000085607 file for file; step 3 removes it.)
:::

```{r related-deposits}
related_deposits <- tibble::tribble(
  ~dataset_id,    ~related_to,     ~relation,                  ~notes,
  "MSV000092756", "MSV000092754",  "same_samples_two_assays",
  "Chagas serum panel: HILIC negative (092756) and HILIC positive (092754).",
  "MSV000092757", "MSV000092754",  "same_samples_two_assays",
  paste("Chagas serum panel: reverse phase positive (092757) and HILIC positive",
        "(092754). The file names are shared but the files are different",
        "measurements, not copies. One study for pooling; never merge on file name."),
  "MSV000092758", "MSV000092754",  "same_samples_two_assays",
  "Chagas serum panel: reverse phase negative (092758) and HILIC positive (092754).",
  "MSV000086158", "MSV000082630",  "same_samples_two_assays",
  paste("The ONR plasma samples of 082630 (Q Exactive) measured again on a maXis;",
        "the table of 086158 gives each file's Q Exactive counterpart (QE_filename).")
)
write.csv(related_deposits, "artifacts/related_deposits.csv", row.names = FALSE)
related_deposits
```

# Step summary

The results of this step, as the chapter's table shows them.

```{r step-summary}
summary_step <- data.frame(
  item  = c("deposits entering", "deposits that may answer the research question",
            "... sample table found by hand", "... information completed by hand",
            "assays with a hit"),
  value = c(length(D_entering),
            length(D_both),
            "--",
            length(intersect(D_both, union(names(constants), names(custom_col_map)))),
            nrow(dataset_metadata)))
# For the discussion's recap only: the files with a hit in those deposits.
summary_step <- rbind(summary_step, data.frame(
  item  = "recap: files",
  value = length(unique(paste(all_hits$ATTRIBUTE_DatasetAccession,
                              basename(all_hits$file_path))))))
write.csv(summary_step, "artifacts/summary_step2.csv", row.names = FALSE)
summary_step
```

# Session info

```{r session-info}
sessionInfo()
```
