---
title: "03 - Curation"
subtitle: "Step 3: deposits dropped only for a stated reason"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
---

# Overview

Deposits that cannot answer the question are removed, each with a recorded
reason, and only the assays that measure every target are kept.

**Input:** `artifacts/hits_assigned.csv`, `artifacts/metadata_entities.csv`,
`artifacts/metadata_inventory.csv`, `artifacts/removed_step2.csv` and
`artifacts/assay_metadata.csv` (all from step 2), plus the list of excluded deposits written in this notebook (manual work).

**Output:** `artifacts/hits_curated.csv` and `artifacts/dataset_metadata_table.csv`,
read by step 4 and every later step, and
`artifacts/exclusions.csv`, the reason for every deposit removed. The figure
`figures/metadata_inventory_kept.png` (and `.pdf`) shows what the kept assays
describe.

```{r setup}
source("../../R/setup.R")
source("../../R/masst.R")
source("../../R/metadata.R")
```

```{r load}
all_hits <- read.csv("artifacts/hits_assigned.csv", check.names = FALSE,
                     stringsAsFactors = FALSE)
metadata <- read.csv("artifacts/metadata_entities.csv", check.names = FALSE,
                     stringsAsFactors = FALSE)
inventory     <- read.csv("artifacts/metadata_inventory.csv", stringsAsFactors = FALSE)
# As text: a file with no row (step 2 removed nothing) would read as logical columns.
removed_step2 <- read.csv("artifacts/removed_step2.csv", colClasses = "character")
assay_meta    <- read.csv("artifacts/assay_metadata.csv", stringsAsFactors = FALSE)
D_both <- unique(all_hits$deposit_id)
D_entering <- D_both
```

# Manual: excluded deposits

::: {.callout-warning title="Manual work"}
Each deposit below was read and excluded by hand, with the reason recorded.

| reason | what it means |
|---|---|
| `duplicate_deposit` | the same files deposited again under another accession |
:::

```{r other-exclusions}
other_exclusions <- dplyr::bind_rows(removed_step2, tibble::tribble(
  ~dataset_id,    ~reason,             ~notes,
  "MSV000085724", "duplicate_deposit", "re-deposits MSV000085607 file for file (.mzML instead of .mzXML)"
))
by_hand <- intersect(D_both, other_exclusions$dataset_id)
D_both  <- setdiff(D_both, other_exclusions$dataset_id)
```

# What the metadata shows

A deposit is kept when its metadata (step 2) shows blood samples from adults with
an age (step 2 sets the adult age), and no species other than human. Otherwise it
is removed with the reason: another species, no blood found, no adult with an age,
no metadata found, or metadata that could not be read when step 2 ran (render
step 2 again). A deposit that records no subject is kept; the analysis then
counts its files.

```{r fit-rule}
inv <- merge(inventory, metadata[, c("dataset_id", "deposit_id")])
dep <- aggregate(cbind(samples, human, other_species, blood, with_age, adult_blood,
                       reachable) ~ deposit_id, data = inv, FUN = max)
dep <- dep[dep$deposit_id %in% D_both, ]
dep$reason <- with(dep, ifelse(!reachable, "metadata_unreachable",
                ifelse(samples == 0, "no_metadata",
                ifelse(other_species > 0 & human == 0, "not_human",
                ifelse(blood == 0, "no_blood_found",
                ifelse(adult_blood == 0, "no_adult_with_age", NA))))))
other_exclusions <- dplyr::bind_rows(other_exclusions, tibble::tibble(
  dataset_id = dep$deposit_id[!is.na(dep$reason)],
  # As text: when no deposit is removed here, ifelse() gives logical NA.
  reason     = as.character(dep$reason[!is.na(dep$reason)]),
  notes      = "step 2 inventory"))
D_both <- setdiff(D_both, dep$deposit_id[!is.na(dep$reason)])
cat("Kept:", length(D_both), "deposits\n")
table(other_exclusions$reason)
```

# Every target in one assay

A deposit can carry every target and still never measure them in the same assay.
Only assays with a hit for every target are kept, and hits marked `_unmatched`
in step 2 are dropped. A deposit left without any assay is added to the
exclusions with the reason `no_assay_with_both_roles` (`no_assay_with_every_target`
with any other number of targets), or `no_hit_matched_to_assay` when step 2 could
match none of its hits to an assay.

```{r entity-coverage}
# One has_<role> column per target, in the order of compound.csv: does the assay
# carry a hit for it? The metabolite's keeps its short name, has_metab.
roles    <- compound_spec$role
has_cols <- sub("^has_metabolite$", "has_metab", paste0("has_", roles))
cover    <- all_hits[all_hits$deposit_id %in% D_both,
                     c("dataset_id", "deposit_id", "assay", "role")]
for (i in seq_along(roles)) cover[[has_cols[i]]] <- cover$role == roles[i]
entity_coverage <- cover |>
  dplyr::group_by(dataset_id, deposit_id, assay) |>
  dplyr::summarise(dplyr::across(dplyr::all_of(has_cols), any),
                   n_hits  = dplyr::n(),
                   .groups = "drop")
entity_coverage$keep <- Reduce(`&`, entity_coverage[has_cols]) &
  !grepl("_unmatched$", entity_coverage$dataset_id)

datasets_with_kept_entity <- unique(
  entity_coverage$deposit_id[entity_coverage$keep])
deposits_lost <- setdiff(D_both, datasets_with_kept_entity)
# A deposit whose hits are all `_unmatched` never had an assay to check.
unmatched_only <- deposits_lost[vapply(deposits_lost, function(d)
  all(grepl("_unmatched$", entity_coverage$dataset_id[entity_coverage$deposit_id == d])),
  logical(1))]

# The reason keeps its two-target name when there are two targets.
no_assay <- if (length(roles) == 2) "no_assay_with_both_roles" else
  "no_assay_with_every_target"
if (length(deposits_lost)) {
  unmatched <- deposits_lost %in% unmatched_only
  new_excl <- tibble::tibble(
    dataset_id = deposits_lost,
    reason     = ifelse(unmatched, "no_hit_matched_to_assay", no_assay),
    notes      = ifelse(unmatched, "no hit file matched to an assay in step 2",
                        paste(paste(roles, collapse = " + "),
                              "MASST hits land in different assays of the same deposit")))
  other_exclusions <- dplyr::bind_rows(other_exclusions, new_excl)
  D_both <- setdiff(D_both, deposits_lost)
}

entities_kept    <- entity_coverage$dataset_id[entity_coverage$keep]
all_hits         <- all_hits[all_hits$dataset_id %in% entities_kept, ]
dataset_metadata <- metadata[metadata$dataset_id %in% entities_kept, ]
```

```{r entity-coverage-summary}
#| echo: false
lost_assay <- setdiff(deposits_lost, unmatched_only)
if (length(lost_assay))
  cat("Deposits dropped (no assay carries ", if (length(roles) == 2) "both " else "all of ",
      paste(roles, collapse = " + "), "):\n  ",
      paste(lost_assay, collapse = "\n  "), "\n", sep = "")
if (length(unmatched_only))
  cat("Deposits dropped (no hit matched to an assay):\n  ",
      paste(unmatched_only, collapse = "\n  "), "\n", sep = "")
cat("Surviving entities:", length(entities_kept),
    " | surviving deposits:", length(datasets_with_kept_entity), "\n")
entity_coverage
```

# What the kept assays describe

The kept assays are described by what the question on the blood of adults needs.

For each assay kept, what its metadata (step 2) describes: how many samples,
how many of them are blood, how many have an age, and how many are blood samples
from adults.

```{r fig-kept-inventory, fig.width = 6.3, fig.height = 5}
FIG_DIR <- "figures"; dir.create(FIG_DIR, showWarnings = FALSE)
figp    <- function(f) file.path(FIG_DIR, f)

kept_inv <- merge(dataset_metadata[, c("dataset_id", "deposit_id", "assay", "chromatography")],
                  inventory)
# Assays of one deposit share its sample table: their labels add the column and the
# polarity, or keep the dataset id where that still does not tell them apart.
pol <- c(positive = "+", negative = "-")[
  assay_meta$polarity[match(paste(kept_inv$deposit_id, kept_inv$assay),
                            paste(assay_meta$dataset_id, assay_meta$assay))]]
kept_inv$label <- ifelse(is.na(kept_inv$assay), kept_inv$deposit_id,
                         paste0(kept_inv$deposit_id, " ", kept_inv$chromatography,
                                ifelse(is.na(pol), "", pol)))
same <- duplicated(kept_inv$label) | duplicated(kept_inv$label, fromLast = TRUE)
kept_inv$label[same] <- kept_inv$dataset_id[same]

bars <- c(samples = "samples described", blood = "blood", with_age = "with an age",
          adult_blood = "adult blood")
kept_long <- tidyr::pivot_longer(kept_inv[, c("label", names(bars))], -label,
                                 names_to = "measure", values_to = "n")
kept_long$measure <- factor(bars[kept_long$measure], levels = bars)
kept_long$label   <- factor(kept_long$label, levels = rev(unique(kept_inv$label)))

p_kept <- ggplot(kept_long, aes(n, label, fill = measure)) +
  geom_col(position = position_dodge(width = 0.8, reverse = TRUE), width = 0.75) +
  scale_fill_manual(values = c("grey82", "grey62", "grey42", "grey15"), name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "samples", y = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top", panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank())
ggsave(figp("metadata_inventory_kept.png"), p_kept, width = 6.3, height = 5, dpi = 200)
ggsave(figp("metadata_inventory_kept.pdf"), p_kept, width = 6.3, height = 5)
p_kept
```

# One set-up per MassIVE deposit

MetaboLights and the Metabolomics Workbench record the set-ups of a deposit as
separate assays; MassIVE does not, so a MassIVE deposit counts as one assay. Each
kept MassIVE deposit is checked here for a mix of set-ups, from what can be known
without the raw files: the chromatography and polarity its metadata records, and the
polarity of its hits (the charge of the adduct each hit matched). A deposit that
mixes set-ups is removed: it cannot be measured as one assay, and its files cannot
be told apart reliably. A file without a hit, in a deposit whose metadata records no
set-up, is not covered by this check.

```{r stray-hits}
# Hits dropped by hand (see the callout below), one row per deposit and adduct.
stray_hits <- tibble::tribble(
  ~deposit_id,    ~adduct_inferred, ~why,
  "MSV000086158", "[M-H]-",         "negative-ion matches in positive-mode stool files with no metadata row"
)
stray <- paste(all_hits$deposit_id, all_hits$adduct_inferred) %in%
  paste(stray_hits$deposit_id, stray_hits$adduct_inferred)
```

::: {.callout-warning title="Manual work"}
One deposit is corrected by hand before the check. MSV000086158 records its set-up
(reverse phase, positive mode) for its plasma files only. `r sum(stray)` of its hits match
kynurenine as a negative ion ([M-H]-, m/z
`r paste(unique(sprintf("%.3f", all_hits$Precursor_MZ[stray])), collapse = " and ")`)
in `r length(unique(all_hits$file_path[stray]))` stool files that have no
metadata row and whose other hits are all positive ions. A file acquired in positive
mode cannot hold a negative ion, so these are stray library matches rather than a
second set-up, and they are dropped. No rule is written for this: it is a reading of
one deposit.
:::

```{r one-setup}
message(sum(stray), " hit(s) dropped by hand as stray matches")
all_hits <- all_hits[!stray, ]

tech  <- read.csv("artifacts/assay_metadata.csv", stringsAsFactors = FALSE)
known <- function(x) sort(unique(x[!is.na(x) & !x %in% c("", "Unknown")]))
msv   <- sort(unique(all_hits$deposit_id[startsWith(all_hits$deposit_id, "MSV")]))
setups <- do.call(rbind, lapply(msv, function(d) {
  t   <- tech[tech$dataset_id == d, ]
  add <- all_hits$adduct_inferred[all_hits$deposit_id == d]
  pol <- ifelse(grepl("[+]$", add), "positive", ifelse(grepl("-$", add), "negative", NA))
  data.frame(deposit           = d,
             chromatography    = paste(known(t$chrom_type), collapse = " and "),
             polarity_recorded = paste(known(t$polarity), collapse = " and "),
             polarity_of_hits  = paste(known(pol), collapse = " and "))
}))
if (is.null(setups))
  setups <- data.frame(deposit = character(), chromatography = character(),
                       polarity_recorded = character(), polarity_of_hits = character())
setups$mixed <- grepl(" and ", paste(setups$chromatography, setups$polarity_recorded,
                                     setups$polarity_of_hits))
mixed <- setups$deposit[setups$mixed]
if (length(mixed)) {
  other_exclusions <- dplyr::bind_rows(other_exclusions, tibble::tibble(
    dataset_id = mixed, reason = "mixed_setups",
    notes      = "MassIVE deposit whose files mix set-ups; MassIVE records no assays"))
  D_both           <- setdiff(D_both, mixed)
  all_hits         <- all_hits[!all_hits$deposit_id %in% mixed, ]
  dataset_metadata <- dataset_metadata[!dataset_metadata$deposit_id %in% mixed, ]
}
setups
```

# Write curated hits

```{r write-artifacts}
write.csv(dataset_metadata, "artifacts/dataset_metadata_table.csv",
          row.names = FALSE)
write.csv(all_hits,         "artifacts/hits_curated.csv",
          row.names = FALSE)
write.csv(other_exclusions, "artifacts/exclusions.csv", row.names = FALSE)
```

```{r write-artifacts-summary}
#| echo: false
cat("artifacts/dataset_metadata_table.csv :",
    nrow(dataset_metadata), "rows\n")
cat("artifacts/hits_curated.csv           :",
    nrow(all_hits), "rows /",
    length(unique(all_hits$dataset_id)), "entities\n")
```

# Step summary

The results of this step, as the chapter's table shows them.

```{r step-summary}
# The removals of this step, by reason (step 2's are in its own table). A reason
# this application does not use reads "--".
step2_ids <- if (exists("removed_step2")) removed_step2$dataset_id else character()
step3 <- other_exclusions[!other_exclusions$dataset_id %in% step2_ids, ]
step3 <- step3[!duplicated(step3$dataset_id), ]
# One unit per table: deposits (the assays kept go to the recap). The rows starting with "..."
# break down the deposits removed.
REASONS <- list(
  "... no or too little metadata" = c("no_metadata", "no_or_limited_metadata"),
  "... not human study samples (another species, in vitro, reference material)" =
    c("not_human", "non_human", "in_vitro", "reference_material"),
  "... do not fit the research question (blood, age)" = c("no_blood_found", "no_adult_with_age"),
  "... files not placed in an assay" = c("no_hit_matched_to_assay", no_assay))
NOT_USED <- c()
n_by <- vapply(REASONS, function(r) sum(step3$reason %in% r), numeric(1))
n_other <- sum(!step3$reason %in% unlist(REASONS))
summary_step <- data.frame(
  item  = c("deposits entering", "deposits removed", names(REASONS), "... other reasons",
            "deposits kept", "recap: assays kept"),
  value = as.character(c(length(D_entering), nrow(step3),
                         ifelse(names(REASONS) %in% NOT_USED, "--", n_by), n_other,
                         length(D_both), nrow(dataset_metadata))))
# For the discussion's recap only: the files with a hit that are kept.
summary_step <- rbind(summary_step, data.frame(
  item  = "recap: files",
  value = as.character(length(unique(paste(all_hits$deposit_id,
                                           basename(all_hits$file_path)))))))
write.csv(summary_step, "artifacts/summary_step3.csv", row.names = FALSE)
summary_step
```

# Session info

```{r session-info}
sessionInfo()
```
