---
title: "04 - Confirmation"
subtitle: "Step 4: every hit re-scored with SIRIUS and CSI:FingerID"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
params:
  run_confirmations: false
---

# Overview

Every curated hit is re-scored with SIRIUS and CSI:FingerID and graded by how
well the target ranks, and only assays in which every target confirms are
kept. An assay that confirms can still be set aside by hand, with its reason,
where its spectra show the rules were not enough (*Assays set aside by hand*).
From the confirmed spectra this step also sets, per assay and target, the
ion step 5 extracts, its m/z, and the retention-time range of the confirmed
spectra, around which step 5 reads the trace and places its window. From the
headers of the raw files with a confirmed spectrum it reads the instrument of
each assay, which sets the m/z tolerance step 5 extracts at.

The SIRIUS runs are done by the notebooks in `confirmation/`, one per
repository, because each repository serves its spectra differently. They are
slow and use one SIRIUS instance, so a normal render of this notebook reads their
cached results. Render with `-P run_confirmations:true` to run them first, one
after the other, each in its own process.

**Input:** `artifacts/hits_curated.csv` and `artifacts/dataset_metadata_table.csv`
(step 3), `artifacts/assay_metadata.csv` (step 2), and the raw files with a
confirmed spectrum, on disk once the confirmation notebooks have run.

**Output:** `artifacts/hits-confirmed-all.csv` (every confirmed spectrum, with its
grade), `artifacts/datasets-confirmed.csv` (one row per assay),
`artifacts/extraction_windows.csv` (one row per assay and target),
`artifacts/file_instruments.csv` (one row per file with a confirmed spectrum: the
instrument and polarity its header records) and `artifacts/assay_instruments.csv`
(one row per assay: its instrument, class and m/z tolerance), read by step 5.

```{r setup}
source("../../R/setup.R")
source("../../R/masst.R")
source("../../R/metadata.R")
source("../../R/ms1.R")   # extraction config, instrument readers
```

# SIRIUS runs (opt-in)

```{r run-confirmations, eval=params$run_confirmations}
for (q in c("confirmation/massIVE-hit-confirmation.qmd",
            "confirmation/metaboLights-confirmation.qmd",
            "confirmation/metabolomics-workbench-confirmation.qmd")) {
  message("Rendering ", q, " ...")
  system2("quarto", c("render", shQuote(q)))
}
```

# Load confirmed features

Each confirmation notebook writes a self-contained CSV (`dataset_id`
deposit-level, `assay`, `role`, `rt`, `mz`, `adduct_inferred`, `source_file`,
`confidence`, plus SIRIUS scoring columns). We bind them, rename the
deposit-level key to `deposit_id`, and attach the combined `dataset_id` and
chromatography from `dataset_metadata_table.csv` via a `(deposit_id, assay)`
join.

Fewer assays can arrive here than `03` curated. A MASST hit is a spectral
similarity, so a deposit can carry features at every target's mass whose best
molecular formula is not the target's at all, and such a deposit confirms
nothing and is simply absent below. The chunk after the summary lists every curated
assay that confirmed no feature together with the formula SIRIUS ranked first
for each role, which is what separates a correct rejection from something to
investigate. A deposit that confirms some roles but not all does arrive, and
drops at the every-target filter further down, where it is reported separately.

```{r load-confirmed}
# One confirmation artifact per repository. A missing one adds no hits instead of
# stopping the render, and is reported: its deposits are absent from all below.
conf_files <- c(
  MassIVE                  = "artifacts/hits-confirmed-massive.csv",
  MetaboLights             = "artifacts/hits-confirmed-metaboLights.csv",
  `Metabolomics Workbench` = "artifacts/hits-confirmed-metabolomics-workbench.csv")
conf_present <- file.exists(conf_files)
if (!any(conf_present))
  stop("no confirmation artifact found; render the confirmation notebooks first")
if (!all(conf_present))
  message("confirmation artifact absent for ",
          paste(names(conf_files)[!conf_present], collapse = ", "),
          ": their deposits are missing from everything below")
conf_all <- do.call(rbind, lapply(conf_files[conf_present], read.csv,
                                  check.names = FALSE))
# A target no longer in compound.csv can still have rows here, from confirmations
# run before a change of targets: they are dropped, and named.
old_roles <- setdiff(unique(conf_all$role), compound_spec$role)
if (length(old_roles)) {
  message("confirmed spectra of roles not in compound.csv, dropped: ",
          paste(old_roles, collapse = ", "))
  conf_all <- conf_all[conf_all$role %in% compound_spec$role, ]
}
rownames(conf_all) <- NULL

names(conf_all)[names(conf_all) == "dataset_id"] <- "deposit_id"

ds_meta_tbl <- read.csv("artifacts/dataset_metadata_table.csv",
                        check.names = FALSE)

conf_all <- dplyr::left_join(
  conf_all,
  ds_meta_tbl[, c("dataset_id", "deposit_id", "assay", "chromatography")],
  by = c("deposit_id", "assay"))

conf_all <- conf_all[!is.na(conf_all$dataset_id), ]
conf_all$confidence <- factor(conf_all$confidence,
                              levels = c("strict", "standard",
                                         "liberal", "not_confirmed"))
# A scan matched by several library spectra can come back once per match: keep
# one row per spectrum and target, at its best tier.
o    <- order(conf_all$confidence)
dup  <- duplicated(conf_all[o, c("dataset_id", "role", "source_file", "rt", "mz",
                                 "adduct_inferred")])
if (any(dup)) message("copies of one spectrum dropped: ", sum(dup))
conf_all <- conf_all[sort(o[!dup]), ]
rownames(conf_all) <- NULL
```

```{r load-confirmed-summary}
#| echo: false
cat("Confirmed features loaded :", nrow(conf_all), "\n")
cat("Assays (combined key)     :", length(unique(conf_all$dataset_id)), "\n")
cat("\nChromatography across confirmed assays:\n")
print(table(unique(conf_all[, c("dataset_id", "chromatography")])$chromatography,
            useNA = "ifany"))
```


# Filter cascade

A hit for every target per assay was already required upstream, so the
remaining filter narrows to what is *interpretable* after SIRIUS: every target
still present after annotation. That is the only condition applied here. Two
conditions deliberately are not: the targets need not share an adduct
(a compound detected as its sodium adduct is still present in the file), and
each (assay, role) is not collapsed to its best tier (see below).

## Every confirmed tier is kept

SIRIUS returns a formula rank and a structure rank per feature, and the
confirmation notebooks bin them into tiers. Their rank limits are set there, and
listed in the table under *Evaluate SIRIUS matches* in
`confirmation/massIVE-hit-confirmation.qmd`:

- **strict**: the target's formula and structure both ranked first;
- **standard**: both within the tighter rank limits;
- **liberal**: both within the looser rank limits;
- **not_confirmed**: beyond liberal; dropped before the confirmed CSVs.

Every confirmed feature is kept, at whatever tier it reached. Collapsing each
(assay, role) to its best tier would mean one `strict` parent feature in an
assay discarding every `standard` parent feature in that same assay,
including rank-1 identifications on other files. That is the right rule for
choosing what step 5 extracts (the ion, its m/z and the stretch of trace it
reads), where the most confident evidence should decide, and the wrong rule for
everything else, where a confirmed feature is a confirmed feature. The tier
restriction lives in `best_tier()`, used by
`derive_extraction_config()` and `extraction_window()`; the
`confidence` column flows through here unchanged so any downstream analysis can
still stratify or weight by tier.

```{r keep-confidence-tier}
CONF_ORDER <- c(strict = 1, standard = 2, liberal = 3, not_confirmed = 4)
conf_best  <- conf_all
```

## `r if (nrow(compound_spec) == 2) "Both roles" else "Every target"` confirmed (any tier above not_confirmed)

A `dataset_id` is kept only if every target in `compound.csv` has at least one
feature at `strict`, `standard`, or `liberal`. The `confidence` column flows
through unchanged so downstream code can stratify or weight by tier in step 6.

```{r filter-every-target}
# A hit confirmed in a QC injection, blank, pooled QC or calibration standard is not
# evidence about a sample, and must not anchor an extraction window either.
n_nonbio_hits <- sum(is_nonbio_file(conf_best$source_file), na.rm = TRUE)
conf_samples <- conf_best |>
  dplyr::filter(confidence %in% c("strict", "standard", "liberal")) |>
  dplyr::filter(!is_nonbio_file(source_file))
conf_final <- conf_samples |>
  dplyr::group_by(dataset_id) |>
  dplyr::filter(all(compound_spec$role %in% role)) |>
  dplyr::ungroup() |>
  as.data.frame()

dropped_ds <- setdiff(unique(conf_all$dataset_id),
                      unique(conf_final$dataset_id))
```

```{r filter-summary}
#| echo: false
cat("Confirmed hits in QC, blank or standard files, dropped:", n_nonbio_hits, "\n")
cat("Surviving assays   :", length(unique(conf_final$dataset_id)), "\n")
cat("Surviving features :", nrow(conf_final), "\n")
if (length(dropped_ds))
  cat(if (nrow(compound_spec) == 2) "Dropped (one role not confirmed at any tier):\n  " else
        "Dropped (not every target confirmed at any tier):\n  ",
      paste(dropped_ds, collapse = "\n  "), "\n", sep = "")
```

A curated assay can leave that surviving set in one of two ways: it confirms
nothing at all, or it confirms some roles and not every one. The discriminating
fact in both cases is the formula SIRIUS ranked first at the target mass: a
different element composition there means the MASST hit was a spectral
look-alike and the re-scoring is doing its job, while the target's own formula
means the deposit really carries that role and fails only for want of the
other targets.

```{r confirmation-attrition}
#| echo: false
.cur <- read.csv("artifacts/hits_curated.csv", stringsAsFactors = FALSE)
.dep_of <- stats::setNames(as.character(.cur$deposit_id), as.character(.cur$dataset_id))
.curated <- unique(as.character(.cur$dataset_id))
.absent  <- setdiff(.curated, unique(as.character(conf_all$dataset_id)))
best_formula <- function(ds, role) {
  f <- file.path(results_dir_for(ds, .dep_of[[ds]]), paste0(role, "_results.csv"))
  if (!file.exists(f)) return("no SIRIUS output")
  d <- read.csv(f, stringsAsFactors = FALSE)
  d <- unique(d[d$rank.x %in% 1, c("xcms_fts", "molecularFormula")])
  if (!nrow(d)) return("no formula ranked")
  tb <- sort(table(d$molecularFormula), decreasing = TRUE)
  sprintf("%s (%d of %d features)", names(tb)[1], tb[1], sum(tb))
}
cat("Curated in 03:", length(.curated),
    "| confirming at least one feature:", length(.curated) - length(.absent),
    if (nrow(compound_spec) == 2) "| surviving both-roles:" else "| confirming every target:",
    length(unique(conf_final$dataset_id)), "\n")
.gone <- c(.absent, setdiff(dropped_ds, .absent))
if (length(.gone)) {
  cat("\nTarget formulas:",
      paste(sprintf("%s %s", compound_spec$role, compound_spec$formula),
            collapse = ", "), "\n")
  for (.d in .gone) {
    cat(" ", .d, if (.d %in% .absent) "(confirmed nothing)" else
          if (nrow(compound_spec) == 2) "(one role only)" else "(not every target)", "\n")
    for (.r in compound_spec$role)
      cat("      ", format(.r, width = max(nchar(compound_spec$role))), "best formula:",
          best_formula(.d, .r), "\n")
  }
}
```

# Assays set aside by hand

The rules above decide which assays confirm, and they are the same for every
assay. Some things only show when an assay's spectra are looked at. An assay
can confirm every target while most of the signal at a target's mass is another
compound of the same formula: the MS2 spectra tell the two apart, but the MS1
trace step 5 measures does not, and no rule here looks for it. Such an assay is
set aside by hand after confirmation, and its reason is written beside it in the
list below. It counts as rejected in the step summary,
and it is left out of everything that follows: the artifacts this step writes,
and therefore steps 5 and 6, do not see it.

```{r set-aside}
# MANUAL WORK: assays set aside by hand after confirmation, with the reason.
# An assay listed here confirmed every target by the rules above.
SET_ASIDE <- character()   # none in this application

# Applied once, here: a set-aside assay leaves conf_final, from which everything
# below and every artifact step 5 reads is built.
ds_set_aside <- intersect(names(SET_ASIDE), unique(conf_final$dataset_id))
if (length(setdiff(names(SET_ASIDE), ds_set_aside)))
  warning("set aside by hand but not confirmed, so nothing to set aside: ",
          paste(setdiff(names(SET_ASIDE), ds_set_aside), collapse = ", "))
conf_final <- conf_final[!conf_final$dataset_id %in% ds_set_aside, ]
```

```{r set-aside-table}
#| echo: false
if (length(ds_set_aside))
  knitr::kable(data.frame(assay = short_ds(ds_set_aside),
                          reason = unname(SET_ASIDE[ds_set_aside])),
               caption = "Assays set aside by hand after confirmation.")
```

Set aside by hand:
`r if (length(ds_set_aside)) paste(short_ds(ds_set_aside), collapse = ", ") else "none"`.
`r length(unique(conf_final$dataset_id))` confirmed assays go on to step 5.

# RT diagnostics

Median and range of the confirmed retention time per role and assay, a check
that each target elutes where step 5 will look for it.

## Per-assay RT ranges

Coloured line is the min-to-max range; point is the median.

```{r plot-rt-windows-raw, fig.width = 9, fig.height = 5}
# One colour per target, in compound.csv order; beyond the second, from a
# qualitative palette.
role_colours <- c("#2166AC", "#D6604D",
                  grDevices::hcl.colors(max(nrow(compound_spec) - 2, 0), "Dark 3"))
role_colours <- setNames(role_colours[seq_len(nrow(compound_spec))], compound_spec$role)

plot_rt_windows <- function(df, title) {
  df |>
    dplyr::filter(!is.na(rt)) |>
    dplyr::group_by(dataset_id, role, chromatography) |>
    dplyr::summarise(rt_lo  = min(rt),
                     rt_hi  = max(rt),
                     rt_med = median(rt),
                     .groups = "drop") |>
    ggplot(aes(y = dataset_id, colour = role)) +
    geom_linerange(aes(xmin = rt_lo, xmax = rt_hi),
                   linewidth = 2, alpha = 0.4,
                   position = position_dodge(width = 0.5)) +
    geom_point(aes(x = rt_med),
               size = 3, position = position_dodge(width = 0.5)) +
    facet_grid(chromatography ~ ., scales = "free_y", space = "free_y") +
    scale_colour_manual(values = role_colours) +
    labs(x = "Retention time (s)", y = NULL, colour = "Role",
         title = title) +
    theme_bw() +
    theme(legend.position  = "bottom",
          strip.background = element_rect(fill = "grey92"))
}

.p_rt <- plot_rt_windows(conf_final, "Per-assay RT ranges - all confirmed features")
# Diagnostic only. Step 5's windows figure (fig-step5-windows) draws these times
# with the windows it sets on the MS1 trace.
.p_rt
```

```{r rt-outliers}
#| echo: false
# Flag an assay whose median confirmed RT is beyond RT_FOLD times (either way)
# the median of all assays of its phase; a phase with one assay flags nothing.
RT_FOLD <- 2
rt_med <- conf_final |>
  dplyr::filter(!is.na(rt)) |>
  dplyr::group_by(dataset_id, chromatography) |>
  dplyr::summarise(rt = stats::median(rt), .groups = "drop") |>
  dplyr::group_by(chromatography) |>
  dplyr::mutate(phase_rt = stats::median(rt)) |>
  dplyr::ungroup() |>
  dplyr::mutate(odd = rt > RT_FOLD * phase_rt | rt < phase_rt / RT_FOLD)
rt_odd <- rt_med[rt_med$odd, ]
# Each flagged assay with its median RT, and where the unflagged assays of its
# phase sit.
rt_odd_text <- if (!nrow(rt_odd)) "none" else
  paste(vapply(seq_len(nrow(rt_odd)), function(i) {
    ph   <- rt_odd$chromatography[i]
    rest <- round(rt_med$rt[rt_med$chromatography %in% ph & !rt_med$odd])
    sprintf("%s (%s) at %.0f s, the unflagged %s assays at %s s",
            rt_odd$dataset_id[i], ph, rt_odd$rt[i], ph,
            if (min(rest) == max(rest)) min(rest)
            else paste(min(rest), "to", max(rest)))
  }, character(1)), collapse = "; ")
```

Chromatography labels come from the deposit / ReDU metadata and paper constants
(`02`), not from RT. Absolute RT is not comparable across assays (it depends on
each method's gradient), so the spread within a phase facet is expected, not
evidence of mislabelling. An assay is worth checking against its source method
when its median confirmed RT is more than `r RT_FOLD` times, or less than
1/`r RT_FOLD` of, the median over all assays of its phase. Flagged this way:
`r rt_odd_text`.

## MS2 trigger RT versus MS1 chromatographic apex

The retention time recorded above is that of the MS2 scan that matched, not of
the compound. Data-dependent acquisition fires as soon as a precursor is intense
enough and can re-fire while it stays intense, so the scans that confirm one
feature scatter across its chromatographic peak and well down its tail. The RT
`05` must centre an extraction window on is the MS1 apex, which is stable where
the trigger time is not.

That is measured rather than assumed, and it is measured in `05`, which is where
the MS1 chromatograms exist. Its *Extraction windows* figure draws, for every
assay it extracts, the confirmed spectra against the window it sets on the MS1
peak they point to. In an application with probe checks
(`_05-probe-checks.qmd`), *Checks that can catch us being wrong* also plots every
confirmed MS2 scan of one assay against the apex of the peak it fired on.

# m/z deviation

The observed precursor m/z should match the inferred adduct's theoretical m/z
within the instrument's mass accuracy; deviation in ppm is `(observed -
theoretical) / theoretical * 1e6`.

```{r mz-deviation, fig.width = 9, fig.height = 4}
PPM_OK   <- 5    # a well-calibrated Orbitrap stays within +/- this
PPM_WARN <- 10   # beyond +/- this: miscalibration or wrong adduct

conf_final$mz_theo <- mapply(
  function(f, a) {
    if (is.na(f) || is.na(a)) return(NA_real_)
    mass2mz(calculateMass(f), adduct = a)[[1]]
  },
  role_formula[conf_final$role],
  conf_final$adduct_inferred,
  USE.NAMES = FALSE)

conf_final$mz_ppm_dev <- (conf_final$mz - conf_final$mz_theo) /
                         conf_final$mz_theo * 1e6

conf_final |>
  dplyr::filter(!is.na(mz_ppm_dev)) |>
  ggplot(aes(y = dataset_id, x = mz_ppm_dev, colour = adduct_inferred)) +
  geom_vline(xintercept = 0,        linetype = 2, colour = "grey40") +
  geom_vline(xintercept = c(-PPM_OK, PPM_OK), linetype = 3, colour = "grey70") +
  geom_jitter(size = 2.5, height = 0.15, alpha = 0.85) +
  facet_grid(. ~ role) +
  labs(y = NULL, x = "m/z deviation (ppm)", colour = "Adduct",
       title = "Precursor m/z deviation vs theoretical adduct mass",
       caption = paste0("Dashed: 0 ppm  ·  Dotted: +/- ", PPM_OK, " ppm")) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r mz-deviation-share}
#| echo: false
dev_ok <- conf_final$mz_ppm_dev[is.finite(conf_final$mz_ppm_dev)]
pct_ok   <- round(100 * mean(abs(dev_ok) <= PPM_OK), 1)
pct_warn <- round(100 * mean(abs(dev_ok) <= PPM_WARN), 1)
```

Of the `r format(length(dev_ok), big.mark = ",")` confirmed spectra with a known ion,
`r pct_ok`% have a precursor m/z within `r PPM_OK` ppm of the ion's exact mass,
where a well-calibrated Orbitrap stays, and `r pct_warn`% within `r PPM_WARN` ppm,
beyond which a calibration issue or a mis-assigned adduct is likely.

# Per-assay summary

```{r dataset-summary-table}
# Per target, in compound.csv order: its confirmed spectra in the assay (n_<role>),
# then their median retention time (rt_<role>_s). One row, as summarise() takes it.
per_target_summary <- function(role, rt) {
  roles   <- compound_spec$role
  n_cols  <- lapply(roles, function(r) sum(role == r))
  rt_cols <- lapply(roles, function(r) round(median(rt[role == r], na.rm = TRUE), 1))
  as.data.frame(c(setNames(n_cols,  paste0("n_", roles)),
                  setNames(rt_cols, paste0("rt_", roles, "_s"))))
}

ds_summary <- conf_final |>
  dplyr::group_by(dataset_id, deposit_id, assay, chromatography) |>
  dplyr::summarise(
    repository      = dplyr::case_when(
      startsWith(deposit_id[1], "MSV")   ~ "MassIVE",
      startsWith(deposit_id[1], "MTBLS") ~ "MetaboLights",
      startsWith(deposit_id[1], "ST")    ~ "Metabolomics Workbench"),
    best_confidence = levels(conf_final$confidence)[
      min(CONF_ORDER[as.character(confidence)])],
    per_target_summary(role, rt),
    .groups         = "drop") |>
  dplyr::left_join(
    ds_meta_tbl[, intersect(c("dataset_id", "const_body_site", "const_country",
                              "const_location_type"),
                            names(ds_meta_tbl))],
    by = "dataset_id")

ds_summary
```

# What the structure ranking prefers

Confirmation places the target among the structures CSI:FingerID retrieves for
its formula. It does not say the fingerprint prefers it, and where the formula
has a near neighbour that fragmentation cannot separate, a role can confirm at a
tier and still be outranked in every spectrum of the corpus. The candidate lists
the confirmation notebooks already wrote answer that directly, so it costs
nothing to look: for each confirmed spectrum, where the target sits among the
structures of its own formula, and the score gap to whatever sits above it.
Scores compare within a spectrum rather than across spectra, so both gaps are
within-spectrum differences summarised over spectra.

A role the ranking never prefers is not thereby refuted: positional isomers of
one formula are the case fingerprint prediction is weakest on, and a small
deficit is a tie rather than a verdict. Its identity then rests on something
other than fragmentation, retention time most often, and that has to be stated
rather than assumed.

```{r structure-ranking}
#| echo: false
SR <- structure_ranks(conf_final)
sr_role <- if (is.null(SR)) data.frame() else
  SR |>
    dplyr::group_by(role) |>
    dplyr::summarise(
      spectra        = dplyr::n(),
      target_first   = sum(rank == 1),
      median_lead    = round(stats::median(lead, na.rm = TRUE), 1),
      min_lead       = if (all(is.na(lead))) NA_real_ else round(min(lead, na.rm = TRUE), 1),
      median_deficit = round(stats::median(deficit, na.rm = TRUE), 1),
      .groups = "drop")
sr_top <- if (is.null(SR)) data.frame() else
  SR |>
    dplyr::filter(rank > 1) |>
    dplyr::count(role, top_name, top_key, name = "spectra") |>
    dplyr::arrange(role, dplyr::desc(spectra))
```

```{r structure-ranking-role}
#| echo: false
knitr::kable(sr_role, caption = paste(
  "Where the target sits among the structures of its own formula, over every",
  "confirmed spectrum. Lead is its CSI:FingerID score ahead of the next",
  "candidate where it ranks first; deficit is how far behind the first it sits",
  "where it does not."))
```

```{r structure-ranking-top}
#| echo: false
knitr::kable(sr_top, caption = paste(
  "What outranks the target where something does, by the structure database's",
  "name and by InChIKey skeleton. A database entry may be an isotope-labelled",
  "standard, which shares the skeleton of the unlabelled compound."))
```

# Write artifacts

Step 5 reads the confirmed features and the per-assay summary.

```{r write-artifacts}
write.csv(conf_final, "artifacts/hits-confirmed-all.csv", row.names = FALSE)
write.csv(ds_summary, "artifacts/datasets-confirmed.csv", row.names = FALSE)
```

# What step 5 extracts

Step 5 measures each target in every file of an assay at one ion and one m/z,
and reads its trace over one stretch of retention time. The three are set here,
per assay and target, from the confirmed spectra (`derive_extraction_config()`
in `R/ms1.R`):

- **ion**: the molecular ion if it was confirmed, otherwise the most frequent
  confirmed adduct;
- **m/z**: the median observed precursor m/z of that ion, which absorbs the
  instrument's own calibration offset;
- **range**: from the earliest to the latest confirmed retention time of that
  ion, at the best tier the target reached in the assay, with `RT_PAD_S` seconds
  added on each side.

The range is where step 5 reads the trace, not the window in which it looks for
the peak. The confirmed spectra were triggered along the whole peak and down its
tail, and can include a second peak of the same m/z (see *MS2 trigger RT versus
MS1 chromatographic apex* above), so the range can be several times wider than
the peak. Step 5 reads the trace over the range, and a margin beyond it, and
places the window there, on the MS1 peak the confirmed spectra point to; its
notebook draws that window. It reads these rows from
`artifacts/extraction_windows.csv` and checks that each target has a
confirmation on one of the files it extracts. If a spectrum that set a range
lies on a file step 5 does not extract, step 5 sets that range again from the
files it does extract, and its log says so.

```{r extraction-windows}
RT_PAD_S <- 5   # seconds added on each side of the confirmed span

target_name <- setNames(compound_spec$name, compound_spec$role)
ids <- sort(unique(conf_final$dataset_id), method = "radix")
# One row per target of the assay, in compound.csv order. The cfg names a target's
# fields by its prefix (adduct_par, mz_met, rt_par_win).
window_rows <- function(ds) {
  cfg <- derive_extraction_config(ds, conf_final, rt_pad_s = RT_PAD_S)
  do.call(rbind, lapply(compound_spec$role, function(r) {
    p <- target_prefix(r)
    data.frame(dataset_id = ds, assay = cfg$assay, role = r,
               target    = target_name[[r]],
               ion       = cfg[[paste0("adduct_", p)]],
               mz        = cfg[[paste0("mz_", p)]],
               rt_lo     = cfg[[paste0("rt_", p, "_win")]][1],
               rt_hi     = cfg[[paste0("rt_", p, "_win")]][2],
               rt_pad_s  = RT_PAD_S,
               n_spectra = sum(conf_final$dataset_id == ds & conf_final$role == r),
               row.names = NULL)
  }))
}
ext_windows <- do.call(rbind, lapply(ids, window_rows))
# No assay has confirmed every target yet: the same table, empty.
if (is.null(ext_windows))
  ext_windows <- data.frame(dataset_id = character(), assay = character(), role = character(),
                            target = character(), ion = character(), mz = numeric(),
                            rt_lo = numeric(), rt_hi = numeric(), rt_pad_s = numeric(),
                            n_spectra = integer())
write.csv(ext_windows, "artifacts/extraction_windows.csv", row.names = FALSE)
ext_windows
```

# The instrument of each assay

Step 5 reads each target's trace within an m/z tolerance, and how wide it must be
depends on the mass analyser. On an Orbitrap the scans across a true peak sit
within a few ppm of its mass; on a time-of-flight instrument the scans on a
peak's edges scatter much further. Step 5 therefore extracts an assay at
`r PPM_BY_CLASS[["orbitrap"]]` ppm on an Orbitrap, at `r PPM_BY_CLASS[["tof"]]`
ppm on a time-of-flight instrument, and at `r PPM_DEFAULT` ppm where the
instrument cannot be told (`PPM_BY_CLASS`, `PPM_DEFAULT` and
`instrument_class()` in `R/ms1.R`).

```{r header-sample-size}
N_HEADER <- 5   # files whose header is read per assay, when they agree
```

The deposit metadata often does not record the instrument, so it is read from
the raw files. Every file with a confirmed spectrum is on disk once the
confirmation notebooks have run, and its header records the manufacturer, the
model, the mass analyser and the ion source; a few of its scans give its
polarity (`read_instrument()`). An assay's files come from one instrument, so
`r N_HEADER` files per assay are read, spread over its file list; when they do
not agree on the instrument class or the polarity, all of the assay's files are
read. Each file read is kept in `artifacts/file_instruments.csv`, so a later
render reads only the files not yet in it. An assay takes the class its files
record, and is `unknown` where they record none or more than one.

```{r file-instruments}
# Instrument and polarity from file headers: N_HEADER files per assay, all of
# them when those disagree. Files already in the table are not read again.
hit_files <- unique(data.frame(dataset_id = conf_final$dataset_id,
                               deposit_id = conf_final$deposit_id,
                               file       = basename(as.character(conf_final$source_file))))
hit_files <- hit_files[!is.na(hit_files$file), ]
hit_files <- hit_files[order(hit_files$dataset_id, hit_files$file), ]
INST_CSV  <- "artifacts/file_instruments.csv"
file_inst <- if (file.exists(INST_CSV)) read.csv(INST_CSV, colClasses = "character") else
  data.frame(dataset_id = character(), file = character(), manufacturer = character(),
             model = character(), analyzer = character(), ion_source = character(),
             polarity = character())
key <- function(d) paste(d$dataset_id, d$file)
read_headers <- function(want) {
  todo <- want[!key(want) %in% key(file_inst), ]
  if (!nrow(todo)) return(file_inst)
  todo$path <- cached_raw_paths(todo$file)
  todo <- todo[!is.na(todo$path), ]
  new  <- lapply(split(todo, todo$deposit_id), function(d) {
    message("[", d$deposit_id[1], "] reading the header of ", nrow(d), " file(s)")
    cbind(d[, c("dataset_id", "file")], do.call(rbind, lapply(d$path, read_instrument)))
  })
  rbind(file_inst, do.call(rbind, new))
}
spread <- function(f) f[unique(round(seq(1, length(f), length.out = min(N_HEADER, length(f)))))]
first_pick <- do.call(rbind, lapply(split(hit_files, hit_files$dataset_id),
                                    function(d) d[d$file %in% spread(d$file), ]))
n_before  <- nrow(file_inst)
file_inst <- read_headers(first_pick)
# An assay whose files read so far disagree is read in full.
agree <- vapply(split(file_inst, file_inst$dataset_id), function(d)
  length(unique(instrument_class(d$model, d$analyzer, d$manufacturer))) <= 1 &&
    length(unique(d$polarity[!is.na(d$polarity)])) <= 1, logical(1))
disagree  <- names(agree)[!agree]
file_inst <- read_headers(hit_files[hit_files$dataset_id %in% disagree, ])
n_read    <- nrow(file_inst) - n_before
# Only the current hit files: a new confirmation can drop one.
file_inst <- file_inst[key(file_inst) %in% key(hit_files), ]
write.csv(file_inst, INST_CSV, row.names = FALSE)
n_not_on_disk <- sum(!key(first_pick) %in% key(file_inst))
```

```{r assay-instruments}
# One row per confirmed assay: instrument, class, the m/z tolerance step 5 uses,
# majority polarity, files that differ, and what the metadata records.
tech <- read.csv("artifacts/assay_metadata.csv", stringsAsFactors = FALSE)
collapse <- function(x) {
  x <- unique(x[!is.na(x) & nzchar(x)])
  if (length(x)) paste(x, collapse = "; ") else NA_character_
}
one_class <- function(cls) {
  cls <- setdiff(unique(cls), "unknown")
  if (length(cls) == 1) cls else "unknown"
}
assay_row <- function(ds) {
  d   <- file_inst[file_inst$dataset_id == ds, ]
  cls <- one_class(instrument_class(d$model, d$analyzer, d$manufacturer))
  pol <- if (any(!is.na(d$polarity))) names(which.max(table(d$polarity))) else NA_character_
  # The metadata of the assay: every row of a MassIVE deposit, which is one assay,
  # and the assay's own row elsewhere.
  s <- ds_summary[ds_summary$dataset_id == ds, ]
  m <- tech[tech$dataset_id == s$deposit_id & (is.na(s$assay) | tech$assay %in% s$assay), ]
  data.frame(dataset_id          = ds,
             instrument          = collapse(paste(d$manufacturer, d$model)),
             analyzer            = collapse(d$analyzer),
             class               = cls,
             ppm                 = if (cls %in% names(PPM_BY_CLASS)) PPM_BY_CLASS[[cls]] else PPM_DEFAULT,
             polarity            = pol,
             n_files             = nrow(d),
             n_other_polarity    = sum(!is.na(d$polarity) & d$polarity != pol),
             metadata_instrument = collapse(c(m$instrument, m$mass_analyzer)),
             metadata_class      = one_class(instrument_class(m$instrument, m$mass_analyzer)),
             metadata_polarity   = collapse(m$polarity))
}
assay_inst <- do.call(rbind, lapply(ids, assay_row))
# No assay has confirmed every target yet: the same table, empty.
if (is.null(assay_inst))
  assay_inst <- data.frame(dataset_id = character(), instrument = character(),
                           analyzer = character(), class = character(), ppm = numeric(),
                           polarity = character(), n_files = integer(),
                           n_other_polarity = integer(), metadata_instrument = character(),
                           metadata_class = character(), metadata_polarity = character())
write.csv(assay_inst, "artifacts/assay_instruments.csv", row.names = FALSE)

# Where the metadata does not tell the instrument or the polarity, or tells another.
meta_check <- with(assay_inst, data.frame(
  assay      = short_ds(dataset_id),
  header     = ifelse(n_files == 0, "no file read",
                      paste0(instrument, " (", class, "), ", polarity)),
  metadata   = paste0(ifelse(is.na(metadata_instrument), "no instrument", metadata_instrument),
                      " (", metadata_class, "), ",
                      ifelse(is.na(metadata_polarity), "no polarity", metadata_polarity)),
  instrument = ifelse(metadata_class == "unknown", "not told",
               ifelse(n_files == 0, "not read",
               ifelse(metadata_class == class, "agrees", "differs"))),
  polarity   = ifelse(is.na(metadata_polarity), "not told",
               ifelse(n_files == 0, "not read",
               ifelse(is.na(polarity), "not in headers",
               ifelse(metadata_polarity == polarity, "agrees", "differs"))))))
meta_off <- meta_check[meta_check$instrument != "agrees" | meta_check$polarity != "agrees", ]
# Assays whose files do not all share one polarity.
mixed <- assay_inst[assay_inst$n_other_polarity > 0, ]
mixed_text <- if (!nrow(mixed)) "none" else
  paste(sprintf("%s (%d of %d files not %s)", short_ds(mixed$dataset_id),
                mixed$n_other_polarity, mixed$n_files, mixed$polarity), collapse = "; ")
```

The headers of `r format(nrow(file_inst), big.mark = ",")` files were read for the
`r nrow(assay_inst)` confirmed assays. Assays whose first files disagreed, and were
read in full: `r if (length(disagree)) paste(short_ds(disagree), collapse = ", ") else "none"`.
Files chosen but not on disk: `r n_not_on_disk`. `r sum(assay_inst$class == "orbitrap")` assays were
recorded on an Orbitrap and are extracted at `r PPM_BY_CLASS[["orbitrap"]]` ppm,
`r sum(assay_inst$class == "tof")` on a time-of-flight instrument at
`r PPM_BY_CLASS[["tof"]]` ppm, and `r sum(assay_inst$class == "unknown")` on an
instrument their headers do not tell, at `r PPM_DEFAULT` ppm.

```{r assay-instruments-table}
#| echo: false
knitr::kable(data.frame(assay = short_ds(assay_inst$dataset_id),
                        instrument = assay_inst$instrument, class = assay_inst$class,
                        ppm = assay_inst$ppm, files = assay_inst$n_files),
             caption = "The instrument of each confirmed assay, read from its files' headers, and the m/z tolerance step 5 extracts it at.")
```

The metadata does not tell the instrument of
`r sum(meta_check$instrument == "not told")` assays and records another class
than the headers for `r sum(meta_check$instrument == "differs")`; it does not
tell the polarity of `r sum(meta_check$polarity == "not told")` and records
another for `r sum(meta_check$polarity == "differs")`. Those assays:

```{r assay-instruments-metadata}
#| echo: false
knitr::kable(meta_off, row.names = FALSE,
             caption = "Assays whose metadata does not tell, or tells another, instrument or polarity than their files' headers.")
```

Step 5 extracts one ion of each target, of one charge, so a file recorded in the
other polarity cannot show it. Assays whose files do not all share one polarity:
`r mixed_text`.

# Step summary

The results of this step, as the chapter's table shows them.

Confirmed spectra of the assays that go on to step 5: `r knitr::combine_words(paste(vapply(compound_spec$role, function(r) format(sum(conf_final$role == r), big.mark = ","), character(1)), "of the", compound_spec$role))`.

```{r step-summary}
# A curated assay without SIRIUS results for every target with a hit in it is
# neither confirmed nor rejected: it is left out of both and named.
.roles_of  <- split(.cur$role, .cur$dataset_id)
sirius_ran <- vapply(.curated, function(ds) all(file.exists(file.path(
  results_dir_for(ds, .dep_of[[ds]]), paste0(unique(.roles_of[[ds]]), "_results.csv")))),
  logical(1))
# The confirmed assays that go on to step 5: an assay set aside by hand
# (ds_set_aside) has already left conf_final, and is counted as rejected.
ds_confirmed    <- unique(conf_final$dataset_id)
ds_not_run      <- setdiff(.curated[!sirius_ran], c(ds_confirmed, ds_set_aside))
# An assay whose confirmed spectra are all in QC, blank or standard files confirms
# no target in a sample, so it counts with those that confirm nothing.
ds_no_target    <- setdiff(.curated, c(conf_samples$dataset_id, ds_not_run))
ds_some_targets <- setdiff(dropped_ds, c(ds_no_target, ds_not_run))
if (length(ds_not_run))
  message("no SIRIUS results yet, neither confirmed nor rejected: ",
          paste(ds_not_run, collapse = ", "))

# One unit per table: assays. The confirmed spectra are given in the text above.
n_targets <- nrow(compound_spec)
summary_step <- data.frame(
  item  = c("assays entering",
            paste0("assays confirmed (", targets_phrase(n_targets), ")"),
            "assays removed",
            "... no target confirmed",
            paste("...", some_targets_phrase(n_targets), "confirmed"),
            "... set aside by hand"),
  value = c(length(.curated),
            length(ds_confirmed),
            length(ds_no_target) + length(ds_some_targets) + length(ds_set_aside),
            length(ds_no_target),
            length(ds_some_targets),
            length(ds_set_aside)))
# For the discussion's recap only: the deposits behind the confirmed assays that go
# on, and their files with a confirmed spectrum.
summary_step <- rbind(summary_step, data.frame(
  item  = c("recap: deposits", "recap: files"),
  value = c(length(unique(conf_final$deposit_id[conf_final$dataset_id %in% ds_confirmed])),
            length(unique(paste(conf_final$deposit_id, basename(as.character(conf_final$source_file))))))))
write.csv(summary_step, "artifacts/summary_step4.csv", row.names = FALSE)
summary_step
```

# Session info

```{r session-info}
sessionInfo()
```
