---
title: "06 - Analysis"
subtitle: "Step 6: tryptophan and kynurenine against age"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
---

# Overview

Tryptophan is broken down to kynurenine by indoleamine 2,3-dioxygenase, an enzyme
that inflammation induces, and the kynurenine to tryptophan ratio is the usual
read-out of its activity. That activity rises with age. The hypothesis tested
here, stated before any result, is therefore that **the kynurenine to tryptophan
ratio rises with age**.

The test is one model per matrix: the log2 ratio against age in adult plasma, and
the same in adult serum. Each model pools the assays of its matrix with a term for
each assay, so that each assay keeps its own level and the slope is read within
assays. Sex and health status enter both models. Plasma and serum are kept apart
because the ratio need not sit at the same level in both, and because a person
sampled in both would count twice in one model over both. Each model is fitted
once more on its healthy units alone. The notebook then asks which assays show
the rise on their own. Before any of this, where several assays measured the same
samples, only one of them is kept.

**Input:** `artifacts/sample_table_perfile.csv` (step 5, one row per file with
each target's detection and peak area), `artifacts/bio_files_per_ds.rds` and
`artifacts/bio_meta_per_ds.rds` (step 5, the extracted files and whether a
metadata source covers each), `artifacts/datasets-confirmed.csv` and
`artifacts/hits-confirmed-all.csv` (step 4), and `artifacts/related_deposits.csv`
(step 2).

**Output:** `artifacts/assay_choice.csv`, the assay kept from each set of
samples; `artifacts/ratio_covariate_pooled.csv` and
`artifacts/ratio_covariate_per_assay.csv`, the fits of each matrix and of each
assay alone in each matrix; `artifacts/summary_step6.csv`;
`figures/ratio_vs_age_by_matrix.png`, one row per matrix; and
`figures/ratio_vs_age_healthy.png` (and `.pdf`), the healthy units alone; and
`figures/loo_assay_by_matrix.png` (and `.pdf`), each matrix's model without each of
its assays in turn.

```{r setup}
#| warning: false
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
})
source("../../R/metadata.R")   # norm_file_key(), as_num(), short_ds()
source("../../R/analysis.R")   # the step 6 helpers

MATRICES      <- c("plasma", "serum")   # blood, one model each
ADULT_MIN_AGE <- 18
# A model is fitted (a matrix, its healthy units, or one assay alone) when it has
# this many units over this many years of age.
MIN_UNITS <- 8
MIN_SPAN  <- 10
# Values that mean healthy, as a list: a pattern such as "healthy|control" also
# matches QC. Chagas uninf_prog is left out: the deposit does not define it.
HEALTHY <- c("healthy", "control", "healthy control", "uninfected", "uninf_non")
# Health values that mark an injection, not a sample (the Chagas panel's QC,
# blanks and standard mix). Such files come from no person and are left out.
NOT_SAMPLE <- c("qc", "blank", "blank_tube", "6mix")
# The three health levels, in the order of the tables.
LEVELS <- c("healthy", "not healthy", "not recorded")

# How the model terms read in tables and titles.
TERM_NAMES <- c(agen = "age", sex2 = "sex", health = "health", study = "assay")

# How a slope and a p-value read in the prose and the step summary.
fmt   <- function(r) sprintf("%+.3f (%.3f to %.3f)", r$slope, r$lo, r$hi)
fmt_p <- function(p) formatC(p, digits = 2, format = "fg", flag = "#")
# A list for the text: "A", "A and B", "A, B and C".
and_list <- function(x) if (length(x) < 3) paste(x, collapse = " and ") else
  paste(paste(head(x, -1), collapse = ", "), "and", tail(x, 1))
# Numbers in the displayed tables, to three significant digits.
rounded <- function(d) { num <- vapply(d, is.double, TRUE); d[num] <- lapply(d[num], signif, 3); d }

FIG_DIR <- "figures"; dir.create(FIG_DIR, showWarnings = FALSE)
figp    <- function(f) file.path(FIG_DIR, f)
```

# Load the inputs

```{r load}
pf        <- read.csv("artifacts/sample_table_perfile.csv", stringsAsFactors = FALSE)
datasets  <- read.csv("artifacts/datasets-confirmed.csv", stringsAsFactors = FALSE)
related   <- read.csv("artifacts/related_deposits.csv", stringsAsFactors = FALSE)
conf      <- read.csv("artifacts/hits-confirmed-all.csv", check.names = FALSE,
                      stringsAsFactors = FALSE)
bio_files <- readRDS("artifacts/bio_files_per_ds.rds")
bio_meta  <- readRDS("artifacts/bio_meta_per_ds.rds")

# Stops if step 5 has no file list for an assay of the per-file table: the two
# come from different renders, and step 6 must wait for both.
pf$has_metadata <- has_metadata(pf, bio_meta)

# Assay labels: the accession, or the full assay id where an accession has several.
lab <- function(x) short_ds(x, datasets$dataset_id)
```

# One assay per set of samples

Some samples were measured more than once, by assays with different LC-MS set-ups.
The assays of one Metabolomics Workbench deposit share their sample names, and
step 2 declares the other sets (`related_deposits.csv`): the Chagas serum panel, run
on a HILIC and on a reverse-phase column, and the plasma samples of MSV000082630,
measured again on another instrument as MSV000086158. Counting each would count
a subject more than once, so one assay of each set is kept: the one in which MS1
detects both targets in the most files; on a tie, the one with more files in which
MS2 confirms both; still tied, the first assay id. The files are counted as step 5
counts them, over every extracted file, before the restriction to files with
metadata.

```{r keep-one}
choice <- keep_one_assay(pf, datasets, related, conf, bio_files)
write.csv(choice, "artifacts/assay_choice.csv", row.names = FALSE)

# The sets with more than one assay, the kept one first.
multi <- choice[choice$group %in% choice$group[duplicated(choice$group)], ]
multi$chromatography <- datasets$chromatography[match(multi$dataset_id, datasets$dataset_id)]
knitr::kable(data.frame(set = lab(multi$group), assay = lab(multi$dataset_id),
                        column = multi$chromatography, files = multi$files,
                        both_ms1 = multi$ms1_both, both_ms2 = multi$ms2_both,
                        kept = multi$kept),
             col.names = c("set", "assay", "chromatography", "files",
                           "files with both targets (MS1)",
                           "files with both targets (MS2)", "kept"),
             caption = "Assays that measured the same samples")

describe <- function(i) sprintf("%s (%s)", lab(multi$dataset_id[i]), multi$chromatography[i])
kept_txt <- vapply(split(seq_len(nrow(multi)), factor(multi$group, unique(multi$group))), function(i)
  sprintf("%s is kept over %s", describe(i[multi$kept[i]]),
          and_list(describe(i[!multi$kept[i]]))), "")
n_out <- sum(!choice$kept)
n_in  <- sum(choice$kept)

pf <- pf[pf$dataset_id %in% choice$dataset_id[choice$kept], ]
```

`r if (nrow(multi)) paste0(paste(kept_txt, collapse = "; "), ".") else "No two assays measured the same samples."`
`r n_out` `r if (n_out == 1) "assay is" else "assays are"` set aside, and
`r n_in` `r if (n_in == 1) "goes" else "go"` on to the analysis.

# Analysis units

Only files a metadata source covers are analysed. A file gives a ratio where both
targets are detected with a peak area; a truncated peak is detected but has no
area, so it gives none. The ratio is kynurenine over tryptophan, on a log2 scale.
Files are then reduced to units: one per subject and matrix, so that a study
sampling each person many times does not outweigh one sampling once. A unit's
ratio is the median over its files that give one; its age, sex and health are
the values its files record, whether or not a file gives a ratio, and are left
missing where its files disagree. Where a file names no subject, that file is
its own unit. A subject sampled in both
plasma and serum therefore gives two units, one in each matrix, and each goes to
the model of its own matrix. The models use
adult blood units: plasma or serum, aged `r ADULT_MIN_AGE` or older.

The deposits write health in their own words. Each value gets a level:
healthy where it is one of the healthy words
(`r and_list(sprintf('"%s"', HEALTHY))`), not recorded where it is missing or a
placeholder such as "not collected", and not healthy for any other value, such
as a disease name or "infected". The Chagas panel's "uninf_prog" is not among
the healthy words: the deposit does not say what it means, so it is not taken
as healthy. A value that marks a QC, blank or mix injection
(`r and_list(sprintf('"%s"', NOT_SAMPLE))`) is not a health status: such a file
comes from no person and is left out. The table lists every value found in the
files analysed, with the level it gets, so the mapping can be checked.

```{r frame}
# Three levels: "healthy" for a value in `healthy`, "not recorded" where missing
# or a placeholder (is_missing_val()), else "not healthy".
health_level <- function(x, healthy) {
  s <- tolower(trimws(as.character(x)))
  ifelse(is_missing_val(x), "not recorded",
         ifelse(s %in% tolower(healthy), "healthy", "not healthy"))
}

pf <- pf[pf$has_metadata, ]
# QC, blank and mix injections are not samples.
not_sample   <- tolower(trimws(pf$health_status)) %in% NOT_SAMPLE
n_not_sample <- sum(not_sample)
pf <- pf[!not_sample, ]
pf$fk     <- norm_file_key(pf$file)
# Kynurenine (the metabolite) over tryptophan (the parent), on step 5's peak
# areas: `area` is "auc", or "auc0" in the check on the area baseline.
file_l2   <- function(d, area = "auc")
  ifelse(d$par_detected %in% TRUE & d$met_detected %in% TRUE,
         log2(d[[paste0("met_", area)]] / d[[paste0("par_", area)]]), NA_real_)
pf$l2     <- file_l2(pf)
pf$site   <- canon_site(pf$body_site)
pf$sex2   <- harmonise_sex(pf$sex)
pf$agen   <- as_num(pf$age)
pf$health <- health_level(pf$health_status, HEALTHY)
pf$subj   <- nz(pf$subject_id)
pf$study  <- pf$dataset_id

# Every health value as recorded, the level it gets, and where it is found.
health_map <- pf |>
  mutate(value = ifelse(is.na(health_status) | !nzchar(trimws(health_status)),
                        "(empty)", health_status)) |>
  group_by(value, level = health) |>
  summarise(files = n(),
            assays = paste(lab(sort(unique(study), method = "radix")), collapse = ", "),
            .groups = "drop") |>
  arrange(level, value)
knitr::kable(health_map, col.names = c("health as recorded", "level", "files", "assays"),
             caption = "The health values of the files analysed, and their level")
```

`r if (n_not_sample) paste0(n_not_sample, if (n_not_sample == 1) " file is" else " files are", " left out as QC, blank or mix injections.") else "No file carries a QC, blank or mix value."`

```{r sleep-study}
# Sleep study, by hand: "unhealthy (NOS)" marks draws under sleep deprivation, so
# only files its own metadata labels healthy are kept, and the person is healthy.
SLEEP_STUDY <- "MSV000082630"
# Each label and the disease the deposit writes with it.
SLEEP_LABELS <- c("healthy" = "no disease reported", "unhealthy (NOS)" = "sleep deprivation")
in_sleep <- pf$study == SLEEP_STUDY
# The deposit's own table, one row per file of the deposit analysed here.
own <- read.delim(file.path("dataset_metadata", paste0(SLEEP_STUDY, "_metadata.tsv")),
                  stringsAsFactors = FALSE)
own <- own[match(pf$fk[in_sleep], norm_file_key(own$filename)), ]
# The per-file health must be the deposit's own HealthStatus, file by file, and
# each label must come with its disease.
if (!identical(nz(own$HealthStatus), nz(pf$health_status[in_sleep])))
  stop(SLEEP_STUDY, ": the per-file health is not the deposit's own HealthStatus")
labelled <- own$HealthStatus %in% names(SLEEP_LABELS)
if (!identical(own$DOIDCommonName[labelled], unname(SLEEP_LABELS[own$HealthStatus[labelled]])))
  stop(SLEEP_STUDY, ": a health label does not come with its disease")
sleep_kept   <- in_sleep & pf$health == "healthy"
n_sleep      <- sum(in_sleep)
n_sleep_kept <- sum(sleep_kept)
sleep_people <- n_distinct(pf$subj[sleep_kept])
# People of the deposit with no file labelled healthy: they drop out.
sleep_lost   <- setdiff(pf$subj[in_sleep], pf$subj[sleep_kept])
pf <- pf[!in_sleep | sleep_kept, ]
```

`r SLEEP_STUDY`, a sleep study of healthy volunteers, labels its files "healthy"
or "unhealthy (NOS)" in its own metadata. Its disease column reads "no disease
reported" with the first label and "sleep deprivation" with the second, so the
second marks the draws taken under sleep deprivation, the study's intervention.
For this deposit only, a person's ratio therefore comes from their files
labelled healthy, and the person is healthy: of its `r n_sleep` files a metadata
source covers, `r n_sleep_kept` are kept, from `r sleep_people` people.
`r if (length(sleep_lost)) paste0(length(sleep_lost), if (length(sleep_lost) == 1) " person has" else " people have", " no file labelled healthy and drop", if (length(sleep_lost) == 1) "s", " out.")`

```{r units}
# One unit per subject and matrix, keeping the subject. Units are formed over all
# files before those without a ratio drop, so age, sex and health use every file.
ru <- unitise(pf, fields = "subj")
# Units whose files disagree on age, sex or health, which they then lack.
ru_disagree <- disagree_text(ru, TERM_NAMES)
ru <- ru[is.finite(ru$l2), ]

# Adult blood units with an age. The models use those with a sex recorded.
aged   <- ru[ru$site %in% MATRICES & is.finite(ru$agen) & ru$agen >= ADULT_MIN_AGE, ]
no_sex <- aged[is.na(aged$sex2), ]
blood  <- aged[!is.na(aged$sex2), ]
# The matrices with units in the models, in the order of MATRICES.
mats   <- intersect(MATRICES, blood$site)

ids   <- sort(unique(ru$study), method = "radix")
count <- function(u) as.integer(table(factor(u$study, levels = ids)))
ages  <- function(s, d = blood) {
  a <- d$agen[d$study == s]
  if (length(a)) sprintf("%g to %g", min(a), max(a)) else "-"
}
per_site <- do.call(cbind, lapply(MATRICES, function(m) count(aged[aged$site == m, ])))
knitr::kable(data.frame(
  assay      = lab(ids),
  with_ratio = count(ru),
  per_site,
  with_sex   = count(blood),
  unit       = vapply(ids, function(s)
                 paste(sort(unique(aged$kind[aged$study == s]), method = "radix"),
                       collapse = " and "), ""),
  ages       = vapply(ids, ages, "", d = aged)),
  row.names = FALSE,
  col.names = c("assay", "units with a ratio",
                sprintf("adult %s units with an age", MATRICES),
                "of which a sex is recorded", "unit", "ages"),
  caption = "Units per assay, from every unit with a ratio to those in the models")

# The same adult blood units by matrix, assay, sex and health.
SH_COLS <- c("female", "male", "sex not recorded",
             "healthy", "not healthy", "health not recorded")
sh_rows <- function(d) {
  s   <- sort(unique(d$study), method = "radix")
  tab <- function(v, lv) unclass(table(factor(d$study, s), factor(v, lv)))
  x   <- cbind(tab(ifelse(is.na(d$sex2), "not recorded", d$sex2),
                   c("female", "male", "not recorded")),
               tab(d$health, LEVELS))
  x   <- rbind(x, colSums(x))
  colnames(x) <- SH_COLS
  data.frame(matrix = d$site[1], assay = c(lab(s), "total"), x,
             check.names = FALSE, row.names = NULL)
}
by_sh <- bind_rows(lapply(split(aged, factor(aged$site, intersect(MATRICES, aged$site))),
                          sh_rows))
knitr::kable(by_sh, row.names = FALSE,
             caption = "Adult blood units with an age, by matrix, sex and health")

n_site   <- table(factor(aged$site, MATRICES))
n_sex    <- table(factor(ifelse(is.na(aged$sex2), "not recorded", aged$sex2),
                         c("female", "male", "not recorded")))
n_health <- table(factor(aged$health, LEVELS))

# What adding sex to the models costs: the units with no sex recorded, and the
# assays left with no unit at all.
lost_assays  <- setdiff(unique(aged$study), blood$study)
sex_loss_txt <- if (nrow(no_sex) == 0) {
  "Every one has a sex recorded, so adding sex to the models loses no unit."
} else {
  n <- table(factor(lab(no_sex$study)))
  paste0("Adding sex to the models loses the ", nrow(no_sex),
         if (nrow(no_sex) == 1) " unit" else " units", " with no sex recorded (",
         paste(sprintf("%s: %d", names(n), n), collapse = ", "), ")",
         if (length(lost_assays)) paste0(", and with them ", and_list(lab(lost_assays)),
                                         ", which records no sex at all") else
           ", and no assay", ".")
}

# People with a unit in both matrices, keyed by deposit and subject: plasma and
# serum of one Workbench deposit can be two assays.
named   <- blood[!is.na(blood$subj), ]
person  <- paste(datasets$deposit_id[match(named$study, datasets$dataset_id)], named$subj)
in_both <- names(which(tapply(named$site, person, n_distinct) >= 2))
both_assays <- sort(unique(named$study[person %in% in_both]), method = "radix")

# The health levels each model holds.
levels_in <- function(d) {
  l <- intersect(LEVELS, d$health)
  if (length(l) == length(LEVELS)) "all three levels" else
    paste(if (length(l) == 1) "the level" else "the levels", and_list(sprintf('"%s"', l)))
}
lv <- vapply(mats, function(m) levels_in(blood[blood$site == m, ]), "")
levels_txt <- if (!length(mats)) "" else if (length(mats) > 1 && length(unique(lv)) == 1) {
  paste0("Both models hold units of ", lv[[1]], ".")
} else {
  paste0("The ", mats[1], " model holds units of ", lv[[1]],
         if (length(mats) > 1) paste0("; the ", mats[-1], " model, units of ", lv[-1],
                                      collapse = ""), ".")
}
# Assays in the models with units that are single files: no subject is named.
file_assays <- sort(unique(blood$study[blood$kind == "file"]), method = "radix")
```

`r if (nzchar(ru_disagree)) paste0("The files of some units disagree on age, sex or health, which these units then lack: ", ru_disagree, ".") else "No unit's files disagree on age, sex or health."`
Of the `r nrow(aged)` adult blood units with an age, from
`r n_distinct(aged$study)` assays, `r n_site[["plasma"]]` are plasma and
`r n_site[["serum"]]` serum; `r n_sex[["female"]]` are women and
`r n_sex[["male"]]` men. `r sex_loss_txt`
`r if (length(in_both)) paste0(length(in_both), if (length(in_both) == 1) " person has" else " people have", " a unit in both plasma and serum (", and_list(lab(both_assays)), "), one in each matrix, so neither model counts them twice.") else "No person has a unit in both plasma and serum."`
By health, `r n_health[["healthy"]]` units are healthy,
`r n_health[["not healthy"]]` not healthy and `r n_health[["not recorded"]]` have
no health recorded. `r levels_txt`
`r if (length(file_assays)) paste0("Where no subject is named, in ", and_list(lab(file_assays)), ", each file is a unit, so repeat files of one person count there as separate units.")`

# The ratio against age

## One model per matrix

The level of the ratio depends on the method as much as on the blood: each
assay's column, instrument and ionisation shift all of its files alike. A term
per assay absorbs that shift, so the slope is estimated within assays and pooled
across them, and a difference between assays is never read as an effect of age.

Plasma and serum get one model each. Serum is what is left once the blood has
clotted, plasma what is left when clotting is blocked, so the ratio need not sit
at the same level in both; and a person sampled in both would count twice in one
model over both matrices. In each matrix, the model, in words: a unit's log2
ratio is the level of its assay, plus a change $b$ for each year of age, plus a
shift for its sex and a shift for its health (healthy, not healthy or not
recorded), plus what the model leaves unexplained. As a formula, for unit $i$ of
one matrix, fitted by least squares:

$$
\log_2\left(\frac{\text{kynurenine}}{\text{tryptophan}}\right)_i
= \text{assay}_i + b \cdot \text{age}_i + \text{sex}_i + \text{health}_i
+ \varepsilon_i
$$

The slope $b$, the change in the log2 ratio per year of age, fitted once in
plasma and once in serum, is the number the hypothesis is about: it predicts
$b > 0$. A matrix measured by one assay only has no assay term: its level is the
model's intercept. Sex and health enter where they vary inside at least one assay
of the matrix. A level found in only some assays is compared inside those; a
level that never varies inside an assay is part of that assay's term. A matrix is
fitted where it has at least `r MIN_UNITS` units spanning at least `r MIN_SPAN`
years of age. The table also holds, for each matrix, the fit on its healthy units
only, where health drops out (see below).

```{r per-matrix}
# A term enters only where it varies inside an assay. Assay comes before health,
# so a level found in one assay only is absorbed by the assay term.
varies    <- function(d, v) any(tapply(d[[v]], d$study, n_distinct) >= 2)
age_terms <- function(d) c("agen", if (varies(d, "sex2")) "sex2",
                           if (n_distinct(d$study) >= 2) "study",
                           if (varies(d, "health")) "health")
# The terms of a fit, in the order of TERM_NAMES.
terms_of <- function(m) TERM_NAMES[intersect(names(TERM_NAMES), attr(terms(m), "term.labels"))]
# Enough units over enough years of age for a slope.
fittable   <- function(d) nrow(d) >= MIN_UNITS && diff(range(d$agen)) >= MIN_SPAN
fit_matrix <- function(d) if (fittable(d)) lm(reformulate(age_terms(d), "l2"), data = d)
fit_row <- function(m, d, mat, label) {
  ci <- confint(m)["agen", ]
  data.frame(matrix = mat, model = label, units = nrow(d), assays = n_distinct(d$study),
             terms = and_list(terms_of(m)),
             slope = coef(m)[["agen"]], lo = ci[[1]], hi = ci[[2]],
             p = summary(m)$coefficients["agen", 4])
}
healthy   <- blood[blood$health == "healthy", ]   # healthy, with a sex recorded
by_mat    <- function(d) setNames(lapply(mats, function(m) d[d$site == m, ]), mats)
units_m   <- by_mat(blood)
healthy_m <- by_mat(healthy)
# One model per matrix on all its units, and one on its healthy units; NULL where
# too few units, or too narrow a span of age, to fit.
m_all     <- lapply(units_m, fit_matrix)
m_healthy <- lapply(healthy_m, fit_matrix)
fit_rows  <- function(ms, ds, label) bind_rows(lapply(names(ms), function(m)
  if (!is.null(ms[[m]])) fit_row(ms[[m]], ds[[m]], m, label)))
pooled <- bind_rows(fit_rows(m_all, units_m, "all units"),
                    fit_rows(m_healthy, healthy_m, "healthy units only"))
write.csv(pooled, "artifacts/ratio_covariate_pooled.csv", row.names = FALSE)
knitr::kable(rounded(pooled), row.names = FALSE,
             col.names = c("matrix", "fit", "units", "assays", "terms", "slope per year",
                           "95% CI low", "95% CI high", "p"),
             caption = "The log2 ratio against age in each matrix, with a term per assay")

# The row of one fit: the matrix and "all units" or "healthy units only".
fit_of <- function(mat, label = "all units") pooled[pooled$matrix == mat & pooled$model == label, ]
direction <- function(r) if (r$lo > 0) {
  "The interval lies above zero: the ratio rises with age, the direction the hypothesis predicts."
} else if (r$hi < 0) {
  "The interval lies below zero: the ratio falls with age, against the hypothesis."
} else {
  "The interval includes zero, so the direction is not resolved."
}
# The shift of one level in a fit, as estimate (95% CI; p); NULL where the assay
# terms absorb it.
shift <- function(m, term) {
  co <- summary(m)$coefficients
  if (!term %in% rownames(co)) return(NULL)
  ci <- confint(m)[term, ]
  sprintf("%+.3f (%.3f to %.3f; p = %s)", co[term, 1], ci[[1]], ci[[2]], fmt_p(co[term, 4]))
}
# What the model of one matrix says: its slope, then the sex and health shifts.
say_matrix <- function(mat) {
  r <- fit_of(mat)
  if (!nrow(r)) return(paste0("In ", mat, ", the ", nrow(units_m[[mat]]),
                              " units are too few, or span too few years, to be fitted."))
  sex_shift <- shift(m_all[[mat]], "sex2male")
  ill_shift <- shift(m_all[[mat]], "healthnot healthy")
  paste0("In ", mat, ", the log2 ratio changes by ", fmt(r),
         " per year of age (slope and 95% confidence interval; p = ", fmt_p(r$p),
         "), from ", r$units, " units in ", r$assays,
         if (r$assays == 1) " assay. " else " assays. ", direction(r),
         if (!is.null(sex_shift)) paste0(" At the same age, in the same assay and of the same health, men differ from women by ", sex_shift, " log2."),
         if (!is.null(ill_shift)) paste0(" Units not healthy differ from healthy ones by ", ill_shift, " log2."))
}
```

`r paste(vapply(mats, say_matrix, ""), collapse = "\n\n")`

## Each assay alone

A slope pooled over assays can be carried by one assay. Each assay is therefore
fitted alone in each matrix where it has at least `r MIN_UNITS` units spanning at
least `r MIN_SPAN` years of age: a least-squares slope of the ratio on age, and
the same slope adjusted for sex and health where these vary inside the assay.
The units are those of the models above.

```{r per-assay}
fit_one <- function(u) {
  if (!fittable(u)) return(NULL)
  adj <- c(sex = if (n_distinct(u$sex2) >= 2) "sex2",
           health = if (n_distinct(u$health) >= 2) "health")
  m   <- lm(l2 ~ agen, data = u)
  co  <- summary(m)$coefficients
  ci  <- confint(m)["agen", ]
  ca  <- summary(lm(reformulate(c("agen", adj), "l2"), data = u))$coefficients
  data.frame(dataset_id = u$study[1], matrix = u$site[1], assay = lab(u$study[1]),
             n = nrow(u), lo_x = min(u$agen), hi_x = max(u$agen),
             slope = co["agen", 1], lo = ci[[1]], hi = ci[[2]], p = co["agen", 4],
             r2 = summary(m)$r.squared,
             adjusted_for = if (length(adj)) and_list(names(adj)) else "none",
             slope_adj = ca["agen", 1], p_adj = ca["agen", 4])
}
# Fit each assay of `d` alone in each matrix, and say where its interval lies
# relative to zero.
fit_each <- function(d) {
  d   <- d[order(match(d$site, MATRICES), d$study, method = "radix"), ]
  key <- paste(d$site, d$study)
  out <- bind_rows(lapply(split(d, factor(key, unique(key))), fit_one))
  if (nrow(out)) out$alone <- ifelse(out$lo > 0, "rise", ifelse(out$hi < 0, "fall", "no"))
  out
}
per_cols <- c("matrix", "assay", "units", "age from", "age to", "slope per year",
              "95% CI low", "95% CI high", "p", "R2", "adjusted for",
              "adjusted slope", "adjusted p", "shown alone")

per_dep <- fit_each(blood)
write.csv(per_dep, "artifacts/ratio_covariate_per_assay.csv", row.names = FALSE)
knitr::kable(rounded(per_dep[, -1]), row.names = FALSE, col.names = per_cols,
             caption = "The log2 ratio against age inside each assay, in each matrix")

# Assays in a matrix, as the text names them: "MSV000085256 (plasma, 59 to 74)",
# with the ages the assay covers there.
pair_txt <- function(mat, s) and_list(sprintf("%s (%s, %s)", lab(s), mat,
  mapply(function(m, x) ages(x, blood[blood$site == m, ]), mat, s)))
said   <- function(r) pair_txt(r$matrix, r$dataset_id)
pick   <- function(r, what) r[r$alone %in% what, ]
rises  <- pick(per_dep, "rise")
falls  <- pick(per_dep, "fall")
others <- pick(per_dep, "no")
pairs  <- unique(blood[order(match(blood$site, MATRICES), blood$study, method = "radix"),
                       c("site", "study")])
unfit  <- pairs[!paste(pairs$site, pairs$study) %in% paste(per_dep$matrix, per_dep$dataset_id), ]
# Assays fitted in both matrices.
twice  <- unique(per_dep$dataset_id[duplicated(per_dep$dataset_id)])
```

`r nrow(rises)` of the `r nrow(per_dep)` assay fit`r if (nrow(per_dep) != 1) "s"`
`r if (nrow(rises) == 1) "shows" else "show"` the rise on
`r if (nrow(rises) == 1) "its" else "their"` own`r if (nrow(rises)) paste0(": ", said(rises))`.
`r if (length(twice)) paste0("An assay that measured both matrices is fitted once in each (", and_list(lab(twice)), ").")`
`r if (nrow(falls)) paste0(said(falls), if (nrow(falls) == 1) " shows" else " show", " a fall with age, against the hypothesis.")`
`r if (nrow(others)) paste0("In ", said(others), ", the interval includes zero.")`
`r if (nrow(unfit)) paste0("Too few units or too narrow a span of age to be fitted alone: ", pair_txt(unfit$site, unfit$study), ".")`
The ages in brackets are those each assay covers in that matrix. The model of a
matrix uses the ages of all its assays together, while each assay keeps its own
level.

The figure has one row per matrix. On the left, every unit of that matrix's
model is shown as its ratio adjusted for the other terms of the model (a
component-plus-residual plot), so the line is the matrix's slope and the band
spans the slopes in its 95% confidence interval. On the right, the slope inside
each assay of that matrix fitted alone, with its interval and the ages it
covers, against the slope of the matrix's model.

```{r age-figure}
#| echo: false
# One figure row for matrix `mat`. Left: units of model `m` on `d`, adjusted ratio,
# slope and 95% band. Right: slopes of assays fitted alone (`per`) vs `est`.
aged_ids  <- sort(unique(aged$study), method = "radix")
assay_col <- setNames(scales::hue_pal()(length(aged_ids)), lab(aged_ids))
age_figure <- function(m, d, per, est, mat, tags, xlim = NULL) {
  b_age <- coef(m)[["agen"]]
  se    <- summary(m)$coefficients["agen", 2]
  x0    <- mean(d$agen)
  y0    <- mean(d$l2)
  d$assay   <- lab(d$study)
  d$partial <- resid(m) + b_age * (d$agen - x0) + y0
  # The band: every slope in the 95% interval, drawn through the mean age.
  tq   <- qt(0.975, df.residual(m))
  xs   <- seq(min(d$agen), max(d$agen), length.out = 100)
  band <- data.frame(x = xs, y = y0 + b_age * (xs - x0),
                     lo = y0 + b_age * (xs - x0) - tq * se * abs(xs - x0),
                     hi = y0 + b_age * (xs - x0) + tq * se * abs(xs - x0))
  adjusted <- setdiff(terms_of(m), "age")
  title_A  <- paste0(tags[1], "   ", mat,
                     if (length(adjusted)) paste(", adjusted for", and_list(adjusted)))

  pA <- ggplot(d, aes(agen, partial)) +
    geom_ribbon(data = band, aes(x = x, ymin = lo, ymax = hi), inherit.aes = FALSE,
                fill = "grey70", alpha = 0.5) +
    geom_point(aes(colour = assay), size = 1.1, alpha = 0.55) +
    scale_colour_manual(values = assay_col) +
    geom_line(data = band, aes(x = x, y = y), inherit.aes = FALSE, linewidth = 0.9) +
    labs(x = "age (years)", y = expression(log[2]*(kynurenine / tryptophan)),
         colour = NULL, title = title_A) +
    theme_bw(base_size = 10) +
    theme(legend.position = "bottom", legend.text = element_text(size = 7),
          panel.grid.minor = element_blank()) +
    guides(colour = guide_legend(ncol = 2, override.aes = list(size = 2, alpha = 1)))

  fb <- rbind(
    if (nrow(per)) data.frame(
      label = sprintf("%s  (n = %d, %g-%g)", per$assay, per$n, per$lo_x, per$hi_x),
      slope = per$slope, lo = per$lo, hi = per$hi, kind = "within one assay"),
    # The chapter counts these units as subjects; the keys keep "units".
    data.frame(label = sprintf("%s model: %s  (n = %d)", mat,
                               sub("units", "subjects", est$model), est$units),
               slope = est$slope, lo = est$lo, hi = est$hi,
               kind = "model of the matrix"))
  fb$label <- factor(fb$label, levels = unique(rev(fb$label)))
  pB <- ggplot(fb, aes(slope, label, colour = kind, shape = kind)) +
    geom_vline(xintercept = 0, linetype = 2, colour = "grey55") +
    geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y", width = 0.25) +
    geom_point(size = 2) +
    scale_colour_manual(values = c("within one assay" = "grey25",
                                   "model of the matrix" = "#B2182B"),
                        breaks = unique(fb$kind), name = NULL) +
    scale_shape_manual(values = c("within one assay" = 16,
                                  "model of the matrix" = 18),
                       breaks = unique(fb$kind), name = NULL) +
    coord_cartesian(xlim = xlim) +
    labs(x = "slope per year of age (log2), 95% CI", y = NULL,
         title = paste0(tags[2], "   ", mat, ", each assay alone")) +
    theme_bw(base_size = 10) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())

  patchwork::wrap_plots(pA, pB, ncol = 2, widths = c(1, 1.15))
}
# The figure: one row per fitted matrix, panels lettered in reading order, all
# slopes on one scale. `label` is "all units" or "healthy units only".
age_rows <- function(ms, ds, per, label) {
  fitted <- names(ms)[!vapply(ms, is.null, TRUE)]
  xlim   <- range(0, per$lo, per$hi, pooled$lo[pooled$model == label],
                  pooled$hi[pooled$model == label])
  patchwork::wrap_plots(lapply(seq_along(fitted), function(i) {
    mat <- fitted[i]
    age_figure(ms[[mat]], ds[[mat]], if (nrow(per)) per[per$matrix == mat, ] else per,
               fit_of(mat, label), mat, LETTERS[2 * i - 1:0], xlim)
  }), ncol = 1)
}
n_rows   <- sum(!vapply(m_all, is.null, TRUE))
n_rows_h <- sum(!vapply(m_healthy, is.null, TRUE))
```

```{r fig-ratio-age, fig.width = 10, fig.height = 5 * max(1, n_rows)}
#| echo: false
if (n_rows) {
  p <- age_rows(m_all, units_m, per_dep, "all units")
  print(p)
  ggsave(figp("ratio_vs_age_by_matrix.png"), p, width = 10, height = 5 * n_rows, dpi = 200)
}
```

## Each assay left out

A pooled slope can also hang on one assay. The model of each matrix is
therefore fitted again once without each of its assays, with the same terms and
rules as above. The table gives the model of each matrix first, then each fit
without one assay.

```{r leave-one-out}
# The model of one matrix, then the same model once without each of its assays
# (left_out is NA for the model itself); a fit whose units are too few is left out.
loo_rows <- function(mat) {
  d <- units_m[[mat]]
  bind_rows(lapply(c(NA_character_, sort(unique(d$study), method = "radix")), function(s) {
    u <- d[!d$study %in% s, ]
    m <- fit_matrix(u)
    assay <- if (is.na(s)) "none" else lab(s)
    if (!is.null(m)) mutate(fit_row(m, u, mat, if (is.na(s)) "all assays" else paste("without", assay)),
                            left_out = s, assay = assay)
  }))
}
fitted_m <- names(m_all)[!vapply(m_all, is.null, TRUE)]
loo <- bind_rows(lapply(fitted_m, loo_rows))
loo$side <- ifelse(loo$lo > 0, "lies above zero", ifelse(loo$hi < 0, "lies below zero", "includes zero"))
# Numbers of this section: three significant digits in the table, two in the text.
sig_n <- function(x, digits = 2) formatC(x, digits = digits, format = "fg", flag = "#")
knitr::kable(data.frame(loo$matrix, loo$assay, loo$units, loo$assays, sig_n(loo$slope, 3),
                        paste(sig_n(loo$lo, 3), "to", sig_n(loo$hi, 3)),
                        formatC(loo$p, digits = 2, format = "g", flag = "#")),
             col.names = c("matrix", "assay left out", "units", "assays", "slope per year",
                           "95% CI", "p"),
             caption = "The model of each matrix, and the same model without each of its assays")

# What leaving out one assay does in one matrix: the range of the slopes, and
# where the interval lies.
say_loo <- function(mat) {
  r     <- loo[loo$matrix == mat, ]
  refit <- r[!is.na(r$left_out), ]
  if (!nrow(refit)) return(paste0("In ", mat, ", no fit is left once an assay is left out."))
  off <- refit[refit$side != "lies above zero", ]
  paste0("In ", mat, ", the slope is ", sig_n(r$slope[is.na(r$left_out)]),
         " per year with all assays and between ", sig_n(min(refit$slope)), " and ",
         sig_n(max(refit$slope)), " with one assay left out; ",
         if (!nrow(off)) paste0("the interval stays above zero in all ", nrow(refit), " of these fits.")
         else paste0("the interval ", paste(unique(off$side), collapse = " or "), " in ", nrow(off),
                     " of these ", nrow(refit), " fits: ", and_list(paste("without", off$assay)), "."))
}
loo_h <- 1.2 + 0.35 * max(1, table(loo$matrix))
```

`r paste(vapply(fitted_m, say_loo, ""), collapse = " ")` In the figure, the
model of each matrix is a square and each fit without one assay a dot, each with
its 95% interval.

```{r fig-leave-one-out, fig.width = 8, fig.height = loo_h}
#| echo: false
# One panel per matrix: the model and each fit without one assay, coloured by
# where the interval lies; dashed at zero, grey at the model's slope.
if (nrow(loo)) {
  fd <- loo
  fd$label    <- factor(fd$model, rev(unique(fd$model[order(!is.na(fd$left_out), fd$left_out,
                                                            method = "radix")])))
  fd$matrix   <- factor(fd$matrix, fitted_m)
  fd$interval <- paste("interval", fd$side)
  fd$main     <- is.na(fd$left_out)
  side_col    <- c("interval lies above zero" = "#2166AC", "interval includes zero" = "#B2182B",
                   "interval lies below zero" = "grey30")
  p_loo <- ggplot(fd, aes(slope, label)) +
    geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
    geom_vline(data = fd[fd$main, ], aes(xintercept = slope), colour = "grey75") +
    geom_errorbar(aes(xmin = lo, xmax = hi, colour = interval), orientation = "y", width = 0.25) +
    geom_point(aes(colour = interval, shape = main), size = 2.2) +
    scale_shape_manual(values = c(`TRUE` = 15, `FALSE` = 16), guide = "none") +
    scale_colour_manual(values = side_col, breaks = intersect(names(side_col), fd$interval),
                        name = NULL) +
    facet_wrap(~ matrix, scales = "free_y", nrow = 1) +
    labs(x = "change in log2 kynurenine/tryptophan per year of age (95% CI)", y = NULL) +
    theme_bw(base_size = 10) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
  print(p_loo)
  ggsave(figp("loo_assay_by_matrix.png"), p_loo, width = 8, height = loo_h, dpi = 200)
  ggsave(figp("loo_assay_by_matrix.pdf"), p_loo, width = 8, height = loo_h)
}
```

## Healthy units only

This section's figure is in the chapter's appendix, with the figure of each
assay left out. Disease can move the ratio on its own, since inflammation induces the
enzyme, and the units that are not healthy are not spread evenly over the
assays or over age. The model of each matrix is therefore repeated on its
healthy units alone, where health drops out.

```{r healthy}
per_h <- fit_each(healthy)
if (nrow(per_h))
  knitr::kable(rounded(per_h[, -1]), row.names = FALSE, col.names = per_cols,
               caption = "Healthy units only: the log2 ratio against age inside each assay, in each matrix")
# Healthy units with no sex: not in these fits.
n_h_nosex <- sum(no_sex$health == "healthy")
h_rises <- pick(per_h, "rise")
h_falls <- pick(per_h, "fall")
# What the healthy units of one matrix say: where they come from, and the slope.
say_healthy <- function(mat) {
  d <- healthy_m[[mat]]
  if (!nrow(d)) return(paste0("In ", mat, ", no unit is healthy."))
  n <- table(factor(d$study, sort(unique(d$study), method = "radix")))
  r <- fit_of(mat, "healthy units only")
  head <- paste0("In ", mat, ", the ", nrow(d), " healthy units come from ", length(n),
                 if (length(n) == 1) " assay (" else " assays (",
                 paste(sprintf("%s: %d", lab(names(n)), n), collapse = ", "), ")")
  if (!nrow(r)) return(paste0(head, ": too few, or spanning too few years, to be fitted."))
  paste0(head, ". On them, the log2 ratio changes by ", fmt(r),
         " per year of age (p = ", fmt_p(r$p), "). ", direction(r))
}
```

`r paste(vapply(mats, say_healthy, ""), collapse = "\n\n")`

`r if (n_h_nosex) paste0("Another ", n_h_nosex, " healthy ", if (n_h_nosex == 1) "unit has no sex and is" else "units have no sex and are", " left out.")`
Of the `r nrow(per_h)` assay fit`r if (nrow(per_h) != 1) "s"` with enough healthy
units, `r nrow(h_rises)` `r if (nrow(h_rises) == 1) "shows" else "show"` the rise
and `r nrow(h_falls)` a fall. The figure is drawn as the one above, on the
healthy units only.

```{r fig-ratio-age-healthy, fig.width = 10, fig.height = 5 * max(1, n_rows_h)}
#| echo: false
if (n_rows_h) {
  p_h <- age_rows(m_healthy, healthy_m, per_h, "healthy units only")
  print(p_h)
  ggsave(figp("ratio_vs_age_healthy.png"), p_h, width = 10, height = 5 * n_rows_h, dpi = 200)
  ggsave(figp("ratio_vs_age_healthy.pdf"), p_h, width = 10, height = 5 * n_rows_h)
}
```

# Step summary

The results of this step, as the chapter's table shows them.

```{r step-summary}
# One value of the fit of a matrix, or "not fitted".
fit_value <- function(mat, label, f) {
  r <- fit_of(mat, label)
  if (nrow(r)) f(r) else "not fitted"
}
per_matrix_items <- function(mat) data.frame(
  item  = paste0(mat, ": ", c(
    "assays in the model",
    "units in the model",
    "change in log2 kynurenine/tryptophan per year of age (95% CI)",
    "p",
    "healthy units",
    "change per year on the healthy units (95% CI)")),
  value = c(n_distinct(units_m[[mat]]$study), nrow(units_m[[mat]]),
            fit_value(mat, "all units", fmt),
            fit_value(mat, "all units", function(r) fmt_p(r$p)),
            nrow(healthy_m[[mat]]),
            fit_value(mat, "healthy units only", fmt)))
summary_step <- rbind(
  data.frame(item  = c("assays set aside (same samples, another set-up)",
                       "people with a unit in both plasma and serum"),
             value = c(sum(!choice$kept), length(in_both))),
  do.call(rbind, lapply(mats, per_matrix_items)),
  data.frame(item  = c("assay fits that show the rise", "assay fits that show a fall"),
             value = c(sprintf("%d of %d", nrow(rises), nrow(per_dep)),
                       sprintf("%d of %d", nrow(falls), nrow(per_dep)))))
write.csv(summary_step, "artifacts/summary_step6.csv", row.names = FALSE)
summary_step
```

# Checks on repeated sampling

These checks are not part of the chapter's results, and the step summary above
does not use them. A unit above takes the median ratio over its person's files.
Where a person was sampled at several time points or visits, the later samples
can follow an illness or a treatment, and the median then mixes them with the
first. Here each person keeps only the files of the first time point, where the
deposit's own metadata record when each sample was taken, and both matrix models
are fitted again.

## First time point only

The order of a person's samples is read only from the deposit's own metadata
table (`dataset_metadata/<deposit>_metadata.tsv`), matched to the per-file table
file by file, never from the file names. MSV000096564 records the date and time
each sample was collected, its time point in step 2
(`artifacts/metadata_entities.csv`). MSV000084556 records the time point in the
subject (PD001_T1, PD001_T2, PD001_T3), which step 2 removes to name the person.
MSV000094097 records the diet phase of each sample in ATTRIBUTE_Diet, a column
step 2 does not map: "Before" is the first time point and "After" the later one.
In these deposits, a unit keeps the files of its earliest time, all of them
where several files share it, and its ratio is again their median. A file with
no time recorded is left out. Units are then formed, and kept, as above.

```{r first-time-point}
# How a deposit's own table orders a person's files: the column it is read from,
# what it records, and how a value becomes a time, earliest first.
step2  <- read.csv("artifacts/metadata_entities.csv", stringsAsFactors = FALSE)
col_of <- function(d, field) step2[[field]][match(d, step2$deposit_id)]
FIRST_TIME <- list(
  # The date and time each sample was collected: step 2's time point.
  MSV000096564 = list(
    column = col_of("MSV000096564", "col_timepoint"),
    how    = "date and time of collection",
    time   = function(x) as.numeric(as.POSIXct(x, format = "%m/%d/%Y %H:%M", tz = "UTC"))),
  # The subject as recorded, which ends in the time point: step 2's subject,
  # before step 2 removes the time point.
  MSV000084556 = list(
    column = col_of("MSV000084556", "col_subject_id"),
    how    = "time point that ends the recorded subject, as in PD001_T2",
    time   = function(x) as_num(sub("^.*_T([0-9]+)$", "\\1", x))),
  # The diet phase of each sample, a column step 2 does not map: "Before" the
  # diet, then "After" it. Any other value is no time.
  MSV000094097 = list(
    column = "ATTRIBUTE_Diet",
    how    = "diet phase, Before then After",
    time   = function(x) unname(c(Before = 1, After = 2)[x])))

# Blood files, each with the value and the time its deposit's own table records
# (NA in the other deposits).
bf <- pf[pf$site %in% MATRICES, ]
bf$deposit  <- datasets$deposit_id[match(bf$study, datasets$dataset_id)]
bf$unit     <- ifelse(is.na(bf$subj), paste("file", bf$study, bf$fk),
                      paste(bf$study, bf$subj, bf$site))
bf$recorded <- NA_character_
bf$time     <- NA_real_
for (d in names(FIRST_TIME)) {
  own <- read.delim(file.path("dataset_metadata", paste0(d, "_metadata.tsv")),
                    stringsAsFactors = FALSE)
  i   <- bf$deposit == d
  bf$recorded[i] <- own[[FIRST_TIME[[d]]$column]][match(bf$fk[i], norm_file_key(own$filename))]
  bf$time[i]     <- FIRST_TIME[[d]]$time(bf$recorded[i])
}
# MSV000084556: the recorded subject, without its time point, is step 2's person.
p84 <- bf$deposit == "MSV000084556"
if (!identical(sub("_T[0-9]+$", "", bf$recorded[p84]), bf$subj[p84]))
  stop("MSV000084556: the recorded subject is not the person step 2 names")
# MSV000094097: files whose diet phase is recorded, and those whose file name
# agrees (Basal before, Day15 after). The order comes from the table only.
p94       <- bf$deposit == "MSV000094097"
n94       <- sum(p94)
n94_diet  <- sum(!is.na(bf$time[p94]))
named94   <- ifelse(grepl("_Basal_", bf$file[p94]), "Before",
                    ifelse(grepl("_Day15_", bf$file[p94]), "After", NA))
n94_agree <- sum(named94 == bf$recorded[p94], na.rm = TRUE)

# In these deposits, a unit keeps the files of its earliest recorded time, and a
# file with no time recorded is left out. Every other unit keeps all its files.
ordered   <- bf$deposit %in% names(FIRST_TIME)
undated   <- ordered & is.na(bf$time)
n_undated <- sum(undated)
earliest  <- ave(bf$time, bf$unit, FUN = function(t) if (all(is.na(t))) NA else min(t, na.rm = TRUE))
first     <- !ordered | (!undated & bf$time == earliest)

# The deposits with repeated samples (a unit of several files): how the first
# time point is found in each, and what the check keeps.
n_files  <- ave(seq_len(nrow(bf)), bf$unit, FUN = length)
repeated <- sort(unique(bf$deposit[n_files > 1]), method = "radix")
as_is    <- setdiff(repeated, names(FIRST_TIME))
time_rows <- bind_rows(lapply(repeated, function(d) {
  i <- bf$deposit == d
  data.frame(deposit = d,
             how     = if (d %in% names(FIRST_TIME))
                         sprintf("%s (%s)", FIRST_TIME[[d]]$how, FIRST_TIME[[d]]$column) else
                         "not read: every file kept, as above",
             people  = n_distinct(bf$subj[i]),
             units   = n_distinct(bf$unit[i]),
             several = sum(table(bf$unit[i]) > 1),
             files   = sum(i),
             kept    = sum(first[i]))
}))
knitr::kable(time_rows, row.names = FALSE,
             col.names = c("deposit", "first time point from", "people", "blood units",
                           "units with several files", "blood files", "files kept"),
             caption = "How the first time point of a person is found, per deposit with repeated samples")

# The sleep study is left unordered: its subject suffix is not labelled a time,
# and its units keep only files labelled healthy (rule by hand above).
if (!identical(as_is, SLEEP_STUDY))
  stop("repeated samples left as they are in ", and_list(as_is), "; the text explains ",
       SLEEP_STUDY, " only")
p82     <- bf$deposit == SLEEP_STUDY
own82   <- read.delim(file.path("dataset_metadata", paste0(SLEEP_STUDY, "_metadata.tsv")),
                      stringsAsFactors = FALSE)
rec82   <- own82$SubjectIdentifierAsRecorded[match(bf$fk[p82], norm_file_key(own82$filename))]
n82     <- sum(p82)
n82_num <- sum(grepl(col_of(SLEEP_STUDY, "const_subject_suffix"), rec82))
if (n82_num != n82) stop(SLEEP_STUDY, ": a recorded subject does not end in a bare number")
if (any(bf$health[p82] != "healthy")) stop(SLEEP_STUDY, ": a file not labelled healthy is in the units")
# The recorded subjects of one person with several files, as an example.
who82 <- sort(unique(bf$subj[p82][duplicated(bf$subj[p82])]))[1]
ex82  <- sort(rec82[bf$subj[p82] == who82])

# Slopes in this section's text, to two significant digits, so that a bound near
# zero keeps its digits.
sig     <- function(x, flag = "#") formatC(x, digits = 2, format = "fg", flag = flag)
fmt_sig <- function(r) sprintf("%s (%s to %s)", sig(r$slope, "#+"), sig(r$lo), sig(r$hi))

# The units of the first time point, filtered as the units of the models above.
fu <- unitise(bf[first, ], fields = "subj")
fu_disagree <- disagree_text(fu, TERM_NAMES)
fu <- fu[is.finite(fu$l2) & is.finite(fu$agen) & fu$agen >= ADULT_MIN_AGE & !is.na(fu$sex2), ]
fu_m <- by_mat(fu)
m_fu <- lapply(fu_m, fit_matrix)
first_pooled <- bind_rows(pooled[pooled$model == "all units", ],
                          fit_rows(m_fu, fu_m, "first time point only"))
first_pooled$model[first_pooled$model == "all units"] <- "all time points (median)"
first_pooled <- first_pooled[order(match(first_pooled$matrix, MATRICES)), ]
knitr::kable(rounded(transform(first_pooled, p = fmt_p(p))), row.names = FALSE,
             col.names = c("matrix", "files of a person", "units", "assays", "terms",
                           "slope per year", "95% CI low", "95% CI high", "p"),
             caption = "The log2 ratio against age in each matrix: all time points and the first only")

# The files set aside (later time points, or no time recorded), the assays whose
# units change, and the units lost because their first time point gives no ratio.
n_later  <- sum(!first & !undated)
later_in <- sort(unique(bf$study[!first & !undated]), method = "radix")
changed  <- sort(unique(bf$study[!first]), method = "radix")
n_lost   <- length(setdiff(blood$unit, fu$unit))
# Units whose health level differs from above: there a unit takes the health all
# its files record, here that of the files of its first time point.
hc <- merge(blood[, c("unit", "study", "health")], fu[, c("unit", "health")],
            by = "unit", suffixes = c("_all", "_first"))
hc <- hc[hc$health_all != hc$health_first, ]
hc_n <- table(sprintf("%s, from %s to %s", lab(hc$study), hc$health_all, hc$health_first))
say_first <- function(mat) {
  a <- fit_of(mat)
  b <- first_pooled[first_pooled$matrix == mat & first_pooled$model == "first time point only", ]
  if (!nrow(b)) return(paste0("In ", mat, ", the units of the first time point are too few to be fitted."))
  paste0("In ", mat, ", on the first time point, the log2 ratio changes by ", fmt_sig(b),
         " per year of age (p = ", fmt_p(b$p), ", ", b$units, " units), against ",
         fmt_sig(a), " (p = ", fmt_p(a$p), ", ", a$units, " units) over all time points. ",
         direction(b))
}
```

`r if (n_undated) paste0(n_undated, if (n_undated == 1) " blood file of " else " blood files of ", and_list(names(FIRST_TIME)), if (n_undated == 1) " has" else " have", " no time recorded and ", if (n_undated == 1) "is" else "are", " left out.") else paste0("Every blood file of ", and_list(names(FIRST_TIME)), " has its time recorded.")`
In MSV000094097, ATTRIBUTE_Diet is filled for
`r if (n94_diet == n94) "all" else paste(n94_diet, "of the")` `r n94` blood
files, and in `r if (n94_agree == n94_diet) "every one of them" else paste(n94_agree, "of them")`
the file name says the same: Basal for "Before", Day15 for "After". Its person
is the one step 2 names, as in the models above.

`r SLEEP_STUDY` is left out of the check, and its units stay as in the models
above: its subject field carries a bare number in all `r n82` of its blood files
kept above (`r and_list(ex82)` for one person), a number its metadata does not
label as a time or a draw, and its people already use only the files its
metadata labels healthy.

`r n_later` blood files of later time points are set aside, in
`r and_list(lab(later_in))`. `r if (n_lost) paste0(n_lost, if (n_lost == 1) " unit of the models drops out" else " units of the models drop out", " because the files of its first time point give no ratio.") else "Every unit of the models keeps a ratio."`
A unit's age, sex and health now come from the files of its first time point
only. `r if (nzchar(fu_disagree)) paste0("These files disagree on age, sex or health in some units, which then lack it: ", fu_disagree, ".") else "No unit's files disagree on them."`
`r if (nrow(hc)) paste0("The health level changes for ", nrow(hc), if (nrow(hc) == 1) " unit (" else " units (", and_list(sprintf("%d in %s", hc_n, names(hc_n))), ").") else "No unit changes health level."`

`r paste(vapply(mats, say_first, ""), collapse = "\n\n")`

The assays whose units change, fitted alone in each matrix, on all their time
points and on the first one only:

```{r first-time-point-per-assay}
per_all   <- per_dep[per_dep$dataset_id %in% changed, ]
per_first <- fit_each(fu[fu$study %in% changed, ])
both_per  <- bind_rows(mutate(per_all, files = "all time points (median)"),
                       mutate(per_first, files = "first time point only"))
both_per  <- both_per[order(match(both_per$matrix, MATRICES), both_per$assay, both_per$files,
                            method = "radix"), ]
knitr::kable(rounded(both_per[, c("matrix", "assay", "files", "n", "lo_x", "hi_x",
                                  "slope", "lo", "hi", "p", "alone")]),
             row.names = FALSE,
             col.names = c("matrix", "assay", "files of a person", "units", "age from",
                           "age to", "slope per year", "95% CI low", "95% CI high", "p",
                           "shown alone"),
             caption = "The log2 ratio against age inside each assay whose units change")
all_rises   <- pick(per_all, "rise")
first_rises <- pick(per_first, "rise")
```

Over all time points, `r nrow(all_rises)` of the `r nrow(per_all)` fits of these
assays `r if (nrow(all_rises) == 1) "shows" else "show"` the rise on
`r if (nrow(all_rises) == 1) "its" else "their"` own`r if (nrow(all_rises)) paste0(": ", said(all_rises))`.
On the first time point only, `r nrow(first_rises)` of `r nrow(per_first)`
`r if (nrow(first_rises) == 1) "does" else "do"``r if (nrow(first_rises)) paste0(": ", said(first_rises))`.

# Check: the area baseline

This check is not part of the chapter's results, and the step summary above
does not use it. Step 5 measures a peak's area above a straight line between its
two boundary scans. Where the scan just outside a boundary is empty, the signal
rises from nothing there, and that line leaves out the part of the peak below
the boundary scan. Step 5 therefore also records the area with the line started
from zero on such a side (`auc0`). Here the model of each matrix, on all its
units and on its healthy units only, is fitted again on log2 ratios built from
that area. The files, the units and the terms stay as above; only the areas
change.

```{r check-baseline}
# The two baselines, as the table names them.
BASELINES <- c("between the boundary scans", "from zero beside an empty scan")
# A per-file table written before step 5 recorded auc0 has no such columns.
has_auc0 <- all(c("par_auc0", "met_auc0") %in% names(pf))
if (has_auc0) {
  # Each file's ratio on auc0, then the units and the fits as above.
  pf0    <- pf
  pf0$l2 <- file_l2(pf0, "auc0")
  n_ratio <- sum(is.finite(pf$l2))
  n_moved <- sum(is.finite(pf$l2) & pf0$l2 != pf$l2)
  u0 <- unitise(pf0, fields = "subj")
  u0 <- u0[is.finite(u0$l2) & u0$site %in% MATRICES & is.finite(u0$agen) &
             u0$agen >= ADULT_MIN_AGE & !is.na(u0$sex2), ]
  # auc0 is missing exactly where the area is, so the units are those of the models.
  stopifnot(setequal(u0$unit, blood$unit))
  units0   <- by_mat(u0)
  healthy0 <- by_mat(u0[u0$health == "healthy", ])
  alt <- bind_rows(fit_rows(lapply(units0, fit_matrix), units0, "all units"),
                   fit_rows(lapply(healthy0, fit_matrix), healthy0, "healthy units only"))
  by_base <- bind_rows(mutate(pooled, baseline = BASELINES[1]),
                       mutate(alt, baseline = BASELINES[2]))
  by_base <- by_base[order(match(by_base$matrix, MATRICES), by_base$model, method = "radix"), ]
  # The slope of one matrix on all its units, on both baselines.
  slope_p  <- function(r) sprintf("%+.3f (%.3f to %.3f; p = %s)", r$slope, r$lo, r$hi, fmt_p(r$p))
  say_base <- function(mat) {
    r <- by_base[by_base$matrix == mat & by_base$model == "all units", ]
    if (nrow(r) < 2) return(paste0("in ", mat, ", too few units to fit"))
    sprintf("by %s in %s with the line between the boundary scans and by %s with the line from zero",
            slope_p(r[1, ]), mat, slope_p(r[2, ]))
  }
  # Where the 95% interval of each fit lies, on each baseline.
  side <- function(lo, hi) ifelse(lo > 0, "above zero", ifelse(hi < 0, "below zero", "across zero"))
  fits <- merge(pooled, alt, by = c("matrix", "model"), suffixes = c("", "0"))
  fits <- fits[order(match(fits$matrix, MATRICES), fits$model, method = "radix"), ]
  fits$side  <- side(fits$lo, fits$hi)
  fits$side0 <- side(fits$lo0, fits$hi0)
  same <- fits$side == fits$side0
  on   <- c("all units" = "on all units", "healthy units only" = "on the healthy units only")
  # The fits whose interval lies the same way on both baselines, grouped by where it lies.
  alike <- fits[same, ]
  grp   <- split(alike$matrix, factor(paste(on[alike$model], alike$side, sep = "|"),
                                      unique(paste(on[alike$model], alike$side, sep = "|"))))
  alike_txt <- paste(vapply(names(grp), function(g) {
    k <- strsplit(g, "|", fixed = TRUE)[[1]]
    sprintf("%s in %s %s", k[2], and_list(grp[[g]]), k[1])
  }, ""), collapse = "; ")
  baseline_txt <- paste0(
    "The line from zero changes the ratio of ", n_moved, " of the ", n_ratio,
    " files that give one. On all units, the log2 ratio changes per year of age ",
    paste(vapply(mats, say_base, ""), collapse = "; "), ". ",
    if (any(same)) paste0("On both baselines, the 95% interval lies ", alike_txt, ". "),
    if (any(!same)) paste0("It lies differently in ",
      and_list(sprintf("%s %s: %s with the line between the boundary scans, %s with the line from zero",
                       fits$matrix[!same], on[fits$model[!same]], fits$side[!same], fits$side0[!same])), "."))
} else {
  baseline_txt <- paste("The per-file table has no area on the second baseline (no par_auc0 or",
                        "met_auc0 column): it was written before step 5 recorded one, so the",
                        "check waits for step 5 to be rendered again.")
}
if (has_auc0)
  knitr::kable(rounded(transform(by_base[, c("matrix", "model", "baseline", "units",
                                             "slope", "lo", "hi", "p")], p = fmt_p(p))),
               row.names = FALSE,
               col.names = c("matrix", "fit", "area baseline", "units", "slope per year",
                             "95% CI low", "95% CI high", "p"),
               caption = "The log2 ratio against age in each matrix, on the two area baselines")
```

`r baseline_txt`

# Session info

```{r session-info}
sessionInfo()
```
