---
title: "MASST Hit Confirmation: RuSirius"
subtitle: "Per-dataset SIRIUS annotation of curated MASST hits (MassIVE)"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
execute:
  warning: false
  message: false
---

# Overview

This document takes the curated MASST hit table produced by
`01-masst-curation.qmd` (DIA datasets and datasets without metadata
already removed) and runs SIRIUS annotation for every remaining
MassIVE dataset.

For each dataset two SIRIUS projects are created:

```
sirius_projects/
  <dataset_id>/
    parent/
    metabolite/
results/
  <dataset_id>/
    parent_results.csv
    metabolite_results.csv
```

Results are written to disk for manual review.

# Version check

```{r}
#| include: false
# This notebook lives in confirmation/ — run it from the repo root so
# source("R/...") and read("artifacts/...") paths resolve.
knitr::opts_knit$set(root.dir = "..")
```

```{r versions}
#source("check-versions.R")
```

# Setup

```{r setup}
library(Spectra)
library(MsBackendMassIVE)
library(MetaboCoreUtils)
library(RuSirius)
library(dplyr)
library(tidyr)
# load_metadata(): the harmonised per-file metadata built in 01. Needed here so
# the SIRIUS cap can be stratified by biofluid rather than silently favouring
# whichever matrix ionises best.
source("R/metadata.R")
```

# Compound targets

```{r compound-spec}
compound_spec <- data.frame(
  role    = c("parent",       "metabolite"),
  name    = c("omeprazole",   "5-hydroxyomeprazole"),
  formula = c("C17H19N3O3S",  "C17H19N3O4S"),
  ikey1   = c("SUBDBMMJDZJVOS", "CMZHQFXXAAIBKE")
)
```

# Load curated hits

`artifacts/hits_curated.csv` is written by `01-masst-curation.qmd`
after the DIA and no-metadata filters have been applied.

```{r load-hits}
hits_curated <- read.csv("artifacts/hits_curated.csv", check.names = FALSE)
```

# Filter to MassIVE datasets

```{r filter-massive}
hits_massive <- hits_curated[
  startsWith(hits_curated$ATTRIBUTE_DatasetAccession, "MSV"), ]

massive_datasets <- unique(hits_massive$ATTRIBUTE_DatasetAccession)
cat(length(massive_datasets), "MassIVE datasets to process\n")
```

# Helper functions

```{r helpers}
ADDUCT_CANDIDATES <- c(
  "[M+H]+",
  "[M+Na]+",
  "[M+K]+",
  "[M+NH4]+",
  "[M-H]-",
  "[M+Cl]-"
)
PPM <- 20
RT_TOL <- 3

# Cap the number of hits sent to SIRIUS, PER (role x biofluid).
#
# Why stratified, and not simply the top-N by cosine per role: the confirmed
# features anchor the retention-time window that `derive_extraction_config()`
# uses for MS1 extraction, and retention time is matrix-dependent. A biofluid
# with no confirmed hit therefore has no anchor, and its files cannot be
# propagated into honestly.
#
# A dataset-wide top-N-by-cosine cap silently decides which biofluids get an
# anchor, because cosine tracks abundance and abundance tracks matrix. On
# MSV000082493 the old cap (top 50 per role, dataset-wide) selected 50 urine and
# 0 plasma for the metabolite, purely because urine spectra score higher. The
# biofluids it dropped then looked like "the compound is absent here", which is
# an artifact of our own selection rather than a fact about the data.
#
# Capping per (role x biofluid) guarantees every biofluid gets an attempt, so a
# biofluid that still ends up with no confirmation is genuine evidence and its
# samples can be excluded on principle rather than by accident.
CAP_PER_ROLE_BIOFLUID <- 25

# Biofluid label for each hit, taken from the harmonised per-file metadata that
# `01` built (col_body_site -> body_site). Never parsed from filenames. Hits whose
# file has no body_site fall into an explicit "unknown" stratum, which still gets
# its own allocation rather than being silently dropped.
hit_biofluid <- function(hits_ds, ds) {
  meta <- tryCatch(load_metadata(ds), error = function(e) NULL)
  if (is.null(meta) || !"body_site" %in% names(meta))
    return(rep("unknown", nrow(hits_ds)))
  key <- basename(as.character(meta$file))
  bs  <- as.character(meta$body_site)
  out <- bs[match(basename(as.character(hits_ds$file_path)), key)]
  ifelse(is.na(out) | !nzchar(out), "unknown", out)
}

infer_adduct <- function(
  obs_mz,
  formula,
  ppm = PPM,
  candidates = ADDUCT_CANDIDATES
) {
  if (is.na(obs_mz) || is.na(formula)) {
    return(NA_character_)
  }
  mass <- calculateMass(formula)
  theo <- vapply(
    candidates,
    function(a) mass2mz(mass, adduct = a)[[1]],
    numeric(1)
  )
  diff_ppm <- abs(obs_mz - theo) / theo * 1e6
  best <- which.min(diff_ppm)
  if (diff_ppm[best] > ppm) NA_character_ else candidates[best]
}

slice_role <- function(hits_ds, sp, sp_file, sp_scan, role) {
  role_formula <- compound_spec$formula[compound_spec$role == role]
  h <- hits_ds[hits_ds$role == role & !is.na(hits_ds$rt), ]
  if (!nrow(h)) {
    return(list())
  }

  lapply(seq_len(nrow(h)), function(i) {
    hi <- h[i, ]
    mz_rng <- hi$Precursor_MZ * c(1 - PPM * 1e-6, 1 + PPM * 1e-6)
    rt_rng <- hi$rt + c(-RT_TOL, RT_TOL)

    sp_f <- filterRt(sp[sp_file == hi$file], rt_rng)
    ms1 <- filterMsLevel(sp_f, 1L)
    ms2 <- filterPrecursorMzRange(filterMsLevel(sp_f, 2L), mz_rng)

    int_at_mz <- vapply(
      peaksData(ms1),
      function(pm) {
        keep <- pm[, "mz"] >= mz_rng[1] & pm[, "mz"] <= mz_rng[2]
        if (any(keep)) max(pm[keep, "intensity"]) else 0
      },
      numeric(1)
    )
    ms1_top <- if (length(ms1) && max(int_at_mz) > 0) {
      ms1[which.max(int_at_mz)]
    } else {
      ms1[integer(0)]
    }

    out <- c(ms1_top, ms2)
    out$adduct <- rep(hi$adduct_inferred, length(out))
    out$group_idx <- rep(i, length(out))
    out
  })
}

run_sirius_role <- function(sp_slices, adduct_list, project_path, study_id) {
  nonempty    <- vapply(sp_slices, length, integer(1)) > 0
  n_dropped   <- sum(!nonempty)
  sp_slices   <- sp_slices[nonempty]
  adduct_list <- adduct_list[nonempty]
  if (n_dropped > 0)
    message("  Dropped ", n_dropped,
            " empty slice(s): MassIVE index out of sync with file content")
  if (!length(sp_slices)) return(data.frame())

  srs <- Sirius(
    projectId = study_id,
    username = Sys.getenv("SIRIUS_USERNAME"),
    path = paste0(getwd(), "/", project_path),
    password = Sys.getenv("SIRIUS_PASSWORD")
  )

  pol   <- ifelse(endsWith(trimws(unlist(adduct_list)), "+"), "pos", "neg")
  first <- TRUE
  for (p in unique(pol)) {
    idx  <- which(pol == p)
    sp_p <- concatenateSpectra(sp_slices[idx])
    srs  <- import(
      sirius = srs,
      spectra = sp_p,
      adduct = adduct_list[idx],
      ms_column_name = "group_idx",
      deleteExistingFeatures = first
    )
    first <- FALSE
  }
  run(
    srs,
    formulaIdParams = formulaIdParam(),
    structureDbSearchParams = structureDbSearchParam(
      structureSearchDbs = "ALL"
    ),
    predictParams = predictParam(),
    wait = TRUE
  )
  res <- results(
    srs,
    result.type = "structureDb",
    return.type = "data.frame",
    topFormula = 15,
    topStructure = 15
  )
  shutdown(srs)
  res
}
```

# Per-dataset SIRIUS annotation

```{r per-dataset-loop}
role_formula <- setNames(compound_spec$formula, compound_spec$role)
hits_rt_list <- list()

for (ds in massive_datasets) {
  message("\n--- Processing ", ds, " ---")
  hits_ds <- hits_massive[hits_massive$ATTRIBUTE_DatasetAccession == ds, ]

  dir_res       <- file.path("results", ds)
  roles_present <- unique(hits_ds$role)
  expected_csv  <- file.path(dir_res, paste0(roles_present, "_results.csv"))
  sirius_done   <- length(expected_csv) && all(file.exists(expected_csv))
  rt_path       <- file.path(dir_res, "hits_rt.csv")
  rt_done       <- file.exists(rt_path) &&
                   all(c("mz", "adduct_inferred", "source_file", "assay") %in%
                       names(read.csv(rt_path, nrows = 1, check.names = FALSE)))

  if (sirius_done && rt_done) {
    message("  Already analyzed, skipping")
    next
  }
  if (sirius_done) message("  SIRIUS results found, rerunning RT mapping only")

  hits_ds$adduct_inferred <- mapply(
    infer_adduct,
    hits_ds$Precursor_MZ,
    role_formula[hits_ds$role],
    USE.NAMES = FALSE
  )
  hits_ds$scan_verified <- NA_character_
  na_adduct <- is.na(hits_ds$adduct_inferred)
  if (any(na_adduct)) {
    hits_ds$scan_verified[na_adduct] <- "no_adduct_match"
    message("  ", sum(na_adduct), " hit(s) dropped: precursor mz unmatched ",
            "to any candidate adduct within ", PPM, " ppm")
    hits_ds <- hits_ds[!na_adduct, ]
  }

  # Cap per (role x biofluid): top-cosine within each stratum, so every biofluid
  # present in the hits gets its own shot at an anchor. See CAP_PER_ROLE_BIOFLUID.
  if (nrow(hits_ds) && is.finite(CAP_PER_ROLE_BIOFLUID)) {
    hits_ds$biofluid <- hit_biofluid(hits_ds, ds)
    before <- table(hits_ds$role, hits_ds$biofluid)
    hits_ds <- do.call(rbind, lapply(
      split(hits_ds, list(hits_ds$role, hits_ds$biofluid), drop = TRUE),
      function(d) if (nrow(d) <= CAP_PER_ROLE_BIOFLUID) d
                  else d[order(-d$Cosine)[seq_len(CAP_PER_ROLE_BIOFLUID)], ]))
    after <- table(hits_ds$role, hits_ds$biofluid)
    message("  Hits per role x biofluid before cap:")
    for (r in rownames(before))
      message("    ", r, ": ", paste(sprintf("%s=%d", colnames(before), before[r, ]),
                                     collapse = "  "))
    message("  Sent to SIRIUS after per-biofluid cap (", CAP_PER_ROLE_BIOFLUID, "/stratum):")
    for (r in rownames(after))
      message("    ", r, ": ", paste(sprintf("%s=%d", colnames(after), after[r, ]),
                                     collapse = "  "))
  }

  dir_par <- file.path("sirius_projects", ds, "parent")
  dir_met <- file.path("sirius_projects", ds, "metabolite")
  dir.create(dir_par, recursive = TRUE, showWarnings = FALSE)
  dir.create(dir_met, recursive = TRUE, showWarnings = FALSE)
  dir.create(dir_res, recursive = TRUE, showWarnings = FALSE)

  wanted <- unique(basename(hits_ds$file_path))
  massive_sync_data_files(ds, fileName = wanted)
  sp       <- Spectra(backendInitialize(MsBackendMassIVE(), massiveId = ds,
                                        offline = TRUE))
  sp_file  <- basename(dataOrigin(sp))
  # Cache filenames are prefixed by the MsBackend (<deposit/zip>_<name>)
  # to avoid cross-dataset basename collisions; reconcile them back to the
  # bare names in `wanted` so the (file, scan) match, the mz-fallback, and
  # slice_role() below key on the same names hits_ds$file uses.
  .u   <- unique(sp_file)
  .map <- setNames(vapply(.u, function(f) {
    hit <- wanted[f == wanted | endsWith(f, paste0("_", wanted))]
    if (length(hit)) hit[which.max(nchar(hit))] else f
  }, character(1)), .u)
  sp_file  <- unname(.map[sp_file])
  sp_scan  <- acquisitionNum(sp)

  hits_ds$file    <- basename(hits_ds$file_path)
  hit_idx         <- match(
    paste(hits_ds$file, hits_ds$scan_id, sep = "::"),
    paste(sp_file,      sp_scan,          sep = "::")
  )
  hits_ds$rt <- rtime(sp)[hit_idx]
  hits_ds$mz <- precursorMz(sp)[hit_idx]
  # Provenance: the ACTUAL cached file (study-ID-prefixed) each rt was read
  # from. If a hit resolves against the wrong deposit's data, this carries
  # the wrong study prefix, making cross-deposit leakage visible instead of
  # hidden behind the (always-correct-by-construction) bare hit filename.
  hits_ds$origin <- basename(dataOrigin(sp))[hit_idx]

  ppm_err  <- abs(hits_ds$mz - hits_ds$Precursor_MZ) / hits_ds$Precursor_MZ * 1e6
  mismatch <- is.na(hit_idx) | is.na(hits_ds$mz) | (!is.na(ppm_err) & ppm_err > PPM)
  hits_ds$scan_verified <- ifelse(mismatch, "mismatch", "ok")

  if (any(mismatch)) {
    message("  ", sum(mismatch), " hit(s) with scan_id/precursorMz mismatch, ",
            "attempting mz fallback")
    for (j in which(mismatch)) {
      hi_j   <- hits_ds[j, ]
      sp_j   <- sp[sp_file == hi_j$file]
      pmz_j  <- precursorMz(sp_j)
      pint_j <- precursorIntensity(sp_j)
      cands  <- which(!is.na(pmz_j) &
                      abs(pmz_j - hi_j$Precursor_MZ) /
                        hi_j$Precursor_MZ * 1e6 < PPM)
      if (!length(cands)) {
        hits_ds$rt[j]            <- NA_real_
        hits_ds$mz[j]            <- NA_real_
        hits_ds$scan_verified[j] <- "mz_not_found"
        message("    row ", j, ": no scan with matching mz, skipping")
      } else {
        # Anchor on the highest-intensity precursor; that scan sits at
        # the apex of the real chromatographic peak. Expanding to N
        # candidates (former behaviour) produced one rt per scan on the
        # peak's tails, which then stretched the cfg window downstream.
        best <- cands[which.max(pint_j[cands])]
        hits_ds$rt[j]            <- rtime(sp_j)[best]
        hits_ds$mz[j]            <- pmz_j[best]
        hits_ds$origin[j]        <- basename(dataOrigin(sp_j))[best]
        hits_ds$scan_verified[j] <- if (length(cands) == 1) "mz_fallback"
                                    else                    "mz_fallback_apex"
        message("    row ", j, ": ", length(cands),
                " cand(s), picked apex at rt = ",
                round(rtime(sp_j)[best], 1), "s")
      }
    }
  }

  rt_valid <- hits_ds[!is.na(hits_ds$rt),
                      c("role", "rt", "mz", "adduct_inferred", "origin", "assay")]
  # source_file now carries the study-ID prefix of the file the rt came
  # from (provenance), so a HILIC hit resolved against RP data would show
  # an "MSV000092757_" prefix here rather than looking identical to a
  # genuine HILIC file.
  names(rt_valid)[names(rt_valid) == "origin"] <- "source_file"
  rt_valid$xcms_fts <- ave(seq_len(nrow(rt_valid)), rt_valid$role,
                           FUN = seq_along)
  hits_rt_list[[ds]] <- rt_valid[, c("role", "xcms_fts", "rt", "mz",
                                     "adduct_inferred", "source_file", "assay")]
  write.csv(hits_rt_list[[ds]], file.path(dir_res, "hits_rt.csv"),
            row.names = FALSE)

  if (!sirius_done) {
    sp_par <- slice_role(hits_ds, sp, sp_file, sp_scan, "parent")
    sp_met <- slice_role(hits_ds, sp, sp_file, sp_scan, "metabolite")

    if (!length(sp_par)) {
      message("  No parent spectra, skipping parent SIRIUS run")
    } else {
      adduct_list_par <- lapply(sp_par, function(s) s$adduct[1])
      res_par <- run_sirius_role(sp_par, adduct_list_par, dir_par, study_id = "parent")
      write.csv(res_par, file.path(dir_res, "parent_results.csv"), row.names = FALSE)
      message("  Parent done: ", nrow(res_par), " rows written")
    }

    if (!length(sp_met)) {
      message("  No metabolite spectra, skipping metabolite SIRIUS run")
    } else {
      adduct_list_met <- lapply(sp_met, function(s) s$adduct[1])
      res_met <- run_sirius_role(sp_met, adduct_list_met, dir_met, study_id = "metabolite")
      write.csv(res_met, file.path(dir_res, "metabolite_results.csv"), row.names = FALSE)
      message("  Metabolite done: ", nrow(res_met), " rows written")
    }
  }
}

message("All datasets processed.")
```

# Evaluate SIRIUS matches

## Helper functions

Each SIRIUS feature is assigned a **confidence level** based on where the
target formula and structure rank in the SIRIUS output:

| Confidence | Formula rank | Structure rank |
|------------|-------------|----------------|
| strict     | 1           | 1              |
| standard   | ≤ 3         | ≤ 5            |
| liberal    | ≤ 15        | ≤ 15           |
| not_confirmed | target formula absent or ranks beyond liberal thresholds | |

```{r resolve-helpers}
library(dplyr)
library(ggplot2)

CONFIDENCE_LEVELS <- c("strict", "standard", "liberal", "not_confirmed")
CONFIDENCE_COLOURS <- c(
  strict      = "#1A7837",
  standard    = "#74C476",
  liberal     = "#FEC44F",
  not_confirmed = "#D7191C"
)

#' Resolve target features in one SIRIUS results table (one dataset, one role)
resolve_features <- function(res,
                             target_formula,
                             target_ikey_prefix) {

  target_rows <- res |>
    filter(molecularFormula == target_formula,
           !is.na(inchiKey),
           startsWith(inchiKey, target_ikey_prefix)) |>
    group_by(xcms_fts) |>
    slice_min(rank.y, n = 1, with_ties = FALSE) |>
    ungroup() |>
    transmute(
      xcms_fts,
      formula_rank   = rank.x,
      formula_score  = round(siriusScoreNormalized, 3),
      structure_rank = rank.y,
      tanimoto       = round(tanimotoSimilarity, 3),
      csi_score      = round(csiScore, 1),
      adduct,
      structure_name = structureName,
      target_present = TRUE
    )

  res |>
    distinct(xcms_fts) |>
    left_join(target_rows, by = "xcms_fts") |>
    mutate(
      target_present = !is.na(formula_rank),
      confidence = case_when(
        formula_rank == 1  & structure_rank == 1  ~ "strict",
        formula_rank <= 3  & structure_rank <= 5  ~ "standard",
        formula_rank <= 15 & structure_rank <= 15 ~ "liberal",
        TRUE                                       ~ "not_confirmed"
      ),
      confidence = factor(confidence, levels = CONFIDENCE_LEVELS)
    ) |>
    arrange(confidence, formula_rank, structure_rank)
}

#' Run resolve_features across all (dataset, role) result CSVs
resolve_all <- function(results_dir = "results", compound_spec) {

  datasets <- list.dirs(results_dir, recursive = FALSE, full.names = FALSE)
  datasets <- datasets[startsWith(datasets, "MSV")]
  all_rows <- list()

  for (ds in datasets) {
    for (i in seq_len(nrow(compound_spec))) {
      role     <- compound_spec$role[i]
      csv_path <- file.path(results_dir, ds, paste0(role, "_results.csv"))
      if (!file.exists(csv_path)) next
      res <- read.csv(csv_path, check.names = FALSE)
      if (!nrow(res)) next

      resolve_features(
        res,
        target_formula     = compound_spec$formula[i],
        target_ikey_prefix = compound_spec$ikey1[i]
      ) |>
        mutate(dataset_id = ds, role = role, .before = xcms_fts) ->
        all_rows[[paste(ds, role, sep = "::")]]
    }
  }

  features <- bind_rows(all_rows)

  summary_tbl <- features |>
    group_by(dataset_id, role) |>
    summarise(
      n_total          = n(),
      n_target_present = sum(target_present, na.rm = TRUE),
      n_strict         = sum(confidence == "strict",      na.rm = TRUE),
      n_standard       = sum(confidence == "standard",    na.rm = TRUE),
      n_liberal        = sum(confidence == "liberal",     na.rm = TRUE),
      .groups = "drop"
    )

  list(features = features, summary = summary_tbl)
}
```

## Run evaluation

```{r resolve-run}
resolved    <- resolve_all(results_dir = "results", compound_spec = compound_spec)
features    <- resolved$features
summary_tbl <- resolved$summary
summary_tbl
```

## Plots

### Confidence levels per dataset and role

```{r plot-confidence, fig.width=9, fig.height=4}
features |>
  count(dataset_id, role, confidence) |>
  ggplot(aes(x = dataset_id, y = n, fill = confidence)) +
  geom_col() +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5),
            size = 3, colour = "white", fontface = "bold") +
  facet_wrap(~role) +
  scale_fill_manual(values = CONFIDENCE_COLOURS, drop = FALSE) +
  labs(x = NULL, y = "Number of features", fill = "Confidence",
       title = "SIRIUS confirmation confidence per dataset") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

### Formula and structure rank coloured by confidence

```{r plot-ranks, fig.width=6, fig.height=5}
features |>
  filter(target_present) |>
  ggplot(aes(x = formula_rank, y = structure_rank,
             colour = confidence, shape = role)) +
  annotate("rect", xmin = 0.5, xmax = 1.5, ymin = 0.5, ymax = 1.5,
           fill = "#1A7837", alpha = 0.08) +
  annotate("rect", xmin = 0.5, xmax = 3.5, ymin = 0.5, ymax = 5.5,
           fill = "#74C476", alpha = 0.06) +
  annotate("rect", xmin = 0.5, xmax = 6, ymin = 0.5, ymax = 5,
           fill = "#FEC44F", alpha = 0.04) +
  geom_jitter(size = 3, width = 0.2, height = 0.2, alpha = 0.9) +
  scale_colour_manual(values = CONFIDENCE_COLOURS, drop = FALSE) +
  scale_x_continuous(breaks = c(1, 3, 5)) +
  scale_y_continuous(breaks = c(1, 5)) +
  labs(x = "Formula rank", y = "Structure rank",
       colour = "Confidence", shape = "Role",
       title = "Rank of target among confirmed features") +
  theme_bw() +
  theme(legend.position = "bottom")
```

### Score scatter coloured by confidence

```{r plot-scores, fig.width=7, fig.height=4}
features |>
  filter(target_present) |>
  ggplot(aes(x = tanimoto, y = formula_score,
             colour = confidence, shape = role)) +
  geom_point(size = 3, alpha = 0.85) +
  scale_colour_manual(values = CONFIDENCE_COLOURS, drop = FALSE) +
  labs(x = "Tanimoto similarity", y = "Normalised formula score",
       colour = "Confidence", shape = "Role",
       title = "Score distribution (target-present features)") +
  theme_bw() +
  theme(legend.position = "bottom")
```


# RT summary

## Join retention times

```{r rt-join}
# Load hits_rt.csv from disk for any deposit the loop above skipped.
# (Partial re-renders are common - e.g. we deleted hits_rt.csv for one
# deposit to force a remap but left the others alone; without this
# top-up the left_join below silently NA-fills source_file/rt/mz for
# every skipped deposit.)
missing <- setdiff(massive_datasets, names(hits_rt_list))
for (ds in missing) {
  p <- file.path("results", ds, "hits_rt.csv")
  if (file.exists(p)) hits_rt_list[[ds]] <- read.csv(p, check.names = FALSE)
}

rt_tbl   <- bind_rows(hits_rt_list, .id = "dataset_id")
features <- features |>
  left_join(rt_tbl, by = c("dataset_id", "role", "xcms_fts"))
```

## RT range per dataset and role

```{r rt-summary-table}
rt_summary <- features |>
  filter(!is.na(rt)) |>
  group_by(dataset_id, role, confidence) |>
  summarise(
    n         = n(),
    rt_min    = round(min(rt), 1),
    rt_median = round(median(rt), 1),
    rt_max    = round(max(rt), 1),
    rt_range  = round(max(rt) - min(rt), 1),
    .groups   = "drop"
  )
rt_summary
```

## RT distribution across datasets

```{r rt-plot, fig.width=9, fig.height=4}
features |>
  filter(!is.na(rt)) |>
  ggplot(aes(x = dataset_id, y = rt,
             colour = confidence, shape = role)) +
  geom_jitter(size = 3, width = 0.15, alpha = 0.85) +
  facet_wrap(~role) +
  scale_colour_manual(values = CONFIDENCE_COLOURS, drop = FALSE) +
  labs(x = NULL, y = "Retention time (s)",
       colour = "Confidence", shape = "Role",
       title = "RT of confirmed features across MassIVE datasets (s)") +
  theme_bw() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

# Write confirmed hits

Hits with `strict`, `standard`, or `liberal` SIRIUS confidence are kept
as MassIVE-confirmed; the `confidence` and `rt` (seconds) columns are
preserved for downstream stratification and RT-window alignment.

```{r write-confirmed}
hits_confirmed_massive <- features |>
  filter(confidence %in% c("strict", "standard", "liberal"))

write.csv(hits_confirmed_massive, "artifacts/hits-confirmed-massive.csv",
          row.names = FALSE)
```


