---
title: "MetaboLights Hit Confirmation: RuSirius"
subtitle: "Per-dataset SIRIUS annotation of curated MASST hits (MetaboLights)"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
execute:
  warning: false
  message: false
---

# Overview

Identical pipeline to `masst-hit-confirmation.qmd`, but restricted to
MetaboLights datasets (`MTBLS*` accessions) and using `MsBackendMetaboLights`
for raw file access.

# Setup

```{r}
#| include: false
# This notebook lives in confirmation/, so run it from the repo root and
# source("R/...") and read("artifacts/...") paths resolve.
knitr::opts_knit$set(root.dir = "..")
```

```{r setup}
library(Spectra)
library(MsBackendMetaboLights)
library(MetaboCoreUtils)
library(RuSirius)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Compound targets

```{r compound-spec}
# compound_spec is defined once in R/masst.R, sourced below. The local copy that
# used to sit here named its InChIKey column `ikey1` while the shared one names it
# `inchikey_block1`, so sourcing silently replaced it and every lookup returned NULL.
```

# Load curated hits

```{r load-hits}
hits_curated <- read.csv("artifacts/hits_curated.csv", check.names = FALSE)
```

# Filter to MetaboLights datasets

```{r filter-mtbls}
hits_mtbls <- hits_curated[
  startsWith(hits_curated$ATTRIBUTE_DatasetAccession, "MTBLS"), ]

mtbls_datasets <- unique(hits_mtbls$ATTRIBUTE_DatasetAccession)
cat(length(mtbls_datasets), "MetaboLights datasets to process\n")
```

# Helper functions

```{r helpers}
# ADDUCT_CANDIDATES is defined once in R/masst.R; the local copy here had drifted.
source("R/masst.R")
PPM    <- 25
RT_TOL <- 3   # seconds around hit RT

# Cap the number of hits per role sent to SIRIUS. The confirmed features are
# used downstream (05, derive_extraction_config) only to compute the per-
# (dataset, role) MEDIAN m/z + RT that anchors the MS1 extraction window. The cap is
# OFF: a capped subset is enough to anchor, but the confirmed set is also the
# downstream evidence of presence, and capping by cosine silently decided which hits
# counted. The is.finite() guard below is what switches it off.
CAP_PER_ROLE <- Inf

infer_adduct <- function(obs_mz, formula,
                         ppm        = PPM,
                         candidates = ADDUCT_CANDIDATES) {
  if (is.na(obs_mz) || is.na(formula)) return(NA_character_)
  mass     <- calculateMass(formula)
  theo     <- vapply(candidates,
                     function(a) mass2mz(mass, adduct = a)[[1]], numeric(1))
  diff_ppm <- abs(obs_mz - theo) / theo * 1e6
  best     <- which.min(diff_ppm)
  if (diff_ppm[best] > ppm) NA_character_ else candidates[best]
}

slice_role <- function(hits_ds, sp, sp_file, role) {
  h <- hits_ds[hits_ds$role == role & !is.na(hits_ds$rt), ]
  if (!nrow(h)) return(list())

  lapply(seq_len(nrow(h)), function(i) {
    hi     <- h[i, ]
    mz_rng <- hi$Precursor_MZ * c(1 - PPM * 1e-6, 1 + PPM * 1e-6)
    rt_rng <- hi$rt + c(-RT_TOL, RT_TOL)

    sp_f <- filterRt(sp[sp_file == hi$file], rt_rng)
    ms1  <- filterMsLevel(sp_f, 1L)
    ms2  <- filterPrecursorMzRange(filterMsLevel(sp_f, 2L), mz_rng)

    int_at_mz <- vapply(peaksData(ms1), function(pm) {
      keep <- pm[, "mz"] >= mz_rng[1] & pm[, "mz"] <= mz_rng[2]
      if (any(keep)) max(pm[keep, "intensity"]) else 0
    }, numeric(1))

    ms1_top <- if (length(ms1) && max(int_at_mz) > 0) {
      ms1[which.max(int_at_mz)]
    } else {
      ms1[integer(0)]
    }

    out            <- c(ms1_top, ms2)
    out$adduct     <- rep(hi$adduct_inferred, length(out))
    out$group_idx  <- rep(i, length(out))
    out
  })
}

run_sirius_role <- function(sp_slices, adduct_list, project_path, study_id) {
  srs <- Sirius(
    projectId = study_id,
    username  = Sys.getenv("SIRIUS_USERNAME"),
    path      = paste0(getwd(), "/", project_path),
    password  = Sys.getenv("SIRIUS_PASSWORD")
  )

  pol   <- ifelse(endsWith(trimws(unlist(adduct_list)), "+"), "pos", "neg")
  first <- TRUE
  for (p in unique(pol)) {
    idx  <- which(pol == p)
    sp_p <- concatenateSpectra(sp_slices[idx])
    srs  <- import(
      sirius                 = srs,
      spectra                = sp_p,
      adduct                 = adduct_list[idx],
      ms_column_name         = "group_idx",
      deleteExistingFeatures = first
    )
    first <- FALSE
  }

  run(
    srs,
    formulaIdParams        = formulaIdParam(),
    structureDbSearchParams = structureDbSearchParam(structureSearchDbs = "ALL"),
    predictParams          = predictParam(),
    wait                   = TRUE
  )
  res <- results(srs,
                 result.type   = "structureDb",
                 return.type   = "data.frame",
                 topFormula    = 15,
                 topStructure  = 15)
  shutdown(srs)
  res
}
```

# Per-dataset SIRIUS annotation

```{r per-dataset-loop}
role_formula  <- setNames(compound_spec$formula, compound_spec$role)
hits_rt_list  <- list()

for (ds in mtbls_datasets) {
  message("\n--- Processing ", ds, " ---")
  hits_ds <- hits_mtbls[hits_mtbls$ATTRIBUTE_DatasetAccession == ds, ]

  dir_res       <- file.path("results", ds)
  roles_present <- unique(hits_ds$role)
  expected_csv  <- file.path(dir_res, paste0(roles_present, "_results.csv"))
  sirius_done   <- length(expected_csv) && all(file.exists(expected_csv))
  rt_path       <- file.path(dir_res, "hits_rt.csv")
  rt_done       <- file.exists(rt_path) &&
                   all(c("mz", "adduct_inferred", "source_file", "assay") %in%
                       names(read.csv(rt_path, nrows = 1, check.names = FALSE)))

  if (sirius_done && rt_done) {
    message("  Already analyzed, skipping")
    next
  }
  if (sirius_done) message("  SIRIUS results found, rerunning RT mapping only")

  hits_ds$adduct_inferred <- mapply(infer_adduct,
                                    hits_ds$Precursor_MZ,
                                    role_formula[hits_ds$role],
                                    USE.NAMES = FALSE)
  hits_ds$scan_verified <- NA_character_
  na_adduct <- is.na(hits_ds$adduct_inferred)
  if (any(na_adduct)) {
    hits_ds$scan_verified[na_adduct] <- "no_adduct_match"
    message("  ", sum(na_adduct), " hit(s) dropped: no adduct match within ",
            PPM, " ppm")
    hits_ds <- hits_ds[!na_adduct, ]
  }

  # Cap to the top-CAP_PER_ROLE highest-cosine hits per role before SIRIUS.
  # Only the confirmed features' median m/z + RT are used downstream, so a
  # capped high-cosine subset is sufficient; this also limits how many raw
  # files are synced + loaded below.
  if (nrow(hits_ds) && is.finite(CAP_PER_ROLE)) {
    hits_ds <- do.call(rbind, lapply(split(hits_ds, hits_ds$role), function(d) {
      if (nrow(d) <= CAP_PER_ROLE) d else d[order(-d$Cosine)[seq_len(CAP_PER_ROLE)], ]
    }))
    n_by_role <- table(hits_ds$role)
    message("  Capped to ", paste(sprintf("%s=%d", names(n_by_role), n_by_role),
            collapse = ", "), " hit(s) for SIRIUS (top cosine)")
  }

  dir_par <- file.path("sirius_projects", ds, "parent")
  dir_met <- file.path("sirius_projects", ds, "metabolite")
  dir_res <- file.path("results", ds)
  dir.create(dir_par, recursive = TRUE, showWarnings = FALSE)
  dir.create(dir_met, recursive = TRUE, showWarnings = FALSE)
  dir.create(dir_res, recursive = TRUE, showWarnings = FALSE)

  wanted <- unique(basename(hits_ds$file_path))
  # Pass assayName so the correct assay's files load. MsBackendMetaboLights
  # (>= 1.7.4, gabri branch) bakes the assay index into the cache filename, so
  # a deposit's pos/neg files with identical basenames no longer collide.
  assay <- unique(hits_ds$assay)
  mtbls_sync_data_files(mtblsId = ds, assayName = assay, fileName = wanted)
  sp <- Spectra(backendInitialize(MsBackendMetaboLights(),
                                  mtblsId   = ds,
                                  assayName = assay,
                                  fileName  = wanted,
                                  offline   = TRUE))
  sp_file <- basename(dataOrigin(sp))
  # Cache filenames are prefixed by the MsBackend (<deposit/zip>_<name>)
  # to avoid cross-dataset basename collisions; reconcile them back to the
  # bare names in `wanted` so the (file, scan) match, the mz-fallback, and
  # slice_role() below key on the same names hits_ds$file uses.
  .u   <- unique(sp_file)
  .map <- setNames(vapply(.u, function(f) {
    hit <- wanted[f == wanted | endsWith(f, paste0("_", wanted))]
    if (length(hit)) hit[which.max(nchar(hit))] else f
  }, character(1)), .u)
  sp_file <- unname(.map[sp_file])
  sp_scan <- acquisitionNum(sp)

  hits_ds$file <- basename(hits_ds$file_path)
  hit_idx      <- match(
    paste(hits_ds$file, hits_ds$scan_id, sep = "::"),
    paste(sp_file,      sp_scan,          sep = "::")
  )
  hits_ds$rt <- rtime(sp)[hit_idx]
  hits_ds$mz <- precursorMz(sp)[hit_idx]
  # Provenance: the ACTUAL cached file (study-ID-prefixed) each rt was read
  # from, so cross-deposit leakage shows a wrong prefix instead of hiding
  # behind the bare hit filename.
  hits_ds$origin <- basename(dataOrigin(sp))[hit_idx]

  ppm_err  <- abs(hits_ds$mz - hits_ds$Precursor_MZ) / hits_ds$Precursor_MZ * 1e6
  mismatch <- is.na(hit_idx) | is.na(hits_ds$mz) |
              (!is.na(ppm_err) & ppm_err > PPM)
  hits_ds$scan_verified <- ifelse(mismatch, "mismatch", "ok")

  if (any(mismatch)) {
    message("  ", sum(mismatch), " hit(s) with scan_id/precursorMz mismatch")

    for (j in which(mismatch)) {
      hi_j   <- hits_ds[j, ]
      sp_j   <- sp[sp_file == hi_j$file]
      pmz_j  <- precursorMz(sp_j)
      pint_j <- precursorIntensity(sp_j)
      cands  <- which(!is.na(pmz_j) &
                      abs(pmz_j - hi_j$Precursor_MZ) /
                        hi_j$Precursor_MZ * 1e6 < PPM)

      if (!length(cands)) {
        hits_ds$rt[j]            <- NA_real_
        hits_ds$mz[j]            <- NA_real_
        hits_ds$scan_verified[j] <- "mz_not_found"
        message("    row ", j, ": no matching mz, skipping")
      } else {
        best <- cands[which.max(pint_j[cands])]
        hits_ds$rt[j]            <- rtime(sp_j)[best]
        hits_ds$mz[j]            <- pmz_j[best]
        hits_ds$origin[j]        <- basename(dataOrigin(sp_j))[best]
        hits_ds$scan_verified[j] <- if (length(cands) == 1) "mz_fallback"
                                    else                    "mz_fallback_apex"
        message("    row ", j, ": ", length(cands),
                " cand(s), picked apex at rt = ",
                round(rtime(sp_j)[best], 1), "s")
      }
    }
  }

  rt_valid <- hits_ds[!is.na(hits_ds$rt),
                      c("role", "rt", "mz", "adduct_inferred", "origin", "assay")]
  # source_file carries the study-ID prefix of the file the rt came from
  # (provenance) so cross-deposit leakage is visible at a glance.
  names(rt_valid)[names(rt_valid) == "origin"] <- "source_file"
  rt_valid$xcms_fts <- ave(seq_len(nrow(rt_valid)), rt_valid$role,
                           FUN = seq_along)
  hits_rt_list[[ds]] <- rt_valid[, c("role", "xcms_fts", "rt", "mz",
                                     "adduct_inferred", "source_file", "assay")]
  write.csv(hits_rt_list[[ds]], file.path(dir_res, "hits_rt.csv"),
            row.names = FALSE)

  if (!sirius_done) {
    sp_par <- slice_role(hits_ds, sp, sp_file, "parent")
    sp_met <- slice_role(hits_ds, sp, sp_file, "metabolite")

    if (!length(sp_par)) {
      message("  No parent spectra, skipping parent SIRIUS run")
    } else {
      adduct_list_par <- lapply(sp_par, function(s) s$adduct[1])
      res_par <- run_sirius_role(sp_par, adduct_list_par,
                                 dir_par, study_id = "parent")
      write.csv(res_par, file.path(dir_res, "parent_results.csv"),
                row.names = FALSE)
      message("  Parent done: ", nrow(res_par), " rows")
    }

    if (!length(sp_met)) {
      message("  No metabolite spectra, skipping metabolite SIRIUS run")
    } else {
      adduct_list_met <- lapply(sp_met, function(s) s$adduct[1])
      res_met <- run_sirius_role(sp_met, adduct_list_met,
                                 dir_met, study_id = "metabolite")
      write.csv(res_met, file.path(dir_res, "metabolite_results.csv"),
                row.names = FALSE)
      message("  Metabolite done: ", nrow(res_met), " rows")
    }
  }
}

message("All MetaboLights datasets processed.")
```

# Evaluate SIRIUS matches

## Helper functions

```{r resolve-helpers}
CONFIDENCE_LEVELS  <- c("strict", "standard", "liberal", "not_confirmed")
CONFIDENCE_COLOURS <- c(
  strict        = "#1A7837",
  standard      = "#74C476",
  liberal       = "#FEC44F",
  not_confirmed = "#D7191C"
)

resolve_features <- function(res, target_formula, target_ikey_prefix) {
  target_rows <- res |>
    filter(molecularFormula == target_formula,
           !is.na(inchiKey),
           startsWith(inchiKey, target_ikey_prefix)) |>
    group_by(xcms_fts) |>
    slice_min(rank.y, n = 1, with_ties = FALSE) |>
    ungroup() |>
    transmute(
      xcms_fts,
      formula_rank   = rank.x,
      formula_score  = round(siriusScoreNormalized, 3),
      structure_rank = rank.y,
      tanimoto       = round(tanimotoSimilarity, 3),
      csi_score      = round(csiScore, 1),
      adduct,
      structure_name = structureName,
      target_present = TRUE
    )

  res |>
    distinct(xcms_fts) |>
    left_join(target_rows, by = "xcms_fts") |>
    mutate(
      target_present = !is.na(formula_rank),
      confidence = case_when(
        formula_rank == 1  & structure_rank == 1  ~ "strict",
        formula_rank <= 3  & structure_rank <= 5  ~ "standard",
        formula_rank <= 15 & structure_rank <= 15 ~ "liberal",
        TRUE                                       ~ "not_confirmed"
      ),
      confidence = factor(confidence, levels = CONFIDENCE_LEVELS)
    ) |>
    arrange(confidence, formula_rank, structure_rank)
}

resolve_all <- function(results_dir = "results", compound_spec) {
  datasets <- list.dirs(results_dir, recursive = FALSE, full.names = FALSE)
  datasets <- datasets[startsWith(datasets, "MTBLS")]
  all_rows <- list()

  for (ds in datasets) {
    for (i in seq_len(nrow(compound_spec))) {
      role     <- compound_spec$role[i]
      csv_path <- file.path(results_dir, ds, paste0(role, "_results.csv"))
      if (!file.exists(csv_path)) next
      res <- read.csv(csv_path, check.names = FALSE)
      if (!nrow(res)) next

      resolve_features(res,
                       target_formula     = compound_spec$formula[i],
                       target_ikey_prefix = compound_spec$inchikey_block1[i]) |>
        mutate(dataset_id = ds, role = role, .before = xcms_fts) ->
        all_rows[[paste(ds, role, sep = "::")]]
    }
  }

  features <- bind_rows(all_rows)
  summary_tbl <- features |>
    group_by(dataset_id, role) |>
    summarise(n_total          = n(),
              n_target_present = sum(target_present, na.rm = TRUE),
              n_strict         = sum(confidence == "strict",   na.rm = TRUE),
              n_standard       = sum(confidence == "standard", na.rm = TRUE),
              n_liberal        = sum(confidence == "liberal",  na.rm = TRUE),
              .groups = "drop")

  list(features = features, summary = summary_tbl)
}
```

## Run evaluation

```{r resolve-run}
resolved    <- resolve_all(results_dir = "results", compound_spec = compound_spec)
features    <- resolved$features
summary_tbl <- resolved$summary
summary_tbl
```

## Plots

### Confidence levels per dataset and role

```{r plot-confidence, fig.width=9, fig.height=4}
features |>
  count(dataset_id, role, confidence) |>
  ggplot(aes(x = dataset_id, y = n, fill = confidence)) +
  geom_col() +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5),
            size = 3, colour = "white", fontface = "bold") +
  facet_wrap(~role) +
  scale_fill_manual(values = CONFIDENCE_COLOURS, drop = FALSE) +
  labs(x = NULL, y = "Number of features", fill = "Confidence",
       title = "SIRIUS confirmation confidence: MetaboLights") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

### Formula and structure rank

```{r plot-ranks, fig.width=6, fig.height=5}
features |>
  filter(target_present) |>
  ggplot(aes(x = formula_rank, y = structure_rank,
             colour = confidence, shape = role)) +
  annotate("rect", xmin = 0.5, xmax = 1.5,  ymin = 0.5, ymax = 1.5,
           fill = "#1A7837", alpha = 0.08) +
  annotate("rect", xmin = 0.5, xmax = 3.5,  ymin = 0.5, ymax = 5.5,
           fill = "#74C476", alpha = 0.06) +
  annotate("rect", xmin = 0.5, xmax = 9, ymin = 0.5, ymax = 5,
           fill = "#FEC44F", alpha = 0.04) +
  geom_jitter(size = 3, width = 0.2, height = 0.2, alpha = 0.9) +
  scale_colour_manual(values = CONFIDENCE_COLOURS, drop = FALSE) +
  scale_x_continuous(breaks = c(1, 3, 5)) +
  scale_y_continuous(breaks = c(1, 5)) +
  labs(x = "Formula rank", y = "Structure rank",
       colour = "Confidence", shape = "Role",
       title = "Rank of target: MetaboLights") +
  theme_bw() +
  theme(legend.position = "bottom")
```

### Score scatter

```{r plot-scores, fig.width=7, fig.height=4}
features |>
  filter(target_present) |>
  ggplot(aes(x = tanimoto, y = formula_score,
             colour = confidence, shape = role)) +
  geom_point(size = 3, alpha = 0.85) +
  scale_colour_manual(values = CONFIDENCE_COLOURS, drop = FALSE) +
  labs(x = "Tanimoto similarity", y = "Normalised formula score",
       colour = "Confidence", shape = "Role",
       title = "Score distribution: MetaboLights") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# RT summary

## Join retention times

Map `xcms_fts` back to the RT recorded during scan verification; within each
dataset × role it equals the hit's position in the RT-valid subset, the same
ordering `slice_role` used when stamping `group_idx`.

```{r rt-join}
# Load hits_rt.csv from disk for any deposit the loop above skipped.
# Prevents partial-render NA-fills in the left_join below.
missing <- setdiff(mtbls_datasets, names(hits_rt_list))
for (ds in missing) {
  p <- file.path("results", ds, "hits_rt.csv")
  if (file.exists(p)) hits_rt_list[[ds]] <- read.csv(p, check.names = FALSE)
}

rt_tbl <- bind_rows(hits_rt_list, .id = "dataset_id")

features <- features |>
  left_join(rt_tbl, by = c("dataset_id", "role", "xcms_fts"))
```

## RT range per dataset and role

```{r rt-summary-table}
rt_summary <- features |>
  filter(!is.na(rt)) |>
  group_by(dataset_id, role, confidence) |>
  summarise(
    n         = n(),
    rt_min    = round(min(rt), 1),
    rt_median = round(median(rt), 1),
    rt_max    = round(max(rt), 1),
    rt_range  = round(max(rt) - min(rt), 1),
    .groups   = "drop"
  )
rt_summary
```

## RT distribution across datasets

```{r rt-plot, fig.width=9, fig.height=4}
features |>
  filter(!is.na(rt)) |>
  ggplot(aes(x = dataset_id, y = rt,
             colour = confidence, shape = role)) +
  geom_jitter(size = 3, width = 0.15, alpha = 0.85) +
  facet_wrap(~role) +
  scale_colour_manual(values = CONFIDENCE_COLOURS, drop = FALSE) +
  labs(x = NULL, y = "Retention time (s)",
       colour = "Confidence", shape = "Role",
       title = "RT of confirmed features across MetaboLights datasets (s)") +
  theme_bw() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

# Write confirmed hits

Hits with `strict`, `standard`, or `liberal` SIRIUS confidence are kept
as MetaboLights-confirmed; the `confidence` and `rt` (seconds) columns are
preserved for downstream stratification and RT-window alignment.

```{r write-confirmed}
hits_confirmed_mtbls <- features |>
  filter(confidence %in% c("strict", "standard", "liberal"))

write.csv(hits_confirmed_mtbls, "artifacts/hits-confirmed-metaboLights.csv",
          row.names = FALSE)
```
