---
title: "Metabolomics Workbench Hit Confirmation: RuSirius"
subtitle: "Per-assay SIRIUS annotation of curated MASST hits (Metabolomics Workbench)"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
---

# Overview

Identical pipeline to `massIVE-hit-confirmation.qmd`, but restricted to
Metabolomics Workbench deposits (`ST*` accessions) and using
`MsBackendMetabolomicsWorkbench` for raw file access. A Workbench deposit often
holds several assays of the same samples, each in its own files, so SIRIUS runs
once per assay, and each assay keeps its own results
(`results/<dataset_id>/`) and SIRIUS projects (`sirius_projects/<dataset_id>/`).

# Setup

```{r}
#| include: false
# This notebook lives in confirmation/; root.dir ".." makes its paths resolve
# from application/<name>/.
knitr::opts_knit$set(root.dir = "..")
```

```{r setup}
library(Spectra)
library(MsBackendMetabolomicsWorkbench)
library(MetaboCoreUtils)
library(RuSirius)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Load curated hits

```{r load-hits}
hits_curated <- read.csv("artifacts/hits_curated.csv", check.names = FALSE)
```

# Filter to Metabolomics Workbench deposits

```{r filter-mwb}
hits_mwb <- hits_curated[
  startsWith(hits_curated$ATTRIBUTE_DatasetAccession, "ST"), ]

mwb_datasets <- unique(hits_mwb$dataset_id)
cat(length(mwb_datasets), "Metabolomics Workbench datasets to process\n")
```

# Helper functions

```{r helpers}
# compound_spec comes from R/masst.R; the adduct of each hit was set in step 1.
source("../../R/masst.R")
PPM    <- 25
RT_TOL <- 3   # seconds around hit RT

# Cap on the hits per role sent to SIRIUS, off by default (CONFIRM_CAP=Inf): the
# confirmed set is also the evidence of presence, and a cap by cosine decides what counts.
CAP_PER_ROLE <- {
  .v <- Sys.getenv("CONFIRM_CAP", "Inf")
  if (identical(tolower(.v), "inf")) Inf else as.numeric(.v)
}

slice_role <- function(hits_ds, sp, sp_file, role) {
  h <- hits_ds[hits_ds$role == role & !is.na(hits_ds$rt), ]
  if (!nrow(h)) return(list())

  # Subset the deposit's spectra once per file, not once per feature.
  sp_by_file <- stats::setNames(
    lapply(unique(h$file), function(f) sp[sp_file == f]), unique(h$file))

  lapply(seq_len(nrow(h)), function(i) {
    hi     <- h[i, ]
    mz_rng <- hi$Precursor_MZ * c(1 - PPM * 1e-6, 1 + PPM * 1e-6)
    rt_rng <- hi$rt + c(-RT_TOL, RT_TOL)

    sp_f <- filterRt(sp_by_file[[hi$file]], rt_rng)
    ms1  <- filterMsLevel(sp_f, 1L)
    ms2  <- filterPrecursorMzRange(filterMsLevel(sp_f, 2L), mz_rng)
    # Drop MS2 scans with no peaks: they carry nothing and crash RuSirius at import.
    ms2 <- ms2[lengths(ms2) > 0]

    int_at_mz <- vapply(peaksData(ms1), function(pm) {
      keep <- pm[, "mz"] >= mz_rng[1] & pm[, "mz"] <= mz_rng[2]
      if (any(keep)) max(pm[keep, "intensity"]) else 0
    }, numeric(1))

    ms1_top <- if (length(ms1) && max(int_at_mz) > 0) {
      ms1[which.max(int_at_mz)]
    } else {
      ms1[integer(0)]
    }

    out           <- c(ms1_top, ms2)
    out$adduct    <- rep(hi$adduct_inferred, length(out))
    out$group_idx <- rep(i, length(out))
    out
  })
}

# Each role runs in batches of CONFIRM_BATCH features (100 unless set, Inf for one piece),
# each checkpointed so a restart never redoes a finished batch; the results do not change.
BATCH_SIZE <- {
  .v <- Sys.getenv("CONFIRM_BATCH", "100")
  if (identical(tolower(.v), "inf")) Inf else as.numeric(.v)
}

# The wall time of each SIRIUS run, downloads excluded, for artifacts/sirius_times.csv.
timed_sirius <- function(sp_slices, adduct_list, project_path, study_id) {
  t0  <- Sys.time()
  res <- run_sirius_once(sp_slices, adduct_list, project_path, study_id)
  f   <- "artifacts/sirius_times.csv"
  row <- data.frame(date = format(t0, "%Y-%m-%d %H:%M"), run = project_path,
                    features = length(sp_slices),
                    seconds = round(as.numeric(difftime(Sys.time(), t0, units = "secs")), 1))
  write.table(row, f, sep = ",", row.names = FALSE, col.names = !file.exists(f),
              append = file.exists(f))
  res
}

run_sirius_role <- function(sp_slices, adduct_list, project_path, study_id) {
  if (!is.finite(BATCH_SIZE))
    return(timed_sirius(sp_slices, adduct_list, project_path, study_id))
  n <- length(sp_slices)
  if (!n) return(data.frame())
  starts <- seq(1, n, by = BATCH_SIZE)
  ids_in <- function(idx)
    as.character(unique(unlist(lapply(sp_slices[idx], function(s) s$group_idx))))
  dir.create(project_path, recursive = TRUE, showWarnings = FALSE)
  parts <- vector("list", length(starts))
  for (b in seq_along(starts)) {
    idx  <- starts[b]:min(starts[b] + BATCH_SIZE - 1, n)
    ckpt <- file.path(project_path, sprintf("batch%03d_results.csv", b))
    if (file.exists(ckpt)) {
      prev <- tryCatch(read.csv(ckpt, colClasses = "character"),
                       error = function(e) data.frame())
      # Reused only if every feature in it belongs to this batch; a checkpoint
      # from a differently cut run is recomputed rather than trusted.
      if (!nrow(prev) || all(prev$xcms_fts %in% ids_in(idx))) {
        message("  ", study_id, " batch ", b, "/", length(starts), ": checkpoint reused")
        parts[[b]] <- prev
        next
      }
    }
    message("  ", study_id, " batch ", b, "/", length(starts),
            ": features ", min(idx), "-", max(idx))
    # The batch's project directory must exist before Sirius() opens it, or import() fails
    # later with an unrelated error.
    bdir <- file.path(project_path, sprintf("batch%03d", b))
    dir.create(bdir, recursive = TRUE, showWarnings = FALSE)
    res_b <- timed_sirius(sp_slices[idx], adduct_list[idx], bdir,
                             sprintf("%s_b%03d", study_id, b))
    write.csv(res_b, ckpt, row.names = FALSE)
    parts[[b]] <- tryCatch(read.csv(ckpt, colClasses = "character"),
                           error = function(e) data.frame())
  }
  parts <- parts[vapply(parts, nrow, integer(1)) > 0]
  if (!length(parts)) return(data.frame())
  dplyr::bind_rows(parts)
}

run_sirius_once <- function(sp_slices, adduct_list, project_path, study_id) {
  # A feature with no MS2 left cannot be annotated (RuSirius would import it at mass 0),
  # so it is not sent.
  has_ms2     <- vapply(sp_slices, function(s) any(msLevel(s) >= 2L), logical(1))
  if (any(!has_ms2))
    message("  ", sum(!has_ms2), " feature(s) without MS2 not sent to SIRIUS")
  sp_slices   <- sp_slices[has_ms2]
  adduct_list <- adduct_list[has_ms2]
  if (!length(sp_slices)) return(data.frame())

  # No credentials are passed: SIRIUS uses the GUI's cached login, which a failed Login()
  # would wipe. Wait a little for the login, and stop if it does not come.
  srs <- Sirius(projectId = study_id, path = paste0(getwd(), "/", project_path))
  logged_in <- FALSE
  for (i in 1:6) {
    logged_in <- isTRUE(tryCatch(srs@api$login_and_account_api$IsLoggedIn(),
                                 error = function(e) FALSE))
    if (logged_in) break
    Sys.sleep(10)
  }
  if (!logged_in) {
    shutdown(srs)
    stop("SIRIUS is not logged in: log in through the SIRIUS GUI, then render again.")
  }

  pol   <- ifelse(endsWith(trimws(unlist(adduct_list)), "+"), "pos", "neg")
  first <- TRUE
  for (p in unique(pol)) {
    idx  <- which(pol == p)
    sp_p <- concatenateSpectra(sp_slices[idx])
    srs  <- import(
      sirius                 = srs,
      spectra                = sp_p,
      adduct                 = adduct_list[idx],
      ms_column_name         = "group_idx",
      deleteExistingFeatures = first
    )
    first <- FALSE
  }

  run(
    srs,
    formulaIdParams         = formulaIdParam(),
    structureDbSearchParams = structureDbSearchParam(structureSearchDbs = "ALL"),
    predictParams           = predictParam(),
    wait                    = TRUE
  )
  res <- results(srs,
                 result.type  = "structureDb",
                 return.type  = "data.frame",
                 topFormula   = 15,
                 topStructure = 15)
  shutdown(srs)
  res
}
```

# Per-assay SIRIUS annotation

```{r per-dataset-loop}
hits_rt_list <- list()

for (ds in mwb_datasets) {
  message("\n--- Processing ", ds, " ---")
  hits_ds <- hits_mwb[hits_mwb$dataset_id == ds, ]
  dep     <- hits_ds$ATTRIBUTE_DatasetAccession[1]

  dir_res       <- file.path("results", ds)
  roles_present <- unique(hits_ds$role)
  expected_csv  <- file.path(dir_res, paste0(roles_present, "_results.csv"))
  sirius_done   <- length(expected_csv) && all(file.exists(expected_csv))
  rt_path       <- file.path(dir_res, "hits_rt.csv")
  rt_done       <- file.exists(rt_path) &&
                   all(c("mz", "adduct_inferred", "source_file", "assay") %in%
                       names(read.csv(rt_path, nrows = 1, check.names = FALSE)))

  if (sirius_done && rt_done) {
    message("  Already analyzed, skipping")
    next
  }
  if (sirius_done) message("  SIRIUS results found, rerunning RT mapping only")

  # A finite CAP_PER_ROLE keeps the highest-cosine hits per role, which also limits the
  # raw files synced and loaded below.
  if (nrow(hits_ds) && is.finite(CAP_PER_ROLE)) {
    hits_ds <- do.call(rbind, lapply(split(hits_ds, hits_ds$role), function(d) {
      if (nrow(d) <= CAP_PER_ROLE) d else d[order(-d$Cosine)[seq_len(CAP_PER_ROLE)], ]
    }))
    n_by_role <- table(hits_ds$role)
    message("  Capped to ", paste(sprintf("%s=%d", names(n_by_role), n_by_role),
            collapse = ", "), " hit(s) for SIRIUS (top cosine)")
  }

  for (r in compound_spec$role)
    dir.create(file.path("sirius_projects", ds, r), recursive = TRUE,
               showWarnings = FALSE)
  dir.create(dir_res, recursive = TRUE, showWarnings = FALSE)

  wanted <- unique(basename(hits_ds$file_path))
  mwb_sync_data_files(mwbId = dep, fileName = wanted)

  sp      <- Spectra(backendInitialize(MsBackendMetabolomicsWorkbench(),
                                       mwbId    = dep,
                                       fileName = wanted,
                                       offline  = TRUE))
  sp_file <- basename(dataOrigin(sp))
  # The MsBackend prefixes cached file names (<deposit/zip>_<name>); map them back to the
  # bare names in `wanted`, the names hits_ds$file uses.
  .u   <- unique(sp_file)
  .map <- setNames(vapply(.u, function(f) {
    hit <- wanted[f == wanted | endsWith(f, paste0("_", wanted))]
    if (length(hit)) hit[which.max(nchar(hit))] else f
  }, character(1)), .u)
  sp_file <- unname(.map[sp_file])
  sp_scan <- acquisitionNum(sp)

  hits_ds$file <- basename(hits_ds$file_path)
  hit_idx      <- match(
    paste(hits_ds$file, hits_ds$scan_id, sep = "::"),
    paste(sp_file,      sp_scan,          sep = "::")
  )
  hits_ds$rt <- rtime(sp)[hit_idx]
  hits_ds$mz <- precursorMz(sp)[hit_idx]
  # Provenance: the cached file (study/zip-prefixed) each rt was read from, so a hit
  # resolved against another deposit's data shows its prefix.
  hits_ds$origin <- basename(dataOrigin(sp))[hit_idx]

  ppm_err  <- abs(hits_ds$mz - hits_ds$Precursor_MZ) / hits_ds$Precursor_MZ * 1e6
  mismatch <- is.na(hit_idx) | is.na(hits_ds$mz) |
              (!is.na(ppm_err) & ppm_err > PPM)
  hits_ds$scan_verified <- ifelse(mismatch, "mismatch", "ok")

  if (any(mismatch)) {
    message("  ", sum(mismatch), " hit(s) with scan_id/precursorMz mismatch")

    for (j in which(mismatch)) {
      hi_j   <- hits_ds[j, ]
      sp_j   <- sp[sp_file == hi_j$file]
      pmz_j  <- precursorMz(sp_j)
      pint_j <- precursorIntensity(sp_j)
      cands  <- which(!is.na(pmz_j) &
                      abs(pmz_j - hi_j$Precursor_MZ) /
                        hi_j$Precursor_MZ * 1e6 < PPM)

      if (!length(cands)) {
        hits_ds$rt[j]            <- NA_real_
        hits_ds$mz[j]            <- NA_real_
        hits_ds$scan_verified[j] <- "mz_not_found"
        message("    row ", j, ": no matching mz, skipping")
      } else {
        best <- cands[which.max(pint_j[cands])]
        hits_ds$rt[j]            <- rtime(sp_j)[best]
        hits_ds$mz[j]            <- pmz_j[best]
        hits_ds$origin[j]        <- basename(dataOrigin(sp_j))[best]
        hits_ds$scan_verified[j] <- if (length(cands) == 1) "mz_fallback"
                                    else                    "mz_fallback_apex"
        message("    row ", j, ": ", length(cands),
                " cand(s), picked apex at rt = ",
                round(rtime(sp_j)[best], 1), "s")
      }
    }
  }

  # Keep one row per distinct scan: many library spectra match one scan and the copies get
  # the same annotation. Precursor_MZ is the library's, so it is left out of the key.
  ok <- !is.na(hits_ds$rt)
  if (any(ok)) {
    dup <- duplicated(hits_ds[ok, c("role", "file", "origin", "rt", "mz",
                                    "adduct_inferred", "assay")])
    if (any(dup)) {
      message("  ", sum(dup), " duplicate row(s) collapsed; ",
              sum(ok) - sum(dup), " distinct scan(s) go to SIRIUS")
      hits_ds <- hits_ds[-which(ok)[dup], ]
    }
  }

  rt_valid <- hits_ds[!is.na(hits_ds$rt),
                      c("role", "rt", "mz", "adduct_inferred", "origin", "assay")]
  # source_file carries the study/zip prefix of the file the rt came from
  # (provenance) so cross-deposit leakage is visible at a glance.
  names(rt_valid)[names(rt_valid) == "origin"] <- "source_file"
  rt_valid$xcms_fts <- ave(seq_len(nrow(rt_valid)), rt_valid$role,
                           FUN = seq_along)
  # Stored SIRIUS results are numbered by this table's rows: stop if those rows changed
  # (fewer, other or reordered scans) rather than attach results to the wrong spectra.
  rt_old <- if (rt_done) read.csv(rt_path, check.names = FALSE)
  for (r in compound_spec$role) {
    res_csv <- file.path(dir_res, paste0(r, "_results.csv"))
    if (!file.exists(res_csv)) next
    res_ids <- read.csv(res_csv)$xcms_fts
    new_r   <- rt_valid[rt_valid$role == r, ]
    same    <- all(res_ids %in% new_r$xcms_fts)
    if (same && rt_done) {
      old_r <- rt_old[rt_old$role == r, ]
      same  <- isTRUE(all.equal(c(old_r$rt, old_r$mz), c(new_r$rt, new_r$mz)))
    }
    if (!same)
      stop(ds, ": hits_rt no longer matches the SIRIUS results in ", dir_res,
           "; move them aside to recompute")
  }
  hits_rt_list[[ds]] <- rt_valid[, c("role", "xcms_fts", "rt", "mz",
                                     "adduct_inferred", "source_file", "assay")]
  write.csv(hits_rt_list[[ds]], file.path(dir_res, "hits_rt.csv"),
            row.names = FALSE)

  # One SIRIUS run per target, in compound.csv order, into results/<id>/<role>_results.csv;
  # a target that has its results keeps them.
  if (!sirius_done) for (r in compound_spec$role) {
    res_csv <- file.path(dir_res, paste0(r, "_results.csv"))
    if (file.exists(res_csv)) next
    sp_r <- slice_role(hits_ds, sp, sp_file, r)
    if (!length(sp_r)) {
      message("  No ", r, " spectra, skipping ", r, " SIRIUS run")
      next
    }
    adduct_list_r <- lapply(sp_r, function(s) s$adduct[1])
    res_r <- run_sirius_role(sp_r, adduct_list_r,
                             file.path("sirius_projects", ds, r), study_id = r)
    write.csv(res_r, res_csv, row.names = FALSE)
    message("  ", toupper(substring(r, 1, 1)), substring(r, 2), " done: ",
            nrow(res_r), " rows")
  }
}

message("All Metabolomics Workbench datasets processed.")
```

# Evaluate SIRIUS matches

## Helper functions

```{r resolve-helpers}
CONFIDENCE_LEVELS  <- c("strict", "standard", "liberal", "not_confirmed")
CONFIDENCE_COLOURS <- c(
  strict        = "#1A7837",
  standard      = "#74C476",
  liberal       = "#FEC44F",
  not_confirmed = "#D7191C"
)

resolve_features <- function(res, target_formula, target_ikey_prefix) {
  target_rows <- res |>
    filter(molecularFormula == target_formula,
           !is.na(inchiKey),
           startsWith(inchiKey, target_ikey_prefix)) |>
    group_by(xcms_fts) |>
    slice_min(rank.y, n = 1, with_ties = FALSE) |>
    ungroup() |>
    transmute(
      xcms_fts,
      formula_rank   = rank.x,
      formula_score  = round(siriusScoreNormalized, 3),
      structure_rank = rank.y,
      tanimoto       = round(tanimotoSimilarity, 3),
      csi_score      = round(csiScore, 1),
      adduct,
      structure_name = structureName,
      target_present = TRUE
    )

  res |>
    distinct(xcms_fts) |>
    left_join(target_rows, by = "xcms_fts") |>
    mutate(
      target_present = !is.na(formula_rank),
      confidence = case_when(
        formula_rank == 1  & structure_rank == 1  ~ "strict",
        formula_rank <= 3  & structure_rank <= 5  ~ "standard",
        formula_rank <= 15 & structure_rank <= 15 ~ "liberal",
        TRUE                                       ~ "not_confirmed"
      ),
      confidence = factor(confidence, levels = CONFIDENCE_LEVELS)
    ) |>
    arrange(confidence, formula_rank, structure_rank)
}

resolve_all <- function(results_dir = "results", compound_spec, datasets) {
  datasets <- intersect(datasets,
                        list.dirs(results_dir, recursive = FALSE, full.names = FALSE))
  all_rows <- list()

  for (ds in datasets) {
    for (i in seq_len(nrow(compound_spec))) {
      role     <- compound_spec$role[i]
      csv_path <- file.path(results_dir, ds, paste0(role, "_results.csv"))
      if (!file.exists(csv_path)) next
      res <- read.csv(csv_path, check.names = FALSE)
      if (!nrow(res)) next

      resolve_features(res,
                       target_formula     = compound_spec$formula[i],
                       target_ikey_prefix = compound_spec$inchikey_block1[i]) |>
        mutate(dataset_id = ds, role = role, .before = xcms_fts) ->
        all_rows[[paste(ds, role, sep = "::")]]
    }
  }

  features <- bind_rows(all_rows)
  summary_tbl <- features |>
    group_by(dataset_id, role) |>
    summarise(n_total          = n(),
              n_target_present = sum(target_present, na.rm = TRUE),
              n_strict         = sum(confidence == "strict",   na.rm = TRUE),
              n_standard       = sum(confidence == "standard", na.rm = TRUE),
              n_liberal        = sum(confidence == "liberal",  na.rm = TRUE),
              .groups = "drop")

  list(features = features, summary = summary_tbl)
}
```

## Run evaluation

```{r resolve-run}
resolved    <- resolve_all(results_dir = "results", compound_spec = compound_spec,
                           datasets = mwb_datasets)
features    <- resolved$features
summary_tbl <- resolved$summary
summary_tbl
```

## Plots

### Confidence levels per assay and role

```{r plot-confidence, fig.width=9, fig.height=4}
features |>
  count(dataset_id, role, confidence) |>
  ggplot(aes(x = dataset_id, y = n, fill = confidence)) +
  geom_col() +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5),
            size = 3, colour = "white", fontface = "bold") +
  facet_wrap(~role) +
  scale_fill_manual(values = CONFIDENCE_COLOURS, drop = FALSE) +
  labs(x = NULL, y = "Number of features", fill = "Confidence",
       title = "SIRIUS confirmation confidence: Metabolomics Workbench") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

### Formula and structure rank

```{r plot-ranks, fig.width=6, fig.height=5}
features |>
  filter(target_present) |>
  ggplot(aes(x = formula_rank, y = structure_rank,
             colour = confidence, shape = role)) +
  annotate("rect", xmin = 0.5, xmax = 1.5,  ymin = 0.5, ymax = 1.5,
           fill = "#1A7837", alpha = 0.08) +
  annotate("rect", xmin = 0.5, xmax = 3.5,  ymin = 0.5, ymax = 5.5,
           fill = "#74C476", alpha = 0.06) +
  annotate("rect", xmin = 0.5, xmax = 15.5, ymin = 0.5, ymax = 15.5,
           fill = "#FEC44F", alpha = 0.04) +
  geom_jitter(size = 3, width = 0.2, height = 0.2, alpha = 0.9) +
  scale_colour_manual(values = CONFIDENCE_COLOURS, drop = FALSE) +
  scale_x_continuous(breaks = c(1, 3, 5, 10, 15)) +
  scale_y_continuous(breaks = c(1, 5, 10, 15)) +
  labs(x = "Formula rank", y = "Structure rank",
       colour = "Confidence", shape = "Role",
       title = "Rank of target: Metabolomics Workbench") +
  theme_bw() +
  theme(legend.position = "bottom")
```

### Score scatter

```{r plot-scores, fig.width=7, fig.height=4}
features |>
  filter(target_present) |>
  ggplot(aes(x = tanimoto, y = formula_score,
             colour = confidence, shape = role)) +
  geom_point(size = 3, alpha = 0.85) +
  scale_colour_manual(values = CONFIDENCE_COLOURS, drop = FALSE) +
  labs(x = "Tanimoto similarity", y = "Normalised formula score",
       colour = "Confidence", shape = "Role",
       title = "Score distribution: Metabolomics Workbench") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# RT summary

## Join retention times

```{r rt-join}
# Load hits_rt.csv from disk for every dataset the loop above skipped, so that
# every dataset in the combined artifact has its retention times.
missing <- setdiff(unique(features$dataset_id), names(hits_rt_list))
for (ds in missing) {
  p <- file.path("results", ds, "hits_rt.csv")
  if (file.exists(p)) hits_rt_list[[ds]] <- read.csv(p, check.names = FALSE)
}

rt_tbl   <- bind_rows(hits_rt_list, .id = "dataset_id")
features <- features |>
  left_join(rt_tbl, by = c("dataset_id", "role", "xcms_fts"))
```

## RT range per assay and role

```{r rt-summary-table}
rt_summary <- features |>
  filter(!is.na(rt)) |>
  group_by(dataset_id, role, confidence) |>
  summarise(
    n         = n(),
    rt_min    = round(min(rt), 1),
    rt_median = round(median(rt), 1),
    rt_max    = round(max(rt), 1),
    rt_range  = round(max(rt) - min(rt), 1),
    .groups   = "drop"
  )
rt_summary
```

## RT distribution across assays

```{r rt-plot, fig.width=9, fig.height=4}
features |>
  filter(!is.na(rt)) |>
  ggplot(aes(x = dataset_id, y = rt,
             colour = confidence, shape = role)) +
  geom_jitter(size = 3, width = 0.15, alpha = 0.85) +
  facet_wrap(~role) +
  scale_colour_manual(values = CONFIDENCE_COLOURS, drop = FALSE) +
  labs(x = NULL, y = "Retention time (s)",
       colour = "Confidence", shape = "Role",
       title = "RT of confirmed features across Metabolomics Workbench datasets (s)") +
  theme_bw() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

# Write confirmed hits

Hits with `strict`, `standard`, or `liberal` SIRIUS confidence are kept as
Metabolomics Workbench-confirmed; the `confidence` and `rt` (seconds) columns
are preserved for downstream stratification and RT-window alignment.

```{r write-confirmed}
hits_confirmed_mwb <- features |>
  filter(confidence %in% c("strict", "standard", "liberal"))

# The artifact keys rows by deposit, as the other repositories' do; `assay` says
# which dataset of the deposit a row belongs to.
dep_of <- setNames(hits_mwb$ATTRIBUTE_DatasetAccession, hits_mwb$dataset_id)
hits_confirmed_mwb$dataset_id <- unname(dep_of[hits_confirmed_mwb$dataset_id])

write.csv(hits_confirmed_mwb, "artifacts/hits-confirmed-metabolomics-workbench.csv",
          row.names = FALSE)
```
