---
title: "02 - Metadata"
subtitle: "Step 2: sample descriptions retrieved and harmonised"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
---

# Overview

Each deposit that step 1 kept is described from its repository's own metadata,
and the descriptions are mapped onto one per-sample schema. Each hit is also
assigned to the assay it belongs to: from here on the unit is the *assay* (one
analysis of a deposit; a MassIVE deposit counts as one), and later steps join on
its identifier, `dataset_id`.

**Input:** `artifacts/hits_search.csv`, plus per-deposit rules written in this
notebook where the automatic route falls short, marked as manual work. The manual
rules read files obtained by hand and kept in `dataset_metadata/` (they are
tracked in git): `MSV000086975_metadata.tsv`, `MSV000085256_panredu.json`,
`MSV000082433_panredu.json`, `s_MTBLS12231.txt` and the positive and negative
mode assay files `a_MTBLS12231_*.txt`.

**Output:** `artifacts/hits_assigned.csv`, `artifacts/metadata_entities.csv`,
`artifacts/assay_metadata.csv` and `artifacts/related_deposits.csv`, plus `artifacts/rescue_samples.csv`
and `artifacts/removed_step2.csv` (the deposits removed by the first pass).

```{r setup}
source("../../R/setup.R")
source("../../R/masst.R")
source("../../R/metadata.R")
```

```{r load-hits}
all_hits <- read.csv("artifacts/hits_search.csv", check.names = FALSE,
                     stringsAsFactors = FALSE)
D_both <- unique(all_hits$ATTRIBUTE_DatasetAccession)
D_entering <- D_both
cat("Deposits:", length(D_both), "\n")
```

# What is already known

Before anything is fetched, the deposits that the search index (ReDU) already
describes are checked, with the same function as in the other application. A
deposit is judged only when ReDU describes every file that carries a hit. The check
then reads every file ReDU describes in the deposit, blanks and QCs set aside, and
removes the deposit when none of these files is human and at least one is of
another species (a file with no species recorded is ignored). This application is
exploratory, so the species is the only thing checked in ReDU: no deposit is
removed for its body site, and none for metadata that is missing. A deposit in a
repository the `MsBackend` packages cannot read is removed, as in the other
application. Everything else goes on to the other sources.

```{r known-wrong}
# Species only: no body-site check and no removal by hand.
pass1 <- redu_first_pass(
  all_hits,
  fasst_db  = Sys.getenv("FASST_DB", "E:/FASSTrecords/db/masst_records.sqlite"),
  redu_out  = "StructureMASST_output/redu_files.csv",
  body_site = NULL,
  by_hand   = NULL)
removed_step2 <- pass1$removed
D_both        <- pass1$deposits
all_hits      <- pass1$hits
write.csv(removed_step2, "artifacts/removed_step2.csv", row.names = FALSE)
cat("ReDU describes every hit file of", pass1$n_described,
    "deposits | removed:", nrow(removed_step2), "| deposits left:", length(D_both), "\n")
table(removed_step2$reason)
```

# Sources

The repositories share no metadata schema. MassIVE deposits are described by the
submitter's own sample table, when there is one, and by Pan-ReDU; MetaboLights
deposits by their ISA-Tab files and Metabolomics Workbench deposits by their
mwTab files, both fetched with functions of `MsBackendMetaboLights` and
`MsBackendMetabolomicsWorkbench`. Most fetched files are cached in
`dataset_metadata/`. Placeholders such as "missing value" are read as missing.

# Manual: metadata found by hand

::: {.callout-warning title="Manual work"}
The deposits listed here needed their metadata found by hand, because no source
reached by the automatic route below describes them well enough. Each block reads
one deposit's metadata from where it was found (a file in the deposit, a Pan-ReDU
export or the ISA-Tab files) and maps its column names by hand.

- **`MSV000082493` and `MSV000084008`**: the omeprazole probe cohort (Jarmusch
  et al. 2020, *Clin Transl Sci*, PMC7485946). Healthy volunteers given a
  modified Cooperstown CYP-probe cocktail on study days 1 and 9, with a 7-day
  cefprozil course in between; plasma, urine, feces and skin. The metadata is in
  neither deposit and in no catalogue: it is in the authors' GitHub repository,
  `Materials/Metadata.txt` at commit `95586620`, and it carries the genotyped
  CYP2C19 phenotype of every subject. The two accessions share this table:
  plasma rows go to 084008, the rest to 082493.
- **`MSV000086975`**: schizophrenia (PMID 34741130). A submitter TSV inside the
  deposit.
- **`MSV000085256`**: Alzheimer's disease (PMID 34244695). A Pan-ReDU export,
  the deposit itself having come up short.
- **`MSV000082433`**: Amerindian communities. A Pan-ReDU export. The one deposit
  whose recorded country describes where the participants come from rather than
  where the study ran.
- **`MTBLS12231`**: COVID-19 (PMC12019401). The ISA-Tab sample file joined to its
  positive and negative mode assay files. ISA-Tab allows duplicate column names,
  so the first occurrence of each is used.

The probe cohort is fetched first, because the automatic route below reads the
files these blocks write into `dataset_metadata/`. All of them are written with
the same columns into `artifacts/rescue_samples.csv`. The columns that exist only
because the probe cohort has them are `timepoint` (study day), `timepoint_mins`
(minutes after the dose), `pharmkinetics_cyp2c19` (the study's own per-subject
CYP2C19 value), `cyp2c19_phenotype` and `cyp2c19_genotype`.
MSV000084008 has no Pan-ReDU record and all its hit files are plasma, so its
metadata is the plasma part of the shared table.
:::

```{r metadata-setup}
MD <- "dataset_metadata"
miss <- function(x) {
  x[x %in% c("", "missing value", "not applicable", "not collected")] <- NA
  x
}
```

```{r fetch-82493}
SRC_COMMIT <- "95586620ead4ac5241dc1e99a98e047c1f99b17b"
SRC_URL <- sprintf(paste0("https://raw.githubusercontent.com/alan-jarmusch/",
                          "UntargetedMS-DrugMetabolism/%s/Materials/Metadata.txt"),
                   SRC_COMMIT)
LOCAL <- "dataset_metadata/MSV000082493_metadata.tsv"
if (!file.exists(LOCAL))
  utils::download.file(SRC_URL, LOCAL, quiet = TRUE, mode = "wb")

jm <- read.delim(LOCAL, check.names = FALSE, stringsAsFactors = FALSE,
                na.strings = c("", "not applicable"))
```

```{r std-schema}
std_cols <- c("deposit_id","file","subject_id","sex","age","health_status",
              "specimen","timepoint","timepoint_mins","pharmkinetics_cyp2c19",
              "cyp2c19_phenotype","cyp2c19_genotype","include")
num_cols <- c("age", "timepoint_mins", "pharmkinetics_cyp2c19")
blank <- function(n) setNames(as.data.frame(matrix(NA_character_, n, length(std_cols))), std_cols)
```

```{r jarmusch}
jm_std <- data.frame(
  file = jm$filename, subject_id = jm$ATTRIBUTE_Subject, sex = jm$ATTRIBUTE_sex,
  age = as_num(jm$age), health_status = "healthy",
  specimen = dplyr::case_when(grepl("plasma", jm$ATTRIBUTE_Body_Site, TRUE) ~ "plasma",
                              grepl("^UA", jm$ATTRIBUTE_Body_Site) ~ "urine",
                              grepl("fecal", jm$ATTRIBUTE_Body_Site, TRUE) ~ "feces",
                              grepl("forehead|skin", jm$ATTRIBUTE_Body_Site, TRUE) ~ "skin",
                              TRUE ~ NA_character_),
  timepoint = jm$ATTRIBUTE_Study_Day,
  timepoint_mins = as_num(jm$ATTRIBUTE_Time_Point_Mins),
  pharmkinetics_cyp2c19 = as_num(jm$PharmKinetics_CYP2C19),
  cyp2c19_phenotype = jm$PharmGeno_Phenotype_2C19,
  cyp2c19_genotype = jm$PharmGeno_CYP2C19, include = jm$Include_Exclude,
  stringsAsFactors = FALSE)
jm_std <- jm_std[jm_std$include %in% "include" & !is.na(jm_std$specimen), ]
d84008 <- cbind(deposit_id = "MSV000084008", jm_std[jm_std$specimen == "plasma", std_cols[-1]])
d82493 <- cbind(deposit_id = "MSV000082493", jm_std[jm_std$specimen != "plasma", std_cols[-1]])
```

```{r probe-summary}
#| echo: false
cat("MSV000082493 + 084008:", nrow(jm_std), "biological samples of", nrow(jm), "files\n")
cat("\nspecimen x study day:\n"); print(table(jm_std$specimen, jm_std$timepoint, useNA = "ifany"))
cat("\nCYP2C19 genotyped phenotype:\n"); print(table(jm_std$cyp2c19_phenotype, useNA = "ifany"))
cat("\nsubjects:", length(unique(jm_std$subject_id)), "\n")
```

```{r derive-84008}
is_plasma <- grepl("plasma", jm$ATTRIBUTE_Body_Site, ignore.case = TRUE)
plasma <- jm[is_plasma, , drop = FALSE]
write.table(plasma, "dataset_metadata/MSV000084008_metadata.tsv",
            sep = "\t", quote = FALSE, row.names = FALSE)
cat("MSV000084008_metadata.tsv <- plasma rows:", nrow(plasma), "of", nrow(jm), "\n")
cat("body sites kept:", paste(sort(unique(plasma$ATTRIBUTE_Body_Site)), collapse = ", "), "\n")
```

```{r sz}
sz <- read.delim(file.path(MD, "MSV000086975_metadata.tsv"), check.names = FALSE, stringsAsFactors = FALSE)
sid <- strsplit(sz$`#SampleID`, "_", fixed = TRUE)
d86975 <- blank(nrow(sz)); d86975$deposit_id <- "MSV000086975"
d86975$file <- sz$Metabolomics_FileName; d86975$subject_id <- sz$`#SampleID`
d86975$sex <- vapply(sid, function(p) if (length(p) >= 2) p[2] else NA_character_, character(1))
d86975$age <- as_num(sz$AGE)
d86975$health_status <- sz$Metadata          # HC / SZ
d86975$specimen <- "plasma"; d86975$include <- "include"
```

```{r az}
pj <- function(f) {
  txt <- paste(readLines(file.path(MD, f), warn = FALSE), collapse = "")
  objs <- regmatches(txt, gregexpr("\\{[^{}]*\\}", txt))[[1]]
  g <- function(o, k) { m <- regmatches(o, regexpr(sprintf('"%s": "[^"]*"', k), o))
    if (length(m)) sub('.*: "', "", sub('"$', "", m)) else NA_character_ }
  data.frame(file = miss(vapply(objs, g, character(1), "filename")),
             subject_id = miss(vapply(objs, g, character(1), "SubjectIdentifierAsRecorded")),
             sex = miss(vapply(objs, g, character(1), "BiologicalSex")),
             age = as_num(miss(vapply(objs, g, character(1), "AgeInYears"))),
             health_status = miss(vapply(objs, g, character(1), "HealthStatus")),
             specimen = miss(vapply(objs, g, character(1), "UBERONBodyPartName")),
             row.names = NULL, stringsAsFactors = FALSE) }
a <- pj("MSV000085256_panredu.json"); d85256 <- blank(nrow(a)); d85256$deposit_id <- "MSV000085256"
d85256[names(a)] <- a; d85256$specimen <- ifelse(grepl("plasma", d85256$specimen, TRUE), "plasma",
                                          ifelse(grepl("cerebro|CSF", d85256$specimen, TRUE), "CSF", d85256$specimen))
d85256$include <- "include"
```

```{r amerindians}
b <- pj("MSV000082433_panredu.json"); d82433 <- blank(nrow(b)); d82433$deposit_id <- "MSV000082433"
d82433[names(b)] <- b; d82433$include <- "include"
```

```{r covid}
col1 <- function(d, nm) d[[ which(names(d) == nm)[1] ]]
s <- read.delim(file.path(MD, "s_MTBLS12231.txt"), check.names = FALSE, stringsAsFactors = FALSE)
read_a <- function(p) { d <- read.delim(
    file.path(MD, sprintf("a_MTBLS12231_LC-MS_%s_reverse-phase_metabolite_profiling.txt", p)),
    check.names = FALSE, stringsAsFactors = FALSE)
  # the raw column is empty here; pick whichever *Spectral Data File column is
  # populated
  fcols <- grep("Spectral Data File", names(d))
  fcol  <- fcols[which.max(vapply(fcols, function(i)
             sum(grepl("\\.(mzML|mzXML|raw|d|wiff)$", d[[i]])), integer(1)))]
  data.frame(sample = col1(d, "Sample Name"), raw = basename(d[[fcol]]),
             stringsAsFactors = FALSE) }
amap <- rbind(read_a("positive"), read_a("negative"))
sref <- data.frame(sample = col1(s, "Sample Name"),
                   genetype = col1(s, "Factor Value[genetype]"),
                   part = col1(s, "Characteristics[Organism part]"), stringsAsFactors = FALSE)
mm <- dplyr::left_join(amap, sref, by = "sample")
d12231 <- blank(nrow(mm)); d12231$deposit_id <- "MTBLS12231"
d12231$file <- basename(mm$raw); d12231$subject_id <- mm$sample
d12231$health_status <- mm$genetype; d12231$specimen <- mm$part
d12231$include <- "include"
```

```{r combine}
frames <- lapply(list(d82493, d84008, d86975, d85256, d82433, d12231), function(f) {
  f$timepoint <- as.character(f$timepoint)
  for (cc in setdiff(std_cols, num_cols)) f[[cc]] <- as.character(f[[cc]])
  for (cc in num_cols) f[[cc]] <- as_num(f[[cc]])
  f[, std_cols]
})
all_std <- dplyr::bind_rows(frames)
write.csv(all_std, "artifacts/rescue_samples.csv", row.names = FALSE)

cat("Rescue samples per deposit x specimen:\n"); print(table(all_std$deposit_id, all_std$specimen, useNA = "ifany"))
cat("\nHealth stratifier availability per deposit:\n")
for (d in unique(all_std$deposit_id)) {
  x <- all_std[all_std$deposit_id == d, ]
  cat(sprintf("  %-14s n=%-4d sex=%-3s age=%-3s health={%s}\n", d, nrow(x),
      ifelse(any(!is.na(x$sex)),"Y","-"), ifelse(any(!is.na(x$age)),"Y","-"),
      paste(sort(unique(na.omit(x$health_status))), collapse="/")))
}
```

`r length(intersect(D_both, all_std$deposit_id))` of the
`r length(unique(all_std$deposit_id))` deposits read by hand are among the
deposits left for the question, and the step summary counts only these.

# Submitter tables in the deposits

Every MassIVE deposit is also looked up for the submitter's own table, as in the
other application: any table file of the deposit, in any folder, with a column
that names the deposit's own MS files; only the rows of those files are kept. A
table found by hand above is used as it is.

```{r submitter-tables}
msv   <- D_both[startsWith(D_both, "MSV")]
found <- vapply(msv, function(d) !is.null(fetch_massive_submitter_table(d)), logical(1))
found <- found[!msv %in% all_std$deposit_id]
cat("MassIVE deposits with a submitter table:", sum(found), "of", length(msv), "\n")
```

# Automatic harvest

For every deposit, technical metadata (chromatography, ionisation, instrument,
per assay) and biological metadata (which columns hold the subject, sex, age and
so on) are read from the sources above. `harvest_metadata_coverage()` maps each
deposit's own column names onto the shared fields.

```{r auto-assess}
#| code-fold: true
tech_raw <- fetch_assay_metadata(D_both)
coverage <- tibble::as_tibble(
  dplyr::bind_rows(lapply(D_both, harvest_metadata_coverage))
)
write.csv(tech_raw, "artifacts/assay_metadata.csv", row.names = FALSE)
```

```{r auto-assess-display}
#| echo: false
coverage[, c("dataset_id", "n_samples", "n_canonical_mapped",
             grep("^col_", names(coverage), value = TRUE))]
```

# Assays

MetaboLights and Workbench deposits often hold several assays of the same
samples under one accession. Each hit is assigned to its assay from the file it
came from; MassIVE has no assay concept, so its deposits stay whole. A hit whose
file matches no known assay is marked `_unmatched`.

::: {.callout-warning title="Manual work"}
The Workbench lists each sample's files without saying which analysis each one
belongs to, so the first name is taken to belong to the first analysis, and so
on. Where a deposit lists them in another order, the order is declared by hand in
`MWB_FILE_ORDER` (`R/metadata.R`), checked against the deposit's folders and the
retention times in its own results tables. File names found in two folders of a
deposit cannot be told apart and are left out.
:::

```{r mwb-file-order}
MWB_FILE_ORDER[intersect(names(MWB_FILE_ORDER), D_both)]
```

```{r assay-assignment}
mtbls_st <- D_both[grepl("^(MTBLS|ST)", D_both)]
assay_lookups  <- setNames(lapply(mtbls_st, assay_lookup_for),
                           mtbls_st)

all_hits$assay <- mapply(function(ds, fp) {
  if (startsWith(ds, "MSV")) return(NA_character_)
  file_to_assay(basename(fp), assay_lookups[[ds]])
}, all_hits$ATTRIBUTE_DatasetAccession, all_hits$file_path,
USE.NAMES = FALSE)

all_hits$deposit_id <- all_hits$ATTRIBUTE_DatasetAccession
all_hits$dataset_id <- make_dataset_id(all_hits$deposit_id, all_hits$assay)

# MASST hits whose file basename doesn't resolve to any known assay
mtbls_st_hits <- all_hits$deposit_id %in% mtbls_st
unresolved    <- mtbls_st_hits & is.na(all_hits$assay)
all_hits$dataset_id[unresolved] <- paste0(all_hits$deposit_id[unresolved],
                                          "_unmatched")
```

```{r write-hits}
write.csv(all_hits, "artifacts/hits_assigned.csv", row.names = FALSE)
```

# Manual: deposit-level facts and column names

::: {.callout-warning title="Manual work"}
Values that no per-file column carries, taken from the publication or the deposit
description (`constants`), and column names that the automatic mapping misses
(`custom_col_map`). `location_type` says whether `country` is where the study ran
(`study_site`) or where the participants come from (`participant_origin`); only
MSV000082433 is the latter.
One fact comes from file names: MSV000094097 has no subject column in any of
its metadata, and its file names are the only place its submitters say which
person a file comes from (`OAD_001_Basal_P`, `OAD_001_Day15_S`), so the person is
read from the start of the file name (`subject_from_file`). Without it, each
person's plasma and serum at two time points would count as four people.
:::

```{r manual-annotations}
constants <- list(
  # No subject column anywhere; the file name starts with the person
  # (OAD_001_Basal_P, OAD_001_Day15_S).
  MSV000094097 = list(subject_from_file = "^([A-Z]*OAD_[0-9]+)_.*$"),
  # --- study-site cohorts: `country` = recruitment location, not participant
  # origin ---
  MTBLS1866    = list(country = "Italy",          location_type = "study_site"),   # Barberis et al. 2020
  MTBLS11656   = list(country = "China", body_site = "blood plasma",
                      location_type = "study_site"),                                # Xinxiang/Zhengzhou 2017 (PMC11823611)
  ST002044     = list(country = "India", body_site = "blood serum",
                      location_type = "study_site"),                                # Kumar et al.; mwTab has no body_site col
  ST002722     = list(country = "United States of America", body_site = "blood serum",
                      location_type = "study_site"),                                # UCSD NAFLD Center 2014-2017
  MSV000088255 = list(country = "United Kingdom", location_type = "study_site"),    # 100 spot urine, Glasgow, NHS 14/LO/188

  # chromatography companions
  MSV000092754 = list(chromatography = "HILIC"),                                    # HILIC companion of 092757
  MSV000092757 = list(chromatography = "RP"),                                       # RP companion of 092754

  # 082493/084008: omeprazole CYP2C19-probe cohort; genotype in
  # rescue_samples.csv
  MSV000082493 = list(country = "United States of America", location_type = "study_site",
                      chromatography = "RP", health_status = "healthy"),
  MSV000084008 = list(country = "United States of America", location_type = "study_site",
                      chromatography = "RP", health_status = "healthy",
                      body_site = "blood plasma"),
  MSV000086975 = list(country = "United States of America", location_type = "study_site",
                      chromatography = "RP", body_site = "blood plasma"),           # schizophrenia (PMID 34741130)
  MSV000085256 = list(country = "United States of America", location_type = "study_site"),  # Alzheimer's (PMID 34244695)

  # the one deposit whose location describes participant origin:
  MSV000082433 = list(country = "Venezuela", location_type = "participant_origin",
                      chromatography = "RP", health_status = "healthy")             # Amerindian communities
)

custom_col_map <- list(
  # the Chagas pair's submitter tables (metadata/ in each deposit)
  MSV000092754 = list(
    subject_id    = "patient_id1",
    sex           = "ATTRIBUTE_sex",
    age           = "age",
    health_status = "ATTRIBUTE_status"),
  MSV000092757 = list(
    subject_id    = "patient_id1",
    sex           = "ATTRIBUTE_sex",
    age           = "age",
    health_status = "ATTRIBUTE_status"),
  # native columns the DRAFT_CANDIDATES heuristic misses (082493/084008 need
  # none)
  MSV000086975 = list(subject_id = "#SampleID", age = "AGE", health_status = "Metadata")
)
```

# One table per assay

One row per assay, with the repository, the chromatography, the column
mappings (`col_`) and the deposit-level facts (`const_`). Step 3 keeps the rows
that survive curation.

```{r build-metadata-table}
#| code-fold: true
entities <- unique(all_hits[!grepl("_unmatched$", all_hits$dataset_id),
                            c("dataset_id", "deposit_id", "assay")])
entities <- entities[order(entities$dataset_id, method = "radix"), ]

tech_renamed <- tech_raw |>
  dplyr::rename(deposit_id     = dataset_id,
                chromatography = chrom_type) |>
  dplyr::select(deposit_id, assay, repository, chromatography)

is_msv <- startsWith(entities$deposit_id, "MSV")
msv_meta <- entities[is_msv, ] |>
  dplyr::left_join(tech_renamed |> dplyr::select(-assay),
                   by = "deposit_id")
# A MassIVE deposit can report a real chromatography for some runs and "Unknown"
# for others, which gives it two rows here. Keep the real one.
unknown  <- is.na(msv_meta$chromatography) | msv_meta$chromatography %in% c("Unknown", "")
msv_meta <- msv_meta[order(msv_meta$dataset_id, unknown, method = "radix"), ]
msv_meta <- msv_meta[!duplicated(msv_meta$dataset_id), ]
nonmsv_meta <- entities[!is_msv, ] |>
  dplyr::left_join(tech_renamed, by = c("deposit_id", "assay"))

dataset_metadata <- dplyr::bind_rows(msv_meta, nonmsv_meta) |>
  dplyr::left_join(coverage |> dplyr::rename(deposit_id = dataset_id),
                   by = "deposit_id")

const_fields <- unique(unlist(lapply(constants, names)))
for (canon in const_fields) {
  col <- paste0("const_", canon)
  dataset_metadata[[col]] <- vapply(
    dataset_metadata$deposit_id,
    function(d) {
      v <- constants[[d]][[canon]]
      if (is.null(v)) NA_character_ else as.character(v)
    },
    character(1))
}

if ("const_chromatography" %in% names(dataset_metadata)) {
  use_const <- !is.na(dataset_metadata$const_chromatography) &
               (is.na(dataset_metadata$chromatography) |
                dataset_metadata$chromatography == "" |
                dataset_metadata$chromatography == "Unknown")
  dataset_metadata$chromatography[use_const] <-
    dataset_metadata$const_chromatography[use_const]
}

for (ds in names(custom_col_map)) {
  ix <- which(dataset_metadata$deposit_id == ds)
  if (!length(ix)) next
  for (canon in names(custom_col_map[[ds]])) {
    col_key <- paste0("col_", canon)
    if (col_key %in% names(dataset_metadata)) {
      dataset_metadata[[col_key]][ix] <- custom_col_map[[ds]][[canon]]
    }
  }
}

dataset_metadata <- dataset_metadata[
  , c("dataset_id", "deposit_id", "assay", "repository", "chromatography",
      "n_samples", "n_canonical_mapped",
      grep("^col_",   names(dataset_metadata), value = TRUE),
      grep("^const_", names(dataset_metadata), value = TRUE))]
write.csv(dataset_metadata, "artifacts/metadata_entities.csv", row.names = FALSE)
dataset_metadata
```

# Manual: related deposits

```{r probe-shared-files}
#| echo: false
# Files with a hit under both accessions of the probe cohort, matched on the file
# name without its folder and extension.
file_stem <- function(p) sub("[.][^.]+$", "", basename(p))
shared_probe <- intersect(
  file_stem(all_hits$file_path[all_hits$deposit_id == "MSV000082493"]),
  file_stem(all_hits$file_path[all_hits$deposit_id == "MSV000084008"]))
```

::: {.callout-warning title="Manual work"}
An accession is not always a study. Accessions can be related in opposite ways,
and the analysis has to know which is which.

**MSV000082493 and MSV000084008 are two accessions of one study** (the same
14-subject crossover, PMC7485946). The same physical files appear under both,
with different paths and extensions, so they must be counted once.
`r length(shared_probe)` of the files with a hit are under both, such as:

```
mzspec:MSV000082493:updates/2019-03-27_alan_jarmusch_6d15f636/peak/blood/DM003_1_360_RG10_01_43808.mzXML
mzspec:MSV000084008:ccms_peak/blood/DM003_1_360_RG10_01_43808.mzML
```

**MSV000092754 and MSV000092757 are the opposite case.** They share their file
names and subjects, but one is the HILIC run and the other reverse phase, so the
same file name refers to a different measurement in each. They count as one
study for pooling and must never be merged on file name.
:::

```{r related-deposits}
related_deposits <- tibble::tribble(
  ~dataset_id,    ~related_to,     ~relation,                  ~notes,
  "MSV000084008", "MSV000082493",  "same_study_different_arm",
  paste("Same 14-subject study (PMC7485946). 084008 re-deposits 082493's blood",
        "arm under other paths and extensions. 082493 contributes feces/skin/urine here,",
        "084008 blood. Do not treat as independent cohorts and do not count the",
        "subjects twice: collapse on file name."),

  "MSV000092757", "MSV000092754",  "same_samples_two_assays",
  paste("Same Chagas serum panel measured twice, HILIC (092754) and",
        "reverse phase (092757). The file names are shared but the files are",
        "different measurements, not copies. One study for pooling; never merge",
        "on file name.")
)
write.csv(related_deposits, "artifacts/related_deposits.csv", row.names = FALSE)
related_deposits
```

# Step summary

The results of this step, as the chapter's table shows them.

```{r step-summary}
# Only the first pass (What is already known) removes deposits here: the deposits
# left for the question are those it keeps.
summary_step <- data.frame(
  item  = c("deposits entering", "deposits that may answer the research question",
            "... sample table found by hand", "... information completed by hand",
            "assays with a hit"),
  value = c(length(D_entering),
            length(D_both),
            length(intersect(D_both, all_std$deposit_id)),
            length(intersect(D_both, union(names(constants), names(custom_col_map)))),
            nrow(dataset_metadata)))
# For the discussion's recap only: the files with a hit in those deposits.
summary_step <- rbind(summary_step, data.frame(
  item  = "recap: files",
  value = length(unique(paste(all_hits$ATTRIBUTE_DatasetAccession,
                              basename(all_hits$file_path))))))
write.csv(summary_step, "artifacts/summary_step2.csv", row.names = FALSE)
summary_step
```

# Session info

```{r session-info}
sessionInfo()
```
