---
title: "03 - Curation"
subtitle: "Step 3: deposits dropped only for a stated reason"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
---

# Overview

Deposits that cannot answer the question are removed, each with a recorded
reason, and only the assays that measure every target are kept.

**Input:** `artifacts/hits_assigned.csv`, `artifacts/metadata_entities.csv`, `artifacts/assay_metadata.csv` and `artifacts/removed_step2.csv` (the deposits step 2's first pass removed), plus the list of excluded deposits written in this notebook (manual work).

**Output:** `artifacts/hits_curated.csv` and `artifacts/dataset_metadata_table.csv`,
read by step 4 and every later step, and
`artifacts/exclusions.csv`, the reason for every deposit removed.

```{r setup}
source("../../R/setup.R")
source("../../R/masst.R")
source("../../R/metadata.R")
```

```{r load}
all_hits <- read.csv("artifacts/hits_assigned.csv", check.names = FALSE,
                     stringsAsFactors = FALSE)
metadata <- read.csv("artifacts/metadata_entities.csv", check.names = FALSE,
                     stringsAsFactors = FALSE)
removed_step2 <- if (file.exists("artifacts/removed_step2.csv"))
  read.csv("artifacts/removed_step2.csv", colClasses = "character") else
  data.frame(dataset_id = character(), reason = character(), notes = character())
D_both <- unique(all_hits$deposit_id)
D_entering <- D_both
```

# Manual: excluded deposits

::: {.callout-warning title="Manual work"}
Each deposit below was read and excluded by hand, with the reason recorded.

| reason | what it means |
|---|---|
| `no_or_limited_metadata` | no usable biological metadata anywhere (deposit, Pan-ReDU, publication), or a single value for every field |
| `access_restricted` | demographics behind a controlled-access portal |
| `dia` | data-independent acquisition; single-scan MASST matching is unreliable |
| `in_vitro` | cell culture or reference-library deposit; the hits come from added standards, not from samples |
| `non_human` | non-human samples, where ReDU carries no taxonomy to catch them |
| `reference_material` | a reference-material suite, not a study cohort |
:::

```{r other-exclusions}
other_exclusions <- dplyr::bind_rows(removed_step2, tibble::tribble(
  ~dataset_id,    ~reason,                  ~notes,
  "MSV000085870", "no_or_limited_metadata", "DDA acquisition confirmed but no Pan-ReDU or deposit metadata",
  "MSV000093072", "no_or_limited_metadata", "",
  "MSV000095726", "no_or_limited_metadata", "",
  "MSV000096359", "no_or_limited_metadata", "",
  "MSV000096884", "no_or_limited_metadata", "",
  "MSV000097110", "no_or_limited_metadata", "",
  "MSV000094515", "no_or_limited_metadata", "stool / Alzheimer's (U19 ADRC NCRAD); demographics NCRAD-controlled",
  "ST002238",     "no_or_limited_metadata", "no metadata file or linked publication found",
  "MSV000094395", "no_or_limited_metadata", "only DOIDCommonName=Alzheimer's, no other strata in MASST or Pan-ReDU",
  "MSV000095143", "no_or_limited_metadata", "submitter TSV documents 264 files with rich metadata but those filenames don't exist on MassIVE; the 1077 actually-hosted files (Pan-ReDU naming, used by MASST + SIRIUS) have all canonical fields as 'missing value' in every source. Deposit cannot be stratified.",
  "MSV000097935", "access_restricted",      "demographics access-restricted ('Partial Public' MassIVE deposit, Rancho Bernardo Study policy)",
  "MTBLS417",     "dia",                    "DIA; duplicate of a DIA MassIVE deposit, poor metadata",
  "MTBLS2746",    "in_vitro",               "drug-treatment cell culture; MASST hits are spike artefacts",
  "ST001715",     "in_vitro",               "in vitro xenobiotic reference library",
  "MSV000082382", "non_human",              "Pseudomonas / Bacillus on melon plants",
  "MSV000086015", "non_human",              "San Diego Zoo rhinoceros serum (Diceros / Ceratotherium)",
  "MSV000087018", "reference_material",     "NIST RM 8231 frozen-plasma reference-material QA/QC suite, not a study cohort",
  "MSV000090544", "reference_material",     "GNPS Matrix-tubes extraction-method validation on high-biomass fecal+saliva (DOI 10.25345/C5QJ7831B); plate/well layout, no per-sample metadata, not a dosed cohort; omeprazole is incidental",
  "MSV000083065", "no_or_limited_metadata", "NOBEL twin plasma cohort; no Pan-ReDU, no accessible per-sample strata",
  "MSV000083593", "no_or_limited_metadata", "S. aureus infection plasma; no accessible per-sample metadata",
  "MSV000083955", "no_or_limited_metadata", "cerebral cavernous malformation plasma; no per-sample metadata",
  "MSV000084716", "no_or_limited_metadata", "Undiagnosed Disease Network; paper (PMID 33883556) confirms no medication/CYP data, part GC-MS, not in Pan-ReDU",
  "MSV000085508", "no_or_limited_metadata", "UDN, same cohort as MSV000084716 (duplicate), no metadata",
  "MSV000086270", "no_or_limited_metadata", "U19 ADNI-Go 1000-plasma cohort; no accessible per-sample strata",
  "MSV000088645", "no_or_limited_metadata", "U19 T1000 serum; no accessible per-sample strata",
  "MSV000090580", "no_or_limited_metadata", "osteoarthritis plasma (Guma); no accessible per-sample metadata",
  "MSV000092926", "no_or_limited_metadata", "plasma cohort (Knight ADC); no accessible per-sample metadata",
  "MSV000093037", "no_or_limited_metadata", "unidentified deposit: no title/summary/keywords/ReDU",
  "MSV000095383", "no_or_limited_metadata", "U19 Wisconsin serum (Alzheimer's); no accessible per-sample strata"
))

by_hand <- intersect(D_both, other_exclusions$dataset_id)
D_both  <- setdiff(D_both, other_exclusions$dataset_id)
```

```{r other-exclusions-summary}
#| echo: false
cat("Other manual exclusions:", nrow(other_exclusions), "deposits\n")
print(table(other_exclusions$reason))
cat("\nSurviving after manual exclusions:", length(D_both), "deposits\n")
```

## The sample table of MSV000095143

MSV000095143's own sample table came with an update of the deposit
(`updates/2024-10-07_jzemlin_40f5b739/metadata/gnps_metadata_OA.csv`, named in its
`ccms_metadata/summary.tsv`), which the file list of the MassIVE backend leaves out,
so step 2 never reads it. It is fetched here once, kept in `dataset_metadata/`, and
its file names are compared with the deposit's MS files.

```{r msv95143-table}
CHECK_95143 <- "MSV000095143"
LOCAL_95143 <- file.path("dataset_metadata", paste0(CHECK_95143, "_update_metadata.csv"))
if (!file.exists(LOCAL_95143)) {
  url <- paste0("https://massive.ucsd.edu/ProteoSAFe/DownloadResultFile?file=f.",
                CHECK_95143, "/updates/2024-10-07_jzemlin_40f5b739/metadata/",
                "gnps_metadata_OA.csv&forceDownload=true")
  tmp <- tempfile(fileext = ".csv")
  # MassIVE answers a burst of requests with "Too Many Requests" (an error, or a page
  # in place of the table): wait and try again, and keep only the table itself.
  got <- FALSE
  for (i in 1:6) {
    got <- isTRUE(tryCatch(utils::download.file(url, tmp, quiet = TRUE, mode = "wb") == 0,
                           error = function(e) FALSE)) &&
           startsWith(readLines(tmp, n = 1, warn = FALSE), "MassiveID")
    if (got) break
    Sys.sleep(120)
  }
  if (!got) stop("MSV000095143's update table did not download after 6 tries ",
                 "(MassIVE answers bursts with Too Many Requests); render again later")
  file.copy(tmp, LOCAL_95143)
}
tab_95143   <- read_submitter_table(LOCAL_95143)
names_95143 <- unique(tab_95143$filename[!is.na(tab_95143$filename)])
keys_95143  <- msv_ms_file_keys(massive_files(CHECK_95143))
in_95143    <- sum(norm_file_key(basename(names_95143)) %in% keys_95143)
```

Its table lists `r length(names_95143)` files, `r in_95143` of which are among the
deposit's `r length(keys_95143)` MS files.

# Every target in one assay

A deposit can carry every target and still never measure them in the same assay.
Only assays with a hit for every target are kept, and hits marked `_unmatched`
in step 2 are dropped. A deposit left without any assay is added to the
exclusions with the reason `no_assay_with_both_roles` (`no_assay_with_every_target`
with any other number of targets), or `no_hit_matched_to_assay` when step 2 could
match none of its hits to an assay.

```{r entity-coverage}
# One has_<role> column per target, in the order of compound.csv: does the assay
# carry a hit for it? The metabolite's keeps its short name, has_metab.
roles    <- compound_spec$role
has_cols <- sub("^has_metabolite$", "has_metab", paste0("has_", roles))
cover    <- all_hits[all_hits$deposit_id %in% D_both,
                     c("dataset_id", "deposit_id", "assay", "role")]
for (i in seq_along(roles)) cover[[has_cols[i]]] <- cover$role == roles[i]
entity_coverage <- cover |>
  dplyr::group_by(dataset_id, deposit_id, assay) |>
  dplyr::summarise(dplyr::across(dplyr::all_of(has_cols), any),
                   n_hits  = dplyr::n(),
                   .groups = "drop")
entity_coverage$keep <- Reduce(`&`, entity_coverage[has_cols]) &
  !grepl("_unmatched$", entity_coverage$dataset_id)

datasets_with_kept_entity <- unique(
  entity_coverage$deposit_id[entity_coverage$keep])
deposits_lost <- setdiff(D_both, datasets_with_kept_entity)
# A deposit whose hits are all `_unmatched` never had an assay to check.
unmatched_only <- deposits_lost[vapply(deposits_lost, function(d)
  all(grepl("_unmatched$", entity_coverage$dataset_id[entity_coverage$deposit_id == d])),
  logical(1))]

# The reason keeps its two-target name when there are two targets.
no_assay <- if (length(roles) == 2) "no_assay_with_both_roles" else
  "no_assay_with_every_target"
if (length(deposits_lost)) {
  unmatched <- deposits_lost %in% unmatched_only
  new_excl <- tibble::tibble(
    dataset_id = deposits_lost,
    reason     = ifelse(unmatched, "no_hit_matched_to_assay", no_assay),
    notes      = ifelse(unmatched, "no hit file matched to an assay in step 2",
                        paste(paste(roles, collapse = " + "),
                              "MASST hits land in different assays of the same deposit")))
  other_exclusions <- dplyr::bind_rows(other_exclusions, new_excl)
  D_both <- setdiff(D_both, deposits_lost)
}

entities_kept    <- entity_coverage$dataset_id[entity_coverage$keep]
all_hits         <- all_hits[all_hits$dataset_id %in% entities_kept, ]
dataset_metadata <- metadata[metadata$dataset_id %in% entities_kept, ]
```

```{r entity-coverage-summary}
#| echo: false
lost_assay <- setdiff(deposits_lost, unmatched_only)
if (length(lost_assay))
  cat("Deposits dropped (no assay carries ", if (length(roles) == 2) "both " else "all of ",
      paste(roles, collapse = " + "), "):\n  ",
      paste(lost_assay, collapse = "\n  "), "\n", sep = "")
if (length(unmatched_only))
  cat("Deposits dropped (no hit matched to an assay):\n  ",
      paste(unmatched_only, collapse = "\n  "), "\n", sep = "")
cat("Surviving entities:", length(entities_kept),
    " | surviving deposits:", length(datasets_with_kept_entity), "\n")
entity_coverage
```

# One set-up per MassIVE deposit

MetaboLights and the Metabolomics Workbench record the set-ups of a deposit as
separate assays; MassIVE does not, so a MassIVE deposit counts as one assay. Each
kept MassIVE deposit is checked here for a mix of set-ups, from what can be known
without the raw files: the chromatography and polarity its metadata records, and the
polarity of its hits (the charge of the adduct each hit matched). A deposit that
mixes set-ups is removed: it cannot be measured as one assay, and its files cannot
be told apart reliably. A file without a hit, in a deposit whose metadata records no
set-up, is not covered by this check.

```{r one-setup}
tech  <- read.csv("artifacts/assay_metadata.csv", stringsAsFactors = FALSE)
known <- function(x) sort(unique(x[!is.na(x) & !x %in% c("", "Unknown")]))
msv   <- sort(unique(all_hits$deposit_id[startsWith(all_hits$deposit_id, "MSV")]))
setups <- do.call(rbind, lapply(msv, function(d) {
  t   <- tech[tech$dataset_id == d, ]
  add <- all_hits$adduct_inferred[all_hits$deposit_id == d]
  pol <- ifelse(grepl("[+]$", add), "positive", ifelse(grepl("-$", add), "negative", NA))
  data.frame(deposit           = d,
             chromatography    = paste(known(t$chrom_type), collapse = " and "),
             polarity_recorded = paste(known(t$polarity), collapse = " and "),
             polarity_of_hits  = paste(known(pol), collapse = " and "))
}))
if (is.null(setups))
  setups <- data.frame(deposit = character(), chromatography = character(),
                       polarity_recorded = character(), polarity_of_hits = character())
setups$mixed <- grepl(" and ", paste(setups$chromatography, setups$polarity_recorded,
                                     setups$polarity_of_hits))
mixed <- setups$deposit[setups$mixed]
if (length(mixed)) {
  other_exclusions <- dplyr::bind_rows(other_exclusions, tibble::tibble(
    dataset_id = mixed, reason = "mixed_setups",
    notes      = "MassIVE deposit whose files mix set-ups; MassIVE records no assays"))
  D_both           <- setdiff(D_both, mixed)
  all_hits         <- all_hits[!all_hits$deposit_id %in% mixed, ]
  dataset_metadata <- dataset_metadata[!dataset_metadata$deposit_id %in% mixed, ]
}
setups
```

# Files deposited twice

A study deposited twice holds some files in both deposits (step 2,
`related_deposits.csv`); their hits are kept once, under the first deposit.

```{r drop-copies}
copy <- is_copy(all_hits$file_path, all_hits$deposit_id,
                function(d) all_hits$file_path[all_hits$deposit_id == d])
cat("hits on copies dropped:", sum(copy), "in",
    length(unique(basename(all_hits$file_path[copy]))), "files\n")
all_hits <- all_hits[!copy, ]
```

# Write curated hits

```{r write-artifacts}
write.csv(dataset_metadata, "artifacts/dataset_metadata_table.csv",
          row.names = FALSE)
write.csv(all_hits,         "artifacts/hits_curated.csv",
          row.names = FALSE)
write.csv(other_exclusions, "artifacts/exclusions.csv", row.names = FALSE)
```

```{r write-artifacts-summary}
#| echo: false
cat("artifacts/dataset_metadata_table.csv :",
    nrow(dataset_metadata), "rows\n")
cat("artifacts/hits_curated.csv           :",
    nrow(all_hits), "rows /",
    length(unique(all_hits$dataset_id)), "entities\n")
```

# Step summary

The results of this step, as the chapter's table shows them.

```{r step-summary}
# The removals of this step, by reason (step 2's are in its own table). A reason
# this application does not use reads "--".
step2_ids <- if (exists("removed_step2")) removed_step2$dataset_id else character()
step3 <- other_exclusions[!other_exclusions$dataset_id %in% step2_ids, ]
step3 <- step3[!duplicated(step3$dataset_id), ]
# One unit per table: deposits (the assays kept go to the recap). The rows starting with "..."
# break down the deposits removed.
REASONS <- list(
  "... no or too little metadata" = c("no_metadata", "no_or_limited_metadata"),
  "... not human study samples (another species, in vitro, reference material)" =
    c("not_human", "non_human", "in_vitro", "reference_material"),
  "... do not fit the research question (blood, age)" = c("no_blood_found", "no_adult_with_age"),
  "... files not placed in an assay" = c("no_hit_matched_to_assay", no_assay))
NOT_USED <- c("... do not fit the research question (blood, age)")
n_by <- vapply(REASONS, function(r) sum(step3$reason %in% r), numeric(1))
n_other <- sum(!step3$reason %in% unlist(REASONS))
summary_step <- data.frame(
  item  = c("deposits entering", "deposits removed", names(REASONS), "... other reasons",
            "deposits kept", "recap: assays kept"),
  value = as.character(c(length(D_entering), nrow(step3),
                         ifelse(names(REASONS) %in% NOT_USED, "--", n_by), n_other,
                         length(D_both), nrow(dataset_metadata))))
# For the discussion's recap only: the files with a hit that are kept.
summary_step <- rbind(summary_step, data.frame(
  item  = "recap: files",
  value = as.character(length(unique(paste(all_hits$deposit_id,
                                           basename(all_hits$file_path)))))))
write.csv(summary_step, "artifacts/summary_step3.csv", row.names = FALSE)
summary_step
```

# Session info

```{r session-info}
sessionInfo()
```
