---
title: "05 - Extraction"
subtitle: "Step 5: each target measured in every file"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
---

# Overview

Step 4 established, per assay, which ion of each target to extract, at what
m/z, and the range of retention times over which its confirmed spectra were
recorded. This notebook takes those back to every biological file of the assay:
for each file and target it reads the ion trace (EIC) at the m/z of the
confirmed spectra over that range and a margin beyond it, places a
retention-time window on the MS1 peak the confirmed spectra point to, looks for
a peak in that window, and integrates it if it passes the criteria below. Files
that were never fragmented are measured as well as the confirmed ones.

**Input:** `artifacts/hits-confirmed-all.csv`, `artifacts/datasets-confirmed.csv`,
`artifacts/extraction_windows.csv` (step 4: per assay and target, the ion,
its m/z and the range of confirmed retention times) and
`artifacts/assay_instruments.csv` (step 4: per assay, the instrument its raw files
record and the m/z tolerance it sets), `artifacts/dataset_metadata_table.csv`
(step 3), and the hand corrections in `curation/` (manual work, optional).

**Output:** `artifacts/sample_table_perfile.csv`, one row per file with, for each
target, whether it was detected, its peak area and the measurements behind both
(among them `auc0`, the area on a second baseline, for a check in step 6);
`artifacts/sample_table_all.csv`, the same collapsed to one row per sample;
`artifacts/positive_control.csv`; `artifacts/ms1_windows.csv`, the window in
which each target's peak was looked for, per assay; the figure
`figures/step5_windows.png` (and `.pdf`); each assay's extraction, cached in
`artifacts/per_dataset/<dataset_id>.rds`; and the lists of extracted files
(`artifacts/bio_*.rds`).

All compute goes through `run_ms1_for_dataset()` (`R/ms1.R`); this notebook
holds the loop, the extraction settings and the cross-dataset summary.
Datasets are
keyed by the combined `dataset_id` from `02`, and the loop pulls the bare
`deposit_id` and `assay` per row so repository downloads see plain accessions.

## Why not just use the MS2 matches?

A missing MS2 spectrum does not mean a missing compound. These deposits were
acquired by data-dependent acquisition, in which the instrument decides in real
time which ions to fragment, mostly by intensity and within a limited duty
cycle. A compound can therefore be present at a perfectly measurable level and
receive no MS2 scan at all, simply because the instrument was busy with
something else; run the same sample twice and the set of fragmented precursors
differs. A search resting on MS2 alone cannot separate absence from
non-selection, and the failure is not random. Selection is intensity-driven, so
the compounds most often missed are the low-abundance ones, often exactly the
observations a reanalysis is after. MS2-only prevalence is a lower bound, and a
biased one.

The two levels are complementary. MS2 gives identity, because fragments
constrain the structure and SIRIUS can tell isomers apart; MS1 gives coverage,
because it is recorded continuously with no selection step. Propagation uses
each for what it is good at: confirm the compound by MS2 in the files where it
happened to be fragmented, which pins both its mass and its retention time, then
measure that mass at that retention time in every file, including the many where
the instrument never fragmented it.

Proteomics met the same problem and solved it the same way. Match-between-runs
identifies a peptide by MS2 in the run where it was fragmented, then transfers
that identification to other runs by accurate mass and aligned retention time,
quantifying at MS1 where the precursor was never selected. It became standard
practice because DDA under-samples exactly as described above. Metabolomics
repository reanalysis has largely not adopted the equivalent step, and this
notebook is that step.

The trade is sensitivity for specificity, and it is real. MS2 can distinguish
compounds that share a mass and MS1 cannot, so propagation extends coverage only
as far as the chromatography preserves identity. Where a target has an isobar,
a retention-time window set too wide silently measures the wrong molecule, so
the method is only as good as that window, which is why so much of this
notebook is about it.

There is a second benefit, less obvious. Comparing deposits by their MS2
detection rates would substantially be comparing instruments: a deposit run on a
faster instrument with a deeper precursor list fragments more, so it looks more
prevalent. MS1 is acquired whatever the instrument chose to fragment, so
propagating to MS1 removes that confound. It does not make deposits fully
comparable, since sensitivity, matrix and chromatography still differ, but it
removes the one that most distorts a raw detection-rate comparison.

```{r setup}
source("../../R/setup.R")
source("../../R/metadata.R")
source("../../R/masst.R")   # compound_spec + PHENO, the domain constants
source("../../R/ms1.R")
library(Chromatograms)
library(MsQuality)
# The targets, in the order of compound.csv, and the prefix of their columns and
# fields (par, met; target_prefix() in R/setup.R).
roles     <- compound_spec$role
prefixes  <- target_prefix(roles)
n_targets <- length(roles)
```

```{r load-artifacts}
conf_final        <- read.csv("artifacts/hits-confirmed-all.csv",
                              check.names = FALSE)
conf_final$confidence <- factor(conf_final$confidence,
                                 levels = c("strict", "standard",
                                            "liberal", "not_confirmed"))
ds_summary        <- read.csv("artifacts/datasets-confirmed.csv", check.names = FALSE)
ds_meta_tbl       <- read.csv("artifacts/dataset_metadata_table.csv",
                              check.names = FALSE)
# Each assay's ion, m/z and range of confirmed retention times per target, set in
# step 4. The window itself is set here, on the MS1 trace.
if (!file.exists("artifacts/extraction_windows.csv"))
  stop("artifacts/extraction_windows.csv not found: render 04-confirmation.qmd first")
ext_windows       <- read.csv("artifacts/extraction_windows.csv", check.names = FALSE)
# Each assay's instrument and the m/z tolerance it sets, read from the raw files'
# headers in step 4.
if (!file.exists("artifacts/assay_instruments.csv"))
  stop("artifacts/assay_instruments.csv not found: render 04-confirmation.qmd first")
assay_inst        <- read.csv("artifacts/assay_instruments.csv", stringsAsFactors = FALSE)
```

# Files to extract

For each surviving `dataset_id` we build the list of biological MS files,
excluding QC injections, blanks, pooled QCs and reference standards. The three
repositories record that differently, so there is one loop per repository below.
`EXCL_ST` in `R/metadata.R` defines the regex of sample-type strings treated as
non-biological in all three.

```{r init-lists}
bio_files_per_ds  <- list()
bio_assays_per_ds <- list()
# Per file, does a metadata source cover it? Files without metadata are extracted
# for the record (only if they carry a confirmed hit) but excluded downstream.
bio_meta_per_ds   <- list()
```

The MassIVE extraction set comes from the **deposit's own file inventory**
(`massive_list_files` via `select_bio_files_massive()`), de-duplicated by
normalised name across all of the deposit's file trees. This matters because a
metadata catalogue alone under-covers a deposit: MSV000082493's blood
collection was added in a later deposit update that Pan-ReDU never indexed, and
MSV000082433's confirmed exposure sits on a plate Pan-ReDU never received.
Sourcing from the inventory keeps the extraction space equal to the MASST search
space.

A file enters the set if it is **either**:

- covered by a metadata source (Pan-ReDU ∪ submitter table, reconciled in `02`),
  with QC injections, blanks and reference standards removed by the *declared*
  sample-type column (`EXCL_ST` on `SampleType` / `MS_SampleType`); **or**
- carrying a confirmed MASST hit, whatever tree it sits in, so no confirmed
  exposure is ever lost, unless the metadata declares that file a blank, QC or
  standard: carryover gives a blank a confirmed hit, and that is not an exposure.
  For this the sample type is read on every row, also on a submitter row with
  too few fields to count as metadata: the sample type alone says what the file
  is.

Deposits also mislabel such files as samples, or leave them out of the metadata
altogether, so one narrow file-name rule (`EXCL_NONBIO`) removes pooled QCs, QC
injections, blanks and calibration standards from every repository's list, and
their confirmed hits from the anchors, whatever the metadata says.

Files that are neither are not downloaded. `bio_meta_per_ds` flags each kept
file: `has_metadata = TRUE` is analysable downstream; a confirmed-hit-only file
(`FALSE`) is extracted for the presence record but excluded from stratified
analysis. The metadata side is a per-deposit TSV from Pan-ReDU, auto-cached from
`redu.gnps2.org`, or a submitter-provided CSV; both carry `SampleType` and
`filename`, and where the CSV holds bare stems they are reconstructed against
the MASST `file_path` directory and extension.

```{r bio-files-massive}
# Loadable basename of a confirmed-hit file (strip the MsBackend cache prefix).
bare_hit_name <- function(x) sub("^MSV[0-9]+[_-]", "", basename(as.character(x)))
# The rows of a sample table whose declared sample type is a blank, QC or standard.
nonbio_rows <- function(nat) {
  nb <- rep(FALSE, nrow(nat))
  for (st in intersect(c("SampleType", "MS_SampleType"), names(nat)))
    nb <- nb | (grepl(EXCL_ST, nat[[st]], perl = TRUE) & !is.na(nat[[st]]))
  nb
}
msv_rows <- ds_summary[startsWith(ds_summary$deposit_id, "MSV"), ]

for (i in seq_len(nrow(msv_rows))) {
  ds  <- msv_rows$dataset_id[i]
  dep <- msv_rows$deposit_id[i]

  # A repository that does not answer stops the render here, rather than leave the
  # deposit without its metadata or its files (is_network_error() in R/metadata.R).
  meta_keys <- tryCatch({
    nat  <- msv_native_sample_table(dep)
    fcol <- intersect(MSV_FILENAME_COLS, names(nat))[1]
    if (is.na(fcol)) character() else norm_file_key(nat[[fcol]][!nonbio_rows(nat)])
  }, error = function(e) {
    if (is_network_error(e)) stop(e)
    character()
  })
  # The sample type is read on every submitter row, also on one too thin to count
  # as metadata above (MSV000082493's plate*_stdmixA files, declared blank_QC).
  nonbio_keys <- tryCatch({
    nat  <- suppressMessages(msv_native_sample_table(dep, min_fields = 0L))
    fcol <- intersect(MSV_FILENAME_COLS, names(nat))[1]
    if (is.na(fcol)) character() else norm_file_key(nat[[fcol]][nonbio_rows(nat)])
  }, error = function(e) {
    if (is_network_error(e)) stop(e)
    character()
  })
  hit_files <- unique(bare_hit_name(
    conf_final$source_file[conf_final$dataset_id == ds &
                           !is.na(conf_final$source_file)]))

  sel <- tryCatch(select_bio_files_massive(dep, meta_keys, hit_files, nonbio_keys),
                  error = function(e) {
                    if (is_network_error(e)) stop(e)
                    warning("[", ds, "] ", conditionMessage(e)); NULL })
  if (is.null(sel) || !nrow(sel)) {
    warning("[", ds, "] no biological files resolved - skipped"); next
  }

  bio_files_per_ds [[ds]] <- sel$file
  bio_meta_per_ds  [[ds]] <- stats::setNames(sel$has_metadata, sel$file)
  bio_assays_per_ds[[ds]] <- NA_character_   # MassIVE backend needs no assay name
  message(sprintf("[%s] %d files to extract (%d with metadata, %d record-only)",
                  ds, nrow(sel), sum(sel$has_metadata), sum(!sel$has_metadata)))
}
```

MetaboLights keeps the two halves apart: the study file `s_{id}.txt` lists
sample names and factors, while the spectral file names live in the assay file
`a_{id}_*.txt`. Each `dataset_id` is already scoped to one assay, so
`mtbls_assay_data()` is called for one assay at a time and restricted to the
sample rows `s_*.txt` marks as biological.

```{r bio-files-mtbls}
mtbls_rows <- ds_summary[startsWith(ds_summary$deposit_id, "MTBLS"), ]

for (i in seq_len(nrow(mtbls_rows))) {
  ds  <- mtbls_rows$dataset_id[i]
  dep <- mtbls_rows$deposit_id[i]
  an  <- mtbls_rows$assay[i]

  s_path <- file.path(META_DIR, paste0("s_", dep, ".txt"))
  if (!file.exists(s_path)) {
    warning("[", ds, "] no s_*.txt - skipped"); next
  }

  samples <- read.delim(s_path, check.names = FALSE)
  st_col  <- grep("sample type", names(samples), value = TRUE,
                  ignore.case = TRUE)[1]
  if (!is.na(st_col))
    samples <- samples[!grepl(EXCL_ST, samples[[st_col]], perl = TRUE), ]
  bio_names <- unique(samples[["Sample Name"]])

  # The cached copy of the assay table (fetch_assay_table() in R/metadata.R).
  adf <- fetch_assay_table(dep, an)
  if (is.null(adf) || !nrow(adf)) {
    warning("[", ds, "] empty assay table - skipped"); next
  }
  fc <- if ("Derived Spectral Data File" %in% names(adf) &&
            any(nzchar(trimws(as.character(adf[["Derived Spectral Data File"]])))))
          "Derived Spectral Data File"
        else "Raw Spectral Data File"

  rows  <- adf[adf[["Sample Name"]] %in% bio_names, , drop = FALSE]
  files <- unique(trimws(as.character(rows[[fc]])))
  files <- files[!is.na(files) & nchar(files) > 0]

  bio_files_per_ds [[ds]] <- files
  # MetaboLights files come from the deposit's own ISA-Tab manifest, so every
  # one already carries metadata (QC excluded above): all TRUE.
  bio_meta_per_ds  [[ds]] <- stats::setNames(rep(TRUE, length(files)), files)
  bio_assays_per_ds[[ds]] <- an
}
```

The Workbench packs it into one file. Each analysis file `<deposit>_AN*.txt`
carries a `SUBJECT_SAMPLE_FACTORS` block in which a subject's `RAW_FILE_NAME` is
a space-separated list, one name per analysis. The names follow the order of
`MS_run` unless the deposit is declared otherwise (`MWB_FILE_ORDER` in
`R/metadata.R`, see step 2). A name that is missing from the deposit, or that
appears in two of its folders, is left out.

```{r bio-files-mwb}
mwb_rows <- ds_summary[startsWith(ds_summary$deposit_id, "ST"), ]

for (i in seq_len(nrow(mwb_rows))) {
  ds  <- mwb_rows$dataset_id[i]
  dep <- mwb_rows$deposit_id[i]
  an_id <- sub(paste0("^", dep, "_"), "",
               sub("\\.txt$", "", mwb_rows$assay[i]))

  mwf <- file.path(META_DIR, mwb_rows$assay[i])
  if (!file.exists(mwf)) {
    warning("[", ds, "] no ", basename(mwf), " - skipped"); next
  }

  ln  <- readLines(mwf, warn = FALSE)
  ssf <- ln[startsWith(ln, "SUBJECT_SAMPLE_FACTORS") & !startsWith(ln, "#")]
  if (!length(ssf)) {
    warning("[", ds, "] no SUBJECT_SAMPLE_FACTORS - skipped"); next
  }

  ssf_parsed <- lapply(strsplit(ssf, "\t"), function(p) {
    raw_str   <- if (length(p) >= 5)
                   regmatches(p[5], regexpr("RAW_FILE_NAME=[^\t]+", p[5]))
                 else character(0)
    raw_str   <- sub("RAW_FILE_NAME=", "",
                     if (length(raw_str)) raw_str else "")
    raw_files <- trimws(unlist(strsplit(raw_str, "\\s+")))
    list(factors   = if (length(p) >= 4) p[4] else NA_character_,
         raw_files = raw_files[nchar(raw_files) > 0])
  })
  keep_qc    <- !vapply(ssf_parsed,
                        function(r) grepl(EXCL_ST, r$factors, perl = TRUE),
                        logical(1))
  ssf_parsed <- ssf_parsed[keep_qc]

  # RAW_FILE_NAME is a positional list, one entry per analysis.
  an_idx <- match(an_id, mwb_analysis_order(dep))
  if (is.na(an_idx))
    stop("[", ds, "] ", an_id, " not in the deposit's analysis list")

  files <- unique(vapply(ssf_parsed, function(r) {
    # A row with a single name lists this analysis's own file.
    if (length(r$raw_files) == 1) r$raw_files[1]
    else if (length(r$raw_files) >= an_idx) r$raw_files[an_idx] else NA_character_
  }, character(1)))
  files <- files[!is.na(files) & nchar(files) > 0]
  files <- files[files %in% mwb_listed_files(dep)]

  bio_files_per_ds [[ds]] <- files
  # Workbench files come from the deposit's own mwTab manifest (QC excluded), so
  # every one carries metadata: all TRUE.
  bio_meta_per_ds  [[ds]] <- stats::setNames(rep(TRUE, length(files)), files)
  bio_assays_per_ds[[ds]] <- an_id
}
```

The name rule, applied once more to every repository's list, since only the
MassIVE selection applies it itself:

```{r bio-files-nonbio}
for (ds in names(bio_files_per_ds)) {
  nb <- is_nonbio_file(bio_files_per_ds[[ds]])
  if (!any(nb)) next
  message(sprintf("[%s] dropped %d QC, blank or standard files", ds, sum(nb)))
  keep <- bio_files_per_ds[[ds]][!nb]
  bio_files_per_ds[[ds]] <- keep
  if (!is.null(bio_meta_per_ds[[ds]]))
    bio_meta_per_ds[[ds]] <- bio_meta_per_ds[[ds]][names(bio_meta_per_ds[[ds]]) %in% keep]
}
```

A study deposited twice holds some files in both deposits; each is extracted
once, under the deposit step 3 kept its hits in (`is_copy()`):

```{r bio-files-copies}
for (ds in names(bio_files_per_ds)) {
  cp <- is_copy(bio_files_per_ds[[ds]], rep(ds, length(bio_files_per_ds[[ds]])),
                function(d) bio_files_per_ds[[d]])
  if (!any(cp)) next
  message(sprintf("[%s] dropped %d files held again by a related deposit", ds, sum(cp)))
  keep <- bio_files_per_ds[[ds]][!cp]
  bio_files_per_ds[[ds]] <- keep
  if (!is.null(bio_meta_per_ds[[ds]]))
    bio_meta_per_ds[[ds]] <- bio_meta_per_ds[[ds]][names(bio_meta_per_ds[[ds]]) %in% keep]
}
```

What the three loops leave, per dataset:

```{r bio-summary}
#| echo: false
bio_counts <- data.frame(
  dataset_id  = names(bio_files_per_ds),
  n_bio_files = vapply(bio_files_per_ds, length, integer(1)),
  assay       = vapply(names(bio_files_per_ds),
                       function(d) as.character(bio_assays_per_ds[[d]]),
                       character(1)),
  row.names   = NULL
) |>
  dplyr::left_join(ds_summary[, c("dataset_id", "deposit_id", "repository",
                                  "chromatography")],
                   by = "dataset_id")
bio_counts
```

```{r write-file-lists}
saveRDS(bio_files_per_ds,  "artifacts/bio_files_per_ds.rds")
saveRDS(bio_meta_per_ds,   "artifacts/bio_meta_per_ds.rds")
# Does a metadata source cover each file? Keyed dataset_id|basename so identical
# basenames in two deposits cannot collide.
has_metadata_lookup <- unlist(lapply(names(bio_meta_per_ds), function(ds) {
  v <- bio_meta_per_ds[[ds]]
  stats::setNames(as.logical(v), paste0(ds, "|", basename(names(v))))
}))
```


# Extraction settings

```{r extraction-settings}
DATASETS_TO_RUN <- names(bio_files_per_ds)

RT_PAD_S <- ext_windows$rt_pad_s[1]   # s padded each side of the confirmed-RT range

# The m/z tolerance of each assay, set in step 4 by the instrument its raw files
# record. An assay step 4 gives no tolerance for is extracted at PPM_DEFAULT.
ids_run <- as.character(DATASETS_TO_RUN)
tol <- data.frame(dataset_id = ids_run,
                  class = assay_inst$class[match(ids_run, assay_inst$dataset_id)],
                  ppm   = assay_inst$ppm[match(ids_run, assay_inst$dataset_id)])
no_inst <- is.na(tol$ppm)
if (any(no_inst))
  message("not in artifacts/assay_instruments.csv, extracted at ", PPM_DEFAULT, " ppm: ",
          paste(tol$dataset_id[no_inst], collapse = ", "))
tol$class[no_inst] <- "unknown"
tol$ppm[no_inst]   <- PPM_DEFAULT
# The target the text below gives the tolerance in Da for: the one of highest m/z.
heaviest <- if (n_targets == 1) "the target's" else
  if (n_targets == 2) "the heavier target's" else "the heaviest target's"
```

Two settings bound the part of the raw data that is read for a target. Both are
set in step 4 and read back here:

- **m/z tolerance**, per assay. When we go looking for a compound at its known
  mass, real instruments are slightly off, so we accept anything within a
  tolerance of it, and how far off they are depends on the mass analyser. Step 4
  reads each assay's instrument from its raw files and sets the tolerance from
  it (`artifacts/assay_instruments.csv`): `r PPM_BY_CLASS[["orbitrap"]]` ppm on
  an Orbitrap (±`r sprintf("%.4f", PPM_BY_CLASS[["orbitrap"]] * 1e-6 * max(compound_spec$mz_pos))` Da
  at `r heaviest` m/z), where the scans across a true peak sit within a
  few ppm of its mass, and `r PPM_BY_CLASS[["tof"]]` ppm on a time-of-flight
  instrument (±`r sprintf("%.4f", PPM_BY_CLASS[["tof"]] * 1e-6 * max(compound_spec$mz_pos))` Da),
  where the scans on a peak's edges scatter much further. An assay whose
  instrument cannot be told gets `r PPM_DEFAULT` ppm. Too narrow a tolerance cuts
  real signal off a peak's edges; too wide a one lets a neighbouring ion of
  nearly the same mass into the trace. Of the `r nrow(tol)` assays extracted
  here, `r sum(tol$class == "orbitrap")` are extracted at
  `r PPM_BY_CLASS[["orbitrap"]]` ppm, `r sum(tol$class == "tof")` at
  `r PPM_BY_CLASS[["tof"]]` ppm and `r sum(tol$class == "unknown")`, whose
  instrument is not known, at `r PPM_DEFAULT` ppm (table below).
- **`RT_PAD_S` = `r RT_PAD_S`**: seconds added on each side of
  step 4's range of confirmed retention times. Step 4 sets it, since it builds
  the range, and it is read back here. It bounds the stretch of trace that is
  read, not the window: the window is set on the MS1 trace and padded by
  `APEX_PAD` (see *The retention-time window* below).

```{r extraction-tolerance}
#| echo: false
knitr::kable(data.frame(assay = short_ds(tol$dataset_id), instrument = tol$class,
                        `m/z tolerance (ppm)` = tol$ppm, check.names = FALSE),
             caption = "The m/z tolerance of each assay extracted here, set by its instrument in step 4.")
```

There is no peak-edge padding: the picker integrates valley to valley, so how
far it reaches is set by the peak's own shape rather than by a fixed margin.


## How the peaks are picked and filtered

The picker lives in `R/ms1.R` and runs in five stages, in this order.

- **Set the window** (`acceptance_window`). Rather than a fixed span, each
  confirmed MS2 hit is anchored to the apex of the MS1 peak it sits on, and the
  window brackets the best-supported cluster of those apexes, padded by
  `APEX_PAD` = `r APEX_PAD` s (see *The retention-time window* below). The picker starts
  from the local maxima inside the window; the apex it settles on may lie up to
  `APEX_PAD` s outside it, and how far the integration then reaches is set by
  the peak.
- **Integrate** (`pick`). Each local maximum is integrated valley to valley,
  walking out from the apex until the trace returns to baseline or meets a deep
  valley, merging shallow shoulders and sub-`r PROX_S` s doublets and stopping at dead
  gaps where the EIC dropped out. Area is trapezoidal above a straight line
  between the two boundary scans; prominence, SNR and the nearest-taller-neighbour
  ratio are computed on the whole peak.
- **Choose which peak.** Where several fall inside the window, the dominant one
  is chosen by area weighted by a gaussian in its distance from the expected RT
  (`RT_SIGMA` = `r RT_SIGMA` s). This is the isomer defence at the selection level: a
  larger interferent elsewhere in the window loses to the on-RT peak. HILIC
  keeps pure area-selection, its windows having been audited peak by peak.
- **Accept or reject** (`accept_peak`), conditional on chromatography. Both
  modes require an apex
  of `r format(MIN_INT, big.mark = ",")` counts, an SNR of `r SNR_MIN_RP` on
  reverse phase or `r SNR_MIN_HILIC` on HILIC, an FWHM inside
  [`r FWHM_BAND_RP[1]`, `r FWHM_BAND_RP[2]`] s on reverse phase or of at least
  `r FWHM_BAND_HILIC[1]` s on HILIC. HILIC also requires return-to-baseline
  and no much-taller neighbour; reverse phase drops both, because they rejected
  sharp peaks truncated on one flank and real peaks sitting beside a larger
  neighbour, and instead requires a 3-point peak to clear its own baseline. The
  reverse-phase floors were set on audited peaks: the SNR floor sits below every
  real peak seen and the FWHM ceiling above the widest.
- **Measure** (`add_msquality`). The picker chooses the peak and its boundaries;
  **MsQuality** then measures it, so the reported width, FWHM and shape are mzQC
  controlled-vocabulary quantities rather than local re-implementations. Handed
  our boundaries, its `peakWidth()` and `xicFwhm()` reproduce the local values
  exactly on reverse phase; on HILIC the two FWHMs differ, because ours takes
  the outermost half-maximum crossings of a merged cluster while MsQuality takes
  50% of the maximum inside the region. `gaussianSimilarity()` is recorded per
  peak but never gated on, since it needs at least 5 points and is NA on the
  sparsest. Apex intensity and SNR stay local: both describe *which* peak was
  selected, and MsQuality has no way of being told that.

A peak whose trace empties before it falls to half height is measured out to
those empty scans and kept as `truncated`. An unresolved neighbouring ion does
this: where the target weakens the instrument merges the two into one centroid
that drifts out of the mass window, so the trace reads zero while the compound
is still eluting. Only a trace shaped like a peak qualifies: no empty scan
inside it, rising to the apex and falling after it. A dip of up to
`r 100 * TRUNC_DIP`% of the apex on the way up or down is allowed, as noise from
one scan to the next. The peak counts as a detection, but its area is a lower
bound, so no area is reported.

Each peak also records a second area, `auc0`, for a check in step 6. Where the
scan just outside a boundary is empty, the signal rises from nothing there, and
a line starting at the boundary scan leaves out the part of the peak below it;
`auc0` starts the line from zero on such a side instead. Where neither side
borders an empty scan, the two areas are the same. `auc0` plays no part in
detection or in choosing the peak, and it is missing wherever the area is.

One further rejection happens after the picker, once every channel exists, and
only for a compound whose `compound.csv` row names an isobar: another compound
of the same mass that elutes where a partner compound does, and that spectral
confirmation may not reject either. The partner is the target the row's
`isobar_at` names; with two targets it may be left blank, and is then the other
one. `resolve_isobars()` drops such a channel's peak when it sits nearer the
*partner's* retention time than its own, each compound's retention time being
the median of the anchors its window rests on. A channel with no declared
isobar is never tested, because a peak drifting toward another target's
retention time is still its own compound. Where either of the two windows is
uncertain (its confirmed spectra point to two equally supported MS1 peaks), the
channel is not tested.

The gates were tuned by auditing peaks in a reverse-phase assay and a HILIC
assay, each peak plotted and labelled by eye as signal, noise or interferent.
Detected peaks carrying no MS2 confirmation of their own were then reviewed
against the panels `R/peak_review_panels.R` saves under `peaks_review_v2/`,
which also name the gate rule behind every rejection. Peaks that survive every
rule while still being wrong, where noise and signal overlap on every shape
metric, are corrected by hand (below). For a genuinely ambiguous peak a
transparent per-peak correction is more honest than a gate no one can fully reason
about. This is inspection-backed calibration rather than a formal benchmark, but
every decision is auditable from the saved panels.

::: {.callout-warning title="Manual work"}
Three optional files in `curation/` hold corrections made by hand after looking at
the peak panels: `rt_apex_overrides.csv`, the retention time to use where the
picker took the wrong apex, `ms1_drop_list.csv`, peaks to drop, and
`ms1_accept_list.csv`, peaks to pick around a given apex instead of in the
window (the peak criteria still apply). Every entry carries its reason. An application without them is extracted without corrections.
:::

```{r peak-criteria}
#| echo: false
tol_text <- sprintf("%g ppm on an Orbitrap, %g ppm on a time-of-flight, %g ppm where the instrument is not known",
                    PPM_BY_CLASS[["orbitrap"]], PPM_BY_CLASS[["tof"]], PPM_DEFAULT)
gate <- data.frame(
  criterion = c("ion extracted",
                "m/z tolerance",
                "retention-time window",
                "peak chosen among several in the window",
                "apex intensity",
                "signal-to-noise ratio",
                "width at half maximum",
                "return to baseline (prominence, fraction of the apex)",
                "taller neighbour",
                "peak that leaves the m/z window before half height"),
  `reverse phase` = c(
    "molecular ion if confirmed, else the most frequent confirmed adduct",
    tol_text,
    sprintf("MS1 apexes of the confirmed spectra (best-supported cluster), %g s either side", APEX_PAD),
    sprintf("largest area, weighted by distance from the window centre (Gaussian, sd %g s)", RT_SIGMA),
    sprintf("at least %g counts", MIN_INT),
    sprintf("at least %g", SNR_MIN_RP),
    sprintf("%g to %g s", FWHM_BAND_RP[1], FWHM_BAND_RP[2]),
    sprintf("at least %g, for 3-point peaks only", PROM_MIN),
    "not tested",
    "detected, no area"),
  HILIC = c(
    "same",
    tol_text,
    sprintf("MS1 apexes of the confirmed spectra (best-supported cluster), %g s either side", APEX_PAD),
    "largest area",
    sprintf("at least %g counts", MIN_INT),
    sprintf("at least %g", SNR_MIN_HILIC),
    sprintf("at least %g s", FWHM_BAND_HILIC[1]),
    sprintf("at least %g", PROM_MIN),
    sprintf("none more than %g times the apex within %g s", NBR_MAX, NEAR_S),
    "detected, no area"),
  check.names = FALSE)
knitr::kable(gate, caption = "Peak criteria, read from `R/ms1.R` and the settings above.")
```

## The retention-time window

Retention time (RT) is *when* a compound comes off the chromatography column. We
only look for a peak inside an RT window, so the window has to sit over the
peak. It is set in two stages.

Step 4 gives, per assay and target, the range of retention times at which the
target's confirmed spectra were recorded: earliest to latest, at the best tier
the target reached, padded each side by `RT_PAD_S`, in
`artifacts/extraction_windows.csv`. That range is not the window. An MS2 scan is
triggered whenever the precursor is intense enough, so the confirmed spectra
scatter along the peak and down its tail, and can include a second peak of the
same m/z; the range can be several times wider than the peak. It says where to
read the trace: each file's trace is read over the range, and `WIDE_EXTRA` =
`r WIDE_EXTRA` s beyond it on each side so the peak walk has data.

```{r most-fragmented}
#| echo: false
# The most times one file was fragmented on one target (distinct confirmed MS2
# retention times), over the assays extracted here, for the example below.
frag <- conf_final[conf_final$dataset_id %in% names(bio_files_per_ds) &
                   !is.na(conf_final$rt), ]
frag <- unique(data.frame(
  key = paste(frag$dataset_id, frag$role, basename(as.character(frag$source_file))),
  rt  = frag$rt))
n_frag_max <- if (nrow(frag)) max(table(frag$key)) else 0L
```

The window is then set on the MS1 trace (`acceptance_window()`). Each confirmed
MS2 retention time is anchored to the apex of the MS1 peak it sits on, in the
trace of the file it was recorded in (`snap_one()`). From the MS1 scan nearest
the MS2 scan, the trace is followed along both flanks with the walk the picker
uses to integrate a peak (*Integrate* above): a shallow shoulder, or a maximum
within `PROX_S` = `r PROX_S` s of the start, is taken in, and a real valley, the
baseline or a gap in the trace ends the walk. The anchor is the highest scan of
that stretch. A spectrum fired on a peak's tail therefore anchors at the top of
its own peak, not at a taller peak beside it. An anchor more than `SNAP_W` =
`r SNAP_W` s from its MS2 retention time is dropped, and a hand override in
`curation/` gives the apex where one exists. Each file then gives one anchor,
its most intense anchored apex, so a file fragmented `r n_frag_max` times on one
target (the most of any file here) weighs no more than a file fragmented once.
The anchors with the most neighbours within
`CLUST_TOL` = `r CLUST_TOL` s mark the centre of the best-supported cluster; the
cluster is every anchor within `CLUST_TOL` of that centre, and the window spans
it, padded by `APEX_PAD` = `r APEX_PAD` s each side. An anchor outside the cluster
is left out, so a single file anchored on another peak cannot stretch the window
over the whole run. It then extends, step by
step, over accepted peak apexes within `COHERE_S` = `r COHERE_S` s beyond its
edge, up to 2 × `APEX_PAD` = `r 2 * APEX_PAD` s from the cluster, so a peak just outside the edge is
not cut off. Where two clusters are equally supported, the assay does not say
which retention time is the compound: the window spans every anchor, both
clusters included, and is flagged uncertain. Where no confirmed spectrum gives
an anchor, the window falls back to the range itself. The window, the number
of files with an anchor (in the kept cluster or not) and the uncertain
flag are kept with each assay's extraction, and *Extraction windows* below
writes and draws them.

Before an assay is extracted, each target must have a confirmation on one of the
files extracted here, or the assay is skipped with the reason
(`extraction_window()`): a range resting only on other files would import
another matrix's retention time. For the same reason, if a spectrum that set
step 4's range lies on a file not extracted here, the range is set again from
the extracted files' spectra alone, and the log says so.

Where a target has an isobar (declared in `compound.csv`), two defences keep the
measurement off it: the proximity-weighted selection above, under which the
on-RT peak beats a larger interferent inside the window, and the window itself,
kept to a modest pad so the isobar rarely enters it. An application with a
declared isobar and an independent ground truth measures what each defence is
worth in its own checks (`_05-probe-checks.qmd`).

# Per-dataset extraction

For each assay, `run_ms1_for_dataset()` loads the MS1 spectra, pulls each
compound's extracted-ion chromatogram (EIC) at the confirmed mass, within the
assay's m/z tolerance, over step 4's (widened) retention-time range, sets the
window on it, then **picks and
integrates the on-RT peak per file** (the picker described above), applies the
manual drop-list, flags files whose peak width is an outlier, joins the sample
metadata, and emits a per-file table
(`sample_tbl`) with, for each target, whether it was detected and its peak
area.

```{r run-all}
#| echo: false
# Cached per assay (delete an assay's RDS to force a re-run), keyed on step 4's ion,
# m/z and range and the m/z tolerance: a change in any of them re-reads the raw files.
CACHE_DIR <- "artifacts/per_dataset"
dir.create(CACHE_DIR, showWarnings = FALSE, recursive = TRUE)

cfg_is_current <- function(cached, want) {
  if (is.null(cached$cfg)) return(FALSE)
  # With two targets: assay, adduct_par, adduct_met, mz_par, mz_met, rt_par_win,
  # rt_met_win, ppm.
  k <- c("assay", paste0("adduct_", prefixes), paste0("mz_", prefixes),
         paste0("rt_", prefixes, "_win"), "ppm")
  isTRUE(all.equal(cached$cfg[k], want[k], tolerance = 1e-6))
}

results <- list()
SKIPPED_DATASETS <- character(0)   # extraction failed
NO_WINDOW        <- character(0)   # no confirmed range on the files extracted here
for (ds in DATASETS_TO_RUN) {
  # Step 4's ion, m/z and range for this assay; an assay with a target unconfirmed on
  # the files extracted here is skipped.
  cfg <- tryCatch(extraction_window(ds, ext_windows, conf_final, bio_files_per_ds[[ds]]),
                  error = function(e) {
                    message("  SKIPPED ", ds, ", no confirmed range on the files extracted here: ",
                            conditionMessage(e))
                    NULL
                  })
  if (is.null(cfg)) { NO_WINDOW <- c(NO_WINDOW, ds); next }
  # The m/z tolerance step 4 set from the assay's instrument.
  cfg$ppm <- tol$ppm[tol$dataset_id == ds]
  cache_path <- file.path(CACHE_DIR, paste0(ds, ".rds"))
  cached <- if (file.exists(cache_path)) readRDS(cache_path) else NULL
  # A cache that records no tolerance was extracted at 20 ppm. Record that, so an
  # assay whose tolerance is 20 ppm is not read again.
  if (!is.null(cached$cfg) && is.null(cached$cfg$ppm)) {
    cached$cfg$ppm <- 20
    if (cfg_is_current(cached, cfg)) saveRDS(cached, cache_path)
  }
  if (!is.null(cached) && !cfg_is_current(cached, cfg)) {
    message("  cache for ", ds, " was built on a different ion, m/z, range or m/z tolerance; re-extracting")
    cached <- NULL
  }
  # Picker, gate, confirmed anchors or hand corrections changed: re-pick from the
  # stored EICs, reading no raw file.
  if (!is.null(cached) &&
      (!identical(cached$picker, PICKER_VERSION) ||
       !identical(cached$is_hilic, row_is_hilic(dataset_meta_row(ds_meta_tbl, ds))) ||
       !identical(cached$anchors, anchor_key(conf_final, ds)) ||
       !identical(cached$curation, curation_key(ds)))) {
    message("  ", ds, ": re-picking from the stored EICs (picker, gate, anchors or hand corrections changed)")
    cached <- repick_cached(cached, conf_final, bio_files_per_ds[[ds]], ds_meta_tbl)
    saveRDS(cached, cache_path)
  }
  if (!is.null(cached)) {
    message("========== ", ds, " (cached) ==========")
    results[[ds]] <- cached
  } else {
    message("\n========== ", ds, " ==========")
    # A repository that does not answer stops the render (the next one resumes from the
    # cache); any other failure skips the dataset, recorded in SKIPPED_DATASETS.
    res <- tryCatch(
      run_ms1_for_dataset(
        dataset_id       = ds,
        conf_final       = conf_final,
        bio_files_per_ds = bio_files_per_ds,
        ds_meta_tbl      = ds_meta_tbl,
        ppm              = cfg$ppm,
        cfg              = cfg),
      error = function(e) {
        if (is_network_error(e)) stop(e)
        message("  SKIPPED ", ds, ", extraction failed: ", conditionMessage(e))
        NULL
      })
    if (is.null(res)) { SKIPPED_DATASETS <<- c(SKIPPED_DATASETS, ds); next }
    results[[ds]] <- res
    saveRDS(results[[ds]], cache_path)
  }
}
if (length(NO_WINDOW))
  message("\n", length(NO_WINDOW),
          " assay(s) SKIPPED (no confirmed range on the files extracted here; ",
          "re-confirm on them): ", paste(NO_WINDOW, collapse = ", "))
if (length(SKIPPED_DATASETS))
  message("\n", length(SKIPPED_DATASETS),
          " dataset(s) SKIPPED (extraction failed, see above): ",
          paste(SKIPPED_DATASETS, collapse = ", "))
```

## Extraction windows

The window in which each target's peak was looked for, per extracted assay, is
written to `artifacts/ms1_windows.csv`: its bounds (`rt_lo`, `rt_hi`), the number
of files whose confirmed spectra snapped to an MS1 peak (`n_anchor_files`; the
window is set from the best-supported cluster of these, so some can lie outside
it), and whether it is uncertain. The range of confirmed retention times the
trace was read over is kept beside it (`ms2_lo`, `ms2_hi`) for the record.

```{r ms1-windows}
# The window of each assay and target, set on the MS1 trace by acceptance_window();
# n_anchor_files is NA where no confirmed spectrum snapped, and the window is the range.
target_name <- setNames(compound_spec$name, compound_spec$role)
ids <- sort(as.character(names(results)), method = "radix")
ms1_window_rows <- function(ds) {
  r     <- results[[ds]]
  stopifnot(prefixes %in% names(r$rt_anchor))
  a     <- r$rt_anchor[prefixes]
  # One cfg field per target, in target order: field("mz_%s") is list(mz_par, mz_met).
  field <- function(k) lapply(prefixes, function(p) r$cfg[[sprintf(k, p)]])
  ms2   <- field("rt_%s_win")
  data.frame(dataset_id     = ds, assay = as.character(r$cfg$assay), role = roles,
             target         = target_name[roles],
             ion            = unlist(field("adduct_%s")),
             mz             = unlist(field("mz_%s")),
             rt_lo          = unlist(lapply(a, function(x) x$win[1])),
             rt_hi          = unlist(lapply(a, function(x) x$win[2])),
             n_anchor_files = vapply(a, function(x)
               if (is.null(x$n_anchor_files)) NA_integer_ else as.integer(x$n_anchor_files),
               integer(1)),
             uncertain      = vapply(a, function(x) isTRUE(x$uncertain), logical(1)),
             ms2_lo         = unlist(lapply(ms2, function(w) w[1])),
             ms2_hi         = unlist(lapply(ms2, function(w) w[2])),
             row.names = NULL)
}
ms1_windows <- do.call(rbind, lapply(ids, ms1_window_rows))
# No assay was extracted: the same table, empty.
if (is.null(ms1_windows))
  ms1_windows <- data.frame(dataset_id = character(), assay = character(), role = character(),
                            target = character(), ion = character(), mz = numeric(),
                            rt_lo = numeric(), rt_hi = numeric(), n_anchor_files = integer(),
                            uncertain = logical(), ms2_lo = numeric(), ms2_hi = numeric())
write.csv(ms1_windows, "artifacts/ms1_windows.csv", row.names = FALSE)
# Height in inches of the figure below: one row per assay, 0.15 in per target,
# and a line of key for each kind of marked window (uncertain, fallback) present.
n_key <- any(ms1_windows$uncertain) + any(is.na(ms1_windows$n_anchor_files))
fig_h <- 1.3 + 0.15 * n_targets * length(ids) + 0.2 * n_key
ms1_windows
```

The figure shows, per assay, the window in which each target's peak was looked
for (A, the shaded band) with the retention times of its confirmed spectra
(points), and how far each confirmed precursor m/z sits from the target's exact
mass (B). The column on the right counts the confirmed spectra of each target,
`r knitr::combine_words(roles, and = " then ")`. The band is set on MS1, the
points are MS2 trigger times. Those scatter along the peak's tail and can fall
on a second peak of the
same m/z, so points can lie outside the band. A detected peak can reach past the
band too: its apex may sit up to `r APEX_PAD` s outside it, and its area runs
valley to valley. An asterisk after an assay's name marks a window to read with
care. A dashed outline is an uncertain window: the confirmed spectra point to two
equally supported MS1 peaks, and the window spans both. A dotted outline is a
fallback: no confirmed spectrum snapped to an MS1 peak, so the band is the range
the trace was read over.

```{r fig-step5-windows, fig.width = 6.3, fig.height = fig_h}
FIG_DIR <- "figures"; dir.create(FIG_DIR, showWarnings = FALSE)
figp    <- function(f) file.path(FIG_DIR, f)

# A window to read with care is marked: uncertain (two equally supported MS1
# peaks, both spanned) or fallback (nothing snapped, so it is the range read).
win <- ms1_windows
win$flag <- ifelse(win$uncertain, "uncertain",
                   ifelse(is.na(win$n_anchor_files), "fallback", ""))

# One row per extracted assay, the first at the top, with a band per target around
# its line; an assay with a marked window is labelled with an asterisk.
flagged <- unique(win$dataset_id[win$flag != ""])
lab   <- function(ds) paste0(short_ds(ds, ids), ifelse(ds %in% flagged, " *", ""))
lev   <- rev(lab(ids))
off   <- ((n_targets + 1) / 2 - seq_len(n_targets)) * 0.72 / n_targets
half  <- 0.28 / n_targets
row_y <- function(ds, role)
  match(lab(ds), lev) + off[match(role, roles)]
# Blue and red for the first two targets, then a palette.
tcol  <- c("#2166AC", "#D6604D", grDevices::hcl.colors(max(n_targets - 2, 0), "Dark 3"))
tcol  <- setNames(tcol[seq_len(n_targets)], target_name[roles])

pts <- conf_final[conf_final$dataset_id %in% ids & !is.na(conf_final$rt), ]
pts$y      <- row_y(pts$dataset_id, pts$role)
pts$target <- factor(target_name[pts$role], names(tcol))
win$y      <- row_y(win$dataset_id, win$role)
win$row    <- match(lab(win$dataset_id), lev)
win$target <- factor(win$target, names(tcol))
win$n_spectra <- vapply(seq_len(nrow(win)), function(i)
  sum(conf_final$dataset_id == win$dataset_id[i] & conf_final$role == win$role[i]),
  integer(1))

rows <- scale_y_continuous(breaks = seq_along(lev), labels = lev,
                           limits = c(0.5, length(lev) + 0.5), expand = c(0, 0))
jit  <- position_jitter(width = 0, height = 0.1 / n_targets, seed = 1)
look <- list(scale_colour_manual(values = tcol, name = NULL, drop = FALSE),
             scale_fill_manual(values = tcol, guide = "none"),
             guides(colour = guide_legend(override.aes = list(size = 2, alpha = 1))),
             theme_minimal(base_size = 9),
             theme(panel.grid.minor = element_blank(),
                   plot.title = element_text(size = 9, face = "bold")))

p_rt <- ggplot() +
  geom_rect(data = win, aes(xmin = rt_lo, xmax = rt_hi, ymin = y - half,
                            ymax = y + half, fill = target), alpha = 0.25) +
  geom_rect(data = win[win$flag != "", ], aes(xmin = rt_lo, xmax = rt_hi, ymin = y - half,
                                              ymax = y + half, colour = target, linetype = flag),
            fill = NA, linewidth = 0.3, show.legend = FALSE) +
  geom_point(data = pts, aes(rt, y, colour = target),
             size = 0.8, alpha = 0.5, stroke = 0.3, position = jit) +
  scale_linetype_manual(values = c(uncertain = 2, fallback = 3), guide = "none") +
  rows + look +
  labs(x = "retention time (s)", y = NULL, title = "A   Retention time")

# Each assay's own m/z tolerance, dotted: 5 ppm on an Orbitrap, 20 ppm otherwise.
tolr <- unique(win[, c("dataset_id", "row")])
tolr$ppm <- tol$ppm[match(tolr$dataset_id, tol$dataset_id)]
tolr <- rbind(transform(tolr, x = -ppm), transform(tolr, x = ppm))
p_mz <- ggplot(pts[!is.na(pts$mz_ppm_dev), ],
               aes(mz_ppm_dev, y, colour = target)) +
  geom_vline(xintercept = 0, colour = "grey50", linewidth = 0.3) +
  geom_segment(data = tolr, aes(x = x, xend = x, y = row - 0.45, yend = row + 0.45),
               inherit.aes = FALSE, colour = "grey60", linetype = 3, linewidth = 0.3) +
  geom_point(size = 0.8, alpha = 0.5, stroke = 0.3, position = jit) +
  rows + look + theme(axis.text.y = element_blank()) +
  labs(x = "m/z deviation (ppm)", y = NULL, title = "B   Precursor m/z")

# The counts get a panel of their own, so they never sit on a point: two targets
# read "first / second" on the row's line, more sit at their target's band.
if (n_targets == 2) {
  p_n <- ggplot(win, aes(x = ifelse(role == roles[1], 0.45, 0.55), y = row,
                         label = prettyNum(n_spectra, big.mark = ","), colour = target,
                         hjust = ifelse(role == roles[1], 1, 0))) +
    geom_text(size = 2.8, show.legend = FALSE) +
    geom_text(data = win[win$role == roles[1], ], aes(x = 0.5, y = row, label = "/"),
              inherit.aes = FALSE, colour = "grey40", size = 2.8)
} else {
  p_n <- ggplot(win, aes(x = 0.5, y = y, label = prettyNum(n_spectra, big.mark = ","),
                         colour = target)) +
    geom_text(size = 2.8, show.legend = FALSE)
}
p_n <- p_n +
  scale_x_continuous(limits = c(0, 1), expand = c(0, 0)) +
  rows + look +
  theme(axis.text = element_blank(), panel.grid = element_blank()) +
  labs(x = NULL, y = NULL, title = "Spectra")

# The key to the marks sits in the figure, so a copy of it carries its own; one
# line per kind present.
key <- c(uncertain = "* dashed: uncertain window, spanning two equally supported MS1 peaks",
         fallback  = "* dotted: fallback window, the range read (no confirmed spectrum snapped to an MS1 peak)")
key <- key[names(key) %in% win$flag]

p_windows <- patchwork::wrap_plots(p_rt, p_mz, p_n, widths = c(1.5, 1, 0.45),
                                   guides = "collect") +
  patchwork::plot_annotation(
    caption = if (length(key)) paste(key, collapse = "\n"),
    theme   = theme(plot.caption = element_text(size = 7, colour = "grey30", hjust = 0))) &
  theme(legend.position = "bottom", legend.box = "vertical",
        legend.margin = margin(0, 0, 0, 0), legend.spacing.y = unit(2, "pt"),
        legend.box.spacing = unit(4, "pt"))

# The note under the figure names each marked window.
say <- function(f, what) {
  r <- win[win$flag == f, , drop = FALSE]
  if (nrow(r)) paste0(what, " here: ", paste(sprintf("%s (%s)", short_ds(r$dataset_id, ids),
                                                    as.character(r$target)), collapse = ", "), ".")
}
unc_note <- paste(c(say("uncertain", "Uncertain windows"), say("fallback", "Fallback windows")),
                  collapse = " ")
if (!nzchar(unc_note)) unc_note <- "No window here is uncertain or a fallback."

if (length(lev)) {
  ggsave(figp("step5_windows.png"), p_windows, width = 6.3, height = fig_h, dpi = 200)
  ggsave(figp("step5_windows.pdf"), p_windows, width = 6.3, height = fig_h)
  p_windows
}
```

The precursor m/z deviations of panel B, by the assays' m/z tolerance:

```{r mz-dev-by-tolerance}
#| echo: false
d <- pts[!is.na(pts$mz_ppm_dev), ]
d$tolerance_ppm <- tol$ppm[match(d$dataset_id, tol$dataset_id)]
do.call(rbind, lapply(split(d, d$tolerance_ppm), function(x) data.frame(
  tolerance_ppm = x$tolerance_ppm[1], assays = length(unique(x$dataset_id)),
  spectra = nrow(x), beyond_5_ppm = sum(abs(x$mz_ppm_dev) > 5),
  max_abs_ppm = round(max(abs(x$mz_ppm_dev)), 1))))
```

`r unc_note`

### Uncertain windows

```{r uncertain-traces}
# The raw traces behind each uncertain window, for the appendix: up to four files
# per window, those with the most confirmed spectra of that target.
unc <- ms1_windows[ms1_windows$uncertain, ]
# The file anchors each uncertain window rests on, for the text.
unc$anchors <- vapply(seq_len(nrow(unc)), function(i) {
  a <- results[[unc$dataset_id[i]]]$rt_anchor[[target_prefix(unc$role[i])]]$anchors
  if (is.null(a)) "not stored" else paste(round(sort(a)), collapse = ", ")
}, character(1))
unc_rows <- lapply(seq_len(nrow(unc)), function(i) {
  ds   <- unc$dataset_id[i]
  chr  <- results[[ds]][[paste0("chr_", target_prefix(unc$role[i]))]]
  bare <- function(x) norm_file_key(to_bare_name(basename(as.character(x)),
                                                 basename(bio_files_per_ds[[ds]])))
  orig <- basename(dataOrigin(chr))
  keyf <- bare(orig)
  hh   <- conf_final[conf_final$dataset_id == ds & conf_final$role == unc$role[i], ]
  n_by <- sort(table(bare(hh$source_file)), decreasing = TRUE)
  pick_f <- head(names(n_by)[names(n_by) %in% keyf], 4)
  lapply(pick_f, function(f) {
    j   <- match(f, keyf)
    pd  <- as.data.frame(Chromatograms::peaksData(chr)[[j]])
    pd  <- pd[is.finite(pd$intensity) &
              pd$rtime >= unc$rt_lo[i] - 30 & pd$rtime <= unc$rt_hi[i] + 30, ]
    lab <- paste0(short_ds(ds), ", ", unc$target[i], "\n", orig[j])
    list(trace = data.frame(panel = lab, rt = pd$rtime, intensity = pd$intensity),
         ms2   = data.frame(panel = lab, rt = as.numeric(hh$rt[bare(hh$source_file) == f])),
         win   = data.frame(panel = lab, lo = unc$rt_lo[i], hi = unc$rt_hi[i]))
  })
})
unc_rows <- unlist(unc_rows, recursive = FALSE)
unc_h    <- 1 + 2 * ceiling(length(unc_rows) / 2)   # inches: two panels per row
```

```{r fig-step5-uncertain, fig.width = 7, fig.height = unc_h}
if (length(unc_rows)) {
  part <- function(k) do.call(rbind, lapply(unc_rows, `[[`, k))
  p_unc <- ggplot(part("trace"), aes(rt, intensity)) +
    geom_rect(data = part("win"), aes(xmin = lo, xmax = hi, ymin = -Inf, ymax = Inf),
              inherit.aes = FALSE, fill = "grey85") +
    geom_line(linewidth = 0.4) +
    geom_vline(data = part("ms2"), aes(xintercept = rt), colour = "firebrick",
               linetype = 2, linewidth = 0.3) +
    facet_wrap(~panel, ncol = 2, scales = "free") +
    labs(x = "retention time (s)", y = "MS1 intensity") +
    theme_bw(base_size = 8)
  ggsave(figp("step5_uncertain_windows.png"), p_unc, width = 7, height = unc_h, dpi = 200)
  ggsave(figp("step5_uncertain_windows.pdf"), p_unc, width = 7, height = unc_h)
  p_unc
}
```

Windows flagged uncertain:
`r if (nrow(unc)) paste(sprintf("%s (%s; file anchors at %s s)", short_ds(unc$dataset_id), unc$target, unc$anchors), collapse = "; ") else "none"`.
In each, the files with a confirmed spectrum do not agree on one MS1 peak: the
window spans the peaks they point to, and the figure shows their traces. Step 5
does not align retention times between files, so it cannot tell these cases
apart from a real second compound.

For each file detected in an uncertain window, the apex and the range step 5
integrates, from the picker run again on the stored trace:

```{r uncertain-bounds}
#| echo: false
# pick() on the stored trace and window, as process_eics() runs it; the bounds are
# kept only where its apex is the stored one, so they belong to the peak measured.
unc_bounds <- do.call(rbind, lapply(seq_len(nrow(unc)), function(i) {
  ds    <- unc$dataset_id[i]
  p     <- target_prefix(unc$role[i])
  r     <- results[[ds]]
  m     <- r[[paste0("metrics_", p)]]
  chr   <- r[[paste0("chr_", p)]]
  files <- basename(dataOrigin(chr))
  win   <- r$rt_anchor[[p]]$win
  det   <- m$file[m[[paste0(p, "_detected")]] %in% TRUE]
  do.call(rbind, lapply(which(files %in% det), function(j) {
    pd     <- Chromatograms::peaksData(chr)[[j]]
    pk     <- pick(pd[, "rtime"], pd[, "intensity"], win[1], win[2], r$is_hilic)
    apex   <- m[[paste0(p, "_apex_rt")]][match(files[j], m$file)]
    same   <- !is.null(pk) && isTRUE(abs(pk$apex_rt - apex) < 0.01)
    data.frame(assay = short_ds(ds), target = unc$target[i], file = files[j],
               apex = round(apex, 1),
               from = if (same) round(pk$lb, 1) else NA,
               to   = if (same) round(pk$rb, 1) else NA)
  }))
}))
if (is.null(unc_bounds) || !nrow(unc_bounds)) cat("No file is detected in an uncertain window.\n") else
  knitr::kable(unc_bounds, row.names = FALSE,
               col.names = c("assay", "target", "file", "apex (s)",
                             "integrated from (s)", "to (s)"))
```

## Checking the tolerance

Each assay is extracted at the tolerance its instrument sets, not at one tuned on
the assay. The check below asks what a wider one would add, on the assays listed
in `curation/tolerance_check.csv`. Each is extracted again at the
time-of-flight tolerance, `r PPM_BY_CLASS[["tof"]]` ppm, with the same ion, m/z,
range, picker and hand corrections (`run_ms1_for_dataset()`), and its raw scans
are read for what a trace does not keep: the m/z of each centroid. The result is
cached in `artifacts/tolerance_check_<dataset_id>.rds` and read again while the
assay's extraction above is unchanged.

```{r tolerance-check}
# The assays to check, listed in curation/tolerance_check.csv; without it the
# section is skipped.
TOL_CHECK     <- if (file.exists("curation/tolerance_check.csv"))
  read.csv("curation/tolerance_check.csv", stringsAsFactors = FALSE)$assay else character()
TOL_CHECK_PPM <- PPM_BY_CLASS[["tof"]]  # the wider tolerance
CAL_PAD_S     <- 15    # s each side of the first target's window (reference ions)
CAL_MIN_INT   <- 1e5   # a reference ion's centroids are at least this intense
CAL_MIN_FRAC  <- 0.9   # ... and it is found in at least this fraction of the files

not_run <- setdiff(TOL_CHECK, names(results))
if (length(not_run))
  message("not extracted above, so not checked: ", paste(not_run, collapse = ", "))
TOL_CHECK <- intersect(TOL_CHECK, names(results))
if (!length(TOL_CHECK))
  cat("No assay is checked: curation/tolerance_check.csv names no assay extracted above.\n")
# The check compares the two targets of a pair, file by file; with any other
# number of targets it is not run.
if (length(TOL_CHECK) && n_targets != 2) {
  cat("No assay is checked: the check is written for two targets, and compound.csv has ",
      n_targets, ".\n", sep = "")
  TOL_CHECK <- character(0)
}

# Cached like the extraction above: read again only when the assay's ion, m/z,
# range, windows, picker, gate, confirmed anchors or hand corrections change.
tol_check <- list()
for (ds in TOL_CHECK) {
  r   <- results[[ds]]
  cfg <- r$cfg
  cfg$ppm <- TOL_CHECK_PPM
  win <- lapply(r$rt_anchor[prefixes], `[[`, "win")   # list(par = , met = )
  key <- list(win = win, picker = PICKER_VERSION, hilic = r$is_hilic,
              anchors = anchor_key(conf_final, ds), curation = curation_key(ds),
              settings = c(CAL_PAD_S, CAL_MIN_INT, CAL_MIN_FRAC))
  path   <- file.path("artifacts", paste0("tolerance_check_", ds, ".rds"))
  cached <- if (file.exists(path)) readRDS(path) else NULL
  if (!is.null(cached) && cfg_is_current(cached, cfg) && identical(cached$key, key)) {
    message("========== ", ds, " at ", TOL_CHECK_PPM, " ppm (cached) ==========")
    tol_check[[ds]] <- cached
    next
  }
  message("\n========== ", ds, " at ", TOL_CHECK_PPM, " ppm ==========")
  # The extraction of run_ms1_for_dataset(), at the wider tolerance.
  row   <- dataset_meta_row(ds_meta_tbl, ds)
  files <- bio_files_per_ds[[ds]]
  cur   <- curation_inputs()
  ms1   <- load_spectra_ms1(row$deposit_id, files, assay = row$assay)
  eics  <- extract_eics(ms1, cfg, cfg$ppm)
  proc  <- process_eics(per_target(eics, "chr"), eics$chr_all, ds, cfg, conf_final,
                        cur$ov, cur$drop_all, row_is_hilic(row), basename(files))
  # The raw scans: each target's centroids in its window above, and the reference ions
  # over the first target's window and CAL_PAD_S either side.
  cent <- dplyr::bind_rows(lapply(setNames(prefixes, roles), function(p)
    scan_centroids(ms1, cfg[[paste0("mz_", p)]], win[[p]], TOL_CHECK_PPM)), .id = "role")
  ions <- common_ions(ms1, win[[1]] + c(-1, 1) * CAL_PAD_S, CAL_MIN_INT, CAL_MIN_FRAC)
  tol_check[[ds]] <- c(list(dataset_id = ds, cfg = cfg, key = key),
                       proc[paste0("metrics_", prefixes)],
                       list(rt_anchor = proc$rt_anchor, centroids = cent, ions = ions))
  saveRDS(tol_check[[ds]], path)
}
```

The files each target is detected in at the assay's own tolerance (the extraction
above) and at `r PPM_BY_CLASS[["tof"]]` ppm, after the isobar rule as in the
sample tables below:

```{r tolerance-check-detections}
# The keys of the files each target is detected in, after the isobar rule.
detected_keys <- function(r, ds) {
  m  <- resolve_isobars(per_target(r, "metrics"), conf_final, ds, r$rt_anchor)
  fb <- basename(bio_files_per_ds[[ds]])
  # One set of keys per target, named by role: list(parent = , metabolite = ).
  setNames(lapply(prefixes, function(p)
    norm_file_key(m[[p]]$file[m[[p]][[paste0(p, "_detected")]] %in% TRUE], fb)), roles)
}
tc_det    <- list()
tc_counts <- list()
for (ds in TOL_CHECK) {
  own  <- detected_keys(results[[ds]], ds)
  wide <- detected_keys(tol_check[[ds]], ds)
  f <- data.frame(file = basename(bio_files_per_ds[[ds]]))
  f$key <- norm_file_key(f$file)
  for (role in roles) {
    f[[paste0(role, "_own")]]  <- f$key %in% own[[role]]
    f[[paste0(role, "_wide")]] <- f$key %in% wide[[role]]
  }
  f$both_own  <- f[[paste0(roles[1], "_own")]] & f[[paste0(roles[2], "_own")]]
  f$both_wide <- f[[paste0(roles[1], "_wide")]] & f[[paste0(roles[2], "_wide")]]
  tc_det[[ds]] <- f
  windows <- function(r) lapply(r$rt_anchor[prefixes], `[[`, "win")
  if (!isTRUE(all.equal(windows(results[[ds]]), windows(tol_check[[ds]]))))
    message(ds, ": the windows at ", TOL_CHECK_PPM, " ppm differ from those above")
  own_c  <- f[paste0(c(roles, "both"), "_own")]
  wide_c <- f[paste0(c(roles, "both"), "_wide")]
  tc_counts[[ds]] <- data.frame(
    assay = short_ds(ds), target = c(target_name[roles], "both"),
    files = nrow(f), `tolerance (ppm)` = tol$ppm[tol$dataset_id == ds],
    `at its tolerance` = colSums(own_c), `at the wider tolerance` = colSums(wide_c),
    `at the wider only` = colSums(wide_c & !own_c),
    `at its tolerance only` = colSums(own_c & !wide_c),
    check.names = FALSE, row.names = NULL)
}
if (length(TOL_CHECK)) do.call(rbind, unname(tc_counts))
```

A file the wider tolerance adds can hold the target, its m/z at the edge of the
narrower tolerance, or another ion of nearly the same mass, and the raw centroids
tell the two apart. In each file, every scan of a target's window gives its most intense
centroid within `r PPM_BY_CLASS[["tof"]]` ppm of the target, and the scan where
that centroid is most intense is the apex: its m/z deviation from the target and
its retention time. A file is **shifted** when the wider tolerance adds it to the
files with both targets and both its apexes lie further from the target than the
assay's own tolerance. Two checks follow. An MS2 spectrum of either target in a
shifted file (a match of step 3, `artifacts/hits_curated.csv`, or a confirmation
of step 4) would point to the target itself. A drift of a file's mass calibration
would move every ion, so the ions found in at least `r 100 * CAL_MIN_FRAC`% of the
files over the `r roles[1]`'s window and `r CAL_PAD_S` s either side (centroids of at
least `r format(CAL_MIN_INT, big.mark = ",", scientific = FALSE)` counts, in
0.01 m/z bins) serve as references. Each file's offset is the median, over these
ions, of its deviation from the ion's median m/z in the files that are not shifted.

```{r tolerance-check-files}
hits <- if (length(TOL_CHECK)) read.csv("artifacts/hits_curated.csv", check.names = FALSE)
tc_files <- list()
tc_ions  <- list()
for (ds in TOL_CHECK) {
  fb  <- basename(bio_files_per_ds[[ds]])
  ppm <- tol$ppm[tol$dataset_id == ds]
  f   <- tc_det[[ds]]
  # The apex of each file and target: the scan with the most intense centroid.
  apex <- tol_check[[ds]]$centroids |>
    dplyr::group_by(role, file) |>
    dplyr::slice_max(int, n = 1, with_ties = FALSE) |>
    dplyr::ungroup()
  apex$key <- norm_file_key(apex$file, fb)
  for (role in roles) {
    a <- apex[apex$role == role, ]
    f[[paste0(role, "_dev")]] <- a$dev[match(f$key, a$key)]
    f[[paste0(role, "_rt")]]  <- a$rt[match(f$key, a$key)]
  }
  # Shifted: both targets added by the wider tolerance, both apexes beyond the assay's own.
  f$shifted <- f$both_wide & !f$both_own &
    abs(f[[paste0(roles[1], "_dev")]]) > ppm & abs(f[[paste0(roles[2], "_dev")]]) > ppm
  f$shifted <- f$shifted %in% TRUE
  # An MS2 spectrum of either target: a match of step 3 or a confirmation of step 4.
  f$ms2 <- f$key %in% c(norm_file_key(hits$file_path[hits$dataset_id == ds]),
                        norm_file_key(conf_final$source_file[conf_final$dataset_id == ds], fb))
  # Each file's calibration offset: per reference ion, its deviation (ppm) from the
  # ion's median m/z in the files that are not shifted; per file, the median over its ions.
  ions <- tol_check[[ds]]$ions
  ions$key     <- norm_file_key(ions$file, fb)
  ions$shifted <- ions$key %in% f$key[f$shifted]
  ions <- ions |>
    dplyr::group_by(ion) |>
    dplyr::mutate(ref = median(mz[!shifted]), dev = 1e6 * (mz - ref) / ref) |>
    dplyr::ungroup()
  offset <- tapply(ions$dev, ions$key, median)
  f$cal_offset <- unname(offset[f$key])
  tc_ions[[ds]]  <- ions
  tc_files[[ds]] <- f
}

# Per assay and target: the shifted files against the other files the target is
# detected in at the assay's own tolerance, by median apex deviation and retention time.
tc_tab <- do.call(rbind, lapply(TOL_CHECK, function(ds) {
  f <- tc_files[[ds]]
  do.call(rbind, lapply(roles, function(role) {
    dev <- f[[paste0(role, "_dev")]]
    rt  <- f[[paste0(role, "_rt")]]
    own <- f[[paste0(role, "_own")]] & !f$shifted
    data.frame(dataset_id = ds, target = target_name[[role]], shifted = sum(f$shifted),
               dev_shifted = median(dev[f$shifted]), dev_own = median(dev[own]),
               rt_shifted = median(rt[f$shifted]), rt_own = median(rt[own]))
  }))
}))
# Per assay, for the text: rounded, one row each.
drops <- curation_inputs()$drop_all
tc_sum <- do.call(rbind, lapply(TOL_CHECK, function(ds) {
  f   <- tc_files[[ds]]
  t   <- tc_tab[tc_tab$dataset_id == ds, ]
  dt  <- t$rt_shifted - t$rt_own
  # The files where a hand drop (curation/ms1_drop_list.csv) removes the detection of
  # either target, its cached file names resolved against the extracted ones.
  dropped <- norm_file_key(drops$file[drops$dataset == ds], f$file)
  # The calibration offsets: the range of the shifted files', the 5th to 95th
  # percentile of the other files'.
  sh  <- f$cal_offset[f$shifted & !is.na(f$cal_offset)]
  cal_sh <- if (length(sh)) round(range(sh), 2) else c(NA, NA)
  cal_ot <- round(quantile(f$cal_offset[!f$shifted], c(0.05, 0.95), na.rm = TRUE, names = FALSE), 2)
  data.frame(assay = short_ds(ds), ppm = tol$ppm[tol$dataset_id == ds],
             n_shifted = sum(f$shifted),
             dev_par = round(t$dev_shifted[1], 1), dev_met = round(t$dev_shifted[2], 1),
             rt_par = sprintf("%.1f s %s", abs(dt[1]), ifelse(dt[1] < 0, "earlier", "later")),
             rt_met = sprintf("%.1f s %s", abs(dt[2]), ifelse(dt[2] < 0, "earlier", "later")),
             n_ions = length(unique(tc_ions[[ds]]$ion)),
             cal_sh_lo = cal_sh[1], cal_sh_hi = cal_sh[2],
             cal_ot_lo = cal_ot[1], cal_ot_hi = cal_ot[2],
             n_ms2 = sum(f$shifted & f$ms2),
             n_drop = sum(f$key %in% dropped))
}))
# The shifted files: apex deviation (ppm) and retention time (s) of each target, and
# the file's calibration offset (ppm).
if (length(TOL_CHECK)) do.call(rbind, lapply(TOL_CHECK, function(ds) {
  f    <- tc_files[[ds]]
  cols <- c(paste0(roles, "_dev"), paste0(roles, "_rt"))
  f    <- f[f$shifted, c("file", cols, "cal_offset", "ms2")]
  if (nrow(f)) cbind(assay = short_ds(ds), f, row.names = NULL) |>
    dplyr::mutate(dplyr::across(dplyr::all_of(cols), ~ round(.x, 1)),
                  cal_offset = round(cal_offset, 2))
}))
```

```{r tolerance-check-table}
#| echo: false
if (length(TOL_CHECK))
  knitr::kable(data.frame(assay = short_ds(tc_tab$dataset_id), target = tc_tab$target,
                          `shifted files` = tc_tab$shifted,
                          `median deviation, shifted (ppm)` = round(tc_tab$dev_shifted, 1),
                          `median deviation, detected (ppm)` = round(tc_tab$dev_own, 1),
                          `median apex RT, shifted (s)` = round(tc_tab$rt_shifted, 1),
                          `median apex RT, detected (s)` = round(tc_tab$rt_own, 1),
                          check.names = FALSE),
               caption = paste("The shifted files of each checked assay against the other files",
                               "where each target is detected at the assay's own tolerance: median",
                               "apex m/z deviation from the target and median apex retention time."))
```

<!-- Shown only where an assay is checked. knitr runs the inline code even when the
div is hidden, so it reads the two targets' names with [ ], which gives NA rather
than an error where compound.csv has fewer. -->
::: {.content-`r if (length(TOL_CHECK)) "visible" else "hidden"`}
In `r tc_sum$n_shifted` files of `r tc_sum$assay`, both targets are detected at
`r TOL_CHECK_PPM` ppm but not both at `r tc_sum$ppm` ppm, and both apexes lie further
than `r tc_sum$ppm` ppm from the target's m/z: a median of `r tc_sum$dev_par` ppm
for `r target_name[roles][1]` and `r tc_sum$dev_met` ppm for
`r target_name[roles][2]`. Their apexes come `r tc_sum$rt_par`
(`r target_name[roles][1]`) and `r tc_sum$rt_met`
(`r target_name[roles][2]`) than in the other files where each target is
detected at `r tc_sum$ppm` ppm. Around the `r roles[1]`'s window, `r tc_sum$n_ions` ions
are found in at least `r 100 * CAL_MIN_FRAC`% of the files. Measured on them, the
offsets of the shifted files lie between `r tc_sum$cal_sh_lo` and
`r tc_sum$cal_sh_hi` ppm, and those of the other files between
`r tc_sum$cal_ot_lo` and `r tc_sum$cal_ot_hi` ppm (5th to 95th percentile). The
shifted files are not off in mass as a whole: only the ions of the two targets
are. `r ifelse(tc_sum$n_ms2 == 0, "None", tc_sum$n_ms2)` of the shifted files
carry an MS2 spectrum matched to either target. At `r TOL_CHECK_PPM` ppm, step 5 would
count each of them as a detection of both. In `r tc_sum$n_drop` more files the
same ion lies close enough to the target to be detected at `r tc_sum$ppm` ppm;
their detections are dropped by hand (`curation/ms1_drop_list.csv`).
:::

## Extraction runtime

```{r}
#| echo: false
# Timing log written by R/peak_review_panels.R; git-ignored, so tolerate its
# absence.
et_path <- "peaks_review_v2/_extract_times.csv"
et <- if (file.exists(et_path)) {
  x <- read.csv(et_path, stringsAsFactors = FALSE)
  # The current assays only (keyed by assay or by deposit), each one's latest run.
  x <- x[x$dataset %in% c(names(results), sub("_.*", "", names(results))), ]
  x <- x[!duplicated(x$dataset, fromLast = TRUE), ]
  x <- x[order(x$extract_seconds), ]
  x$minutes    <- round(x$extract_seconds / 60, 1)
  x$s_per_file <- round(x$extract_seconds / x$n_files, 2)
  x
} else NULL
et_hours <- if (is.null(et)) NA else round(sum(et$extract_seconds) / 3600, 1)
et_files <- if (is.null(et)) NA else format(sum(et$n_files), big.mark = ",")
# The version the log recorded, where it did; else the installed one.
chr_ver  <- if (!is.null(et) && any(nzchar(et$chromatograms))) {
  paste(unique(et$chromatograms[nzchar(et$chromatograms)]), collapse = ", ")
} else as.character(utils::packageVersion("Chromatograms"))
```

Loading each file's spectra and pulling its
`r xfun::numbers_to_words(n_targets)` EIC`r if (n_targets > 1) "s"` is the
pipeline's dominant compute cost, and it runs single-threaded. That is measured rather than assumed:
parallel backends gave no gain once the files are read, because the cost is disk
I/O rather than CPU. Cost tracks file count and file size, not the number of
detections, and the timings are compute-only on cached files (raw-file FTP
download is excluded). The corpus takes about **`r et_hours` hours** over
`r if (is.null(et)) NA else nrow(et)` datasets (`r et_files` files).

These numbers were measured with **Chromatograms `r chr_ver`**. Extraction
performance is strongly version-dependent, so treat the table as valid only for
that version and re-measure after any upgrade of the extraction stack.

```{r}
#| echo: false
#| output: asis
if (is.null(et)) {
  cat("_Timing table skipped: `", et_path, "` is not present (git-ignored). Run ",
      "`Rscript R/peak_review_panels.R <application>` to regenerate it._\n", sep = "")
} else {
  cat(knitr::kable(et[, c("dataset", "n_files", "minutes", "s_per_file")],
                   col.names = c("dataset", "files", "min", "s / file"),
                   align = "lrrr", row.names = FALSE), sep = "\n")
}
```

<!-- Application-specific checks. An application with a positive control of its
own, such as a dosed or genotyped cohort, keeps it in _05-probe-checks.qmd beside
this notebook. Without that file the section is omitted and everything else here
runs unchanged. -->

```{r probe-checks-targets}
#| echo: false
# The probe checks are written for the targets parent and metabolite.
probe_checks <- file.exists("_05-probe-checks.qmd") &&
  identical(compound_spec$role, c("parent", "metabolite"))
if (file.exists("_05-probe-checks.qmd") && !probe_checks)
  message("_05-probe-checks.qmd is skipped: its checks need the targets parent ",
          "and metabolite, in that order, in compound.csv")
```

```{r probe-checks, child = if (probe_checks) "_05-probe-checks.qmd"}
```

# Cross-dataset sample table

```{r combine-tables}
#| echo: false
all_samples <- do.call(rbind, lapply(names(results), function(ds) {
  # The sample table is rebuilt, not read from the cache, so it is never stale. A
  # repository that does not answer stops here, before rbind() fails on the stub.
  meta <- tryCatch(load_metadata(ds), error = function(e) {
    if (is_network_error(e)) stop(e)
    data.frame(file = basename(bio_files_per_ds[[ds]]))
  })
  .m <- resolve_isobars(per_target(results[[ds]], "metrics"),
                        conf_final, ds, results[[ds]]$rt_anchor)
  st <- build_sample_table(.m, basename(bio_files_per_ds[[ds]]), meta, dedup = TRUE)
  # Keep only files still in the current bio_files: a cached extraction can hold files
  # a later QC rule removed.
  keep <- norm_file_key(st$file) %in% norm_file_key(bio_files_per_ds[[ds]])
  cbind(dataset_id = ds, assay = results[[ds]]$assay, st[keep, , drop = FALSE],
        stringsAsFactors = FALSE)
}))

# Files with no metadata are kept for the presence record but excluded from
# every stratified analysis below.
all_samples$has_metadata <- {
  key <- paste0(all_samples$dataset_id, "|",
                basename(as.character(all_samples$file)))
  hit <- has_metadata_lookup[key]
  ifelse(is.na(hit), FALSE, unname(hit))
}

write.csv(all_samples, "artifacts/sample_table_all.csv", row.names = FALSE)

cat("Datasets processed :", length(results), "\n")
cat("Total rows         :", nrow(all_samples),
    "  (with metadata:", sum(all_samples$has_metadata),
    "| record-only:", sum(!all_samples$has_metadata), ")\n")
```

`sample_table_all.csv` is one row per subject (the `build_sample_table()`
dedup). Per-file and longitudinal analyses (`06`) need the rows *before* that
dedup, so we also write a per-file table, keeping `subject_id` (+ a
`subject_key` fallback) so those analyses aggregate to one-per-subject
themselves; the rows are not independent.

```{r combine-tables-perfile}
#| echo: false
getcol <- function(df, nm)
  if (nm %in% names(df)) as.character(df[[nm]]) else rep(NA_character_, nrow(df))

all_samples_perfile <- dplyr::bind_rows(lapply(names(results), function(ds) {
  r    <- results[[ds]]
  meta <- tryCatch(load_metadata(ds), error = function(e) {
    if (is_network_error(e)) stop(e)   # as in combine-tables
    data.frame(file = basename(bio_files_per_ds[[ds]]))
  })
  .m   <- resolve_isobars(per_target(r, "metrics"), conf_final, ds, r$rt_anchor)
  st   <- build_sample_table(.m, basename(bio_files_per_ds[[ds]]), meta, dedup = FALSE)
  st$subject_key <- dplyr::coalesce(getcol(st, "subject_id"),
                                    getcol(st, "sample_name"), as.character(st$file))
  cbind(dataset_id = ds, assay = r$assay, st, stringsAsFactors = FALSE)
}))
write.csv(all_samples_perfile, "artifacts/sample_table_perfile.csv", row.names = FALSE)
cat("One row per raw file:", nrow(all_samples_perfile),
    "| one row per subject-sample after collapsing repeat files:", nrow(all_samples), "\n")
```

## Positive control: files where MS2 already confirmed the compound

Where SIRIUS confirmed a compound's MS2 spectrum in a file, the compound is in
that file, so MS1 extraction must find it there too. A confirmed file it misses
is propagation losing real signal rather than biology, and the table names the
rule behind each miss. A `truncated` detection counts as found: its peak ran
into empty scans before falling to half height, so it is detected but carries
no area.

```{r positive-control}
#| echo: false
#| output: asis
pc <- dplyr::bind_rows(lapply(names(results), function(ds) {
  fb   <- basename(as.character(bio_files_per_ds[[ds]]))
  bare <- function(x) norm_file_key(to_bare_name(basename(as.character(x)), fb))
  st   <- all_samples_perfile[all_samples_perfile$dataset_id == ds, ]
  st$k <- norm_file_key(st$file)
  dplyr::bind_rows(lapply(roles, function(role) {
    pre <- target_prefix(role)
    ck  <- unique(bare(conf_final$source_file[conf_final$dataset_id == ds &
                                              conf_final$role == role]))
    s   <- st[st$k %in% ck, , drop = FALSE]
    if (!nrow(s)) return(NULL)
    det <- s[[paste0(pre, "_detected")]] %in% TRUE
    why <- as.character(s[[paste0(pre, "_gate")]]); why[is.na(why)] <- "no peak in window"
    tab <- sort(table(why[!det]), decreasing = TRUE)
    data.frame(dataset = short_ds(ds, names(results)), compound = role, confirmed = nrow(s),
               detected = sum(det),
               by_hand = sum(det & why == "manual accept"),
               truncated = sum(det & s[[paste0(pre, "_truncated")]] %in% TRUE),
               missed = sum(!det),
               why_missed = if (length(tab)) paste0(names(tab), " (", tab, ")", collapse = "; ") else "",
               stringsAsFactors = FALSE)
  }))
}))
write.csv(pc, "artifacts/positive_control.csv", row.names = FALSE)
cat(knitr::kable(pc, row.names = FALSE,
                 col.names = c("dataset", "compound", "MS2-confirmed files", "MS1 detected",
                               "of which accepted by hand", "of which truncated", "missed",
                               "why missed"),
                 caption = sprintf(paste(
                   "Positive control: %d of %d MS2-confirmed files are detected at MS1,",
                   "%d of them accepted by hand."),
                   sum(pc$detected), sum(pc$confirmed), sum(pc$by_hand))), sep = "\n")
```

```{r fig-pc-misses, fig.width = 7, fig.height = 0.6 + 1.9 * max(1, ceiling(sum(pc$missed) / 4))}
# Every miss of the table above, one panel each: trace, window, MS2 times and apex.
# A file's key within its assay, as the table above matches files.
key_in <- function(ds, x)
  norm_file_key(to_bare_name(basename(as.character(x)),
                             basename(as.character(bio_files_per_ds[[ds]]))))
miss <- as.data.frame(dplyr::bind_rows(lapply(names(results), function(ds) {
  st <- all_samples_perfile[all_samples_perfile$dataset_id == ds, ]
  dplyr::bind_rows(lapply(roles, function(role) {
    pre <- target_prefix(role)
    ck  <- key_in(ds, conf_final$source_file[conf_final$dataset_id == ds &
                                             conf_final$role == role])
    s   <- st[norm_file_key(st$file) %in% ck &
              !(st[[paste0(pre, "_detected")]] %in% TRUE), , drop = FALSE]
    if (!nrow(s)) return(NULL)
    why <- as.character(s[[paste0(pre, "_gate")]]); why[is.na(why)] <- "no peak in window"
    data.frame(dataset_id = ds, role = role, pre = pre, file = s$file, reason = why,
               apex_rt = s[[paste0(pre, "_apex_rt")]])
  }))
})))
stopifnot(nrow(miss) == sum(pc$missed))

# A file name without its accession and extension, cut in the middle if long.
short_file <- function(f) {
  f <- sub("[.][^.]*$", "", sub("^(MSV|MTBLS|ST)[0-9]+[_-]", "", basename(f)))
  ifelse(nchar(f) > 28, paste0(substr(f, 1, 13), "...", substring(f, nchar(f) - 11)), f)
}
miss_rows <- lapply(seq_len(nrow(miss)), function(i) {
  m   <- miss[i, ]
  r   <- results[[m$dataset_id]]
  k   <- norm_file_key(m$file)
  win <- r$rt_anchor[[m$pre]]$win
  ms2 <- unique(as.numeric(conf_final$rt[conf_final$dataset_id == m$dataset_id &
                                         conf_final$role == m$role &
                                         key_in(m$dataset_id, conf_final$source_file) == k]))
  # The window and 25 s either side, wider where a confirmed spectrum or the apex lies outside.
  xr  <- range(win + c(-25, 25), ms2 - 5, ms2 + 5, m$apex_rt, na.rm = TRUE)
  chr <- r[[paste0("chr_", m$pre)]]
  j   <- match(k, key_in(m$dataset_id, dataOrigin(chr)))
  pd  <- if (is.na(j)) data.frame(rtime = numeric(), intensity = numeric()) else
    as.data.frame(Chromatograms::peaksData(chr)[[j]])
  pd  <- pd[pd$rtime >= xr[1] & pd$rtime <= xr[2], ]
  id  <- sprintf("%03d", i)
  pd$id  <- rep(id, nrow(pd))
  # A run of scans with signal is one line; an empty scan breaks it.
  pd$run <- paste(id, cumsum(!is.finite(pd$intensity)))
  on  <- pd[is.finite(pd$intensity), ]
  list(label = paste0(short_ds(m$dataset_id, names(results)), ", ",
                      sub("^L-", "", target_name[[m$role]]),
                      "\n", short_file(m$file), "\n", m$reason),
       trace = on,
       empty = pd[!is.finite(pd$intensity), ],
       # The scan the picker took as apex; none where it found no peak.
       apex  = on[which.min(abs(on$rtime - m$apex_rt)), ],
       ms2   = data.frame(id = rep(id, length(ms2)), rtime = ms2),
       win   = data.frame(id = id, lo = win[1], hi = win[2]),
       lim   = data.frame(id = id, rtime = xr, intensity = 0))
})

by_reason <- sort(table(miss$reason), decreasing = TRUE)
miss_note <- if (nrow(miss))
  paste0("The figure draws each of the ", nrow(miss), " misses, one panel per file and target: ",
         paste0(names(by_reason), " (", by_reason, ")", collapse = ", "), ".") else
  "No MS2-confirmed file is missed, so there is nothing to draw."

miss_h <- 0.6 + 1.9 * max(1, ceiling(nrow(miss) / 4))   # inches: four panels per row
if (nrow(miss)) {
  part <- function(k) do.call(rbind, lapply(miss_rows, `[[`, k))
  p_miss <- ggplot(part("trace"), aes(rtime, intensity)) +
    geom_rect(data = part("win"), aes(xmin = lo, xmax = hi, ymin = -Inf, ymax = Inf),
              inherit.aes = FALSE, fill = "grey88") +
    geom_blank(data = part("lim")) +
    geom_vline(data = part("ms2"), aes(xintercept = rtime), colour = "firebrick",
               linetype = 2, linewidth = 0.3) +
    geom_rug(data = part("empty"), aes(x = rtime), inherit.aes = FALSE, sides = "b",
             colour = "grey55", linewidth = 0.2, length = unit(0.04, "npc")) +
    geom_line(aes(group = run), linewidth = 0.35) +
    geom_point(size = 0.3) +
    geom_point(data = part("apex"), shape = 21, size = 2, stroke = 0.5,
               colour = "#2166AC", fill = NA) +
    facet_wrap(~id, ncol = 4, scales = "free",
               labeller = as_labeller(setNames(vapply(miss_rows, `[[`, "", "label"),
                                               sprintf("%03d", seq_along(miss_rows))))) +
    scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale()),
                       expand = expansion(mult = c(0.05, 0.1))) +
    labs(x = "retention time (s)", y = "MS1 intensity",
         caption = paste("Shaded: the window. Dashed: the file's confirmed MS2 spectra.",
                         "Circle: the apex the picker chose. Ticks: scans with no signal.")) +
    theme_bw(base_size = 7) +
    theme(strip.text = element_text(size = 6, hjust = 0),
          panel.grid.minor = element_blank(),
          plot.caption = element_text(size = 6, colour = "grey30", hjust = 0))
  ggsave(figp("step5_pc_misses.png"), p_miss, width = 7, height = miss_h, dpi = 200)
  ggsave(figp("step5_pc_misses.pdf"), p_miss, width = 7, height = miss_h)
  p_miss
} else {
  # The chapter collects this figure from both applications, so it is written even empty.
  p_none <- ggplot() + annotate("text", x = 0, y = 0, label = miss_note, size = 3) + theme_void()
  ggsave(figp("step5_pc_misses.png"), p_none, width = 7, height = 1, dpi = 200)
  ggsave(figp("step5_pc_misses.pdf"), p_none, width = 7, height = 1)
}
```

`r miss_note`

## Files per assay, by MS2 and by MS1

For each assay and target, the files in which MS2 confirmed the target (the
positive control's rule) and the files in which MS1 detects its peak. The
chapter's examples of what extraction adds are read from this table.

```{r files-per-assay}
#| echo: false
per_assay <- dplyr::bind_rows(lapply(names(results), function(ds) {
  st <- all_samples_perfile[all_samples_perfile$dataset_id == ds, ]
  a  <- short_ds(ds, names(results))
  dplyr::bind_rows(lapply(roles, function(role) {
    pre <- target_prefix(role)
    data.frame(assay = a, compound = role, files = nrow(st),
               ms2 = sum(pc$confirmed[pc$dataset == a & pc$compound == role]),
               ms1 = sum(st[[paste0(pre, "_detected")]] %in% TRUE))
  }))
}))
write.csv(per_assay, "artifacts/files_per_assay.csv", row.names = FALSE)
knitr::kable(per_assay, col.names = c("assay", "compound", "files",
                                      "MS2-confirmed files", "MS1-detected files"))
```

# Step summary

The results of this step, as the chapter's table shows them.

```{r step-summary}
# A file has a target by MS2 where one of its confirmed spectra came from it, and by
# MS1 where its peak was detected; a file counts when it has every target.
ms2_every <- sum(vapply(names(results), function(ds) {
  fb <- basename(as.character(bio_files_per_ds[[ds]]))
  k  <- norm_file_key(all_samples_perfile$file[all_samples_perfile$dataset_id == ds])
  confirmed <- function(role) k %in% norm_file_key(
    conf_final$source_file[conf_final$dataset_id == ds & conf_final$role == role], fb)
  sum(Reduce(`&`, lapply(roles, confirmed)))
}, integer(1)))
ms1_every <- sum(Reduce(`&`, lapply(prefixes, function(p)
  all_samples_perfile[[paste0(p, "_detected")]] %in% TRUE)))
# The hand corrections in curation/, counted for the datasets extracted here.
cur <- curation_inputs()
n_manual <- sum(cur$ov$dataset_id %in% names(results)) +
            sum(cur$drop_all$dataset %in% names(results)) +
            sum(cur$accept$dataset %in% names(results))

# The step's table counts files only; the recap rows (assays, deposits, MS1 over MS2)
# are for the discussion.
summary_step <- data.frame(
  item  = c("files extracted",
            paste("... with", targets_phrase(n_targets), "confirmed by MS2"),
            paste("... with", targets_phrase(n_targets), "detected in MS1"),
            "recap: assays", "recap: deposits",
            paste0("recap: ", targets_phrase(n_targets), ", MS1 over MS2")),
  value = c(nrow(all_samples_perfile), ms2_every, ms1_every,
            length(results),
            length(unique(ds_summary$deposit_id[ds_summary$dataset_id %in% names(results)])),
            sprintf("%.1f", ms1_every / ms2_every)))
write.csv(summary_step, "artifacts/summary_step5.csv", row.names = FALSE)
summary_step
```

Step 5 extracts `r length(results)` assays. Of the `r sum(pc$confirmed)` cases where
MS2 confirmed a target in a file, the positive control finds its MS1 peak in
`r sum(pc$detected)`. `r if (n_manual == 0) "No peak was corrected by hand." else paste(n_manual, if (n_manual == 1) "peak was" else "peaks were", "corrected by hand (curation/).")`

# Session info

```{r session-info}
sessionInfo()
```
