---
title: "06 - Analysis"
subtitle: "Step 6: omeprazole against CYP2C19 genotype"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    df-print: paged
---

# Overview

A study in the corpus gave omeprazole to genotyped volunteers (Jarmusch et
al., *Clin Transl Sci* 2020, PMC7485946). Omeprazole is the standard probe for
the liver enzyme CYP2C19, which turns it into 5-OH-omeprazole, and each
volunteer's CYP2C19 metaboliser class is known from their genotype. The study
was deposited twice: MSV000084008 holds the timed plasma draws, MSV000082493 the
skin, urine and faeces samples (and a second copy of most plasma files).

It is the one place in the corpus where the answer is known before looking, and
it is known twice. Before the dose there is no drug, so the extraction should
find nothing. After the dose, the ratio of omeprazole to 5-OH-omeprazole should
fall as CYP2C19 activity rises. This notebook tests both, then scans the rest
of the corpus: every metadata field that varies inside an assay, tested against
the ratio in blood.

**Input:** `artifacts/sample_table_perfile.csv`, `artifacts/bio_files_per_ds.rds`
and `artifacts/bio_meta_per_ds.rds` (step 5), `artifacts/datasets-confirmed.csv`
and `artifacts/hits-confirmed-all.csv` (step 4), `artifacts/related_deposits.csv`
(step 2), and the dosed study's submitter tables,
`dataset_metadata/MSV000084008_metadata.tsv` and
`dataset_metadata/MSV000082493_metadata.tsv`, which give each sample's
volunteer, study day and time since the dose, and for the plasma the
metaboliser class.

**Output:** `artifacts/assay_choice.csv` (which assay is kept where several
measured the same samples), `artifacts/ratio_scan.csv` (the scan of the rest of
the corpus), `artifacts/summary_step6.csv` and `figures/ratio_volunteer.png`.

```{r setup}
suppressPackageStartupMessages({ library(dplyr); library(ggplot2) })
source("../../R/metadata.R")   # norm_file_key(), as_num(), short_ds()
source("../../R/analysis.R")   # the helpers shared with the kyn-trp notebook
source("../../R/masst.R")      # PHENO: the CYP2C19 classes, low to high activity

# Figures go to this application's own folder.
FIG_DIR <- "figures"; dir.create(FIG_DIR, showWarnings = FALSE)
figp    <- function(f) file.path(FIG_DIR, f)

# The dosed study: its plasma and its skin accession.
PLASMA_DS <- "MSV000084008"
SKIN_DS   <- "MSV000082493"
# Draws taken before the drug can reach the blood: at the dose and 5 min later.
EARLY_MIN <- c(0, 5)

# The rest of the corpus: matrices where the ratio is read, per-file metadata fields
# scanned against it (column = label), and the scan's false discovery rate.
MATRICES    <- c("plasma", "serum")
SCAN_FIELDS <- c(sex = "sex", age = "age", group = "group",
                 health_status = "health status", body_site = "matrix",
                 country = "country", treatment = "treatment",
                 timepoint = "time point")
FDR <- 0.05

# Class labels and colours of the figure.
PH_SHORT <- c("Poor", "Inter.", "Normal", "Rapid", "Ultra")
pal <- c("Poor Metabolizer" = "#B2182B", "Intermediate Metabolizer" = "#EF8A62",
         "Normal Metabolizer" = "#D9D9D9", "Rapid Metabolizer" = "#67A9CF",
         "Ultrarapid Metabolizer" = "#2166AC")

# Numbers and lists for the text.
fmt_p    <- function(p) formatC(p, digits = 2, format = "fg", flag = "#")
pct      <- function(x) sprintf("%.1f%%", 100 * x)
and_list <- function(x) if (length(x) < 3) paste(x, collapse = " and ") else
  paste(paste(head(x, -1), collapse = ", "), "and", tail(x, 1))
```

# Inputs

```{r load}
pf        <- read.csv("artifacts/sample_table_perfile.csv", stringsAsFactors = FALSE)
datasets  <- read.csv("artifacts/datasets-confirmed.csv", stringsAsFactors = FALSE)
related   <- read.csv("artifacts/related_deposits.csv", stringsAsFactors = FALSE)
conf      <- read.csv("artifacts/hits-confirmed-all.csv", stringsAsFactors = FALSE)
bio_files <- readRDS("artifacts/bio_files_per_ds.rds")
bio_meta  <- readRDS("artifacts/bio_meta_per_ds.rds")

# Does a metadata source cover the file? Stops if step 5 has no file list for an
# assay of the table.
pf$has_metadata <- has_metadata(pf, bio_meta)
```

# One assay per set of samples

Some samples were measured more than once, by assays on different columns or in
different ionisation modes, and keeping them all would count the same people more
than once.
So one assay is kept per set of samples: the one where MS1 detects both targets
in the most files; on a tie, the one with more files where MS2 confirms both;
still tied, the first assay id in sort order. A study deposited twice, as the
dosed study is, holds one set of files rather than two measurements and is left
as it is.

```{r keep-one}
# Counted on every extracted file, before any restriction to metadata.
choice <- keep_one_assay(pf, datasets, related, conf, bio_files)
write.csv(choice, "artifacts/assay_choice.csv", row.names = FALSE)

# The sets with more than one assay.
groups <- choice[choice$group %in% choice$group[duplicated(choice$group)], ]
groups$assay <- short_ds(groups$dataset_id, datasets$dataset_id)
groups$chromatography <- datasets$chromatography[match(groups$dataset_id, datasets$dataset_id)]

# One sentence per set for the text below, the kept assay first.
said <- vapply(split(groups, factor(groups$group, levels = unique(groups$group))),
               function(g) {
  g <- g[order(!g$kept), ]
  sprintf("%s (%s) is kept and %s set aside: MS1 detects both targets in %s of their files, MS2 confirms both in %s.",
          g$assay[1], g$chromatography[1],
          and_list(paste0(g$assay[-1], " (", g$chromatography[-1], ")")),
          and_list(g$ms1_both), and_list(g$ms2_both))
}, "")
groups[, c("assay", "chromatography", "files", "ms1_both", "ms2_both", "kept")]
n_out <- sum(!choice$kept)
n_in  <- sum(choice$kept)

pf <- pf[pf$dataset_id %in% choice$dataset_id[choice$kept], ]
```

`r if (length(said)) paste(said, collapse = " ") else "No two assays here measured the same samples."`
`r n_out` `r if (n_out == 1) "assay is" else "assays are"` set aside, and
`r n_in` `r if (n_in == 1) "goes" else "go"` on to the analysis.

# The dosed study

## Plasma draws

`r PLASMA_DS` holds the plasma. The volunteers were drawn at set times after the
dose, on study days before and after a course of antibiotics, and a draw could
be injected more than once. The submitter table gives each file's
volunteer, day, minutes since the dose and metaboliser class, and each file is
matched to its extracted copy by `norm_file_key()`. A target that step 5 did
not detect in a file has an area of zero there: it is below detection, not its
integrated noise.

```{r draws}
p84   <- pf[pf$dataset_id == PLASMA_DS, ]
p84$k <- norm_file_key(p84$file)
# A detection must have an area. A truncated peak (detected, no area) would
# otherwise count as zero, as if it were below detection.
if (any(p84$par_detected %in% TRUE & is.na(p84$par_auc) |
        p84$met_detected %in% TRUE & is.na(p84$met_auc)))
  stop("a detection in ", PLASMA_DS, " has no peak area")

ta   <- read.delim("dataset_metadata/MSV000084008_metadata.tsv",
                   check.names = FALSE, stringsAsFactors = FALSE)
ta$k <- norm_file_key(ta$filename)
i    <- match(ta$k, p84$k)
ta$extracted <- !is.na(i)
ta$par_det   <- p84$par_detected[i] %in% TRUE
ta$met_det   <- p84$met_detected[i] %in% TRUE
ta$par_auc   <- ifelse(ta$par_det, p84$par_auc[i], 0)
ta$met_auc   <- ifelse(ta$met_det, p84$met_auc[i], 0)
ta$tmin      <- as_num(ta$ATTRIBUTE_Time_Point_Mins)
# Volunteer draws only: blanks and QC injections have no draw time.
ta <- ta[ta$MS_SampleType == "sample" & !is.na(ta$tmin), ]
# A draw is one volunteer, study day and time; its files are its injections.
ta$draw <- paste(ta$ATTRIBUTE_Subject, ta$ATTRIBUTE_Study_Day, ta$tmin)
# Study days per volunteer, for the text below.
n_days <- tapply(ta$ATTRIBUTE_Study_Day, ta$ATTRIBUTE_Subject, n_distinct)
```

The table lists `r nrow(ta)` files from `r n_distinct(ta$draw)` draws of
`r n_distinct(ta$ATTRIBUTE_Subject)` volunteers:
`r sum(table(ta$draw) > 1)` draws were injected more than once. Of these files,
`r sum(ta$extracted)` were extracted, and they cover
`r n_distinct(ta$draw[ta$extracted])` of the draws. Of the volunteers,
`r sum(n_days > 1)` were drawn on more than one study day and
`r sum(n_days == 1)` on a single day.

## Before the dose

The draws at the dose and `r EARLY_MIN[2]` minutes later come before the drug
can reach the blood. They are real plasma, read at the true m/z and in the true
retention window, so any detection there is a false one. The forehead skin of
`r SKIN_DS`, swabbed before the first dose, is a second such set. Either target
counts as a detection, and only files that were extracted are counted. The
upper end of an exact binomial interval bounds the rate of false detections,
since finding nothing in a few files says less than finding nothing in many.

```{r negative-control}
# Plasma drawn at the dose or just after, among the files extracted. A draw
# injected more than once counts once, as detected if any of its files is.
pre      <- ta[ta$tmin %in% EARLY_MIN & ta$extracted, ]
pre_draw <- tapply(pre$par_det | pre$met_det, pre$draw, any)

# Forehead skin swabbed before the first dose: day 0, or minute 0 of day 1.
sk <- read.delim("dataset_metadata/MSV000082493_metadata.tsv",
                 check.names = FALSE, stringsAsFactors = FALSE)
sk <- sk[sk$MS_SampleType == "sample" & sk$ATTRIBUTE_Body_Site == "L.forehead" &
           (sk$ATTRIBUTE_Study_Day %in% "0" |
            (sk$ATTRIBUTE_Study_Day %in% "1" & sk$ATTRIBUTE_Time_Point_Mins %in% "0")), ]
p82 <- pf[pf$dataset_id == SKIN_DS, ]
j   <- match(norm_file_key(sk$filename), norm_file_key(p82$file))
j   <- j[!is.na(j)]
skin_det <- p82$par_detected[j] %in% TRUE | p82$met_detected[j] %in% TRUE

neg <- data.frame(
  set      = c(sprintf("plasma draws taken at %s min", and_list(EARLY_MIN)),
               "skin samples swabbed before the first dose"),
  n        = c(length(pre_draw), length(j)),
  detected = c(sum(pre_draw), sum(skin_det)))
# Upper end of the exact 95% interval on the rate of false detections.
neg$upper95 <- mapply(function(x, n) binom.test(x, n)$conf.int[2],
                      neg$detected, neg$n)
neg
```

The extraction detects a target in `r neg$detected[1]` of the `r neg$n[1]` early
plasma draws (`r nrow(pre)` files, since some draws were injected more than once)
and in `r neg$detected[2]` of the `r neg$n[2]` skin samples. The upper ends of the
exact 95% intervals on the rate of false detections are `r pct(neg$upper95[1])` in
plasma and `r pct(neg$upper95[2])` on skin.

## The ratio against genotype

The usual read-out of CYP2C19 activity is this ratio in one draw, taken at a
fixed time after the dose, because the ratio changes across the dosing interval.
Here each volunteer's ratio is taken instead on the areas under the whole
concentration-time curve, which uses every draw. A draw injected more than once
gives the mean of its extracted files, so each draw weighs once and the order
of the files does not matter. For each volunteer and study day, the area of
each target is integrated over the draw times by the trapezoid rule, and the
ratio of the two areas is taken. The study days are integrated separately, since
they straddle the antibiotics, and then averaged per volunteer. A file where a
target was not detected counts as zero, the usual below-detection rule; dropping
it instead would raise the area for exactly the poor metabolisers whose
metabolite is truly absent.

```{r auc}
# One area per draw and target: the mean over the draw's extracted files.
draw_areas <- function(t)
  aggregate(cbind(par_auc, met_auc) ~ ATTRIBUTE_Subject + ATTRIBUTE_Study_Day +
              tmin + PharmGeno_Phenotype_2C19,
            data = t[t$extracted, ], FUN = mean)
dr <- draw_areas(ta)

# Trapezoidal area over the draw times.
trap <- function(x, y) {
  o <- order(x); x <- x[o]; y <- y[o]
  if (length(x) < 4) return(NA_real_)   # too few times for a curve
  sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)
}

# One ratio of areas per volunteer and study day.
day_ratios <- function(dr) {
  a <- do.call(rbind, lapply(
    split(dr, list(dr$ATTRIBUTE_Subject, dr$ATTRIBUTE_Study_Day), drop = TRUE),
    function(d) data.frame(
      subject = d$ATTRIBUTE_Subject[1], day = d$ATTRIBUTE_Study_Day[1],
      auc_par = trap(d$tmin, d$par_auc), auc_met = trap(d$tmin, d$met_auc),
      phenotype = d$PharmGeno_Phenotype_2C19[1], stringsAsFactors = FALSE)))
  a$log2_ratio <- log2(a$auc_par / a$auc_met)
  a <- a[is.finite(a$log2_ratio), ]
  a$rank <- match(a$phenotype, PHENO)
  a
}
auc <- day_ratios(dr)

# One value per volunteer: the mean of the two days.
subj_auc <- aggregate(log2_ratio ~ subject + phenotype + rank, data = auc, FUN = mean)
```

The direction is set before looking. CYP2C19 makes the metabolite, so a poor
metaboliser keeps more omeprazole and less 5-OH-omeprazole, and the ratio should
fall along the classes from poor to ultrarapid. The classes are ranked in that
order, one step per class:
`r and_list(sprintf("%s %d", tolower(sub(" Metabolizer", "", PHENO)), seq_along(PHENO)))`.

The model, in words: a volunteer's log2 ratio of the two areas is a baseline,
plus a change $b$ for each step up in activity class, plus what the model leaves
unexplained. As a formula, one row per volunteer $v$, fitted by least squares:

$$
\log_2\left(\frac{\text{AUC}_{\text{omeprazole}}}{\text{AUC}_{\text{5-OH-omeprazole}}}\right)_v
= a + b \cdot \text{class}_v + \varepsilon_v
$$

The slope $b$ is the change per class, and the prediction is $b < 0$. Reading
the rank as a number assumes the classes are evenly spaced, which nothing
guarantees; it is the price of an effect in units one can read.

```{r genotype-model}
m_auc <- lm(log2_ratio ~ rank, data = subj_auc)
ci    <- confint(m_auc)["rank", ]   # 95% confidence interval
p_auc <- summary(m_auc)$coefficients["rank", 4]
summary(m_auc)
# Classes with a single volunteer, for the text below.
n_single <- sum(table(subj_auc$phenotype) == 1)
```

Over `r nrow(subj_auc)` volunteers, the log2 ratio changes by
`r sprintf("%+.3f", coef(m_auc)[["rank"]])` per class (confidence interval
`r sprintf("%.3f", ci[1])` to `r sprintf("%.3f", ci[2])`, p = `r fmt_p(p_auc)`,
R² = `r sprintf("%.2f", summary(m_auc)$r.squared)`):
`r if (ci[2] < 0) "the predicted fall, with an interval that excludes zero" else if (ci[1] > 0) "a rise, against the prediction" else "the interval includes zero, so no fall is shown"`.

Of the classes, `r n_single` `r if (n_single == 1) "holds" else "hold"` a
single volunteer, so one person could carry the slope. The model is refitted
once without each volunteer.

```{r leave-one-out}
# The slope and its p, fitted once without each volunteer of `d`.
leave_one_out <- function(d) do.call(rbind, lapply(seq_len(nrow(d)), function(i) {
  co <- summary(lm(log2_ratio ~ rank, data = d[-i, ]))$coefficients
  data.frame(dropped = d$subject[i], phenotype = d$phenotype[i],
             beta = co[2, 1], p = co[2, 4])
}))
jk <- leave_one_out(subj_auc)
jk[order(jk$beta), ]
```

Without any one volunteer the slope stays between
`r sprintf("%+.3f", min(jk$beta))` and `r sprintf("%+.3f", max(jk$beta))`,
`r if (all(jk$beta < 0)) "negative every time" else "and changes sign"`, and the
largest p is `r fmt_p(max(jk$p))`. One point per volunteer:

```{r fig-ratio-class, fig.width=6.4, fig.height=4.4}
#| echo: false
vol <- subj_auc
vol$phenotype <- factor(vol$phenotype, levels = PHENO)
fit_line <- data.frame(rank = seq(1, 5, length.out = 50))
fit_line$log2_ratio <- predict(m_auc, fit_line)
nv <- table(vol$phenotype)

p_vol <- ggplot(vol, aes(rank, log2_ratio)) +
  geom_line(data = fit_line, colour = "grey35", linewidth = 0.8) +
  geom_point(aes(fill = phenotype), shape = 21, size = 3.4, stroke = 0.6,
             colour = "grey20") +
  scale_fill_manual(values = pal, guide = "none") +
  scale_x_continuous(breaks = 1:5,
                     labels = sprintf("%s\n(%d)", PH_SHORT, as.integer(nv))) +
  # No title: the fit is given in the text and the chapter's caption.
  labs(x = "CYP2C19 metaboliser class (volunteers)",
       y = "log2(AUC omeprazole / AUC 5-hydroxyomeprazole)") +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank())
ggsave(figp("ratio_volunteer.png"), p_vol, width = 6.4, height = 4.4, dpi = 200)
p_vol
```

# The rest of the corpus

Elsewhere in the corpus nobody's dose or genotype is on record, so there is no
known answer to test. What is left is to ask whether the ratio in blood moves
with anything the deposits record. A unit is one subject in one matrix, so a
study that sampled each person many times does not outweigh one that sampled
once. Where no subject is named, each file is its own unit, and repeat files of
one person then count as separate units. A field of a subject sampled several
times, such as a time point, belongs to the unit only where all its files agree.

```{r rest-of-corpus}
# Files a metadata source covers, outside both accessions of the dosed study.
DOSED <- c(PLASMA_DS, SKIN_DS)
rest  <- pf[pf$has_metadata & !pf$dataset_id %in% DOSED, ]
rest$fk    <- norm_file_key(rest$file)
rest$study <- rest$dataset_id
rest$subj  <- nz(rest$subject_id)
rest$site  <- canon_site(rest$body_site)
# log2(omeprazole / 5-OH-omeprazole) where both were detected; a truncated peak
# has no area, so no ratio.
rest$l2 <- ifelse(rest$par_detected %in% TRUE & rest$met_detected %in% TRUE,
                  log2(rest$par_auc / rest$met_auc), NA_real_)
# The scanned fields: a placeholder is missing, and sex and matrix are written
# one way across deposits.
stopifnot(all(names(SCAN_FIELDS) %in% names(rest)))
for (f in names(SCAN_FIELDS))
  rest[[f]] <- ifelse(is_missing_val(rest[[f]]), NA_character_, trimws(as.character(rest[[f]])))
rest$sex       <- harmonise_sex(rest$sex)
rest$body_site <- rest$site

u     <- unitise(rest, fields = names(SCAN_FIELDS))
# Units whose files disagree on a scanned field, which they then lack.
u_disagree <- disagree_text(u, SCAN_FIELDS)
blood <- u[is.finite(u$l2) & u$site %in% MATRICES, ]

u |>
  group_by(assay = short_ds(study, datasets$dataset_id)) |>
  summarise(units = n(), either_target = sum(either),
            with_ratio = sum(is.finite(l2)),
            blood_with_ratio = sum(is.finite(l2) & site %in% MATRICES),
            blood_single_files = sum(is.finite(l2) & site %in% MATRICES & kind == "file"),
            .groups = "drop")

# Blood units that are single files with no subject named, per assay.
fs <- blood$study[blood$kind == "file"]
file_units <- table(factor(fs, sort(unique(fs), method = "radix")))
```

`r if (nzchar(u_disagree)) paste0("The files of some units disagree on a field, which these units then lack: ", u_disagree, ".") else "No unit's files disagree on any of the fields scanned below."`
Outside the dosed study, `r n_distinct(u$study)` assays have files with
metadata, `r nrow(u)` units in all. Omeprazole or 5-OH-omeprazole is detected in
`r sum(u$either)` of these units, and both, which gives a ratio, in
`r sum(is.finite(u$l2))`; `r nrow(blood)` of those are blood, from
`r n_distinct(blood$study)` assays.
`r if (length(file_units)) paste0("Of these blood units, ", sum(file_units), if (sum(file_units) == 1) " is a single file" else " are single files", " with no subject named (", and_list(sprintf("%s: %d", short_ds(names(file_units), datasets$dataset_id), file_units)), "), so several may come from one person.")`

## Every field against the ratio

The scan takes every metadata field of the per-file table
(`r and_list(SCAN_FIELDS)`) and, inside each assay where the field takes more
than one value among the blood units with a ratio, tests the ratio against it.
The model, in words: inside one assay, a unit's log2 ratio is the assay's level,
plus the effect of the field (a slope where the field is a number, such as age;
a shift for each value otherwise), plus what the model leaves unexplained. As a
formula, for unit $i$ of assay $s$ and field $f$, fitted by least squares:

$$
\log_2\left(\frac{\text{omeprazole}}{\text{5-OH-omeprazole}}\right)_i
= a_s + f_i + \varepsilon_i
$$

Each fit gives one p value, from the F test of the field (for a number, the same
as the test of its slope). Testing inside one assay keeps each assay's method
from being read as an effect. All p values are then corrected together with
Benjamini-Hochberg, and a contrast counts as significant where its corrected p
is below `r FDR`. A field that repeats another inside an assay, as group and health status
often do, is tested once there. A field constant inside every assay but
different between assays is confounded with the assay: its effect cannot be told
apart from the method's, so it cannot be tested.

```{r scan}
blood_ids <- sort(unique(blood$study), method = "radix")
values_in <- function(f, s) { v <- blood[[f]][blood$study == s]; unique(v[!is.na(v)]) }

# Per field: the assays where it varies, and whether it is confounded with the
# assay (one value inside every assay, not the same value in all of them).
fields <- do.call(rbind, lapply(names(SCAN_FIELDS), function(f) {
  n_in <- vapply(blood_ids, function(s) length(values_in(f, s)), 0L)
  data.frame(field = SCAN_FIELDS[[f]], column = f,
             varies_in = sum(n_in >= 2),
             confounded = all(n_in <= 1) && n_distinct(blood[[f]], na.rm = TRUE) >= 2)
}))

# One test per field and assay where the field varies. A field whose values
# split the assay's units exactly as an earlier field's do is the same test.
scan <- list(); repeats <- character(0)
for (s in blood_ids) {
  z0   <- blood[blood$study == s, ]
  seen <- character(0)
  for (f in fields$column[fields$varies_in > 0]) {
    if (length(values_in(f, s)) < 2) next
    split_key <- paste(match(z0[[f]], unique(z0[[f]])), collapse = " ")
    if (split_key %in% seen) {
      repeats <- c(repeats, sprintf("%s repeats %s in %s", SCAN_FIELDS[[f]],
                                    SCAN_FIELDS[[names(seen)[seen == split_key]]],
                                    short_ds(s, datasets$dataset_id)))
      next
    }
    seen[f] <- split_key
    z <- z0[!is.na(z0[[f]]), ]
    x <- if (!anyNA(as_num(z[[f]]))) as_num(z[[f]]) else factor(z[[f]])
    m <- lm(z$l2 ~ x)
    scan[[length(scan) + 1]] <- data.frame(
      field = SCAN_FIELDS[[f]], assay = short_ds(s, datasets$dataset_id),
      units = nrow(z),
      compared = if (is.numeric(x)) sprintf("slope over %g to %g", min(x), max(x)) else
        paste(sprintf("%s (%d)", levels(x), tabulate(x)), collapse = ", "),
      p = if (df.residual(m) > 0) anova(m)[["Pr(>F)"]][1] else NA_real_)
  }
}
scan <- bind_rows(scan)
# A contrast with no unit left over for the error has no test.
untestable <- scan[is.na(scan$p), ]
scan       <- scan[!is.na(scan$p), ]
scan$q     <- p.adjust(scan$p, "BH")
write.csv(scan, "artifacts/ratio_scan.csv", row.names = FALSE)
knitr::kable(scan[order(scan$q, scan$p), ], row.names = FALSE, digits = 3,
             col.names = c("field", "assay", "units", "compared", "p",
                           "p after Benjamini-Hochberg"),
             caption = "The log2 ratio against each field, inside each assay where the field varies")

tested     <- fields$field[fields$varies_in > 0]
confounded <- fields$field[fields$confounded]
left       <- fields$field[fields$varies_in == 0 & !fields$confounded]
n_sig      <- sum(scan$q < FDR)
```

Of the `r nrow(fields)` fields, `r length(tested)` vary inside at least one
assay (`r and_list(tested)`), which gives `r nrow(scan)` contrasts inside assays.
`r if (length(repeats)) paste0("In the scan, ", and_list(repeats), if (length(repeats) == 1) ", so the pair is tested once." else ", so each pair is tested once.")`
`r if (nrow(untestable)) paste0("Too few units leave no test for ", and_list(sprintf("%s in %s (%d units)", untestable$field, untestable$assay, untestable$units)), ".")`
After the correction over all `r nrow(scan)`, `r n_sig`
`r if (n_sig == 1) "survives" else "survive"` at a false discovery rate of
`r 100 * FDR`% (smallest corrected p = `r fmt_p(min(scan$q))`).
`r length(confounded)` `r if (length(confounded) == 1) "field is" else "fields are"`
confounded with the assay`r if (length(confounded)) paste0(" (", and_list(confounded), ")")`.
`r if (length(left)) paste0("The other ", length(left), " (", and_list(left), ") ", if (length(left) == 1) "has" else "have", " no value, or a single value, across the blood units with a ratio.")`
With no dose time on record, the ratio also carries when each person last took
the drug, so these numbers describe the corpus; they do not test a known answer.

# Step summary

The results of this step, as the chapter's table shows them.

```{r step-summary}
summary_step <- data.frame(
  item  = c("assays set aside (same samples, another set-up)",
            "volunteers",
            sprintf("plasma draws at %s min with a detection", and_list(EARLY_MIN)),
            "skin samples before the first dose with a detection",
            "change in log2 omeprazole/5-OH-omeprazole per activity class (95% CI)",
            "p (R2)",
            "slope with each volunteer left out",
            "other assays with a ratio in blood",
            "blood units with a ratio",
            "contrasts tested",
            sprintf("significant at %g%% FDR", 100 * FDR),
            "fields confounded with the assay"),
  value = c(sum(!choice$kept), nrow(subj_auc),
            sprintf("%d of %d", neg$detected[1], neg$n[1]),
            sprintf("%d of %d", neg$detected[2], neg$n[2]),
            sprintf("%+.3f (%.3f to %.3f)", coef(m_auc)[["rank"]], ci[1], ci[2]),
            sprintf("%s (%.2f)", fmt_p(p_auc), summary(m_auc)$r.squared),
            sprintf("%+.3f to %+.3f", min(jk$beta), max(jk$beta)),
            n_distinct(blood$study), nrow(blood), nrow(scan),
            sprintf("%d of %d", n_sig, nrow(scan)),
            if (length(confounded)) sprintf("%d (%s)", length(confounded), and_list(confounded))
            else "0"))
write.csv(summary_step, "artifacts/summary_step6.csv", row.names = FALSE)
summary_step
```

# Checks on repeated sampling

These checks are not part of the chapter's results, and the step summary above
does not use them. Each volunteer in the model of the dosed study gives one
value: the ratio of the areas over all draws, averaged over the two study days.
The checks fit the same model on other values from the same draws: the first
study day alone, and the ratio in a single draw.

```{r checks-setup}
# The model of the dosed study on one value per volunteer (a log2 ratio, with the
# class rank): the volunteers, the slope per class with its 95% CI, p and R2.
class_fit <- function(d, ratio, days) {
  m  <- lm(log2_ratio ~ rank, data = d)
  ci <- confint(m)["rank", ]
  data.frame(ratio = ratio, days = days, volunteers = nrow(d),
             slope = coef(m)[["rank"]], lo = ci[[1]], hi = ci[[2]],
             p = summary(m)$coefficients["rank", 4], r2 = summary(m)$r.squared)
}
# One value per volunteer: the mean over the study days that give one.
per_volunteer <- function(d) aggregate(log2_ratio ~ subject + phenotype + rank, data = d, FUN = mean)
check_cols <- c("ratio", "study days", "volunteers", "slope per class",
                "95% CI low", "95% CI high", "p", "R2")
say_fit <- function(r) sprintf("%+.3f per class (95%% CI %.3f to %.3f, p = %s, R² = %.2f)",
                               r$slope, r$lo, r$hi, fmt_p(r$p), r$r2)
main_fit <- class_fit(subj_auc, "areas over all draws", "mean of days")
```

## First study day only

```{r check-day1}
# The study day before the antibiotics, as the submitter table names it.
day1 <- unique(ta$ATTRIBUTE_Study_Day[ta$ATTRIBUTE_Study_Day_Text == "pre-antibiotic PK"])
stopifnot(length(day1) == 1)
auc_day1  <- auc[auc$day == day1, ]
check_day <- rbind(main_fit, class_fit(auc_day1, "areas over all draws", paste("day", day1)))
knitr::kable(transform(check_day, p = fmt_p(p)), row.names = FALSE, digits = 3,
             col.names = check_cols,
             caption = "The ratio against class: both study days and the first alone")
no_day1 <- setdiff(subj_auc$subject, auc_day1$subject)
```

The submitter table names study day `r day1` as the one before the antibiotics
("pre-antibiotic PK"). On that day alone, over `r nrow(auc_day1)` volunteers,
the log2 ratio of the two areas changes by `r say_fit(check_day[2, ])`, against
`r say_fit(main_fit)` over the mean of both days.
`r if (check_day$hi[2] < 0) paste0("The interval on day ", day1, " alone also lies below zero, the predicted fall.") else paste0("On day ", day1, " alone, the interval includes zero.")`
`r if (length(no_day1)) paste0(and_list(no_day1), if (length(no_day1) == 1) " has" else " have", " no area ratio on day ", day1, " and is left out.")`

## A single draw

```{r check-single-draw}
# The usual index reads the ratio in one draw, about 3 hours after the dose.
SINGLE_MIN <- 180
draw_times <- sort(unique(dr$tmin))
nearest    <- draw_times[abs(draw_times - SINGLE_MIN) == min(abs(draw_times - SINGLE_MIN))]

# The ratio of the two targets in one draw at time t, per volunteer and day. A
# draw where a target is not detected gives no ratio.
at_draw <- function(t) {
  d <- dr[dr$tmin == t, ]
  data.frame(subject = d$ATTRIBUTE_Subject, day = d$ATTRIBUTE_Study_Day,
             phenotype = d$PharmGeno_Phenotype_2C19,
             rank = match(d$PharmGeno_Phenotype_2C19, PHENO),
             log2_ratio = log2(d$par_auc / d$met_auc))
}
draws <- lapply(setNames(nearest, nearest), at_draw)
check_draw <- rbind(main_fit, do.call(rbind, lapply(names(draws), function(t) {
  d <- draws[[t]]
  d <- d[is.finite(d$log2_ratio), ]
  label <- sprintf("single draw at %s min", t)
  rbind(class_fit(d[d$day == day1, ], label, paste("day", day1)),
        class_fit(per_volunteer(d), label, "mean of days"))
})))
knitr::kable(transform(check_draw, p = fmt_p(p)), row.names = FALSE, digits = 3,
             col.names = check_cols,
             caption = "The ratio against class: the areas over all draws and a single draw")
# Draws (volunteer and day) with no ratio at each of the nearest times.
no_ratio <- vapply(draws, function(d) sum(!is.finite(d$log2_ratio)), 0L)
# The rows of one draw time in the table.
draw_rows <- function(t) check_draw[check_draw$ratio == sprintf("single draw at %s min", t), ]
say_draw  <- function(t) {
  r <- draw_rows(t)
  paste0("At ", t, " min, the log2 ratio changes by ", say_fit(r[1, ]), " on day ", day1,
         " (", r$volunteers[1], " volunteers), and by ", say_fit(r[2, ]),
         " over the mean of the days (", r$volunteers[2], " volunteers).")
}
```

The usual index of CYP2C19 activity is the ratio in one draw about
`r SINGLE_MIN` min after the dose. This study drew at `r and_list(draw_times)`
min. `r if (SINGLE_MIN %in% draw_times) paste0("The index is read at ", SINGLE_MIN, " min.") else if (length(nearest) > 1) paste0("No draw falls at ", SINGLE_MIN, " min; the nearest are ", and_list(nearest), " min, equally far from it, so the index is read at each.") else paste0("No draw falls at ", SINGLE_MIN, " min; the index is read at the nearest, ", nearest, " min.")`
The index is the log2 ratio of the two targets' areas in that draw, per volunteer
and day; a draw injected more than once gives the mean of its files, as above. A
draw where a target is not detected gives no ratio and is left out:
`r and_list(sprintf("%d at %s min", no_ratio, names(no_ratio)))`.

`r paste(vapply(names(draws), say_draw, ""), collapse = " ")` Over the areas of
all draws and both days, the slope is `r say_fit(main_fit)`. Of the
`r nrow(check_draw) - 1` single-draw fits, `r sum(check_draw$hi[-1] < 0)` have an
interval below zero, the predicted fall.

# Check across assays

This check is not part of the chapter's results either. The scan above tests
each field inside each assay. Sex, age and matrix are recorded the same way in
every deposit, so they can also be tested across assays: one model over the
assays where the field varies, with each assay's own level ($a_s$ above) and one
effect of the field shared by all of them. The same model without the assay's
level is shown too; it reads each method's offset as part of the field's effect.

```{r check-across}
# The fields recorded the same way in every deposit. Group, health status and the
# others name each deposit's own categories, so they have no shared effect.
SHARED_FIELDS <- c("sex", "age", "body_site")
varies_in <- function(f) blood_ids[vapply(blood_ids, function(s) length(values_in(f, s)) >= 2, TRUE)]
across <- do.call(rbind, lapply(SHARED_FIELDS, function(f) {
  vs <- varies_in(f)
  if (length(vs) < 2) return(NULL)
  z <- blood[blood$study %in% vs & !is.na(blood[[f]]), ]
  x <- if (!anyNA(as_num(z[[f]]))) as_num(z[[f]]) else factor(z[[f]])
  data.frame(field = SCAN_FIELDS[[f]], n_assays = length(vs),
             assays = and_list(short_ds(vs, datasets$dataset_id)),
             units = nrow(z),
             p_with = anova(lm(z$l2 ~ factor(z$study) + x))["x", "Pr(>F)"],
             p_without = anova(lm(z$l2 ~ x))["x", "Pr(>F)"])
}))
knitr::kable(transform(across[, names(across) != "n_assays"],
                       p_with = fmt_p(p_with), p_without = fmt_p(p_without)),
             row.names = FALSE,
             col.names = c("field", "assays", "units", "p, with each assay's level",
                           "p, without"),
             caption = "The log2 ratio against each shared field, across the assays where it varies")
# Shared fields that vary inside a single assay: their test is the one in the scan.
one_assay <- SCAN_FIELDS[SHARED_FIELDS[vapply(SHARED_FIELDS, function(f) length(varies_in(f)) == 1, TRUE)]]
n_below   <- sum(c(across$p_with, across$p_without) < 0.05)
cap       <- function(x) paste0(toupper(substring(x, 1, 1)), substring(x, 2))
```

`r paste(sprintf("Across the %d assays where %s varies (%d units), p = %s with each assay's level and %s without.", across$n_assays, across$field, across$units, fmt_p(across$p_with), fmt_p(across$p_without)), collapse = " ")`
`r if (n_below == 0) "None is below 0.05: the added assays make no field significant across assays either." else paste(n_below, "of these p values are below 0.05.")`
`r if (length(one_assay)) paste0(cap(and_list(unname(one_assay))), if (length(one_assay) == 1) " varies" else " vary", " inside a single assay only, so ", if (length(one_assay) == 1) "its test is" else "their tests are", " the one in the scan above.")`

# Check: the area baseline

This check is not part of the chapter's results, and the step summary above
does not use it. Step 5 measures a peak's area above a straight line between its
two boundary scans. Where the scan just outside a boundary is empty, the signal
rises from nothing there, and that line leaves out the part of the peak below
the boundary scan. Step 5 therefore also records the area with the line started
from zero on such a side (`auc0`). Here the model of the dosed study is fitted
again with each file's areas taken from `auc0`. The draws, the areas over the
draw times and the volunteers are built as above.

```{r check-baseline}
# The two baselines, as the table names them.
BASELINES <- c("between the boundary scans", "from zero beside an empty scan")
# A per-file table written before step 5 recorded auc0 has no such columns.
has_auc0 <- all(c("par_auc0", "met_auc0") %in% names(p84))
if (has_auc0) {
  # Each file's areas on auc0, zero where the target was not detected, as above.
  ta0 <- ta
  j0  <- match(ta0$k, p84$k)
  ta0$par_auc <- ifelse(ta0$par_det, p84$par_auc0[j0], 0)
  ta0$met_auc <- ifelse(ta0$met_det, p84$met_auc0[j0], 0)
  # Extracted draw files with a detection, and those whose areas the line from zero changes.
  n_det   <- sum(ta0$extracted & (ta0$par_det | ta0$met_det))
  n_moved <- sum(ta0$extracted & (ta0$par_auc != ta$par_auc | ta0$met_auc != ta$met_auc))
  subj0 <- per_volunteer(day_ratios(draw_areas(ta0)))
  # auc0 is missing exactly where the area is, so the volunteers are those of the model.
  stopifnot(setequal(subj0$subject, subj_auc$subject))
  # The model's fit and its slope without each volunteer in turn.
  with_loo <- function(d, label) {
    f   <- class_fit(d, label, "mean of days")
    loo <- leave_one_out(d)
    cbind(f[, names(f) != "days"], loo_lo = min(loo$beta), loo_hi = max(loo$beta))
  }
  by_base <- rbind(with_loo(subj_auc, BASELINES[1]), with_loo(subj0, BASELINES[2]))
  say_loo <- function(r) sprintf("%+.3f and %+.3f", r$loo_lo, r$loo_hi)
  baseline_txt <- paste0(
    "The line from zero changes the areas of ", n_moved, " of the ", n_det,
    " extracted draw files with a detection. With it, over ", by_base$volunteers[2],
    " volunteers, the log2 ratio changes by ", say_fit(by_base[2, ]),
    ", against ", say_fit(by_base[1, ]), " with the line between the boundary scans",
    "; without any one volunteer the slope stays between ", say_loo(by_base[2, ]),
    " (", say_loo(by_base[1, ]), " with the line between the boundary scans).")
} else {
  baseline_txt <- paste("The per-file table has no area on the second baseline (no par_auc0 or",
                        "met_auc0 column): it was written before step 5 recorded one, so the",
                        "check waits for step 5 to be rendered again.")
}
if (has_auc0)
  knitr::kable(transform(by_base, p = fmt_p(p)), row.names = FALSE, digits = 3,
               col.names = c("area baseline", "volunteers", "slope per class",
                             "95% CI low", "95% CI high", "p", "R2",
                             "slope without one volunteer, lowest", "highest"),
               caption = "The ratio against class on the two area baselines")
```

`r baseline_txt`

# Session info

```{r session-info}
sessionInfo()
```
