# Checks that can catch us being wrong

A pipeline is only as trustworthy as the things that could contradict it, and
plausible numbers are not evidence if nothing in the design would fail when they
are wrong. The checks below are built in on purpose, in increasing order of how
hard they are to argue with.

**1. The confirmed-file positive control.** Where MS2 has already proved the
compound is present, MS1 extraction must find it; if it does not, propagation is
losing real signal and every prevalence number is an underestimate.
`run_ms1_for_dataset()` reports this as `confirmed files loaded: N | with >=1
peak: M`, where M should equal N and any shortfall is a bug rather than biology.
It is the cheapest hard check here and needs no external data.

**2. A known-negative control.** The complement: run the identical pipeline on
files that *cannot* contain the compound and confirm it finds nothing. The
pre-dose plasma draws of the probe study are exactly that (real matrix, true
m/z, true retention window, drug absent by design), and `06-analysis` reports them. A
synthetic decoy mass is deliberately not used: an untargeted file carries real
compounds at arbitrary masses, so a hit at a decoy mass is usually the picker
correctly finding something else, which bounds the false-positive rate from
above without estimating it.

**3. Report the surface, not the best cell.** Every parameter here is a choice,
and a conclusion that survives only one setting is not a conclusion. The
retention-time sweep below is the pattern to follow: a broad plateau where the
answer is stable, and a cliff with a mechanism that explains exactly where and
why it breaks. A single lucky optimum would not be evidence.

**4. External ground truth, which beats every internal check.** Parameters can
be tuned until an internal check passes; an outside number cannot be. `06-analysis` uses
the one deposit where truth exists independently of us: a cohort in which every
subject was deliberately dosed, with pre-dose samples that are known negatives
and a CYP2C19 genotype per volunteer that the pipeline never sees.

**5. The apex-snap shown rather than asserted.** The retention time a confirmed
hit carries is when the instrument fired, not when the compound eluted, and the
whole window design rests on the second being stable where the first is not.
The dosed cohort shows it best, so it is plotted: in MSV000084008, the files with
the most confirmed metabolite scans, each scan on the EIC of the file it fired in.

```{r fig-dda-offset, fig.width = 9, fig.height = 5.6}
#| echo: false
# The dosed cohort's plasma, and the files with the most confirmed metabolite scans:
# one panel each.
.ds    <- "MSV000084008"
.n_fig <- 15
.r <- results[[.ds]]
.h <- conf_final[conf_final$dataset_id == .ds & conf_final$role == "metabolite", ]
.h$bn <- basename(.h$source_file)
.top <- names(sort(table(norm_file_key(.h$bn)), decreasing = TRUE))[seq_len(min(.n_fig, length(unique(norm_file_key(.h$bn)))))]
.h   <- .h[norm_file_key(.h$bn) %in% .top, ]
.org <- basename(chartr(rawToChar(as.raw(92)), "/",
                        chromData(.r$chr_met)$dataOrigin))
.ix  <- which(norm_file_key(.org) %in% .top)
.lab <- function(x) sub("^MSV[0-9]+_", "", sub("\\.mzX?ML$", "", x))
.win <- .r$rt_anchor$met$win

.tr <- dplyr::bind_rows(lapply(.ix, function(i) data.frame(
  file = .lab(.org[i]),
  rt = rtime(.r$chr_met[i])[[1]], int = intensity(.r$chr_met[i])[[1]])))
.tr <- .tr[is.finite(.tr$int), ]
# The apex inside step 5's window: elsewhere in the trace sits the sulfone.
.ap <- .tr[.tr$rt >= .win[1] & .tr$rt <= .win[2], ] |> dplyr::group_by(file) |>
  dplyr::slice_max(int, n = 1, with_ties = FALSE) |> dplyr::ungroup()
.ms2 <- .h |> dplyr::transmute(file = .lab(bn), ms2 = rt)

ggplot(.tr, aes(rt, int)) +
  geom_line(colour = "grey35", linewidth = 0.4) +
  geom_vline(data = .ms2, aes(xintercept = ms2), colour = "#D6604D",
             linewidth = 0.4, alpha = 0.85) +
  geom_point(data = .ap, aes(rt, int), colour = "#2166AC", size = 1.8) +
  facet_wrap(~ file, scales = "free_y", ncol = 5) +
  coord_cartesian(xlim = .win + c(-25, 25)) +   # zoom, not a filter: nothing is dropped
  labs(x = "Retention time (s)", y = "Intensity",
       title = paste(.ds, "metabolite: MS2 trigger times (red) against the MS1 apex (blue)"),
       subtitle = paste("one panel per file, the", length(.top),
                        "files with the most confirmed metabolite scans")) +
  theme_bw(base_size = 10) +
  theme(strip.background = element_rect(fill = "grey92"),
        panel.grid.minor = element_blank())
```

```{r dda-offset-numbers}
#| echo: false
.j <- dplyr::left_join(.ms2, dplyr::select(.ap, file, apex = rt), by = "file")
cat("files with a confirmed metabolite scan:", nrow(.ap),
    "| confirmed scans:", nrow(.j), "\n")
cat("apex span:", sprintf("%.1f-%.1f s (%.1f s)", min(.ap$rt), max(.ap$rt),
                          diff(range(.ap$rt))), "\n")
cat("trigger span:", sprintf("%.1f-%.1f s (%.1f s)", min(.j$ms2), max(.j$ms2),
                             diff(range(.j$ms2))), "\n")
cat("trigger minus apex:", sprintf("%+.1f to %+.1f s", min(.j$ms2 - .j$apex),
                                   max(.j$ms2 - .j$apex)), "\n")
```

```{r dda-offset-example}
#| echo: false
# The file holding the largest trigger-to-apex offset, for the sentence below:
# how many confirmed scans it contributes and how far apart they fired.
.big   <- .j$file[which.max(.j$ms2 - .j$apex)]
.big_n <- sum(.j$file == .big)
.big_s <- round(diff(range(.j$ms2[.j$file == .big])))
```

The apexes sit in a band a fraction of the width of the triggers, and the
largest offsets are not scattered files but single intense peaks re-fragmented
on the way down: the file with the largest offset contributes `r .big_n`
confirmed scans, spread over `r .big_s` seconds, all belonging to one apex.
Centring a window on the trigger times would
force it to span the union of those tails; centring on the apexes does not. This
is what the window derivation is for, and the figure is the check that it is
warranted.

**6. Which of the two isobars is actually in the parent channel.** The window
argument above says the compounds are chromatographically resolved; this asks
the harder question of whether the peak we integrate is the right molecule.
Omeprazole is a sulfoxide, so it cannot lose $\cdot$SH; hydroxyomeprazole
sulfide, which shares its formula, gives m/z $313.1422$ on exactly that loss.
The share of confirmed parent MS2 carrying that ion therefore reads directly on
how much of the channel is the isobar, and the probe cohort supplies its own
control: the cluster where omeprazole is confirmed should be near zero.

```{r isobar-fragment}
#| echo: false
.ifc <- "artifacts/isobar_fragment_check.csv"
# Reads raw MS2, not the cached EICs, so it is slow and its result is cached. Delete
# the CSV to recompute; a change in the confirmed set is not noticed on its own.
if (!file.exists(.ifc)) {
  SULFIDE_MZ <- 313.1422   # [M+H-SH]+, impossible for a sulfoxide
  SHARED_MZ  <- 198.0581   # substituted pyridine ring, given by both
  TOL <- 0.02; PREC_TOL <- 0.05
  .pc <- function(x) if (!length(x)) NA_real_ else round(100 * mean(x))
  .rows <- list()
  for (ds in names(results)) {
    r <- results[[ds]]
    # only files carrying a confirmed parent hit: reading the whole corpus
    # costs hours and the question is what the confirmed spectra are
    want <- unique(basename(as.character(conf_final$source_file[
      conf_final$dataset_id == ds & conf_final$role == "parent"])))
    if (!length(want)) next
    fs <- unique(chartr(rawToChar(as.raw(92)), "/", chromData(r$chr_par)$dataOrigin))
    fs <- fs[basename(fs) %in% want & file.exists(fs)]
    if (!length(fs)) next
    sp <- tryCatch(filterMsLevel(Spectra(fs), 2L), error = function(e) NULL)
    if (is.null(sp) || !length(sp)) next
    pmz <- precursorMz(sp)
    k <- which(!is.na(pmz) & abs(pmz - compound_spec$mz_pos[compound_spec$role == "parent"]) < PREC_TOL)
    if (!length(k)) next
    sp <- sp[k]; pk <- peaksData(sp); rt <- rtime(sp)
    hasfrag <- function(t) vapply(pk, function(m) any(abs(m[, 1] - t) < TOL), logical(1))
    s313 <- hasfrag(SULFIDE_MZ)
    # the same two centres, and the same stand-down condition, resolve_isobars() uses
    ctr <- function(role) {
      q <- as.numeric(conf_final$rt[conf_final$dataset_id == ds & conf_final$role == role])
      q <- q[is.finite(q)]; if (length(q)) stats::median(q) else NA_real_
    }
    cp <- ctr("parent"); cm <- ctr("metabolite")
    ok <- !isTRUE(r$rt_anchor$par$uncertain) && !isTRUE(r$rt_anchor$met$uncertain) &&
      is.finite(cp) && is.finite(cm) && abs(cp - cm) >= CHAN_SEP
    own <- if (ok) abs(rt - cp) < abs(rt - cm) else rep(NA, length(rt))
    .rows[[length(.rows) + 1]] <- data.frame(
      dataset_id = ds, spectra = length(pk),
      pct_sulfide_313 = .pc(s313), pct_shared_198 = .pc(hasfrag(SHARED_MZ)),
      resolved = ok,
      n_integrated = if (ok) sum(own) else NA_integer_,
      pct_sulfide_integrated = if (ok) .pc(s313[own]) else NA_real_,
      n_rejected = if (ok) sum(!own) else NA_integer_,
      pct_sulfide_rejected = if (ok) .pc(s313[!own]) else NA_real_,
      stringsAsFactors = FALSE)
  }
  write.csv(do.call(rbind, .rows), .ifc, row.names = FALSE)
}
.z <- if (file.exists(.ifc)) read.csv(.ifc, stringsAsFactors = FALSE)
# One cell out of the table, so no number in the prose below is typed twice.
.iso <- function(ds, col) if (is.null(.z)) "--" else .z[[col]][.z$dataset_id == ds]
{
  .z$verdict <- ifelse(.z$pct_sulfide_313 >= 40, "largely the isobar",
                ifelse(.z$pct_sulfide_313 >= 15, "mixed", "mostly omeprazole"))
  .show <- .z[order(-.z$pct_sulfide_313),
              c("dataset_id", "spectra", "pct_sulfide_313",
                "pct_sulfide_integrated", "pct_sulfide_rejected", "verdict")]
  knitr::kable(.show, row.names = FALSE,
    col.names = c("assay", "MS2 at the parent mass", "% with 313.14 (-SH)",
                  "integrated cluster", "rejected cluster", "verdict"),
    caption = paste("Sulfide-diagnostic fragment among confirmed parent spectra.",
                    "The last two columns split that rate by retention cluster,",
                    "classified by the same comparison resolve_isobars() makes;",
                    "they are empty where the anchors are too close for the",
                    "rule to apply."))
}
```

```{r isobar-separation}
#| echo: false
# How far apart a deposit's parent and metabolite anchors sit: the two medians of
# the confirmed retention times that resolve_isobars() compares.
.sep <- function(ds) {
  m <- function(role) stats::median(as.numeric(
    conf_final$rt[conf_final$dataset_id == ds & conf_final$role == role]), na.rm = TRUE)
  abs(m("parent") - m("metabolite"))
}
```

Read the rates as a mixture, not a verdict on a whole deposit. The probe cohort
sits at `r .iso("MSV000084008", "pct_sulfide_313")` per cent pooled, which on its
own says little; split by retention cluster it is
`r .iso("MSV000084008", "pct_sulfide_integrated")` per cent in the cluster the
extraction integrates against
`r .iso("MSV000084008", "pct_sulfide_rejected")` per cent in the one it rejects.
That contrast is the whole case for `resolve_isobars()`: the rule is not
discarding signal, it is discarding a different molecule. MSV000088255, a urine
cohort, does not reach this step: step 4 sets it aside, because there the peak at
omeprazole's mass is in most files the isomer itself (`04-confirmation`). One
deposit is worth naming, and the rule stands down in it, which is why its row has no
split. MTBLS1866 runs at
`r .iso("MTBLS1866_LC-MS_positive_reverse-phase_metabolite_profiling", "pct_sulfide_313")`
per cent and is **not** gated, because its two anchors sit
`r sprintf("%.1f", .sep("MTBLS1866_LC-MS_positive_reverse-phase_metabolite_profiling"))` s
apart, under the `r CHAN_SEP` s the rule needs to tell the channels apart. Its
parent contrast should be read with that attached.

A methodological warning that applies to all of this: **do not choose a
parameter by maximising agreement with the thing you then cite as validation.**
That is circular.
Where a parameter must be tuned, tune it against something structurally
independent of the validation target. The parent compound is useful here
precisely because it has no isobaric twin, so its retention-time drift can be
measured without reference to any downstream result.

**7. What each choice is actually worth.** The warning above is a rule for
choosing parameters; this is the audit of the ones already chosen. Every
decision in the extraction is defensible in prose, which is exactly the problem:
a reader has no way to tell a defensible choice from a convenient one. So each
is turned off in turn, one at a time with everything else left alone, and the
genotype model is refitted from the peaks that survive. MSV000084008 is the only
deposit carrying genotypes, so it is the only place the coefficient can be
recomputed; MSV000082493 supplies the counter-check, its skin samples taken
before the first dose being the one known-negative the intensity floor is
answerable to.

```{r ablation-helpers}
#| echo: false
# ---- ablations: re-run the extraction with one choice changed ---------------
# Re-run role_metrics() from the cached EICs; with defaults they must match the cache.

ablation_setup <- function(ds, bio_files) {
  if (!identical(compound_spec$role, c("parent", "metabolite")))
    stop("ablation_setup(", ds, "): the ablations need exactly two targets, ",
         "the parent and the metabolite")
  r  <- readRDS(sprintf("artifacts/per_dataset/%s.rds", ds))
  fb <- basename(as.character(bio_files))
  ch <- list(par = r$chr_par_narrow, met = r$chr_met_narrow)
  list(ds = ds, cfg = r$cfg, ch = ch,
       metrics = list(par = r$metrics_par, met = r$metrics_met),
       tr = lapply(ch, function(c0) lapply(Chromatograms::peaksData(c0),
              function(pd) list(rt = pd[, "rtime"], it = pd[, "intensity"]))),
       fl = lapply(ch, function(c0) basename(dataOrigin(c0))),
       bare = function(x) norm_file_key(to_bare_name(basename(as.character(x)), fb)))
}

# The consensus the pipeline does not use: every confirmed hit votes, so a file
# fragmented fifteen times outweighs one fragmented five.
ablation_window_per_hit <- function(e, role, conf, old_win) {
  hh <- conf[conf$dataset_id == e$ds & conf$role == role_of_prefix(role), ]
  a  <- suppressWarnings(as.numeric(hh$rt)); a <- a[is.finite(a)]
  if (!length(a)) return(old_win)
  nbrs <- vapply(a, function(z) sum(abs(a - z) <= CLUST_TOL), integer(1))
  top  <- a[nbrs == max(nbrs)]
  inl  <- a[abs(a - stats::median(top)) <= CLUST_TOL]
  w <- c(min(inl) - APEX_PAD, max(inl) + APEX_PAD)
  attr(w, "uncertain") <- diff(range(top)) > CLUST_TOL
  w
}

# Stops where role_metrics() stops, before the isobar rule; ablation_isobars()
# is that step, kept separate so the baseline can be checked against the cache.
ablation_variant <- function(e, conf, floor_val = MIN_INT, per_hit = FALSE,
                             ov, drop_all, accept = curation_inputs()$accept) {
  anchor <- list(); mets <- list()
  for (role in c("par", "met")) {
    old_win <- e$cfg[[paste0("rt_", role, "_win")]]
    win <- if (per_hit) ablation_window_per_hit(e, role, conf, old_win) else
      acceptance_window(e$ds, role_of_prefix(role),
                        list(files = e$fl[[role]], traces = e$tr[[role]]),
                        e$bare, old_win, conf, ov, FALSE)
    anchor[[role]] <- list(win = as.numeric(win),
                           uncertain = isTRUE(attr(win, "uncertain")))
    picks <- lapply(e$tr[[role]], function(t) pick(t$rt, t$it, win[1], win[2], FALSE))
    acc   <- accept[accept$dataset == e$ds & accept$role == role_of_prefix(role), , drop = FALSE]
    hand  <- match(norm_file_key(e$fl[[role]]), norm_file_key(acc$file))
    for (i in which(!is.na(hand)))
      picks[[i]] <- pick(e$tr[[role]][[i]]$rt, e$tr[[role]][[i]]$it,
                         acc$apex_rt[hand[i]] - APEX_PAD, acc$apex_rt[hand[i]] + APEX_PAD, FALSE)
    picks <- add_msquality(e$ch[[role]], picks)
    keep  <- which(!vapply(picks, is.null, logical(1)))
    pk <- picks[keep]; fl <- e$fl[[role]][keep]
    old_floor <- MIN_INT; MIN_INT <<- floor_val
    det <- vapply(seq_along(pk), function(i) accept_peak(pk[[i]], FALSE) &&
                    !is_dropped(e$ds, role_of_prefix(role), fl[i], drop_all), logical(1))
    MIN_INT <<- old_floor
    mets[[role]] <- stats::setNames(
      data.frame(file = fl,
                 auc = vapply(pk, function(p) p$auc, numeric(1)),
                 apex_rt = vapply(pk, function(p) p$apex_rt, numeric(1)),
                 detected = det, stringsAsFactors = FALSE),
      c("file", paste0(role, c("_auc", "_apex_rt", "_detected"))))
  }
  list(ds = e$ds, win = anchor, par = mets$par, met = mets$met)
}

ablation_isobars <- function(v, conf) {
  mr <- resolve_isobars(v[c("par", "met")], conf, v$ds, v$win)
  modifyList(v, list(par = as.data.frame(mr$par), met = as.data.frame(mr$met)))
}
```

```{r ablations}
#| echo: false
rt_ov   <- read.csv("curation/rt_apex_overrides.csv", stringsAsFactors = FALSE)
drop_all <- read.csv("curation/ms1_drop_list.csv", stringsAsFactors = FALSE)
hits_curated <- read.csv("artifacts/hits_curated.csv", check.names = FALSE,
                         stringsAsFactors = FALSE)

E84 <- ablation_setup("MSV000084008", bio_files_per_ds[["MSV000084008"]])
E82 <- ablation_setup("MSV000082493", bio_files_per_ds[["MSV000082493"]])
vary <- function(e, conf = conf_final, ...)
  ablation_variant(e, conf = conf, ov = rt_ov, drop_all = drop_all, ...)

# The genotype refit: one trapezoidal AUC per subject per day, averaged over
# days, regressed on the phenotype rank. The same path 06-analysis takes.
ta84 <- read.delim("dataset_metadata/MSV000084008_metadata.tsv",
                   check.names = FALSE, stringsAsFactors = FALSE)
ta84$k <- norm_file_key(ta84$filename)
ta84$tmin <- as_num(ta84$ATTRIBUTE_Time_Point_Mins)
ta84 <- ta84[ta84$MS_SampleType == "sample" & !is.na(ta84$tmin), ]

trap <- function(x, y) {
  o <- order(x); x <- x[o]; y <- y[o]; y[is.na(y)] <- 0
  if (length(unique(x)) < 4) NA_real_ else
    sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)
}

refit <- function(v) {
  d <- ta84
  mp <- v$par; mm <- v$met
  mp$k <- norm_file_key(mp$file); mm$k <- norm_file_key(mm$file)
  i <- match(d$k, mp$k); j <- match(d$k, mm$k)
  d$par_auc <- ifelse(mp$par_detected[i] %in% TRUE, mp$par_auc[i], NA_real_)
  d$met_auc <- ifelse(mm$met_detected[j] %in% TRUE, mm$met_auc[j], NA_real_)
  auc <- do.call(rbind, lapply(
    split(d, list(d$ATTRIBUTE_Subject, d$ATTRIBUTE_Study_Day), drop = TRUE),
    function(z) data.frame(
      subject = z$ATTRIBUTE_Subject[1],
      rank = match(z$PharmGeno_Phenotype_2C19[1], PHENO),
      log2_ratio = log2(trap(z$tmin, z$par_auc) / trap(z$tmin, z$met_auc)))))
  auc <- auc[is.finite(auc$log2_ratio) & !is.na(auc$rank), ]
  s <- aggregate(log2_ratio ~ subject + rank, data = auc, FUN = mean)
  co <- summary(lm(log2_ratio ~ rank, data = s))$coefficients
  c(beta = co[2, 1], se = co[2, 2], p = co[2, 4], n = nrow(s))
}

# Skin sampled before the first dose: the drug had not been taken and skin has
# no route to it, so anything called here is a false positive.
sk <- read.delim("dataset_metadata/MSV000082493_metadata.tsv",
                 check.names = FALSE, stringsAsFactors = FALSE)
sk <- sk[sk$MS_SampleType == "sample" & sk$ATTRIBUTE_Body_Site == "L.forehead" &
           (sk$ATTRIBUTE_Study_Day %in% "0" |
            (sk$ATTRIBUTE_Study_Day %in% "1" &
             sk$ATTRIBUTE_Time_Point_Mins %in% "0")), ]
sk$k <- norm_file_key(sk$filename)
skin_fp <- function(v) {
  mp <- v$par; mp$k <- norm_file_key(mp$file)
  sum(mp$par_detected[match(sk$k, mp$k)] %in% TRUE)
}

# The second known negative: the dosed cohort's pre-dose and five-minute plasma
# draws, taken before absorption. A detection of either compound is a false positive.
pre84 <- ta84[ta84$tmin %in% c(0, 5), ]
predose_fp <- function(v) {
  mp <- v$par; mm <- v$met
  mp$k <- norm_file_key(mp$file); mm$k <- norm_file_key(mm$file)
  sum((mp$par_detected[match(pre84$k, mp$k)] %in% TRUE) |
      (mm$met_detected[match(pre84$k, mm$k)] %in% TRUE))
}

rows <- list()
add <- function(choice, alternative, note, v84, v82) {
  b <- refit(v84)
  rows[[length(rows) + 1]] <<- data.frame(
    choice = choice, alternative = alternative,
    detections = as.integer(sum(v84$par$par_detected) + sum(v84$met$met_detected)),
    note = note, skin_fp = as.integer(skin_fp(v82)),
    predose_fp = as.integer(predose_fp(v84)),
    beta = round(unname(b["beta"]), 3), se = round(unname(b["se"]), 3),
    p = signif(unname(b["p"]), 2), subjects = as.integer(b["n"]),
    stringsAsFactors = FALSE)
}

R84 <- vary(E84); R82 <- vary(E82)
stopifnot(identical(R84$par$par_detected, E84$metrics$par$par_detected),
          identical(R84$met$met_detected, E84$metrics$met$met_detected),
          identical(R82$par$par_detected, E82$metrics$par$par_detected))
V0 <- ablation_isobars(R84, conf_final); W0 <- ablation_isobars(R82, conf_final)
# Pre-dose denominator: the draws extracted, not those the submitter table declares
# nor the rows a variant's metrics carry (only where the picker found a peak).
n_pre <- sum(pre84$k %in%
             norm_file_key(basename(as.character(bio_files_per_ds[["MSV000084008"]]))))
add("pipeline as reported", "--", "", V0, W0)

for (f in c(1, 100, 250, 1000))
  add(sprintf("apex floor %d counts", MIN_INT), sprintf("floor %d counts", f), "",
      ablation_isobars(vary(E84, floor_val = f), conf_final),
      ablation_isobars(vary(E82, floor_val = f), conf_final))

for (tiers in list("strict", c("strict", "standard"))) {
  cf <- conf_final[conf_final$confidence %in% tiers, ]
  add("anchors at any confidence tier", sprintf("%s only", paste(tiers, collapse = "+")),
      sprintf("%d of %d anchors", nrow(cf), nrow(conf_final)),
      ablation_isobars(vary(E84, conf = cf), cf),
      ablation_isobars(vary(E82, conf = cf), cf))
}

.vp <- ablation_isobars(vary(E84, per_hit = TRUE), conf_final)
add("one RT anchor per file", "one anchor per confirmed hit",
    sprintf("parent window %.0f-%.0f s, not %.0f-%.0f",
            .vp$win$par$win[1], .vp$win$par$win[2],
            V0$win$par$win[1], V0$win$par$win[2]),
    .vp, ablation_isobars(vary(E82, per_hit = TRUE), conf_final))

add("reject cross-channel isobars", "keep every in-window peak", "", R84, R82)

# Adduct policy acts upstream of the window, on which MASST hits exist at all,
# so it has no separate slope: this deposit's own hits are protonated either way.
.ok <- hits_curated$adduct_inferred %in% "[M+H]+"
rows[[length(rows) + 1]] <- data.frame(
  choice = "any plausible adduct", alternative = "protonated only",
  detections = NA_integer_,
  note = sprintf("%d of %d curated hits, %d of %d entities", sum(.ok),
                 nrow(hits_curated), length(unique(hits_curated$dataset_id[.ok])),
                 length(unique(hits_curated$dataset_id))),
  skin_fp = NA_integer_, predose_fp = NA_integer_,
  beta = NA_real_, se = NA_real_, p = NA_real_,
  subjects = NA_integer_, stringsAsFactors = FALSE)

.a <- do.call(rbind, rows)

# Pull one cell out for the prose below, so no number here is typed twice.
.ab <- function(alt, col) .a[[col]][.a$alternative == alt]
.bmax <- sprintf("%.3f", max(abs(.a$beta - .a$beta[1]), na.rm = TRUE))
.pmax <- sprintf("%.5f", max(.a$p, na.rm = TRUE))
.se0  <- sprintf("%.3f", .a$se[1])
.nsk  <- nrow(sk)

.tbl <- .a
# p is formatted here rather than left to kable: `digits` would round every one
# of them to 0.001 and the column would say nothing.
.tbl$p <- ifelse(is.na(.tbl$p), "", sprintf("%.5f", .tbl$p))
knitr::kable(.tbl, row.names = FALSE, digits = 3,
  col.names = c("choice", "alternative", "detections", "note",
                "pre-dose skin", "pre-dose plasma", "slope", "SE", "p", "subjects"),
  caption = paste("Each row changes one decision and refits. Detections are",
                  "MSV000084008's; pre-dose skin counts false positives among",
                  .nsk, "known negatives in MSV000082493, and pre-dose plasma",
                  "among the", n_pre, "draws taken before absorption in",
                  "MSV000084008."))
```

```{r ablation-subjects}
#| echo: false
# Subjects in the refit: the most any variant keeps, and whether every variant
# keeps that many.
.nsub     <- max(.a$subjects, na.rm = TRUE)
.nsub_all <- all(.a$subjects == .nsub, na.rm = TRUE)
```

The first row is the pipeline as reported, and it reproduces the cached
detection flags exactly: the chunk asserts that before it computes anything
else, so a variant that differs does so because of the change and not because
the re-run drifted from the pipeline. Across every other row the slope stays
within `r .bmax` of it, against a standard error on that slope of `r .se0`; the
p-value never rises above `r .pmax`, and `r if (.nsub_all) "all" else "not all"`
`r .nsub` subjects survive every variant. None of these choices carries the
validation: the correlation is not an artefact of the
processing, because the processing can be varied and it does not go away.

The intensity floor is the one row that is not symmetric. Raising it costs
plasma detections and buys nothing, while lowering it admits files that cannot
contain the drug at all: at $100$ counts it accepts
`r .ab("floor 100 counts", "skin_fp")` of the `r .nsk` pre-dose skin samples and
`r .ab("floor 100 counts", "predose_fp")` of the `r n_pre` pre-dose plasma
draws, and a floor of $1$ accepts exactly the same files, so nothing sits
between the two. At $250$ both controls are empty again, which is where the
sweep stops being free. That asymmetry, not a tuned optimum, is why the floor
sits where it does, and the per-matrix breakdown says where its cost falls.

```{r floor-by-matrix}
#| echo: false
mb <- read.delim("dataset_metadata/MSV000082493_metadata.tsv",
                 check.names = FALSE, stringsAsFactors = FALSE)
mb <- mb[mb$MS_SampleType == "sample", ]
mb$k <- norm_file_key(mb$filename)
.mp <- R82$par; .mp$k <- norm_file_key(.mp$file)
.lo <- vary(E82, floor_val = 100)$par
.fb <- do.call(rbind, lapply(split(mb, mb$ATTRIBUTE_Body_Site), function(g) {
  i <- match(g$k, .mp$k); i <- i[!is.na(i)]
  data.frame(body_site = g$ATTRIBUTE_Body_Site[1], files = nrow(g),
             with_peak_in_window = length(i),
             accepted_at_500 = sum(.mp$par_detected[i]),
             accepted_at_100 = sum(.lo$par_detected[i]),
             stringsAsFactors = FALSE)
}))
.fb$rejected_at_500 <- .fb$with_peak_in_window - .fb$accepted_at_500
.fb$floor_only <- .fb$accepted_at_100 - .fb$accepted_at_500
.fm <- function(site, col) .fb[[col]][.fb$body_site == site]
knitr::kable(.fb, row.names = FALSE,
  col.names = c("body site", "files", "peak in window", "accepted at 500",
                "accepted at 100", "rejected at 500", "floor-only"),
  caption = paste("MSV000082493, parent channel, before the isobar rule.",
                  "Floor-only rejections are those a lower threshold would",
                  "rescue; the rest fail another criterion and cannot be."))
```

The gate is a conjunction, so a peak failing any criterion other than the apex
height cannot be rescued by a lower threshold, and the split matters. In faeces
only `r .fm("fecal", "floor_only")` of `r .fm("fecal", "rejected_at_500")`
rejections are the floor's doing: most of that matrix's shortfall is peaks that
do not have the shape of a peak, and lowering the threshold would not recover
them. In skin the proportions invert, `r .fm("L.forehead", "floor_only")` of
`r .fm("L.forehead", "rejected_at_500")`, so the same change admits that many
detections in the one matrix with no plausible route of exposure. The prevalence
figures downstream are lower bounds, most conspicuously in faeces and on skin;
no known-negative exists for those matrices to justify a lower floor, and the
one that does exist argues against it.

# The retention-time window and the sulfone

Each assay's window is set on the MS1 apexes of its confirmed spectra and padded
by `APEX_PAD` = `r APEX_PAD` s on each side (*The retention-time window* above).
How wide that pad may be is set by the metabolite's isobar.

*Why the isomer matters.* 5-hydroxyomeprazole is not the only thing at its mass.
Omeprazole sulfone, the compound CYP3A4 makes from omeprazole, has the identical
formula (`C17H19N3O4S`) and therefore the identical m/z (362.1169). MS1 cannot
tell the two apart. Only chromatography separates them, and the consequences of
losing that separation are severe: the sulfone reports CYP3A4 activity, not the
CYP2C19 activity we are trying to measure, and in a poor CYP2C19 metaboliser the
drug shunts to CYP3A4, so the sulfone is *larger* than the real 5-OH peak. The
proximity-weighted selection described above is the primary defence; the window
is a second, softer one, kept modest so the sulfone rarely even enters it.

```{r sulfone-margin}
#| echo: false
SULFONE_RT  <- 181   # s, sulfone RT in MSV000084008, read off its EICs
SWEEP_PAR_W <- 5     # s, parent half-window held fixed in the pad sweep below
# The metabolite's window in MSV000084008, as this step set it (ms1_windows above).
.w5 <- unlist(ms1_windows[ms1_windows$dataset_id == "MSV000084008" &
                          ms1_windows$role == "metabolite", c("rt_lo", "rt_hi")])
```

The margin is narrower than it looks. In MSV000084008 the metabolite's window,
the MS1 apexes of its confirmed spectra padded by `r APEX_PAD` s, is
[`r round(.w5[1])`, `r round(.w5[2])`] s and is safe, since the sulfone sits at
~`r SULFONE_RT` s. But its upper edge is only `r round(SULFONE_RT - .w5[2])` s
short of the sulfone, so a pad that much wider reaches it.

*This is measured, not assumed.* That deposit's metadata records each subject's
genotyped CYP2C19 phenotype (`PharmGeno_Phenotype_2C19`), which the pipeline
never uses during extraction, so it is an independent target. Re-integrating the
same cached chromatograms of MSV000084008 at a range of metabolite
acceptance-window half-widths (parent fixed at ±`r SWEEP_PAR_W` s; metabolite
centre ~`r round(mean(.w5))` s, sulfone ~`r SULFONE_RT` s) and correlating the
subject-level AUC ratio against that genotyped phenotype gives:

```{r}
#| echo: false
#| output: asis
#| label: tbl-pad-sweep
# From the cached wide EICs of MSV000084008: widen the metabolite window toward the
# sulfone, re-pick with and without proximity weighting, correlate with genotype.
sweep_cache <- "peaks_review_v2/_cache/MSV000084008.rds"
sweep_meta  <- "dataset_metadata/MSV000084008_metadata.tsv"
if (!"MSV000084008" %in% names(bio_files_per_ds)) {
  cat("_Pad sweep not applicable: it needs the genotyped probe cohort ",
      "MSV000084008, which this application does not contain._\n", sep = "")
} else if (file.exists(sweep_cache) && file.exists(sweep_meta)) {
  # PHENO (CYP2C19 activity order) comes from R/masst.R.
  cache <- readRDS(sweep_cache)
  ta <- read.delim(sweep_meta, check.names = FALSE, stringsAsFactors = FALSE)
  ta$tmin <- as.numeric(ta$ATTRIBUTE_Time_Point_Mins)
  ta <- ta[ta$MS_SampleType == "sample" & !is.na(ta$tmin), ]
  ta$k <- norm_file_key(ta$filename)
  fb   <- basename(as.character(bio_files_per_ds[["MSV000084008"]]))
  bare <- function(x) norm_file_key(to_bare_name(basename(as.character(x)), fb))
  ov_sweep <- if (file.exists("curation/rt_apex_overrides.csv"))
    read.csv("curation/rt_apex_overrides.csv", stringsAsFactors = FALSE) else
    data.frame(dataset_id = character(), role = character(),
               ms2_rt = numeric(), apex_rt = numeric())
  ctr  <- function(role, cr) mean(acceptance_window("MSV000084008", role, cr, bare,
                                                    cr$win, conf_final, ov_sweep))
  par_ctr <- ctr("parent", cache$par); met_ctr <- ctr("metabolite", cache$met)
  # hilic = TRUE gives pure area-select, so it doubles as the "no proximity"
  # control.
  auc_of <- function(cr, centre, W, proximity) {
    data.frame(k = norm_file_key(cr$files),
               auc = vapply(cr$traces, function(tr) {
                 p <- pick(tr$rt, tr$it, centre - W, centre + W, hilic = !proximity)
                 if (is.null(p) || !accept_peak(p, FALSE)) NA_real_ else p$auc
               }, numeric(1)))
  }
  trap <- function(x, y) { o <- order(x); x <- x[o]; y <- y[o]; y[is.na(y)] <- 0
    if (length(unique(x)) < 4) NA_real_ else sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2) }
  rho_of <- function(pa, ma) {
    t2 <- ta
    t2$par_auc <- pa$auc[match(t2$k, pa$k)]; t2$met_auc <- ma$auc[match(t2$k, ma$k)]
    a <- do.call(rbind, lapply(
      split(t2, list(t2$ATTRIBUTE_Subject, t2$ATTRIBUTE_Study_Day), drop = TRUE),
      function(d) data.frame(rank = match(d$PharmGeno_Phenotype_2C19[1], PHENO),
                             lr = log2(trap(d$tmin, d$par_auc) / trap(d$tmin, d$met_auc)),
                             subject = d$ATTRIBUTE_Subject[1])))
    a <- a[is.finite(a$lr), ]; s <- aggregate(lr ~ subject + rank, a, mean)
    cor.test(s$rank, s$lr, method = "spearman", exact = FALSE)$estimate
  }
  pa <- auc_of(cache$par, par_ctr, SWEEP_PAR_W, TRUE)   # parent fixed throughout
  sweep <- do.call(rbind, lapply(c(5, 10, 15, 20, 30), function(W) data.frame(
    `metabolite window` = sprintf("[%.0f, %.0f] (+/-%d s)", met_ctr - W, met_ctr + W, W),
    `reaches sulfone?`  = if ((met_ctr + W) >= SULFONE_RT) "yes" else "no",
    `with proximity`    = sprintf("%+.3f", rho_of(pa, auc_of(cache$met, met_ctr, W, TRUE))),
    `without proximity` = sprintf("%+.3f", rho_of(pa, auc_of(cache$met, met_ctr, W, FALSE))),
    check.names = FALSE)))
  cat(knitr::kable(sweep, align = "lccc", row.names = FALSE), sep = "\n")
} else {
  cat("_Pad sweep skipped: needs `", sweep_cache, "` (git-ignored). Run ",
      "`Rscript R/peak_review_panels.R omeprazole MSV000084008` to rebuild it._\n", sep = "")
}
```

With the proximity weighting the correlation is unchanged at every width;
without it, the ratio collapses once the window admits the sulfone. So the
isomer defence is the on-RT selection, not a tight window; the ±`r APEX_PAD` s
pad only buys drift tolerance.
